---
title: "Object Creation for Lab Comparison Analysis"
format:
  html:
    embed-resources: true
---

# Introduction

This document creates all the data objects needed for the lab comparison analysis.
These objects are saved to the `object/` directory and can be loaded by other
analysis scripts.

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(here)

base_path <- here("5_downstream_analysis_manual")
source(file.path(base_path, "helpers.R"))
```

```{r cores-cluster}
# cluster parallel setting
bpparam <- MulticoreParam(workers = parallel::detectCores() - 1)
```

# Full Data Objects

## Load Full Data

Note: 

- the data has been filtered to include only the m/z range of 50 - 1100 Da
  So that the raw signal is more comparable. !!!

```{r load-full-data}
if (!file.exists(file.path(base_path, "object/sp_full.RData"))) {
  sp_afekta <- load_data(lab = "afekta", study_group = "HE")
  sp_hmgu <- load_data(lab = "hmgu", study_group = "HE")
  sp_icl <- load_data(lab = "icl", study_group = "HE")
  sp_cembio <- load_data(lab = "cembio", study_group = "HE")
  sp_full <- c(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
  sp_full <- filterMzRange(sp_full, mz = c(50, 1100))
  sp_full <- setBackend(sp_full, MsBackendMemory())
  sp_full <- applyProcessing(sp_full)
  save(sp_full, file = file.path(base_path, "object/sp_full.RData"))
  rm(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
} else {
  load(file.path(base_path, "object/sp_full.RData"))
}
```

## Create Base Peak Chromatogram (BPC) - Full Data

```{r load-bpc-full}
if(!file.exists(file.path(base_path, "object/bpc_full.RData"))) {
  bpc_full <- Chromatograms(sp_full, summarize.method = "max",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab"))
  bpc_full <- setBackend(bpc_full, ChromBackendMemory())
  save(file = file.path(base_path, "object/bpc_full.RData"), bpc_full)
} else {
  load(file.path(base_path, "object/bpc_full.RData"))
}
```

## Create Total Ion Current (TIC) - Full Data

```{r load-tic-full}
if (!file.exists(file.path(base_path, "object/tic_full.RData"))) {
  tic_full <- Chromatograms(sp_full, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab"))
  tic_full <- setBackend(tic_full, ChromBackendMemory())
  save(file = file.path(base_path, "object/tic_full.RData"), tic_full)
} else {
  load(file.path(base_path, "object/tic_full.RData"))
}
```

# Detected Signal Objects

## Load Detected Signal Data

```{r load-detect}
if (!file.exists(file.path(base_path, "object/sp_full_detect.RData"))) {
  bpparam <- MulticoreParam(8) # for the cluster
  sp_afekta <- detect_signal(
    lab = "afekta",
    study_group = "HE",
    annotated = FALSE,
    bpparam = bpparam
  )
  save(sp_afekta, file = file.path(base_path, "object/sp_afekta.RData"))
  rm(sp_afekta)

  sp_hmgu <- detect_signal(
    lab = "hmgu",
    study_group = "HE",
    annotated = FALSE,
    bpparam = bpparam
  )
  save(sp_hmgu, file = file.path(base_path, "object/sp_hmgu.RData"))
  rm(sp_hmgu)

  sp_icl <- detect_signal(
    lab = "icl",
    study_group = "HE",
    annotated = FALSE,
    bpparam = bpparam
  )
  save(sp_icl, file = file.path(base_path, "object/sp_icl.RData"))
  rm(sp_icl)

  sp_cembio <- detect_signal(
    lab = "cembio",
    study_group = "HE",
    annotated = FALSE,
    bpparam = bpparam
  )
  save(sp_cembio, file = file.path(base_path, "object/sp_cembio.RData"))

  load(file.path(base_path, "object/sp_hmgu.RData"))
  load(file.path(base_path, "object/sp_afekta.RData"))
  load(file.path(base_path, "object/sp_icl.RData"))

  sp_full_detect <- c(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
  sp_full_detect <- filterMzRange(sp_full_detect, mz = c(100, 1100))
  sp_full_detect <- setBackend(sp_full_detect, MsBackendMemory())
  sp_full_detect <- applyProcessing(sp_full_detect)
  save(
    sp_full_detect,
    file = file.path(base_path, "object/sp_full_detect.RData")
  )
  rm(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
} else {
  load(file.path(base_path, "object/sp_full_detect.RData"))
}
```

## Create Base Peak Chromatogram (BPC) - Detected Signal

```{r load-bpc-detect}
if (!file.exists(file.path(base_path, "object/bpc_detect.RData"))) {
  bpc_detect <- Chromatograms(sp_full_detect, summarize.method = "max",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("lab", "mixture"))
  bpc_detect <- setBackend(bpc_detect, ChromBackendMemory())
  save(file = file.path(base_path, "object/bpc_detect.RData"), bpc_detect)
} else {
  load(file.path(base_path, "object/bpc_detect.RData"))
}
```

## Create Total Ion Current (TIC) - Detected Signal

```{r load-tic-detect}
if (!file.exists(file.path(base_path, "object/tic_detect.RData"))) {
  tic_detect <- Chromatograms(sp_full_detect, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab"))
  tic_detect <- setBackend(tic_detect, ChromBackendMemory())
  save(file = file.path(base_path, "object/tic_detect.RData"), tic_detect)
} else {
  load(file.path(base_path, "object/tic_detect.RData"))
}
```


# Annotated Signal

The annotated level no longer materialises in this file. The producer
`3_annotation_manual/HE/3_Sirius_curation/sirius_annotation_all_ms2.qmd`
writes `5_downstream_analysis/object/annotation_chrom_metrics.csv` directly; the
bulk annotated TIC (P4) is `sum(areaUnderTic)` per `(lab, mixture,
polarity)` over that table, computed inline in
`7_reproducibility_results.qmd`.

# Summary

Public objects (consumed by `7_reproducibility_results.qmd`,
`4_consensus_peaks.qmd`):

- `sp_full.RData` — full spectra, all labs
- `bpc_full.RData`, `tic_full.RData` — BPC / TIC, full signal
- `sp_full_detect.RData` — combined detected signal
- `bpc_detect.RData`, `tic_detect.RData` — BPC / TIC, detected signal

Internal resume checkpoints (per-lab; only loaded inside this file to
recover from a partial detect_signal() build):
`sp_afekta.RData`, `sp_hmgu.RData`, `sp_icl.RData`, `sp_cembio.RData`.

The annotated level lives in the SIRIUS producer's
`5_downstream_analysis/object/annotation_chrom_metrics.csv` — consumers derive bulk
and per-compound annotated views from that CSV. Stale RData files from
prior runs (`sp_full_ann`, `bpc_ann`, `tic_ann`, `eics_ann`,
`tic_full_spectra`, `eics_detected`) can be deleted from `object/`.
