---
title: "Peak Feature Calculation — Detected Peaks"
format:
  html:
    toc: true
    embed-resources: true
execute:
  dir: project
---

```{r}
library(BiocParallel)
bp <- SnowParam(workers = 4)
register(bp)
```

# Introduction

This document builds the cross-lab **consensus peak table** from the
per-lab xcms-detected extracted-ion chromatograms (EICs) and runs the
downstream reproducibility comparisons on it.

The approach:

- Start from **all** detected peaks across the four labs (no annotation
  filter); the chromatogram metadata records which peaks are
  SIRIUS-annotated.
- Match peaks across labs (m/z + NAPS-aligned RT) and keep the
  cross-lab-overlapping features to form a common consensus feature set.

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)
library(UpSetR)
library(MsQuality)
library(igraph)

base_path <- here::here("5_downstream_analysis_manual")
source(file.path(base_path, "helpers.R"))
source(here::here("1_preprocessing/setup.R"))
```


# Create Consensus Peak Table Across Labs

```{r}
pt_afekta <- read.csv(file.path(
  base_path,
  "object/detected_peaks_afekta_HE.csv"
))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(
  base_path,
  "object/detected_peaks_cembio_HE.csv"
))
lab_counts <- c(
  afekta = nrow(pt_afekta),
  hmgu = nrow(pt_hmgu),
  icl = nrow(pt_icl),
  cembio = nrow(pt_cembio)
)
barplot(
  lab_counts,
  main = "Number of Detected Peaks per Lab",
  ylab = "Number of detected peaks",
  xlab = "Labs",
  col = "steelblue"
)
```

## Method

Correspondence is computed per sample (one mixture × polarity), on
**all detected peaks** from the four labs (no annotation filter). For
each sample:

1. Peaks from the four labs are pooled and sorted by m/z.
2. A graph is built where peaks are nodes and an edge connects two peaks
   whenever |Δmz| ≤ `PPM_TOL` × m/z × 10⁻⁶ **and** |Δrt| ≤ `RT_TOL` s.
   The edge list is constructed by a two-pointer sweep over the
   m/z-sorted peaks.
3. Connected components of the graph are the consensus peaks; every
   peak in the same component shares a candidate match with at least one
   other peak in the component.
4. When a single lab contributes more than one peak to a component, the
   peak closest to the centroid (anchored on labs that contribute
   exactly one peak) is kept, where distance is mz+rt distance
   normalised by `PPM_TOL` and `RT_TOL`.

Only m/z and RT are used to define correspondence — no FWHM or other
shape-based information.

```{r build-consensus-peaks}
PPM_TOL <- 20    # m/z tolerance (parts per million)
RT_TOL  <- 15    # RT tolerance (seconds)
LABS    <- c("afekta", "hmgu", "icl", "cembio")

# Pool peaks from the four labs, tag with lab id, drop NAs in intb.
pt_all <- rbind(
  transform(pt_afekta, lab = "afekta"),
  transform(pt_hmgu,   lab = "hmgu"),
  transform(pt_icl,    lab = "icl"),
  transform(pt_cembio, lab = "cembio")
)
pt_all <- pt_all[!is.na(pt_all$intb), ]

# Apply per-lab NAPS-spline alignment (from 3_rt_alignment.qmd) BEFORE
# the cross-lab matching, then fall back to raw rt for peaks outside the
# NAPS anchor range. Peaks stay in the matching pool either way.
# Aligned-fallback recovers ~34 % more 4-lab consensus features vs raw
# matching (2598 -> 3490; see the comparison table below) by both
# tightening intra-group spread (6.35 -> 3.39 s median) and keeping
# out-of-range peaks instead of dropping them.
load(file.path(base_path, "object/rt_alignment_splines.RData"))
anchors_for_range <- read.csv(
  file.path(base_path, "object/rt_alignment_anchors.csv"),
  stringsAsFactors = FALSE)

apply_splines <- function(df, splines_map, anchors_df, rt_col = "rt") {
  df$key        <- paste(df$lab, df$polarity, sep = "_")
  df$aligned_rt <- NA_real_
  for (key_i in unique(df$key)) {
    f <- splines_map[[key_i]]
    if (is.null(f)) next
    anchor_range <- anchors_df |>
      dplyr::filter(paste(lab, polarity, sep = "_") == key_i) |>
      dplyr::pull(rt_local) |> range(na.rm = TRUE)
    idx       <- which(df$key == key_i)
    rt_lab    <- df[[rt_col]][idx]
    in_range  <- !is.na(rt_lab) &
                 rt_lab >= anchor_range[1] & rt_lab <= anchor_range[2]
    df$aligned_rt[idx[in_range]] <- as.numeric(f(rt_lab[in_range]))
  }
  df$key <- NULL
  # Fallback: peaks outside NAPS range keep their raw rt for matching.
  df$match_rt <- ifelse(is.na(df$aligned_rt), df[[rt_col]], df$aligned_rt)
  df
}
pt_all <- apply_splines(pt_all, splines, anchors_for_range, rt_col = "rt")
cat(sprintf("Peaks aligned via NAPS: %d / %d (%.1f%%). Remainder use raw rt.\n",
            sum(!is.na(pt_all$aligned_rt)), nrow(pt_all),
            100 * mean(!is.na(pt_all$aligned_rt))))

# Per-sample mz+rt graph grouping (two-pointer sweep + connected components).
# rt_col controls which RT column to match on:
#   "match_rt" (default) — NAPS-aligned with raw fallback (main pipeline)
#   "rt"                 — raw rt (used by the alignment-effect comparison)
group_sample <- function(pt_all, sample_id,
                         ppm = PPM_TOL, rt_tol = RT_TOL,
                         rt_col = "match_rt") {
  sub <- pt_all[pt_all$sample == sample_id, ]
  sub <- sub[order(sub$mz), ]
  sub$peak_idx <- seq_len(nrow(sub))

  n      <- nrow(sub)
  mz_vec <- sub$mz
  rt_vec <- sub[[rt_col]]

  cap  <- 1024L # what is this cap ?
  from <- integer(cap); to <- integer(cap); k <- 0L

  for (i in seq_len(n - 1)) {
    tol_i <- mz_vec[i] * ppm * 1e-6
    j     <- i + 1L
    while (j <= n && mz_vec[j] - mz_vec[i] <= tol_i) {
      if (abs(rt_vec[j] - rt_vec[i]) <= rt_tol) {
        k <- k + 1L
        if (k > cap) {
          cap <- cap * 2L
          length(from) <- cap; length(to) <- cap
        }
        from[k] <- i; to[k] <- j
      }
      j <- j + 1L
    }
  }
  edges <- data.frame(from = from[seq_len(k)], to = to[seq_len(k)])

  g <- graph_from_data_frame(edges, directed = FALSE,
                             vertices = data.frame(name = sub$peak_idx))
  sub$consensus_id <- components(g)$membership
  sub
}

# Cache filename includes "aligned" so the legacy raw-matching cache
# (correspondence_grouped.rds) doesn't shadow the new aligned-fallback
# output.
corr_cache <- file.path(base_path, "object/correspondence_grouped_aligned.rds")
if (file.exists(corr_cache)) {
  grouped <- readRDS(corr_cache)
} else {
  samples <- sort(unique(pt_all$sample))
  grouped <- do.call(rbind, lapply(samples, function(s) {
    g <- group_sample(pt_all, s)
    g$consensus_id <- paste(s, g$consensus_id, sep = "_")
    g
  }))
  grouped$consensus_id <- as.integer(factor(grouped$consensus_id))
  saveRDS(grouped, corr_cache)
}
cat("Pooled peaks:", nrow(grouped),
    "in", length(unique(grouped$consensus_id)), "consensus groups\n")
```

```{r disambiguate-and-build-ct}
# Disambiguate: when one lab contributes >1 peak to a group, pick the peak
# closest to the centroid defined by the anchor labs (those contributing
# exactly one peak). Disambiguation runs on the same axis as matching
# (match_rt = aligned where possible, raw fallback otherwise).
#
# data.table rewrite: dplyr's group_by(consensus_id) |> mutate(if any())
# over ~810k consensus groups was taking >1 h due to per-group dispatch
# overhead. data.table's by= operations are vectorized in compiled C
# and run in seconds.
suppressPackageStartupMessages(library(data.table))
gdt <- as.data.table(grouped)
gdt[, n_per_lab := .N, by = .(consensus_id, lab)]
gdt[, anchor := n_per_lab == 1L]
# Per-consensus reference (mz, match_rt): median over anchor rows if any,
# else median over all rows.
gdt[, ref_mz := if (any(anchor)) median(mz[anchor]) else median(mz),
    by = consensus_id]
gdt[, ref_rt := if (any(anchor)) median(match_rt[anchor]) else median(match_rt),
    by = consensus_id]
gdt[, score := abs(mz - ref_mz) / (ref_mz * PPM_TOL * 1e-6) +
                abs(match_rt - ref_rt) / RT_TOL]
# Per (consensus_id, lab), keep the row with the smallest score.
disambig <- gdt[gdt[, .I[which.min(score)], by = .(consensus_id, lab)]$V1]
disambig[, c("n_per_lab", "anchor", "ref_mz", "ref_rt",
             "score", "peak_idx") := NULL]
disambig <- as.data.frame(disambig)
rm(gdt); gc(full = TRUE, verbose = FALSE)

# Pivot to wide consensus-table format expected downstream:
# one row per (sample, consensus_id), one column per lab.
ct_all_samples <- disambig |>
  select(sample, consensus_id, lab, chrom_peak_id) |>
  pivot_wider(
    id_cols     = c(sample, consensus_id),
    names_from  = lab,
    values_from = chrom_peak_id,
    names_prefix = "chrom_peak_id_"
  ) |>
  mutate(
    idx_afekta = chrom_peak_id_afekta,
    idx_hmgu   = chrom_peak_id_hmgu,
    idx_icl    = chrom_peak_id_icl,
    idx_cembio = chrom_peak_id_cembio,
    n_labs_matched = (!is.na(chrom_peak_id_afekta)) +
                    (!is.na(chrom_peak_id_hmgu))   +
                    (!is.na(chrom_peak_id_icl))    +
                    (!is.na(chrom_peak_id_cembio))
  ) |>
  as.data.frame()

cat("Consensus peaks (any lab count):", nrow(ct_all_samples), "\n")
cat("Consensus peaks present in all 4 labs:",
    sum(ct_all_samples$n_labs_matched == 4), "\n")
```


```{r}
# Add detection flags to consensus_matches
ct_all_samples <- ct_all_samples |>
  mutate(
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  )

cat("Peak counts by number of labs detected:\n")
print(table(ct_all_samples$n_labs_matched))
```

```{r}
# Replace mixture column from 1.1 to 1_1
meta$mixture <- gsub("\\.", "_", meta$Mixture)
nc_mix <- table(meta$mixture)
# This shows the number of compounds per mixture.

# Create mapping of sample to mixture name using pt_* data
sample_to_mixture <- pt_afekta |>
  select(sample, mixture) |>
  distinct()

# Below we can summarize the number of peaks detected per sample with compound count
mixture_summary <- lapply(split(ct_all_samples, ct_all_samples$sample), function(df) {
  sample_id <- unique(df$sample)
  mixture_name <- sample_to_mixture$mixture[sample_to_mixture$sample == sample_id]

  res <- data.frame(
    sample = sample_id,
    mixture = mixture_name,
    detected_4labs = sum(df$n_labs_matched == 4),
    detected_3labs = sum(df$n_labs_matched == 3),
    detected_2labs = sum(df$n_labs_matched == 2),
    detected_1lab = sum(df$n_labs_matched == 1),
    total_peaks = nrow(df),
    n_compounds = as.integer(nc_mix[mixture_name])
  )
  res
})

mixture_summary_df <- do.call(rbind, mixture_summary)
rownames(mixture_summary_df) <- NULL

knitr::kable(mixture_summary_df, caption = "Peak Detection Summary by Mixture")
```


## Cross-reference with SIRIUS-annotated panel

Sanity-check the consensus matching against ground truth: how many of
the SIRIUS-confirmed spike-in compounds (every annotated row with a
reported `mz_observed` in `annotation_identity.csv`) land in a
consensus feature? And how many of the cross-lab consensus features
correspond to a known compound vs are unidentified background
chemistry? (We deliberately do not filter on `confidence_tier` — that
is set by `6_confidence_tier.qmd`, which renders after this file.)

We match each annotated `(mixture, lab, compound, polarity)` row
against the consensus feature at the same `sample` (m/z within 20 ppm,
`match_rt` within 15 s of the consensus group's lab-mean centroid).

```{r annotated-to-consensus}
#| message: false
producer_dir <- here::here("5_downstream_analysis_manual", "object")
ann_chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                      stringsAsFactors = FALSE,
                      colClasses = c(mixture = "character"))
ann_id    <- read.csv(file.path(producer_dir, "annotation_identity.csv"),
                      stringsAsFactors = FALSE,
                      colClasses = c(mixture = "character"))
# Normalize mixture IDs to underscore form (defensive).
ann_chrom$mixture <- gsub("\\.", "_", ann_chrom$mixture)
ann_id$mixture    <- gsub("\\.", "_", ann_id$mixture)

# Ground-truth set: every SIRIUS-annotated row with a reported
# mz_observed. We do NOT filter on confidence_tier here — that column is
# populated downstream by 6_confidence_tier.qmd, which renders AFTER this
# QMD, so at this point it is still the "unfiltered" placeholder.
# Filtering on A/B/C here would drop every row (the bug that produced
# all-zero cross-reference tables).
ann_id <- ann_id |>
  filter(!is.na(mz_observed))
# Per (mixture, lab, compound, polarity), take the SIRIUS row with the
# highest confidenceApproxMatch as the canonical m/z observation.
ann_id$confidenceApproxMatch[is.na(ann_id$confidenceApproxMatch)] <- 0
ann_id_top <- ann_id |>
  group_by(mixture, lab, compound, polarity) |>
  slice_max(confidenceApproxMatch, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(mixture, lab, compound, polarity, mz_observed)

ann_dom <- ann_chrom |>
  filter(detection == "detected", !is.na(apex_rt), !is.na(maxIntensity)) |>
  group_by(mixture, lab, compound, polarity) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup() |>
  inner_join(ann_id_top,
             by = c("mixture", "lab", "compound", "polarity"))

cat(sprintf("Annotated (mz_observed + own detected peak) lab observations: %d\n",
            nrow(ann_dom)))

# Per (consensus_id × lab) representative — `disambig` already ensures
# at most one peak per lab per consensus, so each row IS the centroid.
consensus_rep <- disambig |>
  transmute(consensus_id, lab, sample,
            mz_rep = mz, rt_rep = match_rt)

# n_labs per consensus
n_labs_per_cid <- ct_all_samples |>
  select(consensus_id, n_labs_matched)

# Convert annotated mixture "2.10" → "2_10" to match consensus side.
ann_dom <- ann_dom |>
  mutate(mixture_us = gsub("\\.", "_", mixture))

# Sample → mixture mapping
sample_to_mixture_pol <- pt_afekta |>
  distinct(sample, mixture, polarity)

# Resolve annotated → sample
ann_dom <- ann_dom |>
  inner_join(sample_to_mixture_pol,
             by = c("mixture_us" = "mixture", "polarity"))

# Restrict consensus_rep to per (sample, lab, mixture, polarity)
PPM_TOL_MATCH <- 20
RT_TOL_MATCH  <- 15
ann_matched <- ann_dom |>
  inner_join(consensus_rep, by = c("sample", "lab"),
             relationship = "many-to-many") |>
  mutate(mz_ok = abs(mz_observed - mz_rep) <=
                  mz_observed * PPM_TOL_MATCH * 1e-6,
         rt_ok = abs(apex_rt - rt_rep) <= RT_TOL_MATCH) |>
  filter(mz_ok, rt_ok) |>
  mutate(dist = abs(mz_observed - mz_rep) /
                (mz_observed * PPM_TOL_MATCH * 1e-6) +
                abs(apex_rt - rt_rep) / RT_TOL_MATCH) |>
  group_by(mixture, lab, compound, polarity) |>
  slice_min(dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  inner_join(n_labs_per_cid, by = "consensus_id")

cat(sprintf("Annotated rows mapped to a consensus feature: %d / %d (%.1f%%)\n",
            nrow(ann_matched), nrow(ann_dom),
            100 * nrow(ann_matched) / nrow(ann_dom)))

# Distribution by consensus n_labs
knitr::kable(
  ann_matched |> count(n_labs_matched) |> arrange(n_labs_matched),
  caption = "Mapped annotated rows by consensus cross-lab support")

# Per-compound aggregation
per_cpd <- ann_matched |>
  group_by(mixture, compound, polarity) |>
  summarise(n_obs = n(),
            n_in_4lab = sum(n_labs_matched == 4),
            n_in_3plus = sum(n_labs_matched >= 3),
            .groups = "drop")
cat(sprintf("\nDistinct (mixture, compound, polarity) spike-ins observed: %d\n",
            nrow(per_cpd)))
cat(sprintf("  ALL observations in 4-lab consensus: %d (%.1f%%)\n",
            sum(per_cpd$n_in_4lab == per_cpd$n_obs),
            100 * mean(per_cpd$n_in_4lab == per_cpd$n_obs)))
cat(sprintf("  ALL observations in 3+ lab consensus: %d (%.1f%%)\n",
            sum(per_cpd$n_in_3plus == per_cpd$n_obs),
            100 * mean(per_cpd$n_in_3plus == per_cpd$n_obs)))

# Reverse: of 4-lab consensus features, how many are SIRIUS-identified?
total_4lab <- sum(ct_all_samples$n_labs_matched == 4)
known_4lab <- n_distinct(ann_matched$consensus_id[
  ann_matched$n_labs_matched == 4])
cat(sprintf("\nOf %d four-lab consensus features:\n", total_4lab))
cat(sprintf("  with a SIRIUS-identified spike-in: %d (%.1f%%)\n",
            known_4lab, 100 * known_4lab / total_4lab))
cat(sprintf("  unidentified background chemistry: %d (%.1f%%)\n",
            total_4lab - known_4lab,
            100 * (total_4lab - known_4lab) / total_4lab))
```

**Interpretation (pre-QC, all consensus features).** This baseline uses
*every* consensus feature, including ones that later fail the peak-
quality filters. For the **headline numbers used in the presentation**
— the same mapping computed against the **QC-passed consensus set** —
see the dedicated section below ("Cross-reference with SIRIUS panel —
QC-passed features only"), which is what should be quoted on slides.
This block is retained because the alignment-effect comparison below
needs `ann_dom`, `consensus_rep`, and `n_labs_per_cid` as built here.


## Effect of NAPS alignment on consensus matching

The main pipeline above matches peaks on `match_rt` (NAPS-aligned with
raw-`rt` fallback for peaks outside the NAPS anchor range). To quantify
what alignment buys, we rerun the same igraph two-pointer matching on
**raw `rt`** and compare:

- How many cross-lab consensus features survive (especially the
  strict all-4-lab subset)?
- How tight are the matched groups in RT — i.e. do peaks at the same
  true compound across labs actually land within a few seconds of
  each other?

The raw-rt matching uses identical PPM_TOL = 20 and RT_TOL = 15 s.
Only the RT column being matched on differs.

```{r raw-matching-comparison}
#| message: false
raw_cache <- file.path(base_path, "object/correspondence_grouped_raw.rds")
if (file.exists(raw_cache)) {
  grouped_raw <- readRDS(raw_cache)
} else {
  samples_r <- sort(unique(pt_all$sample))
  grouped_raw <- do.call(rbind, lapply(samples_r, function(s) {
    g <- group_sample(pt_all, s, rt_col = "rt")
    g$consensus_id <- paste(s, g$consensus_id, sep = "_")
    g
  }))
  grouped_raw$consensus_id <- as.integer(factor(grouped_raw$consensus_id))
  saveRDS(grouped_raw, raw_cache)
}
cat("Raw-rt matching: ", nrow(grouped_raw), " peaks in ",
    length(unique(grouped_raw$consensus_id)), " consensus groups\n", sep = "")
```

```{r alignment-effect-stats}
#| message: false
# Per-consensus stats for both passes: n_labs + intra-group spread on the
# axis the matching ran on.
stat_pass <- function(grp, rt_col, label) {
  dt <- as.data.table(grp)
  dt[, c("rt_use", "label") := list(get(rt_col), label)]
  nlabs <- dt[, .(n_labs = uniqueN(lab)), by = consensus_id]
  spread <- dt[nlabs[n_labs >= 2], on = "consensus_id",
               .(spr = max(rt_use) - min(rt_use)), by = consensus_id]
  list(n_groups   = uniqueN(dt$consensus_id),
       n_4lab     = sum(nlabs$n_labs == 4),
       n_3plus    = sum(nlabs$n_labs >= 3),
       median_spr = median(spread$spr),
       p75_spr    = quantile(spread$spr, 0.75),
       p95_spr    = quantile(spread$spr, 0.95),
       max_spr    = max(spread$spr),
       spread_vec = spread$spr,
       nlabs_vec  = nlabs$n_labs)
}

st_raw <- stat_pass(grouped_raw, "rt",       "RAW (rt)")
st_aln <- stat_pass(grouped,     "match_rt", "ALIGNED (match_rt)")

cmp <- data.frame(
  Metric = c("Consensus groups (any n_labs)",
             "Groups in all 4 labs",
             "Groups in 3+ labs",
             "Median intra-group RT spread (s)",
             "p75 intra-group RT spread (s)",
             "p95 intra-group RT spread (s)",
             "Max intra-group RT spread (s)"),
  RAW_rt        = c(st_raw$n_groups, st_raw$n_4lab, st_raw$n_3plus,
                    round(st_raw$median_spr, 2),
                    round(st_raw$p75_spr,   2),
                    round(st_raw$p95_spr,   2),
                    round(st_raw$max_spr,   2)),
  ALIGNED_match = c(st_aln$n_groups, st_aln$n_4lab, st_aln$n_3plus,
                    round(st_aln$median_spr, 2),
                    round(st_aln$p75_spr,   2),
                    round(st_aln$p95_spr,   2),
                    round(st_aln$max_spr,   2)),
  Delta_pct     = round(100 *
                        (c(st_aln$n_groups, st_aln$n_4lab, st_aln$n_3plus,
                           st_aln$median_spr, st_aln$p75_spr,
                           st_aln$p95_spr,   st_aln$max_spr) /
                         c(st_raw$n_groups, st_raw$n_4lab, st_raw$n_3plus,
                           st_raw$median_spr, st_raw$p75_spr,
                           st_raw$p95_spr,   st_raw$max_spr) - 1), 1))
knitr::kable(cmp, col.names = c("Metric", "RAW (rt)",
                                "ALIGNED (match_rt)", "Δ %"),
             caption = "Effect of NAPS-spline alignment on consensus matching")
```

```{r alignment-effect-histogram}
#| fig-cap: "Distribution of intra-group RT spread across consensus features detected in ≥ 2 labs, on a log10 axis. Aligned matching shifts the bulk of features to a tighter RT spread (~ 3-4 s median) compared with raw matching (~ 6 s)."
#| fig-height: 4
spr_df <- rbind(
  data.frame(spr_s = st_raw$spread_vec, pass = "RAW (rt)"),
  data.frame(spr_s = st_aln$spread_vec, pass = "ALIGNED (match_rt)"))
spr_df$pass <- factor(spr_df$pass,
                      levels = c("RAW (rt)", "ALIGNED (match_rt)"))

ggplot(spr_df, aes(x = spr_s, fill = pass)) +
  geom_histogram(bins = 60, alpha = 0.7, position = "identity") +
  scale_x_log10(labels = scales::label_number(accuracy = 0.1)) +
  scale_fill_manual(values = c("RAW (rt)" = "grey55",
                                "ALIGNED (match_rt)" = "#1F77B4")) +
  labs(x = "Intra-group RT spread (s, log)", y = "Number of consensus groups",
       fill = NULL,
       title = "Intra-group RT spread: raw vs aligned matching") +
  theme_bw() +
  theme(legend.position = "top")
```

```{r alignment-effect-nlabs-distribution}
#| fig-cap: "Distribution of consensus group sizes (n_labs_matched) under raw vs aligned matching. Aligned matching recovers ~ 35 % more all-4-lab features by bringing peaks just outside the 15 s window inside it."
#| fig-height: 3.5
nlab_df <- rbind(
  data.frame(n_labs = st_raw$nlabs_vec, pass = "RAW (rt)"),
  data.frame(n_labs = st_aln$nlabs_vec, pass = "ALIGNED (match_rt)"))
nlab_df$pass <- factor(nlab_df$pass,
                       levels = c("RAW (rt)", "ALIGNED (match_rt)"))

ggplot(nlab_df, aes(x = factor(n_labs), fill = pass)) +
  geom_bar(position = "dodge", alpha = 0.85) +
  scale_y_log10() +
  scale_fill_manual(values = c("RAW (rt)" = "grey55",
                                "ALIGNED (match_rt)" = "#1F77B4")) +
  labs(x = "n_labs_matched", y = "Consensus groups (log)",
       fill = NULL,
       title = "Consensus group size distribution: raw vs aligned") +
  theme_bw() +
  theme(legend.position = "top")
```

**What alignment does and doesn't do.** The matching gains
(+~ 35 % four-lab features, intra-group spread halved) tell us
alignment is doing real work at the *grouping* level — peaks that
were just outside the 15 s tolerance under raw rt now fall inside it
after the NAPS spline corrects the per-lab drift. But the **intensity
CV on the four-lab subset stays around 0.94-0.96 either way** (Phase 5
of [`7_reproducibility_results.qmd`](7_reproducibility_results.qmd)),
which is the cleanest possible statement that the residual cross-lab
abundance disagreement is structural — per-lab MS detector response —
and not something alignment can rescue.

### Does alignment improve the SIRIUS-panel mapping?

The cross-reference above (SIRIUS-annotated → consensus) was run on
the aligned matching only. Here we repeat it under raw matching and
report the delta — the test of whether alignment is also recovering
more *identified* compounds in the cross-lab consensus, not just more
unknown features.

```{r sirius-mapping-raw-vs-aligned}
#| message: false
# Re-run the SIRIUS panel mapping on the RAW grouping.
grouped_raw_dt <- as.data.table(grouped_raw)
consensus_rep_raw <- grouped_raw_dt[
  , .(mz_rep = mean(mz), rt_rep = mean(rt)),
    by = .(consensus_id, lab, sample)]
nlabs_raw <- grouped_raw_dt[
  , .(n_labs_matched = uniqueN(lab)), by = consensus_id]

map_sirius <- function(ann_dom, consensus_rep, nlabs,
                       ppm_tol = 20, rt_tol = 15) {
  hits <- ann_dom |>
    inner_join(as.data.frame(consensus_rep),
               by = c("sample", "lab"),
               relationship = "many-to-many") |>
    mutate(mz_ok = abs(mz_observed - mz_rep) <=
                    mz_observed * ppm_tol * 1e-6,
           rt_ok = abs(apex_rt - rt_rep) <= rt_tol) |>
    filter(mz_ok, rt_ok) |>
    mutate(dist = abs(mz_observed - mz_rep) /
                  (mz_observed * ppm_tol * 1e-6) +
                  abs(apex_rt - rt_rep) / rt_tol) |>
    group_by(mixture, lab, compound, polarity) |>
    slice_min(dist, n = 1, with_ties = FALSE) |>
    ungroup() |>
    inner_join(as.data.frame(nlabs), by = "consensus_id")

  per_cpd <- hits |>
    group_by(mixture, compound, polarity) |>
    summarise(n_obs       = n(),
              n_in_4lab   = sum(n_labs_matched == 4),
              n_in_3plus  = sum(n_labs_matched >= 3),
              .groups = "drop")
  list(n_rows_mapped     = nrow(hits),
       n_rows_total      = nrow(ann_dom),
       n_compounds       = nrow(per_cpd),
       n_compounds_4lab  = sum(per_cpd$n_in_4lab == per_cpd$n_obs),
       n_compounds_3plus = sum(per_cpd$n_in_3plus == per_cpd$n_obs))
}

map_raw <- map_sirius(ann_dom, consensus_rep_raw, nlabs_raw)
map_aln <- map_sirius(ann_dom, consensus_rep,    n_labs_per_cid)

sirius_cmp <- data.frame(
  Metric = c("Annotated lab observations mapped to a consensus feature",
             "  ... mapping rate",
             "Distinct spike-ins observed in any consensus group",
             "Spike-ins with ALL lab observations in a 4-lab consensus",
             "  ... fraction",
             "Spike-ins with ALL lab observations in a 3+ lab consensus",
             "  ... fraction"),
  RAW = c(sprintf("%d / %d", map_raw$n_rows_mapped, map_raw$n_rows_total),
          sprintf("%.1f%%",  100 * map_raw$n_rows_mapped / map_raw$n_rows_total),
          map_raw$n_compounds,
          map_raw$n_compounds_4lab,
          sprintf("%.1f%%",  100 * map_raw$n_compounds_4lab / map_raw$n_compounds),
          map_raw$n_compounds_3plus,
          sprintf("%.1f%%",  100 * map_raw$n_compounds_3plus / map_raw$n_compounds)),
  ALIGNED = c(sprintf("%d / %d", map_aln$n_rows_mapped, map_aln$n_rows_total),
              sprintf("%.1f%%",  100 * map_aln$n_rows_mapped / map_aln$n_rows_total),
              map_aln$n_compounds,
              map_aln$n_compounds_4lab,
              sprintf("%.1f%%",  100 * map_aln$n_compounds_4lab / map_aln$n_compounds),
              map_aln$n_compounds_3plus,
              sprintf("%.1f%%",  100 * map_aln$n_compounds_3plus / map_aln$n_compounds)))
knitr::kable(sirius_cmp,
             caption = "SIRIUS spike-in panel → consensus feature mapping: raw vs aligned matching")
```

If aligned matching is genuinely recovering structure (not just adding
noise), the spike-in-coverage numbers should move in the same
direction as the bulk numbers above: more identified compounds in the
4-lab consensus, more in the 3+ lab consensus.

## Future work

The consensus feature table built above contains ~ 3 500 four-lab
chromatographic features, of which only ~ 2 % are SIRIUS-identified
spike-ins. The remainder is reproducible-but-unannotated chemistry. A
few natural follow-ups become possible once both tables are in hand:

### 1. ISF + adduct discovery via co-elution with annotated compounds

For each SIRIUS-confirmed compound at apex RT *t* and parent
m/z *m*, neighbouring consensus features that elute at the same RT
(*t* ± δ_rt) but at predictable mass offsets are candidate
**in-source fragments** (-H₂O = -18.011, -CO₂ = -43.990, -NH₃ = -17.027),
**adducts** (M+Na - M+H = +21.982, M+K - M+H = +37.956,
M-H₂O+H = -18.011), or **isotopologues** (+1.003 m/z per ¹³C).

Implementation sketch:

- For each (mixture, polarity, compound) at known apex RT, find all
  consensus features (4-lab subset) within ± 3 s of that RT.
- Compute Δm/z to the parent m/z; flag matches within ± 5 ppm of any
  known offset.
- Aggregate per compound: how many ISF / adduct / isotopologue
  candidates explain unknown 4-lab features near each spike-in?

This converts a fraction of the "unidentified background" into
structural relatives of identified compounds — and tightens the
estimate of how much truly novel chemistry sits in the 4-lab
consensus.

### 2. Diagnostic for annotated peaks that fail to map

About a third of annotated lab observations don't find a consensus
feature. Likely causes, ordered by expected frequency:

- **xcms missed the peak** at that lab's apex RT — SIRIUS confirmed
  via MS2 fragments but the underlying MS1 ion was below the per-lab
  peak picker's detection floor. Cross-checkable against
  `detected_peaks_<lab>_HE.csv` row count near (mz_observed, apex_rt).
- **m/z mismatch** — the SIRIUS top candidate's reported `mz_observed`
  differs from the xcms-detected ion m/z by > 20 ppm. Worth flagging
  as candidate (mis)identification.
- **RT mismatch** — `apex_rt` from the SNR-aware chromExtract differs
  from any xcms peak's RT by > 15 s. Should be rare after the cluster
  algorithm fix in §7.1 of the producer.

A diagnostic table joining annotated `(mixture, lab, compound)` to
the per-lab detected-peaks CSV would categorize each unmapped row and
surface systematic gaps (e.g. is hmgu over-represented among the
"xcms missed" category, consistent with the DDA-acquisition story?).


## Upset Plot - Lab Overlap

```{r upset-plot, message = FALSE, warning = FALSE}
# Prepare binary matrix for upset plot
upset_data <- ct_all_samples |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(
    afekta = as.integer(afekta_detected),
    hmgu = as.integer(hmgu_detected),
    icl = as.integer(icl_detected),
    cembio = as.integer(cembio_detected)
  ) |>
  select(afekta, hmgu, icl, cembio)

# Convert to data frame with rownames for upset function
upset_data_df <- as.data.frame(upset_data)
rownames(upset_data_df) <- paste0("peak_", seq_len(nrow(upset_data_df)))

# Only plot intersections with at least 1 peak (exclude empty combinations)
upset(upset_data_df,
      sets = c("afekta", "hmgu", "icl", "cembio"),
      main.bar.color = "steelblue",
      sets.bar.color = "darkred",
      text.scale = c(2, 2, 2, 1.5, 2, 2),
      point.size = 3,
      line.size = 1.2)
```

## Pairwise Lab Comparisons

```{r pairwise-overlap}
# Create pairwise overlap summary
labs <- c("afekta", "hmgu", "icl", "cembio")
pairwise_summary <- data.frame()

for (i in seq_along(labs)) {
  for (j in seq_along(labs)) {
    if (i < j) {
      lab1 <- labs[i]
      lab2 <- labs[j]

      col1 <- paste0(lab1, "_detected")
      col2 <- paste0(lab2, "_detected")

      both <- sum(ct_all_samples[[col1]] & ct_all_samples[[col2]])
      only_lab1 <- sum(ct_all_samples[[col1]] & !ct_all_samples[[col2]])
      only_lab2 <- sum(!ct_all_samples[[col1]] & ct_all_samples[[col2]])

      total_lab1 <- sum(ct_all_samples[[col1]])
      total_lab2 <- sum(ct_all_samples[[col2]])

      pairwise_summary <- rbind(pairwise_summary, data.frame(
        lab1 = lab1,
        lab2 = lab2,
        both_detected = both,
        only_lab1 = only_lab1,
        only_lab2 = only_lab2,
        total_lab1 = total_lab1,
        total_lab2 = total_lab2,
        jaccard_index = round(100 * both / (both + only_lab1 + only_lab2), 1),
        overlap_percent_lab1 = round(100 * both / total_lab1, 1),
        overlap_percent_lab2 = round(100 * both / total_lab2, 1)
      ))
    }
  }
}

knitr::kable(pairwise_summary, caption = "Pairwise Lab Overlap - Detailed Metrics")
```

As expected, afekta and cembio have the highest overlap.

```{r}
ct_filtered <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Consensus peaks retained:", nrow(ct_filtered), "\n")
cat(
  "Percentage of detected peaks:",
  round(100 * nrow(ct_filtered) / nrow(ct_all_samples), 1),
  "%\n"
)
```

Build the final consensus peak table with chrom_peak_id mapping:
Save the consensus table for downstream analysis:

```{r save-consensus}
write.csv(
  ct_all_samples,
  file = file.path(base_path, "object/ct_all_samples.csv"),
  row.names = FALSE
)
write.csv(
  ct_filtered,
  file = file.path(base_path, "object/ct_filtered4.csv"),
  row.names = FALSE
)
```

## Summary Tables

### Overlap: Cembio vs Afekta

```{r cembio-afekta-overlap}
cembio_afekta_summary <- data.frame(
  category = c("Both detected", "Only afekta", "Only cembio", "Neither"),
  count = c(
    sum(ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected)
  )
)

cembio_afekta_summary <- cembio_afekta_summary |>
  mutate(percent = round(100 * count / sum(count), 1))

knitr::kable(cembio_afekta_summary,
             caption = "Peak Overlap: Cembio vs Afekta")
```

### Overlap: All 4 Labs

```{r all-labs-overlap}
all_labs_summary <- data.frame(
  detection_pattern = character(),
  n_peaks = integer(),
  percent = numeric()
)

for (combo in 0:15) {
  afekta <- as.logical(bitwAnd(combo, 1))
  hmgu <- as.logical(bitwAnd(combo, 2))
  icl <- as.logical(bitwAnd(combo, 4))
  cembio <- as.logical(bitwAnd(combo, 8))

  n <- sum(
    ct_all_samples$afekta_detected == afekta &
    ct_all_samples$hmgu_detected == hmgu &
    ct_all_samples$icl_detected == icl &
    ct_all_samples$cembio_detected == cembio
  )

  if (n > 0) {
    pattern <- paste0(
      if_else(afekta, "A", "-"),
      if_else(hmgu, "H", "-"),
      if_else(icl, "I", "-"),
      if_else(cembio, "C", "-")
    )

    all_labs_summary <- rbind(all_labs_summary, data.frame(
      detection_pattern = pattern,
      n_peaks = n,
      percent = round(100 * n / nrow(ct_all_samples), 1)
    ))
  }
}

all_labs_summary <- all_labs_summary |>
  arrange(desc(n_peaks))

knitr::kable(all_labs_summary,
             caption = "Peak Detection Patterns: All 4 Labs (A=afekta, H=hmgu, I=icl, C=cembio)")
```



# Peak Feature Summary by Lab (Detected)

## Prepare subset for downstream analysis

```{r prepare-subsets}
# Peaks present in ALL 4 labs (main analysis subset)
all_four_labs <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Peaks in all 4 labs:", nrow(all_four_labs), "\n")
```

# Creating eics
## 4 labs

```{r}
pt <- all_four_labs |>
  mutate(consensus_id = row_number()) |>
  select(consensus_id, sample, chrom_peak_id_afekta, chrom_peak_id_hmgu,
  chrom_peak_id_icl, chrom_peak_id_cembio) |>
  pivot_longer(
    cols = c(chrom_peak_id_afekta, chrom_peak_id_hmgu, chrom_peak_id_icl,
    chrom_peak_id_cembio),
    names_to = "lab",
    names_prefix = "chrom_peak_id_",
    values_to = "chrom_peak_id"
  ) |>
  filter(!is.na(chrom_peak_id))
```

```{r}
# Combine all lab peak tables with lab identifier for simple merge
pt_all <- rbind(
  transform(pt_afekta, lab = "afekta"),
  transform(pt_hmgu, lab = "hmgu"),
  transform(pt_icl, lab = "icl"),
  transform(pt_cembio, lab = "cembio")
)

# Join peak info using merge instead of case_when
join_cols <- c("chrom_peak_id", "lab", "mzmin", "mzmax", "rtmin", "rtmax",
               "mz", "polarity", "mixture", "into", "intb", "maxo")
pt <- merge(pt, pt_all[, join_cols], by = c("chrom_peak_id", "lab"), all.x = TRUE)

# Rename and adjust boundaries
names(pt)[names(pt) == "mzmin"] <- "mzMin"
names(pt)[names(pt) == "mzmax"] <- "mzMax"
names(pt)[names(pt) == "rtmin"] <- "rtMin"
names(pt)[names(pt) == "rtmax"] <- "rtMax"

pt$rtMin <- pt$rtMin - 2
pt$rtMax <- pt$rtMax + 2
pt$mzMin <- pt$mzMin - 0.02
pt$mzMax <- pt$mzMax + 0.02
pt$polarity <- ifelse(pt$sample %in% 1:19, 1L, 0L)

write.csv(
  pt,
  file = file.path(base_path, "object/peak_table_detected_all4labs.csv"),
  row.names = FALSE
)
```

```{r}
library(Chromatograms)

if (!file.exists(file.path(base_path, "object/eics_detected_all4labs.RData"))) {
  load(file.path(base_path, "object/sp_full.RData"))
  tic_full <- Chromatograms(
    sp_full,
    summarize.method = "sum",
    factorize.by = "dataOrigin",
    spectraVariables = c("mixture", "lab")
  )
  eics <- chromExtract(
    tic_full,
    peak.table = pt,
    by = c("polarity", "mixture", "lab")
  )
  nrow(pt) == length(eics)
  eics <- setBackend(eics, ChromBackendMemory())
  save(eics, file = file.path(base_path, "object/eics_detected_all4labs.RData"))
} else {
  load(file.path(base_path, "object/eics_detected_all4labs.RData"))
}

eics
```

## Compute metrics


```{r compute-metrics}
# Step 1: peak boundaries — now in Chromatograms, no longer a calculateMetrics metric
cat("Computing peak boundaries...\n")
pb <- Chromatograms::peakBoundary(eics)
cat("peakBoundary columns:", paste(names(pb), collapse = ", "), "\n")
chromData(eics) <- cbind(chromData(eics), pb)

# Step 2: quality metrics used for filtering and downstream doc-3 analysis
# Removed: peakBoundary (above), gaussianSimilarity (computed on narrow EICs below)
# Removed: f = seq_len() — not a valid parameter in current MsQuality
pk <- calculateMetrics(
  eics,
  metrics = c(
    "peakCount", "xicFwhm", "maxIntensity",
    "baselineIntensity", "signalToNoiseRatio",
    "peakWidth", "peakProminence"
  ),
  na.rm = TRUE,
  filterEmptyObject = FALSE,
  probs = 0.05
)

chromData(eics) <- cbind(chromData(eics), pk)
cat("Metrics computed for", length(eics), "EICs\n")
```

## Filter bad peaks

Remove consensus peaks with poor quality metrics in a single step.

```{r filter-bad-peaks}
# Consolidate all filtering criteria
bad_ids <- unique(c(
  eics$consensus_id[is.na(eics$xicFwhm)],           # NA FWHM
  eics$consensus_id[eics$peakProminence < 1.5],     # Low prominence
  eics$consensus_id[eics$peakWidth < 1],            # Too narrow
  eics$consensus_id[eics$peakWidth > 30]            # Too wide
))

cat("Removing", length(bad_ids), "consensus peaks with poor metrics\n")
eics <- eics[!eics$consensus_id %in% bad_ids]
cat("Remaining EICs:", length(eics), "\n")
cat("Unique consensus IDs:", length(unique(eics$consensus_id)), "\n")
```

## Cross-reference with SIRIUS panel — QC-passed features only

The earlier "Cross-reference with SIRIUS-annotated panel" section ran
the annotation-to-consensus mapping against **every** consensus
feature, including ones that fail the peak-quality filter (FWHM /
prominence / width). For a presentation, we want the mapping against
the **QC-passed subset** — the consensus features actually used in
all downstream reproducibility analyses. This chunk redoes the same
mapping restricted to `qc_passed_ids`.

```{r annotated-to-consensus-qc-passed}
#| message: false
ids_4lab        <- ct_all_samples |> filter(n_labs_matched == 4L) |>
  pull(consensus_id) |> unique()
qc_passed_ids   <- setdiff(ids_4lab, bad_ids)
consensus_rep_qc <- consensus_rep |>
  filter(consensus_id %in% qc_passed_ids)
n_labs_per_cid_qc <- n_labs_per_cid |>
  filter(consensus_id %in% qc_passed_ids)

cat(sprintf("Four-lab consensus features: %d (pre-QC) → %d (post-QC, lost %d to FWHM / prominence / width filters)\n",
            length(ids_4lab), length(qc_passed_ids),
            length(ids_4lab) - length(qc_passed_ids)))

ann_matched_qc <- ann_dom |>
  inner_join(consensus_rep_qc, by = c("sample", "lab"),
             relationship = "many-to-many") |>
  mutate(mz_ok = abs(mz_observed - mz_rep) <=
                  mz_observed * PPM_TOL_MATCH * 1e-6,
         rt_ok = abs(apex_rt - rt_rep) <= RT_TOL_MATCH) |>
  filter(mz_ok, rt_ok) |>
  mutate(dist = abs(mz_observed - mz_rep) /
                (mz_observed * PPM_TOL_MATCH * 1e-6) +
                abs(apex_rt - rt_rep) / RT_TOL_MATCH) |>
  group_by(mixture, lab, compound, polarity) |>
  slice_min(dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  inner_join(n_labs_per_cid_qc, by = "consensus_id")

cat(sprintf("\nAnnotated lab observations mapped to a QC-passed consensus feature: %d / %d (%.1f%%)\n",
            nrow(ann_matched_qc), nrow(ann_dom),
            100 * nrow(ann_matched_qc) / nrow(ann_dom)))

knitr::kable(
  ann_matched_qc |> count(n_labs_matched) |> arrange(n_labs_matched),
  caption = "Mapped annotated rows (QC-passed only) by consensus cross-lab support.")

per_cpd_qc <- ann_matched_qc |>
  group_by(mixture, compound, polarity) |>
  summarise(n_obs       = n(),
            n_in_4lab   = sum(n_labs_matched == 4),
            n_in_3plus  = sum(n_labs_matched >= 3),
            .groups     = "drop")
cat(sprintf("\nDistinct (mixture, compound, polarity) spike-ins with at least one QC-passed mapping: %d\n",
            nrow(per_cpd_qc)))
cat(sprintf("  ALL observations in 4-lab consensus: %d (%.1f%%)\n",
            sum(per_cpd_qc$n_in_4lab == per_cpd_qc$n_obs),
            100 * mean(per_cpd_qc$n_in_4lab == per_cpd_qc$n_obs)))
cat(sprintf("  ALL observations in 3+ lab consensus: %d (%.1f%%)\n",
            sum(per_cpd_qc$n_in_3plus == per_cpd_qc$n_obs),
            100 * mean(per_cpd_qc$n_in_3plus == per_cpd_qc$n_obs)))

# Reverse: of QC-passed 4-lab consensus features, how many are SIRIUS-identified?
total_4lab_qc <- ct_all_samples |>
  filter(consensus_id %in% qc_passed_ids, n_labs_matched == 4) |>
  nrow()
known_4lab_qc <- n_distinct(ann_matched_qc$consensus_id[
  ann_matched_qc$n_labs_matched == 4])
cat(sprintf("\nOf %d QC-passed four-lab consensus features:\n", total_4lab_qc))
cat(sprintf("  with a SIRIUS-identified spike-in: %d (%.1f%%)\n",
            known_4lab_qc, 100 * known_4lab_qc / total_4lab_qc))
cat(sprintf("  unidentified background chemistry: %d (%.1f%%)\n",
            total_4lab_qc - known_4lab_qc,
            100 * (total_4lab_qc - known_4lab_qc) / total_4lab_qc))
```

These are the numbers to quote in the presentation — they describe the
mapping against the **same feature set the reproducibility analyses
use** (sp_consensus / tic_consensus).

### Concordant peaks per mixture

```{r concordant-per-mixture, fig.width=10, fig.height=4}
#| fig-cap: "Number of consensus peaks detected in all 4 labs per mixture (post-filter)."
mix_counts <- as.data.frame(chromData(eics)) |>
  distinct(consensus_id, mixture, polarity) |>
  count(mixture, polarity, name = "n_peaks") |>
  mutate(polarity = ifelse(polarity == 1L, "Positive", "Negative"))

ggplot(mix_counts, aes(x = mixture, y = n_peaks)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  geom_text(aes(label = n_peaks), vjust = -0.3, size = 3) +
  facet_wrap(~polarity, ncol = 1, scales = "free_y") +
  labs(x = "Mixture", y = "Concordant peaks (all 4 labs)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Re-extract EICs using peak boundaries

```{r re-extract}
# NOTE: delete the cached RData below if you want to force re-extraction
# (necessary after changes to boundary computation or imputation).
rdata_path <- file.path(base_path, "object/eics_filtered_detected_all4labs.RData")

if (!file.exists(rdata_path)) {
  # Trim each EIC to its peak boundaries + small margin, then impute edge NAs.
  # peakBoundary column names are printed above — adjust if different.
  final_eics <- lapply(seq_along(eics), function(i) {
    eic  <- eics[i]
    rt1  <- eic$left_boundary
    rt2  <- eic$right_boundary
    if (is.na(rt1) || is.na(rt2)) return(eic)
    eic  <- filterPeaksData(eic, variables = "rtime",
                            ranges = c(rt1 - 0.5, rt2 + 0.5))
    applyProcessing(eic)
  })
  final_eics <- concatenateChromatograms(final_eics)
  # Impute residual NAs (boundary edge artefacts from scan-grid misalignment)
  final_eics <- imputePeaksData(final_eics, method = "linear")
  save(final_eics, file = rdata_path)
} else {
  load(rdata_path)
}

stopifnot(all(final_eics$consensus_id == eics$consensus_id))
cat("Re-extracted EICs:", length(final_eics), "\n")

# Patch remaining edge NAs — linear imputation cannot fill leading/trailing NAs
# because there is no boundary value to interpolate toward.  At peak boundaries
# the signal approaches baseline, so 0 is the appropriate fill value.
# This block also runs on a loaded cache, so stale objects self-heal.
n_na <- sum(is.na(unlist(intensity(final_eics))))
cat("NAs after linear imputation:", n_na, "\n")
if (n_na > 0) {
  cat("  -> filling", n_na, "edge NAs with 0 (signal at peak boundary ≈ baseline)...\n")
  pd <- peaksData(final_eics)
  pd <- lapply(pd, function(df) {
    df$intensity[is.na(df$intensity)] <- 0
    df
  })
  peaksData(final_eics) <- pd
  cat("NAs after 0-fill:", sum(is.na(unlist(intensity(final_eics)))), "\n")
  save(final_eics, file = rdata_path)  # patch the cache so next load is already clean
} else {
  cat("  -> no edge NAs found.\n")
}
```

```{r compute-metrics-final}
# Second metrics round — computed on boundary-trimmed, imputed EICs.
# Only the three metrics needed by the downstream ICC analysis.
# (areaUnderTic and xicFwhm on narrow EICs are more accurate than on wide EICs.)
# Guard: skip if metrics are already stored in the loaded RData.
if (!("areaUnderTic" %in% names(chromData(final_eics)))) {
  cat("Computing tight-EIC metrics (areaUnderTic, xicFwhm, gaussianSimilarity)...\n")
  pk2 <- calculateMetrics(
    final_eics,
    metrics = c("areaUnderTic", "xicFwhm", "gaussianSimilarity"),
    na.rm = TRUE,
    filterEmptyObject = FALSE
  )
  # Store metrics in chromData(final_eics) so doc 4 can access them after
  # loading the same RData object it needs for compareChromatograms anyway.
  chromData(final_eics) <- cbind(chromData(final_eics), pk2)
  save(final_eics, file = rdata_path)
  cat("Metrics stored in chromData and RData re-saved.\n")
} else {
  cat("Tight-EIC metrics already in chromData — skipping calculateMetrics.\n")
  pk2 <- as.data.frame(chromData(final_eics))[,
    intersect(c("areaUnderTic", "xicFwhm",
                "gaussianSimilarity.gaussian_similarity",
                "gaussianSimilarity.gaussian_residuals"),
              names(chromData(final_eics)))
  ]
}

# Build the full analysis table for doc-3 analysis: tight-EIC metrics +
# QC metrics from the first round.
# left_boundary/right_boundary column names must match what peakBoundary() added.
pk_filtered <- cbind(
  chromData(eics)[, c(
    "lab", "consensus_id", "mixture", "polarity", "mz", "chrom_peak_id",
    "peakCount", "maxIntensity", "baselineIntensity", "signalToNoiseRatio",
    "peakWidth", "peakProminence", "left_boundary", "right_boundary"
  )],
  pk2
)

cat("pk_filtered columns:", paste(names(pk_filtered), collapse = ", "), "\n")
cat("Rows:", nrow(pk_filtered), "\n")

write.csv(
  pk_filtered,
  file = file.path(base_path, "object/peak_metrics_detected_all4labs.csv"),
  row.names = FALSE
)
```

## Create Spectra - Consensus Signal (post-filter)

Spectra filtered to the m/z × RT rectangles of the **filtered +
boundary-trimmed** consensus peaks, per (lab, mixture, polarity)
sample. Same pattern as `detect_signal()` in
`helpers.R` — `filterPeaksRanges()` keeps only mass peaks
inside any of that sample's consensus windows.

```{r build-sp-consensus}
library(Spectra)

if (!file.exists(file.path(base_path, "object/sp_consensus.RData"))) {
  load(file.path(base_path, "object/sp_full.RData"))

  # Per-(lab, mixture, polarity) consensus peak windows, boundary-trimmed
  cd <- as.data.frame(chromData(final_eics))
  cd <- cd[!is.na(cd$left_boundary) & !is.na(cd$right_boundary), ]

  # Match spectra ↔ consensus windows on (lab, mixture, polarity)
  sp_key <- paste(sp_full$lab, sp_full$mixture, polarity(sp_full), sep = "_")
  cd_key <- paste(cd$lab, cd$mixture, cd$polarity, sep = "_")

  # Per-sample: filter mass peaks to consensus rectangles, then apply
  # the lazy processing queue so concatenateSpectra() will accept the
  # result (mirrors detect_signal() at helpers.R:104-117).
  filtered_list <- lapply(unique(sp_key), function(key) {
    s <- sp_full[sp_key == key]
    w <- cd[cd_key == key, , drop = FALSE]
    if (nrow(w) == 0) {
      return(Spectra::applyProcessing(s[FALSE]))
    }
    s <- Spectra::filterPeaksRanges(
      s,
      mz = as.matrix(w[, c("mzMin", "mzMax")]),
      rtime = as.matrix(w[, c("left_boundary", "right_boundary")]),
      keep = TRUE
    )
    Spectra::applyProcessing(s)
  })
  sp_consensus <- lapply(filtered_list, applyProcessing)

  sp_consensus <- concatenateSpectra(sp_consensus)
  sp_consensus <- filterEmptySpectra(sp_consensus)
  save(sp_consensus, file = file.path(base_path, "object/sp_consensus.RData"))
} else {
  load(file.path(base_path, "object/sp_consensus.RData"))
}
```

## Create Total Ion Current (TIC) - Consensus Signal

Same `Chromatograms()` pattern used for `tic_full` / `tic_detect` in
`1_full_detected_objects.qmd`, applied to `sp_consensus`.

```{r build-tic-consensus}
if (!file.exists(file.path(base_path, "object/tic_consensus.RData"))) {
  tic_consensus <- Chromatograms(sp_consensus,
                                 summarize.method = "sum",
                                 factorize.by     = "dataOrigin",
                                 spectraVariables = c("mixture", "lab"))
  tic_consensus <- setBackend(tic_consensus, ChromBackendMemory())
  save(tic_consensus,
       file = file.path(base_path, "object/tic_consensus.RData"))
} else {
  load(file.path(base_path, "object/tic_consensus.RData"))
}

cat("tic_consensus:", length(tic_consensus), "chromatograms\n")
```

# Peak Quality Diagnostics

## Distributions per Lab

Per-lab density of one intensity metric (`areaUnderTic`, log2-scaled)
and four shape/quality metrics (`peakWidth`, `signalToNoiseRatio`,
`xicFwhm`, `gaussianSimilarity`), shown separately for positive and
negative polarity. Each panel has its own independent x and y range
so distributions are not squashed by metrics on a wider scale.

```{r quality-distributions, fig.width=12, fig.height=6}
#| fig-cap: "Peak-quality metric distributions per lab, by polarity. Each panel has its own x/y range. Filtered consensus peaks only."
keep_metrics <- c("log2(areaUnderTic)", "peakWidth",
                  "signalToNoiseRatio", "xicFwhm", "gaussianSimilarity")

quality_long <- pk_filtered |>
  mutate(polarity = ifelse(polarity == 1L, "Positive", "Negative")) |>
  transmute(
    lab, polarity,
    `log2(areaUnderTic)` = log2(areaUnderTic),
    peakWidth, signalToNoiseRatio, xicFwhm,
    gaussianSimilarity   = `gaussianSimilarity.gaussian_similarity`
  ) |>
  pivot_longer(-c(lab, polarity),
               names_to = "metric", values_to = "value") |>
  filter(!is.na(value), is.finite(value)) |>
  mutate(metric = factor(metric, levels = keep_metrics))

ggplot(quality_long, aes(x = value, fill = lab, colour = lab)) +
  geom_density(alpha = 0.25, linewidth = 0.5) +
  facet_wrap(vars(polarity, metric), scales = "free", ncol = 5) +
  scale_fill_viridis_d(option = "turbo") +
  scale_colour_viridis_d(option = "turbo") +
  labs(title = "Peak-quality distributions per lab",
       subtitle = paste0("Consensus peaks retained after filter ",
                         "(n = ", n_distinct(pk_filtered$consensus_id), ")"),
       x = NULL, y = "Density") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Inter-Lab Relative Abundance Correlation

Absolute `areaUnderTic` is not directly comparable across labs —
instrument sensitivity, ionisation efficiency and injection volume
shift the whole intensity scale. To make labs comparable, for each
peak we compute a log2 fold change of its AUC against the lab's
own median AUC, within each polarity:

`rel_abund[peak, lab] = log2( AUC[peak, lab] / median(AUC[, lab]) )`

The lab median becomes 0, a peak twice as intense as the lab median is
+1, half-median is −1, regardless of the lab's absolute scale.

For each consensus peak we then have a 4-vector
`(rel_abund_afekta, rel_abund_cembio, rel_abund_hmgu, rel_abund_icl)`.
Stacking those vectors across all consensus peaks gives, per lab pair,
two columns of relative abundances over the same set of peaks.
**Pearson r between those two columns is the inter-lab relative
abundance correlation** — it asks whether peaks that are above-median
in lab A are also above-median in lab B.

Polarities are kept separate because positive and negative AUCs are
on different scales and would otherwise need their own normalisation.

```{r abundance-correlation}
LABS <- c("afekta", "cembio", "hmgu", "icl")

abund_wide <- pk_filtered |>
  mutate(polarity = ifelse(polarity == 1L, "Positive", "Negative")) |>
  filter(!is.na(areaUnderTic), areaUnderTic > 0) |>
  group_by(lab, polarity) |>
  mutate(rel_abund = log2(areaUnderTic / median(areaUnderTic))) |>
  ungroup() |>
  select(consensus_id, polarity, lab, rel_abund) |>
  group_by(consensus_id, polarity) |>
  filter(n() == 4) |>
  ungroup() |>
  pivot_wider(names_from = lab, values_from = rel_abund) |>
  drop_na(all_of(LABS))

lab_pairs <- combn(LABS, 2, simplify = FALSE)

pair_corrs <- do.call(rbind, lapply(c("Positive", "Negative"), function(pol) {
  sub <- abund_wide[abund_wide$polarity == pol, ]
  do.call(rbind, lapply(lab_pairs, function(p) {
    v <- sub[, p]
    data.frame(
      polarity      = pol,
      pair          = paste(p, collapse = ":"),
      same_platform = identical(sort(p), c("afekta", "cembio")),
      r             = cor(v[[1]], v[[2]]),
      n             = nrow(v)
    )
  }))
}))
pair_corrs <- pair_corrs[order(pair_corrs$polarity, -pair_corrs$r), ]

knitr::kable(transform(pair_corrs, r = round(r, 3)),
  caption = "Inter-lab Pearson r on log2(AUC / lab median), by polarity. Consensus peaks present in all 4 labs.")
```

```{r abundance-scatter, fig.width=11, fig.height=8}
#| fig-cap: "Inter-lab log2(AUC / lab median) per lab pair. Red dashed = identity; black = OLS fit. Same-platform afekta:cembio highlighted."
abund_pairs <- do.call(rbind, lapply(lab_pairs, function(p) {
  data.frame(
    consensus_id  = abund_wide$consensus_id,
    polarity      = abund_wide$polarity,
    x             = abund_wide[[p[1]]],
    y             = abund_wide[[p[2]]],
    pair          = paste(p, collapse = ":"),
    same_platform = identical(sort(p), c("afekta", "cembio"))
  )
}))

ggplot(abund_pairs, aes(x = x, y = y, colour = same_platform)) +
  geom_abline(slope = 1, intercept = 0,
              colour = "red", linetype = "dashed") +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "black", linewidth = 0.5) +
  scale_colour_manual(values = c(`FALSE` = "grey40",
                                 `TRUE`  = "#E64B35"),
                      labels = c("Cross-platform", "Same-platform"),
                      name = NULL) +
  facet_grid(polarity ~ pair, scales = "free") +
  labs(title = "Inter-lab relative abundance (log2 AUC / lab median)",
       subtitle = "Consensus peaks present in all 4 labs",
       x = "Lab 1 (log2 AUC / lab median)",
       y = "Lab 2 (log2 AUC / lab median)") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Note on EIC profile correlation

We do **not** compute inter-lab EIC profile correlation (e.g.
apex-aligned Pearson on the full chromatographic shape) in this
document. Earlier exploration showed that profile correlation between
matched consensus peaks is high in absolute terms, but it is **equally
high against random peaks** drawn from the same sample (AUROC ≈ 0.5
against an RT-gap-controlled null). A score that does not separate
matched from non-matched peaks cannot be used for correspondence,
quality assessment, or any downstream filter — so the metric is
omitted here.

# Next Steps

1. Peak metrics saved to `object/peak_metrics_detected_all4labs.csv`.
2. Filtered EICs saved to `object/eics_filtered_detected_all4labs.RData`.
3. Post-filter consensus Spectra saved to `object/sp_consensus.RData`.
4. Post-filter consensus TIC saved to `object/tic_consensus.RData`.
5. Downstream variance and normalisation analysis continues in
   `5_normalization.qmd` and the ICC reproducibility ladder in
   `7_reproducibility_results.qmd`.
