---
title: "Reproducibility results — data-level progression and pairwise agreement"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
        embed-resources: true
---

# Introduction

The same 19 mixtures × 4 labs × 2 polarities are observable at **four
different data levels**, each built by a different stage of the
processing pipeline:

| Level | Built by | Feature space | Per sample summary |
|---|---|---|---|
| **Full**       | mzML, no peak detection | all MS1 ions | TIC, BPC |
| **Detected**   | xcms (per-lab, per-sample peak picking) | per-injection detected peaks | TIC, BPC restricted to detected ions |
| **Annotated**  | SIRIUS MS2 + SNR-aware chromExtract | the spike-in panel (~250 (mix × compound) tuples detected in ≥1 lab) | sum/max of maxIntensity over the panel |
| **Consensus**  | `4_consensus_peaks.qmd` cross-lab feature matching | NAPS-aligned features matched in all 4 labs | sample × consensus-peak matrix |

This file asks **one question**: *does each successive layer recover
biological signal from technical variance?* The answer is yes — see the
Phase 1 variance decomposition and the data-level ICC ladder for the
headline story, and the Phase 5 Consensus-peak analysis for the per-peak picture
on the broader feature space (1.8k features vs ~250 in Annotated).

Other axes of the analysis live in dedicated files:

- **Normalization strategies + full ICC ladder + all-strategies
  pairwise ICC heatmap** → `5_normalization.qmd` §P3.1–§P3.6
- **NAPS RT alignment** → `3_rt_alignment.qmd`
- **Per-compound identity / RT / abundance / m/z diagnostics** →
  `8_per_compound_diagnostics.qmd`
- **Confidence tiering (A / B / C / D)** → `6_confidence_tier.qmd`

Data: 19 standard mixtures (HUMAN endosome) from 4 labs with harmonized
LC systems. No technical replicates; lab identity is confounded with
instrument platform.

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)
library(pheatmap)

base_path     <- here("5_downstream_analysis_manual")
producer_dir  <- here("5_downstream_analysis_manual", "object")
source(file.path(base_path, "helpers.R"))

# SIRIUS-confirmed compound set (best_structure_rank not NA -> SIRIUS returned the
# expected structure, a COSMIC score exists). The whole downstream is restricted to
# these (matching QMD 8): every read of annotation_chrom_metrics below is
# semi_joined to this set; the per-compound pairwise/Spearman use the normalized
# CSV, which QMD 5 already wrote confirmed-only. The untargeted Full/Detected/
# Consensus levels (peak_metrics) are NOT annotation-scoped and stay unfiltered.
.confirmed_compounds <- read.csv(
    file.path(producer_dir, "annotation_identity.csv"),
    stringsAsFactors = FALSE, colClasses = c(mixture = "character")) |>
  dplyr::filter(!is.na(best_structure_rank)) |>
  dplyr::mutate(mixture = gsub("\\.", "_", mixture)) |>
  dplyr::distinct(mixture, compound)
cat("SIRIUS-confirmed compounds (QMD 7 scope):", nrow(.confirmed_compounds), "\n")
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`, `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) %>%
  distinct() %>%
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) %>%
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

Each lab's protocol is unique; lab identity is completely confounded with
instrumental setup. Notable variations:

- MS1 scan rate
- Injection volume
- Solvent brand.

## Metrics Definition

Three main chromatographic metrics:

- **TIC**: Sum of all ion intensities
- **BPC**: Max intensity per scan

Three data levels (least → most processed):

- **Full Signal**: raw profile including baseline (`bpc_full` / `tic_full`)
- **Detected Signal**: peaks from centWave; isolates compound signal from
  baseline (`bpc_detect` / `tic_detect`)
- **Annotated levels** (both shown side by side everywhere below): derived
  from `annotation_chrom_metrics.csv` (EIC re-extraction over the
  SIRIUS-confirmed curated peaks), restricted to `(mixture, compound)` with
  non-NA `maxIntensity` in **all 4 labs** at that polarity (recomputed each
  render). Per-sample bulk: `tic = sum(areaUnderTic)`; `bpc = max(maxIntensity)`.
  Two variants, by `source`:
  - **Annotated (MS2)** — `source = "own"` only: the lab itself MS2-curated
    the compound. The strict "this lab independently identified it" signal.
  - **Annotated + MS1** — own **plus** MS1-rescued (`source = "consensus"`:
    the lab did *not* curate it but its EIC at the cross-lab consensus RT
    shows a peak). The rescued half carries no MS2 confirmation (only a
    co-eluting MS1 peak), so it is noisier — comparing the two shows how much
    the rescue inflates coverage and how it shifts agreement. The split is
    very uneven by lab: afekta / cembio / hmgu are mostly own MS2, whereas
    **icl (Waters)** curated few compounds, so its "+MS1" leans heavily on
    rescue (requiring 4-lab MS2 makes the MS2-only set much smaller, icl
    especially).

Two polarities: Positive and Negative ionization modes.

```{r load-data-both-modes}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "object/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]
    load(file.path(base_path, "object/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "object/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]
    load(file.path(base_path, "object/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = ifelse(data_level == "full", "Full Signal", "Detected Signal")) %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm = bpc / median_bpc,
      tic_norm = tic / median_tic
    ) %>%
    ungroup()
}

compute_metrics_consensus <- function(polarity, polarity_name) {
  # Consensus Signal — TIC of the cross-lab matched feature signal per
  # (lab, mixture). Loaded from tic_consensus.RData (built in
  # 4_consensus_peaks.qmd / 1_full_detected_objects.qmd). BPC here is
  # derived as the per-chromatogram max-intensity (parallel to TIC = sum).
  load(file.path(base_path, "object/tic_consensus.RData"))
  tic <- tic_consensus[tic_consensus$polarity == polarity]
  ints <- intensity(tic)
  data.frame(
    lab = tic$lab,
    mixture = tic$mixture,
    polarity = polarity_name,
    bpc = vapply(ints, max, numeric(1), na.rm = TRUE),
    tic = vapply(ints, sum, numeric(1), na.rm = TRUE)
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = "Consensus Signal") %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm   = bpc / median_bpc,
      tic_norm   = tic / median_tic
    ) %>%
    ungroup()
}

compute_metrics_annotated <- function(polarity, polarity_name, src = c("all", "own")) {
  src <- match.arg(src)
  pol_str <- if (polarity == 1L) "pos" else "neg"

  # mixture forced to character — read.csv would otherwise coerce quoted
  # "2.10" → 2.1 (numeric), silently colliding with mixture "2.1".
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))
  chrom$mixture <- gsub("\\.", "_", chrom$mixture)
  chrom <- chrom %>% dplyr::semi_join(.confirmed_compounds, by = c("mixture", "compound"))
  # "own"  = MS2-identified peaks only (the lab curated the compound).
  # "all"  = own + MS1-rescued (source == "consensus": the lab's EIC at the
  #          cross-lab consensus RT, even when it did not curate the compound).
  if (src == "own") chrom <- chrom %>% filter(source == "own")

  well_detected <- chrom %>%
    filter(polarity == pol_str, !is.na(maxIntensity)) %>%
    distinct(mixture, compound, lab) %>%
    count(mixture, compound) %>%
    filter(n == 4L) %>%
    select(mixture, compound)

  bulk <- chrom %>%
    filter(polarity == pol_str, !is.na(maxIntensity)) %>%
    inner_join(well_detected, by = c("mixture", "compound")) %>%
    group_by(lab, mixture, compound) %>%
    slice_max(maxIntensity, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    group_by(lab, mixture) %>%
    summarise(bpc = max(maxIntensity, na.rm = TRUE),
              tic = sum(areaUnderTic, na.rm = TRUE),
              .groups = "drop")

  data.frame(
    lab = bulk$lab, mixture = bulk$mixture,
    polarity = polarity_name,
    bpc = bulk$bpc, tic = bulk$tic
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = if (src == "own") "Annotated (MS2)" else "Annotated + MS1") %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm   = bpc / median_bpc,
      tic_norm   = tic / median_tic
    ) %>%
    ungroup()
}

metrics_full      <- run_for_polarities(compute_metrics, "full")
metrics_detect    <- run_for_polarities(compute_metrics, "detect")
metrics_annotated_ms2 <- run_for_polarities_names(function(p, n) compute_metrics_annotated(p, n, "own"))
metrics_annotated_all <- run_for_polarities_names(function(p, n) compute_metrics_annotated(p, n, "all"))
metrics_consensus <- run_for_polarities_names(compute_metrics_consensus)

# Recompute the "275-set" size (per polarity) for the prose / sanity check.
.well_n <- function(p) {
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))
  chrom$mixture <- gsub("\\.", "_", chrom$mixture)
  chrom <- chrom %>% dplyr::semi_join(.confirmed_compounds, by = c("mixture", "compound"))
  chrom %>% filter(polarity == p, !is.na(maxIntensity)) %>%
    distinct(mixture, compound, lab) %>%
    count(mixture, compound) %>% filter(n == 4L) %>% nrow()
}
cat("well-detected (mixture, compound) pairs — pos:", .well_n("pos"),
    " | neg:", .well_n("neg"), "\n")

metrics_combined <- bind_rows(metrics_full, metrics_detect, metrics_consensus,
                              metrics_annotated_ms2, metrics_annotated_all) %>%
  mutate(data_level = factor(data_level,
                             # Display order = Full -> Detected -> Consensus ->
                             # Annotated (MS2) -> Annotated + MS1. The untargeted
                             # TIC levels appear first; the two annotated levels
                             # (MS2-only vs MS2 + MS1-rescue) are kept rightmost.
                             levels = c("Full Signal",
                                        "Detected Signal",
                                        "Consensus Signal",
                                        "Annotated (MS2)",
                                        "Annotated + MS1"))) %>%
  group_by(lab, polarity, data_level) %>%
  mutate(
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE)
  ) %>%
  ungroup()

# Reusable ordered character vector for the few loops that iterate levels.
# Order: Full -> Detected -> Consensus -> Annotated (Annotated rightmost).
data_levels_all <- c("Full Signal", "Detected Signal", "Consensus Signal",
                     "Annotated (MS2)", "Annotated + MS1")
```

---

# Phase 1: The Problem — Raw Data Assessment {#phase1}

This phase characterizes the raw data to understand the magnitude of
inter-laboratory variability before any corrections are applied.

## 1.1 Distribution of Raw Metrics

### 1.1.1 Violin Plots: Overall Intensity Distribution

```{r violin-plots}
#| fig-cap: "BPC and TIC distributions by lab, polarity, and data level (Full → Detected → Consensus → Annotated, rightmost)."
#| fig-width: 18
#| fig-height: 7
for (metric in c("bpc", "tic")) print(plot_violin(metrics_combined, metric))
```

Detected signal recenters the negative and positive distributions.
The bimodal shape is evident.

### 1.1.2 Density Plots: Distribution Shape by Lab

```{r density-plots}
ggplot(metrics_combined %>% filter(data_level == "Full Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_bw() +
  labs(title = "TIC Distribution by Lab (Full Signal)",
       x = "log2(TIC)", y = "Density")

ggplot(metrics_combined %>% filter(data_level == "Detected Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_bw() +
  labs(title = "TIC Distribution by Lab (Detected Signal)",
       x = "log2(TIC)", y = "Density")
```

Full signal is more compressed than detected signal, likely because
each mixture has different numbers of compounds.

### 1.1.3 Mixture-Level Agreement Across Labs

```{r mixture-distribution}
#| fig-width: 14
#| fig-height: 7
plot_mixture_distribution <- function(data, metric) {
  if (metric %in% c("bpc", "tic")) {
    data <- data |> mutate(across(all_of(metric), log2))
    title_suffix <- "(log2)"
  } else {
    title_suffix <- ""
  }

  ggplot(data, aes(x = mixture, y = .data[[metric]], color = lab, group = lab)) +
    geom_line(alpha = 0.7) +
    facet_grid(data_level ~ polarity) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(title = paste("Mixture Distribution Across Labs -", metric, title_suffix),
         x = "Mixture", y = paste(metric, title_suffix), color = "Laboratory")
}

for (metric in c("bpc", "tic")) {
  print(plot_mixture_distribution(metrics_combined, metric))
}
```

There is good agreement between Cembio and Afekta, and some with ICL.
However, HMGU is consistently lower intensity.

## 1.2 Variance Decomposition: Quantifying the Problem

### 1.2.1 Linear Model Analysis

Let's run linear models to evaluate mixture effects and batch effects.
The goal is to quantify how much of the variance is due to lab vs mixture.
We use ANOVA to aggregate coefficients into Sum of Squares (SS), which
quantifies the relative power of each factor.

We run this analysis across all 4 conditions (positive/negative, full/detected).

**The model** (`helpers.R::check_metric_quality`): for each (polarity × data
level) we fit

> `lm(log2(tic) ~ lab + mixture)`

and partition the ANOVA sum-of-squares into **Lab %**, **Mixture %**, and
**Residual %**. The `log2` makes the multiplicative cross-lab intensity scale
additive. The goal is a *low* Lab % and *high* Mixture % — biology (mixture)
should explain the signal, not the instrument (lab). Note this is an ordinary
fixed-effects `lm` and the percentages come from the ANOVA SS table; the ICC
ladder in §1.2.4 instead uses an explicit variance-component ratio (between-
mixture / total) on the same `log2` bulk values.

```{r variance-decomposition-lm}
results_table <- data.frame()
residual_plots <- list()
all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in data_levels_all) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")

    residual_plots[[paste(pol, level, sep = " - ")]] <- result$residual_plot
    all_coefficients <- rbind(all_coefficients,
      result$coefficients %>% mutate(Polarity = pol, Data_Level = level))
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

### 1.2.2 Lab Effect Coefficients

```{r lab-coefficients}
#| fig-cap: "Lab effect coefficients from linear models. Deviation from reference lab (Afekta); negative = lower intensity."
lab_coefs <- all_coefficients %>%
  filter(type == "Lab Effect") %>%
  mutate(lab_name = gsub("lab", "", coefficient))

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_hue() +
  labs(title = "Lab Effect Coefficients from Linear Models",
       subtitle = "Deviation from reference lab (afekta); negative = lower intensity",
       x = "Laboratory", y = "Coefficient (log2 scale)", fill = "Laboratory") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.2.3 Variance Decomposition Summary

```{r variance-stacked-bar}
#| fig-cap: "Variance decomposition of TIC across all four data levels (Full → Detected → Consensus → Annotated) and both polarities. Lab variance dominates in raw data; subsequent steps progressively shift variance away from lab towards mixture and residual."
#| fig-width: 14
#| fig-height: 7
variance_long <- results_table %>%
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) %>%
  pivot_longer(cols = ends_with("_Pct"), names_to = "Source", values_to = "Percentage") %>%
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Data_Level = factor(Data_Level, levels = data_levels_all),
    R2_label = paste("R² =", round(R_squared, 3))
  )

text_labels <- variance_long %>% distinct(Polarity, Data_Level, R2_label)

ggplot(variance_long,
       aes(x = Data_Level, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.85, width = 0.7) +
  geom_text(aes(x = Data_Level, label = R2_label, y = 107), data = text_labels,
            size = 3.6, inherit.aes = FALSE) +
  facet_wrap(~ Polarity, nrow = 1) +
  labs(title = "Variance decomposition: TIC across all four data levels",
       subtitle = "Raw data shows massive lab variance — subsequent steps shift variance away from lab",
       x = NULL, y = "Variance contribution (%)", fill = "Source") +
  theme_bw(base_size = 13) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylim(0, 115)
```

### 1.2.4 ICC across the four data levels

The variance decomposition above splits total variance into Lab,
Mixture, and Residual components — informative but not directly
comparable to the per-strategy bulk ICC reported in §P3.5 of
`5_normalization.qmd`. Here we compute the **same ICC statistic**
(variance-component ratio on log2-bulk TIC) at each data level so the
progression Full → Detected → Consensus → Annotated is on one axis.

```{r icc-across-data-levels}
#| fig-cap: "Reproducibility ICC of bulk TIC across the four data levels. Higher = labs agree on a mixture's value as tightly as the mixtures themselves differ. Computed with the same variance-component formula used by 5_normalization.qmd §P3.5."
#| fig-height: 4
icc_one <- function(df) {
  if (!nrow(df) || all(is.na(df$tic))) return(NA_real_)
  df$lv <- log2(pmax(df$tic, 1e-9))
  between <- var(tapply(df$lv, df$mixture, mean,
                         na.rm = TRUE), na.rm = TRUE)
  within  <- mean(tapply(df$lv, df$mixture, var,
                          na.rm = TRUE), na.rm = TRUE)
  if (!is.finite(between + within) || (between + within) == 0)
    return(NA_real_)
  between / (between + within)
}

icc_by_level <- metrics_combined |>
  filter(!is.na(tic), tic > 0) |>
  group_by(polarity, data_level) |>
  group_modify(~ tibble(ICC = icc_one(.x))) |>
  ungroup() |>
  mutate(data_level = factor(data_level, levels = data_levels_all))

ggplot(icc_by_level, aes(x = data_level, y = ICC, fill = data_level)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f", ICC)),
            vjust = -0.4, size = 4) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  facet_wrap(~ polarity) +
  scale_fill_brewer(palette = "Set2") +
  ylim(0, 1.05) +
  labs(title = "Reproducibility (ICC) by data level",
       subtitle = "Bulk TIC, variance-component ICC. Red = ICC 0.5 moderate-reproducibility landmark.",
       x = NULL, y = "ICC") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))

knitr::kable(icc_by_level |>
               mutate(ICC = round(ICC, 3)) |>
               tidyr::pivot_wider(names_from = polarity,
                                   values_from = ICC),
             caption = "ICC of bulk TIC across the four data levels.")
```

### 1.2.5 Residual Diagnostics

```{r residual-plots}
#| fig-height: 10
residual_data <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in data_levels_all) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    residual_data <- rbind(residual_data, data.frame(
      Fitted = fitted(result$model), Residuals = residuals(result$model),
      Polarity = pol, Data_Level = level,
      Lab = data_subset$lab, Mixture = as.factor(data_subset$mixture)
    ))
  }
}

ggplot(residual_data %>%
         mutate(Data_Level = factor(Data_Level, levels = data_levels_all)),
       aes(x = Fitted, y = Residuals, color = Mixture, shape = Lab)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_color_hue() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(title = "Fitted vs Residuals: TIC Across All Conditions",
       subtitle = "Points should scatter randomly around zero for a good model fit",
       x = "Fitted Values", y = "Residuals", color = "Mixture", shape = "Laboratory") +
  theme_bw() +
  theme(legend.position = "bottom")

results_table
```

**Key Finding:** Overall variance is dominated by lab effects.
The residual plot clearly shows separation between labs vs HMGU and
still lab separation in others. Lab separation remains unexplained by
mixture. We cannot compare labs directly without normalization.

---

# Phase 5: Peak Feature Analysis — Consensus Peaks {#phase5}

This phase extends the global analysis by incorporating peak-level metrics
computed from EICs of consensus peaks detected across all 4 laboratories.
This provides a more refined view of inter-lab reproducibility at the
individual peak level.

**Scope note.** Phase 5 reads `peak_metrics_detected_all4labs.csv` from
`4_consensus_peaks.qmd`, which matches **xcms-detected** peaks across
labs without any spike-in identity. It is independent of the SIRIUS
producer cluster-RT algorithm — re-running the producer does **not**
require re-rendering `4_consensus_peaks.qmd`. Phase 5 answers a
different question: "for the same chromatographic feature detected in
all 4 labs, how reproducible is the peak shape / intensity?" — the
flip side of the Annotated row, where Phase 5 measures *detection-level*
agreement and the Annotated row measures *identity-level* abundance.

**Matching basis.** The peak_metrics table is produced by
`4_consensus_peaks.qmd` under **NAPS-aligned matching with raw-RT
fallback** (20 ppm / 15 s), which finds 3 490 four-lab features; after
the peak-quality filters here, 1 787 four-lab consensus peaks remain.
This is the detection-level (xcms peak shape/intensity) view; the
identity-level (SIRIUS-annotated, normalized) ICC ladder is reported
in `5_normalization.qmd` §P3.5–§P3.6.

## 5.1 Data Overview

The peak feature data comes from `4_consensus_peaks.qmd`, which:

1. Created a consensus peak table matching peaks across labs (mz tol = 20 ppm, rt tol = 15 s, on NAPS-aligned RT)
2. Filtered to peaks detected in all 4 labs (1787 consensus peaks after QC; 7148 lab rows)
3. Extracted EICs and computed peak metrics (FWHM, area, SNR, etc.)
4. Applied quality filters (prominence, peak width, FWHM validity)

```{r load-peak-metrics}
pk_metrics <- read.csv(file.path(base_path, "object/peak_metrics_detected_all4labs.csv"))

cat("Peak metrics dimensions:", nrow(pk_metrics), "rows x", ncol(pk_metrics), "columns\n")
cat("Unique consensus peaks:", length(unique(pk_metrics$consensus_id)), "\n")
cat("Labs represented:", paste(unique(pk_metrics$lab), collapse = ", "), "\n")
```

## 5.2 Spearman Intensity Rank Correlation

We use Spearman (rank-based) correlation to test whether peaks that are
relatively intense in one lab are also relatively intense in another.
Spearman is robust to the non-linear scale differences across labs without
requiring explicit normalization — it compares rank order, not absolute
values.

### 5.2.1 Intensity Correlation Matrix

```{r peak-metrics-correlation}
# Pivot to wide format for correlation
intensity_wide <- pk_metrics |>
  select(consensus_id, lab, areaUnderTic, maxIntensity, mixture, polarity) |>
  pivot_wider(
    id_cols = c(consensus_id, mixture, polarity),
    names_from = lab,
    values_from = c(areaUnderTic, maxIntensity)
  )

# Spearman correlation for areaUnderTic
area_cols <- paste0("areaUnderTic_", c("afekta", "hmgu", "icl", "cembio"))
area_matrix <- intensity_wide[, area_cols]
names(area_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_area <- cor(area_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("Spearman Correlation - Area Under TIC:\n")
print(round(cor_area, 3))

# Spearman correlation for maxIntensity
max_cols <- paste0("maxIntensity_", c("afekta", "hmgu", "icl", "cembio"))
max_matrix <- intensity_wide[, max_cols]
names(max_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_max <- cor(max_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("\nSpearman Correlation - Max Intensity:\n")
print(round(cor_max, 3))
```

### 5.2.2 Correlation Heatmap Visualization

```{r peak-correlation-heatmap}
#| fig-cap: "Spearman rank correlation heatmaps for recalculated peak intensities across 4 labs."
#| fig-height: 5
library(pheatmap)

pheatmap(cor_area,
         main = "Area Under TIC (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)

pheatmap(cor_max,
         main = "Max Intensity (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)
```

Afekta and Cembio show the strongest rank correlation (ρ ≈ 0.76),
consistent with their shared Agilent Q-TOF platform. All other pairs
are moderate (ρ ≈ 0.4–0.5), indicating that relative peak intensity
rankings are partially conserved across labs but with substantial
instrument-specific variation.

## 5.3 Peak Metric Distributions Across Labs

The following metrics are computed for each EIC from the filtered consensus
peaks:

- **Peak Width**: Time span (seconds) between the detected peak boundaries.
  Reflects chromatographic resolution and column performance.
- **FWHM** (Full Width at Half Maximum): Width of the peak at half its
  maximum intensity. A narrower, more symmetric measure of peak shape than
  peak width.
- **Signal-to-Noise Ratio (SNR)**: Ratio of peak maximum intensity to the
  estimated baseline. Higher SNR indicates a cleaner, more confident
  detection.

### 5.3.1 Peak Width Distribution

```{r peak-width-distribution}
ggplot(pk_metrics |> filter(!is.na(peakWidth)),
       aes(x = lab, y = peakWidth, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Peak Width Distribution by Lab",
   subtitle = "Consensus peaks detected in all 4 labs",
   x = "Laboratory",
   y = "Peak Width (seconds)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

### 5.3.2 FWHM Distribution

```{r fwhm-distribution}
ggplot(pk_metrics |> filter(!is.na(xicFwhm)), aes(x = lab, y = xicFwhm, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "FWHM Distribution by Lab",
   subtitle = "Full Width at Half Maximum of detected peaks",
   x = "Laboratory",
   y = "FWHM (seconds)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

### 5.3.3 Signal-to-Noise Ratio Distribution

```{r snr-distribution}
ggplot(pk_metrics |> filter(!is.na(signalToNoiseRatio)),
       aes(x = lab, y = log10(signalToNoiseRatio + 1), fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Signal-to-Noise Ratio Distribution by Lab",
   subtitle = "log10(SNR + 1) scale",
   x = "Laboratory",
   y = "log10(SNR + 1)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

## 5.4 Inter-Lab Variability for Consensus Peaks

For each consensus peak (detected in all 4 labs), we compute the coefficient
of variation (CV = standard deviation / mean) of each metric across the 4
labs. CV is inherently scale-normalized: it measures relative spread
regardless of absolute intensity differences, so no prior normalization of
intensity metrics is needed.

- **CV < 0.3 (30%)**: Good reproducibility
- **CV 0.3–0.5**: Moderate variability
- **CV > 0.5**: High variability

**IQR** (Interquartile Range): the difference between the 75th and 25th
percentiles of the CV distribution, measuring the spread of the middle 50%
of values.

### 5.4.1 CV Summary Table

```{r cv-summary-table}
cv_by_consensus <- pk_metrics |>
  group_by(consensus_id) |>
  summarise(
    n_labs = n(),
    cv_area = sd(areaUnderTic, na.rm = TRUE) / mean(areaUnderTic, na.rm = TRUE),
    cv_max = sd(maxIntensity, na.rm = TRUE) / mean(maxIntensity, na.rm = TRUE),
    cv_fwhm = sd(xicFwhm, na.rm = TRUE) / mean(xicFwhm, na.rm = TRUE),
    cv_peakwidth = sd(peakWidth, na.rm = TRUE) / mean(peakWidth, na.rm = TRUE),
    cv_snr = sd(signalToNoiseRatio, na.rm = TRUE) / mean(signalToNoiseRatio, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_labs == 4)

cv_summary <- cv_by_consensus |>
  summarise(
    across(starts_with("cv_"), list(
      median = ~median(.x, na.rm = TRUE),
      q25 = ~quantile(.x, 0.25, na.rm = TRUE),
      q75 = ~quantile(.x, 0.75, na.rm = TRUE)
    ))
  )

cv_table <- data.frame(
  Metric = c("Area Under TIC", "Max Intensity", "FWHM", "Peak Width", "SNR"),
  Median_CV = c(cv_summary$cv_area_median, cv_summary$cv_max_median,
                cv_summary$cv_fwhm_median, cv_summary$cv_peakwidth_median,
                cv_summary$cv_snr_median),
  IQR = c(cv_summary$cv_area_q75 - cv_summary$cv_area_q25,
          cv_summary$cv_max_q75 - cv_summary$cv_max_q25,
          cv_summary$cv_fwhm_q75 - cv_summary$cv_fwhm_q25,
          cv_summary$cv_peakwidth_q75 - cv_summary$cv_peakwidth_q25,
          cv_summary$cv_snr_q75 - cv_summary$cv_snr_q25)
)

knitr::kable(cv_table,
             caption = "Inter-Lab CV for Consensus Peaks (median and interquartile range)",
             digits = 3)
```

### 5.4.2 CV Distribution Visualization

```{r cv-distribution}
#| fig-cap: "CV distribution across consensus peaks for each metric. Red dashed line = 30% CV threshold."
cv_long <- cv_by_consensus |>
 pivot_longer(
   cols = starts_with("cv_"),
   names_to = "metric",
   names_prefix = "cv_",
   values_to = "cv"
 ) |>
 mutate(metric = case_when(
   metric == "area" ~ "Area Under TIC",
   metric == "max" ~ "Max Intensity",
   metric == "fwhm" ~ "FWHM",
   metric == "peakwidth" ~ "Peak Width",
   metric == "snr" ~ "SNR"
 ))

ggplot(cv_long |> filter(!is.na(cv) & cv < 2), aes(x = metric, y = cv, fill = metric)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 geom_hline(yintercept = 0.3, linetype = "dashed", color = "red") +
 annotate("text", x = 5.3, y = 0.35, label = "CV = 30%", color = "red", size = 3) +
 labs(
   title = "Inter-Lab Variability for Consensus Peaks",
   subtitle = "Lower CV = better reproducibility across labs",
   x = "Metric",
   y = "Coefficient of Variation"
 ) +
 theme_bw() +
 theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

Shape metrics (FWHM, peak width) show lower CV than intensity metrics
(area, max intensity), confirming that chromatographic behavior is more
reproducible across labs than ionization response. SNR shows the highest
variability, reflecting differences in baseline estimation across
instruments.

---

# Summary {#summary}

This file shows the **data-level progression** story:

1. **Lab variance dominates raw data**. At the Full Signal level,
   97–100 % of TIC variance is lab-attributable (variance decomposition,
   §1.2.3) and the data-level ICC (§1.2.4) is near zero.
2. **Peak detection recovers some biology**. Detected Signal shifts
   variance away from lab toward mixture; xcms peak picking strips most
   of the lab-specific background and the ICC rises.
3. **Annotation concentrates the biology further**. Both annotated levels
   — Annotated (MS2) and Annotated + MS1 — tighten cross-lab agreement
   (higher ICC, lower within-mixture variance); SIRIUS-keyed features
   carry identity-aware abundance.
4. **Consensus peaks deliver the cleanest picture**. The cross-lab
   matched peak set (~1.8 k features) is the broadest reproducible space;
   the per-peak inter-lab CV distribution shows it behaves like the
   spike-in panel.

For the **normalization-strategy evaluation** (Raw / Aligned / Lab-
median / Global / Decile / LOESS with the full ICC ladder × pairwise
heatmap), see `5_normalization.qmd` §P3.4–§P3.6. For the **RT
alignment**, see `3_rt_alignment.qmd`. For the **per-compound
diagnostics**, see `8_per_compound_diagnostics.qmd`.

---

# Pairwise reproducibility on the winning normalization

The full strategy comparison (six normalizations × both metrics × both
compound-set policies) lives in `5_normalization.qmd` §P3.5–§P3.6.
Below: the same pairwise structure, **restricted to Raw vs Lab-median
only** (the §P3.4 winner for Annotated Signal), and computed on
**`areaUnderTic`** (the AUC — integrated peak area) which is the
better-matched bulk metric for a presentation summary. Two views:

1. **Bulk-level pairwise ICC** — per (lab pair × strategy), how well
   do the two labs agree on each mixture's total annotated AUC?
2. **Per-compound pairwise Spearman ρ** — same lab pairs, but on
   *individual* (mixture × compound) AUC values rather than summed
   bulk.

**Reading the ICC.** It's pairwise (only 2 labs at a time) and pools
*across the 19 mixtures* in a specific way: for each of the 19
mixtures we have 2 values (one per lab), so the comparison is
**mix 1_1 lab A vs mix 1_1 lab B, then mix 1_2 lab A vs mix 1_2 lab
B**, etc. — same mixture in each pair, never across-mixture mismatch.
`compute_icc()` then splits the spread of all 38 values into

- **between-mixture** = how different the per-mixture *averages* are
  from each other across the 19 mixtures (the biological signal), and
- **within-mixture** = how much the 2 labs disagree on the same
  mixture, averaged across mixtures (cross-lab noise).

ICC = between / (between + within). Mix 1_1 is only ever compared
with mix 1_1 — the formula never pools different mixtures into one
bucket.

**Reading the Spearman.** Spearman ρ is the rank correlation of
per-(mixture × compound) AUC between two labs. **It is rank-based, so
multiplying everything in one lab by a constant doesn't change it** —
that's why we don't normalize here. The number answers: *"if lab A
ranks compound X above compound Y, does lab B agree?"* ρ → 1 means
labs rank the spike-ins the same way; ρ near 0 means compound-level
rankings disagree.

What Spearman ρ does **not** tell you is *which compounds a lab
favours over another*. That's a per-compound bias question
(e.g. log₂(labA / labB) per compound), not a single rank-correlation
number per lab pair. The values here summarise overall ranking
agreement, no more.

**Why afekta + cembio (★) agree best — and how to read this.** afekta and
cembio are the **only same-platform pair**: both run an Agilent 1290 UHPLC into
an Agilent Q-ToF (6546 / 6545), while icl is Waters (Synapt) and hmgu is Sciex
(ZenoTOF). So afekta↔cembio carries the *least* cross-platform technical variance
and **should** sit at the top — it is the **ceiling reference (★)**, a positive
control, not a suspicious result. The informative contrast is how far the
cross-platform pairs (anything vs icl or hmgu) fall *below* that ceiling.

**Caveats — read these as a coarse summary, not a definitive statistic.**
(1) Only **19 mixtures**, so each pairwise ICC is estimated from 19 paired
points — small N, wide intervals. (2) It is **pairwise** (2 labs at a time), not
a joint 4-lab model. (3) Both annotated signals are shown as separate rows —
**Annotated (MS2)** (own peaks only) and **Annotated + MS1** (own + rescue,
which is noisier); compare the two to see how MS1 rescue shifts agreement.
(4) Bulk ICC sums each mixture to one number, hiding
per-compound disagreement. Treat the ★ pair as the achievable ceiling and the
cross-platform pairs as the realistic floor; for per-compound detail prefer the
Spearman panel and `8_per_compound_diagnostics.qmd`.

```{r reprod-load-normalized}
#| message: false
source(here("5_downstream_analysis_manual", "helpers.R"))

producer_dir <- here("5_downstream_analysis_manual", "object")
ann_norm <- read.csv(
  file.path(producer_dir, "annotation_chrom_metrics_normalized.csv"),
  stringsAsFactors = FALSE,
  colClasses = c(mixture = "character"))
ann_norm$mixture <- gsub("\\.", "_", ann_norm$mixture)

# Strict subset (4/4 labs, all aligned), built per SIGNAL source so every
# pairwise plot below shows both: "Annotated (MS2)" = own (MS2-curated) peaks only,
# "Annotated + MS1" = own + MS1-rescue (source == "consensus").
.sig_levels <- c("Annotated (MS2)", "Annotated + MS1")
build_well_aligned_w <- function(src) {
  x <- if (src == "own") ann_norm |> filter(source == "own") else ann_norm
  wd <- x |> filter(!is.na(maxIntensity), maxIntensity > 0) |>
    distinct(mixture, lab, compound, polarity) |>
    group_by(mixture, compound, polarity) |> filter(n_distinct(lab) == 4L) |>
    ungroup() |> distinct(mixture, compound, polarity)
  aw <- x |> inner_join(wd, by = c("mixture", "compound", "polarity")) |>
    group_by(lab, mixture, polarity, compound) |>
    slice_max(maxIntensity, n = 1, with_ties = FALSE) |> ungroup()
  al <- aw |> filter(!is.na(aligned_rt)) |>
    count(mixture, compound, polarity, name = "n") |> filter(n == 4L) |>
    select(mixture, compound, polarity)
  aw |> inner_join(al, by = c("mixture", "compound", "polarity")) |>
    mutate(signal = if (src == "own") "Annotated (MS2)" else "Annotated + MS1")
}
ann_well_aligned_w <- bind_rows(build_well_aligned_w("own"),
                                build_well_aligned_w("all")) |>
  mutate(signal = factor(signal, levels = .sig_levels))
well_aligned_w <- ann_well_aligned_w |> distinct(signal, mixture, compound, polarity)

cat("4-lab aligned (mixture, compound, polarity) per signal:\n")
print(well_aligned_w |> count(signal, name = "n_tuples"))
```

```{r reprod-pairwise-bulk-raw-vs-labmed}
#| fig-cap: "Pairwise ICC on each lab pair's per-mixture summed AUC (areaUnderTic). Rows = the two annotated signals (MS2-only vs MS2 + MS1-rescue); columns = polarity; x = Raw vs Lab-median bulk. ★ = same-platform pair (afekta + cembio), the ceiling reference. Brighter blue = the pair agrees on each mixture's total as tightly as the mixtures differ."
#| fig-height: 8

lab_pairs_w <- list(
  c("afekta", "cembio"), c("afekta", "hmgu"), c("afekta", "icl"),
  c("cembio", "hmgu"),   c("cembio", "icl"),  c("hmgu",   "icl"))
agilent_w <- c("afekta", "cembio")

# Bulk per (lab, mixture, polarity, strategy=Raw|Lab-median).
# Bulk = sum(areaUnderTic) — integrated peak area, summed across the
# Strict compound set. The same number is used for both the same-mix
# cross-lab agreement and the across-mix biological spread.
bulk_raw <- ann_well_aligned_w |>
  group_by(signal, lab, mixture, polarity) |>
  summarise(bulk = sum(areaUnderTic, na.rm = TRUE),
            .groups = "drop") |>
  mutate(strategy = "Raw")
lab_med <- bulk_raw |>
  group_by(signal, lab, polarity) |>
  summarise(med = median(bulk, na.rm = TRUE), .groups = "drop")
bulk_labmed <- bulk_raw |>
  left_join(lab_med, by = c("signal", "lab", "polarity")) |>
  mutate(bulk = bulk / med, strategy = "Lab-median") |>
  select(-med)
bulk_two <- bind_rows(bulk_raw, bulk_labmed)

icc_pairs_w <- bind_rows(lapply(lab_pairs_w, function(p) {
  pair_label <- paste(sort(p), collapse = " + ")
  bind_rows(lapply(.sig_levels, function(sig) {
    bulk_pair <- bulk_two |> filter(lab %in% p, signal == sig)
    bind_rows(lapply(c("pos", "neg"), function(pol_str) {
      pol_name <- if (pol_str == "pos") "Positive" else "Negative"
      bind_rows(lapply(c("Raw", "Lab-median"), function(s) {
        df <- bulk_pair |> filter(polarity == pol_str, strategy == s)
        if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
        res <- compute_icc(df, "bulk")
        res$strategy      <- s
        res$polarity      <- pol_name
        res$pair          <- pair_label
        res$signal        <- sig
        res$same_platform <- all(p %in% agilent_w)
        res
      }))
    }))
  }))
}))

ggplot(icc_pairs_w |>
         mutate(strategy = factor(strategy, levels = c("Raw", "Lab-median")),
                signal = factor(signal, levels = .sig_levels),
                pair_label = ifelse(same_platform, paste(pair, "★"), pair)),
       aes(x = strategy, y = pair_label, fill = ICC)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", ICC)), size = 3.0) +
  facet_grid(signal ~ polarity) +
  scale_fill_gradient2(low = "white", mid = "#FDC086", high = "#1F77B4",
                       midpoint = 0.5, limits = c(0, 1), name = "ICC") +
  labs(x = NULL, y = "Lab pair (★ = same platform)",
       title = "Pairwise reproducibility — bulk AUC",
       subtitle = "areaUnderTic; all-4-lab set per signal") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

## Per-compound ranking agreement (Spearman)

The plot above computes pairwise ICC on **summed** bulk (one number
per mixture for each lab). Spearman ρ instead asks whether two labs
**rank the individual compounds in the same order** by AUC, across
all (mixture × compound) measurements.

```{r reprod-per-compound-pairwise}
#| fig-cap: "Spearman ρ between each lab pair on per-(mixture × compound) areaUnderTic (AUC), faceted by signal: Annotated (MS2) vs Annotated + MS1 (own + rescue). ρ → 1 means the labs rank the spike-ins the same way; ρ near 0 means the compound-level rankings disagree. Rank-based, so normalization does not change these values."
#| fig-height: 5

wide_per_compound_w <- ann_well_aligned_w |>
  select(signal, mixture, compound, polarity, lab, areaUnderTic) |>
  pivot_wider(names_from = lab, values_from = areaUnderTic,
              names_prefix = "auc_")

per_compound_pair_cor_w <- bind_rows(lapply(lab_pairs_w, function(p) {
  bind_rows(lapply(.sig_levels, function(sig) {
    bind_rows(lapply(c("pos", "neg"), function(pol_str) {
      pol_name <- if (pol_str == "pos") "Positive" else "Negative"
      df <- wide_per_compound_w |> filter(signal == sig, polarity == pol_str)
      a <- df[[paste0("auc_", p[1])]]
      b <- df[[paste0("auc_", p[2])]]
      ok <- is.finite(a) & a > 0 & is.finite(b) & b > 0
      if (sum(ok) < 5L) return(NULL)
      tibble(pair          = paste(sort(p), collapse = " + "),
             polarity      = pol_name,
             signal        = sig,
             n             = sum(ok),
             spearman      = cor(a[ok], b[ok], method = "spearman"),
             same_platform = all(p %in% agilent_w))
    }))
  }))
}))

ggplot(per_compound_pair_cor_w |>
         mutate(signal = factor(signal, levels = .sig_levels),
                pair_label = ifelse(same_platform, paste(pair, "★"), pair)),
       aes(x = polarity, y = pair_label, fill = spearman)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", spearman)),
            colour = "black", size = 3.0) +
  facet_wrap(~ signal) +
  scale_fill_gradient2(low = "#B40426", mid = "white", high = "#3B4CC0",
                       midpoint = 0.5, limits = c(0, 1), name = "Spearman ρ") +
  labs(x = NULL, y = "Lab pair (★ = same platform)",
       title = "Per-compound ranking agreement — AUC",
       subtitle = "areaUnderTic, Spearman ρ") +
  theme_bw()

knitr::kable(per_compound_pair_cor_w |>
               transmute(signal, polarity, pair,
                          Spearman = round(spearman, 3), n) |>
               arrange(signal, polarity, pair),
             caption = "Per-compound pairwise Spearman ρ (AUC), by signal.")
```

The plot above ranks each compound by its **single most-intense adduct**
(`slice_max`). Below is the same Spearman but **summing `areaUnderTic` across
all detected adducts** per `(mixture, compound, lab)` — so a compound seen as
`[M+H]+` in one lab and `[M-H]-` in another contributes its total ion current
instead of one adduct. Same all-4-labs compound set.

```{r reprod-per-compound-pairwise-sum}
#| fig-cap: "Spearman ρ per lab pair on per-(mixture × compound) AUC SUMMED across all detected adducts (compare with the best-adduct version above), faceted by signal: Annotated (MS2) vs Annotated + MS1 (own + rescue). ★ = same-platform pair."
#| fig-height: 5
sum_per_compound_w <- bind_rows(lapply(.sig_levels, function(sig) {
  x  <- if (sig == "Annotated (MS2)") ann_norm |> filter(source == "own") else ann_norm
  cs <- well_aligned_w |> filter(signal == sig) |>
    distinct(mixture, compound, polarity)
  x |> inner_join(cs, by = c("mixture", "compound", "polarity")) |>
    filter(!is.na(areaUnderTic)) |>
    group_by(mixture, compound, polarity, lab) |>
    summarise(areaUnderTic = sum(areaUnderTic, na.rm = TRUE), .groups = "drop") |>
    mutate(signal = sig)
}))
wide_sum_w <- sum_per_compound_w |>
  pivot_wider(names_from = lab, values_from = areaUnderTic, names_prefix = "auc_")

per_compound_pair_cor_sum_w <- bind_rows(lapply(lab_pairs_w, function(p) {
  bind_rows(lapply(.sig_levels, function(sig) {
    bind_rows(lapply(c("pos", "neg"), function(pol_str) {
      pol_name <- if (pol_str == "pos") "Positive" else "Negative"
      df <- wide_sum_w |> filter(signal == sig, polarity == pol_str)
      a <- df[[paste0("auc_", p[1])]]; b <- df[[paste0("auc_", p[2])]]
      ok <- is.finite(a) & a > 0 & is.finite(b) & b > 0
      if (sum(ok) < 5L) return(NULL)
      tibble(pair = paste(sort(p), collapse = " + "), polarity = pol_name,
             signal = sig, n = sum(ok),
             spearman = cor(a[ok], b[ok], method = "spearman"),
             same_platform = all(p %in% agilent_w))
    }))
  }))
}))

ggplot(per_compound_pair_cor_sum_w |>
         mutate(signal = factor(signal, levels = .sig_levels),
                pair_label = ifelse(same_platform, paste(pair, "★"), pair)),
       aes(x = polarity, y = pair_label, fill = spearman)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", spearman)), colour = "black", size = 3.0) +
  facet_wrap(~ signal) +
  scale_fill_gradient2(low = "#B40426", mid = "white", high = "#3B4CC0",
                       midpoint = 0.5, limits = c(0, 1), name = "Spearman ρ") +
  labs(x = NULL, y = "Lab pair (★ = same platform)",
       title = "Per-compound ranking agreement — AUC summed across adducts",
       subtitle = "areaUnderTic summed across adducts, Spearman ρ") +
  theme_bw()

knitr::kable(per_compound_pair_cor_sum_w |>
               transmute(signal, polarity, pair, Spearman = round(spearman, 3), n) |>
               arrange(signal, polarity, pair),
             caption = "Per-compound pairwise Spearman ρ (AUC summed across adducts), by signal.")
```

---

# Consensus-feature coverage

How the cross-lab consensus feature space (annotation-independent, built in
`4_consensus_peaks.qmd`) relates to the curated panel.

```{r consensus-n-labs}
#| fig-cap: "Consensus features by the number of labs that detected them — the cross-lab reproducibility funnel."
ct_all <- read.csv(file.path(base_path, "object", "ct_all_samples.csv"),
                   stringsAsFactors = FALSE)
knitr::kable(as.data.frame(table(`# labs detected` = ct_all$n_labs_matched)),
             col.names = c("# labs detected", "consensus features"),
             caption = "Consensus features by number of labs that detected them. The 4-lab set is QC-filtered down to 1787 (peak_metrics_detected_all4labs.csv).")
```

```{r consensus-mapping}
#| message: false
suppressPackageStartupMessages({ library(readxl); library(stringr) })
.labs4 <- c("afekta", "cembio", "hmgu", "icl")
pm_qc <- read.csv(file.path(base_path, "object", "peak_metrics_detected_all4labs.csv"),
                  stringsAsFactors = FALSE)
.cur_base <- here("3_annotation_manual", "HE", "3_Sirius_curation")
.cur <- lapply(setNames(.labs4, .labs4), function(l) {
  d <- as.data.frame(read_xlsx(file.path(.cur_base, paste0(l, "_lab_annotations_sirius_curated.xlsx"))))
  # SIRIUS-confirmed only: keep curated rows where SIRIUS confirmed the structure
  # (best_structurePerIdRank not NA -> the same definition the rest of this
  # downstream uses). Maps the consensus features to confirmed panel peaks only.
  d <- d[!is.na(suppressWarnings(as.numeric(d$best_structurePerIdRank))), ]
  unique(str_trim(unlist(str_split(as.character(d$chrom_peak_id), fixed("|"))))) })
pm_qc$is_cur <- mapply(function(l, id) as.character(id) %in% .cur[[l]], pm_qc$lab, pm_qc$chrom_peak_id)
mapped_qc <- pm_qc |> group_by(consensus_id) |> summarise(mapped = any(is_cur), .groups = "drop")
cat(sprintf("QC-passed consensus features (detected + QC'd in all 4 labs): %d\n", nrow(mapped_qc)))
cat(sprintf("  ... mapped to a SIRIUS-confirmed panel compound (by chrom_peak_id): %d (%.1f%%)\n",
            sum(mapped_qc$mapped), 100 * mean(mapped_qc$mapped)))
```

The bulk of the QC-passed consensus features are **not** spike-ins — they are
endogenous / matrix features the consortium reproducibly detects but that the
panel does not target. The curated panel maps onto a small, high-confidence
slice of that space.

---

# Future directions

- Expand the four-level progression to the other HUMAN matrices
  (FE, Others) beyond HE endosome.
- Add a "post-normalization" data level (Annotated normalized) to the
  data-level progression once the §8 normalization output is the default
  consumer input.


