---
title: "SIRIUS annotation, all MS2: the finalized per-lab strategy"
format:
  html:
    fig-width: 12
    fig-height: 8
    embed-resources: true
---

## Goal

Annotate **all MS2-bearing features** from every lab, without requiring
cross-lab consensus, and emit the two SIRIUS-side canonical CSVs consumed by
the downstream RT-alignment pipeline:

- `annotation_identity.csv` — one row per `(mixture, lab, compound,
  polarity)` with best ranks, adducts, COSMIC scores, and a
  `confidence_tier` placeholder (`"unfiltered"`; P5 overwrites).
- `feat_rts.csv` — one row per identified feature with the MS2-trigger
  RT, `ionMass`, `adduct`, `cosmic`. The downstream MS1 anchoring +
  cluster + EIC steps in
  [`5_downstream_analysis_pg/3_rt_alignment.qmd`](../../../5_downstream_analysis_pg/3_rt_alignment.qmd)
  §P2.9-§P2.10 consume this and produce `annotation_chrom_metrics.csv`.

This file is a **pure producer**: no EIC extraction, no peak picking,
no cross-lab comparison. Those moved to `3_rt_alignment.qmd` so MS1
anchoring runs before clustering. The producer is now SIRIUS only.

## The finalized strategy

Every lab shares one **grouping** step: MS2 are merged by precursor m/z within
20 ppm, then split wherever the retention-time gap exceeds 15 s, giving one
feature per real precursor peak. The precursor of the representative is the
feature's median m/z.

Labs differ only in how a feature's N raw MS2 scans become the spectra SIRIUS
annotates, and in whether an MS1 isotope pattern is attached:

| lab | representative | MS1 | why |
|---|---|---|---|
| **afekta** | richest scan **per collision energy** (~3 spectra) | yes | clean 10/20/40 eV ladder, so energy-resolved trees help; worth +22 over single-richest |
| **cembio** | **all scans** (capped at 40) | **yes, from its separate MS1 injection** | MS2-only injection with weak group-2/3 spectra, where pooling every scan rebuilds recovery (+53); MS1 adds a further +13 |
| **icl** | **all scans** (capped at 40) | yes | Waters DDA, over-split features whose richest scan is the noisiest; precursor m/z must be re-derived first |
| **hmgu** | single richest scan | yes | single-CE with ~72% single-scan features, so every representative collapses to this |

Two lab-specific pre-steps matter. **icl** must re-derive the precursor m/z with
`estimatePrecursorMz()`, because the Waters mzML value is unreliable and recovery
is otherwise zero. **hmgu** applies a precursor-abundance filter (`prec_thr` 300)
using the maximum precursor intensity across the feature's scans, not a single
median-RT scan, which previously sampled off-apex dips and dropped real compounds.

### cembio's MS1, the one non-obvious case

cembio's MS2 injection carries no usable isotope pattern, so it was long run
without MS1. Its MS1 scans hold the feature precursor only about **74%** of the
time. But cembio also acquires a **separate MS1 injection** which holds it about
**100%** of the time and pairs unambiguously by polarity and run number. The two
injections are **55 days apart** (different batches), so no retention-time
correspondence is assumed: each feature is searched only in its own RT window
(rule 1) and the pattern is kept only if M+1 is genuinely present (rule 3).
Attaching it is worth **+13** and takes cembio past the semi-automated workflow.

Three rules keep the attachment honest, each learned from a compound it lost:

1. Take the precursor **apex across the feature's own RT span**, not the scan
   nearest its median RT: features can span 85 to 125 s, so the median points at
   the wrong part of the chromatogram.
2. **Trim to the isotope window.** A full survey scan is roughly 600 peaks and
   SIRIUS then fits the wrong envelope in crowded low-m/z regions.
3. **Attach nothing unless M+1 is genuinely present** above the noise floor at a
   plausible ratio. A weak precursor with no M+1 tells SIRIUS the molecule
   contains no carbon-13, which penalises carbon-rich formulas and demotes the
   correct one. **No MS1 beats a false MS1.**

Recovery numbers and the full method comparison live in
[`results_summary.qmd`](results_summary.html).

## Workflow overview

```{mermaid}
flowchart TB
  A["Raw mzML per (lab, mixture)"] --> B["Pre-steps: icl re-derive precursor;<br/>hmgu drop weak precursors;<br/>cembio attach MS1 from its MS1 injection"]
  B --> C["Group: merge precursors within 20 ppm,<br/>split at a 15 s RT gap (all labs)"]
  C --> D{"Representative<br/>per lab"}
  D -->|afekta| E["richest scan<br/>per collision energy"]
  D -->|cembio, icl| F["every scan, cap 40<br/>(SIRIUS consensus)"]
  D -->|hmgu| G["single richest scan"]
  E --> H["SIRIUS: formula ID + CSI:FingerID, BIO db"]
  F --> H
  G --> H
  H --> I["Match to the 20 spiked compounds"]
  I --> J["annotation_identity.csv + feat_rts.csv"]
  J -.->|consumed by| K["3_rt_alignment.qmd: MS1 anchoring + EIC"]
```

afekta, cembio and icl also run a second pass, the single richest scan for every lab, and the
two are unioned. hmgu has only the one run.

## Setup

```{r}
#| label: setup
#| message: false
library(Spectra)
library(MsExperiment)
library(xcms)
library(RuSirius)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(MsQuality)
library(tidytext)

source(here("1_preprocessing", "setup.R"))
```

## Config

```{r}
#| label: config
labs        <- c("afekta", "hmgu", "icl", "cembio")   # all four use the strategy below
study_group <- "HE"
# shared precursor-m/z window: the full mass range of the curated annotation, for cross-lab comparability.
mz_range    <- c(30, 1050)
# underscore mixture IDs throughout: "2.10" would round-trip through CSV as numeric 2.1 and collide with "2.1".
mixtures    <- c("1_1", "1_2", "1_3", "1_4", "1_5", "1_6",
                 "2_1", "2_2", "2_3", "2_4", "2_5",
                 "2_6", "2_7", "2_8", "2_9", "2_10",
                 "3_1", "3_2", "3_3")

# "primary" = each lab's FINALIZED representative (afekta richest-per-CE, cembio and icl
# all-scans, hmgu richest-scan), which is what this document describes. Setting it to
# "richest" instead forces single-richest for every lab, the union-baseline pass that
# lives in results/richest.
representative <- "primary"

# hmgu alone applies a precursor-abundance filter to drop weak-precursor DDA noise.
prec_thr_by_lab <- c(hmgu = 300, afekta = 0, icl = 0, cembio = 0)

output_dir <- here("2_annotation", "sirius_all_ms2", "results", representative)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

::: {.callout-note}
## This document renders from cached results

Every `(lab, mixture)` whose `sirius_summary_*` and `sirius_full_*` CSVs already exist
is loaded from disk rather than re-annotated, so rendering does **not** call SIRIUS and
does not need the cluster. `results/primary` is complete at 76/76, so a render is a
read-only reproduction of the finalized annotation. Delete a pair of CSVs to force that
one mixture to be recomputed.
:::

## 1. Extraction helper

`extract_all_ms2()` loads the raw mzML for one (lab, mixture) and returns the
spectra SIRIUS annotates, split by polarity. It implements the grouping,
representative and per-lab pre-steps described under
[The finalized strategy](#the-finalized-strategy) above; the code below is the
authoritative version of those steps.

Two facts the code does not restate:

- **Noise filter.** MS2 peaks are kept if they are at least 5% of the base peak
  or 1000 counts. This cuts the noisy late-RT hmgu spectra about fourfold (median
  91 to 21 peaks) while leaving clean spectra essentially untouched.
- **Feature RT is corrected downstream.** The grouping sets a feature's RT to its
  MS2-trigger time, not its chromatographic peak. `5_downstream_analysis_pg/3_rt_alignment.qmd`
  rewrites each `rt_sec` to the nearest MS1 peak at the precursor m/z, which fixes
  cembio's roughly 18 s schedule offset and the occasional DDA trigger far from
  the real peak.

```{r}
extract_all_ms2 <- function(lab_name, study_group, mixture) {
  study_path <- here("1_preprocessing", lab_name, study_group)
  mzml_dir   <- file.path(study_path, "mzml")
  mix_label  <- mixture

  empty_result <- function(sp = Spectra(backend = MsBackendMemory())) {
    list(spectra_pos = sp, spectra_neg = sp,
         n_ms2_pos = 0L, n_ms2_neg = 0L,
         n_features_pos = 0L, n_features_neg = 0L)
  }

  seq_pos <- read_xlsx(file.path(study_path, "seq_pos.xlsx")) |> as.data.frame()
  seq_neg <- read_xlsx(file.path(study_path, "seq_neg.xlsx")) |> as.data.frame()
  seq_pos$polarity <- "pos"; seq_neg$polarity <- "neg"
  seq <- rbind(seq_pos, seq_neg)
  seq$filename <- paste0(seq$`Data File`, ".mzML")
  seq <- seq[seq$filename %in% list.files(mzml_dir), ]
  seq$mixture <- gsub("\\.", "_",
                      sub("target$", "",
                          sub(".*_", "",
                              sub("_MS2$", "", seq$`Data File`))))

  has_ms2_files <- any(grepl("MS2", seq$filename, fixed = TRUE))
  seq <- if (has_ms2_files) {
    seq[grepl("MS2", seq$filename, fixed = TRUE) & seq$mixture == mix_label, ]
  } else {
    seq[seq$mixture == mix_label, ]
  }
  if (nrow(seq) == 0L) return(empty_result())

  sp <- Spectra(file.path(mzml_dir, seq$filename), backend = MsBackendMzR())
  sp <- setBackend(sp, MsBackendMemory())

  # icl: the Waters precursor m/z is unreliable, so re-derive it (recovery is 0 otherwise).
  if (lab_name == "icl") sp$precursorMz <- estimatePrecursorMz(sp)

  # afekta's primary pass uses a cleaner prep (5%-only filter, no cap, >=2 peaks); every other
  # lab and the richest pass use the production prep (5%|1000, cap 60, >=3 peaks).
  is_afekta <- lab_name == "afekta" && representative == "primary"

  sp <- filterIntensity(sp, intensity = function(x) {
    if (length(x) == 0L) return(logical(0))
    bp <- max(x, na.rm = TRUE)
    if (!is.finite(bp) || bp <= 0) return(rep(TRUE, length(x)))
    if (is_afekta) x > bp * 0.05 else x >= bp * 0.05 | x >= 1000
  }, msLevel. = 2L)
  sp <- applyProcessing(sp)
  sp <- sp[sp$msLevel != 2L | lengths(sp) >= (if (is_afekta) 2L else 3L)]

  # cap each MS2 to its 60 richest peaks: panel compounds have <=19, noise spectra up to ~250.
  cap_top <- function(x, n = 60L) {
    if (nrow(x) <= n) return(x)
    x[sort(order(x[, "intensity"], decreasing = TRUE)[seq_len(n)]), , drop = FALSE]
  }

  # grouping (all labs): merge by precursor within 20 ppm, split at a >15 s RT gap.
  group_precursor_rt <- function(file, rt, pmz, ppm = 20, rt_gap = 15) {
    g <- integer(length(rt)); ng <- 0L
    for (f in unique(file)) {
      idx <- which(file == f); o <- idx[order(pmz[idx])]; ps <- pmz[o]
      mzg <- cumsum(c(TRUE, (diff(ps) / ps[-length(ps)] * 1e6) > ppm))
      for (m in unique(mzg)) {
        ii <- o[mzg == m]; ii <- ii[order(rt[ii])]
        sub <- cumsum(c(TRUE, diff(rt[ii]) > rt_gap))
        for (s in unique(sub)) { ng <- ng + 1L; g[ii[sub == s]] <- ng }
      }
    }
    g
  }

  # representative (2a): richest scan, one MS2 per feature; precursor := feature median.
  reduce_richest <- function(ms2) {
    grp_f <- factor(ms2$group_idx)
    npk   <- lengths(ms2)
    g_med <- tapply(ms2$precursorMz, grp_f, median)
    sel   <- unlist(tapply(seq_along(ms2), grp_f,
                           function(ii) ii[which.max(npk[ii])]))
    keep  <- ms2[sel]
    keep$precursorMz <- as.numeric(g_med[as.character(keep$group_idx)])
    keep
  }

  # representative (2b, afekta): richest scan at each collision energy (~3 per feature).
  reduce_richest_perCE <- function(ms2) {
    ce  <- suppressWarnings(round(as.numeric(ms2$collisionEnergy)))
    ce[is.na(ce)] <- -1L
    fce <- interaction(factor(ms2$group_idx), factor(ce), drop = TRUE)
    npk <- lengths(ms2)
    sel <- sort(unlist(tapply(seq_along(ms2), fce,
                              function(ii) ii[which.max(npk[ii])]), use.names = FALSE))
    keep  <- ms2[sel]
    g_med <- tapply(ms2$precursorMz, factor(ms2$group_idx), median)
    keep$precursorMz <- as.numeric(g_med[as.character(keep$group_idx)])
    keep
  }
  # representative (2c, cembio + icl): every scan (cap 40 richest); SIRIUS builds its own consensus.
  reduce_allscans <- function(ms2, cap = 40L) {
    grp_f <- factor(ms2$group_idx)
    g_med <- tapply(ms2$precursorMz, grp_f, median)
    npk   <- lengths(ms2)
    sel   <- unlist(tapply(seq_along(ms2), grp_f, function(ii)
                    if (length(ii) <= cap) ii else ii[order(npk[ii], decreasing = TRUE)[seq_len(cap)]]),
                    use.names = FALSE)
    keep  <- ms2[sort(sel)]
    keep$precursorMz <- as.numeric(g_med[as.character(keep$group_idx)])
    keep
  }
  # the richest pass forces single-richest for every lab; the primary pass is per-lab (see strategy).
  reduce_fun <- if (representative == "richest") reduce_richest
                else if (lab_name %in% c("cembio", "icl")) reduce_allscans
                else if (is_afekta) reduce_richest_perCE
                else reduce_richest

  ms2 <- sp[sp$msLevel == 2L]
  ms2 <- filterEmptySpectra(ms2)
  if (length(ms2) == 0L) return(empty_result())
  if (!is_afekta) ms2 <- applyProcessing(addProcessing(ms2, cap_top, n = 60L))

  # keep only features inside the shared precursor-m/z window (mz_range), for cross-lab comparability.
  ms2 <- ms2[!is.na(ms2$precursorMz) &
             ms2$precursorMz >= mz_range[1] & ms2$precursorMz <= mz_range[2]]
  if (length(ms2) == 0L) return(empty_result())

  ms2$group_idx <- group_precursor_rt(as.character(ms2$dataOrigin),
                                      ms2$rtime, ms2$precursorMz)

  # MS1 for the isotope pattern (and hmgu's precursor filter). cembio takes it from a SEPARATE MS1
  # injection paired by polarity + run number; the "cembio: recovering the isotope pattern" section
  # of results_summary.qmd explains why and gives the five attachment rules implemented below.
  ms1_injection <- lab_name == "cembio"
  ms1 <- sp[sp$msLevel == 1L]
  ms2_to_ms1_file <- NULL
  if (ms1_injection) {
    inj <- list.files(mzml_dir, full.names = TRUE)
    inj <- inj[!grepl("MS2", basename(inj), fixed = TRUE)]
    runno  <- function(b) { m <- regmatches(b, regexpr("_[0-9]{3}_", b)); if (length(m)) m else NA_character_ }
    pol_of <- function(b) if (grepl("_pos_", b, fixed = TRUE)) "_pos_" else "_neg_"
    ms2_files <- unique(as.character(ms2$dataOrigin))
    map <- setNames(rep(NA_character_, length(ms2_files)), ms2_files)
    for (f in ms2_files) {
      b <- basename(f); rn <- runno(b); if (is.na(rn)) next
      hit <- inj[grepl(rn, basename(inj), fixed = TRUE) & grepl(pol_of(b), basename(inj), fixed = TRUE)]
      if (length(hit)) map[[f]] <- hit[1]
    }
    got <- sum(!is.na(map))
    message("  MS1 injection: paired ", got, "/", length(ms2_files), " MS2 files")
    if (got > 0) {
      ms1 <- setBackend(Spectra(unique(na.omit(unname(map))), backend = MsBackendMzR()), MsBackendMemory())
      ms1 <- ms1[ms1$msLevel == 1L]
      ms2_to_ms1_file <- map
    } else ms1_injection <- FALSE
  }
  use_ms1 <- (lab_name != "cembio" || ms1_injection) && length(ms1) > 0L

  if (use_ms1) {
    grp_f  <- factor(ms2$group_idx)
    g_ids  <- as.integer(levels(grp_f))
    g_file <- tapply(as.character(ms2$dataOrigin), grp_f, `[`, 1L)
    g_rt   <- tapply(ms2$rtime, grp_f, median)
    g_mz   <- tapply(ms2$precursorMz, grp_f, median)
    if (ms1_injection) {
      # key by basename: Spectra reports dataOrigin with backslashes, list.files with forward slashes.
      by_ms1file <- split(seq_along(ms1), basename(as.character(ms1$dataOrigin)))
      ms1_by_file <- setNames(lapply(names(ms2_to_ms1_file), function(f) {
        t <- ms2_to_ms1_file[[f]]
        if (is.na(t)) return(integer(0))
        v <- by_ms1file[[basename(t)]]; if (is.null(v)) integer(0) else v }),
        names(ms2_to_ms1_file))
    } else {
      ms1_by_file <- split(seq_along(ms1), as.character(ms1$dataOrigin))
    }
    # pick the precursor apex within the feature's own rt span, then keep it only if the isotope
    # pattern is real (M+0 above the floor, M+1 present at a plausible ratio). See results_summary.qmd.
    g_rtmin <- tapply(ms2$rtime, grp_f, min); g_rtmax <- tapply(ms2$rtime, grp_f, max)
    ms1_pd_all <- if (ms1_injection) peaksData(ms1) else NULL
    RT_PAD <- 10; PPM <- 20; MS1_MIN_INT <- 1000
    pick <- vapply(seq_along(g_ids), function(i) {
      cand <- ms1_by_file[[ g_file[i] ]]
      if (is.null(cand) || !length(cand)) return(NA_integer_)
      if (!ms1_injection) return(cand[which.min(abs(ms1$rtime[cand] - g_rt[i]))])
      rt <- ms1$rtime[cand]
      near <- cand[rt >= g_rtmin[i] - RT_PAD & rt <= g_rtmax[i] + RT_PAD]
      if (!length(near)) return(NA_integer_)
      mz <- g_mz[i]; tol <- mz * PPM / 1e6
      ints <- vapply(near, function(k) {
        p <- ms1_pd_all[[k]]; if (!nrow(p)) return(0)
        h <- abs(p[, "mz"] - mz) <= tol
        if (any(h)) max(p[h, "intensity"]) else 0
      }, numeric(1))
      if (all(ints <= 0)) return(NA_integer_)
      k <- near[which.max(ints)]
      p <- ms1_pd_all[[k]]
      h0 <- abs(p[, "mz"] - mz) <= tol
      if (!any(h0)) return(NA_integer_)
      i0 <- max(p[h0, "intensity"]); if (i0 < MS1_MIN_INT) return(NA_integer_)
      h1 <- abs(p[, "mz"] - (mz + 1.00336)) <= tol * 2
      if (!any(h1)) return(NA_integer_)
      r1 <- max(p[h1, "intensity"]) / i0
      if (r1 < 0.005 || r1 > 1) return(NA_integer_)
      k
    }, integer(1))
    if (ms1_injection)
      message("  MS1 attached to ", sum(!is.na(pick)), "/", length(g_ids), " features")
    keep_g  <- !is.na(pick)
    ms1_sel <- ms1[pick[keep_g]]
    ms1_sel$group_idx <- g_ids[keep_g]

    # trim attached MS1 to the precursor isotope window M-1.5 to M+7.5 (via addProcessing, not a hand-built Spectra).
    if (ms1_injection && length(ms1_sel)) {
      ms1_sel$prec_mz <- as.numeric(g_mz[keep_g])
      iso_trim <- function(x, prec_mz, ...) {
        if (!nrow(x) || is.na(prec_mz[1])) return(x)
        x[x[, "mz"] >= prec_mz[1] - 1.5 & x[, "mz"] <= prec_mz[1] + 7.5, , drop = FALSE]
      }
      ms1_sel <- applyProcessing(addProcessing(ms1_sel, iso_trim, spectraVariables = "prec_mz"))
    }

    # hmgu-only precursor-abundance filter: keep intensity = MAX precursor peak over the feature's scans.
    prec_thr <- prec_thr_by_lab[[lab_name]]
    if (length(prec_thr) && !is.na(prec_thr) && prec_thr > 0) {
      ms1_pd <- peaksData(ms1)
      fo <- as.character(ms2$dataOrigin)
      scan_int <- vapply(seq_along(ms2), function(k) {
        cand <- ms1_by_file[[ fo[k] ]]
        if (is.null(cand) || !length(cand)) return(NA_real_)
        s <- cand[which.min(abs(ms1$rtime[cand] - ms2$rtime[k]))]
        p <- ms1_pd[[s]]
        w <- p[abs(p[, "mz"] - ms2$precursorMz[k]) <= 0.01, "intensity"]
        if (length(w)) max(w) else 0
      }, numeric(1))
      g_max <- tapply(scan_int, grp_f,
                      function(v) if (all(is.na(v))) NA_real_ else max(v, na.rm = TRUE))
      keep_ids <- as.integer(names(g_max))[is.na(g_max) | g_max >= prec_thr]
      ms1_sel  <- ms1_sel[ms1_sel$group_idx %in% keep_ids]
      ms2      <- ms2[ms2$group_idx %in% keep_ids]
      if (length(ms2) == 0L) return(empty_result())
    }

    ms2     <- reduce_fun(ms2)                           # richest scan (or richest-per-CE for afekta)
    ms1_sel <- ms1_sel[ms1_sel$group_idx %in% unique(ms2$group_idx)]
    sp      <- c(ms1_sel, ms2)
  } else {
    sp <- reduce_fun(ms2)                                # no MS1 available for this feature set
  }
  if (length(sp) == 0L) return(empty_result(sp))

  list(
    spectra_pos    = sp[sp$polarity == 1L],
    spectra_neg    = sp[sp$polarity == 0L],
    n_ms2_pos      = sum(sp$msLevel == 2L & sp$polarity == 1L),
    n_ms2_neg      = sum(sp$msLevel == 2L & sp$polarity == 0L),
    n_features_pos = length(unique(sp$group_idx[sp$polarity == 1L])),
    n_features_neg = length(unique(sp$group_idx[sp$polarity == 0L]))
  )
}
```

## 2. SIRIUS helpers

Path helpers for the per-(lab, mixture) cache files, and a single list of
ID columns that SIRIUS returns inconsistently typed (sometimes numeric,
sometimes character). We coerce all of them to character on every read /
write so `bind_rows` across batches and mixtures never fails on type
mismatch. Add new column names to `sirius_id_cols` if a future error
surfaces one.

```{r}
#| label: helpers
sirius_sum_path       <- function(lab, m) file.path(output_dir, sprintf("sirius_summary_%s_mix_%s.csv",   lab, m))
sirius_full_path      <- function(lab, m) file.path(output_dir, sprintf("sirius_full_%s_mix_%s.csv",      lab, m))
extraction_count_path <- function(lab, m) file.path(output_dir, sprintf("extraction_counts_%s_mix_%s.csv", lab, m))

sirius_id_cols <- c("alignedFeatureId", "compoundId", "externalFeatureId",
                    "formulaId", "xcms_fts", "sirius_fts",
                    "molecularFormulaId", "structureCandidateInChIKey")
```

```{r}
#| label: sirius-login
sirius_user <- "plouail@eurac.edu"
sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")
```

### `run_sirius_for_lab()`

Runs SIRIUS over one (lab, mixture)'s spectra in a **single pass**: all
features are imported at once into **one project per (lab, mixture)**, which
is kept on disk (the `.sirius` space can be reopened in the SIRIUS GUI). On
the cluster we allocate enough memory to hold a whole mixture, so there is no
batching or project-rolling.

After import the function cross-checks `featuresId(srs, type = "xcms")`
against the features intended for import and re-imports anything silently
dropped (up to two retries; bail when retries stop making progress).

Pos and neg features go into the same project so RuSirius assigns polarity
correctly. `numberOfCandidates = 5` / `2` (per ionization) is a deliberate
cut from the default 15/5 - each candidate triggers a CSI:FingerID + BIO
lookup, the dominant per-feature cost. De-novo formula generation is off and
`ilpTimeout` bounds the ILP tree step (the per-instance/per-decomposition
second caps were removed).

```{r}
#| label: run-sirius-for-lab
run_sirius_for_lab <- function(sp_pos, sp_neg,
                               username, password, projectId, path) {
  if (length(sp_pos) == 0L && length(sp_neg) == 0L)
    return(list(summary = NULL, full = NULL))

  # One SIRIUS project per (lab, mixture), kept on disk so the .sirius space can
  # be reopened later in the SIRIUS GUI. No batching / project-rolling: import the
  # whole mixture at once and run SIRIUS in a single pass.
  srs <- Sirius(username = username, password = password,
                projectId = projectId, path = path)
  on.exit(try(shutdown(srs), silent = TRUE), add = TRUE)  # keep the project dir

  groups_pos <- unique(sp_pos$group_idx)
  groups_neg <- unique(sp_neg$group_idx)
  pol_lookup <- c(
    setNames(rep("pos", length(groups_pos)), as.character(groups_pos)),
    setNames(rep("neg", length(groups_neg)), as.character(groups_neg))
  )

  n_pos <- length(groups_pos); n_neg <- length(groups_neg)
  message("  importing ", n_pos, " pos + ", n_neg, " neg features (single pass)")

  if (n_pos > 0)
    srs <- import(srs, spectra = sp_pos, ms_column_name = "group_idx",
                  deleteExistingFeatures = TRUE)
  if (n_neg > 0)
    srs <- import(srs, spectra = sp_neg, ms_column_name = "group_idx",
                  deleteExistingFeatures = n_pos == 0)

  # cross-check featuresId vs intended and re-import anything silently dropped
  expected_ids <- as.character(c(groups_pos, groups_neg))
  missing_ids  <- setdiff(expected_ids, as.character(featuresId(srs, type = "xcms")))
  for (retry in 1:2) {
    if (length(missing_ids) == 0L) break
    message("    !! ", length(missing_ids), " missing - retry ", retry)
    miss_pos <- sp_pos[sp_pos$group_idx %in% missing_ids]
    miss_neg <- sp_neg[sp_neg$group_idx %in% missing_ids]
    if (length(miss_pos) > 0L)
      srs <- import(srs, spectra = miss_pos, ms_column_name = "group_idx", deleteExistingFeatures = FALSE)
    if (length(miss_neg) > 0L)
      srs <- import(srs, spectra = miss_neg, ms_column_name = "group_idx", deleteExistingFeatures = FALSE)
    prev_n      <- length(missing_ids)
    missing_ids <- setdiff(expected_ids, as.character(featuresId(srs, type = "xcms")))
    if (length(missing_ids) >= prev_n) { message("    !! retry made no progress - giving up"); break }
  }
  if (length(missing_ids) > 0L)
    warning("    !! ", length(missing_ids), " feature(s) never imported: ",
            paste(head(missing_ids, 10), collapse = ","))

  run(srs,
    formulaIdParams = formulaIdParam(
      numberOfCandidates              = 5,
      instrument                      = "QTOF",
      numberOfCandidatesPerIonization = 2,
      massAccuracyMS2ppm              = 10,
      filterByIsotopePattern          = FALSE,
      isotopeMs2Settings              = "SCORE",
      # De novo formula generation is combinatorial; off => bottom-up search only
      # (fast, bounded) restricted to BIO formulas - still finds the spike-in
      # panel. ilpTimeout caps the ILP tree step.
      performDeNovoBelowMz            = 0,
      formulaSearchDBs                = c("BIO"),
      ilpTimeout                      = TRUE),
    predictParams           = predictParam(),
    structureDbSearchParams = structureDbSearchParam(structureSearchDbs = c("BIO")),
    recompute = TRUE, wait = TRUE)

  sums <- summary(srs, result.type = "structure") |>
    mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
           across(any_of(sirius_id_cols), as.character),
           polarity = pol_lookup[as.character(externalFeatureId)])

  fulls <- results(srs, result.type = "structureDb",
                   topFormula = 8, topStructure = 20,
                   return.type = "data.frame") |>
    mutate(across(any_of(sirius_id_cols), as.character),
           polarity = pol_lookup[as.character(xcms_fts)])

  list(summary = sums, full = fulls)
}
```

## 3. Extract + run SIRIUS per (mixture, lab)

Each (mixture, lab) is processed end-to-end before moving to the next:
extract → run SIRIUS → write cache → drop spectra. Memory stays bounded
to one lab's spectra at a time, and a partial run resumes simply by
re-rendering - cached pairs skip both extraction and SIRIUS.

```{r}
#| label: extract-and-sirius
summary_results    <- setNames(vector("list", length(mixtures)), mixtures)
full_results       <- setNames(vector("list", length(mixtures)), mixtures)
extraction_summary <- list()

for (m in mixtures) {
  summary_results[[m]] <- list()
  full_results[[m]]    <- list()
  for (lab in labs) {
    sum_path  <- sirius_sum_path(lab, m)
    full_path <- sirius_full_path(lab, m)

    if (file.exists(sum_path) && file.exists(full_path)) {
      message("Cached: ", lab, " / mix ", m)
      summary_results[[m]][[lab]] <- read.csv(sum_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
               across(any_of(c(sirius_id_cols, "lab", "polarity", "adduct")), as.character))
      full_results[[m]][[lab]] <- read.csv(full_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c(sirius_id_cols, "lab", "polarity", "adduct")), as.character))
      if (file.exists(extraction_count_path(lab, m)))
        extraction_summary[[paste(m, lab)]] <-
          read.csv(extraction_count_path(lab, m),
                   stringsAsFactors = FALSE,
                   colClasses = c(mixture = "character",
                                  lab     = "character")) |>
          as_tibble()
      next
    }

    message("Extracting: ", lab, " / mix ", m)
    sp <- extract_all_ms2(lab, study_group, m)
    extraction_summary[[paste(m, lab)]] <- tibble(
      mixture = m, lab = lab,
      n_features_pos = sp$n_features_pos, n_features_neg = sp$n_features_neg,
      n_ms2_pos      = sp$n_ms2_pos,      n_ms2_neg      = sp$n_ms2_neg)
    write.csv(extraction_summary[[paste(m, lab)]],
              extraction_count_path(lab, m), row.names = FALSE)

    if (length(sp$spectra_pos) == 0L && length(sp$spectra_neg) == 0L) {
      message("  no MS2 - writing empty cache")
      summary_results[[m]][[lab]] <- tibble(lab = character(0))
      full_results[[m]][[lab]]    <- tibble(lab = character(0))
      write.csv(summary_results[[m]][[lab]], sum_path,  row.names = FALSE)
      write.csv(full_results[[m]][[lab]],    full_path, row.names = FALSE)
      rm(sp); gc(); next
    }

    if (!exists("sirius_user")) sirius_user <- Sys.getenv("SIRIUS_USERNAME")
    if (!exists("sirius_pass")) sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")

    res <- run_sirius_for_lab(
      sp_pos       = sp$spectra_pos,
      sp_neg       = sp$spectra_neg,
      username     = sirius_user, password = sirius_pass,
      projectId    = paste0("ring_trial_allms2_", lab, "_mix_", m),
      path         = output_dir)

    summary_results[[m]][[lab]] <-
      if (is.null(res$summary)) tibble(lab = character(0))
      else res$summary |> mutate(lab = lab)
    full_results[[m]][[lab]] <-
      if (is.null(res$full)) tibble(lab = character(0))
      else res$full |> mutate(lab = lab)

    write.csv(summary_results[[m]][[lab]], sum_path,  row.names = FALSE)
    write.csv(full_results[[m]][[lab]],    full_path, row.names = FALSE)

    rm(sp, res); gc()
  }
}

bind_rows(extraction_summary)
```

Extraction-summary bar plot per lab × polarity × mixture.

```{r}
#| label: plot-extraction
#| fig-cap: "SIRIUS features per lab × polarity × mixture (MS1+MS2 group; MS2 only for cembio)"
if (length(extraction_summary) > 0) {
  bind_rows(extraction_summary) |>
    select(lab, mixture, pos = n_features_pos, neg = n_features_neg) |>
    pivot_longer(c(pos, neg), names_to = "polarity", values_to = "n") |>
    ggplot(aes(x = lab, y = n, fill = polarity)) +
    geom_col(position = "dodge") +
    facet_wrap(~ mixture) +
    scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
    labs(x = NULL, y = "SIRIUS features", fill = "Polarity") +
    theme_bw()
}
```

Aggregate to long tables (`all_summaries`, `all_full`) keyed by mixture
for everything downstream.

```{r}
#| label: aggregate
# drop 0-row tibbles before bind_rows: header-only caches read back as logical columns and would crash the bind.
agg <- function(results_list) {
  bind_rows(lapply(mixtures, function(m) {
    non_empty <- Filter(\(df) nrow(df) > 0L, results_list[[m]])
    if (length(non_empty) == 0L) return(NULL)
    bind_rows(non_empty) |> mutate(mixture = m)
  }))
}
all_summaries <- agg(summary_results)
all_full      <- agg(full_results)

write.csv(all_summaries, file.path(output_dir, "sirius_summary_all_labs_all_mixtures.csv"), row.names = FALSE)
write.csv(all_full,      file.path(output_dir, "sirius_full_all_labs_all_mixtures.csv"),    row.names = FALSE)
```

### Feature-import verification

`run_sirius_for_lab()` already verifies each batch in-flight. This is the
end-to-end check: per (lab, mixture, polarity) we compare features
**extracted** upstream vs features SIRIUS produced **any** result for
(structure summary ∪ structureDb result). A residual `missing > 0` row
means SIRIUS imported the feature but produced no annotation - or, if a
warning fired during the loop, the retry import never succeeded.

```{r}
#| label: verify-import
#| include: false
extracted_long <- bind_rows(extraction_summary) |>
  select(mixture, lab, pos = n_features_pos, neg = n_features_neg) |>
  pivot_longer(c(pos, neg), names_to = "polarity", values_to = "extracted")

annotated_long <- bind_rows(
  all_summaries |> filter(!is.na(externalFeatureId)) |>
    transmute(mixture, lab, polarity, feature_id = externalFeatureId),
  all_full      |> filter(!is.na(xcms_fts)) |>
    transmute(mixture, lab, polarity, feature_id = xcms_fts)
) |>
  group_by(mixture, lab, polarity) |>
  summarise(annotated = n_distinct(feature_id), .groups = "drop")

verify_df <- extracted_long |>
  left_join(annotated_long, by = c("mixture", "lab", "polarity")) |>
  mutate(annotated = coalesce(annotated, 0L),
         missing   = extracted - annotated,
         pct_kept  = ifelse(extracted == 0, NA_real_,
                            round(100 * annotated / extracted, 1)))

write.csv(verify_df,
          file.path(output_dir, "feature_import_verification.csv"),
          row.names = FALSE)

verify_df |> arrange(desc(missing)) |> knitr::kable()

cat("Entries with missing > 0:", sum(verify_df$missing > 0, na.rm = TRUE),
    "  |  total extracted:",  sum(verify_df$extracted, na.rm = TRUE),
    "  annotated:",            sum(verify_df$annotated, na.rm = TRUE),
    "  missing:",              sum(verify_df$missing,   na.rm = TRUE), "\n")
```

## 4. Match to expected compounds (all mixtures)

```{r}
#| label: expected
expected_all <- meta |>
  mutate(Mixture = gsub("\\.", "_", Mixture)) |>   # normalize dot → underscore at entry
  filter(Mixture %in% mixtures) |>
  transmute(mixture              = Mixture,
            compound             = `ChEBI name`,
            expected_formula     = formula,
            expected_inchikey    = InChIKey,
            expected_inchikey_2d = substr(InChIKey, 1, 14))

expected_all |>
  count(mixture, name = "n_expected") |>
  knitr::kable(caption = "Expected compounds per mixture")
```

### Ambiguous compounds in the expected list

Two patterns are checked here before the coverage analysis.

**(a) Within-mixture stereoisomers (same 2D InChIKey).** SIRIUS returns
only the 2D InChIKey, so two compounds sharing those 14 characters in
the same mixture cannot be told apart from the annotation. We **drop
both compounds** from `expected_all` rather than over-counting one of
them.

**(b) Same compound listed in multiple mixtures.** Informational only -
the join below keys on `mixture`, so each annotation is attributed to
its own mixture and these duplicates don't double count.

```{r}
#| label: expected-stereoisomers
exp_dups_2d <- expected_all |>
  group_by(mixture, expected_inchikey_2d) |>
  filter(n() > 1) |>
  arrange(mixture, expected_inchikey_2d, compound) |>
  ungroup()

knitr::kable(exp_dups_2d,
  caption = "(a) Stereoisomer pairs - DROPPED from expected_all")

expected_all <- expected_all |>
  anti_join(exp_dups_2d, by = c("mixture", "expected_inchikey_2d"))
```

```{r}
#| label: expected-cross-mixture
exp_dups_mix <- expected_all |>
  group_by(expected_inchikey) |>
  filter(n_distinct(mixture) > 1) |>
  arrange(expected_inchikey, mixture) |>
  ungroup()

knitr::kable(exp_dups_mix,
  caption = "(b) Same compound across multiple mixtures (kept)")

all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  filter(!is.na(inchiKey)) |>
  semi_join(exp_dups_mix |> distinct(mixture, expected_inchikey_2d),
            by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  count(mixture, lab, polarity, inchikey_2d) |>
  knitr::kable(caption = "Were the cross-mixture duplicates annotated?")
```

A SIRIUS hit counts as recovery of an expected compound when its 2D
InChIKey matches the expected one **for the same mixture**. Per
(mixture, lab, compound, polarity) we record the best formula rank, best
structure rank and the union of adducts seen.

```{r}
#| label: match-structure
structure_hits_all <- all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  inner_join(
    expected_all |>
      select(mixture, expected_inchikey_2d, compound,
             expected_formula, expected_inchikey),
    by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  group_by(mixture, lab, inchikey_2d, polarity) |>
  summarise(
    best_formula_rank   = suppressWarnings(min(rank.x, na.rm = TRUE)),
    best_structure_rank = suppressWarnings(min(rank.y, na.rm = TRUE)),
    adducts             = paste(sort(unique(na.omit(adduct))), collapse = ";"),
    molecularFormula    = first(molecularFormula),
    compound            = first(compound),
    expected_formula    = first(expected_formula),
    expected_inchikey   = first(expected_inchikey),
    .groups = "drop") |>
  mutate(across(c(best_formula_rank, best_structure_rank),
                \(x) ifelse(is.infinite(x), NA_real_, x)))

structure_hits_all
```

## 5. Coverage

```{r}
#| label: coverage
covered_all <- expected_all |>
  left_join(structure_hits_all |>
              distinct(mixture, inchikey_2d) |>
              mutate(structure_recovered = TRUE),
            by = c("mixture", "expected_inchikey_2d" = "inchikey_2d")) |>
  mutate(structure_recovered = !is.na(structure_recovered))

covered_all |>
  group_by(mixture) |>
  summarise(n_recovered = sum(structure_recovered),
            n_expected  = n(), .groups = "drop") |>
  knitr::kable(caption = "Structure recovery per mixture")
```

```{r}
#| label: plot-coverage-by-lab
#| fig-cap: "Structure hits per lab × polarity, faceted by mixture"
#| fig-height: 8
structure_hits_all |>
  count(mixture, lab, polarity) |>
  ggplot(aes(x = lab, y = n, fill = polarity)) +
  geom_col(position = "dodge") +
  facet_wrap(~ mixture) +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "Structure hits", fill = "Polarity") +
  theme_bw()
```

## 6. RT extraction for EIC pipeline

RT per feature is extracted from the MS2 spectra and cached per (lab,
mixture) so raw files are only read once. `hit_rt` joins the RT onto
expected-compound matches and is the input to §7.1's `eic_params`. No
cross-lab RT analysis here — that lives in `8_per_compound_diagnostics.qmd`.

```{r}
#| label: extract-rt
#| message: false
rt_all <- bind_rows(lapply(mixtures, function(m) {
  bind_rows(lapply(labs, function(lab) {
    path <- file.path(output_dir, paste0(lab, "_mix_", m, "_rt.csv"))
    if (file.exists(path)) {
      df <- read.csv(path, stringsAsFactors = FALSE)
      if ("rt_min" %in% names(df) && !"rt_sec" %in% names(df))
        df <- mutate(df, rt_sec = rt_min * 60)
      return(df |> mutate(lab = lab, mixture = m))
    }
    message("Extracting RT for: ", lab, " / mix ", m)
    res <- extract_all_ms2(lab, study_group, m)
    make_rt_tbl <- function(sp) {
      if (length(sp) == 0L)
        tibble(group_idx = integer(), msLevel = integer(), rt_sec = numeric())
      else
        tibble(group_idx = sp$group_idx, msLevel = sp$msLevel, rt_sec = sp$rtime)
    }
    rt_df <- bind_rows(make_rt_tbl(res$spectra_pos),
                       make_rt_tbl(res$spectra_neg)) |>
      filter(msLevel == 2L) |>
      group_by(group_idx) |>
      summarise(rt_sec = median(rt_sec, na.rm = TRUE), .groups = "drop") |>
      mutate(lab = lab, mixture = m)
    write.csv(rt_df, path, row.names = FALSE)
    rt_df
  }))
}))
```

```{r}
#| label: rt-expected-compounds
hit_rt <- all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14),
         xcms_fts    = as.character(xcms_fts)) |>
  filter(!is.na(inchiKey)) |>
  inner_join(
    expected_all |> select(mixture, expected_inchikey_2d, compound,
                           expected_formula, expected_inchikey),
    by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  left_join(rt_all |> mutate(group_idx = as.character(group_idx)),
            by = c("lab", "mixture", "xcms_fts" = "group_idx")) |>
  filter(!is.na(rt_sec))

cat("hit_rt rows (per-feature RT for expected-compound matches):",
    nrow(hit_rt), "\n")
```

## 7. Candidate pool for the downstream RT-alignment / EIC pipeline

This producer does **no** EIC extraction. It emits one candidate row per
annotated feature and writes the two canonical CSVs the downstream reads. All
chromatography work, MS1-EIC extraction, cross-lab RT alignment, the SNR-aware
peak picker, and the cluster-consensus RT, lives in
[`5_downstream_analysis_pg/3_rt_alignment.qmd`](../../../5_downstream_analysis_pg/3_rt_alignment.qmd),
which consumes `feat_rts.csv` and writes `annotation_chrom_metrics.csv`.

### 7.1 Candidate pool (`eic_candidates`)

One row per feature per `(mixture, lab, polarity, compound, adduct)`, keyed on
the MS2-trigger RT from section 6. No collapse across features, so
inclusion-list-shattered peaks stay separate until the downstream Stage-2
similarity rule snaps them to one consensus RT.

```{r}
#| label: eic-candidates
#| message: false
# one row per feature per (mixture, lab, polarity, compound, adduct): the raw
# candidate pool for the consensus clustering in 3_rt_alignment.qmd. No collapse here,
# so inclusion-list-shattered peaks stay separate until Stage 2 snaps them to one RT.
eic_candidates <- hit_rt |>
  select(mixture, lab, polarity, xcms_fts, compound, inchikey_2d, rt_sec) |>
  distinct() |>
  left_join(all_summaries |>
              select(mixture, lab, externalFeatureId, ionMass, adduct,
                     confidenceExactMatch) |>
              mutate(externalFeatureId = as.character(externalFeatureId)),
            by = c("mixture", "lab", "xcms_fts" = "externalFeatureId")) |>
  filter(!is.na(ionMass), !is.na(rt_sec)) |>
  rename(cosmic = confidenceExactMatch)
```

### 7.2 Write feature-RT table for downstream consumption

The per-feature MS2-trigger RTs in `eic_candidates` are the input to
the alignment-based cluster algorithm in `5_downstream_analysis_pg/3_rt_alignment.qmd`
(`§P2.9` onwards). From there the EIC extraction + SNR-aware picker
writes `5_downstream_analysis_pg/object/annotation_chrom_metrics.csv`.

This producer no longer does EIC extraction — the SIRIUS run + the
expected-compound matching above are its only responsibilities. All
downstream chromatography work — alignment, cluster RT, EIC, picker,
metrics — lives in `3_rt_alignment.qmd`.

```{r}
#| label: write-feat-rts
feat_rts <- eic_candidates |>
  select(mixture, lab, polarity, xcms_fts, compound, inchikey_2d,
         rt_sec, ionMass, adduct, cosmic)

dir.create(here::here("5_downstream_analysis_pg", "object"),
           recursive = TRUE, showWarnings = FALSE)
write.csv(feat_rts,
          here::here("5_downstream_analysis_pg", "object", "feat_rts.csv"),
          row.names = FALSE)
cat(sprintf("feat_rts.csv -- %d rows (per-feature MS2-trigger RTs)\n",
            nrow(feat_rts)))
```

## 8. Write canonical CSVs

This producer writes the **identity table** to
`5_downstream_analysis_pg/object/annotation_identity.csv`. The
chromatography metrics (`annotation_chrom_metrics.csv`) are produced
downstream in `3_rt_alignment.qmd`.

### 8.1 `annotation_identity.csv`

`structure_hits_all` is augmented with the COSMIC scores and the
observed `ionMass` from `all_summaries`, plus a placeholder
`confidence_tier = "unfiltered"`. The join picks the best-`csiScore`
candidate per `(mixture, lab, inchikey_2d, polarity)` — identity is
one-per-compound, so we collapse the candidate dimension first; the
resulting `ionMass` is the precursor m/z of that winning candidate and
is exposed downstream as `mz_observed` (basis for the inter-lab m/z
deviation analysis in `8_per_compound_diagnostics.qmd`).

```{r}
#| label: build-annotation-identity
id_summary <- all_summaries |>
  filter(!is.na(inchiKey)) |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  group_by(mixture, lab, inchikey_2d, polarity) |>
  slice_max(csiScore, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(mixture, lab, inchikey_2d, polarity,
         confidenceExactMatch, confidenceApproxMatch,
         mz_observed = ionMass)

annotation_identity <- structure_hits_all |>
  left_join(id_summary,
            by = c("mixture", "lab", "inchikey_2d", "polarity")) |>
  mutate(confidence_tier = "unfiltered") |>
  select(mixture, lab, compound, inchikey_2d, polarity,
         best_formula_rank, best_structure_rank, adducts,
         molecularFormula, expected_formula, expected_inchikey,
         mz_observed, confidenceExactMatch, confidenceApproxMatch,
         confidence_tier)

write.csv(annotation_identity,
          here::here("5_downstream_analysis_pg", "object",
                     "annotation_identity.csv"),
          row.names = FALSE)
cat("annotation_identity.csv  -- ", nrow(annotation_identity), " rows\n",
    "  with COSMIC scores: ",
    sum(!is.na(annotation_identity$confidenceExactMatch)),
    "  with mz_observed: ",
    sum(!is.na(annotation_identity$mz_observed)), "\n", sep = "")
```

