---
title: "All-MS2 SIRIUS: results"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{css}
.rt { border-collapse:collapse; font-size:13px; font-variant-numeric:tabular-nums }
.rt th, .rt td { padding:5px 9px; text-align:center; border:1px solid #e2e8ea }
.rt thead th { font-size:11px; text-transform:uppercase; color:#4a5a63; background:#f6f9fa }
.rt thead th.lab { border-bottom:2px solid #0d7186; color:#0d7186 }
.rt tfoot td, .rt tfoot th { border-top:2px solid #cfd8db; font-weight:700 }
/* Quarto adds Bootstrap's .table-striped, whose row rule sets a background on every cell and
   outranks a bare .hi selector, so the colour bands need the extra specificity and !important. */
.rt td.hi   { background-color:#238a5e1f !important; color:#17603f; font-weight:600 }
.rt td.mid  { background-color:#b5771f22 !important; color:#8a5a15; font-weight:600 }
.rt td.lo   { background-color:#c5433f1e !important; color:#963330; font-weight:600 }
.rt td.na   { color:#9aa8af }
.rt td.semi { color:#7a8b93 }
.rt td.tot, .rt tfoot td { color:#0d7186 }
.rt td.best { box-shadow:inset 0 0 0 2px #0d7186; border-radius:3px }
.up { color:#238a5e; font-weight:700 }
.dn { color:#c5433f; font-weight:700 }
```

```{r setup}
suppressWarnings(suppressMessages({
  source(here::here("1_preprocessing", "setup.R")); library(knitr)
}))
meta$mix <- chartr(".", "_", meta$Mixture)
two  <- function(x) substr(x, 1, 14)
RES  <- here::here("2_annotation_auto", "sirius_all_ms2", "results")
CUR  <- here::here("3_annotation_manual", "HE", "3_Sirius_curation")
MIX  <- c(paste0("1_", 1:6), paste0("2_", 1:10), paste0("3_", 1:3))
LABS <- c("afekta", "cembio", "icl", "hmgu")
METHOD <- c(afekta = "per-CE", cembio = "all-scans + MS1", icl = "all-scans", hmgu = "single run")

panelk <- function(mx) unique(two(meta$InChIKey[meta$mix == mx & !is.na(meta$InChIKey)]))
cpd    <- function(k) meta[["ChEBI name"]][match(k, two(meta$InChIKey))]

## one reader for every pass. top1 = SIRIUS's first structure per feature, any = anywhere in the
## candidate list. Both unfiltered by panel, because the hmgu shift test scores a file against a
## DIFFERENT mixture's panel and pre-filtering would force every cross-panel count to zero.
read_pass <- local({
  cache <- new.env(parent = emptyenv())
  function(pass, lab, mx) {
    key <- paste(pass, lab, mx)
    if (!is.null(cache[[key]])) return(cache[[key]]$v)
    f <- file.path(RES, pass, sprintf("sirius_full_%s_mix_%s.csv", lab, mx))
    v <- if (!file.exists(f)) NULL else {
      d <- read.csv(f, stringsAsFactors = FALSE, check.names = FALSE)
      d <- d[!is.na(d$inchiKey) & nzchar(d$inchiKey), ]
      if (!nrow(d)) list(top1 = character(0), any = character(0)) else {
        o <- d[order(d$xcms_fts, -suppressWarnings(as.numeric(d$csiScore))), ]
        list(top1 = unique(two(o$inchiKey[!duplicated(o$xcms_fts)])), any = unique(two(o$inchiKey)))
      }
    }
    assign(key, list(v = v), envir = cache); v
  }
})
got <- function(pass, lab, mx, how = "top1") {
  v <- read_pass(pass, lab, mx); if (is.null(v)) NULL else intersect(panelk(mx), v[[how]])
}
n_or_na <- function(v) if (is.null(v)) NA_integer_ else length(v)

## semi-automated reference. best_structurePerIdRank is the rank SIRIUS gave the curated target;
## NA means nothing was annotated from MS2 for it, so it is not an annotation and drops out.
semi_tbl <- local({
  cache <- new.env(parent = emptyenv())
  function(lab) {
    if (!is.null(cache[[lab]])) return(cache[[lab]]$v)
    f <- file.path(CUR, paste0(lab, "_lab_annotations_sirius_curated.xlsx"))
    v <- if (!file.exists(f)) NULL else {
      d <- as.data.frame(read_xlsx(f, .name_repair = "minimal"), check.names = FALSE)
      data.frame(mix = chartr(".", "_", as.character(d$mixture)), k2 = two(d$target_InChIKey),
                 rank = suppressWarnings(as.numeric(d$best_structurePerIdRank)))
    }
    assign(lab, list(v = v), envir = cache); v
  }
})
semi_got <- function(lab, mx, maxrank = 1) {
  d <- semi_tbl(lab); if (is.null(d)) return(NULL)
  intersect(panelk(mx), unique(d$k2[d$mix == mx & !is.na(d$rank) & d$rank <= maxrank]))
}
semi <- function(lab, mx) n_or_na(semi_got(lab, mx))

## colour a count by strength; used with kable(escape = FALSE)
band <- function(v) ifelse(is.na(v), "na", ifelse(v >= 12, "hi", ifelse(v >= 6, "mid", "lo")))
fmt  <- function(v) ifelse(is.na(v), '<span class="na">-</span>',
                           sprintf('<span class="%s">%d</span>', band(v), v))
dfmt <- function(d) ifelse(d == 0, "=", sprintf('<span class="%s">%+d</span>',
                                                ifelse(d > 0, "up", "dn"), d))
```

Each mixture was spiked with a panel of 20 known standards. Recovery counts how many
of those 20 SIRIUS named correctly, matched on the skeleton part of the InChIKey so
stereochemistry is not required. The semi-automated curated workflow is the reference
throughout, and a run that does not exist is shown as a dash rather than counted as zero.

## How the annotation was run

afekta, cembio and icl are each annotated **twice** and the two runs are combined. hmgu is
annotated **once**, for a reason given in step 3.

**Step 1, lab-specific pre-steps.** Applied to the raw scans before anything else.

- *icl* re-derives the precursor m/z with `estimatePrecursorMz()`. The Waters mzML value is
  unreliable, and recovery is zero without this.
- *hmgu* drops features whose precursor is too weak to be real. For each MS2 scan, the nearest
  MS1 scan in time is taken and the precursor's intensity read from it within 0.01 m/z. A feature
  is kept if the **highest** of those readings, across all its scans, reaches **300 counts**.
  Taking the maximum rather than a single mid-retention scan matters: the mid-scan estimate lands
  off-apex and discarded real compounds.
- *cembio* attaches an MS1 isotope pattern from its separate MS1 injection, described in its own
  section below.
- *afekta* needs no pre-step.

**Step 2, grouping. Identical for every lab.**

- Merge MS2 scans whose precursor m/z agree within 20 ppm.
- Split a group wherever the retention-time gap between consecutive scans exceeds 15 s.
- The result is one feature per real precursor peak, with the feature's precursor set to the
  median m/z of its scans.

**Step 3, choosing which scans SIRIUS sees.** A feature can hold one scan or dozens, so something
must choose. This is the only step the two runs disagree about.

- **The secondary run** keeps the single scan with the most fragment peaks. Identical for every
  lab, so it is the common baseline. It is stored under `results/richest`, after what it does.
- **The primary run** picks whatever suits how that lab acquires:
  - *afekta*, the richest scan at each collision energy, giving about three energy-resolved
    spectra per feature. Its clean 10/20/40 eV ladder means the rungs carry complementary
    fragments.
  - *cembio* and *icl*, every scan of the feature, capped at the 40 richest, letting SIRIUS build
    its own consensus. Both have weak or noisy individual scans, and pooling recovers evidence no
    single scan carries.
- *hmgu is annotated only once.* It is single-collision-energy with about 72% single-scan
  features, so every possible representative reduces to the same single richest scan. There is no
  second run to make and nothing to combine. Its column in the tables is that one run.

**Step 4, combining.** For the three labs with two runs, a compound counts as recovered if
**either** run finds it. Neither run is a superset of the other, so the union is the real ceiling
of these methods.

## Recovery by lab

Counts are top-1 identity out of 20. Colour marks strength, green 12 or more, amber 6 to 11,
red 5 or fewer.

```{r recovery}
## grouped header (lab spanning its own columns), coloured cells, teal outline on the better
## all-MS2 pass for that mixture. Built directly because a grouped header needs the markup.
cell <- function(v, best = FALSE) {
  if (is.na(v)) return('<td class="na">-</td>')
  sprintf('<td class="%s%s">%d</td>', band(v), if (best) " best" else "", v)
}
scell <- function(v) if (is.na(v)) '<td class="na">-</td>' else sprintf('<td class="semi">%d</td>', v)

## hmgu's group-2 vials are mislabelled by one, so scoring a file against the panel its NAME
## claims understates it. The corrected column scores it against the panel it actually contains.
true_mix <- function(mx) {
  if (mx == "2_10") return("2_2")
  p <- as.integer(strsplit(mx, "_")[[1]])
  if (p[1] == 2L && p[2] >= 2L && p[2] <= 9L) return(sprintf("2_%d", p[2] + 1L))
  mx
}
hmgu_corr <- function(mx) {
  v <- read_pass("primary", "hmgu", mx); if (is.null(v)) return(NA_integer_)
  length(intersect(panelk(true_mix(mx)), v$top1))
}

tot <- setNames(lapply(LABS, function(l) {
  keep <- Filter(function(mx) !is.null(got("primary", l, mx)) &&
                   (l == "hmgu" || !is.null(got("richest", l, mx))), MIX)
  list(n = length(keep),
       S = sum(vapply(keep, function(mx) semi(l, mx), 1L)),
       P = sum(vapply(keep, function(mx) length(got("primary", l, mx)), 1L)),
       C = if (l != "hmgu") NA_integer_ else sum(vapply(keep, hmgu_corr, 1L)),
       R = if (l == "hmgu") NA_integer_ else sum(vapply(keep, function(mx) length(got("richest", l, mx)), 1L)))
}), LABS)

h1 <- paste0('<tr><th rowspan="2">mixture</th>',
             paste(vapply(LABS, function(l)
               sprintf('<th class="lab" colspan="3">%s</th>', l), ""), collapse = ""), '</tr>')
h2 <- paste0('<tr>', paste(vapply(LABS, function(l)
  if (l == "hmgu") '<th>semi</th><th>as labelled</th><th>corrected</th>'
  else sprintf('<th>semi</th><th>secondary</th><th>%s</th>', METHOD[[l]]), ""), collapse = ""), '</tr>')

body <- vapply(MIX, function(mx) {
  cells <- vapply(LABS, function(l) {
    p <- n_or_na(got("primary", l, mx))
    if (l == "hmgu") {
      cc <- hmgu_corr(mx)
      return(paste0(scell(semi(l, mx)), cell(p, !is.na(p) && !is.na(cc) && p >= cc),
                    cell(cc, !is.na(cc) && !is.na(p) && cc > p)))
    }
    r <- n_or_na(got("richest", l, mx))
    bp <- !is.na(p) && (is.na(r) || p >= r); br <- !is.na(r) && (is.na(p) || r > p)
    paste0(scell(semi(l, mx)), cell(r, br), cell(p, bp))
  }, "")
  paste0('<tr><th>', mx, '</th>', paste(cells, collapse = ""), '</tr>')
}, "")

foot <- paste0('<tr><th>total</th>', paste(vapply(LABS, function(l) {
  s <- sprintf('<td class="semi">%d</td>', tot[[l]]$S)
  if (l == "hmgu") paste0(s, sprintf('<td>%d</td><td class="tot">%d</td>', tot[[l]]$P, tot[[l]]$C))
  else paste0(s, sprintf('<td>%d</td><td class="tot">%d</td>', tot[[l]]$R, tot[[l]]$P))
}, ""), collapse = ""), '</tr>')

knitr::asis_output(paste0('<table class="rt"><thead>', h1, h2, '</thead><tbody>',
                          paste(body, collapse = ""), '</tbody><tfoot>', foot, '</tfoot></table>'))
```

Totals cover only the mixtures where a lab's passes both exist, so the columns compare like
with like: `r paste(sprintf("%s %d", LABS, vapply(LABS, function(l) tot[[l]]$n, 1L)), collapse = ", ")`.

hmgu carries two columns instead of a second run. **As labelled** scores each file against the
panel its filename claims, which is how every other lab is scored. **Corrected** scores it against
the panel it actually contains, following the vial map established in the last section. The gap
between them is the mislabel, not a method effect, and the corrected column is hmgu's real
performance.

## Union of the two passes

```{r union}
u <- do.call(rbind, lapply(c("afekta", "cembio", "icl"), function(l) {
  keep <- Filter(function(mx) !is.null(got("primary", l, mx)) && !is.null(got("richest", l, mx)), MIX)
  P <- sum(vapply(keep, function(mx) length(got("primary", l, mx)), 1L))
  R <- sum(vapply(keep, function(mx) length(got("richest", l, mx)), 1L))
  U <- sum(vapply(keep, function(mx) length(union(got("primary", l, mx), got("richest", l, mx))), 1L))
  data.frame(lab = l, mixtures = length(keep), primary = P, secondary = R, union = U,
             `union gain` = dfmt(U - max(P, R)), check.names = FALSE)
}))
kable(u, format = "html", escape = FALSE, row.names = FALSE, table.attr = 'class="rt"')
```

The gain is what the union adds over the better single pass, so it is the cost of shipping
only one representative.

The gain comes out as **+7 in all three labs**, which looks like an error and is worth
explaining. It is not. In every lab the primary pass has the larger total, so the union is
simply primary plus the compounds that only the secondary run finds, so the gain *equals* that
secondary-only count. Those counts happen to be 7, 7 and 7. The arithmetic was checked per mixture
in both directions, and the compounds themselves are almost entirely different between labs,
sharing only homocysteine thiolactone between afekta and cembio. So it is a coincidence of three
small numbers, not a shared cause.

## What each pass contributes

The two passes do not just differ in quantity, they favour different chemistry, and the
split is consistent across labs.

A caveat on how the classes are assigned. The panel metadata carries ChEBI, KEGG, HMDB and
LipidMaps identifiers but no class or ontology column, so the groupings below come from
**matching the compound name** (anything ending in "acid" is counted as an acid, names containing
"glucos" or "ribos" as sugars, and so on). That is a rough indication of a trend, not a formal
classification, and a handful of compounds land in "other" because the name gives nothing away.
The table after it uses only the molecular formula, so it involves no judgement at all, and it
points the same way.

```{r classes}
classify <- function(n) {
  n <- tolower(ifelse(is.na(n), "", n))
  if (grepl("cholic|cholan|sterol|estr|pregn|androst", n))                    return("steroid, bile acid")
  if (grepl("phosphocholine|glycero|acyl-sn|ceramide|sphingo", n))            return("lipid")
  if (grepl("adenosin|guanosin|cytidin|uridin|inosin|nad|amp|adp|atp|xanthin", n)) return("nucleotide, cofactor")
  if (grepl("glucos|galactos|mannos|fructos|sucros|ribos|saccharide|glcp", n))return("sugar")
  if (grepl("glycine|alanine|valine|leucine|serine|threonine|proline|lysine|arginine|histidine|tryptophan|tyrosine|phenylalanine|glutam|aspart|cystein|methionin|asparagin|homocystein", n)) return("amino acid")
  if (grepl("acid$|acid |oate$|oic", n))                                      return("small organic acid")
  "other"
}
sides <- do.call(rbind, lapply(c("afekta", "cembio", "icl"), function(l) {
  po <- ro <- character(0)
  for (mx in MIX) {
    p <- got("primary", l, mx); r <- got("richest", l, mx)
    if (is.null(p) || is.null(r)) next
    po <- c(po, setdiff(p, r)); ro <- c(ro, setdiff(r, p))
  }
  rbind(data.frame(side = "only primary finds it", class = vapply(cpd(po), classify, "")),
        data.frame(side = "only secondary finds it", class = vapply(cpd(ro), classify, "")))
}))
ct <- as.data.frame.matrix(table(sides$class, sides$side))
ct <- ct[order(-rowSums(ct)), , drop = FALSE]
kable(cbind(class = rownames(ct), ct), format = "html", row.names = FALSE, table.attr = 'class="rt"')
```

Now the same two sets described only by molecular formula, with no naming judgement involved.
Mass and carbon count say how big the molecules are. The O/C ratio is oxygen atoms divided by
carbon atoms, a standard way to say how oxygen-rich a molecule is: a fatty acid sits near 0.1,
a small organic acid nearer 0.5, and a sugar reaches 1.0 because it carries one oxygen per
carbon. It is used here only as an objective stand-in for "sugar-like or acid-like" that does
not depend on reading names.

```{r props}
nel <- function(f, el) {
  if (is.na(f)) return(NA_integer_)
  m <- regmatches(f, gregexpr(paste0(el, "[0-9]*(?![a-z])"), f, perl = TRUE))[[1]]
  if (!length(m)) return(0L)
  n <- sub(paste0("^", el), "", m); sum(ifelse(n == "", 1L, as.integer(n)))
}
po <- ro <- character(0)
for (l in c("afekta", "cembio", "icl")) for (mx in MIX) {
  p <- got("primary", l, mx); r <- got("richest", l, mx)
  if (is.null(p) || is.null(r)) next
  po <- c(po, setdiff(p, r)); ro <- c(ro, setdiff(r, p))
}
side_props <- function(k) {
  i <- match(k, two(meta$InChIKey)); f <- meta$formula[i]
  C <- vapply(f, nel, 1L, "C"); O <- vapply(f, nel, 1L, "O")
  list(n = length(k),
       mass = median(suppressWarnings(as.numeric(meta$M[i])), na.rm = TRUE),
       C = median(C, na.rm = TRUE), OC = median(O / pmax(C, 1), na.rm = TRUE),
       P = mean(vapply(f, nel, 1L, "P") > 0, na.rm = TRUE),
       S = mean(vapply(f, nel, 1L, "S") > 0, na.rm = TRUE))
}
A <- side_props(po); B <- side_props(ro)
kable(data.frame(
  ` ` = c("compounds", "median mass", "median carbon count",
          "median O/C ratio", "contains phosphorus", "contains sulfur"),
  `only primary finds it` = c(A$n, round(A$mass), A$C, sprintf("%.2f", A$OC),
                              sprintf("%.0f%%", 100*A$P), sprintf("%.0f%%", 100*A$S)),
  `only secondary finds it` = c(B$n, round(B$mass), B$C, sprintf("%.2f", B$OC),
                              sprintf("%.0f%%", 100*B$P), sprintf("%.0f%%", 100*B$S)),
  check.names = FALSE), format = "html", row.names = FALSE, table.attr = 'class="rt"')
```

Reading the two together:

- The finalized representatives mostly rescue small organic acids and amino acids, in all three
  labs. Those fragment poorly, so a single scan rarely carries enough evidence, and pooling
  scans or resolving them by collision energy supplies it.
- What the secondary run uniquely keeps is a much smaller set, but it is chemically distinct rather than
  random. Sugars are the one class it wins outright, and the formula table shows it is markedly
  richer in phosphorus and more oxygen-rich.
- This is not a size effect. Median mass and carbon count are nearly identical between the two
  sets, so the difference is composition, not how big the molecules are.
- Phosphorus-bearing compounds, meaning nucleotides and cofactors such as NAD and NADH plus the
  phosphocholines, fragment readily. Averaging across scans appears to dilute the informative
  spectrum or favour over-fragmented rungs of the energy ladder, which is why the single best
  scan still wins for them.

That is a mechanistic reason to run both passes, rather than a quirk of any one lab.

## Top-1 versus any candidate, on both workflows

SIRIUS returns a ranked candidate list in both workflows, and the curated data records the rank
it gave each target. Relaxing "ranked first" to "anywhere in the list" therefore has to be applied
to both sides, otherwise the comparison is rigged.

```{r anyrank}
## rank of each panel compound: position among a feature's structure candidates ordered by score,
## which mirrors the curated best_structurePerIdRank. Best rank the compound reached anywhere.
ranks <- function(pass, lab, mx) {
  f <- file.path(RES, pass, sprintf("sirius_full_%s_mix_%s.csv", lab, mx))
  if (!file.exists(f)) return(NULL)
  d <- read.csv(f, stringsAsFactors = FALSE, check.names = FALSE)
  d <- d[!is.na(d$inchiKey) & nzchar(d$inchiKey), ]
  if (!nrow(d)) return(data.frame(k2 = character(0), rank = numeric(0)))
  o <- d[order(d$xcms_fts, -suppressWarnings(as.numeric(d$csiScore))), ]
  o$rk <- ave(seq_len(nrow(o)), o$xcms_fts, FUN = seq_along)
  o$k2 <- two(o$inchiKey); o <- o[o$k2 %in% panelk(mx), ]
  if (!nrow(o)) return(data.frame(k2 = character(0), rank = numeric(0)))
  a <- aggregate(rk ~ k2, o, min); data.frame(k2 = a$k2, rank = a$rk)
}
semi_ranks <- function(lab, mx) {
  d <- semi_tbl(lab); if (is.null(d)) return(NULL)
  d <- d[d$mix == mx & d$k2 %in% panelk(mx) & !is.na(d$rank), ]
  if (!nrow(d)) return(data.frame(k2 = character(0), rank = numeric(0)))
  a <- aggregate(rank ~ k2, d, min); data.frame(k2 = a$k2, rank = a$rank)
}

rowsA <- vapply(LABS, function(l) {
  keep <- Filter(function(mx) !is.null(got("primary", l, mx)) && !is.null(semi_got(l, mx)), MIX)
  f <- function(fn) sum(vapply(keep, fn, 1L))
  S1 <- f(function(mx) length(semi_got(l, mx)));       SA <- f(function(mx) length(semi_got(l, mx, Inf)))
  T1 <- f(function(mx) length(got("primary", l, mx))); TA <- f(function(mx) length(got("primary", l, mx, "any")))
  cs <- ca <- numeric(0)
  for (mx in keep) {
    sr <- semi_ranks(l, mx); ar <- ranks("primary", l, mx)
    cm <- intersect(sr$k2, ar$k2)                       # compounds BOTH workflows found
    if (length(cm)) { cs <- c(cs, sr$rank[match(cm, sr$k2)]); ca <- c(ca, ar$rank[match(cm, ar$k2)]) }
  }
  md  <- function(v) if (!length(v)) "-" else sprintf("%g", median(v))
  mn  <- function(v) if (!length(v)) "-" else sprintf("%.1f", mean(v))
  amn <- length(ca) && length(cs) && mean(ca) <= mean(cs)      # highlight the mean, where the difference shows
  sprintf(paste0('<tr><th>%s</th><td>%d</td><td class="semi">%d</td><td class="semi">%d</td>',
                 '<td>%d</td><td class="tot">%d</td><td>%s</td><td>%s</td>',
                 '<td class="semi">%s</td><td class="semi">%s</td><td class="semi">%s</td><td class="%s">%s</td></tr>'),
          l, length(keep), S1, SA, T1, TA, dfmt(T1 - S1), dfmt(TA - SA),
          md(cs), mn(cs), md(ca), if (amn) "up" else "dn", mn(ca))
}, "")

knitr::asis_output(paste0(
  '<table class="rt"><thead>',
  '<tr><th rowspan="2">lab</th><th rowspan="2">mixtures</th>',
  '<th class="lab" colspan="2">semi-auto</th><th class="lab" colspan="2">all-MS2</th>',
  '<th class="lab" colspan="2">all-MS2 minus semi</th>',
  '<th class="lab" colspan="4">rank, shared compounds</th></tr>',
  '<tr><th>rank-1</th><th>any rank</th><th>top-1</th><th>any candidate</th>',
  '<th>at rank-1</th><th>at any rank</th>',
  '<th>semi median</th><th>semi mean</th><th>all-MS2 median</th><th>all-MS2 mean</th></tr>',
  '</thead><tbody>', paste(rowsA, collapse = ""), '</tbody></table>'))
```

Three things to read off this:

- Relaxing the rank helps the curated workflow just as much, so comparing our any-candidate
  against its rank-1 would overstate the gain by roughly 30 compounds a lab.
- The per-lab verdict survives either way. icl and hmgu win clearly, afekta narrows, and cembio
  depends on MS1.
- The last block is the median and mean rank of the compounds both workflows find, which is the
  only fair basis since they find different sets. Read them together: the **median is 1 on both
  sides in every lab**, so most compounds are ranked first either way, and the workflows are tied
  at the head of the list. The **mean** is where they part, because it feels the tail: all-MS2 is
  lower (better) everywhere except hmgu, sharply so for cembio (3.8 versus 1.7). Semi-auto's
  higher mean is driven by a few compounds ranked very deep, one as far down as 187, so the
  difference is that all-MS2 rarely buries a compound, not that it ranks the easy ones any higher.

A candidate sitting deep in a list is not an identification, so top-1 remains the number to
ship. The any-rank columns measure what better ranking or review could reach.

## cembio: recovering the isotope pattern from the MS1 injection

cembio was annotated without any isotope pattern at all, so SIRIUS had nothing but fragments to
rank formulas with. The reason looked sound: the MS1 scans riding inside cembio's MS2 injection
contain the feature's precursor only about **74%** of the time, and attaching a scan that lacks
the precursor gives SIRIUS a meaningless pattern.

cembio does however acquire a **separate MS1 injection**, and that one is usable. The catch is
that it is not a companion run: the mzML headers put the MS1 injections on 19 March 2025 and the
MS2 injections on 13 May 2025, **55 days apart**, so they are different batches and their
retention times cannot be assumed to correspond.

Measuring the offset directly, on ions strong enough to track in both runs, the median is close
to zero (about 1 to 3 s across test mixtures), so the chromatography is reproducible on the
whole. But the spread is wide: in two of three mixtures the lower quartile sits beyond 100 s.
Some of that is genuine drift and some is ambiguity in choosing an apex when one m/z has several
peaks, and the two cannot be separated from these data.

No global alignment is applied or assumed. Instead, for each feature, five steps:

1. **Pair** the MS2 injection with the MS1 injection of the same polarity and run number.
2. **Search** only the feature's own retention window, padded by 10 s.
3. **Pick** the scan in that window where the precursor is most intense (within 20 ppm); if none
   contains it, attach nothing.
4. **Trim** to the isotope window, from 1.5 Da below the precursor to 7.5 Da above, which drops a
   survey scan from about 600 peaks to about 30.
5. **Keep it only if it is real:** the precursor peak above 1000 counts and an M+1 peak present at
   0.5 to 100% of its height. Otherwise attach nothing.

Two of these need a word. The isotope pattern (step 4) is the precursor M and its M+1, M+2 peaks
from carbon-13; their heights relative to M scale with the carbon count, which is how SIRIUS uses
them to pin the formula. Trimming keeps only these, so a neighbouring ion cannot be misread as
this compound's M+1. And step 5 matters because a false pattern is worse than none: a precursor
with no M+1 tells SIRIUS the molecule has no carbon, demoting the correct formula.

Steps 3 and 5 are also what make the 55-day batch gap tolerable. A retention time that does not
correspond simply yields no precursor in the window, and the feature falls back to fragments. The
pattern attached to 82 to 96% of features per mixture, and the recovery gain below shows it
usually lands on the right scan. Where the two batches genuinely disagree, the window can hold a
different compound at the same mass, which is one plausible reading of the single failure.

The 2_1 mixture shows why the pairing must not be done by editing the filename:

| | filename |
|---|---|
| MS2 injection | `D_1_pos_MS2_010_Human_Endosome_2.1target.mzML` |
| MS1 injection | `D_1_pos_010_Human_Endosome_2.1.mzML` |

Deleting `MS2_` from the first gives a name that does not exist, because of the `target` suffix.
Only mixture 2_1 has it, so it is easy to miss.

```{r ms1}
NOMS1 <- file.path(RES, "primary_noms1")   # the pre-MS1 cembio result, kept for comparison
pre <- function(mx) {
  f <- file.path(NOMS1, sprintf("sirius_full_cembio_mix_%s.csv", mx))
  if (!file.exists(f)) return(NULL)
  d <- read.csv(f, stringsAsFactors = FALSE, check.names = FALSE)
  d <- d[!is.na(d$inchiKey) & nzchar(d$inchiKey), ]; if (!nrow(d)) return(character(0))
  o <- d[order(d$xcms_fts, -suppressWarnings(as.numeric(d$csiScore))), ]
  intersect(panelk(mx), unique(two(o$inchiKey[!duplicated(o$xcms_fts)])))
}
keep <- Filter(function(mx) !is.null(pre(mx)) && !is.null(got("primary", "cembio", mx)), MIX)
m <- do.call(rbind, lapply(keep, function(mx) {
  a <- length(pre(mx)); b <- length(got("primary", "cembio", mx))
  data.frame(mixture = mx, semi = fmt(semi("cembio", mx)),
             `all-scans` = fmt(a), `all-scans + MS1` = fmt(b),
             change = dfmt(b - a), check.names = FALSE)
}))
A <- sum(vapply(keep, function(mx) length(pre(mx)), 1L))
B <- sum(vapply(keep, function(mx) length(got("primary", "cembio", mx)), 1L))
S <- sum(vapply(keep, function(mx) semi("cembio", mx), 1L))
kable(rbind(m, data.frame(mixture = "total", semi = as.character(S), `all-scans` = as.character(A),
                          `all-scans + MS1` = as.character(B), change = dfmt(B - A),
                          check.names = FALSE)),
      format = "html", escape = FALSE, row.names = FALSE, table.attr = 'class="rt"')
```

Attaching it is worth `r sprintf("%+d", B - A)`, taking cembio from `r A` to `r B` and past the
curated workflow's `r S`. It is now cembio's method; the pre-MS1 result is kept in
`results/primary_noms1`.

One compound is still lost, citraconic acid in 2_4. Its feature merges a 92 s window at a single
m/z, so the isotope pattern we attach is a perfectly good one that belongs to a co-eluting
isobar, and SIRIUS follows it to a different formula. That is a limit of the feature grouping
rather than of the MS1 attachment, and fixing it would mean changing the 15 s split shared by
every lab.

# The hmgu group-2 vial mislabel

hmgu's group-2 files do not contain the mixture their name claims. Each holds the next mixture
(`2_N` contains `2_(N+1)`), and `2_10` holds the missing `2_2`. Nothing is lost, and the fix is
to relabel rather than re-run.

## Evidence 1: the annotations

Each file scored against the panel its name claims, and against the next mixture's panel.

```{r hmgu_annot}
nxt <- function(mx) { p <- as.integer(strsplit(mx, "_")[[1]]); sprintf("%d_%d", p[1], p[2] + 1) }
h <- do.call(rbind, lapply(MIX, function(mx) {
  s <- read_pass("primary", "hmgu", mx); if (is.null(s)) return(NULL)
  nx <- if (mx == "2_10") "2_2" else nxt(mx); pk <- panelk(nx)
  own <- length(intersect(panelk(mx), s$top1))
  oth <- if (!length(pk)) NA_integer_ else length(intersect(pk, s$top1))
  shift <- !is.na(oth) && oth > own
  data.frame(`file` = mx, semi = fmt(semi("hmgu", mx)),
             `its own panel` = fmt(own), `the next panel` = fmt(oth),
             `which that is` = if (!length(pk)) "-" else nx,
             `reads as` = if (shift) sprintf("contains %s", nx) else "own label, correct",
             check.names = FALSE)
}))
kable(h, format = "html", escape = FALSE, row.names = FALSE, table.attr = 'class="rt"')
```

Group-2 rows score near zero on their own label and far higher on the next one, and `2_10`
scores highest on `2_2`. Group-1, the `2_1` control and group-3 behave the opposite way, which
is what a correctly labelled file looks like. The curated column agrees throughout, so this is a
real vial mislabel and not an artefact of the all-MS2 pipeline.

## Evidence 2: the raw MS2, independent of the pipeline

This check uses raw MS2 only, with no SIRIUS, no grouping and no part of the annotation
pipeline, so it cannot be an artefact of our code. For each of a mixture's 20 expected
compounds we take its exact mass, add the common adducts, and ask whether any MS2 scan in a
file was triggered on that mass with a real spectrum behind it. Compounds shared between
panels are dropped, so each count belongs to one mixture only.

```{r hmgu_raw, cache = TRUE}
sp_path  <- here::here("1_preprocessing", "hmgu", "HE")
mzml_dir <- file.path(sp_path, "mzml")
seq <- rbind(as.data.frame(read_xlsx(file.path(sp_path, "seq_pos.xlsx"))),
             as.data.frame(read_xlsx(file.path(sp_path, "seq_neg.xlsx"))))
seq$filename <- paste0(seq$`Data File`, ".mzML")
seq <- seq[seq$filename %in% list.files(mzml_dir), ]
seq$label <- chartr(".", "_", sub("target$", "", sub(".*_", "", sub("_MS2$", "", seq$`Data File`))))
mixes <- sprintf("2_%d", 1:10)
p <- 1.007276
adducts <- c(p, 22.989218, 18.033823, 38.963158, -p, 34.969402, 44.998201)

panel_masses <- function(mx) {
  mm <- meta[meta$mix == mx & !is.na(meta$InChIKey), ]
  e2 <- substr(mm$InChIKey, 1, 14)
  dup <- e2[duplicated(e2) | duplicated(e2, fromLast = TRUE)]
  keep <- !(e2 %in% dup)
  as.numeric(mm$M[keep][!duplicated(e2[keep])])
}
load_ms2 <- function(label) {
  fs <- seq$filename[seq$label == label]
  s <- setBackend(Spectra(file.path(mzml_dir, fs), backend = MsBackendMzR()), MsBackendMemory())
  m <- s[s$msLevel == 2L]; m <- m[!is.na(m$precursorMz)]
  list(prec = m$precursorMz,
       bpk = vapply(peaksData(m), function(pp) if (nrow(pp)) max(pp[, "intensity"]) else 0, numeric(1)))
}
strong <- function(d, M) {
  any(vapply(M + adducts, function(a) {
    hit <- which(abs(d$prec - a) <= a * 15 / 1e6)
    length(hit) > 0 && max(d$bpk[hit]) > 1e4
  }, logical(1)))
}
panels <- setNames(lapply(mixes, panel_masses), mixes)
grid <- expand.grid(file = mixes, panel = mixes, stringsAsFactors = FALSE)
grid$n <- NA_integer_
for (fl in mixes) {
  d <- load_ms2(fl)
  for (pn in mixes)
    grid$n[grid$file == fl & grid$panel == pn] <- sum(vapply(panels[[pn]], function(M) strong(d, M), logical(1)))
}
grid$file <- factor(grid$file, levels = mixes); grid$panel <- factor(grid$panel, levels = mixes)
best <- do.call(rbind, lapply(split(grid, grid$file), function(g) g[which.max(g$n), ]))
```

```{r heatmap, fig.width=7, fig.height=6}
suppressMessages(library(ggplot2))
ggplot(grid, aes(panel, file, fill = n)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = n), size = 3) +
  geom_tile(data = best, fill = NA, color = "red", linewidth = 1) +
  scale_fill_gradient(low = "white", high = "#2166AC", name = "expected\ncompounds\nfound") +
  scale_y_discrete(limits = rev(mixes)) +
  labs(x = "panel it was scored against", y = "raw file, as named",
       title = "Red box marks the mixture each file really contains",
       subtitle = "Correct labels would put every red box on the diagonal") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold"))
```

The red boxes sit one step off the diagonal: `2_1` is correct, `2_2` to `2_9` each hold the next
mixture, and `2_10` wraps back to `2_2`.

## Why: the samples ran in the right order

Acquisition timestamps read from the mzML headers, sorted by real time.

```{r injorder}
acq <- function(label) {
  f <- file.path(mzml_dir, seq$filename[seq$label == label & grepl("_pos_", seq$filename)][1])
  con <- file(f, "r"); on.exit(close(con))
  txt <- readLines(con, n = 500, warn = FALSE)
  m <- regmatches(txt, regexpr('startTimeStamp="[^"]+"', txt)); m <- m[nzchar(m)]
  if (length(m)) sub('startTimeStamp="([^"]+)"', "\\1", m[1]) else NA_character_
}
ord <- data.frame(file = mixes, acquired = vapply(mixes, acq, ""),
                  `true mixture` = as.character(best$panel[match(mixes, best$file)]),
                  check.names = FALSE)
ord <- ord[order(ord$acquired), ]
kable(cbind(`order` = seq_len(nrow(ord)), ord), format = "html", row.names = FALSE,
      table.attr = 'class="rt"')
```

Sorted by acquisition time the true mixture column reads `2_1, 2_2, 2_3 ... 2_10`, perfectly
sequential. The instrument ran the mixtures in the correct order and only the vial labels were
permuted: the second injection was labelled `2.10` instead of `2.2`, pushing every later label
down by one. `build_spectral_library.R` already applies the corrected map.
