---
title: "Curated SIRIUS annotations across labs"
format: html
---

The goal here is to create a table with the compound name (adducts separated
by rows) as rows and the lab as columns; the cell content is the
`chrom_peak_id`. The input is the `*_curated.xlsx` sheet for each lab.

```{r}
#| label: setup
#| message: false
library(readxl)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(here)
library(writexl)
```

```{r}
#| label: config
labs <- c("afekta", "cembio", "hmgu", "icl")

curated_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation")
output_dir  <- file.path(curated_dir, "results")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load and unnest

Each curated row carries pipe-separated `chrom_peak_id` and `adduct` values
of equal length. We split them into one row per (compound, adduct, peak).

```{r}
#| label: load-curated
read_curated <- function(lab) {
    f <- file.path(curated_dir,
                   paste0(lab, "_lab_annotations_sirius_curated.xlsx"))
    read_xlsx(f) |>
        mutate(lab = lab)
}

curated <- map_dfr(labs, read_curated)

long <- curated |>
    select(lab, compound = target_ChEBI.name,
           chrom_peak_id, adduct) |>
    mutate(chrom_peak_id = str_split(chrom_peak_id, fixed("|")),
           adduct        = str_split(adduct,        fixed("|"))) |>
    mutate(n_cp = lengths(chrom_peak_id),
           n_ad = lengths(adduct))

# Sanity check: lengths must match within each row
stopifnot(all(long$n_cp == long$n_ad))

long <- long |>
    select(-n_cp, -n_ad) |>
    unnest(c(chrom_peak_id, adduct)) |>
    mutate(across(c(compound, adduct, chrom_peak_id), str_trim))
```

## Pivot to wide

Rows: compound + adduct. Columns: lab. Values: `chrom_peak_id`
(collapsed with `,` if more than one peak per cell).

```{r}
#| label: pivot
wide <- long |>
    group_by(compound, adduct, lab) |>
    summarise(chrom_peak_id = paste(unique(chrom_peak_id), collapse = ","),
              .groups = "drop") |>
    pivot_wider(names_from = lab,
                values_from = chrom_peak_id) |>
    arrange(compound, adduct)

wide
```

## Save

```{r}
#| label: save
out_file <- file.path(output_dir, "curated_annotation_all_labs.xlsx")
write_xlsx(wide, out_file)
out_file
```

