---
title: "SIRIUS Compound Annotation Across Labs"
format: html
---

## Introduction

This document annotates the manually curated chromatographic peaks from all four
labs using SIRIUS via the `RuSirius` R package. The workflow:

1. Load the combined annotation table and filter per lab
2. Extract MS1 and MS2 spectra for annotated peaks
3. Import into SIRIUS and run formula ID + structure DB search
4. Extract results: top-rank (`summary()`) and full-rank (`results()`) per lab
5. Compare SIRIUS annotations across labs against known identities

**Prerequisites:**

- [SIRIUS 6.3+](https://github.com/sirius-ms/sirius/releases) installed
  and running
- A SIRIUS account (free at [bright-giant.com](https://bright-giant.com/))
- `RuSirius` and `RSirius` packages installed
  (see [RuSirius README](https://github.com/rformassspectrometry/RuSirius))

## Setup

```{r}
#| label: setup
#| message: false
library(Spectra)
library(MsExperiment)
library(xcms)
library(RSirius)
library(RuSirius)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

# Source preprocessing helpers (automate_matching_ms2, etc.)
source(here("1_preprocessing", "setup.R"))
```

## Configuration

```{r}
#| label: config
labs <- c("afekta", "cembio", "hmgu", "icl")
study_group <- "HE"

# Waters data flag (ICL uses Waters Synapt → needs estimatePrecursorMz)
waters_labs <- c("icl")

# Path to combined annotation table
ann_path <- here("3_annotation_manual", "HE", "2_combine_annotation",
                 "all_lab_annotations.xlsx")

# Output directory
output_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation", "results")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load Annotations

```{r}
#| label: load-annotations
ann <- read_xlsx(ann_path)

# Clean lab_id column (remove leading underscore from lab_id values)
ann <- ann |>
  mutate(lab = sub("^_", "", lab_id))

# Overview of peaks per lab and polarity
ann |>
  count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n)
```

## Part 1: Spectra Extraction

### Helper function

Extract MS1 + MS2 `Spectra` for a single lab, ready for SIRIUS import.
MS1 spectra are extracted with `chromPeakSpectra()` from `mse`.
MS2 spectra are matched to annotated peaks using `automate_matching_ms2()`
from `mse2`.

```{r}
#| label: helper-extract
extract_spectra_for_sirius <- function(lab_name,
                                       ann,
                                       study_group,
                                       is_waters = FALSE) {
  study_path <- here("1_preprocessing", lab_name, study_group)

  # --- Filter annotations for this lab ---
  lab_ann <- ann |>
    filter(lab == lab_name)

  # --- Load MS1 XcmsExperiment ---
  mse <- readMsObject(
    XcmsExperiment(),
    AlabasterParam(path = file.path(study_path, "mse")),
    spectraPath = file.path(study_path, "mzml")
  )
  sampleData(mse)$mixture <- sub(".*_", "", sampleData(mse)$Sample.Name)
  sampleData(mse)$mixture <- gsub("\\.", "_", sampleData(mse)$mixture)

  # --- Expand pipe-separated chrom_peak_id into individual rows ---
  # all_lab_annotations.xlsx stores compound-level rows with pipe-separated

  # peak IDs (e.g. "CP396003|CP395863|CP636221"). We need one row per peak.
  cp <- chromPeaks(mse)
  sd <- as.data.frame(sampleData(mse))

  peak_df <- lab_ann |>
    mutate(peak_id_list = strsplit(chrom_peak_id, "\\|")) |>
    tidyr::unnest(peak_id_list) |>
    rename(compound_peak_id = chrom_peak_id,
           chrom_peak_id = peak_id_list)

  # Keep only peaks that exist in chromPeaks
  peak_df <- peak_df |>
    filter(chrom_peak_id %in% rownames(cp))

  # Add mz/rt info from chromPeaks table
  cp_info <- as.data.frame(cp[peak_df$chrom_peak_id, ])
  peak_df$mz <- cp_info$mz
  peak_df$mzmin <- cp_info$mzmin
  peak_df$mzmax <- cp_info$mzmax
  peak_df$rt <- cp_info$rt
  peak_df$rtmin <- cp_info$rtmin
  peak_df$rtmax <- cp_info$rtmax

  # Add per-peak polarity and mixture from sample metadata
  peak_df$polarity <- sd$polarity[cp_info$sample]
  peak_df$mixture <- sd$mixture[cp_info$sample]

  peak_df <- as.data.frame(peak_df)
  rownames(peak_df) <- peak_df$chrom_peak_id

  # --- Extract MS1 spectra for annotated peaks ---
  peak_ids <- peak_df$chrom_peak_id
  ms1 <- chromPeakSpectra(
    mse, msLevel = 1L,
    peaks = peak_ids,
    method = "closest_rt"
  )
  # chromPeakSpectra already sets ms1$chrom_peak_id
  # Switch to in-memory backend and materialize processing queue
  ms1 <- setBackend(ms1, MsBackendMemory())
  ms1 <- applyProcessing(ms1)

  # --- Load MS2 XcmsExperiment ---
  # Some labs (e.g. hmgu) use single injection (no separate mse2);
  # in that case, MS2 spectra live in the same mse object.
  mse2_path <- file.path(study_path, "mse2")
  if (dir.exists(mse2_path)) {
    mse2 <- readMsObject(
      XcmsExperiment(),
      AlabasterParam(path = mse2_path),
      spectraPath = file.path(study_path, "mzml")
    )
  } else {
    mse2 <- mse
  }
  # Strip _MS2 suffix (Waters/ICL naming) before extracting mixture
  sn2 <- sub("_MS2$", "", sampleData(mse2)$Sample.Name)
  sampleData(mse2)$mixture <- sub(".*_", "", sn2)
  sampleData(mse2)$mixture <- gsub("\\.", "_", sampleData(mse2)$mixture)

  # --- Extract MS2 spectra matched to annotated peaks ---
  # automate_matching_ms2 needs per-peak rows with: mixture, polarity,
  # rtmin, rtmax, mzmin, mzmax (as rownames = chrom_peak_id)
  ms2 <- automate_matching_ms2(
    mse2 = mse2,
    match_res = peak_df,
    waters_data = is_waters
  )

  # Filter low-intensity MS2 fragments (keep > 5% of base peak)
  low_int <- function(x, ...) x > max(x, na.rm = TRUE) * 0.05
  ms2 <- filterIntensity(ms2, intensity = low_int)
  ms2 <- ms2[lengths(ms2) > 1]
  # Switch to in-memory backend and materialize processing queue
  ms2 <- setBackend(ms2, MsBackendMemory())
  ms2 <- applyProcessing(ms2)

  # --- Combine MS1 and MS2 ---
  ms1_ms2 <- c(ms1, ms2)

  list(
    spectra = ms1_ms2,
    n_ms1 = length(ms1),
    n_ms2 = length(ms2),
    n_peaks = length(peak_ids)
  )
}
```

### Run extraction

```{r}
#| label: extract-spectra
#| message: false
spectra_per_lab <- list()
extraction_summary <- list()

for (lab_name in labs) {
  message("Extracting spectra for: ", lab_name)
  res <- extract_spectra_for_sirius(
    lab_name = lab_name,
    ann = ann,
    study_group = study_group,
    is_waters = lab_name %in% waters_labs
  )
  spectra_per_lab[[lab_name]] <- res$spectra
  extraction_summary[[lab_name]] <- tibble(
    lab = lab_name,
    n_peaks = res$n_peaks,
    n_ms1 = res$n_ms1,
    n_ms2 = res$n_ms2
  )
  message(sprintf("  %s: %d peaks, %d MS1, %d MS2 spectra",
                  lab_name, res$n_peaks, res$n_ms1, res$n_ms2))
}

bind_rows(extraction_summary)
```

## Part 2: SIRIUS Annotation

**NOTE:** SIRIUS must be running before executing this section.
Start SIRIUS 6.3+ and ensure you are logged in.

Each lab gets its own SIRIUS project. Annotation includes:

- Formula identification (SIRIUS)
- Molecular fingerprint prediction (CSI:FingerID)
- Structure database search (BIO database)
- Compound class prediction (CANOPUS)

```{r}
#| label: sirius-annotate
#| eval: false
summary_results <- list()
full_results <- list()

for (lab_name in labs) {
  message("=== Processing: ", lab_name, " ===")

  # --- Create / open SIRIUS project ---
  srs <- Sirius(
    projectId = paste0("ring_trial_", lab_name),
    path = output_dir
  )

  # --- Import spectra ---
  srs <- import(
    sirius = srs,
    spectra = spectra_per_lab[[lab_name]],
    ms_column_name = "chrom_peak_id",
    deleteExistingFeatures = TRUE
  )
  message(sprintf("  Imported %d features", nrow(featuresInfo(srs))))

  # --- Run annotation ---
  # All labs use Q-ToF instruments
  job_id <- run(
    srs,
    fallbackAdducts = c("[M + H]+", "[M + Na]+", "[M + K]+",
                        "[M - H]-", "[M + Cl]-"),
    formulaIdParams = formulaIdParam(
      numberOfCandidates = 5,
      instrument = "QTOF",
      numberOfCandidatesPerIonization = 2,
      massAccuracyMS2ppm = 10,
      filterByIsotopePattern = FALSE,
      isotopeMs2Settings = "SCORE",
      performDeNovoBelowMz = 600
    ),
    predictParams = predictParam(),
    structureDbSearchParams = structureDbSearchParam(
      structureSearchDbs = c("BIO")
    ),
    recompute = TRUE,
    wait = TRUE
  )
  message(sprintf("  Job %s completed", job_id))

  # --- Extract top-rank summary ---
  summary_results[[lab_name]] <- summary(srs, result.type = "structure") |>
    mutate(lab = lab_name)

  # --- Extract full results (top 5 formulas × top 3 structures) ---
  full_results[[lab_name]] <- results(
    srs,
    result.type = "structureDb",
    topFormula = 5,
    topStructure = 3,
    return.type = "data.frame"
  ) |>
    mutate(lab = lab_name)

  # --- Save per-lab CSVs ---
  write.csv(
    summary_results[[lab_name]],
    file.path(output_dir, paste0("sirius_summary_", lab_name, ".csv")),
    row.names = FALSE
  )
  write.csv(
    full_results[[lab_name]],
    file.path(output_dir, paste0("sirius_full_", lab_name, ".csv")),
    row.names = FALSE
  )

  # Shut down this project's connection
  shutdown(srs)
  message(sprintf("  Results saved for %s\n", lab_name))
}
```

## Part 3: Combine and Compare

### Merge with ground truth

```{r}
#| label: combine-results
#| eval: false
all_summaries <- bind_rows(summary_results)
all_full <- bind_rows(full_results)

# Build a peak-level lookup from the pipe-separated annotation table
ann_expanded <- ann |>
  mutate(peak_id_list = strsplit(chrom_peak_id, "\\|")) |>
  tidyr::unnest(peak_id_list) |>
  select(lab, chrom_peak_id = peak_id_list,
         target_ChEBI.name, target_formula,
         target_InChIKey, adduct, polarity, mixture)

# Map SIRIUS externalFeatureId back to the original annotation info
all_summaries <- all_summaries |>
  left_join(
    ann_expanded,
    by = c("lab", "externalFeatureId" = "chrom_peak_id")
  )

# Save combined tables
write.csv(all_summaries,
          file.path(output_dir, "sirius_summary_all_labs.csv"),
          row.names = FALSE)
write.csv(all_full,
          file.path(output_dir, "sirius_full_all_labs.csv"),
          row.names = FALSE)
```

### Annotation success rate

Check whether SIRIUS top-rank molecular formula matches the known formula.

```{r}
#| label: success-rate
#| eval: false
comparison <- all_summaries |>
  mutate(
    confidenceApproxMatch = as.numeric(confidenceApproxMatch),
    confidenceExactMatch = as.numeric(confidenceExactMatch),
    formula_match = molecularFormula == target_formula,
    has_structure = !is.na(structureName)
  )

comparison |>
  group_by(lab) |>
  summarise(
    n_features = n(),
    n_formula_match = sum(formula_match, na.rm = TRUE),
    pct_formula = round(mean(formula_match, na.rm = TRUE) * 100, 1),
    n_with_structure = sum(has_structure, na.rm = TRUE),
    mean_confidence = round(
      mean(confidenceApproxMatch[is.finite(confidenceApproxMatch)],
           na.rm = TRUE), 3),
    .groups = "drop"
  )
```

### Confidence score distribution

```{r}
#| label: confidence-plot
#| eval: false
ggplot(
  comparison |> filter(is.finite(confidenceApproxMatch)),
  aes(x = lab, y = confidenceApproxMatch, fill = lab)
) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "SIRIUS Confidence Scores by Lab",
    x = "Lab",
    y = "Approximate Confidence"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Compound-level cross-lab agreement

For each standard compound, check whether all labs produce the same top
molecular formula prediction.

```{r}
#| label: cross-lab-agreement
#| eval: false
compound_agreement <- comparison |>
  group_by(target_ChEBI.name) |>
  summarise(
    n_labs = n_distinct(lab),
    formulas = paste(unique(na.omit(molecularFormula)), collapse = "; "),
    n_unique_formulas = n_distinct(molecularFormula, na.rm = TRUE),
    all_agree = n_distinct(molecularFormula, na.rm = TRUE) == 1,
    mean_confidence = round(mean(confidenceApproxMatch, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  filter(n_labs > 1) |>
  arrange(desc(n_labs), desc(all_agree))

compound_agreement
```


## Notes for collaborators

- The spectra extraction (Part 1) does not require SIRIUS. It can be run
  independently to verify that spectra are correctly matched.
- SIRIUS annotation (Part 2) requires a running SIRIUS instance and may take
  significant time depending on the number of features.
- The `eval: false` chunks in Parts 2–3 must be set to `eval: true` (or run
  interactively) once SIRIUS is available.
- Per-lab results are saved as CSVs in the `results/` subfolder.
- The `automate_matching_ms2()` function from `setup.R` matches MS2 spectra
  to annotated MS1 peaks by m/z and RT proximity. It works for all labs
  (both separate and single injection workflows).

## Session Info

```{r}
sessionInfo()
```
