---
title: "SIRIUS annotation - all MS2 per lab (strategy 3)"
format:
  html:
    fig-width: 12
    fig-height: 8
---

## Goal

Annotate **all MS2-bearing features** from every lab, without requiring
cross-lab consensus, and emit the two canonical CSVs defined in
[`5_downstream_analysis/CANONICAL_SCHEMAS.md`](../../../5_downstream_analysis/CANONICAL_SCHEMAS.md):

- `annotation_identity.csv` — one row per `(mixture, lab, compound,
  polarity)` with best ranks, adducts, COSMIC scores, and a
  `confidence_tier` placeholder (`"unfiltered"`; P5 overwrites).
- `annotation_chrom_metrics.csv` — one row per `(mixture, lab,
  polarity, compound, adduct, source)` with `areaUnderTic, xicFwhm,
  maxIntensity, apex_rt` from the two-pass EIC pipeline.

This file is a **pure producer**: no cross-lab comparison plots, no
abundance / RT / presence analysis. Those live in
`5_downstream_analysis/9_per_compound_diagnostics.qmd`.

## Workflow overview

```{mermaid}
flowchart LR
  A[Raw mzML] --> B[MS2 noise filter<br/>5% base peak OR 1000 cnt]
  B --> C{Lab type}
  C -->|cembio| D[Group MS2 by<br/>precursor m/z]
  C -->|DDA<br/>afekta hmgu icl| E[fragmentGroupIndex<br/>MS1 with co-eluting MS2]
  D --> F[Batch 200 features]
  E --> F
  F --> G[SIRIUS<br/>formula ID + CSI:FingerID + BIO DB]
  G --> H[Per batch CSV cache]
  H --> I[Aggregate across<br/>labs and mixtures]
  I --> J[Match to<br/>expected compounds]
  J --> K[MS1 EIC extraction<br/>two-pass]
  K --> L[Write canonical CSVs<br/>identity + chrom_metrics]
```

## Setup

```{r}
#| label: setup
#| message: false
library(Spectra)
library(MsExperiment)
library(xcms)
library(RuSirius)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(MsQuality)
library(tidytext)

source(here("1_preprocessing", "setup.R"))
```

## Config

```{r}
#| label: config
labs        <- c("afekta", "cembio", "hmgu", "icl")
study_group <- "HE"
mixtures    <- c("1.1", "1.2", "1.3", "1.4", "1.5", "1.6",
                 "2.1", "2.2", "2.3", "2.4", "2.5",
                 "2.6", "2.7", "2.8", "2.9", "2.10",
                 "3.1", "3.2", "3.3")

mix_us     <- function(m) gsub("\\.", "_", m)
output_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                   "results_all_ms2")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Extraction helper

`extract_all_ms2()` loads the raw mzML files for one (lab, mixture) and
returns the MS1+MS2 spectra ready for SIRIUS, split by polarity.

**Acquisition layouts.** afekta, cembio and icl have dedicated `_MS2`
files; HMGU acquires MS1+MS2 together in the regular sample file. For
DDA labs (afekta, hmgu, icl) features are formed by `fragmentGroupIndex()`
(one MS1 + its triggered MS2). Cembio acquires MS1 and MS2 in temporally
disjoint blocks within one file with a targeted inclusion list, so
co-elution-based pairing is impossible. Cembio features are formed by
**RT-proximity grouping**: two MS2 of the same precursor m/z belong to
the same feature iff their RT gap is ≤ `cembio_block_gap_s` (default
30 s). This replaces a legacy `rle(precursorMz)` approach that started
a new feature every time another inclusion-list precursor fired
between two MS2 of the same precursor — shattering one chromatographic
peak into many features (e.g. ritalinic acid in mix 2.9 pos: one
inclusion-list scan window at m/z 220.1335 fired 11 MS2 at 5-9 s
intervals interleaved with other precursors → legacy `rle` made
11 features; proximity grouping correctly makes 1).

**Noise filter.** After loading we apply a per-MS2 intensity filter:
keep peaks with intensity ≥ 5% of the base peak OR ≥ 1000 counts. This
cuts MS2 peak counts ~4× on the noisy late-RT hmgu batches (median 91
→ 21 peaks/MS2) while leaving normal spectra essentially untouched. The
OR floor rescues mid-intensity fragments when the base peak is unusually
tall.

```{r}
#| label: helper-extract-all
extract_all_ms2 <- function(lab_name, study_group, mixture) {
  study_path <- here("1_preprocessing", lab_name, study_group)
  mzml_dir   <- file.path(study_path, "mzml")
  mix_label  <- mix_us(mixture)

  empty_result <- function(sp = Spectra(backend = MsBackendMemory())) {
    list(spectra_pos = sp, spectra_neg = sp,
         n_ms2_pos = 0L, n_ms2_neg = 0L,
         n_features_pos = 0L, n_features_neg = 0L)
  }

  seq_pos <- read_xlsx(file.path(study_path, "seq_pos.xlsx")) |> as.data.frame()
  seq_neg <- read_xlsx(file.path(study_path, "seq_neg.xlsx")) |> as.data.frame()
  seq_pos$polarity <- "pos"; seq_neg$polarity <- "neg"
  seq <- rbind(seq_pos, seq_neg)
  seq$filename <- paste0(seq$`Data File`, ".mzML")
  seq <- seq[seq$filename %in% list.files(mzml_dir), ]
  # Mixture is the last underscore-delimited token of the file basename,
  # after stripping a trailing `_MS2` and any `target` suffix (used by
  # cembio for targeted-MS2 acquisitions, e.g. "..._2.1target.mzML").
  seq$mixture <- gsub("\\.", "_",
                      sub("target$", "",
                          sub(".*_", "",
                              sub("_MS2$", "", seq$`Data File`))))

  has_ms2_files <- any(grepl("MS2", seq$filename, fixed = TRUE))
  seq <- if (has_ms2_files) {
    seq[grepl("MS2", seq$filename, fixed = TRUE) & seq$mixture == mix_label, ]
  } else {
    seq[seq$mixture == mix_label, ]
  }
  if (nrow(seq) == 0L) return(empty_result())

  sp <- Spectra(file.path(mzml_dir, seq$filename), backend = MsBackendMzR())
  sp <- setBackend(sp, MsBackendMemory())
  sp <- applyProcessing(sp)

  sp <- filterIntensity(sp, intensity = function(x) {
    if (length(x) == 0L) return(logical(0))
    bp <- max(x, na.rm = TRUE)
    if (!is.finite(bp) || bp <= 0) return(rep(TRUE, length(x)))
    x >= bp * 0.05 | x >= 1000
  }, msLevel. = 2L)

  if (lab_name == "cembio") {
    sp <- sp[sp$msLevel == 2L]
    sp <- filterEmptySpectra(sp)
    if (length(sp) == 0L) return(empty_result(sp))

    # Proximity-aware grouping for cembio's block-MS2 acquisition.
    #
    # Legacy `rle(precursorMz)` started a new feature every time another
    # inclusion-list precursor fired in between two MS2 of the same
    # precursor, shattering one chromatographic peak into many "features".
    # E.g. cembio mix 2.9 ritalinic acid pos: one inclusion-list scan
    # window at m/z 220.1335 produced 11 MS2 firings 5-9 s apart between
    # MS2 of other precursors → 11 separate group_idx values.
    #
    # New rule: sort by (file, precursorMz, RT). Two MS2 of the same
    # precursor m/z in the same file are in the same group iff their
    # RT gap is <= `cembio_block_gap_s`. This correctly collapses
    # interleaved firings of the same precursor while still splitting
    # truly distinct chromatographic peaks of the same m/z at well-
    # separated RTs.
    cembio_block_gap_s <- 30

    key       <- paste0(sp$dataOrigin, "|", sprintf("%.4f", sp$precursorMz))
    ord       <- order(key, sp$rtime)
    key_s     <- key[ord]
    rt_s      <- sp$rtime[ord]

    starts_new_group <- c(TRUE,
                          key_s[-1L] != key_s[-length(key_s)] |
                          (rt_s[-1L] - rt_s[-length(rt_s)]) > cembio_block_gap_s)
    grp_sorted <- cumsum(starts_new_group)

    grp <- integer(length(sp))
    grp[ord] <- grp_sorted
    sp$group_idx <- grp
  } else {
    keep <- unlist(lapply(split(seq_along(sp), sp$dataOrigin), function(idx) {
      first_ms1 <- which(sp$msLevel[idx] == 1L)[1L]
      if (is.na(first_ms1)) integer(0) else idx[first_ms1:length(idx)]
    }))
    sp <- sp[sort(keep)]
    sp <- filterEmptySpectra(sp)
    if (length(sp) == 0L) return(empty_result(sp))
    sp$group_idx <- fragmentGroupIndex(sp)
    sp <- sp[duplicated(sp$group_idx) | duplicated(sp$group_idx, fromLast = TRUE)]
    if (length(sp) == 0L) return(empty_result(sp))
  }

  list(
    spectra_pos    = sp[sp$polarity == 1L],
    spectra_neg    = sp[sp$polarity == 0L],
    n_ms2_pos      = sum(sp$msLevel == 2L & sp$polarity == 1L),
    n_ms2_neg      = sum(sp$msLevel == 2L & sp$polarity == 0L),
    n_features_pos = length(unique(sp$group_idx[sp$polarity == 1L])),
    n_features_neg = length(unique(sp$group_idx[sp$polarity == 0L]))
  )
}
```

## 2. SIRIUS helpers

Path helpers for the per-(lab, mixture) cache files, and a single list of
ID columns that SIRIUS returns inconsistently typed (sometimes numeric,
sometimes character). We coerce all of them to character on every read /
write so `bind_rows` across batches and mixtures never fails on type
mismatch. Add new column names to `sirius_id_cols` if a future error
surfaces one.

```{r}
#| label: helpers
sirius_sum_path       <- function(lab, m) file.path(output_dir, sprintf("sirius_summary_%s_mix_%s.csv",  lab, mix_us(m)))
sirius_full_path      <- function(lab, m) file.path(output_dir, sprintf("sirius_full_%s_mix_%s.csv",     lab, mix_us(m)))
extraction_count_path <- function(lab, m) file.path(output_dir, sprintf("extraction_counts_%s_mix_%s.csv", lab, mix_us(m)))

sirius_id_cols <- c("alignedFeatureId", "compoundId", "externalFeatureId",
                    "formulaId", "xcms_fts", "sirius_fts",
                    "molecularFormulaId", "structureCandidateInChIKey")
```

```{r}
#| label: sirius-login
sirius_user <- "plouail@eurac.edu"
sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")
```

### `run_sirius_for_lab()`

Runs SIRIUS over one (lab, mixture)'s spectra in batches of `batch_size`
features, caching per-batch results to CSV so re-renders skip completed
batches. The function manages the SIRIUS server lifecycle internally and
**restarts the JVM every `restart_every` live batches** - a long-running
JVM accumulates state that eventually crashes `import()`, so bouncing the
process clears it (the on-disk project survives the restart).

Each batch also cross-checks `featuresId(srs, type = "xcms")` against the
features intended for import and re-imports anything silently dropped
(up to two retries; bail when retries stop making progress).

Pos and neg features go into the same project per batch so RuSirius
assigns polarity correctly. `numberOfCandidates = 10`/`3` (per ionization)
is a deliberate cut from the default 15/5 - each candidate triggers a
CSI:FingerID + BIO lookup, the dominant per-feature cost.

```{r}
#| label: run-sirius-for-lab
run_sirius_for_lab <- function(sp_pos, sp_neg,
                               username, password, projectId, path,
                               batch_size    = 200,
                               cache_dir     = NULL,
                               cache_prefix  = NULL,
                               restart_every = 7L) {
  if (length(sp_pos) == 0L && length(sp_neg) == 0L)
    return(list(summary = NULL, full = NULL))

  start_sirius <- function() Sirius(username = username, password = password,
                                    projectId = projectId, path = path)
  srs <- start_sirius()
  on.exit(try(shutdown(srs), silent = TRUE), add = TRUE)

  groups_pos <- unique(sp_pos$group_idx)
  groups_neg <- unique(sp_neg$group_idx)
  pol_lookup <- c(
    setNames(rep("pos", length(groups_pos)), as.character(groups_pos)),
    setNames(rep("neg", length(groups_neg)), as.character(groups_neg))
  )

  all_groups <- c(groups_pos, groups_neg)
  batches    <- split(all_groups, ceiling(seq_along(all_groups) / batch_size))
  has_cache  <- !is.null(cache_dir) && !is.null(cache_prefix)

  sums         <- vector("list", length(batches))
  fulls        <- vector("list", length(batches))
  live_batches <- 0L

  for (i in seq_along(batches)) {
    batch_sum_path  <- if (has_cache) file.path(cache_dir, sprintf("%s_batch%d_sum.csv",  cache_prefix, i))
    batch_full_path <- if (has_cache) file.path(cache_dir, sprintf("%s_batch%d_full.csv", cache_prefix, i))

    message("  batch ", i, "/", length(batches))

    if (has_cache && file.exists(batch_sum_path) && file.exists(batch_full_path)) {
      message("    → cached, skipping")
      sums[[i]]  <- read.csv(batch_sum_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
               across(any_of(sirius_id_cols), as.character))
      fulls[[i]] <- read.csv(batch_full_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(sirius_id_cols), as.character))
      next
    }

    if (live_batches > 0L && live_batches %% restart_every == 0L) {
      message("    → restarting SIRIUS after ", live_batches, " live batches")
      try(shutdown(srs), silent = TRUE)
      srs <- start_sirius()
    }

    b_pos <- sp_pos[sp_pos$group_idx %in% batches[[i]]]
    b_neg <- sp_neg[sp_neg$group_idx %in% batches[[i]]]
    n_pos <- length(unique(b_pos$group_idx))
    n_neg <- length(unique(b_neg$group_idx))
    message("    ", n_pos, " pos + ", n_neg, " neg features")

    if (n_pos > 0)
      srs <- import(srs, spectra = b_pos, ms_column_name = "group_idx",
                    deleteExistingFeatures = TRUE)
    if (n_neg > 0)
      srs <- import(srs, spectra = b_neg, ms_column_name = "group_idx",
                    deleteExistingFeatures = n_pos == 0)

    expected_ids <- as.character(c(unique(b_pos$group_idx),
                                   unique(b_neg$group_idx)))
    project_ids  <- as.character(featuresId(srs, type = "xcms"))
    missing_ids  <- setdiff(expected_ids, project_ids)
    for (retry in 1:2) {
      if (length(missing_ids) == 0L) break
      message("    !! ", length(missing_ids), " missing - retry ", retry)
      miss_pos <- b_pos[b_pos$group_idx %in% missing_ids]
      miss_neg <- b_neg[b_neg$group_idx %in% missing_ids]
      if (length(miss_pos) > 0L)
        srs <- import(srs, spectra = miss_pos, ms_column_name = "group_idx",
                      deleteExistingFeatures = FALSE)
      if (length(miss_neg) > 0L)
        srs <- import(srs, spectra = miss_neg, ms_column_name = "group_idx",
                      deleteExistingFeatures = FALSE)
      prev_n      <- length(missing_ids)
      missing_ids <- setdiff(expected_ids,
                             as.character(featuresId(srs, type = "xcms")))
      if (length(missing_ids) >= prev_n) {
        message("    !! retry made no progress - giving up")
        break
      }
    }
    if (length(missing_ids) > 0L)
      warning("    !! ", length(missing_ids), " feature(s) never imported: ",
              paste(head(missing_ids, 10), collapse = ","))

    run(srs,
      formulaIdParams = formulaIdParam(
        numberOfCandidates              = 10,
        instrument                      = "QTOF",
        numberOfCandidatesPerIonization = 3,
        massAccuracyMS2ppm              = 10,
        filterByIsotopePattern          = FALSE,
        isotopeMs2Settings              = "SCORE",
        performDeNovoBelowMz            = 600),
      predictParams           = predictParam(),
      structureDbSearchParams = structureDbSearchParam(structureSearchDbs = c("BIO")),
      recompute = TRUE, wait = TRUE)

    sums[[i]] <- summary(srs, result.type = "structure") |>
      mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
             across(any_of(sirius_id_cols), as.character),
             polarity = pol_lookup[as.character(externalFeatureId)])

    fulls[[i]] <- results(srs, result.type = "structureDb",
                          topFormula = 8, topStructure = 20,
                          return.type = "data.frame") |>
      mutate(across(any_of(sirius_id_cols), as.character),
             polarity = pol_lookup[as.character(xcms_fts)])

    if (has_cache) {
      write.csv(sums[[i]],  batch_sum_path,  row.names = FALSE)
      write.csv(fulls[[i]], batch_full_path, row.names = FALSE)
    }
    live_batches <- live_batches + 1L
  }

  non_empty <- function(x) Filter(\(df) nrow(df) > 0L, x)
  list(summary = bind_rows(non_empty(sums)),
       full    = bind_rows(non_empty(fulls)))
}
```

## 3. Extract + run SIRIUS per (mixture, lab)

Each (mixture, lab) is processed end-to-end before moving to the next:
extract → run SIRIUS → write cache → drop spectra. Memory stays bounded
to one lab's spectra at a time, and a partial run resumes simply by
re-rendering - cached pairs skip both extraction and SIRIUS.

```{r}
#| label: extract-and-sirius
summary_results    <- setNames(vector("list", length(mixtures)), mixtures)
full_results       <- setNames(vector("list", length(mixtures)), mixtures)
extraction_summary <- list()

for (m in mixtures) {
  summary_results[[m]] <- list()
  full_results[[m]]    <- list()
  for (lab in labs) {
    sum_path  <- sirius_sum_path(lab, m)
    full_path <- sirius_full_path(lab, m)

    if (file.exists(sum_path) && file.exists(full_path)) {
      message("Cached: ", lab, " / mix ", m)
      summary_results[[m]][[lab]] <- read.csv(sum_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
               across(any_of(c(sirius_id_cols, "lab", "polarity", "adduct")), as.character))
      full_results[[m]][[lab]] <- read.csv(full_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c(sirius_id_cols, "lab", "polarity", "adduct")), as.character))
      if (file.exists(extraction_count_path(lab, m)))
        extraction_summary[[paste(m, lab)]] <-
          read.csv(extraction_count_path(lab, m),
                   stringsAsFactors = FALSE,
                   colClasses = c(mixture = "character",
                                  lab     = "character")) |>
          as_tibble()
      next
    }

    message("Extracting: ", lab, " / mix ", m)
    sp <- extract_all_ms2(lab, study_group, m)
    extraction_summary[[paste(m, lab)]] <- tibble(
      mixture = m, lab = lab,
      n_features_pos = sp$n_features_pos, n_features_neg = sp$n_features_neg,
      n_ms2_pos      = sp$n_ms2_pos,      n_ms2_neg      = sp$n_ms2_neg)
    write.csv(extraction_summary[[paste(m, lab)]],
              extraction_count_path(lab, m), row.names = FALSE)

    if (length(sp$spectra_pos) == 0L && length(sp$spectra_neg) == 0L) {
      message("  no MS2 - writing empty cache")
      summary_results[[m]][[lab]] <- tibble(lab = character(0))
      full_results[[m]][[lab]]    <- tibble(lab = character(0))
      write.csv(summary_results[[m]][[lab]], sum_path,  row.names = FALSE)
      write.csv(full_results[[m]][[lab]],    full_path, row.names = FALSE)
      rm(sp); gc(); next
    }

    if (!exists("sirius_user")) sirius_user <- Sys.getenv("SIRIUS_USERNAME")
    if (!exists("sirius_pass")) sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")

    res <- run_sirius_for_lab(
      sp_pos       = sp$spectra_pos,
      sp_neg       = sp$spectra_neg,
      username     = sirius_user, password = sirius_pass,
      projectId    = paste0("ring_trial_allms2_", lab, "_mix_", mix_us(m)),
      path         = output_dir,
      cache_dir    = output_dir,
      cache_prefix = paste0(lab, "_mix_", mix_us(m)))

    summary_results[[m]][[lab]] <-
      if (is.null(res$summary)) tibble(lab = character(0))
      else res$summary |> mutate(lab = lab)
    full_results[[m]][[lab]] <-
      if (is.null(res$full)) tibble(lab = character(0))
      else res$full |> mutate(lab = lab)

    write.csv(summary_results[[m]][[lab]], sum_path,  row.names = FALSE)
    write.csv(full_results[[m]][[lab]],    full_path, row.names = FALSE)
    rm(sp, res); gc()
  }
}

bind_rows(extraction_summary)
```

Extraction-summary bar plot per lab × polarity × mixture.

```{r}
#| label: plot-extraction
#| fig-cap: "SIRIUS features per lab × polarity × mixture (MS1+MS2 group; MS2 only for cembio)"
if (length(extraction_summary) > 0) {
  bind_rows(extraction_summary) |>
    select(lab, mixture, pos = n_features_pos, neg = n_features_neg) |>
    pivot_longer(c(pos, neg), names_to = "polarity", values_to = "n") |>
    ggplot(aes(x = lab, y = n, fill = polarity)) +
    geom_col(position = "dodge") +
    facet_wrap(~ mixture) +
    scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
    labs(x = NULL, y = "SIRIUS features", fill = "Polarity") +
    theme_bw()
}
```

Aggregate to long tables (`all_summaries`, `all_full`) keyed by mixture
for everything downstream.

```{r}
#| label: aggregate
# Skip 0-row tibbles inside each mixture's per-lab list before bind_rows: an
# empty cache from a (lab, mixture) with no MS2 features writes a header-only
# CSV, which read.csv parses with logical-typed columns. Combining those with
# real character/numeric columns would crash bind_rows.
agg <- function(results_list) {
  bind_rows(lapply(mixtures, function(m) {
    non_empty <- Filter(\(df) nrow(df) > 0L, results_list[[m]])
    if (length(non_empty) == 0L) return(NULL)
    bind_rows(non_empty) |> mutate(mixture = m)
  }))
}
all_summaries <- agg(summary_results)
all_full      <- agg(full_results)

write.csv(all_summaries, file.path(output_dir, "sirius_summary_all_labs_all_mixtures.csv"), row.names = FALSE)
write.csv(all_full,      file.path(output_dir, "sirius_full_all_labs_all_mixtures.csv"),    row.names = FALSE)
```

### Feature-import verification

`run_sirius_for_lab()` already verifies each batch in-flight. This is the
end-to-end check: per (lab, mixture, polarity) we compare features
**extracted** upstream vs features SIRIUS produced **any** result for
(structure summary ∪ structureDb result). A residual `missing > 0` row
means SIRIUS imported the feature but produced no annotation - or, if a
warning fired during the loop, the retry import never succeeded.

```{r}
#| label: verify-import
#| include: false
extracted_long <- bind_rows(extraction_summary) |>
  select(mixture, lab, pos = n_features_pos, neg = n_features_neg) |>
  pivot_longer(c(pos, neg), names_to = "polarity", values_to = "extracted")

annotated_long <- bind_rows(
  all_summaries |> filter(!is.na(externalFeatureId)) |>
    transmute(mixture, lab, polarity, feature_id = externalFeatureId),
  all_full      |> filter(!is.na(xcms_fts)) |>
    transmute(mixture, lab, polarity, feature_id = xcms_fts)
) |>
  group_by(mixture, lab, polarity) |>
  summarise(annotated = n_distinct(feature_id), .groups = "drop")

verify_df <- extracted_long |>
  left_join(annotated_long, by = c("mixture", "lab", "polarity")) |>
  mutate(annotated = coalesce(annotated, 0L),
         missing   = extracted - annotated,
         pct_kept  = ifelse(extracted == 0, NA_real_,
                            round(100 * annotated / extracted, 1)))

write.csv(verify_df,
          file.path(output_dir, "feature_import_verification.csv"),
          row.names = FALSE)

verify_df |> arrange(desc(missing)) |> knitr::kable()

cat("Entries with missing > 0:", sum(verify_df$missing > 0, na.rm = TRUE),
    "  |  total extracted:",  sum(verify_df$extracted, na.rm = TRUE),
    "  annotated:",            sum(verify_df$annotated, na.rm = TRUE),
    "  missing:",              sum(verify_df$missing,   na.rm = TRUE), "\n")
```

## 4. Match to expected compounds (all mixtures)

```{r}
#| label: expected
expected_all <- meta |>
  filter(Mixture %in% mixtures) |>
  transmute(mixture              = Mixture,
            compound             = `ChEBI name`,
            expected_formula     = formula,
            expected_inchikey    = InChIKey,
            expected_inchikey_2d = substr(InChIKey, 1, 14))

expected_all |>
  count(mixture, name = "n_expected") |>
  knitr::kable(caption = "Expected compounds per mixture")
```

### Ambiguous compounds in the expected list

Two patterns are checked here before the coverage analysis.

**(a) Within-mixture stereoisomers (same 2D InChIKey).** SIRIUS returns
only the 2D InChIKey, so two compounds sharing those 14 characters in
the same mixture cannot be told apart from the annotation. We **drop
both compounds** from `expected_all` rather than over-counting one of
them.

**(b) Same compound listed in multiple mixtures.** Informational only -
the join below keys on `mixture`, so each annotation is attributed to
its own mixture and these duplicates don't double count.

```{r}
#| label: expected-stereoisomers
exp_dups_2d <- expected_all |>
  group_by(mixture, expected_inchikey_2d) |>
  filter(n() > 1) |>
  arrange(mixture, expected_inchikey_2d, compound) |>
  ungroup()

knitr::kable(exp_dups_2d,
  caption = "(a) Stereoisomer pairs - DROPPED from expected_all")

expected_all <- expected_all |>
  anti_join(exp_dups_2d, by = c("mixture", "expected_inchikey_2d"))
```

```{r}
#| label: expected-cross-mixture
exp_dups_mix <- expected_all |>
  group_by(expected_inchikey) |>
  filter(n_distinct(mixture) > 1) |>
  arrange(expected_inchikey, mixture) |>
  ungroup()

knitr::kable(exp_dups_mix,
  caption = "(b) Same compound across multiple mixtures (kept)")

all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  filter(!is.na(inchiKey)) |>
  semi_join(exp_dups_mix |> distinct(mixture, expected_inchikey_2d),
            by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  count(mixture, lab, polarity, inchikey_2d) |>
  knitr::kable(caption = "Were the cross-mixture duplicates annotated?")
```

A SIRIUS hit counts as recovery of an expected compound when its 2D
InChIKey matches the expected one **for the same mixture**. Per
(mixture, lab, compound, polarity) we record the best formula rank, best
structure rank and the union of adducts seen.

```{r}
#| label: match-structure
structure_hits_all <- all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  inner_join(
    expected_all |>
      select(mixture, expected_inchikey_2d, compound,
             expected_formula, expected_inchikey),
    by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  group_by(mixture, lab, inchikey_2d, polarity) |>
  summarise(
    best_formula_rank   = suppressWarnings(min(rank.x, na.rm = TRUE)),
    best_structure_rank = suppressWarnings(min(rank.y, na.rm = TRUE)),
    adducts             = paste(sort(unique(na.omit(adduct))), collapse = ";"),
    molecularFormula    = first(molecularFormula),
    compound            = first(compound),
    expected_formula    = first(expected_formula),
    expected_inchikey   = first(expected_inchikey),
    .groups = "drop") |>
  mutate(across(c(best_formula_rank, best_structure_rank),
                \(x) ifelse(is.infinite(x), NA_real_, x)))

structure_hits_all
```

## 5. Coverage

```{r}
#| label: coverage
covered_all <- expected_all |>
  left_join(structure_hits_all |>
              distinct(mixture, inchikey_2d) |>
              mutate(structure_recovered = TRUE),
            by = c("mixture", "expected_inchikey_2d" = "inchikey_2d")) |>
  mutate(structure_recovered = !is.na(structure_recovered))

covered_all |>
  group_by(mixture) |>
  summarise(n_recovered = sum(structure_recovered),
            n_expected  = n(), .groups = "drop") |>
  knitr::kable(caption = "Structure recovery per mixture")
```

```{r}
#| label: plot-coverage-by-lab
#| fig-cap: "Structure hits per lab × polarity, faceted by mixture"
#| fig-height: 8
structure_hits_all |>
  count(mixture, lab, polarity) |>
  ggplot(aes(x = lab, y = n, fill = polarity)) +
  geom_col(position = "dodge") +
  facet_wrap(~ mixture) +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "Structure hits", fill = "Polarity") +
  theme_bw()
```

## 6. RT extraction for EIC pipeline

RT per feature is extracted from the MS2 spectra and cached per (lab,
mixture) so raw files are only read once. `hit_rt` joins the RT onto
expected-compound matches and is the input to §7.1's `eic_params`. No
cross-lab RT analysis here — that lives in `9_per_compound_diagnostics.qmd`.

```{r}
#| label: extract-rt
#| message: false
rt_all <- bind_rows(lapply(mixtures, function(m) {
  bind_rows(lapply(labs, function(lab) {
    path <- file.path(output_dir, paste0(lab, "_mix_", mix_us(m), "_rt.csv"))
    if (file.exists(path)) {
      df <- read.csv(path, stringsAsFactors = FALSE)
      if ("rt_min" %in% names(df) && !"rt_sec" %in% names(df))
        df <- mutate(df, rt_sec = rt_min * 60)
      return(df |> mutate(lab = lab, mixture = m))
    }
    message("Extracting RT for: ", lab, " / mix ", m)
    res <- extract_all_ms2(lab, study_group, m)
    make_rt_tbl <- function(sp) {
      if (length(sp) == 0L)
        tibble(group_idx = integer(), msLevel = integer(), rt_sec = numeric())
      else
        tibble(group_idx = sp$group_idx, msLevel = sp$msLevel, rt_sec = sp$rtime)
    }
    rt_df <- bind_rows(make_rt_tbl(res$spectra_pos),
                       make_rt_tbl(res$spectra_neg)) |>
      filter(msLevel == 2L) |>
      group_by(group_idx) |>
      summarise(rt_sec = median(rt_sec, na.rm = TRUE), .groups = "drop") |>
      mutate(lab = lab, mixture = m)
    write.csv(rt_df, path, row.names = FALSE)
    rt_df
  }))
}))
```

```{r}
#| label: rt-expected-compounds
hit_rt <- all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14),
         xcms_fts    = as.character(xcms_fts)) |>
  filter(!is.na(inchiKey)) |>
  inner_join(
    expected_all |> select(mixture, expected_inchikey_2d, compound,
                           expected_formula, expected_inchikey),
    by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  left_join(rt_all |> mutate(group_idx = as.character(group_idx)),
            by = c("lab", "mixture", "xcms_fts" = "group_idx")) |>
  filter(!is.na(rt_sec))

cat("hit_rt rows (per-feature RT for expected-compound matches):",
    nrow(hit_rt), "\n")
```

## 7. EICs for annotated features - `chromExtract()` workflow

For every expected compound we want one MS1 EIC per lab, **including labs
that did not annotate it**, so we can visually check whether the signal
exists but MS2 missed it. One flat peak table → one `chromExtract()` call.

Two row types in the peak table:

* `source = "own"` - the lab annotated this compound via SIRIUS; use its
  ionMass + RT.
* `source = "consensus"` - the lab did not annotate it; use the cross-lab
  median ionMass + median raw RT.

For the MS1 file paths: afekta/cembio/icl have dedicated MS1-only files
(no `_MS2` in name) which give cleaner survey traces than the MS2 file's
interleaved MS1. HMGU acquires everything in one file.

```{r}
#| label: ms1-files-per-lab
ms1_files_for_lab <- function(lab_name, study_group = "HE", mixture) {
  study_path <- here("1_preprocessing", lab_name, study_group)
  mzml_dir   <- file.path(study_path, "mzml")
  seq_pos <- read_xlsx(file.path(study_path, "seq_pos.xlsx")) |> as.data.frame()
  seq_neg <- read_xlsx(file.path(study_path, "seq_neg.xlsx")) |> as.data.frame()
  seq_pos$polarity <- "pos"; seq_neg$polarity <- "neg"
  seq <- rbind(seq_pos, seq_neg)
  seq$filename <- paste0(seq$`Data File`, ".mzML")
  seq <- seq[seq$filename %in% list.files(mzml_dir), ]
  seq$mixture <- gsub("\\.", "_",
                      sub("target$", "",
                          sub(".*_", "",
                              sub("_MS2$", "", seq$`Data File`))))
  is_ms2   <- grepl("MS2", seq$`Data File`, fixed = TRUE)
  separate <- any(is_ms2)
  seq <- if (separate) seq[!is_ms2 & seq$mixture == mix_us(mixture), ]
         else          seq[          seq$mixture == mix_us(mixture), ]
  tibble(lab = lab_name, mixture = mixture, polarity = seq$polarity,
         path = file.path(mzml_dir, seq$filename),
         separate_ms1_ms2 = separate)
}

lab_files <- bind_rows(lapply(mixtures, function(m)
  bind_rows(lapply(labs, ms1_files_for_lab, study_group = study_group, mixture = m))))
```

### 7.1 Build the flat peak table

EICs are extracted with a generous **±60 s** symmetric window around the
annotation RT. The wide window serves two purposes:

1. **Noise context.** §7.4's SNR-aware peak picker estimates baseline +
   noise from the EIC tails (|rt − apex| > 25 s); we need real tail
   region to compute MAD / IQR of intensity.
2. **Tolerance to cross-lab RT drift.** ±60 s easily covers cembio ↔
   icl gradient differences and early-RT instabilities, so the single
   pass below handles what the legacy two-pass approach handled.

The ±20 s pass-1 of the legacy pipeline (and its ±80 s pass-2 retry)
have been removed — they were a workaround for `peakBoundary()`'s habit
of locking onto the tallest local maximum even when no real peak is
present (see §7.4 rationale).

The current build also addresses three MS2-hit-level issues identified
in `9_per_compound_diagnostics.qmd`:

* **All adducts get their own EIC.** The legacy code did `adduct =
  first(adduct)` per `(mixture, lab, polarity, compound)`, silently
  dropping every adduct beyond the first. We now group by adduct as
  well, so `[M+H]+`, `[M+Na]+`, `[M+K]+`, `[M-H+formic]-` etc. each
  emit their own row in the canonical CSV.
* **All features get a vote (no median collapse).** The legacy code
  did `rt_sec = median(rt_sec)` per `(mixture, lab, polarity, compound,
  adduct)`. For cembio's block-MS2 mode, `extract_all_ms2()`'s
  `rle(precursorMz)` can shatter ONE chromatographic peak into many
  features when other inclusion-list precursors interleave — e.g.
  ritalinic acid in cembio mix 2.9 pos has 11 features spanning RT
  198–258 s. The median (227 s) isn't any of them. We now keep every
  feature as a candidate so the clustering step decides which RTs
  belong to one peak.
* **Cluster-based consensus RT** combines cross-lab, cross-polarity,
  and cross-feature evidence to pick the "true" RT for every
  `(mixture, compound)`. The algorithm:

  1. Pool all candidate RTs from every `(lab, polarity, adduct, feature)`
     tuple that SIRIUS annotated for this compound.
  2. Greedy 1-D clustering with a 30 s window (configurable via
     `rt_cluster_window`).
  3. Score each cluster as `n_labs × 10 + n_polarities × 5 +
     sum_cosmic`. Highest score wins.
  4. The winning cluster's mean RT becomes the **consensus RT** used
     for every lab's chromExtract of that compound, in both polarities.
  5. Labs whose own MS2 fell in the winning cluster get `source =
     "own"`; labs whose MS2 fell in a non-winning cluster (DDA hit the
     wrong peak) get `source = "consensus"` — their wrong RT no longer
     contaminates the propagation. Labs that never annotated this
     compound also get `source = "consensus"`.

  This naturally fixes the cross-polarity disagreement, the cembio
  feature-shattering, and the wrong-peak DDA-trigger problem in one
  pass.

```{r}
#| label: peak-table
#| message: false
library(Chromatograms)

sps_all <- Spectra(lab_files$path, backend = MsBackendMzR())
sps_all <- sps_all[msLevel(sps_all) == 1L]
canon   <- unique(dataOrigin(sps_all))
lab_files$dataOrigin <- canon[match(basename(lab_files$path), basename(canon))]
stopifnot(all(!is.na(lab_files$dataOrigin)))

# --- eic_candidates: one row per FEATURE per (mixture, lab, polarity,
# compound, adduct). No collapse across features within (lab, polarity,
# compound, adduct). This is the raw candidate pool that cluster-based
# consensus operates on.
#
# Note: a single (lab, polarity, compound, adduct) tuple can have many
# feature rows here — e.g. cembio mix 2.9 ritalinic acid pos [M+H]+
# has 11 features at RTs 198–258 s because the block-MS2 `rle(...)`
# grouping splits one chromatographic peak into many pieces when other
# precursors interleave. Keeping all of them lets cross-lab + cross-
# polarity clustering decide whether they're 11 separate peaks or one
# peak that was shattered.
eic_candidates <- hit_rt |>
  select(mixture, lab, polarity, xcms_fts, compound, inchikey_2d, rt_sec) |>
  distinct() |>
  left_join(all_summaries |>
              select(mixture, lab, externalFeatureId, ionMass, adduct,
                     confidenceExactMatch) |>
              mutate(externalFeatureId = as.character(externalFeatureId)),
            by = c("mixture", "lab", "xcms_fts" = "externalFeatureId")) |>
  filter(!is.na(ionMass), !is.na(rt_sec)) |>
  rename(cosmic = confidenceExactMatch)

# --- RT clustering per (mixture, compound)
#
# For each (mixture, compound), pool ALL candidate RTs from every
# (lab, polarity, adduct, feature) tuple. Cluster them with a greedy
# 1-D agglomeration: sort RTs, walk the sorted list, start a new
# cluster whenever the gap between consecutive RTs exceeds
# `rt_cluster_window`. This naturally collapses cembio's 11
# ritalinic-acid pos features (5–6 s apart) into one cluster while
# separating the wrong-peak co-eluting contaminants (≥ 30 s away).
rt_cluster_window <- 30   # s

cluster_rts_greedy <- function(rt) {
  if (length(rt) == 0L) return(integer(0))
  ord <- order(rt); s <- rt[ord]
  out <- integer(length(rt))
  cl <- 1L; out[ord[1L]] <- cl
  if (length(s) >= 2L) for (i in 2:length(s)) {
    if (s[i] - s[i - 1L] > rt_cluster_window) cl <- cl + 1L
    out[ord[i]] <- cl
  }
  out
}

eic_candidates <- eic_candidates |>
  group_by(mixture, compound) |>
  mutate(rt_cluster = cluster_rts_greedy(rt_sec)) |>
  ungroup()

# --- Score clusters and pick the winning RT per (mixture, compound)
#
# Score weights:
#   n_labs       × 10   (cross-lab agreement is the strongest signal)
#   n_polarities × 5    (cross-polarity confirmation is good evidence)
#   sum_cosmic         (COSMIC tiebreaker; NA → 0)
#
# Highest-score cluster per (mixture, compound) is the "true" RT.
cluster_scores <- eic_candidates |>
  group_by(mixture, compound, rt_cluster) |>
  summarise(n_labs       = n_distinct(lab),
            n_polarities = n_distinct(polarity),
            n_features   = n(),
            sum_cosmic   = sum(cosmic, na.rm = TRUE),
            # median is more robust than mean against asymmetric feature
            # distributions (e.g. cembio's shattering produces 11 evenly-
            # spaced features but they need not be symmetric around the
            # true apex)
            rt_center    = median(rt_sec),
            rt_min       = min(rt_sec),
            rt_max       = max(rt_sec),
            .groups      = "drop") |>
  mutate(score = n_labs * 10 + n_polarities * 5 + sum_cosmic)

winning_cluster <- cluster_scores |>
  group_by(mixture, compound) |>
  slice_max(score, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(mixture, compound,
         winning_cluster   = rt_cluster,
         consensus_rt      = rt_center,
         consensus_n_labs  = n_labs,
         consensus_n_pol   = n_polarities,
         consensus_features= n_features,
         consensus_score   = score)

# --- Single-annotation diagnostic
#
# How many (mixture, compound) had only ONE cluster (clean — single
# annotation = single RT, no cross-feature splits, no cross-polarity
# disagreement)? How many had multiple clusters?
n_clusters_per_cpd <- cluster_scores |>
  count(mixture, compound, name = "n_clusters")

cluster_summary <- n_clusters_per_cpd |>
  count(n_clusters, name = "n_compounds")

cat("Cluster-based consensus RT — annotation evaluation\n")
cat("===================================================\n")
cat("Compounds with N RT clusters (mixture × compound):\n")
print(as.data.frame(cluster_summary))

cat("\nWinning-cluster cross-lab / cross-polarity support:\n")
winning_cluster |>
  count(consensus_n_labs, consensus_n_pol) |>
  arrange(desc(consensus_n_labs), desc(consensus_n_pol)) |>
  as.data.frame() |> print()

cat("\nMulti-cluster compounds: how many of the candidate features fell into\n",
    "the non-winning ('alternate') cluster? These are the rows whose MS2\n",
    "trigger landed at the wrong chromatographic peak.\n", sep = "")
alt_features <- eic_candidates |>
  inner_join(winning_cluster |>
               select(mixture, compound, winning_cluster),
             by = c("mixture", "compound")) |>
  mutate(in_winning = rt_cluster == winning_cluster) |>
  group_by(mixture, compound) |>
  summarise(n_winning = sum(in_winning),
            n_alt     = sum(!in_winning),
            .groups   = "drop")
cat(sprintf("  Compounds with any alternate-cluster features: %d / %d (%.1f%%)\n",
            sum(alt_features$n_alt > 0L), nrow(alt_features),
            100 * mean(alt_features$n_alt > 0L)))
cat(sprintf("  Total alternate-cluster features: %d\n",
            sum(alt_features$n_alt)))

# --- Build eic_params: per (mixture, lab, polarity, compound, adduct)
# Each row uses the lab's OWN rt_sec if any of its features for this
# (lab, polarity, adduct) fell into the winning cluster; otherwise the
# row is dropped here (it gets a "consensus" extraction later).
#
# Adducts are first-class — every (lab, polarity, compound, adduct)
# tuple where the lab has at least one feature in the winning cluster
# gets its own own-row.
eic_params <- eic_candidates |>
  inner_join(winning_cluster |> select(mixture, compound, winning_cluster),
             by = c("mixture", "compound")) |>
  filter(rt_cluster == winning_cluster) |>
  group_by(mixture, lab, polarity, compound, adduct) |>
  summarise(ionMass    = median(ionMass),
            # median over the lab's features in the winning cluster
            rt_sec     = median(rt_sec),
            # rt_spread tells §7.5 how wide to make the picker's apex
            # search window. 0 for DDA labs (one MS2 per feature); for
            # cembio the spread can be 30–60 s when the inclusion-list
            # scan window spans multiple cycles.
            rt_spread  = max(rt_sec) - min(rt_sec),
            n_features = n(),
            .groups    = "drop")

# --- eic_consensus: the cluster's RT, broadcast to every (mixture,
# polarity, compound, adduct) combination that any lab annotated for
# this compound. So a lab that didn't annotate this compound in this
# polarity still gets a `consensus` extraction at the winning RT.
adduct_inventory <- eic_candidates |>
  distinct(mixture, polarity, compound, adduct, ionMass) |>
  group_by(mixture, polarity, compound, adduct) |>
  summarise(ionMass = median(ionMass), .groups = "drop")

eic_consensus <- adduct_inventory |>
  inner_join(winning_cluster |> select(mixture, compound, consensus_rt),
             by = c("mixture", "compound")) |>
  rename(rt_sec = consensus_rt)

ppm_tol  <- 20
pad_wide <- 60   # symmetric ±60 s — see §7.1 rationale

peak_table <- tidyr::expand_grid(
    lab             = labs,
    mix_pol_cpd_add = eic_consensus |>
                        distinct(mixture, polarity, compound, adduct)) |>
  tidyr::unnest(mix_pol_cpd_add) |>
  left_join(eic_params,    by = c("lab", "mixture", "polarity",
                                  "compound", "adduct")) |>
  left_join(eic_consensus, by = c("mixture", "polarity",
                                  "compound", "adduct"),
            suffix = c("", "_cons")) |>
  mutate(source = if_else(is.na(ionMass), "consensus", "own"),
         rt     = coalesce(rt_sec,  rt_sec_cons),
         mz     = coalesce(ionMass, ionMass_cons)) |>
  filter(!is.na(rt), !is.na(mz)) |>
  left_join(lab_files |> select(lab, mixture, polarity, dataOrigin),
            by = c("lab", "mixture", "polarity")) |>
  filter(!is.na(dataOrigin)) |>
  transmute(mzMin     = mz * (1 - ppm_tol / 1e6),
            mzMax     = mz * (1 + ppm_tol / 1e6),
            rtMin     = pmax(0, rt - pad_wide),
            rtMax     = rt + pad_wide,
            msLevel   = 1L,
            dataOrigin, lab, mixture, polarity, compound, adduct, source,
            mz_center = mz, rt_center = rt)

peak_table |>
  count(mixture, lab, source) |>
  tidyr::pivot_wider(names_from = source, values_from = n, values_fill = 0)
```

### 7.2 One `chromExtract()` call

```{r}
#| label: chromextract
#| message: false
chr_base <- Chromatograms(sps_all)
chr_all  <- chromExtract(chr_base, peak.table = peak_table,
                         by = c("msLevel", "dataOrigin"))
chr_all
```

### 7.3 EICs centred at annotation RT - sanity-check

Diagnostic plot for **one display mixture only** (`display_mixture`,
default `"1.1"`) to keep the panel count manageable. Sections 7.4–7.6
operate on `chr_all` across all mixtures; cross-mixture comparisons live
in `9_per_compound_diagnostics.qmd`.

```{r}
#| label: plot-eics
#| fig-height: 14
#| fig-cap: "Pass-1 MS1 EICs (tight ±20 s) for the display mixture. Solid = lab's own annotation; dashed = consensus borrow."
display_mixture <- "1.1"
disp_idx <- which(chromData(chr_all)$mixture == display_mixture)

eic_df <- bind_rows(lapply(disp_idx, function(i) {
  cd <- chromData(chr_all)[i, , drop = FALSE]
  pd <- peaksData(chr_all[i])[[1L]]
  if (!nrow(pd)) return(NULL)
  tibble(rt_sec = pd[, "rtime"], intensity = pd[, "intensity"],
         lab = cd$lab, mixture = cd$mixture, polarity = cd$polarity,
         compound = cd$compound, source = cd$source, rt_center = cd$rt_center)
})) |>
  mutate(rt_rel = rt_sec - rt_center)

eic_df_disp <- eic_df |>
  filter(abs(rt_rel) <= 100) |>
  group_by(lab, polarity, source, compound) |>
  filter(n() >= 2L) |>
  ungroup()

center_dots <- eic_df_disp |>
  filter(source == "own") |>
  group_by(lab, compound, polarity) |>
  slice_min(abs(rt_rel), n = 1, with_ties = FALSE) |>
  ungroup()

ggplot(eic_df_disp,
       aes(x = rt_rel, y = intensity, colour = lab, linetype = source,
           group = interaction(lab, polarity, source))) +
  geom_line(linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_point(data = center_dots,
             aes(x = rt_rel, y = intensity, colour = lab),
             shape = 21, size = 2, fill = "white", stroke = 1,
             inherit.aes = FALSE) +
  facet_wrap(~ compound, scales = "free_y", ncol = 4) +
  scale_linetype_manual(values = c(own = "solid", consensus = "dashed")) +
  labs(x = "RT - annotation RT (s)", y = "Intensity",
       colour = NULL, linetype = NULL,
       title = sprintf("EICs centred at annotation RT - mixture %s (%d ppm)",
                       display_mixture, ppm_tol)) +
  theme_bw() +
  theme(strip.text = element_text(size = 7), legend.position = "bottom")
```

### 7.4 SNR-aware peak picking — rationale

The previous pipeline used `Chromatograms::peakBoundary()` to find the
tallest local maximum in each EIC, then re-extracted on its boundary.
For real peaks this works fine, but it has a critical failure mode:
**when the compound is below detection at a given lab, the EIC is
mostly noise, and `peakBoundary()` happily returns the tallest noise
spike as the apex**. The two-pass widening retry doesn't help — it just
gives the noise more room to find a louder spike.

We saw this in raw-file diagnostics: hmgu's NAPS standards are the
strongest of any lab (geo-mean 3.5 × 10⁵ counts in pos), but ~70 % of
its annotated spike-in rows reported maxIntensity in the 1-100 count
range with apex_rt scattered across the full chromatogram. Those
"measurements" were noise — the chromatographic peak simply wasn't
there at the dilution used. Sciex ESI source ionisation efficiency
appears compound-specific; a panel tuned for Agilent Jet Stream falls
under hmgu's per-compound floor for a chunk of the panel.

The new picker applies three **independent** rejection guards. A row
must pass all three to be reported as `detection = "detected"`:

1. **SNR.** apex_int − baseline must exceed `snr_threshold × noise`,
   where baseline = median of intensities at |rt − apex| > 25 s and
   noise = max(MAD, IQR/2, 1) of the **low-intensity portion of the
   tail only** (intensities below the tail's 50 th percentile). The
   bottom-half restriction matters because at the 20 ppm m/z window
   used in §7.1, EICs frequently contain other chromatographic peaks
   in the tail region (isomers, late-eluting contaminants at the same
   m/z). Including those in the noise estimate makes "noise" the size
   of a peak, not the size of background, and crushes the SNR of
   real apex peaks (e.g. cembio L-phenylalanine got noise = 894 k,
   apex = 4.6 M → SNR ≈ 0; bottom-half tail gives noise ≈ 75 k,
   SNR ≈ 60). The IQR floor matters for sparse-positive data: MAD
   goes to 0 when the tail is mostly exact zeros, IQR catches the
   spread of low-count noise spikes that MAD misses.

2. **Shape.** The apex's immediate neighbours (apex index ±2, raw, not
   smoothed) must also be elevated. We require ≥ 2 of those 5 scans
   to be ≥ 30 % of apex_int. A single-scan spike with zero-neighbours
   = noise. A real chromatographic peak has a rising and falling
   shoulder spanning at least the next-scan ↔ apex pair.

3. **Absolute floor (only when noise estimate is at its hard floor).**
   If both MAD and IQR are exactly 0, the tail region had no useful
   noise context (a "flat-zero EIC, except for one tiny spike at the
   prior"). We can't trust SNR there; require apex_int ≥ 30 counts
   absolute. This kills the residual hmgu noise grabs that the SNR
   guard alone passes because (5.7 − 0) / (floor-noise=1) = 5.7.

Each row also gets a **source-aware RT prior**:

- `prior_kind = "own"`: the lab annotated this compound via SIRIUS
  *and* its COSMIC ≥ 0.1 *and* its own apex_rt agrees with the
  cross-lab consensus within 30 s. Search ±8 s around the lab's own
  RT (tight — SIRIUS already told us where the peak is).
- `prior_kind = "consensus"`: everything else. Search ±15 s around the
  cross-lab median of "own" RTs (wider — accommodates inter-lab
  chromatographic drift). This includes "own" rows the demotion
  guards disagreed with (missing COSMIC, or RT > 30 s off consensus —
  a clue the SIRIUS hit was spurious).

Returns include the metrics (`maxIntensity`, `areaUnderTic`,
`xicFwhm`), the apex_rt, and the diagnostic columns (`snr`,
`baseline`, `noise`, `n_scans_in_shape`, `prior_kind`, `detection`).

### 7.5 Per-row RT priors

```{r}
#| label: per-row-priors
## Cross-lab + cross-polarity + cross-feature consensus RT for each
## (mixture, compound). Source = §7.1's `winning_cluster`, the highest-
## scoring RT cluster from the greedy clustering. Broadcast to (mixture,
## polarity, compound) so the existing per-row prior logic below stays
## one-step.
rt_consensus_per_cpd <- winning_cluster |>
  select(mixture, compound, rt_consensus = consensus_rt) |>
  tidyr::expand_grid(polarity = c("pos", "neg"))

## COSMIC scores per row (used by the demotion guard).
cosmic_lookup <- annotation_identity_seed <-
  ## Pull straight from the candidate-best id_summary (built in §8.1).
  ## Materialise it here on demand to avoid forward-reference fragility.
  all_summaries |>
    filter(!is.na(inchiKey)) |>
    mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
    group_by(mixture, lab, inchikey_2d, polarity) |>
    slice_max(csiScore, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(mixture, lab, inchikey_2d, polarity,
           cosmic = confidenceExactMatch)

cosmic_per_row <- structure_hits_all |>
  select(mixture, lab, compound, inchikey_2d, polarity) |>
  left_join(cosmic_lookup,
            by = c("mixture","lab","inchikey_2d","polarity")) |>
  group_by(mixture, lab, compound, polarity) |>
  summarise(cosmic = suppressWarnings(max(cosmic, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(cosmic = ifelse(is.infinite(cosmic), NA_real_, cosmic))

## Build the per-row RT prior. eic_params has the lab's own median RT
## within the winning cluster and the rt_spread (max - min) of the
## lab's features in that cluster.
##
## apex_pad logic:
##   "own" rows  : pad must cover the lab's feature spread plus a 5 s
##                 safety margin. For DDA labs (1 MS2 per feature) the
##                 spread is 0, so pad = 8 s (the tight DDA default).
##                 For cembio's block-MS2 the spread can be 30–60 s,
##                 so pad grows to ~35 s; the SNR picker still
##                 rejects noise spikes within this wider window.
##   "consensus" : pad = 15 s, the default for lab×polarity tuples
##                 where the lab itself did not annotate this compound.
prior_per_row <- chromData(chr_all) |>
  as.data.frame() |>
  select(mixture, lab, polarity, compound, adduct, source) |>
  # eic_params keys on adduct too (one row per (mixture, lab, polarity,
  # compound, adduct) since the all-adducts refactor in §7.1). The join
  # MUST include adduct or one chrom_metrics row gets multiple matches
  # and the row count blows up (apply-picker's
  # `stopifnot(nrow(cd_all) == nrow(prior_per_row))` then fails).
  left_join(eic_params |>
              select(mixture, lab, polarity, compound, adduct,
                     rt_own = rt_sec, rt_spread_own = rt_spread),
            by = c("mixture","lab","polarity","compound","adduct")) |>
  left_join(rt_consensus_per_cpd,
            by = c("mixture","polarity","compound")) |>
  left_join(cosmic_per_row,
            by = c("mixture","lab","polarity","compound")) |>
  mutate(prior_kind = case_when(
           source == "own" &
             !is.na(rt_own) & !is.na(cosmic) & cosmic >= 0.1 &
             !is.na(rt_consensus) & abs(rt_own - rt_consensus) <= 30 ~ "own",
           TRUE ~ "consensus"),
         rt_prior  = ifelse(prior_kind == "own", rt_own, rt_consensus),
         apex_pad  = ifelse(prior_kind == "own",
                            pmax(8, replace_na(rt_spread_own, 0) / 2 + 5),
                            15))

cat("Prior kinds (after own→consensus demotions):\n")
prior_per_row |> count(source, prior_kind) |>
  tidyr::pivot_wider(names_from = prior_kind, values_from = n,
                     values_fill = 0L) |>
  knitr::kable()
```

### 7.6 Picker function and per-row extraction

```{r}
#| label: snr-picker
extract_one <- function(rt, intensity, rt_prior,
                        apex_search_pad = 10,
                        tail_pad        = 25,
                        snr_threshold   = 5,
                        abs_min_counts  = 30) {
  ok <- !is.na(intensity)
  rt <- rt[ok]; intensity <- intensity[ok]
  if (length(rt) < 5L)
    return(list(apex_rt = NA_real_, maxIntensity = NA_real_,
                areaUnderTic = NA_real_, xicFwhm = NA_real_,
                snr = NA_real_, baseline = NA_real_, noise = NA_real_,
                n_scans_in_shape = 0L, detection = "no_data"))

  ## centred 3-point moving average (smoothed trace for valley/noise only)
  sm <- as.numeric(stats::filter(intensity, rep(1, 3) / 3, sides = 2))
  sm[is.na(sm)] <- intensity[is.na(sm)]

  ## (1) apex on RAW intensity inside prior ± apex_pad
  search_idx <- which(rt >= rt_prior - apex_search_pad &
                      rt <= rt_prior + apex_search_pad)
  if (!length(search_idx))
    return(list(apex_rt = NA_real_, maxIntensity = NA_real_,
                areaUnderTic = NA_real_, xicFwhm = NA_real_,
                snr = NA_real_, baseline = NA_real_, noise = NA_real_,
                n_scans_in_shape = 0L, detection = "no_data"))
  apex_local <- search_idx[which.max(intensity[search_idx])]
  apex_rt    <- rt[apex_local]
  apex_int   <- intensity[apex_local]

  ## (2) noise + baseline from tail (|rt - apex| > tail_pad)
  tail_idx <- which(abs(rt - apex_rt) > tail_pad)
  if (length(tail_idx) < 10L) {
    n <- length(rt); tail_idx <- c(seq_len(round(0.25 * n)),
                                    (n - round(0.25 * n) + 1L):n)
  }
  tail_int <- intensity[tail_idx]
  ## Use ONLY the low-intensity portion of the tail (bottom 50 %) for
  ## noise/baseline. Otherwise other chromatographic peaks at the same
  ## m/z window (isomers, late-eluting contaminants) get treated as
  ## "noise" and crush the SNR of real peaks. See §7.4 guard #1.
  tail_low <- tail_int[tail_int <= quantile(tail_int, 0.5, na.rm = TRUE)]
  baseline <- median(tail_low, na.rm = TRUE)
  noise_mad <- mad(tail_low, na.rm = TRUE)
  q15 <- as.numeric(quantile(tail_low, 0.15, na.rm = TRUE))
  q85 <- as.numeric(quantile(tail_low, 0.85, na.rm = TRUE))
  raw_noise <- max(noise_mad, (q85 - q15) / 2, na.rm = TRUE)
  noise_at_floor <- !is.finite(raw_noise) || raw_noise <= 0
  noise <- max(raw_noise, 1, na.rm = TRUE)
  snr   <- (apex_int - baseline) / noise

  ## (3) valleys for peak boundaries on the smoothed trace
  v_idx <- tryCatch(MsCoreUtils::valleys(sm), error = function(e) integer(0))
  if (length(v_idx) >= 2L) {
    left  <- suppressWarnings(max(v_idx[v_idx <= apex_local], na.rm = TRUE))
    right <- suppressWarnings(min(v_idx[v_idx >= apex_local], na.rm = TRUE))
    if (!is.finite(left))  left  <- 1L
    if (!is.finite(right)) right <- length(rt)
  } else {
    left <- apex_local; right <- apex_local
    while (left  > 1L         && sm[left  - 1L] > baseline + noise) left  <- left  - 1L
    while (right < length(rt) && sm[right + 1L] > baseline + noise) right <- right + 1L
  }
  pk_rt <- rt[left:right]; pk_int <- intensity[left:right]
  area  <- if (length(pk_rt) >= 2L)
             sum(diff(pk_rt) * (pk_int[-1] + pk_int[-length(pk_int)]) / 2)
           else NA_real_
  hm    <- (apex_int - baseline) / 2 + baseline
  above <- which(pk_int >= hm)
  fwhm  <- if (length(above) >= 2L) diff(range(pk_rt[above])) else NA_real_

  ## (4) shape sanity: at least 2 of 5 scans around apex >= 30% of apex
  nb_lo  <- max(1L, apex_local - 2L); nb_hi <- min(length(rt), apex_local + 2L)
  n_shape <- sum(intensity[nb_lo:nb_hi] >= 0.3 * apex_int, na.rm = TRUE)

  ## (5) three-guard detection classification
  detection <- if (apex_int <= baseline) "below_detection"
               else if (snr < snr_threshold) "below_detection"
               else if (n_shape < 2L) "below_detection"
               else if (noise_at_floor && apex_int < abs_min_counts) "below_detection"
               else "detected"

  list(apex_rt = apex_rt,
       maxIntensity = if (detection == "detected") apex_int else NA_real_,
       areaUnderTic = if (detection == "detected") area     else NA_real_,
       xicFwhm     = if (detection == "detected") fwhm     else NA_real_,
       snr = snr, baseline = baseline, noise = noise,
       n_scans_in_shape = n_shape, detection = detection)
}
```

```{r}
#| label: apply-picker
#| message: false
cd_all <- as.data.frame(chromData(chr_all))
stopifnot(nrow(cd_all) == nrow(prior_per_row))

metrics_rows <- vector("list", nrow(cd_all))
pkData <- peaksData(chr_all)
for (i in seq_len(nrow(cd_all))) {
  pd <- pkData[[i]]
  prior_i <- prior_per_row[i, ]
  if (is.null(pd) || !nrow(pd) || is.na(prior_i$rt_prior)) {
    metrics_rows[[i]] <- list(apex_rt = NA_real_, maxIntensity = NA_real_,
                              areaUnderTic = NA_real_, xicFwhm = NA_real_,
                              snr = NA_real_, baseline = NA_real_,
                              noise = NA_real_, n_scans_in_shape = 0L,
                              detection = "no_data")
    next
  }
  metrics_rows[[i]] <- extract_one(pd[, "rtime"], pd[, "intensity"],
                                   rt_prior        = prior_i$rt_prior,
                                   apex_search_pad = prior_i$apex_pad)
}

metrics_df <- bind_cols(
  cd_all |> select(mixture, lab, polarity, compound, adduct, source),
  prior_per_row |> select(prior_kind),
  bind_rows(lapply(metrics_rows, as.data.frame))
)

cat("Picker done — ", nrow(metrics_df), " rows\n",
    "  detected:        ", sum(metrics_df$detection == "detected"), "\n",
    "  below_detection: ", sum(metrics_df$detection == "below_detection"), "\n",
    "  no_data:         ", sum(metrics_df$detection == "no_data"), "\n",
    sep = "")
```

### 7.7 Detection summary by lab × polarity

The detection fraction per lab tells us **per-lab sensitivity** for the
annotated set — directly comparable across instruments.

```{r}
#| label: detection-summary
metrics_df |>
  count(lab, polarity, detection) |>
  tidyr::pivot_wider(names_from = detection, values_from = n,
                     values_fill = 0L) |>
  mutate(total = detected + below_detection + no_data,
         pct_detected = round(100 * detected / total, 1)) |>
  knitr::kable(caption = "Detection counts by lab × polarity. pct_detected is the per-lab sensitivity for the annotated spike-in panel.")
```

```{r}
#| label: detection-by-prior-kind
metrics_df |>
  count(prior_kind, source, detection) |>
  tidyr::pivot_wider(names_from = detection, values_from = n,
                     values_fill = 0L) |>
  knitr::kable(caption = "Detection × (prior_kind, source). 'own' rows that survive the demotion guards should detect near 100 %; 'consensus' detection rate is the cross-lab presence statistic.")
```

```{r}
#| label: snr-distribution
#| fig-height: 4
#| fig-cap: "SNR distribution for detected rows, by lab × polarity. log10 scale."
metrics_df |>
  filter(detection == "detected", is.finite(snr), snr > 0) |>
  ggplot(aes(x = lab, y = snr, fill = lab)) +
  geom_violin(scale = "width", linewidth = 0.3, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.size = 0.5, alpha = 0.5) +
  scale_y_log10() +
  facet_wrap(~ polarity, nrow = 1) +
  scale_fill_manual(values = c(afekta = "#E64B35", cembio = "#4DBBD5",
                                hmgu = "#00A087", icl = "#3C5488")) +
  labs(x = NULL, y = "SNR (log10)") +
  theme_bw() + theme(legend.position = "none")
```

### 7.8 Sample below-detection EICs — sanity check

Inspect 16 EICs that the picker flagged below_detection. We expect to
see flat-zero traces with at most a few low-count spikes. If we see
visible peaks, the guards are too strict.

```{r}
#| label: below-detection-eics
#| fig-height: 9
#| fig-cap: "EICs the new picker flagged below_detection. Expected: flat noise with occasional sub-30-count spikes. A visible peak here = the guards are too tight."
bd_idx <- which(metrics_df$detection == "below_detection")
n_show <- min(16L, length(bd_idx))
if (n_show > 0L) {
  show_idx <- sample(bd_idx, n_show)
  bd_eic <- bind_rows(lapply(show_idx, function(j) {
    pd <- pkData[[j]]; if (is.null(pd) || !nrow(pd)) return(NULL)
    md <- metrics_df[j, ]
    tibble(rt = pd[, "rtime"], intensity = pd[, "intensity"],
           panel = sprintf("mix %s | %s | %s [%s/%s]",
                           md$mixture, md$lab, substr(md$compound, 1, 18),
                           md$polarity, md$source))
  }))
  ggplot(bd_eic, aes(x = rt, y = intensity)) +
    geom_line(colour = "#ff7f0e", linewidth = 0.4) +
    geom_point(size = 0.6, alpha = 0.5) +
    facet_wrap(~ panel, scales = "free", ncol = 4) +
    labs(x = "RT (s)", y = "Intensity") +
    theme_bw() +
    theme(strip.text = element_text(size = 6.5),
          axis.text  = element_text(size = 6))
} else {
  message("Nothing flagged below_detection — that's suspicious; inspect manually.")
}
```

## 8. Write canonical CSVs

Two CSVs land at `output_dir` per the contract in
[`CANONICAL_SCHEMAS.md`](../../../5_downstream_analysis/CANONICAL_SCHEMAS.md).
Downstream code (`5_downstream_analysis/9_per_compound_diagnostics.qmd`,
`8_data_level_progression.qmd`) consumes only these.

### 8.1 `annotation_identity.csv`

`structure_hits_all` is augmented with the COSMIC scores and the
observed `ionMass` from `all_summaries`, plus a placeholder
`confidence_tier = "unfiltered"`. The join picks the best-`csiScore`
candidate per `(mixture, lab, inchikey_2d, polarity)` — identity is
one-per-compound, so we collapse the candidate dimension first; the
resulting `ionMass` is the precursor m/z of that winning candidate and
is exposed downstream as `mz_observed` (basis for the inter-lab m/z
deviation analysis in `9_per_compound_diagnostics.qmd`).

```{r}
#| label: build-annotation-identity
id_summary <- all_summaries |>
  filter(!is.na(inchiKey)) |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  group_by(mixture, lab, inchikey_2d, polarity) |>
  slice_max(csiScore, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(mixture, lab, inchikey_2d, polarity,
         confidenceExactMatch, confidenceApproxMatch,
         mz_observed = ionMass)

annotation_identity <- structure_hits_all |>
  left_join(id_summary,
            by = c("mixture", "lab", "inchikey_2d", "polarity")) |>
  mutate(confidence_tier = "unfiltered") |>
  select(mixture, lab, compound, inchikey_2d, polarity,
         best_formula_rank, best_structure_rank, adducts,
         molecularFormula, expected_formula, expected_inchikey,
         mz_observed, confidenceExactMatch, confidenceApproxMatch,
         confidence_tier)

write.csv(annotation_identity,
          file.path(output_dir, "annotation_identity.csv"),
          row.names = FALSE)
cat("annotation_identity.csv  -- ", nrow(annotation_identity), " rows\n",
    "  with COSMIC scores: ",
    sum(!is.na(annotation_identity$confidenceExactMatch)),
    "  with mz_observed: ",
    sum(!is.na(annotation_identity$mz_observed)), "\n", sep = "")
```

### 8.2 `annotation_chrom_metrics.csv`

`metrics_df` from §7.6 already carries the metrics + diagnostics for
every row. The canonical schema (see `CANONICAL_SCHEMAS.md`) keeps the
original metric columns and adds the SNR-aware diagnostic fields. Rows
classified `below_detection` carry **NA** for `maxIntensity`,
`areaUnderTic`, `xicFwhm` — so the historical consumer filter
`!is.na(maxIntensity)` still works for "real signal only" and the
detection flag is available for explicit downstream gating.

```{r}
#| label: build-annotation-chrom-metrics
annotation_chrom_metrics <- metrics_df |>
  select(mixture, lab, polarity, compound, adduct, source,
         areaUnderTic, xicFwhm, maxIntensity, apex_rt,
         snr, baseline, noise, n_scans_in_shape,
         prior_kind, detection)

write.csv(annotation_chrom_metrics,
          file.path(output_dir, "annotation_chrom_metrics.csv"),
          row.names = FALSE)

cat("annotation_chrom_metrics.csv  -- ",
    nrow(annotation_chrom_metrics), " rows\n",
    "  detected:        ",
    sum(annotation_chrom_metrics$detection == "detected"), "\n",
    "  below_detection: ",
    sum(annotation_chrom_metrics$detection == "below_detection"), "\n",
    "  no_data:         ",
    sum(annotation_chrom_metrics$detection == "no_data"), "\n",
    sep = "")
```
