---
title: "SIRIUS annotation — all MS2 per lab (strategy 3)"
format: html
---

## Goal

Annotate **all MS2-bearing features** from every lab, without requiring
cross-lab consensus. Each lab's `mse2` object (or `mse` for HMGU, which
uses a single-injection setup) is used directly. MS1 spectra are linked to
MS2 via `fragmentGroupIndex()` from `Spectra`. Results are then grouped by
compound and compared to the consensus-based annotations from
`sirius_annotation_consensus.qmd`.

## Setup

```{r}
#| label: setup
#| message: false
library(Spectra)
library(MsExperiment)
library(xcms)
library(RuSirius)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

source(here("1_preprocessing", "setup.R"))
```

## Config

```{r}
#| label: config
labs        <- c("afekta", "cembio", "hmgu", "icl")
study_group <- "HE"

obj_dir    <- here("5_downstream_analysis", "object")
output_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                   "results_all_ms2")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Extraction helper

For each lab, load the raw mzML files directly as a `Spectra` object.
Labs with separate MS2 acquisition runs (afekta, cembio, icl) have files
labelled with `_MS2` in their name; these DDA files contain interleaved MS1
survey scans and MS2 fragmentation spectra. HMGU uses a single-injection DDA
setup so the regular sample files already contain both levels.
`fragmentGroupIndex()` then pairs MS1 with co-eluting MS2 within each file.

```{r}
#| label: helper-extract-all
extract_all_ms2 <- function(lab_name, study_group) {
  study_path <- here("1_preprocessing", lab_name, study_group)
  mzml_dir   <- file.path(study_path, "mzml")

  # Build filename table from seq files (polarity already a column)
  seq_pos <- read_xlsx(file.path(study_path, "seq_pos.xlsx")) |> as.data.frame()
  seq_neg <- read_xlsx(file.path(study_path, "seq_neg.xlsx")) |> as.data.frame()
  seq_pos$polarity <- "pos"
  seq_neg$polarity <- "neg"
  seq <- rbind(seq_pos, seq_neg)
  seq$filename <- paste0(seq$`Data File`, ".mzML")

  # Keep only files present on disk
  seq <- seq[seq$filename %in% list.files(mzml_dir), ]

  # Derive mixture label: strip trailing _MS2 before extracting the last token
  base_name    <- sub("_MS2$", "", seq$`Data File`)
  seq$mixture  <- gsub("\\.", "_", sub(".*_", "", base_name))

  # For labs with separate MS2 run files use those; otherwise (hmgu) use
  # the regular sample files which are already DDA (MS1+MS2 in one file)
  has_ms2_files <- any(grepl("MS2", seq$filename, fixed = TRUE))
  if (has_ms2_files) {
    seq <- seq[grepl("MS2", seq$filename, fixed = TRUE) & seq$mixture == "1_1", ]
  } else {
    seq <- seq[seq$mixture == "1_1", ]
  }

  # Load raw mzML directly — DDA files contain both MS1 and MS2 spectra
  sp <- Spectra(file.path(mzml_dir, seq$filename), backend = MsBackendMzR())
  sp <- setBackend(sp, MsBackendMemory())
  sp <- applyProcessing(sp)

  if (lab_name == "cembio") {
    # Cembio uses scheduled targeted MS2: MS1 and MS2 are acquired in
    # temporally disjoint blocks within the same file, so co-elution-based
    # pairing is impossible. Keep MS2 only and group consecutive scans that
    # share the same precursor m/z into one SIRIUS feature.
    sp <- sp[sp$msLevel == 2L]
    sp <- filterEmptySpectra(sp)
    key <- paste0(sp$dataOrigin, "|", sprintf("%.4f", sp$precursorMz))
    rl  <- rle(key)
    sp$group_idx <- rep(seq_along(rl$lengths), rl$lengths)
  } else {
    # DDA labs: drop MSn appearing before the first MS1 per file, then
    # pair MS1 with co-eluting MS2 via fragmentGroupIndex().
    keep <- unlist(lapply(split(seq_along(sp), sp$dataOrigin), function(idx) {
      lvl <- sp$msLevel[idx]
      first_ms1 <- which(lvl == 1L)[1L]
      if (is.na(first_ms1)) return(integer(0))
      idx[first_ms1:length(idx)]
    }))
    sp <- sp[sort(keep)]
    sp <- filterEmptySpectra(sp)
    sp$group_idx <- fragmentGroupIndex(sp)
    sp <- sp[duplicated(sp$group_idx) | duplicated(sp$group_idx, fromLast = TRUE)]
  }

  list(
    spectra_pos       = sp[sp$polarity == 1L],
    spectra_neg       = sp[sp$polarity == 0L],
    n_ms2_pos         = sum(sp$msLevel == 2L & sp$polarity == 1L),
    n_ms2_neg         = sum(sp$msLevel == 2L & sp$polarity == 0L),
    n_features_pos    = length(unique(sp$group_idx[sp$polarity == 1L])),
    n_features_neg    = length(unique(sp$group_idx[sp$polarity == 0L]))
  )
}
```

## 2. Run extraction

```{r}
#| label: extract
#| message: false
spectra_all <- list()
extraction_summary <- list()

for (lab in labs) {
  sum_path  <- file.path(output_dir, paste0("sirius_summary_", lab, ".csv"))
  full_path <- file.path(output_dir, paste0("sirius_full_",    lab, ".csv"))
  if (file.exists(sum_path) && file.exists(full_path)) {
    message("Skipping extraction for ", lab, " (SIRIUS results already exist)")
    next
  }
  message("Extracting all MS2 for: ", lab)
  res <- extract_all_ms2(
    lab_name    = lab,
    study_group = study_group
  )
  spectra_all[[lab]] <- res
  extraction_summary[[lab]] <- tibble(
    lab               = lab,
    n_features_pos    = res$n_features_pos,
    n_features_neg    = res$n_features_neg,
    n_ms2_pos         = res$n_ms2_pos,
    n_ms2_neg         = res$n_ms2_neg
  )
}

bind_rows(extraction_summary)
```

```{r}
#| label: plot-extraction
#| fig-cap: "SIRIUS features (MS1+MS2 group; MS2 only for cembio) per lab and polarity"
bind_rows(extraction_summary) |>
  select(lab, pos = n_features_pos, neg = n_features_neg) |>
  pivot_longer(c(pos, neg), names_to = "polarity", values_to = "n") |>
  ggplot(aes(x = lab, y = n, fill = polarity)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "SIRIUS features", fill = "Polarity") +
  theme_bw()
```

## 3. SIRIUS annotation

```{r}
#| label: sirius-login
#| eval: false
sirius_user <- "plouail@eurac.edu"
sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")
```

```{r}
#| label: sirius
summary_results <- list()
full_results    <- list()

# Import + run + collect for a single-polarity Spectra in batches of
# `batch_size` features. Each batch clears the project (delete=TRUE) so
# SIRIUS never holds more than one batch worth of features at a time.
run_sirius_for_polarity <- function(srs, sp, polarity_label,
                                    batch_size = 200) {
  if (length(sp) == 0L)
    return(list(summary = NULL, full = NULL))

  groups  <- unique(sp$group_idx)
  batches <- split(groups, ceiling(seq_along(groups) / batch_size))

  sums  <- vector("list", length(batches))
  fulls <- vector("list", length(batches))
  for (i in seq_along(batches)) {
    sp_b <- sp[sp$group_idx %in% batches[[i]]]
    message("  ", polarity_label, " batch ", i, "/", length(batches),
            " — ", length(batches[[i]]), " features")

    srs <- import(srs,
      spectra                = sp_b,
      ms_column_name         = "group_idx",
      deleteExistingFeatures = TRUE)

    run(srs,
      formulaIdParams = formulaIdParam(
        numberOfCandidates               = 15,
        instrument                       = "QTOF",
        numberOfCandidatesPerIonization  = 5,
        massAccuracyMS2ppm               = 10,
        filterByIsotopePattern           = FALSE,
        isotopeMs2Settings               = "SCORE",
        performDeNovoBelowMz             = 600
      ),
      predictParams           = predictParam(),
      structureDbSearchParams = structureDbSearchParam(
        structureSearchDbs = c("BIO")
      ),
      recompute = TRUE,
      wait      = TRUE
    )

    sums[[i]] <- summary(srs, result.type = "structure") |>
      mutate(across(c(confidenceExactMatch, confidenceApproxMatch), as.numeric))
    fulls[[i]] <- results(srs,
      result.type  = "structureDb",
      topFormula   = 8,
      topStructure = 20,
      return.type  = "data.frame")
  }
  list(summary = bind_rows(sums), full = bind_rows(fulls))
}

for (lab in labs) {
  sum_path  <- file.path(output_dir, paste0("sirius_summary_", lab, ".csv"))
  full_path <- file.path(output_dir, paste0("sirius_full_",    lab, ".csv"))

  if (file.exists(sum_path) && file.exists(full_path)) {
    message("Loading cached SIRIUS results for: ", lab)
    summary_results[[lab]] <- read.csv(sum_path, stringsAsFactors = FALSE) |>
      mutate(across(c(confidenceExactMatch, confidenceApproxMatch), as.numeric))
    full_results[[lab]] <- read.csv(full_path, stringsAsFactors = FALSE)
    next
  }

  message("=== SIRIUS: ", lab, " ===")
  if (!exists("sirius_user")) sirius_user <- Sys.getenv("SIRIUS_USERNAME")
  if (!exists("sirius_pass")) sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")

  srs <- Sirius(
    username  = sirius_user,
    password  = sirius_pass,
    projectId = paste0("ring_trial_allms2_", lab),
    path      = output_dir
  )

  res_pos <- run_sirius_for_polarity(srs, spectra_all[[lab]]$spectra_pos, "pos")
  res_neg <- run_sirius_for_polarity(srs, spectra_all[[lab]]$spectra_neg, "neg")

  summary_results[[lab]] <- bind_rows(res_pos$summary, res_neg$summary) |>
    mutate(lab = lab)
  full_results[[lab]]    <- bind_rows(res_pos$full,    res_neg$full)    |>
    mutate(lab = lab)

  write.csv(summary_results[[lab]], sum_path,  row.names = FALSE)
  write.csv(full_results[[lab]],    full_path, row.names = FALSE)
  shutdown(srs)
}

all_summaries <- bind_rows(lapply(summary_results, function(df) {
  mutate(df, across(c(confidenceExactMatch, confidenceApproxMatch), as.numeric))
}))
all_full <- bind_rows(full_results)

write.csv(all_summaries,
          file.path(output_dir, "sirius_summary_all_labs.csv"), row.names = FALSE)
write.csv(all_full,
          file.path(output_dir, "sirius_full_all_labs.csv"),    row.names = FALSE)
```

## 4. Match to expected compounds (mixture 1.1)

```{r}
#| label: expected
expected <- meta |>
  filter(Mixture == "1.1") |>
  transmute(
    compound             = `ChEBI name`,
    expected_formula     = formula,
    expected_inchikey    = InChIKey,
    expected_inchikey_2d = substr(InChIKey, 1, 14)
  )

cat("Expected compounds in mixture 1.1:", nrow(expected), "\n")
```

```{r}
#| label: match-structure
expected_keys_2d <- unique(expected$expected_inchikey_2d)

structure_hits_all <- all_full |>
  mutate(
    inchikey_2d     = substr(inchiKey, 1, 14),
    structure_match = inchikey_2d %in% expected_keys_2d,
    polarity        = ifelse(grepl("^\\[M[-+]", adduct) &
                               grepl("-\\]", adduct), "neg", "pos")
  ) |>
  filter(structure_match) |>
  group_by(lab, inchikey_2d, polarity) |>
  summarise(
    best_formula_rank   = if (all(is.na(rank.x))) NA_integer_ else min(rank.x, na.rm = TRUE),
    best_structure_rank = if (all(is.na(rank.y))) NA_integer_ else min(rank.y, na.rm = TRUE),
    adducts             = paste(sort(unique(na.omit(adduct))), collapse = ";"),
    molecularFormula    = first(molecularFormula),
    .groups = "drop"
  ) |>
  left_join(expected, by = c("inchikey_2d" = "expected_inchikey_2d"))

structure_hits_all
```

## 5. Coverage

```{r}
#| label: coverage
covered_all <- expected |>
  mutate(
    structure_recovered = expected_inchikey_2d %in% structure_hits_all$inchikey_2d
  )

cat("Structure recovered:", sum(covered_all$structure_recovered),
    "/", nrow(covered_all), "\n")
covered_all
```

```{r}
#| label: plot-coverage-by-lab
#| fig-cap: "Structure hits per lab and polarity (all-MS2 strategy)"
structure_hits_all |>
  count(lab, polarity) |>
  ggplot(aes(x = lab, y = n, fill = polarity)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "Structure hits", fill = "Polarity") +
  theme_bw()
```

## 6. Comparison to consensus-based annotations

Load the structure hits from `sirius_annotation_consensus.qmd` and compare
compound recovery side by side.

```{r}
#| label: compare
consensus_path <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                       "results_consensus_sample1_pos",
                       "consensus_match_summary.csv")

if (file.exists(consensus_path)) {
  consensus_summary <- read.csv(consensus_path, stringsAsFactors = FALSE)

  # Compounds recovered in each strategy
  recovered_consensus <- expected$expected_inchikey_2d[
    expected$expected_inchikey_2d %in%
      substr(consensus_summary$inchikey_2d, 1, 14)
  ]

  recovered_allms2 <- structure_hits_all$inchikey_2d

  comparison <- expected |>
    mutate(
      in_consensus = expected_inchikey_2d %in% recovered_consensus,
      in_all_ms2   = expected_inchikey_2d %in% recovered_allms2
    )

  cat("Consensus strategy recovered:  ",
      sum(comparison$in_consensus), "/", nrow(expected), "\n")
  cat("All-MS2 strategy recovered:    ",
      sum(comparison$in_all_ms2),   "/", nrow(expected), "\n")
  cat("Recovered by both:             ",
      sum(comparison$in_consensus & comparison$in_all_ms2), "\n")
  cat("Only in consensus:             ",
      sum(comparison$in_consensus & !comparison$in_all_ms2), "\n")
  cat("Only in all-MS2:               ",
      sum(!comparison$in_consensus & comparison$in_all_ms2), "\n")

  comparison
} else {
  message("Run sirius_annotation_consensus.qmd first to generate the comparison file.")
}
```
