---
title: "SIRIUS annotation of consensus peaks (all mixtures)"
format: html
---

## Goal

Annotate the chromatographic peaks that were detected by **all four labs**
(from `ct_filtered4.csv`) using SIRIUS via `RuSirius`, then check whether
the annotations are consistent with the compounds expected in each mixture
(`standards.xlsx`).

Scope:

- All 38 samples (19 pos + 19 neg) covering mixtures 1_1–3_3.
- Only peaks present in **all four labs** (`ct_filtered4`).
- One SIRIUS project per lab (all mixtures together); results then merged
  on `consensus_id`.
- Match SIRIUS results to per-mixture standards on **formula** first, then
  on **InChIKey** (first block = 2D structure).

## Setup

```{r}
#| label: setup
#| message: false
library(Spectra)
library(MsExperiment)
library(xcms)
library(RuSirius)
library(readxl)
library(dplyr)
library(tidyr)
library(here)

source(here("1_preprocessing", "setup.R"))   # automate_matching_ms2, meta
```

## Config

```{r}
#| label: config
labs <- c("afekta", "cembio", "hmgu", "icl")
study_group <- "HE"
waters_labs <- c("icl")

# Samples 1-19 = pos, 20-38 = neg (same 19 mixtures in each polarity)
mixtures <- c("1_1","1_2","1_3","1_4","1_5","1_6",
              "2_1","2_10","2_2","2_3","2_4","2_5","2_6","2_7","2_8","2_9",
              "3_1","3_2","3_3")
sample_map <- tibble(
  sample   = c(1:19, 20:38),
  polarity = c(rep("pos", 19), rep("neg", 19)),
  mixture  = rep(mixtures, 2)
)

obj_dir    <- here("5_downstream_analysis", "object")
ct_path    <- file.path(obj_dir, "ct_filtered4.csv")
output_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                   "results_consensus_all")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Load all consensus peaks (all samples, all mixtures)

```{r}
#| label: load-ct
ct <- read.csv(ct_path, stringsAsFactors = FALSE)
ct <- left_join(ct, sample_map, by = "sample")
```

`ct` is now one row per (consensus peak, sample) with polarity and mixture
added from `sample_map`. From here we build a per-lab table of
`chrom_peak_id` ↔ `consensus_id` so we can re-join SIRIUS results to the
consensus later.

```{r}
#| label: per-lab-peak-tables
peaks_by_lab <- lapply(labs, function(lab) {
  data.frame(
    consensus_id  = ct$consensus_id,
    chrom_peak_id = ct[[paste0("idx_", lab)]],
    polarity      = ct$polarity,
    mixture       = ct$mixture,
    stringsAsFactors = FALSE
  )
})
names(peaks_by_lab) <- labs
```

## 2. Spectra extraction helper

Adapted from `sirius_annotation.qmd::extract_spectra_for_sirius`. Difference:
instead of expanding pipe-separated peak IDs from a curated annotation table,
we take a plain `data.frame` of `chrom_peak_id`s (with the
mixture/polarity columns needed by `automate_matching_ms2()`).

```{r}
#| label: helper-extract
extract_spectra_from_peaks <- function(lab_name,
                                       peak_df,
                                       study_group,
                                       is_waters = FALSE) {
  study_path <- here("1_preprocessing", lab_name, study_group)

  mse <- readMsObject(
    XcmsExperiment(),
    AlabasterParam(path = file.path(study_path, "mse")),
    spectraPath = file.path(study_path, "mzml")
  )
  sampleData(mse)$mixture <- sub(".*_", "", sampleData(mse)$Sample.Name)
  sampleData(mse)$mixture <- gsub("\\.", "_", sampleData(mse)$mixture)

  cp <- chromPeaks(mse)
  peak_df <- peak_df[peak_df$chrom_peak_id %in% rownames(cp), ]

  # Pull mz/rt ranges from chromPeaks
  cp_info <- as.data.frame(cp[peak_df$chrom_peak_id, , drop = FALSE])
  peak_df$mz    <- cp_info$mz
  peak_df$mzmin <- cp_info$mzmin
  peak_df$mzmax <- cp_info$mzmax
  peak_df$rt    <- cp_info$rt
  peak_df$rtmin <- cp_info$rtmin
  peak_df$rtmax <- cp_info$rtmax
  rownames(peak_df) <- peak_df$chrom_peak_id

  # MS1
  ms1 <- chromPeakSpectra(
    mse, msLevel = 1L,
    peaks = peak_df$chrom_peak_id,
    method = "closest_rt"
  )
  ms1 <- setBackend(ms1, MsBackendMemory())
  ms1 <- applyProcessing(ms1)
  ms1 <- filterEmptySpectra(ms1)

  # MS2 (some labs have a separate mse2)
  mse2_path <- file.path(study_path, "mse2")
  if (dir.exists(mse2_path)) {
    mse2 <- readMsObject(
      XcmsExperiment(),
      AlabasterParam(path = mse2_path),
      spectraPath = file.path(study_path, "mzml")
    )
  } else {
    mse2 <- mse
  }
  sn2 <- sub("_MS2$", "", sampleData(mse2)$Sample.Name)
  sampleData(mse2)$mixture <- sub(".*_", "", sn2)
  sampleData(mse2)$mixture <- gsub("\\.", "_", sampleData(mse2)$mixture)

  ms2 <- automate_matching_ms2(
    mse2 = mse2,
    match_res = peak_df,
    waters_data = is_waters
  )
  low_int <- function(x, ...) x > max(x, na.rm = TRUE) * 0.05
  ms2 <- filterIntensity(ms2, intensity = low_int)
  ms2 <- ms2[lengths(ms2) > 1]
  ms2 <- setBackend(ms2, MsBackendMemory())
  ms2 <- applyProcessing(ms2)
  ms2 <- filterEmptySpectra(ms2)

  pos_ids <- peak_df$chrom_peak_id[peak_df$polarity == "pos"]
  neg_ids <- peak_df$chrom_peak_id[peak_df$polarity == "neg"]

  # Count MS1 before filtering (sanity check: should be ~1 per peak)
  n_ms1_pos_all <- sum(ms1$chrom_peak_id %in% pos_ids)
  n_ms1_neg_all <- sum(ms1$chrom_peak_id %in% neg_ids)

  # Keep only MS1 for peaks that have at least one MS2 (= features sent to SIRIUS)
  ms1 <- ms1[ms1$chrom_peak_id %in% unique(ms2$chrom_peak_id)]

  ms2_peak_ids <- unique(ms2$chrom_peak_id)
  consensus_w_ms2 <- unique(
    peak_df$consensus_id[peak_df$chrom_peak_id %in% ms2_peak_ids]
  )

  list(
    spectra              = c(ms1, ms2),
    n_ms1_pos            = n_ms1_pos_all,
    n_ms1_neg            = n_ms1_neg_all,
    n_ms2_pos            = sum(ms2$chrom_peak_id %in% pos_ids),
    n_ms2_neg            = sum(ms2$chrom_peak_id %in% neg_ids),
    n_peaks_w_ms2_pos    = length(intersect(ms2_peak_ids, pos_ids)),
    n_peaks_w_ms2_neg    = length(intersect(ms2_peak_ids, neg_ids)),
    n_consensus_w_ms2    = length(consensus_w_ms2),
    n_peaks              = nrow(peak_df)
  )
}
```

## 3 & 4. Extraction + SIRIUS per mixture per lab

For each mixture, spectra are extracted and SIRIUS is run for each lab.
A (mixture, lab) combination is skipped entirely if both result files already
exist. Results are saved under `results_consensus_all/<mix_slug>/`.

SIRIUS 6.3+ must be running. Credentials from `.Renviron`
(`SIRIUS_USERNAME`, `SIRIUS_PASSWORD`).

```{r}
#| label: sirius-login
#| eval: false
sirius_user <- "plouail@eurac.edu"
sirius_pass <- "Terriere85!!"
```

```{r}
#| label: extract-and-sirius
#| message: false
if (!exists("sirius_user")) sirius_user <- Sys.getenv("SIRIUS_USERNAME")
if (!exists("sirius_pass")) sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")

summary_results    <- list()
full_results       <- list()
extraction_summary <- list()

for (mix in mixtures) {
  mix_slug <- mix
  mix_dir  <- file.path(output_dir, mix_slug)
  dir.create(mix_dir, recursive = TRUE, showWarnings = FALSE)

  for (lab in labs) {
    sum_path  <- file.path(mix_dir, paste0("sirius_summary_", lab, ".csv"))
    full_path <- file.path(mix_dir, paste0("sirius_full_",    lab, ".csv"))

    if (file.exists(sum_path) && file.exists(full_path)) {
      message("Cached: mix ", mix, " / ", lab)
      sum_df <- tryCatch(read.csv(sum_path, stringsAsFactors = FALSE), error = function(e) data.frame())
      if (nrow(sum_df) == 0) next   # empty sentinel = no MS2
      summary_results[[mix]][[lab]] <- sum_df |>
        mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
               across(any_of("alignedFeatureId"), as.character),
               mixture = mix, lab = lab)
      full_results[[mix]][[lab]] <- tryCatch(read.csv(full_path, stringsAsFactors = FALSE), error = function(e) data.frame()) |>
        mutate(across(any_of("alignedFeatureId"), as.character),
               mixture = mix, lab = lab)
      next
    }

    # Extract spectra for this mixture only
    message("Extracting: mix ", mix, " / ", lab)
    peak_df_mix <- peaks_by_lab[[lab]][peaks_by_lab[[lab]]$mixture == mix, ]
    res <- extract_spectra_from_peaks(
      lab_name    = lab,
      peak_df     = peak_df_mix,
      study_group = study_group,
      is_waters   = lab %in% waters_labs
    )
    extraction_summary[[mix]][[lab]] <- tibble(
      mixture           = mix,
      lab               = lab,
      n_peaks           = res$n_peaks,
      n_ms1_pos         = res$n_ms1_pos,
      n_ms1_neg         = res$n_ms1_neg,
      n_ms2_pos         = res$n_ms2_pos,
      n_ms2_neg         = res$n_ms2_neg,
      n_peaks_w_ms2_pos = res$n_peaks_w_ms2_pos,
      n_peaks_w_ms2_neg = res$n_peaks_w_ms2_neg,
      n_consensus_w_ms2 = res$n_consensus_w_ms2
    )

    sp          <- res$spectra
    ms2_ids_pos <- unique(sp[sp$msLevel == 2L & sp$polarity == 1L]$chrom_peak_id)
    ms2_ids_neg <- unique(sp[sp$msLevel == 2L & sp$polarity == 0L]$chrom_peak_id)

    if (length(ms2_ids_pos) == 0 && length(ms2_ids_neg) == 0) {
      message("No MS2 for mix ", mix, " / ", lab, " — skipping SIRIUS")
      writeLines(character(0), sum_path)
      writeLines(character(0), full_path)
      next
    }

    message("SIRIUS: mix ", mix, " / ", lab)
    srs <- Sirius(
      username  = sirius_user,
      password  = sirius_pass,
      projectId = paste0("ring_trial_", mix_slug, "_", lab),
      path      = mix_dir
    )

    # Import pos and neg separately (RuSirius needs this for correct polarity)
    if (length(ms2_ids_pos) > 0)
      srs <- import(srs,
        spectra                = sp[sp$chrom_peak_id %in% ms2_ids_pos],
        ms_column_name         = "chrom_peak_id",
        deleteExistingFeatures = TRUE)
    if (length(ms2_ids_neg) > 0)
      srs <- import(srs,
        spectra                = sp[sp$chrom_peak_id %in% ms2_ids_neg],
        ms_column_name         = "chrom_peak_id",
        deleteExistingFeatures = length(ms2_ids_pos) == 0)

    run(
      srs,
      formulaIdParams = formulaIdParam(
        numberOfCandidates              = 15,
        instrument                      = "QTOF",
        numberOfCandidatesPerIonization = 5,
        massAccuracyMS2ppm              = 10,
        filterByIsotopePattern          = FALSE,
        isotopeMs2Settings              = "SCORE",
        performDeNovoBelowMz            = 600
      ),
      predictParams           = predictParam(),
      structureDbSearchParams = structureDbSearchParam(
        structureSearchDbs = c("BIO")
      ),
      recompute = TRUE,
      wait      = TRUE
    )

    summary_results[[mix]][[lab]] <- summary(srs, result.type = "structure") |>
      mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
             across(any_of("alignedFeatureId"), as.character),
             mixture = mix, lab = lab)

    full_results[[mix]][[lab]] <- results(
      srs,
      result.type  = "structureDb",
      topFormula   = 8,
      topStructure = 20,
      return.type  = "data.frame"
    ) |>
      mutate(across(any_of("alignedFeatureId"), as.character),
             mixture = mix, lab = lab)

    write.csv(summary_results[[mix]][[lab]], sum_path,  row.names = FALSE)
    write.csv(full_results[[mix]][[lab]],    full_path, row.names = FALSE)

    shutdown(srs)
  }
}

all_summaries <- bind_rows(lapply(summary_results, bind_rows))
all_full      <- bind_rows(lapply(full_results,    bind_rows))

write.csv(all_summaries,
          file.path(output_dir, "sirius_summary_all.csv"), row.names = FALSE)
write.csv(all_full,
          file.path(output_dir, "sirius_full_all.csv"),    row.names = FALSE)
```

```{r}
#| label: plot-extraction-summary
#| fig-cap: "Peaks with MS2 per lab and polarity (all mixtures combined)"
library(ggplot2)

extr_df <- bind_rows(lapply(extraction_summary, bind_rows))
if (nrow(extr_df) == 0) {
  message("extraction_summary empty (all results cached) — skipping plot")
} else {
  extr_df |>
    group_by(lab) |>
    summarise(pos = sum(n_peaks_w_ms2_pos), neg = sum(n_peaks_w_ms2_neg),
              .groups = "drop") |>
    tidyr::pivot_longer(c(pos, neg), names_to = "polarity", values_to = "n") |>
    ggplot(aes(x = lab, y = n, fill = polarity)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
    labs(x = NULL, y = "Features with MS2", fill = "Polarity") +
    theme_bw()
}
```

## 5. Re-attach `consensus_id` to SIRIUS output

SIRIUS keys results on `externalFeatureId` (= `chrom_peak_id`) in
`summary()` and on `xcms_fts` in `results()`. Both are per-lab; we need
`consensus_id` to compare the four labs row-by-row. `mixture` is already
present from the extraction loop.

```{r}
#| label: attach-consensus
peaks_long <- bind_rows(lapply(labs, function(lab) {
  peaks_by_lab[[lab]] |>
    mutate(lab = lab) |>
    select(lab, chrom_peak_id, consensus_id, polarity, mixture)
}))

all_summaries <- all_summaries |>
  left_join(peaks_long,
            by = c("lab", "mixture", "externalFeatureId" = "chrom_peak_id"))

all_full <- all_full |>
  left_join(peaks_long,
            by = c("lab", "mixture", "xcms_fts" = "chrom_peak_id"))
```

## 5b. RT per chrom_peak_id

SIRIUS output doesn't carry retention time, so we look it up from
`chromPeaks(mse)` per lab. Cached per lab so it only runs once.

```{r}
#| label: peak-rt
#| message: false
peak_rt_per_lab <- list()

for (lab in labs) {
  rt_path <- file.path(output_dir, paste0("peak_rt_", lab, ".csv"))

  if (file.exists(rt_path)) {
    peak_rt_per_lab[[lab]] <- read.csv(rt_path, stringsAsFactors = FALSE)
    next
  }

  message("Loading chromPeaks for: ", lab)
  study_path <- here("1_preprocessing", lab, study_group)
  mse <- readMsObject(
    XcmsExperiment(),
    AlabasterParam(path = file.path(study_path, "mse")),
    spectraPath = file.path(study_path, "mzml")
  )
  cp <- chromPeaks(mse)

  pl <- peaks_by_lab[[lab]]
  pl <- pl[pl$chrom_peak_id %in% rownames(cp), ]
  pl$rt <- cp[pl$chrom_peak_id, "rt"]

  df <- data.frame(
    lab           = lab,
    consensus_id  = pl$consensus_id,
    chrom_peak_id = pl$chrom_peak_id,
    polarity      = pl$polarity,
    mixture       = pl$mixture,
    rt            = pl$rt,
    stringsAsFactors = FALSE
  )
  write.csv(df, rt_path, row.names = FALSE)
  peak_rt_per_lab[[lab]] <- df
}

peak_rt_long <- bind_rows(peak_rt_per_lab)
```

## 6. Expected compounds per mixture

`meta` was loaded by `setup.R`. Columns of interest:
`Name`, `ChEBI name`, `Mixture`, `M`, `formula`, `InChIKey`, `SMILES`.

```{r}
#| label: expected
expected_all <- meta |>
  filter(gsub("\\.", "_", Mixture) %in% mixtures) |>
  transmute(
    mixture              = gsub("\\.", "_", Mixture),
    compound             = `ChEBI name`,
    expected_formula     = formula,
    expected_inchikey    = InChIKey,
    expected_inchikey_2d = substr(InChIKey, 1, 14)
  )

cat("Expected compounds across all mixtures:", nrow(expected_all), "\n")
expected_all |> count(mixture)
```

## 7. Match formulas

For every SIRIUS candidate (across all ranks), flag whether its
`molecularFormula` matches an expected formula **for that mixture**, and
record the SIRIUS rank at which the match occurred.

```{r}
#| label: match-formula
# rank.x = formula rank (1 = top), rank.y = structure rank
formula_hits <- all_full |>
  left_join(
    select(expected_all, mixture, compound, expected_formula, expected_inchikey_2d),
    by = c("mixture", "molecularFormula" = "expected_formula")
  ) |>
  filter(!is.na(compound)) |>
  group_by(lab, consensus_id, mixture, molecularFormula, polarity) |>
  summarise(
    best_formula_rank = if (all(is.na(rank.x))) NA_integer_ else min(rank.x, na.rm = TRUE),
    adducts           = paste(sort(unique(na.omit(adduct))), collapse = ";"),
    .groups = "drop"
  )

formula_hits
```

## 8. Match InChIKey (2D structure)

The first 14 chars of an InChIKey encode the 2D skeleton. Matching is done
per mixture so that a compound only counts as a hit in the mixture where it
is expected.

```{r}
#| label: match-inchikey
structure_hits <- all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  left_join(
    select(expected_all, mixture, compound, expected_formula, expected_inchikey_2d),
    by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")
  ) |>
  filter(!is.na(compound)) |>
  group_by(lab, consensus_id, mixture, inchikey_2d, polarity) |>
  summarise(
    best_formula_rank   = if (all(is.na(rank.x))) NA_integer_ else min(rank.x, na.rm = TRUE),
    best_structure_rank = if (all(is.na(rank.y))) NA_integer_ else min(rank.y, na.rm = TRUE),
    adducts             = paste(sort(unique(na.omit(adduct))), collapse = ";"),
    structureName       = first(structureName),
    molecularFormula    = first(molecularFormula),
    compound            = first(compound),
    .groups = "drop"
  )

structure_hits
```

```{r}
#| label: plot-annotation-coverage
#| fig-cap: "Structure hits per lab and polarity (all mixtures)"
structure_hits |>
  count(lab, polarity) |>
  ggplot(aes(x = lab, y = n, fill = polarity)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "Structure hits", fill = "Polarity") +
  theme_bw()
```

```{r}
#| label: plot-annotation-coverage-per-mixture
#| fig-cap: "Structure hits per mixture and polarity"
structure_hits |>
  count(mixture, polarity) |>
  ggplot(aes(x = mixture, y = n, fill = polarity)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "Structure hits", fill = "Polarity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: plot-adduct-distribution
#| fig-cap: "Adduct distribution across structure hits"
structure_hits |>
  tidyr::separate_rows(adducts, sep = ";") |>
  filter(!is.na(adducts), adducts != "") |>
  count(polarity, adducts) |>
  ggplot(aes(x = reorder(adducts, n), y = n, fill = polarity)) +
  geom_col() +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = "Polarity") +
  theme_bw()
```

## 9. Per-compound summary across labs and polarities

One row per (mixture, compound): how many labs detected it in each polarity,
which adducts were seen, best ranks, and — when detected in both pos and neg
— the RT difference (in seconds) between the two polarity peaks.

```{r}
#| label: per-consensus-summary
by_polarity <- structure_hits |>
  left_join(peak_rt_long,
            by = c("lab", "consensus_id", "polarity", "mixture")) |>
  group_by(mixture, compound, molecularFormula, inchikey_2d, polarity) |>
  summarise(
    n_labs              = n_distinct(lab),
    adducts             = paste(sort(unique(adducts)), collapse = ";"),
    rt_s                = mean(rt, na.rm = TRUE),
    best_formula_rank   = if (all(is.na(best_formula_rank)))   NA_integer_ else min(best_formula_rank,   na.rm = TRUE),
    best_structure_rank = if (all(is.na(best_structure_rank))) NA_integer_ else min(best_structure_rank, na.rm = TRUE),
    .groups = "drop"
  )

summary_table <- by_polarity |>
  mutate(polarity = factor(polarity, levels = c("pos", "neg"))) |>
  pivot_wider(
    names_from   = polarity,
    values_from  = c(n_labs, adducts, rt_s, best_formula_rank, best_structure_rank),
    names_sep    = "_",
    names_expand = TRUE
  ) |>
  mutate(rt_diff_s = abs(rt_s_pos - rt_s_neg)) |>
  arrange(mixture, compound)

summary_table

write.csv(summary_table,
          file.path(output_dir, "consensus_match_summary.csv"),
          row.names = FALSE)
```

## 10. Coverage check

How many of the expected compounds per mixture were recovered (at any rank,
any lab)?

```{r}
#| label: coverage
covered <- expected_all |>
  mutate(
    formula_recovered   = paste(mixture, expected_formula)     %in%
                          paste(formula_hits$mixture,   formula_hits$molecularFormula),
    structure_recovered = paste(mixture, expected_inchikey_2d) %in%
                          paste(structure_hits$mixture, structure_hits$inchikey_2d)
  ) |>
  arrange(mixture, compound)

covered

cat("\nFormula recovered:   ", sum(covered$formula_recovered),
    "/", nrow(covered), "\n")
cat("Structure recovered: ", sum(covered$structure_recovered),
    "/", nrow(covered), "\n")

covered |>
  group_by(mixture) |>
  summarise(
    n_expected          = n(),
    n_formula_recovered = sum(formula_recovered),
    n_struct_recovered  = sum(structure_recovered),
    .groups = "drop"
  )
```

```{r}
#| label: plot-coverage-per-mixture
#| fig-cap: "Compounds found vs expected per mixture"
covered |>
  group_by(mixture) |>
  summarise(
    Expected  = n(),
    Formula   = sum(formula_recovered),
    Structure = sum(structure_recovered),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(c(Expected, Formula, Structure),
                      names_to = "type", values_to = "n") |>
  mutate(type = factor(type, levels = c("Expected", "Formula", "Structure"))) |>
  ggplot(aes(x = mixture, y = n, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Expected = "#aec7e8", Formula = "#1f77b4", Structure = "#d62728")) +
  labs(x = "Mixture", y = "Number of compounds", fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 11. GNPS library lookup for unannotated compounds

For compounds that SIRIUS did not annotate (no structure hit in any lab),
check whether the GNPS reference library contains a spectrum with the same
2D InChIKey. A compound *present* in GNPS but missing from our annotation
is a SIRIUS / MS2 failure; a compound *absent* from GNPS could never have
been annotated by MS2 library matching.

```{r}
#| label: gnps-unannotated
#| message: false
library(DBI)
library(RSQLite)
library(MsBackendSql)

unannotated <- expected_all |>
  filter(!paste(mixture, expected_inchikey_2d) %in%
           paste(structure_hits$mixture, structure_hits$inchikey_2d))

cat("Unannotated compounds (all mixtures): ",
    nrow(unannotated), "/", nrow(expected_all), "\n", sep = "")

dbfile <- here("2_annotation_auto",
               "MsBackendSql.GNPS.matchms.cleaned.v1.sqlite")
con <- dbConnect(SQLite(), dbfile)
ms2_ref <- Spectra(con, source = MsBackendSql())
gnps_keys_2d <- unique(sub("-.*", "", ms2_ref$inchikey))
gnps_keys_2d <- gnps_keys_2d[!is.na(gnps_keys_2d) & nzchar(gnps_keys_2d)]
dbDisconnect(con)

unannotated <- unannotated |>
  mutate(in_gnps = expected_inchikey_2d %in% gnps_keys_2d) |>
  arrange(mixture, desc(in_gnps), compound)

cat("Present in GNPS (SIRIUS gap, not library gap): ",
    sum(unannotated$in_gnps), "/", nrow(unannotated), "\n", sep = "")
cat("Absent from GNPS (library gap):                ",
    sum(!unannotated$in_gnps), "/", nrow(unannotated), "\n", sep = "")

unannotated |>
  select(mixture, compound, expected_formula, expected_inchikey_2d, in_gnps)

write.csv(unannotated,
          file.path(output_dir, "unannotated_vs_gnps.csv"),
          row.names = FALSE)
```

## 12. Total compounds found

### 12a. Comparison with per-lab curated annotations

```{r}
#| label: total-compounds-found
#| message: false
total_expected  <- nrow(expected_all)
total_formula   <- sum(covered$formula_recovered)
total_structure <- sum(covered$structure_recovered)

RT_THRESHOLD <- 15   # seconds — same tolerance as consensus peak grouping

# --- Per-lab counts from curated XLSXs (same source as 6_annotation_analysis.qmd) ---
curated_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation")
perlab_files <- file.path(curated_dir,
  paste0(labs, "_lab_annotations_sirius_curated.xlsx"))

if (all(file.exists(perlab_files))) {
  ct4 <- read.csv(ct_path, stringsAsFactors = FALSE)
  peaks_in_ct4 <- lapply(labs, function(lab) {
    as.character(na.omit(ct4[[paste0("idx_", lab)]]))
  })
  names(peaks_in_ct4) <- labs

  # Mean RT per consensus_id across all labs (reference RT)
  consensus_rt <- peak_rt_long |>
    group_by(consensus_id) |>
    summarise(rt_consensus = mean(rt, na.rm = TRUE), .groups = "drop")

  rank_lab_all <- bind_rows(lapply(labs, function(lab) {
    x <- as.data.frame(readxl::read_xlsx(perlab_files[labs == lab]),
                       check.names = FALSE)
    peak_ids <- strsplit(as.character(x$chrom_peak_id), "|", fixed = TRUE)
    # First ct4 peak_id per row (for RT join)
    first_ct4 <- vapply(peak_ids, function(ids) {
      m <- trimws(ids)[trimws(ids) %in% peaks_in_ct4[[lab]]]
      if (length(m)) m[1L] else NA_character_
    }, character(1))
    in_ct4 <- !is.na(first_ct4)
    data.frame(
      lab            = lab,
      compound       = trimws(x$target_ChEBI.name),
      structure_rank = suppressWarnings(as.numeric(x$best_structurePerIdRank)),
      rtmed          = suppressWarnings(as.numeric(x$rtmed)),
      chrom_peak_id  = first_ct4,
      in_ct4         = in_ct4,
      stringsAsFactors = FALSE
    )
  })) |>
    filter(!is.na(compound), compound != "") |>
    # join → consensus_id → consensus mean RT
    left_join(peaks_long |> select(lab, chrom_peak_id, consensus_id),
              by = c("lab", "chrom_peak_id")) |>
    left_join(consensus_rt, by = "consensus_id") |>
    mutate(rt_diff = abs(rtmed - rt_consensus))

  # Restrict: peak in ct_filtered4 AND rtmed within RT_THRESHOLD of consensus RT
  rank_ct4 <- rank_lab_all |>
    filter(in_ct4, is.na(rt_diff) | rt_diff <= RT_THRESHOLD)

  found_compounds_ct4 <- rank_ct4 |>
    filter(!is.na(structure_rank)) |>
    pull(compound) |> unique()
  n_found_ct4 <- sum(expected_all$compound %in% found_compounds_ct4)

  perlab_row <- tibble(
    approach    = "Per-lab curated (ct_filtered4 peaks, RT ≤ 15 s from consensus)",
    n_expected  = total_expected,
    n_structure = n_found_ct4,
    pct         = round(n_found_ct4 / total_expected * 100, 1)
  )
} else {
  message("Curated lab XLSXs not found — skipping per-lab comparison")
  perlab_row <- NULL
}

comparison_tbl <- bind_rows(
  perlab_row,
  tibble(approach    = "Consensus peaks (this document)",
         n_expected  = total_expected,
         n_structure = total_structure,
         pct         = round(total_structure / total_expected * 100, 1))
)
print(comparison_tbl)

covered |>
  filter(structure_recovered) |>
  select(mixture, compound, expected_formula, expected_inchikey_2d) |>
  arrange(mixture, compound)
```

### 12b. Per-lab annotation results (consensus peaks)

```{r}
#| label: perlab-annotation-table
perlab_ann <- structure_hits |>
  group_by(lab, polarity) |>
  summarise(n_compounds = n_distinct(compound), .groups = "drop")

perlab_ann_mix <- structure_hits |>
  group_by(lab, polarity, mixture) |>
  summarise(n_compounds = n_distinct(compound), .groups = "drop")

cat("Overall structure hits per lab and polarity:\n")
print(perlab_ann)
```

```{r}
#| label: plot-perlab-overall
#| fig-cap: "Structure hits per lab and polarity (all mixtures)"
ggplot(perlab_ann, aes(x = lab, y = n_compounds, fill = polarity)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = n_compounds),
            position = position_dodge(0.9), vjust = -0.3, size = 3) +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "Unique compounds", fill = "Polarity") +
  theme_bw()
```

```{r}
#| label: plot-perlab-per-mixture
#| fig-cap: "Structure hits per lab, polarity and mixture"
#| fig-height: 7
perlab_ann_mix |>
  mutate(mixture = factor(mixture, levels = mixtures)) |>
  ggplot(aes(x = lab, y = n_compounds, fill = polarity)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  facet_wrap(~ mixture, ncol = 4) +
  labs(x = NULL, y = "Unique compounds", fill = "Polarity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text  = element_text(size = 7))
```

## 13. Consensus-peak funnel summary

How many consensus peaks we started with, how many had MS2 in 1/2/3/4 labs,
and how many were annotated (structure hit) in the end.

```{r}
#| label: consensus-funnel
n_start <- n_distinct(ct$consensus_id)

# MS2 per (consensus, lab): a consensus_id appears in all_full for a given lab
# iff that lab had MS2 for one of its chrom_peak_ids and SIRIUS produced results.
ms2_by_consensus <- all_full |>
  filter(!is.na(consensus_id)) |>
  distinct(consensus_id, lab) |>
  count(consensus_id, name = "n_labs_ms2")

ms2_dist <- ms2_by_consensus |>
  count(n_labs_ms2, name = "n_consensus") |>
  arrange(n_labs_ms2)

n_any_ms2 <- nrow(ms2_by_consensus)
n_no_ms2  <- n_start - n_any_ms2

n_annotated_peaks     <- n_distinct(structure_hits$consensus_id)
n_annotated_compounds <- sum(covered$structure_recovered)

cat("Starting consensus peaks:                  ", n_start, "\n", sep = "")
cat("With MS2 in >= 1 lab:                      ", n_any_ms2,
    " (", round(100 * n_any_ms2 / n_start, 1), "%)\n", sep = "")
cat("With NO MS2 in any lab:                    ", n_no_ms2,
    " (", round(100 * n_no_ms2 / n_start, 1), "%)\n", sep = "")
cat("Annotated consensus peaks (any lab):       ", n_annotated_peaks,
    " (", round(100 * n_annotated_peaks / n_start, 1), "%)\n", sep = "")
cat("  → distinct expected compounds recovered: ", n_annotated_compounds,
    " / ", nrow(expected_all), "\n\n", sep = "")

cat("MS2 coverage breakdown:\n")
print(ms2_dist)
```

```{r}
#| label: plot-consensus-funnel
#| fig-cap: "Consensus peaks by number of labs with MS2 (annotated portion in dark)"
annotated_ids <- unique(structure_hits$consensus_id)

ms2_by_consensus |>
  mutate(annotated = consensus_id %in% annotated_ids) |>
  count(n_labs_ms2, annotated) |>
  ggplot(aes(x = factor(n_labs_ms2), y = n, fill = annotated)) +
  geom_col() +
  scale_fill_manual(values = c(`FALSE` = "#cccccc", `TRUE` = "#1f77b4"),
                    labels  = c(`FALSE` = "Not annotated", `TRUE` = "Annotated")) +
  labs(x = "Number of labs with MS2", y = "Consensus peaks", fill = NULL) +
  theme_bw()
```

## Notes

- Pipeline mirrors `sirius_annotation.qmd` but is driven by
  `ct_filtered4.csv` instead of the manually curated annotation table.
- One SIRIUS project per (mixture, lab) stored under
  `results_consensus_all/<mix_slug>/`.
- Formula and structure matching is scoped per mixture: a SIRIUS hit only
  counts if the compound is expected in the sample's mixture.
