---
title: "SIRIUS annotation - all MS2 per lab (strategy 3)"
format:
  html:
    fig-width: 12
    fig-height: 8
    embed-resources: true
---

## Goal

Annotate **all MS2-bearing features** from every lab, without requiring
cross-lab consensus, and emit the two SIRIUS-side canonical CSVs
defined in
[`5_downstream_analysis/CANONICAL_SCHEMAS.md`](../../../5_downstream_analysis/CANONICAL_SCHEMAS.md):

- `annotation_identity.csv` — one row per `(mixture, lab, compound,
  polarity)` with best ranks, adducts, COSMIC scores, and a
  `confidence_tier` placeholder (`"unfiltered"`; P5 overwrites).
- `feat_rts.csv` — one row per identified feature with the MS2-trigger
  RT, `ionMass`, `adduct`, `cosmic`. The downstream MS1 anchoring +
  cluster + EIC steps in
  [`5_downstream_analysis/3_rt_alignment.qmd`](../../../5_downstream_analysis/3_rt_alignment.qmd)
  §P2.9-§P2.10 consume this and produce `annotation_chrom_metrics.csv`.

This file is a **pure producer**: no EIC extraction, no peak picking,
no cross-lab comparison. Those moved to `3_rt_alignment.qmd` so MS1
anchoring runs before clustering. The producer is now SIRIUS only.

## Workflow overview

```{mermaid}
flowchart LR
  A["Raw mzML"] --> B["MS2 noise filter<br/>5 pct base peak OR 1000 cnt"]
  B --> C{"Lab type"}
  C -->|cembio| D["rle group MS2 by<br/>precursor m/z<br/>per stepped-CE cycle"]
  C -->|"DDA: afekta hmgu icl"| E["fragmentGroupIndex<br/>MS1 with co-eluting MS2"]
  D --> F["Batch 200 features"]
  E --> F
  F --> G["SIRIUS<br/>formula ID + CSI:FingerID + BIO DB"]
  G --> H["Per batch CSV cache"]
  H --> I["Aggregate across<br/>labs and mixtures"]
  I --> J["Match to<br/>expected compounds"]
  J --> K["Extract MS2-trigger RT<br/>per feature"]
  K --> L["Write annotation_identity.csv<br/>and feat_rts.csv"]
  L -.->|consumed by| M["3_rt_alignment.qmd P2.9<br/>MS1 anchoring + cluster + EIC"]
```

## Setup

```{r}
#| label: setup
#| message: false
library(Spectra)
library(MsExperiment)
library(xcms)
library(RuSirius)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(MsQuality)
library(tidytext)

source(here("1_preprocessing", "setup.R"))
```

## Config

```{r}
#| label: config
labs        <- c("afekta", "cembio", "hmgu", "icl")
study_group <- "HE"
mixtures    <- c("1.1", "1.2", "1.3", "1.4", "1.5", "1.6",
                 "2.1", "2.2", "2.3", "2.4", "2.5",
                 "2.6", "2.7", "2.8", "2.9", "2.10",
                 "3.1", "3.2", "3.3")

mix_us     <- function(m) gsub("\\.", "_", m)
output_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                   "results_all_ms2")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Extraction helper

`extract_all_ms2()` loads the raw mzML files for one (lab, mixture) and
returns the MS1+MS2 spectra ready for SIRIUS, split by polarity.

**Acquisition layouts.** afekta, cembio and icl have dedicated `_MS2`
files; HMGU acquires MS1+MS2 together in the regular sample file. For
DDA labs (afekta, hmgu, icl) features are formed by `fragmentGroupIndex()`
(one MS1 + its triggered MS2). Cembio acquires MS1 and MS2 in temporally
disjoint blocks within one file with a targeted inclusion list (stepped
CE 10/20/40 eV per precursor), so co-elution-based pairing is impossible.
Cembio features are formed by **`rle(precursorMz)` grouping**: each
contiguous run of MS2 at the same precursor m/z (in acquisition order)
becomes one feature. For stepped-CE acquisition this naturally yields
one feature per 3-CE cycle (clean per-CE fragmentation, ideal SIRIUS
input). A previous RT-proximity merge (`cembio_block_gap_s = 30`)
collapsed all schedule fires of the same precursor into one feature
with mixed CE levels; this hurt SIRIUS identification (~87% → 73%
spike-in coverage) because the merged spectrum smears clean per-CE
ladders. We restore the shattered-per-cycle approach and handle the
resulting multi-feature-per-peak issue **downstream**: an MS1-anchoring
step in `5_downstream_analysis/3_rt_alignment.qmd` §P2.9 rewrites every
lab's feature `rt_sec` (cembio + DDA) to the nearest MS1 chromatographic
peak at the precursor m/z. For cembio this corrects the ~18 s schedule
offset; for DDA labs it corrects coincidental SIRIUS hits on in-source
matrix fragments triggered far from the real peak (see §P2.9.1 case
study).

**Noise filter.** After loading we apply a per-MS2 intensity filter:
keep peaks with intensity ≥ 5% of the base peak OR ≥ 1000 counts. This
cuts MS2 peak counts ~4× on the noisy late-RT hmgu batches (median 91
→ 21 peaks/MS2) while leaving normal spectra essentially untouched. The
OR floor rescues mid-intensity fragments when the base peak is unusually
tall.

```{r}
#| label: helper-extract-all
extract_all_ms2 <- function(lab_name, study_group, mixture) {
  study_path <- here("1_preprocessing", lab_name, study_group)
  mzml_dir   <- file.path(study_path, "mzml")
  mix_label  <- mix_us(mixture)

  empty_result <- function(sp = Spectra(backend = MsBackendMemory())) {
    list(spectra_pos = sp, spectra_neg = sp,
         n_ms2_pos = 0L, n_ms2_neg = 0L,
         n_features_pos = 0L, n_features_neg = 0L)
  }

  seq_pos <- read_xlsx(file.path(study_path, "seq_pos.xlsx")) |> as.data.frame()
  seq_neg <- read_xlsx(file.path(study_path, "seq_neg.xlsx")) |> as.data.frame()
  seq_pos$polarity <- "pos"; seq_neg$polarity <- "neg"
  seq <- rbind(seq_pos, seq_neg)
  seq$filename <- paste0(seq$`Data File`, ".mzML")
  seq <- seq[seq$filename %in% list.files(mzml_dir), ]
  # Mixture is the last underscore-delimited token of the file basename,
  # after stripping a trailing `_MS2` and any `target` suffix (used by
  # cembio for targeted-MS2 acquisitions, e.g. "..._2.1target.mzML").
  seq$mixture <- gsub("\\.", "_",
                      sub("target$", "",
                          sub(".*_", "",
                              sub("_MS2$", "", seq$`Data File`))))

  has_ms2_files <- any(grepl("MS2", seq$filename, fixed = TRUE))
  seq <- if (has_ms2_files) {
    seq[grepl("MS2", seq$filename, fixed = TRUE) & seq$mixture == mix_label, ]
  } else {
    seq[seq$mixture == mix_label, ]
  }
  if (nrow(seq) == 0L) return(empty_result())

  sp <- Spectra(file.path(mzml_dir, seq$filename), backend = MsBackendMzR())
  sp <- setBackend(sp, MsBackendMemory())
  sp <- applyProcessing(sp)

  sp <- filterIntensity(sp, intensity = function(x) {
    if (length(x) == 0L) return(logical(0))
    bp <- max(x, na.rm = TRUE)
    if (!is.finite(bp) || bp <= 0) return(rep(TRUE, length(x)))
    x >= bp * 0.05 | x >= 1000
  }, msLevel. = 2L)

  if (lab_name == "cembio") {
    sp <- sp[sp$msLevel == 2L]
    sp <- filterEmptySpectra(sp)
    if (length(sp) == 0L) return(empty_result(sp))

    # `rle(precursorMz)` grouping for cembio's block-MS2 acquisition.
    #
    # The inclusion-list cycles through precursors with stepped CE
    # (10/20/40 eV). In acquisition order, this produces runs of 3
    # consecutive MS2 at the same precursor m/z (one per CE), then the
    # next precursor, etc. We group each contiguous run as one feature
    # → one feature per 3-CE cycle, with clean per-CE fragmentation
    # ideal for SIRIUS.
    #
    # Multiple cycles of the same precursor (the schedule cycles back
    # every ~15-30 s) produce MULTIPLE features at different RTs along
    # the gradient. The MS2-trigger RT of each feature is the
    # schedule-fire time, which is NOT the chromatographic peak RT.
    # The downstream MS1-anchoring step in 3_rt_alignment.qmd §P2.9
    # rewrites every lab's rt_sec to the nearest MS1 peak (uniform
    # across cembio + DDA labs, since all separate-injection labs
    # have some offset between MS2-trigger RT and MS1 peak RT).
    ord       <- order(sp$dataOrigin, sp$rtime)
    key_s     <- paste0(sp$dataOrigin[ord], "|",
                        sprintf("%.4f", sp$precursorMz[ord]))

    starts_new_group <- c(TRUE,
                          key_s[-1L] != key_s[-length(key_s)])
    grp_sorted <- cumsum(starts_new_group)

    grp <- integer(length(sp))
    grp[ord] <- grp_sorted
    sp$group_idx <- grp
  } else {
    keep <- unlist(lapply(split(seq_along(sp), sp$dataOrigin), function(idx) {
      first_ms1 <- which(sp$msLevel[idx] == 1L)[1L]
      if (is.na(first_ms1)) integer(0) else idx[first_ms1:length(idx)]
    }))
    sp <- sp[sort(keep)]
    sp <- filterEmptySpectra(sp)
    if (length(sp) == 0L) return(empty_result(sp))
    sp$group_idx <- fragmentGroupIndex(sp)
    sp <- sp[duplicated(sp$group_idx) | duplicated(sp$group_idx, fromLast = TRUE)]
    if (length(sp) == 0L) return(empty_result(sp))
  }

  list(
    spectra_pos    = sp[sp$polarity == 1L],
    spectra_neg    = sp[sp$polarity == 0L],
    n_ms2_pos      = sum(sp$msLevel == 2L & sp$polarity == 1L),
    n_ms2_neg      = sum(sp$msLevel == 2L & sp$polarity == 0L),
    n_features_pos = length(unique(sp$group_idx[sp$polarity == 1L])),
    n_features_neg = length(unique(sp$group_idx[sp$polarity == 0L]))
  )
}
```

## 2. SIRIUS helpers

Path helpers for the per-(lab, mixture) cache files, and a single list of
ID columns that SIRIUS returns inconsistently typed (sometimes numeric,
sometimes character). We coerce all of them to character on every read /
write so `bind_rows` across batches and mixtures never fails on type
mismatch. Add new column names to `sirius_id_cols` if a future error
surfaces one.

```{r}
#| label: helpers
sirius_sum_path       <- function(lab, m) file.path(output_dir, sprintf("sirius_summary_%s_mix_%s.csv",  lab, mix_us(m)))
sirius_full_path      <- function(lab, m) file.path(output_dir, sprintf("sirius_full_%s_mix_%s.csv",     lab, mix_us(m)))
extraction_count_path <- function(lab, m) file.path(output_dir, sprintf("extraction_counts_%s_mix_%s.csv", lab, mix_us(m)))

sirius_id_cols <- c("alignedFeatureId", "compoundId", "externalFeatureId",
                    "formulaId", "xcms_fts", "sirius_fts",
                    "molecularFormulaId", "structureCandidateInChIKey")
```

```{r}
#| label: sirius-login
sirius_user <- "plouail@eurac.edu"
sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")
```

### `run_sirius_for_lab()`

Runs SIRIUS over one (lab, mixture)'s spectra in batches of `batch_size`
features, caching per-batch results to CSV so re-renders skip completed
batches. The function manages the SIRIUS server lifecycle internally and
**restarts the JVM every `restart_every` live batches** - a long-running
JVM accumulates state that eventually crashes `import()`, so bouncing the
process clears it (the on-disk project survives the restart).

Each batch also cross-checks `featuresId(srs, type = "xcms")` against the
features intended for import and re-imports anything silently dropped
(up to two retries; bail when retries stop making progress).

Pos and neg features go into the same project per batch so RuSirius
assigns polarity correctly. `numberOfCandidates = 10`/`3` (per ionization)
is a deliberate cut from the default 15/5 - each candidate triggers a
CSI:FingerID + BIO lookup, the dominant per-feature cost.

```{r}
#| label: run-sirius-for-lab
run_sirius_for_lab <- function(sp_pos, sp_neg,
                               username, password, projectId, path,
                               batch_size    = 200,
                               cache_dir     = NULL,
                               cache_prefix  = NULL,
                               restart_every = 7L) {
  if (length(sp_pos) == 0L && length(sp_neg) == 0L)
    return(list(summary = NULL, full = NULL))

  start_sirius <- function() Sirius(username = username, password = password,
                                    projectId = projectId, path = path)
  srs <- start_sirius()
  on.exit(try(shutdown(srs), silent = TRUE), add = TRUE)

  groups_pos <- unique(sp_pos$group_idx)
  groups_neg <- unique(sp_neg$group_idx)
  pol_lookup <- c(
    setNames(rep("pos", length(groups_pos)), as.character(groups_pos)),
    setNames(rep("neg", length(groups_neg)), as.character(groups_neg))
  )

  all_groups <- c(groups_pos, groups_neg)
  batches    <- split(all_groups, ceiling(seq_along(all_groups) / batch_size))
  has_cache  <- !is.null(cache_dir) && !is.null(cache_prefix)

  sums         <- vector("list", length(batches))
  fulls        <- vector("list", length(batches))
  live_batches <- 0L

  for (i in seq_along(batches)) {
    batch_sum_path  <- if (has_cache) file.path(cache_dir, sprintf("%s_batch%d_sum.csv",  cache_prefix, i))
    batch_full_path <- if (has_cache) file.path(cache_dir, sprintf("%s_batch%d_full.csv", cache_prefix, i))

    message("  batch ", i, "/", length(batches))

    if (has_cache && file.exists(batch_sum_path) && file.exists(batch_full_path)) {
      message("    → cached, skipping")
      sums[[i]]  <- read.csv(batch_sum_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
               across(any_of(sirius_id_cols), as.character))
      fulls[[i]] <- read.csv(batch_full_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(sirius_id_cols), as.character))
      next
    }

    if (live_batches > 0L && live_batches %% restart_every == 0L) {
      message("    → restarting SIRIUS after ", live_batches, " live batches")
      try(shutdown(srs), silent = TRUE)
      srs <- start_sirius()
    }

    b_pos <- sp_pos[sp_pos$group_idx %in% batches[[i]]]
    b_neg <- sp_neg[sp_neg$group_idx %in% batches[[i]]]
    n_pos <- length(unique(b_pos$group_idx))
    n_neg <- length(unique(b_neg$group_idx))
    message("    ", n_pos, " pos + ", n_neg, " neg features")

    if (n_pos > 0)
      srs <- import(srs, spectra = b_pos, ms_column_name = "group_idx",
                    deleteExistingFeatures = TRUE)
    if (n_neg > 0)
      srs <- import(srs, spectra = b_neg, ms_column_name = "group_idx",
                    deleteExistingFeatures = n_pos == 0)

    expected_ids <- as.character(c(unique(b_pos$group_idx),
                                   unique(b_neg$group_idx)))
    project_ids  <- as.character(featuresId(srs, type = "xcms"))
    missing_ids  <- setdiff(expected_ids, project_ids)
    for (retry in 1:2) {
      if (length(missing_ids) == 0L) break
      message("    !! ", length(missing_ids), " missing - retry ", retry)
      miss_pos <- b_pos[b_pos$group_idx %in% missing_ids]
      miss_neg <- b_neg[b_neg$group_idx %in% missing_ids]
      if (length(miss_pos) > 0L)
        srs <- import(srs, spectra = miss_pos, ms_column_name = "group_idx",
                      deleteExistingFeatures = FALSE)
      if (length(miss_neg) > 0L)
        srs <- import(srs, spectra = miss_neg, ms_column_name = "group_idx",
                      deleteExistingFeatures = FALSE)
      prev_n      <- length(missing_ids)
      missing_ids <- setdiff(expected_ids,
                             as.character(featuresId(srs, type = "xcms")))
      if (length(missing_ids) >= prev_n) {
        message("    !! retry made no progress - giving up")
        break
      }
    }
    if (length(missing_ids) > 0L)
      warning("    !! ", length(missing_ids), " feature(s) never imported: ",
              paste(head(missing_ids, 10), collapse = ","))

    run(srs,
      formulaIdParams = formulaIdParam(
        numberOfCandidates              = 10,
        instrument                      = "QTOF",
        numberOfCandidatesPerIonization = 3,
        massAccuracyMS2ppm              = 10,
        filterByIsotopePattern          = FALSE,
        isotopeMs2Settings              = "SCORE",
        performDeNovoBelowMz            = 600),
      predictParams           = predictParam(),
      structureDbSearchParams = structureDbSearchParam(structureSearchDbs = c("BIO")),
      recompute = TRUE, wait = TRUE)

    sums[[i]] <- summary(srs, result.type = "structure") |>
      mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
             across(any_of(sirius_id_cols), as.character),
             polarity = pol_lookup[as.character(externalFeatureId)])

    fulls[[i]] <- results(srs, result.type = "structureDb",
                          topFormula = 8, topStructure = 20,
                          return.type = "data.frame") |>
      mutate(across(any_of(sirius_id_cols), as.character),
             polarity = pol_lookup[as.character(xcms_fts)])

    if (has_cache) {
      write.csv(sums[[i]],  batch_sum_path,  row.names = FALSE)
      write.csv(fulls[[i]], batch_full_path, row.names = FALSE)
    }
    live_batches <- live_batches + 1L
  }

  non_empty <- function(x) Filter(\(df) nrow(df) > 0L, x)
  list(summary = bind_rows(non_empty(sums)),
       full    = bind_rows(non_empty(fulls)))
}
```

## 3. Extract + run SIRIUS per (mixture, lab)

Each (mixture, lab) is processed end-to-end before moving to the next:
extract → run SIRIUS → write cache → drop spectra. Memory stays bounded
to one lab's spectra at a time, and a partial run resumes simply by
re-rendering - cached pairs skip both extraction and SIRIUS.

```{r}
#| label: extract-and-sirius
summary_results    <- setNames(vector("list", length(mixtures)), mixtures)
full_results       <- setNames(vector("list", length(mixtures)), mixtures)
extraction_summary <- list()

for (m in mixtures) {
  summary_results[[m]] <- list()
  full_results[[m]]    <- list()
  for (lab in labs) {
    sum_path  <- sirius_sum_path(lab, m)
    full_path <- sirius_full_path(lab, m)

    if (file.exists(sum_path) && file.exists(full_path)) {
      message("Cached: ", lab, " / mix ", m)
      summary_results[[m]][[lab]] <- read.csv(sum_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c("confidenceExactMatch", "confidenceApproxMatch")), as.numeric),
               across(any_of(c(sirius_id_cols, "lab", "polarity", "adduct")), as.character))
      full_results[[m]][[lab]] <- read.csv(full_path, stringsAsFactors = FALSE) |>
        mutate(across(any_of(c(sirius_id_cols, "lab", "polarity", "adduct")), as.character))
      if (file.exists(extraction_count_path(lab, m)))
        extraction_summary[[paste(m, lab)]] <-
          read.csv(extraction_count_path(lab, m),
                   stringsAsFactors = FALSE,
                   colClasses = c(mixture = "character",
                                  lab     = "character")) |>
          as_tibble()
      next
    }

    message("Extracting: ", lab, " / mix ", m)
    sp <- extract_all_ms2(lab, study_group, m)
    extraction_summary[[paste(m, lab)]] <- tibble(
      mixture = m, lab = lab,
      n_features_pos = sp$n_features_pos, n_features_neg = sp$n_features_neg,
      n_ms2_pos      = sp$n_ms2_pos,      n_ms2_neg      = sp$n_ms2_neg)
    write.csv(extraction_summary[[paste(m, lab)]],
              extraction_count_path(lab, m), row.names = FALSE)

    if (length(sp$spectra_pos) == 0L && length(sp$spectra_neg) == 0L) {
      message("  no MS2 - writing empty cache")
      summary_results[[m]][[lab]] <- tibble(lab = character(0))
      full_results[[m]][[lab]]    <- tibble(lab = character(0))
      write.csv(summary_results[[m]][[lab]], sum_path,  row.names = FALSE)
      write.csv(full_results[[m]][[lab]],    full_path, row.names = FALSE)
      rm(sp); gc(); next
    }

    if (!exists("sirius_user")) sirius_user <- Sys.getenv("SIRIUS_USERNAME")
    if (!exists("sirius_pass")) sirius_pass <- Sys.getenv("SIRIUS_PASSWORD")

    res <- run_sirius_for_lab(
      sp_pos       = sp$spectra_pos,
      sp_neg       = sp$spectra_neg,
      username     = sirius_user, password = sirius_pass,
      projectId    = paste0("ring_trial_allms2_", lab, "_mix_", mix_us(m)),
      path         = output_dir,
      cache_dir    = output_dir,
      cache_prefix = paste0(lab, "_mix_", mix_us(m)))

    summary_results[[m]][[lab]] <-
      if (is.null(res$summary)) tibble(lab = character(0))
      else res$summary |> mutate(lab = lab)
    full_results[[m]][[lab]] <-
      if (is.null(res$full)) tibble(lab = character(0))
      else res$full |> mutate(lab = lab)

    write.csv(summary_results[[m]][[lab]], sum_path,  row.names = FALSE)
    write.csv(full_results[[m]][[lab]],    full_path, row.names = FALSE)
    rm(sp, res); gc()
  }
}

bind_rows(extraction_summary)
```

Extraction-summary bar plot per lab × polarity × mixture.

```{r}
#| label: plot-extraction
#| fig-cap: "SIRIUS features per lab × polarity × mixture (MS1+MS2 group; MS2 only for cembio)"
if (length(extraction_summary) > 0) {
  bind_rows(extraction_summary) |>
    select(lab, mixture, pos = n_features_pos, neg = n_features_neg) |>
    pivot_longer(c(pos, neg), names_to = "polarity", values_to = "n") |>
    ggplot(aes(x = lab, y = n, fill = polarity)) +
    geom_col(position = "dodge") +
    facet_wrap(~ mixture) +
    scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
    labs(x = NULL, y = "SIRIUS features", fill = "Polarity") +
    theme_bw()
}
```

Aggregate to long tables (`all_summaries`, `all_full`) keyed by mixture
for everything downstream.

```{r}
#| label: aggregate
# Skip 0-row tibbles inside each mixture's per-lab list before bind_rows: an
# empty cache from a (lab, mixture) with no MS2 features writes a header-only
# CSV, which read.csv parses with logical-typed columns. Combining those with
# real character/numeric columns would crash bind_rows.
agg <- function(results_list) {
  bind_rows(lapply(mixtures, function(m) {
    non_empty <- Filter(\(df) nrow(df) > 0L, results_list[[m]])
    if (length(non_empty) == 0L) return(NULL)
    bind_rows(non_empty) |> mutate(mixture = m)
  }))
}
all_summaries <- agg(summary_results)
all_full      <- agg(full_results)

write.csv(all_summaries, file.path(output_dir, "sirius_summary_all_labs_all_mixtures.csv"), row.names = FALSE)
write.csv(all_full,      file.path(output_dir, "sirius_full_all_labs_all_mixtures.csv"),    row.names = FALSE)
```

### Feature-import verification

`run_sirius_for_lab()` already verifies each batch in-flight. This is the
end-to-end check: per (lab, mixture, polarity) we compare features
**extracted** upstream vs features SIRIUS produced **any** result for
(structure summary ∪ structureDb result). A residual `missing > 0` row
means SIRIUS imported the feature but produced no annotation - or, if a
warning fired during the loop, the retry import never succeeded.

```{r}
#| label: verify-import
#| include: false
extracted_long <- bind_rows(extraction_summary) |>
  select(mixture, lab, pos = n_features_pos, neg = n_features_neg) |>
  pivot_longer(c(pos, neg), names_to = "polarity", values_to = "extracted")

annotated_long <- bind_rows(
  all_summaries |> filter(!is.na(externalFeatureId)) |>
    transmute(mixture, lab, polarity, feature_id = externalFeatureId),
  all_full      |> filter(!is.na(xcms_fts)) |>
    transmute(mixture, lab, polarity, feature_id = xcms_fts)
) |>
  group_by(mixture, lab, polarity) |>
  summarise(annotated = n_distinct(feature_id), .groups = "drop")

verify_df <- extracted_long |>
  left_join(annotated_long, by = c("mixture", "lab", "polarity")) |>
  mutate(annotated = coalesce(annotated, 0L),
         missing   = extracted - annotated,
         pct_kept  = ifelse(extracted == 0, NA_real_,
                            round(100 * annotated / extracted, 1)))

write.csv(verify_df,
          file.path(output_dir, "feature_import_verification.csv"),
          row.names = FALSE)

verify_df |> arrange(desc(missing)) |> knitr::kable()

cat("Entries with missing > 0:", sum(verify_df$missing > 0, na.rm = TRUE),
    "  |  total extracted:",  sum(verify_df$extracted, na.rm = TRUE),
    "  annotated:",            sum(verify_df$annotated, na.rm = TRUE),
    "  missing:",              sum(verify_df$missing,   na.rm = TRUE), "\n")
```

## 4. Match to expected compounds (all mixtures)

```{r}
#| label: expected
expected_all <- meta |>
  filter(Mixture %in% mixtures) |>
  transmute(mixture              = Mixture,
            compound             = `ChEBI name`,
            expected_formula     = formula,
            expected_inchikey    = InChIKey,
            expected_inchikey_2d = substr(InChIKey, 1, 14))

expected_all |>
  count(mixture, name = "n_expected") |>
  knitr::kable(caption = "Expected compounds per mixture")
```

### Ambiguous compounds in the expected list

Two patterns are checked here before the coverage analysis.

**(a) Within-mixture stereoisomers (same 2D InChIKey).** SIRIUS returns
only the 2D InChIKey, so two compounds sharing those 14 characters in
the same mixture cannot be told apart from the annotation. We **drop
both compounds** from `expected_all` rather than over-counting one of
them.

**(b) Same compound listed in multiple mixtures.** Informational only -
the join below keys on `mixture`, so each annotation is attributed to
its own mixture and these duplicates don't double count.

```{r}
#| label: expected-stereoisomers
exp_dups_2d <- expected_all |>
  group_by(mixture, expected_inchikey_2d) |>
  filter(n() > 1) |>
  arrange(mixture, expected_inchikey_2d, compound) |>
  ungroup()

knitr::kable(exp_dups_2d,
  caption = "(a) Stereoisomer pairs - DROPPED from expected_all")

expected_all <- expected_all |>
  anti_join(exp_dups_2d, by = c("mixture", "expected_inchikey_2d"))
```

```{r}
#| label: expected-cross-mixture
exp_dups_mix <- expected_all |>
  group_by(expected_inchikey) |>
  filter(n_distinct(mixture) > 1) |>
  arrange(expected_inchikey, mixture) |>
  ungroup()

knitr::kable(exp_dups_mix,
  caption = "(b) Same compound across multiple mixtures (kept)")

all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  filter(!is.na(inchiKey)) |>
  semi_join(exp_dups_mix |> distinct(mixture, expected_inchikey_2d),
            by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  count(mixture, lab, polarity, inchikey_2d) |>
  knitr::kable(caption = "Were the cross-mixture duplicates annotated?")
```

A SIRIUS hit counts as recovery of an expected compound when its 2D
InChIKey matches the expected one **for the same mixture**. Per
(mixture, lab, compound, polarity) we record the best formula rank, best
structure rank and the union of adducts seen.

```{r}
#| label: match-structure
structure_hits_all <- all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  inner_join(
    expected_all |>
      select(mixture, expected_inchikey_2d, compound,
             expected_formula, expected_inchikey),
    by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  group_by(mixture, lab, inchikey_2d, polarity) |>
  summarise(
    best_formula_rank   = suppressWarnings(min(rank.x, na.rm = TRUE)),
    best_structure_rank = suppressWarnings(min(rank.y, na.rm = TRUE)),
    adducts             = paste(sort(unique(na.omit(adduct))), collapse = ";"),
    molecularFormula    = first(molecularFormula),
    compound            = first(compound),
    expected_formula    = first(expected_formula),
    expected_inchikey   = first(expected_inchikey),
    .groups = "drop") |>
  mutate(across(c(best_formula_rank, best_structure_rank),
                \(x) ifelse(is.infinite(x), NA_real_, x)))

structure_hits_all
```

## 5. Coverage

```{r}
#| label: coverage
covered_all <- expected_all |>
  left_join(structure_hits_all |>
              distinct(mixture, inchikey_2d) |>
              mutate(structure_recovered = TRUE),
            by = c("mixture", "expected_inchikey_2d" = "inchikey_2d")) |>
  mutate(structure_recovered = !is.na(structure_recovered))

covered_all |>
  group_by(mixture) |>
  summarise(n_recovered = sum(structure_recovered),
            n_expected  = n(), .groups = "drop") |>
  knitr::kable(caption = "Structure recovery per mixture")
```

```{r}
#| label: plot-coverage-by-lab
#| fig-cap: "Structure hits per lab × polarity, faceted by mixture"
#| fig-height: 8
structure_hits_all |>
  count(mixture, lab, polarity) |>
  ggplot(aes(x = lab, y = n, fill = polarity)) +
  geom_col(position = "dodge") +
  facet_wrap(~ mixture) +
  scale_fill_manual(values = c(pos = "#4e79a7", neg = "#f28e2b")) +
  labs(x = NULL, y = "Structure hits", fill = "Polarity") +
  theme_bw()
```

## 6. RT extraction for EIC pipeline

RT per feature is extracted from the MS2 spectra and cached per (lab,
mixture) so raw files are only read once. `hit_rt` joins the RT onto
expected-compound matches and is the input to §7.1's `eic_params`. No
cross-lab RT analysis here — that lives in `8_per_compound_diagnostics.qmd`.

```{r}
#| label: extract-rt
#| message: false
rt_all <- bind_rows(lapply(mixtures, function(m) {
  bind_rows(lapply(labs, function(lab) {
    path <- file.path(output_dir, paste0(lab, "_mix_", mix_us(m), "_rt.csv"))
    if (file.exists(path)) {
      df <- read.csv(path, stringsAsFactors = FALSE)
      if ("rt_min" %in% names(df) && !"rt_sec" %in% names(df))
        df <- mutate(df, rt_sec = rt_min * 60)
      return(df |> mutate(lab = lab, mixture = m))
    }
    message("Extracting RT for: ", lab, " / mix ", m)
    res <- extract_all_ms2(lab, study_group, m)
    make_rt_tbl <- function(sp) {
      if (length(sp) == 0L)
        tibble(group_idx = integer(), msLevel = integer(), rt_sec = numeric())
      else
        tibble(group_idx = sp$group_idx, msLevel = sp$msLevel, rt_sec = sp$rtime)
    }
    rt_df <- bind_rows(make_rt_tbl(res$spectra_pos),
                       make_rt_tbl(res$spectra_neg)) |>
      filter(msLevel == 2L) |>
      group_by(group_idx) |>
      summarise(rt_sec = median(rt_sec, na.rm = TRUE), .groups = "drop") |>
      mutate(lab = lab, mixture = m)
    write.csv(rt_df, path, row.names = FALSE)
    rt_df
  }))
}))
```

```{r}
#| label: rt-expected-compounds
hit_rt <- all_full |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14),
         xcms_fts    = as.character(xcms_fts)) |>
  filter(!is.na(inchiKey)) |>
  inner_join(
    expected_all |> select(mixture, expected_inchikey_2d, compound,
                           expected_formula, expected_inchikey),
    by = c("mixture", "inchikey_2d" = "expected_inchikey_2d")) |>
  left_join(rt_all |> mutate(group_idx = as.character(group_idx)),
            by = c("lab", "mixture", "xcms_fts" = "group_idx")) |>
  filter(!is.na(rt_sec))

cat("hit_rt rows (per-feature RT for expected-compound matches):",
    nrow(hit_rt), "\n")
```

## 7. EICs for annotated features - `chromExtract()` workflow

For every expected compound we want one MS1 EIC per lab, **including labs
that did not annotate it**, so we can visually check whether the signal
exists but MS2 missed it. One flat peak table → one `chromExtract()` call.

Two row types in the peak table:

* `source = "own"` - the lab annotated this compound via SIRIUS; use its
  ionMass + RT.
* `source = "consensus"` - the lab did not annotate it; use the cross-lab
  median ionMass + median raw RT.

For the MS1 file paths: afekta/cembio/icl have dedicated MS1-only files
(no `_MS2` in name) which give cleaner survey traces than the MS2 file's
interleaved MS1. HMGU acquires everything in one file.

```{r}
#| label: ms1-files-per-lab
ms1_files_for_lab <- function(lab_name, study_group = "HE", mixture) {
  study_path <- here("1_preprocessing", lab_name, study_group)
  mzml_dir   <- file.path(study_path, "mzml")
  seq_pos <- read_xlsx(file.path(study_path, "seq_pos.xlsx")) |> as.data.frame()
  seq_neg <- read_xlsx(file.path(study_path, "seq_neg.xlsx")) |> as.data.frame()
  seq_pos$polarity <- "pos"; seq_neg$polarity <- "neg"
  seq <- rbind(seq_pos, seq_neg)
  seq$filename <- paste0(seq$`Data File`, ".mzML")
  seq <- seq[seq$filename %in% list.files(mzml_dir), ]
  seq$mixture <- gsub("\\.", "_",
                      sub("target$", "",
                          sub(".*_", "",
                              sub("_MS2$", "", seq$`Data File`))))
  is_ms2   <- grepl("MS2", seq$`Data File`, fixed = TRUE)
  separate <- any(is_ms2)
  seq <- if (separate) seq[!is_ms2 & seq$mixture == mix_us(mixture), ]
         else          seq[          seq$mixture == mix_us(mixture), ]
  tibble(lab = lab_name, mixture = mixture, polarity = seq$polarity,
         path = file.path(mzml_dir, seq$filename),
         separate_ms1_ms2 = separate)
}

lab_files <- bind_rows(lapply(mixtures, function(m)
  bind_rows(lapply(labs, ms1_files_for_lab, study_group = study_group, mixture = m))))
```

### 7.1 Build the flat peak table

EICs are extracted with a generous **±60 s** symmetric window around the
annotation RT. The wide window serves two purposes:

1. **Noise context.** §7.4's SNR-aware peak picker estimates baseline +
   noise from the EIC tails (|rt − apex| > 25 s); we need real tail
   region to compute MAD / IQR of intensity.
2. **Tolerance to cross-lab RT drift.** ±60 s easily covers cembio ↔
   icl gradient differences and early-RT instabilities, so the single
   pass below handles what the legacy two-pass approach handled.

The ±20 s pass-1 of the legacy pipeline (and its ±80 s pass-2 retry)
have been removed — they were a workaround for `peakBoundary()`'s habit
of locking onto the tallest local maximum even when no real peak is
present (see §7.4 rationale).

The current build also addresses three MS2-hit-level issues identified
in `8_per_compound_diagnostics.qmd`:

* **All adducts get their own EIC.** The legacy code did `adduct =
  first(adduct)` per `(mixture, lab, polarity, compound)`, silently
  dropping every adduct beyond the first. We now group by adduct as
  well, so `[M+H]+`, `[M+Na]+`, `[M+K]+`, `[M-H+formic]-` etc. each
  emit their own row in the canonical CSV.
* **All features get a vote (no median collapse).** The legacy code
  did `rt_sec = median(rt_sec)` per `(mixture, lab, polarity, compound,
  adduct)`. For cembio's block-MS2 mode, `extract_all_ms2()`'s
  `rle(precursorMz)` deliberately shatters each schedule fire into its
  own feature (one feature per 3-CE cycle) — clean per-CE input for
  SIRIUS. The downstream MS1-anchoring step in
  `5_downstream_analysis/3_rt_alignment.qmd` §P2.9 rewrites every
  lab's `rt_sec` to the nearest MS1 chromatographic peak; the cluster
  algorithm in §P2.10 then sees real RTs and picks the correct peak
  per compound.
* **Cluster-based consensus RT** combines cross-lab, cross-polarity,
  and cross-feature evidence to pick the "true" RT for every
  `(mixture, compound)`. The algorithm:

  1. Pool all candidate RTs from every `(lab, polarity, adduct, feature)`
     tuple that SIRIUS annotated for this compound.
  2. **Within-lab Stage 1** (`ms1_anchor.R::find_ms1_candidates`):
     extract the MS1 EIC at the precursor *m/z* per row, run
     `find_all_peaks` (SNR ≥ 5), and validate each peak with the same
     six guards Stage 3 uses (per-lab intensity floor, peak shape ≥ 2
     scans, real-scan support, signal concentration ≥ 0.45). Keep
     peaks ≥ 50 % of the surviving max as candidates.
  3. **Cross-lab Stage 2** (`ms1_anchor.R::anchor_rts`): forward-align
     every candidate to the consensus axis via the NAPS splines, then
     apply a ±8 s **similarity rule** — for each candidate RT count
     distinct `(lab, polarity)` within ±8 s; among the max-support
     candidates take the median of their aligned RTs as the centre;
     consensus RT = median of all candidates within ±8 s of that
     centre. Score the supporting set as
     `n_labs × 10 + n_polarities × 5 + sum_cosmic`.
  4. **Cross-compound isobaric resolution** runs in the same Stage 2:
     per mixture, pairs of compounds with any-adduct *m/z* within
     20 ppm AND consensus RT within 5 s collapse — keep the
     higher-scoring compound, drop the loser.
  5. **Snap + inverse-align** (Stage 2, end): the consensus RT is
     inverse-aligned to each lab's local time, and each lab's row
     snaps to the candidate nearest that local target. Labs whose own
     candidate survived get `source = "own"`; labs that never
     annotated this compound get `source = "consensus"` (the EIC will
     be extracted at the inverse-aligned consensus RT in Stage 3).

  This naturally fixes the cross-polarity disagreement, the cembio
  feature-shattering, and the wrong-peak DDA-trigger problem in one
  pass.

```{r}
#| label: eic-candidates
#| message: false
# --- eic_candidates: one row per FEATURE per (mixture, lab, polarity,
# compound, adduct). No collapse across features within (lab, polarity,
# compound, adduct). This is the raw candidate pool that the
# cluster-based consensus algorithm in 3_rt_alignment.qmd operates on.
#
# Note: a single (lab, polarity, compound, adduct) tuple can have many
# feature rows here — e.g. cembio mix 2_9 ritalinic acid pos [M+H]+
# has multiple features (one per inclusion-list MS2 trigger) before the
# Stage 2 similarity rule snaps them to one consensus RT. Keeping all
# of them lets
# cross-lab + cross-polarity clustering decide whether they're separate
# peaks or one peak that was shattered by the inclusion-list schedule.
eic_candidates <- hit_rt |>
  select(mixture, lab, polarity, xcms_fts, compound, inchikey_2d, rt_sec) |>
  distinct() |>
  left_join(all_summaries |>
              select(mixture, lab, externalFeatureId, ionMass, adduct,
                     confidenceExactMatch) |>
              mutate(externalFeatureId = as.character(externalFeatureId)),
            by = c("mixture", "lab", "xcms_fts" = "externalFeatureId")) |>
  filter(!is.na(ionMass), !is.na(rt_sec)) |>
  rename(cosmic = confidenceExactMatch)
```

### 7.2 Write feature-RT table for downstream consumption

The per-feature MS2-trigger RTs in `eic_candidates` are the input to
the alignment-based cluster algorithm in `5_downstream_analysis/3_rt_alignment.qmd`
(`§P2.9` onwards). From there the EIC extraction + SNR-aware picker
writes `5_downstream_analysis/object/annotation_chrom_metrics.csv`.

This producer no longer does EIC extraction — the SIRIUS run + the
expected-compound matching above are its only responsibilities. All
downstream chromatography work — alignment, cluster RT, EIC, picker,
metrics — lives in `3_rt_alignment.qmd`.

```{r}
#| label: write-feat-rts
feat_rts <- eic_candidates |>
  select(mixture, lab, polarity, xcms_fts, compound, inchikey_2d,
         rt_sec, ionMass, adduct, cosmic)

write.csv(feat_rts,
          here::here("5_downstream_analysis", "object", "feat_rts.csv"),
          row.names = FALSE)
cat(sprintf("feat_rts.csv -- %d rows (per-feature MS2-trigger RTs)\n",
            nrow(feat_rts)))
```

## 8. Write canonical CSVs

This producer writes the **identity table** to
`5_downstream_analysis/object/annotation_identity.csv`. The
chromatography metrics (`annotation_chrom_metrics.csv`) are produced
downstream in `3_rt_alignment.qmd`.

### 8.1 `annotation_identity.csv`

`structure_hits_all` is augmented with the COSMIC scores and the
observed `ionMass` from `all_summaries`, plus a placeholder
`confidence_tier = "unfiltered"`. The join picks the best-`csiScore`
candidate per `(mixture, lab, inchikey_2d, polarity)` — identity is
one-per-compound, so we collapse the candidate dimension first; the
resulting `ionMass` is the precursor m/z of that winning candidate and
is exposed downstream as `mz_observed` (basis for the inter-lab m/z
deviation analysis in `8_per_compound_diagnostics.qmd`).

```{r}
#| label: build-annotation-identity
id_summary <- all_summaries |>
  filter(!is.na(inchiKey)) |>
  mutate(inchikey_2d = substr(inchiKey, 1, 14)) |>
  group_by(mixture, lab, inchikey_2d, polarity) |>
  slice_max(csiScore, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(mixture, lab, inchikey_2d, polarity,
         confidenceExactMatch, confidenceApproxMatch,
         mz_observed = ionMass)

annotation_identity <- structure_hits_all |>
  left_join(id_summary,
            by = c("mixture", "lab", "inchikey_2d", "polarity")) |>
  mutate(confidence_tier = "unfiltered") |>
  select(mixture, lab, compound, inchikey_2d, polarity,
         best_formula_rank, best_structure_rank, adducts,
         molecularFormula, expected_formula, expected_inchikey,
         mz_observed, confidenceExactMatch, confidenceApproxMatch,
         confidence_tier)

write.csv(annotation_identity,
          here::here("5_downstream_analysis", "object",
                     "annotation_identity.csv"),
          row.names = FALSE)
cat("annotation_identity.csv  -- ", nrow(annotation_identity), " rows\n",
    "  with COSMIC scores: ",
    sum(!is.na(annotation_identity$confidenceExactMatch)),
    "  with mz_observed: ",
    sum(!is.na(annotation_identity$mz_observed)), "\n", sep = "")
```

