---
title: "Global Analysis: Inter-Laboratory LC-MS Reproducibility"
format: 
    html:
        toc: true
        toc-location: right
        code-fold: true
---

# Introduction

This analysis evaluates QC metrics across 4 laboratories using positive
and negative ionization modes. Instrumental setup, sample preparation, and
data processing all affect chromatographic metrics on both full and
detected signal data.

Data: 19 standard mixtures (HUMAN endosome, different compounds) from 4
labs with harmonized LC systems.

Limitations: 

- No technical replicates
- Laboratory identity is confounded with instrumental parameters.

=> Observed differences reflect combined effects
of lab environment, operator expertise, and equipment configuration.

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)

base_path <- here("5_downstream_analysis")
source(file.path(base_path, "global_analysis_helpers.R"))
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`, `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) %>%
  distinct() %>%
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) %>%
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

Each lab's protocol is unique; lab identity is completely confounded with
instrumental setup. Notable variations: 

- MS1 scan rate
- Injection volume
- Solvent brand.

## Metrics Definition

Three main chromatographic metrics:

- **TIC**: Sum of all ion intensities
- **BPC**: Max intensity per scan

Two data levels:

- **Full Signal**: Raw profile including baseline
- **Detected Signal**: Peaks from centWave; isolates compound signal from
  baseline

Two polarities: Positive and Negative ionization modes.

```{r load-data-both-modes}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "object/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]
    load(file.path(base_path, "object/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "object/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]
    load(file.path(base_path, "object/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = ifelse(data_level == "full", "Full Signal", "Detected Signal")) %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm = bpc / median_bpc,
      tic_norm = tic / median_tic
    ) %>%
    ungroup()
}

run_for_polarities_names <- function(fn) {
  bind_rows(fn(1L, "Positive"), fn(0L, "Negative"))
}

metrics_full <- run_for_polarities(compute_metrics, "full")
metrics_detect <- run_for_polarities(compute_metrics, "detect")

metrics_combined <- bind_rows(metrics_full, metrics_detect) %>%
  group_by(lab, polarity, data_level) %>%
  mutate(
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE)
  ) %>%
  ungroup()
```

---

# Phase 1: The Problem — Raw Data Assessment {#phase1}

This phase characterizes the raw data to understand the magnitude of 
inter-laboratory variability before any corrections are applied.

## 1.1 Distribution of Raw Metrics

### 1.1.1 Violin Plots: Overall Intensity Distribution

```{r violin-plots}
for (metric in c("bpc", "tic")) print(plot_violin(metrics_combined, metric))
```

Detected signal recenters the negative and positive distributions. 
The bimodal shape is evident.

### 1.1.2 Density Plots: Distribution Shape by Lab

```{r density-plots}
ggplot(metrics_combined %>% filter(data_level == "Full Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_minimal() +
  labs(title = "TIC Distribution by Lab (Full Signal)",
       x = "log2(TIC)", y = "Density")

ggplot(metrics_combined %>% filter(data_level == "Detected Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_minimal() +
  labs(title = "TIC Distribution by Lab (Detected Signal)",
       x = "log2(TIC)", y = "Density")
```

Full signal is more compressed than detected signal, likely because
each mixture has different numbers of compounds.

### 1.1.3 Mixture-Level Agreement Across Labs

```{r mixture-distribution}
plot_mixture_distribution <- function(data, metric) {
  if (metric %in% c("bpc", "tic")) {
    data <- data |> mutate(across(all_of(metric), log2))
    title_suffix <- "(log2)"
  } else {
    title_suffix <- ""
  }
  
  ggplot(data, aes(x = mixture, y = .data[[metric]], color = lab, group = lab)) +
    geom_line(alpha = 0.7) +
    facet_grid(data_level ~ polarity) +
    theme_minimal() +
    labs(title = paste("Mixture Distribution Across Labs -", metric, title_suffix),
         x = "Mixture", y = paste(metric, title_suffix), color = "Laboratory")
}

for (metric in c("bpc", "tic")) {
  print(plot_mixture_distribution(metrics_combined, metric))
}
```

There is good agreement between Cembio and Afekta, and some with ICL.
However, HMGU is consistently lower intensity.

## 1.2 Variance Decomposition: Quantifying the Problem

### 1.2.1 Linear Model Analysis

Let's run linear models to evaluate mixture effects and batch effects.
The goal is to quantify how much of the variance is due to lab vs mixture.
We use ANOVA to aggregate coefficients into Sum of Squares (SS), which 
quantifies the relative power of each factor.

We run this analysis across all 4 conditions (positive/negative, full/detected):

```{r variance-decomposition-lm}
results_table <- data.frame()
residual_plots <- list()
all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")

    residual_plots[[paste(pol, level, sep = " - ")]] <- result$residual_plot
    all_coefficients <- rbind(all_coefficients, 
      result$coefficients %>% mutate(Polarity = pol, Data_Level = level))
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

### 1.2.2 Lab Effect Coefficients

```{r lab-coefficients}
lab_coefs <- all_coefficients %>%
  filter(type == "Lab Effect") %>%
  mutate(lab_name = gsub("lab", "", coefficient))

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "Lab Effect Coefficients from Linear Models",
       subtitle = "Deviation from reference lab (afekta); negative = lower intensity",
       x = "Laboratory", y = "Coefficient (log2 scale)", fill = "Laboratory") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.2.3 Variance Decomposition Summary

```{r variance-stacked-bar}
variance_long <- results_table %>%
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) %>%
  pivot_longer(cols = ends_with("_Pct"), names_to = "Source", values_to = "Percentage") %>%
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Label = paste(Polarity, Data_Level, sep = "\n"),
    R2_label = paste("R² =", round(R_squared, 3))
  )

text_labels <- variance_long %>% distinct(Label, R2_label)

ggplot(variance_long, aes(x = Label, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(x = Label, label = R2_label, y = 105), data = text_labels, 
            size = 3, inherit.aes = FALSE) +
  labs(title = "Variance Decomposition: TIC Across All Conditions",
       subtitle = "Raw data shows massive lab variance — the core problem",
       x = "Polarity & Data Level", y = "Variance Contribution (%)", fill = "Source") +
  theme_minimal() +
  theme(legend.position = "top") +
  ylim(0, 110)
```

### 1.2.4 Residual Diagnostics

```{r residual-plots}
#| fig-height: 10
residual_data <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    residual_data <- rbind(residual_data, data.frame(
      Fitted = fitted(result$model), Residuals = residuals(result$model),
      Polarity = pol, Data_Level = level, 
      Lab = data_subset$lab, Mixture = as.factor(data_subset$mixture)
    ))
  }
}

ggplot(residual_data, aes(x = Fitted, y = Residuals, color = Mixture, shape = Lab)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ paste(Polarity, Data_Level, sep = "\n"), nrow = 2, ncol = 2) +
  scale_color_viridis_d(option = "turbo") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(title = "Fitted vs Residuals: TIC Across All Conditions",
       subtitle = "Points should scatter randomly around zero for a good model fit",
       x = "Fitted Values", y = "Residuals", color = "Mixture", shape = "Laboratory") +
  theme_minimal() +
  theme(legend.position = "bottom")

results_table
```

**Key Finding:** Overall variance is dominated by lab effects.
The residual plot clearly shows separation between labs vs HMGU and
still lab separation in others. Lab separation remains unexplained by
mixture. We cannot compare labs directly without normalization.

---

# Phase 2: The Solution — Normalization & Data Processing {#phase2}

This phase tests different normalization strategies to reduce 
inter-laboratory variability.

## 2.1 Normalization Strategy Selection

### 2.1.1 Rationale for Testing Different Approaches

What could be tested:

- **Median scaling**: Adjusting each lab's data to have the same median could
 help account for systematic differences. In that case we take the median of 
 that measure by that lab to correct each mixture.
    - This is safe as the experimental design is balanced. Every lab measures
     the same 38 mixtures, so their intensity distributions should be identical in 
     shape, just shifted up or down.
    - This would be wrong if the mixtures were different across labs.
- **Auto scaling**: Forces the lab to have the same dynamic change.
    - Stronger scaling. Will remove any instrument bias.

What we are NOT testing:

- **Max scaling**: ESI is subject to random electrical spike/detector saturation, 
       therefore max could be an outlier and crush other signal to 0.
    - Generally not used for entire run.

Decision criteria based on violin plot shape:

- If similar: median scaling should be fine
- If not: auto-scaling may be required

The violin plot and density plot above are a bit hard to be definitive
about. The positive data is fairly similar in shape, but the negative data
shows more variation. We test both and see how the models are affected.

### 2.1.2 Visual Comparison: Before vs After Scaling

```{r scaling-comparison-visual}
plot_scaling_comparison <- function(data, metric_raw, metric_scaled, metric_label) {
  plot_data <- data %>%
    select(lab, polarity, data_level, mixture, raw = all_of(metric_raw), scaled = all_of(metric_scaled)) %>%
    pivot_longer(cols = c(raw, scaled), names_to = "Scaling", values_to = "Value") %>%
    mutate(Scaling = ifelse(Scaling == "raw", "Unscaled", "Median Scaled"))

  ggplot(plot_data, aes(x = lab, y = log2(Value), fill = polarity)) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    geom_jitter(aes(color = as.factor(mixture)), width = 0.2, size = 1.5, alpha = 0.7) +
    scale_color_viridis_d(option = "turbo") +
    facet_grid(Scaling ~ data_level, scales = "free_y") +
    theme_minimal() +
    labs(title = paste("Before vs After Median Scaling:", metric_label),
         subtitle = "Distribution shape should remain similar, but scale shifts",
         x = "Laboratory", y = paste(metric_label, "(log2)"), color = "Mixture") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

print(plot_scaling_comparison(metrics_combined, "tic", "tic_median_scaled", "TIC"))
print(plot_scaling_comparison(metrics_combined, "bpc", "bpc_median_scaled", "BPC"))
```

In the detected signal we seem to be keeping some dynamic range, but quite hard
to see on the full signal.

## 2.2 Normalization Impact on Variance

### 2.2.1 Variance Decomposition by Normalization Method

```{r normalization-variance-comparison}
test_data <- metrics_combined %>%
  filter(polarity == "Positive", data_level == "Detected Signal")

result_raw <- check_metric_quality(test_data, "tic")
result_median_scaled <- check_metric_quality(test_data, "tic_median_scaled")
result_autoscaled <- check_metric_quality(test_data, "tic_autoscaled")

normalization_results <- bind_rows(
  result_raw$variance_table %>% mutate(Normalization = "Raw TIC", R_squared = result_raw$r_squared),
  result_median_scaled$variance_table %>% mutate(Normalization = "Median Scaled", R_squared = result_median_scaled$r_squared),
  result_autoscaled$variance_table %>% mutate(Normalization = "Autoscaled (MAD)", R_squared = result_autoscaled$r_squared)
)

normalization_results %>%
  ggplot(aes(x = Normalization, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  labs(title = "Effect of Normalization on Variance Decomposition (TIC)",
       subtitle = "Positive Mode, Detected Signal",
       x = "Normalization Method", y = "Variance Contribution (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 2.2.2 Absolute Noise Comparison

```{r normalization-noise-comparison}
mod_raw <- lm(log2(tic) ~ lab + factor(mixture), data = test_data)
mod_median_scaled <- lm(log2(tic_median_scaled) ~ lab + factor(mixture), data = test_data)
mod_autoscaled <- lm(tic_autoscaled ~ lab + factor(mixture), data = test_data)

ss_raw <- anova(mod_raw)$"Sum Sq"
ss_median_scaled <- anova(mod_median_scaled)$"Sum Sq"
ss_autoscaled <- anova(mod_autoscaled)$"Sum Sq"

cat("Raw Absolute Noise:", round(ss_raw[3], 2), "\n")
cat("Median Scaled Absolute Noise:", round(ss_median_scaled[3], 2), "\n")
cat("Autoscaled Absolute Noise:", round(ss_autoscaled[3], 2), "\n")
cat("\nRaw Total Variance:", round(sum(ss_raw), 2), "\n")
cat("Median Scaled Total Variance:", round(sum(ss_median_scaled), 2), "\n")
cat("Autoscaled Total Variance:", round(sum(ss_autoscaled), 2), "\n")
```

### 2.2.3 Residual Plots by Normalization Method

```{r normalization-residual-comparison}
#| fig-height: 14
plot_residuals(mod_raw, "Raw TIC (log2)", test_data) / 
  plot_residuals(mod_median_scaled, "Median Scaled TIC", test_data) / 
  plot_residuals(mod_autoscaled, "Autoscaled TIC (MAD)", test_data) +
  plot_annotation(title = "Residual Plots: Comparison of Normalization Methods")
```

**Findings:**

- Median scaling removes the residual pattern we could see before. 
- Autoscaling, in contrast, seems to introduce some pattern in the residuals.
- Lab variance drops significantly with median scaling. 
- Absolute noise values do not change; their overall proportion increases because we got rid of the lab variance.

Noise likely reflects injection variability and ionization
fluctuation—factors we cannot isolate with this experimental design.

**Conclusions:**

- Use median scaling
- Noise level acceptable; does not significantly impact biological variance
- We "hope" to focus on biological signals without technical variability obscuring them

## 2.3 Pure Instrument Variability (Exploratory)

Just an idea: even though the biological variance is small, we could try to remove it
to focus on instrument variability. We remove the mixture effect:

```{r instrument-variability}
inst_focus_data <- metrics_combined %>%
  filter(data_level == "Detected Signal") %>%
  group_by(mixture, polarity) %>%
  mutate(
    tic_mix_centered = tic - median(tic, na.rm = TRUE),
    tic_mix_centered = log2(tic_mix_centered + median(tic, na.rm = TRUE))
  ) %>%
  ungroup()

model <- lm(tic_mix_centered ~ lab + factor(mixture), data = inst_focus_data)
an <- anova(model)
total_ss <- sum(an$`Sum Sq`)

data.frame(
  Source = c("Instrument (Lab)", "Biology (Mixture)", "Noise (Residual)"),
  Percentage = c(
    an["lab", "Sum Sq"] / total_ss * 100,
    an["factor(mixture)", "Sum Sq"] / total_ss * 100,
    an["Residuals", "Sum Sq"] / total_ss * 100
  )
)

inst_comparison <- inst_focus_data %>%
  mutate(raw_tic_log2 = log2(tic)) %>%
  select(lab, polarity, mixture, raw_tic_log2, tic_mix_centered) %>%
  pivot_longer(cols = c(raw_tic_log2, tic_mix_centered), names_to = "Scaling", values_to = "Value") %>%
  mutate(Scaling = factor(Scaling, 
    levels = c("raw_tic_log2", "tic_mix_centered"),
    labels = c("Raw TIC (log2)", "Mixture-Centered (log2)")))

ggplot(inst_comparison, aes(x = lab, y = Value, fill = polarity)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_jitter(aes(color = as.factor(mixture)), width = 0.2, size = 1.5, alpha = 0.7) +
  scale_color_viridis_d(option = "turbo") +
  facet_grid(Scaling ~ polarity) +
  theme_minimal() +
  labs(title = "Instrument Variability: Before and After Mixture Centering",
       subtitle = "Removing biological signal reveals pure laboratory effects",
       x = "Laboratory", y = "TIC (log2 scale)", color = "Mixture") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Mixture-centered analysis isolates instrument effects. Not sure if I should 
do anything with this.

---

# Phase 3: Reproducibility & Platform Effects {#phase3}

This phase quantifies reproducibility using standard metrics (ICC) and 
investigates platform-specific effects.

## 3.1 Intraclass Correlation Coefficient (ICC)

### 3.1.1 Understanding ICC

Normalization addresses sensitivity differences across labs, but
lab-dependent noise remains. The intraclass correlation coefficient (ICC)
tests this. ICC repeats earlier variance analysis but provides scale-
invariant comparability across papers—unlike raw Sum of Squares, it's
more meaningful for visualization.

**ICC Formula:**
$$ICC = \frac{\sigma^2_{between}}{\sigma^2_{between} + \sigma^2_{within}}$$

where:

- $\sigma^2_{between}$ = variance in metric values **between** different mixtures (biological signal)
- $\sigma^2_{within}$ = variance in metric values **within** the same mixture across labs (technical noise)

**Interpretation in our context:**

- **High ICC (→1)**: Most variance comes from mixture differences (biology). Labs agree on which mixtures have high/low signal.
- **Low ICC (→0)**: Most variance comes from lab-to-lab differences. Technical noise obscures biological signal.
- **ICC = 0.5**: Equal biological and technical variance (threshold for acceptable reproducibility).

High ICC confirms measurements reflect biological differences, not
technical noise.

### 3.1.2 ICC Computation: Absolute vs Normalized Metrics

```{r icc-computation}
icc_all_results <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    
    results <- bind_rows(
      compute_icc(data_subset, "tic") %>% mutate(Type = "Absolute"),
      compute_icc(data_subset, "bpc") %>% mutate(Type = "Absolute"),
      compute_icc(data_subset, "tic_norm") %>% mutate(Type = "Normalized", Metric = "tic"),
      compute_icc(data_subset, "bpc_norm") %>% mutate(Type = "Normalized", Metric = "bpc")
    ) %>%
      mutate(Polarity = pol, Data_Level = level,
             Metric_Label = case_when(Metric == "tic" ~ "TIC", Metric == "bpc" ~ "BPC", TRUE ~ Metric))
    
    icc_all_results <- bind_rows(icc_all_results, results)
  }
}

icc_all_results %>%
  select(Polarity, Data_Level, Metric_Label, Type, ICC) %>%
  arrange(desc(ICC))
```

### 3.1.3 ICC Comparison Plot

```{r icc-comparison-plot}
ggplot(icc_all_results, aes(x = Metric_Label, y = ICC, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  facet_grid(Data_Level ~ Polarity) +
  labs(title = "Reproducibility Comparison: Absolute vs Normalized Metrics",
       subtitle = "ICC by Polarity and Data Level; dashed line = 0.5 threshold",
       x = "Metric", y = "Intraclass Correlation Coefficient (ICC)", fill = "Metric Type") +
  theme_minimal() +
  theme(legend.position = "top")
```

### 3.1.4 ICC Variance Decomposition

```{r icc-variance-decomposition}
variance_plot_intensity <- icc_all_results %>%
  select(Polarity, Data_Level, Metric_Label, Type, Between_mixture_var, Within_mixture_var) %>%
  pivot_longer(cols = c(Between_mixture_var, Within_mixture_var),
               names_to = "Source", values_to = "Variance") %>%
  mutate(Source = case_when(
    Source == "Between_mixture_var" ~ "Between mixtures (signal)",
    Source == "Within_mixture_var" ~ "Within mixtures (lab noise)",
    TRUE ~ Source
  ))

ggplot(variance_plot_intensity, aes(x = Metric_Label, y = Variance, fill = Source)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.7) +
  facet_grid(Data_Level + Type ~ Polarity, scales = "free_y") +
  labs(title = "Variance Decomposition: TIC and BPC",
       subtitle = "Higher 'between mixtures' = more signal; Higher 'within mixtures' = more lab noise",
       x = "Metric", y = "Variance (log2 scale)", fill = "Source of Variation") +
  scale_fill_manual(values = c("Between mixtures (signal)" = "#619CFF",
                                "Within mixtures (lab noise)" = "#F8766D")) +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 30, hjust = 1))
```

The overall goal is that we end up with Value = Biology.

Starting point: Value = (Biology × Sensitivity) + Background Noise

With detected signal and normalization: Value ≈ Biology

## 3.2 Impact of Peak Detection on Reproducibility

### 3.2.1 Full Signal vs Detected Signal Comparison

```{r peak-detection-impact}
icc_comparison_tic <- icc_all_results %>%
  filter(Type == "Normalized", Metric == "tic") %>%
  select(Polarity, Data_Level, ICC) %>%
  pivot_wider(names_from = Data_Level, values_from = ICC) %>%
  mutate(
    ICC_Improvement = `Detected Signal` - `Full Signal`,
    Percent_Gain = (`Detected Signal` - `Full Signal`) / `Full Signal` * 100
  )

icc_comparison_tic
```

**Findings:** Normalized TIC shows highest ICC. Detected signal (ICC 0.5–0.6)
substantially beats full signal (ICC 0.3) because peak detection removes
vendor-specific solvent noise. Median normalization improved ICC from 0.05
to 0.55—crossing the threshold from poor to moderate reproducibility. The
remaining ~0.5 ICC gap likely reflects ionization preferences and
platform-specific sensitivities.

### 3.2.2 Before & After Visualization

```{r peak-detection-visualization}
ggplot(icc_all_results %>% filter(Type == "Normalized", Metric == "tic"),
       aes(x = Polarity, y = ICC, fill = Data_Level)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(ICC, 2)), position = position_dodge(0.9), vjust = -0.5) +
  labs(title = "Impact of Peak Detection on Reproducibility",
       subtitle = "Comparison of Normalized TIC Stability (Full vs. Detected)",
       y = "Intraclass Correlation Coefficient (ICC)", x = "Polarity",
       fill = "Processing Level",
       caption = "Higher ICC = Better Agreement between labs.\nPeak detection doubles reproducibility in Negative Mode.") +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```

Peak detection substantially improves reproducibility:

- Negative mode: ICC increased from 0.275 to 0.556 (102% improvement)
- Positive mode: ICC increased from 0.375 to 0.503 (34% improvement)

**Conclusion:** All subsequent analyses use detected signal exclusively.
Peak detection removes instrument-specific baseline noise obscuring
biological signal. Full signal contains too much vendor-specific noise
for meaningful inter-lab comparisons.

## 3.3 Platform Effects: Agilent Labs Comparison

### 3.3.1 "Ceiling" Estimate: Same-Platform Labs

For comparison, two labs (CEMBIO and AFEKTA) share the same LC-MS
platform (Agilent Q-TOF). ICC of 0.83 (TIC) and 0.92 (BPC) for these labs
proves selectivity matches when vendor matches. The 0.5–0.8 gap reflects
instrument-specific and ionization differences. 

```{r agilent-ceiling-icc}
agilent_data <- metrics_combined %>%
  filter(lab %in% c("afekta", "cembio"), polarity == "Positive", data_level == "Detected Signal")

icc_agilent_tic <- compute_icc(agilent_data, "tic_norm")
icc_agilent_bpc <- compute_icc(agilent_data, "bpc_norm")

cat("=== Ceiling ICC Estimate (Agilent Labs Only, Positive Mode, Detected Signal) ===\n")
cat("ICC for Normalized TIC:", round(icc_agilent_tic$ICC, 3), "\n")
cat("ICC for Normalized BPC:", round(icc_agilent_bpc$ICC, 3), "\n")
```

### 3.3.2 Effect Size Between Agilent Labs

**What is effect size and why use it here?**

Effect size quantifies the *magnitude* of difference between groups, independent 
of sample size. While ICC tells us how reproducible measurements are, effect size 
tells us *how much* the labs actually differ. We compute log2 fold-changes 
between Cembio and Afekta for each mixture, then summarize across all mixtures.

This complements ICC by answering: "When labs disagree, by how much?"

```{r agilent-effect-size}
agilent_effect_size <- metrics_combined %>%
  filter(lab %in% c("afekta", "cembio"), data_level == "Detected Signal") %>%
  select(lab, mixture, polarity, tic_norm, bpc_norm) %>%
  pivot_wider(names_from = lab, values_from = c(tic_norm, bpc_norm)) %>%
  mutate(
    log_diff_tic = log2(tic_norm_cembio / tic_norm_afekta),
    log_diff_bpc = log2(bpc_norm_cembio / bpc_norm_afekta)
  )

agilent_effect_size %>%
  group_by(polarity) %>%
  summarise(
    TIC_Mean_Log = mean(log_diff_tic, na.rm = TRUE),
    TIC_Median_Log = median(log_diff_tic, na.rm = TRUE),
    TIC_Percent_Diff = (2^abs(TIC_Mean_Log) - 1) * 100,
    BPC_Mean_Log = mean(log_diff_bpc, na.rm = TRUE),
    BPC_Median_Log = median(log_diff_bpc, na.rm = TRUE),
    BPC_Percent_Diff = (2^abs(BPC_Mean_Log) - 1) * 100,
    .groups = "drop"
  )
```

**Key findings:**

- **Positive mode**: Median log2 difference ≈ 0 (labs agree within ~1%); mean 
  pulled by outliers (~6.6% difference)
- **Negative mode**: Consistent ~9% systematic bias that median normalization 
  cannot fully remove

Summary table:
| Metric   | Pattern           | Explanation |
|----------|-------------------|---------------------------------------------|
| TIC_norm | Afekta Higher     | 6546 superior dynamic range; integrates |
|          |                   | more low-abundance signals |
| BPC_norm | Cembio Higher     | 6545 distribution more top-heavy; peaks |
|          |                   | appear taller after normalization |
| Ratio    | Cembio Higher     | 6545 misses background; lower TIC |
|          |                   | denominator inflates ratio |

The two Agilent Q-TOF generations (6546 vs 6545) can be compared for trends 
(high ICC), but would need non-linear correction for absolute quantitation.

## 3.4 Ionization Preference Hypothesis

### 3.4.1 Peak Detection Counts by Lab

One hypothesis for different ICC values across labs is that
different instruments may prefer different ionization pathways. This would lead to
systematic differences in peak intensities across labs, which would reduce
the ICC values.

```{r peak-counts-by-lab}
pt_afekta <- read.csv(file.path(base_path, "object/detected_peaks_afekta_HE.csv"))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(base_path, "object/detected_peaks_cembio_HE.csv"))

lab_counts <- c(afekta = nrow(pt_afekta), hmgu = nrow(pt_hmgu), 
                icl = nrow(pt_icl), cembio = nrow(pt_cembio))
barplot(lab_counts, main = "Number of Detected Peaks per Lab", 
        ylab = "Number of detected peaks", xlab = "Labs", col = "steelblue")
```

### 3.4.2 Intensity Rank Correlation (Spearman)

We run a Spearman analysis, which ranks the compound 
intensities and checks how well the ranks are preserved across labs.
This is run on a consensus table of detected peaks that are present in all 4 labs.

```{r intensity-correlation-setup}
ct_filtered <- read.csv(file.path(base_path, "object/ct_filtered4.csv"))

intensity_table <- ct_filtered %>%
  select(consensus_id, sample, idx_afekta, chrom_peak_id_afekta, 
         chrom_peak_id_icl, chrom_peak_id_hmgu, chrom_peak_id_cembio) %>%
  left_join(pt_afekta %>% select(chrom_peak_id, into, rt),
            by = c("chrom_peak_id_afekta" = "chrom_peak_id")) %>%
  rename(intensity_afekta = into, rt_afekta = rt) %>%
  left_join(pt_icl %>% select(chrom_peak_id, into, rt),
            by = c("chrom_peak_id_icl" = "chrom_peak_id")) %>%
  rename(intensity_icl = into, rt_icl = rt) %>%
  left_join(pt_hmgu %>% select(chrom_peak_id, into, rt),
            by = c("chrom_peak_id_hmgu" = "chrom_peak_id")) %>%
  rename(intensity_hmgu = into, rt_hmgu = rt) %>%
  left_join(pt_cembio %>% select(chrom_peak_id, into, rt),
            by = c("chrom_peak_id_cembio" = "chrom_peak_id")) %>%
  rename(intensity_cembio = into, rt_cembio = rt)

head(intensity_table %>% select(consensus_id, sample, starts_with("intensity"), starts_with("rt")))
```

```{r spearman-correlation}
intensity_matrix <- intensity_table %>% select(starts_with("intensity"))
cor(intensity_matrix, method = "spearman", use = "pairwise.complete.obs")
```

**Finding:** Cembio and Afekta are the most similar, with a Spearman
correlation of 0.77. HMGU has the lowest correlation with all the labs.

### 3.4.3 Peak Overlap Across Labs (UpSet Plot)

```{r upset-plot}
ct_all_sample <- read.csv(file.path(base_path, "object/ct_all_samples.csv"))

upset_data <- ct_all_sample |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(afekta = as.integer(afekta_detected), hmgu = as.integer(hmgu_detected),
         icl = as.integer(icl_detected), cembio = as.integer(cembio_detected)) |>
  select(afekta, hmgu, icl, cembio)

upset_data_df <- as.data.frame(upset_data)
rownames(upset_data_df) <- paste0("peak_", seq_len(nrow(upset_data_df)))

library(UpSetR)
upset(upset_data_df, sets = c("afekta", "hmgu", "icl", "cembio"),
      main.bar.color = "steelblue", sets.bar.color = "darkred",
      text.scale = c(2, 2, 2, 1.5, 2, 2), point.size = 3, line.size = 1.2)
```

---

# Phase 4: Granular Nuance — Retention Time & Biology {#phase4}

This phase explores retention time-dependent effects and biological
signal variations across the chromatographic gradient.

## 4.1 Retention Time Range

All the labs had a retention time gradient of 15min. After preprocessing we 
reduced to a common range:

```{r rt-range}
load(file.path(base_path, "object/tic_full.RData"))
range(rtime(tic_full[1L])[[1]])
```

How do the labs agree across the retention time gradient? Do some labs
show systematic sensitivity drops at early or late elution times?

## 4.2 RT-Binned Correlation Analysis

### 4.2.1 Full Signal: Pairwise Correlation by Mixture

```{r rt-correlation-full-signal}
load(file.path(base_path, "object/tic_full.RData"))
rt_range <- range(rtime(tic_full[1L])[[1]])
breaks <- seq(rt_range[1], rt_range[2], by = 5)

polarities <- unique(tic_full$polarity)
mixtures <- unique(tic_full$mixture)
correlation_results_full <- list()

for (pol in polarities) {
  for (mix in mixtures) {
    subset_tic <- tic_full[which(tic_full$polarity == pol & tic_full$mixture == mix)]
    if (length(subset_tic) == 0) next
    
    intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
      icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
      cembio = numeric(length(breaks) - 1))
    
    for (i in seq_along(subset_tic)) {
      lab_name <- subset_tic$lab[i]
      peaks <- peaksData(subset_tic[i])[[1]]
      intensity_matrix[[lab_name]] <- tapply(peaks$intensity, 
        cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
    }
    
    cor_results <- cor(intensity_matrix, method = "pearson", use = "pairwise.complete.obs")
    polarity_label <- ifelse(pol == 1, "pos", "neg")
    correlation_results_full[[paste0(polarity_label, "_", mix)]] <- cor_results
  }
}

pairs <- list(c("afekta", "cembio"), c("afekta", "hmgu"), c("afekta", "icl"),
              c("cembio", "hmgu"), c("cembio", "icl"), c("hmgu", "icl"))

pairwise_cors_full <- data.frame()
for (result_name in names(correlation_results_full)) {
  cor_mat <- correlation_results_full[[result_name]]
  for (pair in pairs) {
    pairwise_cors_full <- rbind(pairwise_cors_full, data.frame(
      mixture = result_name, pair = paste(pair[1], pair[2], sep = ":"),
      correlation = cor_mat[pair[1], pair[2]]))
  }
}

cor_heatmap_data_full <- pairwise_cors_full |>
  separate(mixture, into = c("polarity", "mix"), sep = "_", extra = "merge") |>
  mutate(polarity = factor(polarity, levels = c("neg", "pos")))

ggplot(cor_heatmap_data_full, aes(x = pair, y = mix, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0.5, limits = c(0, 1)) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "TIC Correlation Heatmap Across All Mixtures (Full Signal)",
       x = "Lab Pair", y = "Mixture", fill = "Pearson r")
```

### 4.2.2 Full Signal: Correlation Distribution

```{r rt-correlation-full-distribution}
library(ggbeeswarm)

ggplot(cor_heatmap_data_full, aes(x = pair, y = correlation, color = pair)) +
  geom_violin(alpha = 0.3, show.legend = FALSE) +
  geom_beeswarm(size = 2, alpha = 0.7, show.legend = FALSE) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", show.legend = FALSE) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of TIC Correlations Across 19 Mixtures (Full Signal)",
       subtitle = "Each point represents one mixture", x = "Lab Pair", y = "Pearson Correlation")
```

### 4.2.3 Full Signal: RT-Dependent Correlation (Single Mixture Example)

```{r rt-dependent-correlation-full}
compute_rt_bin_correlation_full <- function(polarity, mixture) {
  subset_tic <- tic_full[which(tic_full$polarity == polarity & tic_full$mixture == mixture)]
  if (length(subset_tic) == 0) return(NULL)
  
  rt_range <- range(rtime(tic_full[1L])[[1]])
  breaks <- seq(rt_range[1], rt_range[2], by = 5)
  
  intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
    icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
    cembio = numeric(length(breaks) - 1))
  
  for (i in seq_along(subset_tic)) {
    lab_name <- subset_tic$lab[i]
    peaks <- peaksData(subset_tic[i])[[1]]
    intensity_matrix[[lab_name]] <- tapply(peaks$intensity, cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
  }
  
  rt_centers <- (breaks[-1] + breaks[-length(breaks)]) / 2
  window_size <- 10
  pairs <- list(c("afekta", "cembio"), c("afekta", "hmgu"), c("afekta", "icl"),
                c("cembio", "hmgu"), c("cembio", "icl"), c("hmgu", "icl"))
  bin_cors <- list()
  
  for (bin_idx in 1:nrow(intensity_matrix)) {
    start_idx <- max(1, bin_idx - window_size %/% 2)
    end_idx <- min(nrow(intensity_matrix), bin_idx + window_size %/% 2)
    window_data <- intensity_matrix[start_idx:end_idx, ]
    if (nrow(window_data) < 3) next
    cor_mat <- cor(window_data, method = "pearson", use = "pairwise.complete.obs")
    if (any(is.na(cor_mat))) next
    for (pair in pairs) {
      bin_cors[[length(bin_cors) + 1]] <- data.frame(
        rt_center = rt_centers[bin_idx], pair = paste(pair[1], pair[2], sep = ":"),
        correlation = cor_mat[pair[1], pair[2]])
    }
  }
  do.call(rbind, bin_cors)
}

rt_cor_example_full <- compute_rt_bin_correlation_full(1, "1_1")

if (!is.null(rt_cor_example_full)) {
  ggplot(rt_cor_example_full, aes(x = rt_center, y = pair, fill = correlation)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(title = "RT-Dependent Correlation: Positive Mode, Mixture 1_1 (Full Signal)",
         subtitle = "Using 5-second RT bins with 50-second sliding window",
         x = "Retention Time (s)", y = "Lab Pair", fill = "Pearson r")
}
```

Note: This is only one mixture and one polarity. The x-axis should start at the 
actual RT range start (~17s).

### 4.2.4 Detected Signal: Pairwise Correlation by Mixture

```{r rt-correlation-detected-signal}
load(file.path(base_path, "object/tic_detect.RData"))
rt_range <- range(rtime(tic_detect[1L])[[1]])
breaks <- seq(rt_range[1], rt_range[2], by = 5)

polarities <- unique(tic_detect$polarity)
mixtures <- unique(tic_detect$mixture)
correlation_results <- list()

for (pol in polarities) {
  for (mix in mixtures) {
    subset_tic <- tic_detect[which(tic_detect$polarity == pol & tic_detect$mixture == mix)]
    if (length(subset_tic) == 0) next
    
    intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
      icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
      cembio = numeric(length(breaks) - 1))
    
    for (i in seq_along(subset_tic)) {
      lab_name <- subset_tic$lab[i]
      peaks <- peaksData(subset_tic[i])[[1]]
      intensity_matrix[[lab_name]] <- tapply(peaks$intensity, cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
    }
    
    cor_results <- cor(intensity_matrix, method = "pearson", use = "pairwise.complete.obs")
    polarity_label <- ifelse(pol == 1, "pos", "neg")
    correlation_results[[paste0(polarity_label, "_", mix)]] <- cor_results
  }
}

pairwise_cors <- data.frame()
for (result_name in names(correlation_results)) {
  cor_mat <- correlation_results[[result_name]]
  for (pair in pairs) {
    pairwise_cors <- rbind(pairwise_cors, data.frame(
      mixture = result_name, pair = paste(pair[1], pair[2], sep = ":"),
      correlation = cor_mat[pair[1], pair[2]]))
  }
}

cor_heatmap_data <- pairwise_cors |>
  separate(mixture, into = c("polarity", "mix"), sep = "_", extra = "merge") |>
  mutate(polarity = factor(polarity, levels = c("neg", "pos")))

ggplot(cor_heatmap_data, aes(x = pair, y = mix, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0.5, limits = c(0, 1)) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "TIC Correlation Heatmap Across All Mixtures (Detected Signal)",
       x = "Lab Pair", y = "Mixture", fill = "Pearson r")
```

### 4.2.5 Detected Signal: Correlation Distribution

```{r rt-correlation-detected-distribution}
ggplot(cor_heatmap_data, aes(x = pair, y = correlation, color = pair)) +
  geom_violin(alpha = 0.3, show.legend = FALSE) +
  geom_beeswarm(size = 2, alpha = 0.7, show.legend = FALSE) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", show.legend = FALSE) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of TIC Correlations Across 19 Mixtures (Detected Signal)",
       subtitle = "Each point represents one mixture", x = "Lab Pair", y = "Pearson Correlation")
```

Note: How can some correlations be negative?

### 4.2.6 Detected Signal: RT-Dependent Correlation (Single Mixture Example)

```{r rt-dependent-correlation-detected}
compute_rt_bin_correlation <- function(polarity, mixture) {
  subset_tic <- tic_detect[which(tic_detect$polarity == polarity & tic_detect$mixture == mixture)]
  if (length(subset_tic) == 0) return(NULL)
  
  rt_range <- range(rtime(tic_detect[1L])[[1]])
  breaks <- seq(rt_range[1], rt_range[2], by = 5)
  
  intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
    icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
    cembio = numeric(length(breaks) - 1))
  
  for (i in seq_along(subset_tic)) {
    lab_name <- subset_tic$lab[i]
    peaks <- peaksData(subset_tic[i])[[1]]
    intensity_matrix[[lab_name]] <- tapply(peaks$intensity, cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
  }
  
  rt_centers <- (breaks[-1] + breaks[-length(breaks)]) / 2
  window_size <- 10
  bin_cors <- list()
  
  for (bin_idx in 1:nrow(intensity_matrix)) {
    start_idx <- max(1, bin_idx - window_size %/% 2)
    end_idx <- min(nrow(intensity_matrix), bin_idx + window_size %/% 2)
    window_data <- intensity_matrix[start_idx:end_idx, ]
    if (nrow(window_data) < 3) next
    cor_mat <- cor(window_data, method = "pearson", use = "pairwise.complete.obs")
    if (any(is.na(cor_mat))) next
    for (pair in pairs) {
      bin_cors[[length(bin_cors) + 1]] <- data.frame(
        rt_center = rt_centers[bin_idx], pair = paste(pair[1], pair[2], sep = ":"),
        correlation = cor_mat[pair[1], pair[2]])
    }
  }
  do.call(rbind, bin_cors)
}

rt_cor_example <- compute_rt_bin_correlation(1, "1_1")

if (!is.null(rt_cor_example)) {
  ggplot(rt_cor_example, aes(x = rt_center, y = pair, fill = correlation)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(title = "RT-Dependent Correlation: Positive Mode, Mixture 1_1 (Detected Signal)",
         subtitle = "Using 5-second RT bins with 50-second sliding window",
         x = "Retention Time (s)", y = "Lab Pair", fill = "Pearson r")
}
```

### 4.2.7 Comparison: Full vs Detected Signal Correlations

```{r full-vs-detected-correlation}
comparison_data <- bind_rows(
  cor_heatmap_data_full %>% mutate(signal_type = "Full Signal"),
  cor_heatmap_data %>% mutate(signal_type = "Detected Signal")
)

ggplot(comparison_data, aes(x = pair, y = correlation, fill = signal_type)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Comparison: Full vs Detected Signal Correlations",
       subtitle = "Across all 19 mixtures and 6 lab pairs",
       x = "Lab Pair", y = "Pearson Correlation", fill = "Signal Type")
```

## 4.3 RT-Dependent Normalization Assessment

### 4.3.1 Rationale for Local Normalization

Median scaling using the overall median: What fraction of total metabolic
payload elutes in Q1 vs Q2? This preserves quartile relationships and tests
if labs agree on gradient "shape." This only works if metabolite
distribution is similar across gradient—check this below.

The alternative is local normalization, which computes median scaling
factors independently for each RT quartile. This corrects for labs with
non-linear sensitivity biases across the gradient.

Raw data check: Look at instrument sensitivity profile.

- Flat line: global normalization fine
- Sloped line: local normalization needed

### 4.3.2 Instrument Sensitivity Profile Across RT

```{r sensitivity-profile-computation}
compute_consortium_sensitivity <- function(polarity) {
  load(file.path(base_path, "object/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  breaks <- quantile(unlist(rtime(tic_data)), probs = seq(0, 1, 0.25), na.rm = TRUE)

  raw_stats <- t(vapply(seq_along(tic_data), function(i) {
    rt <- rtime(tic_data[i])[[1]]; int <- intensity(tic_data[i])[[1]]
    vapply(1:4, function(q) sum(int[rt >= breaks[q] & rt < breaks[q+1]], na.rm = TRUE), numeric(1))
  }, numeric(4)))

  df <- data.frame(lab = tic_data$lab, mixture = tic_data$mixture,
                   Q1 = raw_stats[,1], Q2 = raw_stats[,2], Q3 = raw_stats[,3], Q4 = raw_stats[,4])
  grand_medians <- df %>% 
    summarise(across(starts_with("Q"), \(x) median(x, na.rm = TRUE))) %>%
    pivot_longer(everything(), names_to = "Quartile", values_to = "Consensus")

  df %>%
    pivot_longer(cols = starts_with("Q"), names_to = "Quartile", values_to = "Intensity") %>%
    left_join(grand_medians, by = "Quartile") %>%
    mutate(
      Log2_Deviation = log2(Intensity / Consensus),
      Polarity = ifelse(polarity == 1, "Positive", "Negative")
    )
}

# Compute sensitivity profiles for both polarities
all_profiles <- run_for_polarities_simple(compute_consortium_sensitivity)
```

### 4.3.3 Raw TIC by Quartile

```{r sensitivity-profile-plots}
# 5a. Raw TIC values per quartile (absolute scale)
all_profiles %>%
  mutate(Log2_TIC = log2(Intensity)) %>%
  group_by(Quartile, lab, Polarity) %>%
  summarise(median_log2_tic = median(Log2_TIC, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Quartile, y = median_log2_tic, group = lab, color = lab)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~Polarity) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Raw TIC Distribution Across Retention Time Quartiles",
    subtitle = "Absolute intensities show both lab differences and RT-dependent patterns",
    y = "log2(TIC Sum per Quartile)",
    x = "Quartile",
    color = "Laboratory"
  ) +
  theme_minimal()

# 5b. Deviation from consortium median (relative scale)
all_profiles %>%
  mutate(Log2_TIC = log2(Intensity)) %>%
  ggplot(aes(x = Quartile, y = Log2_TIC, color = lab, fill = lab)) +
  geom_boxplot(alpha = 0.3, outlier.alpha = 0.4, position = position_dodge(width = 0.8)) +
  stat_summary(aes(group = lab), fun = median, geom = "line", linewidth = 1, position = position_dodge(width = 0.8)) +
  stat_summary(aes(group = lab), fun = median, geom = "point", size = 2.5, position = position_dodge(width = 0.8)) +
  facet_wrap(~Polarity) +
  scale_color_viridis_d(option = "turbo") +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Lab Variation Across Retention Time Quartiles (Raw Data with Distribution)",
    subtitle = "Distribution shows variation across mixtures; non-parallel medians justify local normalization",
    y = "log2(TIC Sum per Quartile)",
    x = "Quartile",
    color = "Laboratory",
    fill = "Laboratory"
  ) +
  theme_minimal()
```

Note: This might be problematic to evaluate like this since every mixture is quite different,
so we have massive spread within each quartile.

**Findings:** Sensitivity slopes are non-linear: ICL shows progressive loss from Q1
(+0.23) to Q4 (-0.78); Agilents show inverse trend. Cembio exhibits
distinct hyper-sensitivity to lipids in Q3 (+1.55 log deviation). These
retention-time-specific biases cannot be corrected with global
normalization. 

**Solution:** Apply Local Median Normalization, computing
scaling factors independently for each RT quartile.

## 4.4 Variance Decomposition by RT Quartile

### 4.4.1 Linear Model Analysis by Quartile

Distribution of biological signal (TIC and BPC) across RT quartiles:

```{r quartile-variance-analysis}
compute_rt_quartiles_biological <- function(polarity, polarity_name) {
  load(file.path(base_path, "object/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  load(file.path(base_path, "object/bpc_detect.RData"))
  bpc_data <- bpc_detect[bpc_detect$polarity == polarity]

  tic_medians <- data.frame(lab = tic_data$lab,
    tic_total = vapply(intensity(tic_data), sum, numeric(1), na.rm = TRUE)) %>%
    group_by(lab) %>% summarise(median_tic = median(tic_total, na.rm = TRUE), .groups = "drop")

  bpc_medians <- data.frame(lab = bpc_data$lab,
    bpc_total = vapply(intensity(bpc_data), sum, numeric(1), na.rm = TRUE)) %>%
    group_by(lab) %>% summarise(median_bpc = median(bpc_total, na.rm = TRUE), .groups = "drop")

  all_rts <- unlist(rtime(tic_data))
  if (length(all_rts) == 0) return(NULL)
  global_breaks <- quantile(all_rts, probs = seq(0, 1, 0.25), na.rm = TRUE)

  compute_quartile_sums <- function(int_vec, rt_vec, breaks, lab_median) {
    vapply(1:4, function(q) sum(int_vec[rt_vec >= breaks[q] & rt_vec < breaks[q+1]], na.rm = TRUE) / lab_median, numeric(1))
  }

  compute_window_stats <- function(tic_chr, bpc_chr, breaks, lab_name, tic_meds, bpc_meds) {
    rt <- rtime(tic_chr)[[1]]
    c(compute_quartile_sums(intensity(tic_chr)[[1]], rt, breaks, tic_meds$median_tic[tic_meds$lab == lab_name]),
      compute_quartile_sums(intensity(bpc_chr)[[1]], rt, breaks, bpc_meds$median_bpc[bpc_meds$lab == lab_name]))
  }

  stats <- t(vapply(seq_along(tic_data), function(i) {
    compute_window_stats(tic_data[i], bpc_data[i], global_breaks, tic_data$lab[i], tic_medians, bpc_medians)
  }, numeric(8)))

  colnames(stats) <- c(
    paste0("TIC_Frac_Q", 1:4),
    paste0("BPC_Frac_Q", 1:4)
  )

  result <- data.frame(
    lab = tic_data$lab,
    mixture = tic_data$mixture,
    polarity = polarity_name,
    stats
  )

  return(result)
}

# Compute biological quartile distributions for both polarities
bio_combined_quartiles <- run_for_polarities_names(compute_rt_quartiles_biological)

quartile_results <- expand.grid(Polarity = c("Positive", "Negative"),
                                Metric = c(paste0("TIC_Frac_Q", 1:4), paste0("BPC_Frac_Q", 1:4))) %>%
  rowwise() %>%
  mutate(var_result = list(get_variance(filter(bio_combined_quartiles, polarity == Polarity), Metric))) %>%
  unnest(var_result)

quartile_results
```

### 4.4.2 Variance Visualization by Quartile

```{r quartile-variance-plots}
print(plot_quartile_variance(quartile_results, "TIC"))
print(plot_quartile_variance(quartile_results, "BPC"))
```

**Finding:** Local normalization by RT quartile reveals instrument-specific biases
in Q3/Q4 (30–50% residual lab variance). Hydrophobic compounds show
irreducible differences in selectivity and dynamic range that median
scaling cannot eliminate. 

## 4.5 Ionization Preference Across RT

### 4.5.1 Quartile Assignment for Peak Intensities

Now that we showed that quartiles varied highly in lab variance, we can check if
the ionization preference is also varying across RT quartile.

```{r quartile-assignment}
load(file.path(base_path, "object/tic_full.RData"))
global_breaks <- quantile(unlist(rtime(tic_full[tic_full$polarity == 1L])),
                          probs = seq(0, 1, 0.25), na.rm = TRUE)

intensity_table <- intensity_table %>%
  mutate(
    mean_rt = rowMeans(select(., starts_with("rt_")), na.rm = TRUE),
    quartile = cut(mean_rt, breaks = global_breaks, labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE)
  )

cat("Quartile Breaks:\n"); print(round(global_breaks, 2))
cat("\nPeaks per quartile:\n"); table(intensity_table$quartile, useNA = "ifany")
```

### 4.5.2 Spearman Correlation by Quartile

```{r spearman-by-quartile}
quartile_correlations <- intensity_table %>%
  filter(!is.na(quartile)) %>%
  group_by(quartile) %>%
  summarise(
    n_peaks = n(),
    afekta_icl = cor(intensity_afekta, intensity_icl, method = "spearman", use = "pairwise.complete.obs"),
    afekta_hmgu = cor(intensity_afekta, intensity_hmgu, method = "spearman", use = "pairwise.complete.obs"),
    afekta_cembio = cor(intensity_afekta, intensity_cembio, method = "spearman", use = "pairwise.complete.obs"),
    icl_hmgu = cor(intensity_icl, intensity_hmgu, method = "spearman", use = "pairwise.complete.obs"),
    icl_cembio = cor(intensity_icl, intensity_cembio, method = "spearman", use = "pairwise.complete.obs"),
    hmgu_cembio = cor(intensity_hmgu, intensity_cembio, method = "spearman", use = "pairwise.complete.obs"),
    .groups = "drop"
  )

quartile_correlations
```

Notes: There's an NA in one of the rows. Need to recalculate nlab in the other script.
This also suggests that normalizing by quartile and redoing the global analysis might be valuable.

---

# Summary {#summary}

This analysis assessed LC-MS reproducibility across 4 laboratories
analyzing 19 standard mixtures. Key findings:

1. **Peak detection improves reproducibility**: Detected signal ICC >0.5
   vs full signal ICC <0.1. Isolating compound signal from baseline noise
   is essential.

2. **Same-platform instruments agree better**: Agilent Q-TOF labs
   (Afekta, Cembio) show ICC 0.83–0.92; different platforms show lower
   agreement. Suggests instrument design dominates variability.

3. **Normalization strategy matters**: Median normalization improves ICC
   from 0.05 to 0.55; volume normalization increases differences.
   Naturally normalized metrics (ratios, percentages) are more stable
   across labs.

4. **Negative mode bias**: Needs further investigation.

5. **Local normalization**: RT-quartile-specific normalization handles
   non-linear instrument sensitivity.

---

# Future Directions

- Combine with peak shape results
- Incorporate more MsQuality metrics (baseline intensity, rtIqrChromatograms)
- Expand to additional chemical classes: essentially need to run peak detection
- Test in-house vs. commercial reverse-phase columns within same lab
    - Afekta is the only lab that has that proper setup for this.
- Annotated data? We will have it only for the HE data, so maybe not for now.

---

# Appendix: Technical Notes

Volume normalization increased differences; platform-independent
intensity scaling is ineffective. Median scaling is preferred. Naturally
normalized metrics (sensitivity, ratios) remain stable across labs.

Lab-dependent correlations indicate the ratio cannot serve as a universal
metric for cross-lab comparisons.


