---
title: "Global Signal Analysis - Positive Mode"
format: html
editor_options:
  chunk_output_type: console
---

# Introduction

In this simplified analysis, I compute key metrics from preprocessed LC-MS data to compare 
across different labs and instrument setups. The analysis focuses on:

- **Base Peak Chromatogram (BPC)**: Sum of peak apices
- **Total Ion Current (TIC)**: Sum of all peak areas
- **BPC/TIC Ratio**: Metric for peak purity/complexity
- **TIC Quantiles**: Distribution assessment

For each metric, I evaluate the data distribution to determine if transformation is needed
before statistical analysis, then fit linear models: `Metric ~ lab + mixture + setup`

The analysis is performed on three data levels:
1. **Full Data**: All detected signal
2. **Detected Data**: Signal from peak picking (excludes baseline noise)
3. **Annotated Data**: Only annotated standard compounds

Setup information is derived from the LC and MS instruments in the lab metadata.


## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggbeeswarm)

# For distribution assessment
library(nortest)
library(car)

# For linear models
library(broom)
```

```{r, cores-cluster}
# cluster parallel setting
bpparam <- MulticoreParam(workers = parallel::detectCores() - 1)
```

## Setup Information

Load and prepare the setup metadata:

```{r load-setup}
setup <- read.csv("lab_setup_phase1.csv", skip = 1)

# Create a combined setup variable from LC and MS instruments
setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC-Instrument`, `MS-Instrument`) %>%
  distinct() %>%
  mutate(
    lc_short = case_when(
      grepl("Agilent 1290", `LC-Instrument`) ~ "Agilent1290",
      grepl("Acquity", `LC-Instrument`) ~ "Acquity",
      grepl("ExionLC", `LC-Instrument`) ~ "ExionLC",
      grepl("Bruker", `LC-Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_short = case_when(
      grepl("6540|6545", `MS-Instrument`) ~ "Agilent",
      grepl("Synapt", `MS-Instrument`) ~ "Waters",
      grepl("Zeno", `MS-Instrument`) ~ "Sciex",
      grepl("TIMS", `MS-Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    setup = paste(lc_short, ms_short, sep = "_")
  ) %>%
  select(lab, setup)

setup_simplified
```

# Full Data Analysis

```{r}
pol <- 1L  # positive polarity
```

Load the full preprocessed data:

```{r load-full-data}
load(file.path("object/sp_full.RData"))
sp_full <- sp_full[sp_full$polarity == pol]

load("object/bpc_full.RData")
bpc_full <- bpc_full[bpc_full$polarity == pol]

load("object/tic_full.RData")
tic_full <- tic_full[tic_full$polarity == pol]
```

## Compute Metrics

```{r compute-full-metrics}
metrics_full <- data.frame(
  lab = bpc_full$lab,
  mixture = bpc_full$mixture,
  bpc = vapply(intensity(bpc_full), sum, numeric(1), na.rm = TRUE),
  tic = vapply(intensity(tic_full), sum, numeric(1), na.rm = TRUE)
)

metrics_full$bpc_tic_ratio <- metrics_full$bpc / metrics_full$tic

# Add TIC quantiles
tic_quantiles <- lapply(intensity(tic_full), function(x) {
  quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
})
metrics_full$tic_q25 <- sapply(tic_quantiles, `[`, 1)
metrics_full$tic_q50 <- sapply(tic_quantiles, `[`, 2)
metrics_full$tic_q75 <- sapply(tic_quantiles, `[`, 3)

# Add setup information
metrics_full <- metrics_full %>%
  left_join(setup_simplified, by = "lab")

head(metrics_full)
```

## Distribution Assessment

For each metric, assess normality to determine if transformation is needed:

```{r dist-assessment-full}
metrics_to_assess <- c("bpc", "tic", "bpc_tic_ratio", "tic_q25", "tic_q50", "tic_q75")

distribution_summary_full <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in metrics_to_assess) {
  data_vec <- metrics_full[[m]]
  data_vec <- data_vec[!is.na(data_vec)]
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)  # Add 1 to handle potential zeros
  shapiro_log <- shapiro.test(log_data)
  
  # Recommend transform if log is more normal (p > 0.05)
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_full <- rbind(distribution_summary_full, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = shapiro_test$p.value,
    log_shapiro_p = shapiro_log$p.value,
    transform_recommended = transform_rec
  ))
}

distribution_summary_full
```

## Linear Models - Full Data

```{r lm-full}
# BPC model
lm_bpc_full <- lm(log(bpc) ~ lab + mixture + setup, data = metrics_full)
broom::tidy(lm_bpc_full)
summary(lm_bpc_full)

# TIC model
lm_tic_full <- lm(log(tic) ~ lab + mixture + setup, data = metrics_full)
broom::tidy(lm_tic_full)
summary(lm_tic_full)

# BPC/TIC ratio model
lm_ratio_full <- lm(bpc_tic_ratio ~ lab + mixture + setup, data = metrics_full)
broom::tidy(lm_ratio_full)
summary(lm_ratio_full)

# TIC quantiles
lm_tic_q50_full <- lm(log(tic_q50) ~ lab + mixture + setup, data = metrics_full)
broom::tidy(lm_tic_q50_full)
summary(lm_tic_q50_full)
```

```{r clean-full}
rm(tic_full, bpc_full, sp_full)
```

# Detected signal analysis

This signal is essentially keeping the rt and mz area from peak detection.
Now it reflects total detected signal (sensitivity of detection).

Easier to compare labs because baseline noise is excluded.
Differences now relate more to how well peaks were detected and instrument
sensitivity rather than background noise.

Load the results sheet below

```{r load-detect}
load("object/sp_full_detect.RData")
sp_full_detect <- sp_full_detect[sp_full_detect$polarity == pol]
```

Summarize the number of MS1 spectra per lab:

```{r nb-spectra-detect}
table(sp_full_detect$lab) ## compare with sp_full
```

Summarize the rtime range per lab (in seconds):

```{r rt-range-detect}
lapply(split(rtime(sp_full_detect), sp_full_detect$lab) , range)
```

## Basic metrics

Now baseline noise and spurious fluctuations are largely removed.

### Base Peak Chromatogram (BPC) area

We generate the BPC from the detected signal.

```{r load-bpc-detect}
load("object/bpc_detect.RData")
bpc_detect <- bpc_detect[bpc_detect$polarity == pol]
```

We summarize the BPC area per mixture per lab by summing the intensity values.

Inference from Detected BPC
This is a much more meaningful metric than the "full" BPC. It represents
the sum of the apex intensities of all detected peaks. This is a direct proxy
for instrument sensitivity and peak detection efficiency. A lab with a higher
detected BPC is either detecting more peaks or detecting the
same peaks at a higher intensity. The CV per mixture  quantifies the inter-lab
variability in this sensitivity.

```{r sum-bpc-detect}
sbpc <- vapply(intensity(bpc_detect), sum, numeric(1), na.rm = TRUE)

plot_data_bpc_det <- data.frame(
  Lab = bpc_detect$lab,
  Mixture = bpc_detect$mixture,
  BPC_Area = sbpc
)

ggplot(plot_data_bpc_det, aes(x = Lab, y = log2(BPC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Detected BPC Area Across Laboratories (Positive Mode)",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "Log2(Detected BPC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Next, we compute the coefficient of variation (CV) per mixture across labs.

```{r cv-detect}
cv_per_mix_det <- tapply(sbpc, bpc_detect$mixture,
function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
plot(cv_per_mix_det, main = "CV of Detected BPC per Mixture")
```

### Total Ion Current (TIC) area

```{r load-tic-detect}
load("object/tic_detect.RData")
tic_detect <- tic_detect[tic_detect$polarity == pol]
```

We summarize the TIC area per mixture per lab by summing the intensity values.

Inference from Detected TIC
This represents the sum of the areas of all detected peaks. It is the
best measure of total quantifiable signal. Unlike the "full" TIC, this
excludes the baseline. Labs with higher detected TIC are capturing more total
analyte signal, again pointing to higher sensitivity or better peak detection.

```{r int-tic-detect}
stic <- vapply(intensity(tic_detect), sum, numeric(1), na.rm = TRUE)

plot_data_tic_det <- data.frame(
  Lab = tic_detect$lab,
  Mixture = tic_detect$mixture,
  TIC_Area = stic
)

ggplot(plot_data_tic_det, aes(x = Lab, y = log2(TIC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Detected TIC Area Across Laboratories (Positive Mode)",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "Log2(Detected TIC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### ratio BPC / TIC

Now that we should have gotten rid of baseline and noise signal, the ratio
should be properly describing the complexity of the detected signal.

Inference from Detected BPC/TIC Ratio:

This measures the **complexity of the detected features**.

A ratio closer to 1 would imply that the detected peaks
are "pure" (BPC $\approx$ TIC).

A low ratio means that even
within the detected peak boundaries, the signal is spread across many ions.
This could be due to co-elution, fragmentation, adducts, or the peak-picking
algorithm including noisy parts of the peak.

Comparing labs, afekta seems to have a lower ratio, suggesting their detected
peaks are more complex(more co-elution/fragments) than other labs.

```{r ratio-detect-across}
ratio_bpc_tic <- sbpc / stic

plot_data_ratio_det <- data.frame(
  Lab = tic_detect$lab,
  Mixture = tic_detect$mixture,
  BPC_TIC_Ratio = ratio_bpc_tic
)

ggplot(plot_data_ratio_det, aes(x = Lab, y = BPC_TIC_Ratio, fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Detected Data BPC/TIC Ratio Across Laboratories (Positive Mode)",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "BPC/TIC Ratio (Detected Signal)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Tic/Bpc ratio along rt

Gives us some kind of chromatographic fingerprint per lab.
Instead of baseline shape, you’re summarizing where peaks are detected along
the gradient.

Inference from Detected BPC/TIC Ratio along RT

This is a "fingerprint of detected signal."

BPC/TIC Ratio Plot: This shows where in the gradient the detected signal is
"clean" (high ratio) or "complex" (low ratio).


```{r ratio-detect, eval = FALSE, fig.width=12, fig.height=9}
# Preallocate containers
ratio_all <- sapply(seq_along(tic_detect), function(i) {
  rt <- rtime(tic_detect[i])[[1]]
  int_tic <- intensity(tic_detect[i])[[1]]
  int_bpc <- intensity(bpc_detect[i])[[1]]

  compute_bin_ratio(rt, int_tic, int_bpc, n_bins = 30)
})

coverage_all <- sapply(seq_along(tic_detect), function(i) {
  rt <- rtime(tic_detect[i])[[1]]
  compute_bin_count(rt, n_bins = 30) ## could adjust this to lower nb of bins
})

# Assign names
colnames(ratio_all) <- paste(tic_detect$lab, tic_detect$mixture, "pol:",
 tic_detect$polarity, sep = "_")
colnames(coverage_all) <- colnames(ratio_all)

ratio_df <- as.data.frame(ratio_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "BPC/TIC Ratio")

coverage_df <- as.data.frame(coverage_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "Data Point Coverage")

sample_info <- data.frame(
  sample_id = colnames(ratio_all),
  lab = tic_detect$lab,
  mixture = tic_detect$mixture
)

plot_data <- bind_rows(ratio_df, coverage_df) %>%
  left_join(sample_info, by = "sample_id")

## plot
plot_data$metric <- factor(plot_data$metric,
                           levels = c("BPC/TIC Ratio", "Data Point Coverage"))

ggplot(plot_data, aes(x = rt_bin, y = value, color = lab, group = sample_id)) +
  geom_line(alpha = 0.7, linewidth = 1) +
  facet_grid(metric ~ mixture, scales = "free_y") +
  labs(
    title = "Chromatographic Fingerprint Comparison (Detected Data - Positive Mode)",
    subtitle = "BPC/TIC Ratio and Data Point Coverage by Lab and Mixture",
    x = "Retention Time Bin",
    y = "Value (Ratio or Count)",
    color = "Laboratory"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12)
  )

## for later similarity analysis i need to remove NA
ratio_all[is.na(ratio_all)] <- 0
```


## Chromatogram  Similarity

### ratio along retention time.

Inference from Detected Ratio DTW

This is a novel metric. It compares the "complexity fingerprint" of only the
detected signal. Clustering here would mean that labs have a similar pattern
of complexity for the peaks they detect across the gradient. This is a
high-level comparison of the quality of the detected features.

```{r, dtw-ratio, eval =eval_sim, fig.width=10, fig.height=10}
library(pheatmap)
library(proxy)
library(dplyr)
library(tidyr)

# Transpose ratio_all so samples are rows
sim_matrix_med <- as.matrix(proxy::dist(t(ratio_all),
                                        method = "dtw_lb",
                                        window.type = "sakoechiba",
                                        window.size = 5))
rownames(sim_matrix_med) <- colnames(ratio_all)
colnames(sim_matrix_med) <- colnames(ratio_all)

annotation_df <- data.frame(
  Lab = tic_detect$lab,
  Mixture = tic_detect$mixture
)
rownames(annotation_df) <- colnames(ratio_all) # CRITICAL: names must match


pheatmap(
  sim_matrix_med,
  main = "DTW Similarity of BPC/TIC Ratio (Detected Signal)",

  # Add the annotation bars
  annotation_row = annotation_df,
  annotation_col = annotation_df,

  # Hide the unreadable text labels
  show_rownames = FALSE,
  show_colnames = FALSE,

  # Improve clustering and color
  clustering_distance_rows = as.dist(sim_matrix_med),
  clustering_distance_cols = as.dist(sim_matrix_med),
  clustering_method = "ward.D2",
  color = hcl.colors(50, "YlOrRd", rev = TRUE)
)
```

!!! adding a column of color by lab would be nice.


```{r , rm-detect}
## delete the memory hungry object
rm(tic_detect, bpc_detect, sp_full_detect)
```

# Annotated signal

Here we will ONLY keep the annotated signal.

```{r an_signal_sp}
load("object/sp_full_ann.RData")
sp_full_ann <- sp_full_ann[sp_full_ann$polarity == pol]
```

```{r, rm-ann}
rm(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
```

Summarize the number of MS1 spectra per lab:

```{r nb-spectra-ann}
table(sp_full_ann$lab)
```

Summarize the rtime range per lab (in seconds):

```{r rt-range-ann}
lapply(split(rtime(sp_full_ann), sp_full_ann$lab) , range)
```

## Basic metrics

### Base Peak Chromatogram (BPC) area

```{r load-bpc-ann}
load("object/bpc_ann.RData")
bpc_ann <- bpc_ann[bpc_ann$polarity == pol]
```

We summarize the BPC area per mixture per lab by summing the intensity values.

Inference from Annotated BPC:

This measures the sum of peak heights (apices) for only the annotated
standards. This is a direct measure of quantitative sensitivity for the
target compounds. The high variability shows that for the same compounds,
some labs get much stronger signals than others. This is a key finding for an
inter-lab comparison.

```{r sum-bpc-ann}
sbpc <- vapply(intensity(bpc_ann), sum, numeric(1))

plot_data_bpc_ann <- data.frame(
  Lab = bpc_ann$lab,
  Mixture = bpc_ann$mixture,
  BPC_Area = sbpc
)

ggplot(plot_data_bpc_ann, aes(x = Lab, y = log2(BPC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Annotated BPC Area Across Laboratories (Positive Mode)",
    subtitle = "Sum of annotated peak heights. Points colored by mixture type.",
    x = "Laboratory",
    y = "Log2(Annotated BPC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Next, we compute the coefficient of variation (CV) per mixture across labs.

```{r cv-ann}
cv_per_mix_ann <- tapply(sbpc, bpc_ann$mixture, function(x)
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
plot(cv_per_mix_ann, main = "CV of Annotated BPC per Mixture")
```

### Total Ion Current (TIC) area

```{r load-tic-ann}
load("object/tic_ann.RData")
tic_ann <- tic_ann[tic_ann$polarity == pol]
```

We summarize the TIC area per mixture per lab by summing the intensity values.

This measures the sum of peak areas for only the annotated standards. This is
the best metric for overall quantitative agreement.

```{r int-tic-ann}
stic <- vapply(intensity(tic_ann), sum, numeric(1))

plot_data_tic_ann <- data.frame(
  Lab = tic_ann$lab,
  Mixture = tic_ann$mixture,
  TIC_Area = stic
)

ggplot(plot_data_tic_ann, aes(x = Lab, y = log2(TIC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Annotated TIC Area Across Laboratories (Positive Mode)",
    subtitle = "Sum of annotated peak areas. Points colored by mixture type.",
    x = "Laboratory",
    y = "Log2(Annotated TIC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### ratio BPC / TIC

Now we get the ratio for the spiked compound that were detected.

Inference from Annotated BPC/TIC Ratio
This is a critical metric for peak purity. This ratio (Apex Height / Peak Area,
scaled) for known compounds tells us how "clean" those peaks are.

A high ratio suggests a sharp, pure peak where the apex dominates the area.

A low ratio suggests a broader, tailing, or co-eluting peak, where the apex
intensity is low relative to the total peak area.


```{r ratio-ann-across}
ratio_bpc_tic <- sbpc / stic

plot_data_ratio_ann <- data.frame(
  Lab = tic_ann$lab,
  Mixture = tic_ann$mixture,
  BPC_TIC_Ratio = ratio_bpc_tic
)

ggplot(plot_data_ratio_ann, aes(x = Lab, y = BPC_TIC_Ratio, fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Annotated Data BPC/TIC Ratio Across Laboratories (Positive Mode)",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "BPC/TIC Ratio (Annotated Signal)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Chromatogram  Similarity

This compares the "purity fingerprint" (from the binned BPC/TIC ratio plot)
between labs. Labs that cluster here have a similar pattern of peak purity
across the gradient for the annotated standards.

### ratio along retention time.

```{r ratio-sim , eval=eval_sim}
sim_matrix_med <- as.matrix(proxy::dist(t(ratio_all),
                                        method = "dtw_lb",
                                        window.type = "sakoechiba",
                                        window.size = 5))

labels <- colnames(ratio_all)
rownames(sim_matrix_med) <- labels
colnames(sim_matrix_med) <- labels

# clustering
hc_med <- hclust(as.dist(sim_matrix_med))
plot(hc_med, labels = labels, main = "DTW clustering of BPC/TIC ratio (Annotated)")

# heatmap
heatmap(sim_matrix_med, labRow = labels, labCol = labels,
        main = "DTW similarity of BPC/TIC ratio (Annotated)")
```

Plot summary plots:

```{r}
png("figure/tic_comparison_plot_pos.png", width = 10, height = 6,
    units = "in", res = 300)
plot_signal_comparison("TIC", 1)
dev.off()

png("figure/bpc_comparison_plot_pos.png", width = 10, height = 6,
    units = "in", res = 300)
plot_signal_comparison("BPC", 1)
dev.off()
```


# Background/other signal

Here i need to load previous object and generate summary of "noise"

-   total signal - all detected peak signal

This value (dif_int2) represents the total baseline signal
(chemical and electronic noise) that was excluded by the peak picking
algorithm. A lab with a high value here (like 'afekta') has a much "noisier"
baseline that was successfully removed.

```{r noise-all}
load("object/tic_full.RData")
tic_full <- tic_full[tic_full$polarity == pol]
load("object/tic_detect.RData")
tic_detect <- tic_detect[tic_detect$polarity == pol]

stic_full <- vapply(intensity(tic_full), sum, numeric(1))
stic_detect <- vapply(intensity(tic_detect), sum, numeric(1))

bg <- stic_full - stic_detect
ratio <- stic_detect/ bg

ggplot(data.frame(Lab = tic_full$lab, Ratio =log2(ratio)), aes(x = Lab, y = Ratio, fill = Lab)
) +
  geom_violin(alpha = 0.5, trim = TRUE) +
  labs(
    title = "Detected vs Background TIC Area Ratio Across Laboratories (Positive Mode)",
    x = "Laboratory",
    y = "Detected / Background TIC Area Rati (log2)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

-   total signal - annotated peak signal

This value (dif_int3) represents all signal except the annotated standards.
This includes the baseline noise and all the unannotated peaks.

```{r noise-ann}
load("object/tic_ann.RData")
tic_ann <- tic_ann[tic_ann$polarity == pol]
stic_ann <- vapply(intensity(tic_ann), sum, numeric(1))
dif_int3 <- stic_full - stic_ann
ratio <- stic_ann/ dif_int3
ggplot(data.frame(Lab = tic_full$lab, Ratio =log2(ratio)), aes(x = Lab, y = Ratio, fill = Lab)
) +
  geom_violin(alpha = 0.5, trim = TRUE) +
  labs(
    title = "Annotated vs Other TIC Area Ratio Across Laboratories (Positive Mode)",
    x = "Laboratory",
    y = "Annotated / Other TIC Area Ratio (log2)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

- detected - annotated peak signal

This (dif_int4) is a very insightful metric for my thesis. It represents the
total signal from unannotated peaks. This is the "unknown" but "real" chemical
signal that was detected.

A lab with a high value (like 'icl') is detecting more "other" compounds.
This could be due to:
Higher sensitivity, allowing it to see more low-level unknown compounds.
Different fragmentation/adduct formation, meaning it's detecting more variants
of the annotated standards, which are not themselves annotated.

A lab with a low value (like 'hmgu') is detecting less signal outside of the
annotated standards.

```{r noise-detect-ann}
ratio <- stic_ann/ (stic_detect - stic_ann)
ggplot(data.frame(Lab = tic_full$lab, Ratio =log2(ratio)),
       aes(x = Lab, y = Ratio, fill = Lab)
) +
  geom_violin(alpha = 0.5, trim = TRUE) +
  labs(
    title = "Annotated vs Other Detected TIC Area Ratio Across Laboratories (Positive Mode)",
    x = "Laboratory",
    y = "Annotated / Other Detected TIC Area Ratio (log2)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


# Peak Analysis

Peak feature calculation and analysis is performed in `peak_feature_calculation.qmd`.

Load the peak report:

```{r load-peak-report}
load("object/full_peak_report.RData")
```

Summary of peak features by lab:

```{r peak-feature-summary}
cat("Number of peaks per lab:\n")
table(full_peak_report$lab)

cat("\nEntropy summary by lab:\n")
split(full_peak_report$entropy, full_peak_report$lab) |>
  lapply(summary)

cat("\nFWHM summary by lab:\n")
split(full_peak_report$fwhm, full_peak_report$lab) |>
  lapply(summary)
```

See `peak_feature_calculation.qmd` for detailed peak feature analysis including:

- EIC extraction and preprocessing
- Peak metric calculation (entropy, FWHM, tailing factor, etc.)
- Retention time stability analysis
- Quantitative robustness assessment
- PCA on peak features

Next steps for peak similarity analysis:

- Mass accuracy check
- Blind clustering to group adducts
- Similarity comparison using zeros grouping, RT scan order differences
