---
title: "Global Analysis: Inter-Laboratory LC-MS Reproducibility"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
---

# Introduction

This analysis evaluates QC metrics across 4 laboratories using positive
and negative ionization modes. Instrumental setup, sample preparation, and
data processing all affect chromatographic metrics on both full and
detected signal data.

Data: 19 standard mixtures (HUMAN endosome, different compounds) from 4
labs with harmonized LC systems.

Limitations:

- No technical replicates
- Laboratory identity is confounded with instrumental parameters.

=> Observed differences reflect combined effects
of lab environment, operator expertise, and equipment configuration.

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)
library(pheatmap)

base_path <- here("5_downstream_analysis")
source(file.path(base_path, "global_analysis_helpers.R"))
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`, `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) %>%
  distinct() %>%
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) %>%
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

Each lab's protocol is unique; lab identity is completely confounded with
instrumental setup. Notable variations:

- MS1 scan rate
- Injection volume
- Solvent brand.

## Metrics Definition

Three main chromatographic metrics:

- **TIC**: Sum of all ion intensities
- **BPC**: Max intensity per scan

Two data levels:

- **Full Signal**: Raw profile including baseline
- **Detected Signal**: Peaks from centWave; isolates compound signal from
  baseline

Two polarities: Positive and Negative ionization modes.

```{r load-data-both-modes}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "object/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]
    load(file.path(base_path, "object/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "object/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]
    load(file.path(base_path, "object/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = ifelse(data_level == "full", "Full Signal", "Detected Signal")) %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm = bpc / median_bpc,
      tic_norm = tic / median_tic
    ) %>%
    ungroup()
}

run_for_polarities_names <- function(fn) {
  bind_rows(fn(1L, "Positive"), fn(0L, "Negative"))
}

metrics_full <- run_for_polarities(compute_metrics, "full")
metrics_detect <- run_for_polarities(compute_metrics, "detect")

metrics_combined <- bind_rows(metrics_full, metrics_detect) %>%
  group_by(lab, polarity, data_level) %>%
  mutate(
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE)
  ) %>%
  ungroup()
```

---

# Phase 1: The Problem — Raw Data Assessment {#phase1}

This phase characterizes the raw data to understand the magnitude of
inter-laboratory variability before any corrections are applied.

## 1.1 Distribution of Raw Metrics

### 1.1.1 Violin Plots: Overall Intensity Distribution

```{r violin-plots}
#| fig-cap: "BPC and TIC distributions by lab, polarity, and data level."
for (metric in c("bpc", "tic")) print(plot_violin(metrics_combined, metric))
```

Detected signal recenters the negative and positive distributions.
The bimodal shape is evident.

### 1.1.2 Density Plots: Distribution Shape by Lab

```{r density-plots}
ggplot(metrics_combined %>% filter(data_level == "Full Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_minimal() +
  labs(title = "TIC Distribution by Lab (Full Signal)",
       x = "log2(TIC)", y = "Density")

ggplot(metrics_combined %>% filter(data_level == "Detected Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_minimal() +
  labs(title = "TIC Distribution by Lab (Detected Signal)",
       x = "log2(TIC)", y = "Density")
```

Full signal is more compressed than detected signal, likely because
each mixture has different numbers of compounds.

### 1.1.3 Mixture-Level Agreement Across Labs

```{r mixture-distribution}
plot_mixture_distribution <- function(data, metric) {
  if (metric %in% c("bpc", "tic")) {
    data <- data |> mutate(across(all_of(metric), log2))
    title_suffix <- "(log2)"
  } else {
    title_suffix <- ""
  }

  ggplot(data, aes(x = mixture, y = .data[[metric]], color = lab, group = lab)) +
    geom_line(alpha = 0.7) +
    facet_grid(data_level ~ polarity) +
    theme_minimal() +
    labs(title = paste("Mixture Distribution Across Labs -", metric, title_suffix),
         x = "Mixture", y = paste(metric, title_suffix), color = "Laboratory")
}

for (metric in c("bpc", "tic")) {
  print(plot_mixture_distribution(metrics_combined, metric))
}
```

There is good agreement between Cembio and Afekta, and some with ICL.
However, HMGU is consistently lower intensity.

## 1.2 Variance Decomposition: Quantifying the Problem

### 1.2.1 Linear Model Analysis

Let's run linear models to evaluate mixture effects and batch effects.
The goal is to quantify how much of the variance is due to lab vs mixture.
We use ANOVA to aggregate coefficients into Sum of Squares (SS), which
quantifies the relative power of each factor.

We run this analysis across all 4 conditions (positive/negative, full/detected):

```{r variance-decomposition-lm}
results_table <- data.frame()
residual_plots <- list()
all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")

    residual_plots[[paste(pol, level, sep = " - ")]] <- result$residual_plot
    all_coefficients <- rbind(all_coefficients,
      result$coefficients %>% mutate(Polarity = pol, Data_Level = level))
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

### 1.2.2 Lab Effect Coefficients

```{r lab-coefficients}
#| fig-cap: "Lab effect coefficients from linear models. Deviation from reference lab (Afekta); negative = lower intensity."
lab_coefs <- all_coefficients %>%
  filter(type == "Lab Effect") %>%
  mutate(lab_name = gsub("lab", "", coefficient))

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "Lab Effect Coefficients from Linear Models",
       subtitle = "Deviation from reference lab (afekta); negative = lower intensity",
       x = "Laboratory", y = "Coefficient (log2 scale)", fill = "Laboratory") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.2.3 Variance Decomposition Summary

```{r variance-stacked-bar}
#| fig-cap: "Variance decomposition of TIC across all conditions. Lab variance dominates in raw data."
variance_long <- results_table %>%
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) %>%
  pivot_longer(cols = ends_with("_Pct"), names_to = "Source", values_to = "Percentage") %>%
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Label = paste(Polarity, Data_Level, sep = "\n"),
    R2_label = paste("R² =", round(R_squared, 3))
  )

text_labels <- variance_long %>% distinct(Label, R2_label)

ggplot(variance_long, aes(x = Label, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(x = Label, label = R2_label, y = 105), data = text_labels,
            size = 3, inherit.aes = FALSE) +
  labs(title = "Variance Decomposition: TIC Across All Conditions",
       subtitle = "Raw data shows massive lab variance — the core problem",
       x = "Polarity & Data Level", y = "Variance Contribution (%)", fill = "Source") +
  theme_minimal() +
  theme(legend.position = "top") +
  ylim(0, 110)
```

### 1.2.4 Residual Diagnostics

```{r residual-plots}
#| fig-height: 10
residual_data <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    residual_data <- rbind(residual_data, data.frame(
      Fitted = fitted(result$model), Residuals = residuals(result$model),
      Polarity = pol, Data_Level = level,
      Lab = data_subset$lab, Mixture = as.factor(data_subset$mixture)
    ))
  }
}

ggplot(residual_data, aes(x = Fitted, y = Residuals, color = Mixture, shape = Lab)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ paste(Polarity, Data_Level, sep = "\n"), nrow = 2, ncol = 2) +
  scale_color_viridis_d(option = "turbo") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(title = "Fitted vs Residuals: TIC Across All Conditions",
       subtitle = "Points should scatter randomly around zero for a good model fit",
       x = "Fitted Values", y = "Residuals", color = "Mixture", shape = "Laboratory") +
  theme_minimal() +
  theme(legend.position = "bottom")

results_table
```

**Key Finding:** Overall variance is dominated by lab effects.
The residual plot clearly shows separation between labs vs HMGU and
still lab separation in others. Lab separation remains unexplained by
mixture. We cannot compare labs directly without normalization.

---

# Phase 2: The Solution — Normalization & Data Processing {#phase2}

This phase tests different normalization strategies to reduce
inter-laboratory variability.

## 2.1 Normalization Strategy Selection

### 2.1.1 Rationale for Testing Different Approaches

What could be tested:

- **Median scaling**: Adjusting each lab's data to have the same median could
 help account for systematic differences. In that case we take the median of
 that measure by that lab to correct each mixture.
    - This is safe as the experimental design is balanced. Every lab measures
     the same 38 mixtures, so their intensity distributions should be identical in
     shape, just shifted up or down.
    - This would be wrong if the mixtures were different across labs.
- **Auto scaling**: Forces the lab to have the same dynamic change.
    - Stronger scaling. Will remove any instrument bias.

What we are NOT testing:

- **Max scaling**: ESI is subject to random electrical spike/detector saturation,
       therefore max could be an outlier and crush other signal to 0.
    - Generally not used for entire run.

Decision criteria based on violin plot shape:

- If similar: median scaling should be fine
- If not: auto-scaling may be required

The violin plot and density plot above are a bit hard to be definitive
about. The positive data is fairly similar in shape, but the negative data
shows more variation. We test both and see how the models are affected.

### 2.1.2 Visual Comparison: Before vs After Scaling

```{r scaling-comparison-visual}
plot_scaling_comparison <- function(data, metric_raw, metric_scaled, metric_label) {
  plot_data <- data %>%
    select(lab, polarity, data_level, mixture, raw = all_of(metric_raw), scaled = all_of(metric_scaled)) %>%
    pivot_longer(cols = c(raw, scaled), names_to = "Scaling", values_to = "Value") %>%
    mutate(Scaling = ifelse(Scaling == "raw", "Unscaled", "Median Scaled"))

  ggplot(plot_data, aes(x = lab, y = log2(Value), fill = polarity)) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    geom_jitter(aes(color = as.factor(mixture)), width = 0.2, size = 1.5, alpha = 0.7) +
    scale_color_viridis_d(option = "turbo") +
    facet_grid(Scaling ~ data_level, scales = "free_y") +
    theme_minimal() +
    labs(title = paste("Before vs After Median Scaling:", metric_label),
         subtitle = "Distribution shape should remain similar, but scale shifts",
         x = "Laboratory", y = paste(metric_label, "(log2)"), color = "Mixture") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

print(plot_scaling_comparison(metrics_combined, "tic", "tic_median_scaled", "TIC"))
print(plot_scaling_comparison(metrics_combined, "bpc", "bpc_median_scaled", "BPC"))
```

In the detected signal we seem to be keeping some dynamic range, but quite hard
to see on the full signal.

## 2.2 Normalization Impact on Variance

### 2.2.1 Variance Decomposition by Normalization Method

```{r normalization-variance-comparison}
#| fig-cap: "Effect of normalization on variance decomposition for positive mode detected signal."
test_data <- metrics_combined %>%
  filter(polarity == "Positive", data_level == "Detected Signal")

result_raw <- check_metric_quality(test_data, "tic")
result_median_scaled <- check_metric_quality(test_data, "tic_median_scaled")
result_autoscaled <- check_metric_quality(test_data, "tic_autoscaled")

normalization_results <- bind_rows(
  result_raw$variance_table %>% mutate(Normalization = "Raw TIC", R_squared = result_raw$r_squared),
  result_median_scaled$variance_table %>% mutate(Normalization = "Median Scaled", R_squared = result_median_scaled$r_squared),
  result_autoscaled$variance_table %>% mutate(Normalization = "Autoscaled (MAD)", R_squared = result_autoscaled$r_squared)
)

normalization_results %>%
  ggplot(aes(x = Normalization, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  labs(title = "Effect of Normalization on Variance Decomposition (TIC)",
       subtitle = "Positive Mode, Detected Signal",
       x = "Normalization Method", y = "Variance Contribution (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 2.2.2 Absolute Noise Comparison

```{r normalization-noise-comparison}
mod_raw <- lm(log2(tic) ~ lab + factor(mixture), data = test_data)
mod_median_scaled <- lm(log2(tic_median_scaled) ~ lab + factor(mixture), data = test_data)
mod_autoscaled <- lm(tic_autoscaled ~ lab + factor(mixture), data = test_data)

ss_raw <- anova(mod_raw)$"Sum Sq"
ss_median_scaled <- anova(mod_median_scaled)$"Sum Sq"
ss_autoscaled <- anova(mod_autoscaled)$"Sum Sq"

cat("Raw Absolute Noise:", round(ss_raw[3], 2), "\n")
cat("Median Scaled Absolute Noise:", round(ss_median_scaled[3], 2), "\n")
cat("Autoscaled Absolute Noise:", round(ss_autoscaled[3], 2), "\n")
cat("\nRaw Total Variance:", round(sum(ss_raw), 2), "\n")
cat("Median Scaled Total Variance:", round(sum(ss_median_scaled), 2), "\n")
cat("Autoscaled Total Variance:", round(sum(ss_autoscaled), 2), "\n")
```

### 2.2.3 Residual Plots by Normalization Method

```{r normalization-residual-comparison}
#| fig-height: 14
plot_residuals(mod_raw, "Raw TIC (log2)", test_data) /
  plot_residuals(mod_median_scaled, "Median Scaled TIC", test_data) /
  plot_residuals(mod_autoscaled, "Autoscaled TIC (MAD)", test_data) +
  plot_annotation(title = "Residual Plots: Comparison of Normalization Methods")
```

**Findings:**

- Median scaling removes the residual pattern we could see before.
- Autoscaling, in contrast, seems to introduce some pattern in the residuals.
- Lab variance drops significantly with median scaling.
- Absolute noise values do not change; their overall proportion increases because we got rid of the lab variance.

Noise likely reflects injection variability and ionization
fluctuation—factors we cannot isolate with this experimental design.

**Conclusions:**

- Use median scaling
- Noise level acceptable; does not significantly impact biological variance
- We "hope" to focus on biological signals without technical variability obscuring them

---

# Phase 3: Reproducibility & Platform Effects {#phase3}

This phase quantifies reproducibility using standard metrics (ICC) and
investigates platform-specific effects.

## 3.1 Intraclass Correlation Coefficient (ICC)

### 3.1.1 Understanding ICC

Normalization addresses sensitivity differences across labs, but
lab-dependent noise remains. The intraclass correlation coefficient (ICC)
tests this. ICC repeats earlier variance analysis but provides scale-
invariant comparability across papers—unlike raw Sum of Squares, it's
more meaningful for visualization.

**ICC Formula:**
$$ICC = \frac{\sigma^2_{between}}{\sigma^2_{between} + \sigma^2_{within}}$$

where:

- $\sigma^2_{between}$ = variance in metric values **between** different mixtures (biological signal)
- $\sigma^2_{within}$ = variance in metric values **within** the same mixture across labs (technical noise)

**Interpretation in our context:**

- **High ICC (→1)**: Most variance comes from mixture differences (biology). Labs agree on which mixtures have high/low signal.
- **Low ICC (→0)**: Most variance comes from lab-to-lab differences. Technical noise obscures biological signal.
- **ICC = 0.5**: Equal biological and technical variance (threshold for acceptable reproducibility).

High ICC confirms measurements reflect biological differences, not
technical noise.

### 3.1.2 ICC Computation: Absolute vs Normalized Metrics

```{r icc-computation}
icc_all_results <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)

    results <- bind_rows(
      compute_icc(data_subset, "tic") %>% mutate(Type = "Absolute"),
      compute_icc(data_subset, "bpc") %>% mutate(Type = "Absolute"),
      compute_icc(data_subset, "tic_norm") %>% mutate(Type = "Normalized", Metric = "tic"),
      compute_icc(data_subset, "bpc_norm") %>% mutate(Type = "Normalized", Metric = "bpc")
    ) %>%
      mutate(Polarity = pol, Data_Level = level,
             Metric_Label = case_when(Metric == "tic" ~ "TIC", Metric == "bpc" ~ "BPC", TRUE ~ Metric))

    icc_all_results <- bind_rows(icc_all_results, results)
  }
}

icc_all_results %>%
  select(Polarity, Data_Level, Metric_Label, Type, ICC) %>%
  arrange(desc(ICC))
```

### 3.1.3 ICC Comparison Plot

```{r icc-comparison-plot}
#| fig-cap: "ICC comparison: absolute vs normalized metrics by polarity and data level. Dashed line = 0.5 threshold."
ggplot(icc_all_results, aes(x = Metric_Label, y = ICC, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  facet_grid(Data_Level ~ Polarity) +
  labs(title = "Reproducibility Comparison: Absolute vs Normalized Metrics",
       subtitle = "ICC by Polarity and Data Level; dashed line = 0.5 threshold",
       x = "Metric", y = "Intraclass Correlation Coefficient (ICC)", fill = "Metric Type") +
  theme_minimal() +
  theme(legend.position = "top")
```

### 3.1.4 ICC Variance Decomposition

```{r icc-variance-decomposition}
variance_plot_intensity <- icc_all_results %>%
  select(Polarity, Data_Level, Metric_Label, Type, Between_mixture_var, Within_mixture_var) %>%
  pivot_longer(cols = c(Between_mixture_var, Within_mixture_var),
               names_to = "Source", values_to = "Variance") %>%
  mutate(Source = case_when(
    Source == "Between_mixture_var" ~ "Between mixtures (signal)",
    Source == "Within_mixture_var" ~ "Within mixtures (lab noise)",
    TRUE ~ Source
  ))

ggplot(variance_plot_intensity, aes(x = Metric_Label, y = Variance, fill = Source)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.7) +
  facet_grid(Data_Level + Type ~ Polarity, scales = "free_y") +
  labs(title = "Variance Decomposition: TIC and BPC",
       subtitle = "Higher 'between mixtures' = more signal; Higher 'within mixtures' = more lab noise",
       x = "Metric", y = "Variance (log2 scale)", fill = "Source of Variation") +
  scale_fill_manual(values = c("Between mixtures (signal)" = "#619CFF",
                                "Within mixtures (lab noise)" = "#F8766D")) +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 30, hjust = 1))
```

The overall goal is that we end up with Value = Biology.

Starting point: Value = (Biology × Sensitivity) + Background Noise

With detected signal and normalization, we hope: Value ≈ Biology

## 3.2 Impact of Peak Detection on Reproducibility

### 3.2.1 Full Signal vs Detected Signal Comparison

```{r peak-detection-impact}
icc_comparison_tic <- icc_all_results %>%
  filter(Type == "Normalized", Metric == "tic") %>%
  select(Polarity, Data_Level, ICC) %>%
  pivot_wider(names_from = Data_Level, values_from = ICC) %>%
  mutate(
    ICC_Improvement = `Detected Signal` - `Full Signal`,
    Percent_Gain = (`Detected Signal` - `Full Signal`) / `Full Signal` * 100
  )

icc_comparison_tic
```

**Findings:** Normalized TIC shows highest ICC. Detected signal (ICC 0.50–0.56)
substantially beats full signal (ICC 0.27–0.36) because peak detection removes
vendor-specific solvent noise. Median normalization improved ICC from ~0.04
to ~0.53—crossing the threshold from poor to moderate reproducibility. The
remaining ~0.3 ICC gap relative to the Agilent ceiling (~0.84) likely reflects
ionization preferences and platform-specific sensitivities.

### 3.2.2 Before & After Visualization

```{r peak-detection-visualization}
#| fig-cap: "Impact of peak detection on ICC. Detected signal substantially improves reproducibility."
ggplot(icc_all_results %>% filter(Type == "Normalized", Metric == "tic"),
       aes(x = Polarity, y = ICC, fill = Data_Level)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(ICC, 2)), position = position_dodge(0.9), vjust = -0.5) +
  labs(title = "Impact of Peak Detection on Reproducibility",
       subtitle = "Comparison of Normalized TIC Stability (Full vs. Detected)",
       y = "Intraclass Correlation Coefficient (ICC)", x = "Polarity",
       fill = "Processing Level",
       caption = "Higher ICC = Better Agreement between labs.\nPeak detection doubles reproducibility in Negative Mode.") +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```

Peak detection substantially improves reproducibility:

- Negative mode: ICC increased from 0.27 to 0.56 (~105% improvement)
- Positive mode: ICC increased from 0.36 to 0.50 (~40% improvement)

**Conclusion:** All subsequent analyses use detected signal exclusively.
Peak detection removes instrument-specific baseline noise obscuring
biological signal. Full signal contains too much vendor-specific noise
for meaningful inter-lab comparisons.

## 3.3 Platform Effects: Agilent Labs Comparison

### 3.3.1 "Ceiling" Estimate: Same-Platform Labs

For comparison, two labs (CEMBIO and AFEKTA) share the same LC-MS
platform (Agilent Q-TOF). ICC of 0.83 (TIC) and 0.92 (BPC) for these labs
proves selectivity matches when vendor matches. The previously seen 0.5–0.8 gap might therefore reflects
instrument-specific and ionization differences.

```{r agilent-ceiling-icc}
agilent_data <- metrics_combined %>%
  filter(lab %in% c("afekta", "cembio"), polarity == "Positive", data_level == "Detected Signal")

icc_agilent_tic <- compute_icc(agilent_data, "tic_norm")
icc_agilent_bpc <- compute_icc(agilent_data, "bpc_norm")

cat("=== Ceiling ICC Estimate (Agilent Labs Only, Positive Mode, Detected Signal) ===\n")
cat("ICC for Normalized TIC:", round(icc_agilent_tic$ICC, 3), "\n")
cat("ICC for Normalized BPC:", round(icc_agilent_bpc$ICC, 3), "\n")
```

### 3.3.2 Effect Size Between Agilent Labs

**What is effect size?**

Effect size measures the *practical magnitude* of a difference between two
groups. Here, we compute log2 fold-changes between Cembio and Afekta for
each mixture (i.e., $\log_2(\text{Cembio} / \text{Afekta})$), then
summarize across all mixtures.

- A log2 fold-change of **0** means the labs produce identical values.
- A log2 fold-change of **±0.5** corresponds to a ~41% difference.
- A log2 fold-change of **±1** corresponds to a 2-fold difference.

**How does this complement ICC?**

ICC answers *"are measurements reproducible?"* — it captures the ratio of
biological signal to technical noise across all labs. Effect size answers
*"by how much do two specific labs disagree?"* — it gives a concrete,
directional estimate of the systematic bias between the two Agilent labs.

```{r agilent-effect-size}
agilent_effect_size <- metrics_combined %>%
  filter(lab %in% c("afekta", "cembio"), data_level == "Detected Signal") %>%
  select(lab, mixture, polarity, tic_norm, bpc_norm) %>%
  pivot_wider(names_from = lab, values_from = c(tic_norm, bpc_norm)) %>%
  mutate(
    log_diff_tic = log2(tic_norm_cembio / tic_norm_afekta),
    log_diff_bpc = log2(bpc_norm_cembio / bpc_norm_afekta)
  )

agilent_effect_size %>%
  group_by(polarity) %>%
  summarise(
    TIC_Mean_Log = mean(log_diff_tic, na.rm = TRUE),
    TIC_Median_Log = median(log_diff_tic, na.rm = TRUE),
    TIC_Percent_Diff = (2^abs(TIC_Mean_Log) - 1) * 100,
    BPC_Mean_Log = mean(log_diff_bpc, na.rm = TRUE),
    BPC_Median_Log = median(log_diff_bpc, na.rm = TRUE),
    BPC_Percent_Diff = (2^abs(BPC_Mean_Log) - 1) * 100,
    .groups = "drop"
  )
```

**Key findings:**

- **Positive mode**: Median log2 difference ≈ 0 (labs agree within ~1%); mean
  pulled by outliers (~6.6% difference)
- **Negative mode**: Consistent ~9% systematic bias that median normalization
  cannot fully remove

```{r agilent-effect-size-table}
effect_size_summary <- agilent_effect_size %>%
  group_by(polarity) %>%
  summarise(
    TIC_Mean_Log = mean(log_diff_tic, na.rm = TRUE),
    TIC_Median_Log = median(log_diff_tic, na.rm = TRUE),
    TIC_Percent_Diff = (2^abs(TIC_Mean_Log) - 1) * 100,
    BPC_Mean_Log = mean(log_diff_bpc, na.rm = TRUE),
    BPC_Median_Log = median(log_diff_bpc, na.rm = TRUE),
    BPC_Percent_Diff = (2^abs(BPC_Mean_Log) - 1) * 100,
    .groups = "drop"
  )

knitr::kable(effect_size_summary,
             caption = "Effect Size Summary: Cembio vs Afekta (log2 fold-change)",
             digits = 3)
```

The two Agilent Q-TOF generations (6546 vs 6545) can be compared for trends
(high ICC), but would need non-linear correction for absolute quantitation.

---

# Phase 4: Granular Nuance — Retention Time & Biology {#phase4}

This phase explores retention time-dependent effects and biological
signal variations across the chromatographic gradient.

## 4.1 Retention Time Range

All the labs had a retention time gradient of 15min. After preprocessing we
reduced to a common range:

```{r rt-range}
load(file.path(base_path, "object/tic_full.RData"))
range(rtime(tic_full[1L])[[1]])
```

How do the labs agree across the retention time gradient? Do some labs
show systematic sensitivity drops at early or late elution times?

## 4.2 RT-Binned Correlation Analysis

### 4.2.1 Full Signal: Pairwise Correlation by Mixture

```{r rt-correlation-full-signal}
load(file.path(base_path, "object/tic_full.RData"))
rt_range <- range(rtime(tic_full[1L])[[1]])
breaks <- seq(rt_range[1], rt_range[2], by = 5)

polarities <- unique(tic_full$polarity)
mixtures <- unique(tic_full$mixture)
correlation_results_full <- list()

for (pol in polarities) {
  for (mix in mixtures) {
    subset_tic <- tic_full[which(tic_full$polarity == pol & tic_full$mixture == mix)]
    if (length(subset_tic) == 0) next

    intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
      icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
      cembio = numeric(length(breaks) - 1))

    for (i in seq_along(subset_tic)) {
      lab_name <- subset_tic$lab[i]
      peaks <- peaksData(subset_tic[i])[[1]]
      intensity_matrix[[lab_name]] <- tapply(peaks$intensity,
        cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
    }

    cor_results <- cor(intensity_matrix, method = "pearson", use = "pairwise.complete.obs")
    polarity_label <- ifelse(pol == 1, "pos", "neg")
    correlation_results_full[[paste0(polarity_label, "_", mix)]] <- cor_results
  }
}

pairs <- list(c("afekta", "cembio"), c("afekta", "hmgu"), c("afekta", "icl"),
              c("cembio", "hmgu"), c("cembio", "icl"), c("hmgu", "icl"))

pairwise_cors_full <- data.frame()
for (result_name in names(correlation_results_full)) {
  cor_mat <- correlation_results_full[[result_name]]
  for (pair in pairs) {
    pairwise_cors_full <- rbind(pairwise_cors_full, data.frame(
      mixture = result_name, pair = paste(pair[1], pair[2], sep = ":"),
      correlation = cor_mat[pair[1], pair[2]]))
  }
}

cor_heatmap_data_full <- pairwise_cors_full |>
  separate(mixture, into = c("polarity", "mix"), sep = "_", extra = "merge") |>
  mutate(polarity = factor(polarity, levels = c("neg", "pos")))

ggplot(cor_heatmap_data_full, aes(x = pair, y = mix, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0.5, limits = c(0, 1)) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "TIC Correlation Heatmap Across All Mixtures (Full Signal)",
       x = "Lab Pair", y = "Mixture", fill = "Pearson r")
```

### 4.2.2 Full Signal: Correlation Distribution

```{r rt-correlation-full-distribution}
library(ggbeeswarm)

ggplot(cor_heatmap_data_full, aes(x = pair, y = correlation, color = pair)) +
  geom_violin(alpha = 0.3, show.legend = FALSE) +
  geom_beeswarm(size = 2, alpha = 0.7, show.legend = FALSE) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", show.legend = FALSE) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of TIC Correlations Across 19 Mixtures (Full Signal)",
       subtitle = "Each point represents one mixture", x = "Lab Pair", y = "Pearson Correlation")
```

### 4.2.3 Full Signal: RT-Dependent Correlation (Single Mixture Example)

```{r rt-dependent-correlation-full}
compute_rt_bin_correlation_full <- function(polarity, mixture) {
  subset_tic <- tic_full[which(tic_full$polarity == polarity & tic_full$mixture == mixture)]
  if (length(subset_tic) == 0) return(NULL)

  rt_range <- range(rtime(tic_full[1L])[[1]])
  breaks <- seq(rt_range[1], rt_range[2], by = 5)

  intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
    icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
    cembio = numeric(length(breaks) - 1))

  for (i in seq_along(subset_tic)) {
    lab_name <- subset_tic$lab[i]
    peaks <- peaksData(subset_tic[i])[[1]]
    intensity_matrix[[lab_name]] <- tapply(peaks$intensity, cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
  }

  rt_centers <- (breaks[-1] + breaks[-length(breaks)]) / 2
  window_size <- 10
  pairs <- list(c("afekta", "cembio"), c("afekta", "hmgu"), c("afekta", "icl"),
                c("cembio", "hmgu"), c("cembio", "icl"), c("hmgu", "icl"))
  bin_cors <- list()

  for (bin_idx in 1:nrow(intensity_matrix)) {
    start_idx <- max(1, bin_idx - window_size %/% 2)
    end_idx <- min(nrow(intensity_matrix), bin_idx + window_size %/% 2)
    window_data <- intensity_matrix[start_idx:end_idx, ]
    if (nrow(window_data) < 3) next
    cor_mat <- cor(window_data, method = "pearson", use = "pairwise.complete.obs")
    if (any(is.na(cor_mat))) next
    for (pair in pairs) {
      bin_cors[[length(bin_cors) + 1]] <- data.frame(
        rt_center = rt_centers[bin_idx], pair = paste(pair[1], pair[2], sep = ":"),
        correlation = cor_mat[pair[1], pair[2]])
    }
  }
  do.call(rbind, bin_cors)
}

rt_cor_example_full <- compute_rt_bin_correlation_full(1, "1_1")

if (!is.null(rt_cor_example_full)) {
  ggplot(rt_cor_example_full, aes(x = rt_center, y = pair, fill = correlation)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(title = "RT-Dependent Correlation: Positive Mode, Mixture 1_1 (Full Signal)",
         subtitle = "Using 5-second RT bins with 50-second sliding window",
         x = "Retention Time (s)", y = "Lab Pair", fill = "Pearson r")
}
```

Note: This is only one mixture and one polarity. The x-axis should start at the
actual RT range start (~17s).

### 4.2.4 Detected Signal: Pairwise Correlation by Mixture

```{r rt-correlation-detected-signal}
load(file.path(base_path, "object/tic_detect.RData"))
rt_range <- range(rtime(tic_detect[1L])[[1]])
breaks <- seq(rt_range[1], rt_range[2], by = 5)

polarities <- unique(tic_detect$polarity)
mixtures <- unique(tic_detect$mixture)
correlation_results <- list()

for (pol in polarities) {
  for (mix in mixtures) {
    subset_tic <- tic_detect[which(tic_detect$polarity == pol & tic_detect$mixture == mix)]
    if (length(subset_tic) == 0) next

    intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
      icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
      cembio = numeric(length(breaks) - 1))

    for (i in seq_along(subset_tic)) {
      lab_name <- subset_tic$lab[i]
      peaks <- peaksData(subset_tic[i])[[1]]
      intensity_matrix[[lab_name]] <- tapply(peaks$intensity, cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
    }

    cor_results <- cor(intensity_matrix, method = "pearson", use = "pairwise.complete.obs")
    polarity_label <- ifelse(pol == 1, "pos", "neg")
    correlation_results[[paste0(polarity_label, "_", mix)]] <- cor_results
  }
}

pairwise_cors <- data.frame()
for (result_name in names(correlation_results)) {
  cor_mat <- correlation_results[[result_name]]
  for (pair in pairs) {
    pairwise_cors <- rbind(pairwise_cors, data.frame(
      mixture = result_name, pair = paste(pair[1], pair[2], sep = ":"),
      correlation = cor_mat[pair[1], pair[2]]))
  }
}

cor_heatmap_data <- pairwise_cors |>
  separate(mixture, into = c("polarity", "mix"), sep = "_", extra = "merge") |>
  mutate(polarity = factor(polarity, levels = c("neg", "pos")))

ggplot(cor_heatmap_data, aes(x = pair, y = mix, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0.5, limits = c(0, 1)) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "TIC Correlation Heatmap Across All Mixtures (Detected Signal)",
       x = "Lab Pair", y = "Mixture", fill = "Pearson r")
```

### 4.2.5 Detected Signal: Correlation Distribution

```{r rt-correlation-detected-distribution}
ggplot(cor_heatmap_data, aes(x = pair, y = correlation, color = pair)) +
  geom_violin(alpha = 0.3, show.legend = FALSE) +
  geom_beeswarm(size = 2, alpha = 0.7, show.legend = FALSE) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", show.legend = FALSE) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of TIC Correlations Across 19 Mixtures (Detected Signal)",
       subtitle = "Each point represents one mixture", x = "Lab Pair", y = "Pearson Correlation")
```

Note: How can some correlations be negative?

### 4.2.6 Detected Signal: RT-Dependent Correlation (Single Mixture Example)

```{r rt-dependent-correlation-detected}
compute_rt_bin_correlation <- function(polarity, mixture) {
  subset_tic <- tic_detect[which(tic_detect$polarity == polarity & tic_detect$mixture == mixture)]
  if (length(subset_tic) == 0) return(NULL)

  rt_range <- range(rtime(tic_detect[1L])[[1]])
  breaks <- seq(rt_range[1], rt_range[2], by = 5)

  intensity_matrix <- data.frame(afekta = numeric(length(breaks) - 1),
    icl = numeric(length(breaks) - 1), hmgu = numeric(length(breaks) - 1),
    cembio = numeric(length(breaks) - 1))

  for (i in seq_along(subset_tic)) {
    lab_name <- subset_tic$lab[i]
    peaks <- peaksData(subset_tic[i])[[1]]
    intensity_matrix[[lab_name]] <- tapply(peaks$intensity, cut(peaks$rtime, breaks = breaks), sum, na.rm = TRUE)
  }

  rt_centers <- (breaks[-1] + breaks[-length(breaks)]) / 2
  window_size <- 10
  bin_cors <- list()

  for (bin_idx in 1:nrow(intensity_matrix)) {
    start_idx <- max(1, bin_idx - window_size %/% 2)
    end_idx <- min(nrow(intensity_matrix), bin_idx + window_size %/% 2)
    window_data <- intensity_matrix[start_idx:end_idx, ]
    if (nrow(window_data) < 3) next
    cor_mat <- cor(window_data, method = "pearson", use = "pairwise.complete.obs")
    if (any(is.na(cor_mat))) next
    for (pair in pairs) {
      bin_cors[[length(bin_cors) + 1]] <- data.frame(
        rt_center = rt_centers[bin_idx], pair = paste(pair[1], pair[2], sep = ":"),
        correlation = cor_mat[pair[1], pair[2]])
    }
  }
  do.call(rbind, bin_cors)
}

rt_cor_example <- compute_rt_bin_correlation(1, "1_1")

if (!is.null(rt_cor_example)) {
  ggplot(rt_cor_example, aes(x = rt_center, y = pair, fill = correlation)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5, limits = c(0, 1)) +
    theme_minimal() +
    labs(title = "RT-Dependent Correlation: Positive Mode, Mixture 1_1 (Detected Signal)",
         subtitle = "Using 5-second RT bins with 50-second sliding window",
         x = "Retention Time (s)", y = "Lab Pair", fill = "Pearson r")
}
```

### 4.2.7 Comparison: Full vs Detected Signal Correlations

This plot directly compares the pairwise TIC correlation distributions from
the full signal and detected signal analyses above. For each lab pair, the
boxplots show the spread of Pearson correlations across all 19 mixtures. If
peak detection successfully removes instrument-specific baseline noise, we
expect detected signal correlations to be higher and less variable.

```{r full-vs-detected-correlation}
#| fig-cap: "Pairwise TIC correlations across 19 mixtures: full signal vs detected signal. Higher medians and tighter distributions in the detected signal indicate that peak detection improves inter-lab agreement."
comparison_data <- bind_rows(
  cor_heatmap_data_full %>% mutate(signal_type = "Full Signal"),
  cor_heatmap_data %>% mutate(signal_type = "Detected Signal")
)

ggplot(comparison_data, aes(x = pair, y = correlation, fill = signal_type)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~polarity, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Comparison: Full vs Detected Signal Correlations",
       subtitle = "Across all 19 mixtures and 6 lab pairs",
       x = "Lab Pair", y = "Pearson Correlation", fill = "Signal Type")
```

## 4.3 RT-Dependent Normalization Assessment

### 4.3.1 Rationale for Local Normalization

Median scaling using the overall median: What fraction of total metabolic
payload elutes in Q1 vs Q2? This preserves quartile relationships and tests
if labs agree on gradient "shape." This only works if metabolite
distribution is similar across gradient—check this below.

The alternative is local normalization, which computes median scaling
factors independently for each RT quartile. This corrects for labs with
non-linear sensitivity biases across the gradient.

Raw data check: Look at instrument sensitivity profile.

- Flat line: global normalization fine
- Sloped line: local normalization needed

### 4.3.2 Instrument Sensitivity Profile Across RT

```{r sensitivity-profile-computation}
compute_consortium_sensitivity <- function(polarity) {
  load(file.path(base_path, "object/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  breaks <- quantile(unlist(rtime(tic_data)), probs = seq(0, 1, 0.25), na.rm = TRUE)

  raw_stats <- t(vapply(seq_along(tic_data), function(i) {
    rt <- rtime(tic_data[i])[[1]]; int <- intensity(tic_data[i])[[1]]
    vapply(1:4, function(q) sum(int[rt >= breaks[q] & rt < breaks[q+1]], na.rm = TRUE), numeric(1))
  }, numeric(4)))

  df <- data.frame(lab = tic_data$lab, mixture = tic_data$mixture,
                   Q1 = raw_stats[,1], Q2 = raw_stats[,2], Q3 = raw_stats[,3], Q4 = raw_stats[,4])
  grand_medians <- df %>%
    summarise(across(starts_with("Q"), \(x) median(x, na.rm = TRUE))) %>%
    pivot_longer(everything(), names_to = "Quartile", values_to = "Consensus")

  df %>%
    pivot_longer(cols = starts_with("Q"), names_to = "Quartile", values_to = "Intensity") %>%
    left_join(grand_medians, by = "Quartile") %>%
    mutate(
      Log2_Deviation = log2(Intensity / Consensus),
      Polarity = ifelse(polarity == 1, "Positive", "Negative")
    )
}

# Compute sensitivity profiles for both polarities
all_profiles <- run_for_polarities_simple(compute_consortium_sensitivity)
```

### 4.3.3 Raw TIC by Quartile

```{r sensitivity-profile-plots}
#| fig-cap: "Instrument sensitivity profiles across RT quartiles. Non-parallel lines indicate RT-dependent lab biases."
# 5a. Raw TIC values per quartile (absolute scale)
all_profiles %>%
  mutate(Log2_TIC = log2(Intensity)) %>%
  group_by(Quartile, lab, Polarity) %>%
  summarise(median_log2_tic = median(Log2_TIC, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Quartile, y = median_log2_tic, group = lab, color = lab)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~Polarity) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Raw TIC Distribution Across Retention Time Quartiles",
    subtitle = "Absolute intensities show both lab differences and RT-dependent patterns",
    y = "log2(TIC Sum per Quartile)",
    x = "Quartile",
    color = "Laboratory"
  ) +
  theme_minimal()

# 5b. Deviation from consortium median (relative scale)
all_profiles %>%
  mutate(Log2_TIC = log2(Intensity)) %>%
  ggplot(aes(x = Quartile, y = Log2_TIC, color = lab, fill = lab)) +
  geom_boxplot(alpha = 0.3, outlier.alpha = 0.4, position = position_dodge(width = 0.8)) +
  stat_summary(aes(group = lab), fun = median, geom = "line", linewidth = 1, position = position_dodge(width = 0.8)) +
  stat_summary(aes(group = lab), fun = median, geom = "point", size = 2.5, position = position_dodge(width = 0.8)) +
  facet_wrap(~Polarity) +
  scale_color_viridis_d(option = "turbo") +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Lab Variation Across Retention Time Quartiles (Raw Data with Distribution)",
    subtitle = "Distribution shows variation across mixtures; non-parallel medians justify local normalization",
    y = "log2(TIC Sum per Quartile)",
    x = "Quartile",
    color = "Laboratory",
    fill = "Laboratory"
  ) +
  theme_minimal()
```

Note: This might be problematic to evaluate like this since every mixture is quite different,
so we have massive spread within each quartile.

```{r sensitivity-findings-table}
# Summary table: median log2 deviation from consortium median per lab per quartile
findings_table <- all_profiles %>%
  mutate(Log2_Dev = log2(Intensity / Consensus)) %>%
  group_by(lab, Quartile, Polarity) %>%
  summarise(
    Median_Log2_Dev = round(median(Log2_Dev, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = Quartile, values_from = Median_Log2_Dev)

knitr::kable(findings_table,
             caption = "Median log2 deviation from consortium median by lab and RT quartile")
```

**Findings:** Sensitivity slopes are non-linear: in positive mode, ICL shows
progressive loss from Q1 (+0.24) to Q4 (−0.78); Agilent labs show the
inverse trend. In negative mode, Cembio exhibits distinct hyper-sensitivity
in Q3 (+1.55 log2 deviation from consensus). These retention-time-specific
biases cannot be corrected with global normalization.

**Solution:** Apply Local Median Normalization, computing
scaling factors independently for each RT quartile.

## 4.4 Variance Decomposition by RT Quartile

### 4.4.1 Linear Model Analysis by Quartile

Section 4.3 showed that sensitivity profiles are non-linear: labs respond
differently across the chromatographic gradient. This raises the question:
**after global median normalization, how much residual lab variance remains
in each RT region?**

We split the detected signal TIC and BPC into four RT quartiles, then
apply the same linear model decomposition from Phase 1 to each quartile
independently. Intensities are median-normalized **within each lab** before
the split, so this tests whether the *relative* distribution of signal
across the gradient differs between labs — not whether absolute intensities
differ. High lab variance in a particular quartile would confirm that
global normalization is insufficient and local (per-quartile) correction is
needed.

Distribution of biological signal (TIC and BPC) across RT quartiles:

```{r quartile-variance-analysis}
compute_rt_quartiles_biological <- function(polarity, polarity_name) {
  load(file.path(base_path, "object/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  load(file.path(base_path, "object/bpc_detect.RData"))
  bpc_data <- bpc_detect[bpc_detect$polarity == polarity]

  tic_medians <- data.frame(lab = tic_data$lab,
    tic_total = vapply(intensity(tic_data), sum, numeric(1), na.rm = TRUE)) %>%
    group_by(lab) %>% summarise(median_tic = median(tic_total, na.rm = TRUE), .groups = "drop")

  bpc_medians <- data.frame(lab = bpc_data$lab,
    bpc_total = vapply(intensity(bpc_data), sum, numeric(1), na.rm = TRUE)) %>%
    group_by(lab) %>% summarise(median_bpc = median(bpc_total, na.rm = TRUE), .groups = "drop")

  all_rts <- unlist(rtime(tic_data))
  if (length(all_rts) == 0) return(NULL)
  global_breaks <- quantile(all_rts, probs = seq(0, 1, 0.25), na.rm = TRUE)

  compute_quartile_sums <- function(int_vec, rt_vec, breaks, lab_median) {
    vapply(1:4, function(q) sum(int_vec[rt_vec >= breaks[q] & rt_vec < breaks[q+1]], na.rm = TRUE) / lab_median, numeric(1))
  }

  compute_window_stats <- function(tic_chr, bpc_chr, breaks, lab_name, tic_meds, bpc_meds) {
    rt <- rtime(tic_chr)[[1]]
    c(compute_quartile_sums(intensity(tic_chr)[[1]], rt, breaks, tic_meds$median_tic[tic_meds$lab == lab_name]),
      compute_quartile_sums(intensity(bpc_chr)[[1]], rt, breaks, bpc_meds$median_bpc[bpc_meds$lab == lab_name]))
  }

  stats <- t(vapply(seq_along(tic_data), function(i) {
    compute_window_stats(tic_data[i], bpc_data[i], global_breaks, tic_data$lab[i], tic_medians, bpc_medians)
  }, numeric(8)))

  colnames(stats) <- c(
    paste0("TIC_Frac_Q", 1:4),
    paste0("BPC_Frac_Q", 1:4)
  )

  result <- data.frame(
    lab = tic_data$lab,
    mixture = tic_data$mixture,
    polarity = polarity_name,
    stats
  )

  return(result)
}

# Compute biological quartile distributions for both polarities
bio_combined_quartiles <- run_for_polarities_names(compute_rt_quartiles_biological)

quartile_results <- expand.grid(Polarity = c("Positive", "Negative"),
                                Metric = c(paste0("TIC_Frac_Q", 1:4), paste0("BPC_Frac_Q", 1:4))) %>%
  rowwise() %>%
  mutate(var_result = list(get_variance(filter(bio_combined_quartiles, polarity == Polarity), Metric))) %>%
  unnest(var_result)

quartile_results
```

### 4.4.2 Variance Visualization by Quartile

```{r quartile-variance-plots}
print(plot_quartile_variance(quartile_results, "TIC"))
print(plot_quartile_variance(quartile_results, "BPC"))
```

**Finding:** Local normalization by RT quartile reveals instrument-specific biases
in Q3/Q4 (30–50% residual lab variance). Hydrophobic compounds show
irreducible differences in selectivity and dynamic range that median
scaling cannot eliminate.

---

# Phase 5: Peak Feature Analysis — Consensus Peaks {#phase5}

This phase extends the global analysis by incorporating peak-level metrics
computed from EICs of consensus peaks detected across all 4 laboratories.
This provides a more refined view of inter-lab reproducibility at the
individual peak level.

## 5.1 Data Overview

The peak feature data comes from `3a_peak_feature_calculation_detected.qmd`,
which:

1. Created a consensus peak table matching peaks across labs (mz tol = 10 ppm, rt tol = 10s)
2. Filtered to peaks detected in all 4 labs (1209 consensus peaks → 871 after QC)
3. Extracted EICs and computed peak metrics (FWHM, area, SNR, etc.)
4. Applied quality filters (prominence, peak width, FWHM validity)

```{r load-peak-metrics}
pk_metrics <- read.csv(file.path(base_path, "object/peak_metrics_detected_all4labs.csv"))

cat("Peak metrics dimensions:", nrow(pk_metrics), "rows x", ncol(pk_metrics), "columns\n")
cat("Unique consensus peaks:", length(unique(pk_metrics$consensus_id)), "\n")
cat("Labs represented:", paste(unique(pk_metrics$lab), collapse = ", "), "\n")
```

## 5.2 Spearman Intensity Rank Correlation

We use Spearman (rank-based) correlation to test whether peaks that are
relatively intense in one lab are also relatively intense in another.
Spearman is robust to the non-linear scale differences across labs without
requiring explicit normalization — it compares rank order, not absolute
values.

### 5.2.1 Intensity Correlation Matrix

```{r peak-metrics-correlation}
# Pivot to wide format for correlation
intensity_wide <- pk_metrics |>
  select(consensus_id, lab, areaUnderTic, maxIntensity, mixture, polarity) |>
  pivot_wider(
    id_cols = c(consensus_id, mixture, polarity),
    names_from = lab,
    values_from = c(areaUnderTic, maxIntensity)
  )

# Spearman correlation for areaUnderTic
area_cols <- paste0("areaUnderTic_", c("afekta", "hmgu", "icl", "cembio"))
area_matrix <- intensity_wide[, area_cols]
names(area_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_area <- cor(area_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("Spearman Correlation - Area Under TIC:\n")
print(round(cor_area, 3))

# Spearman correlation for maxIntensity
max_cols <- paste0("maxIntensity_", c("afekta", "hmgu", "icl", "cembio"))
max_matrix <- intensity_wide[, max_cols]
names(max_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_max <- cor(max_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("\nSpearman Correlation - Max Intensity:\n")
print(round(cor_max, 3))
```

### 5.2.2 Correlation Heatmap Visualization

```{r peak-correlation-heatmap}
#| fig-cap: "Spearman rank correlation heatmaps for recalculated peak intensities across 4 labs."
#| fig-height: 5
library(pheatmap)

pheatmap(cor_area,
         main = "Area Under TIC (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)

pheatmap(cor_max,
         main = "Max Intensity (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)
```

Afekta and Cembio show the strongest rank correlation (ρ ≈ 0.76),
consistent with their shared Agilent Q-TOF platform. All other pairs
are moderate (ρ ≈ 0.4–0.5), indicating that relative peak intensity
rankings are partially conserved across labs but with substantial
instrument-specific variation.

## 5.3 Peak Metric Distributions Across Labs

The following metrics are computed for each EIC from the filtered consensus
peaks:

- **Peak Width**: Time span (seconds) between the detected peak boundaries.
  Reflects chromatographic resolution and column performance.
- **FWHM** (Full Width at Half Maximum): Width of the peak at half its
  maximum intensity. A narrower, more symmetric measure of peak shape than
  peak width.
- **Signal-to-Noise Ratio (SNR)**: Ratio of peak maximum intensity to the
  estimated baseline. Higher SNR indicates a cleaner, more confident
  detection.

### 5.3.1 Peak Width Distribution

```{r peak-width-distribution}
ggplot(pk_metrics, aes(x = lab, y = peakWidth, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Peak Width Distribution by Lab",
   subtitle = "Consensus peaks detected in all 4 labs",
   x = "Laboratory",
   y = "Peak Width (seconds)"
 ) +
 theme_minimal() +
 theme(legend.position = "none")
```

### 5.3.2 FWHM Distribution

```{r fwhm-distribution}
ggplot(pk_metrics |> filter(!is.na(xicFwhm)), aes(x = lab, y = xicFwhm, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "FWHM Distribution by Lab",
   subtitle = "Full Width at Half Maximum of detected peaks",
   x = "Laboratory",
   y = "FWHM (seconds)"
 ) +
 theme_minimal() +
 theme(legend.position = "none")
```

### 5.3.3 Signal-to-Noise Ratio Distribution

```{r snr-distribution}
ggplot(pk_metrics, aes(x = lab, y = log10(signalToNoiseRatio + 1), fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Signal-to-Noise Ratio Distribution by Lab",
   subtitle = "log10(SNR + 1) scale",
   x = "Laboratory",
   y = "log10(SNR + 1)"
 ) +
 theme_minimal() +
 theme(legend.position = "none")
```

## 5.4 Inter-Lab Variability for Consensus Peaks

For each consensus peak (detected in all 4 labs), we compute the coefficient
of variation (CV = standard deviation / mean) of each metric across the 4
labs. CV is inherently scale-normalized: it measures relative spread
regardless of absolute intensity differences, so no prior normalization of
intensity metrics is needed.

- **CV < 0.3 (30%)**: Good reproducibility
- **CV 0.3–0.5**: Moderate variability
- **CV > 0.5**: High variability

**IQR** (Interquartile Range): the difference between the 75th and 25th
percentiles of the CV distribution, measuring the spread of the middle 50%
of values.

### 5.4.1 CV Summary Table

```{r cv-summary-table}
cv_by_consensus <- pk_metrics |>
  group_by(consensus_id) |>
  summarise(
    n_labs = n(),
    cv_area = sd(areaUnderTic, na.rm = TRUE) / mean(areaUnderTic, na.rm = TRUE),
    cv_max = sd(maxIntensity, na.rm = TRUE) / mean(maxIntensity, na.rm = TRUE),
    cv_fwhm = sd(xicFwhm, na.rm = TRUE) / mean(xicFwhm, na.rm = TRUE),
    cv_peakwidth = sd(peakWidth, na.rm = TRUE) / mean(peakWidth, na.rm = TRUE),
    cv_snr = sd(signalToNoiseRatio, na.rm = TRUE) / mean(signalToNoiseRatio, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_labs == 4)

cv_summary <- cv_by_consensus |>
  summarise(
    across(starts_with("cv_"), list(
      median = ~median(.x, na.rm = TRUE),
      q25 = ~quantile(.x, 0.25, na.rm = TRUE),
      q75 = ~quantile(.x, 0.75, na.rm = TRUE)
    ))
  )

cv_table <- data.frame(
  Metric = c("Area Under TIC", "Max Intensity", "FWHM", "Peak Width", "SNR"),
  Median_CV = c(cv_summary$cv_area_median, cv_summary$cv_max_median,
                cv_summary$cv_fwhm_median, cv_summary$cv_peakwidth_median,
                cv_summary$cv_snr_median),
  IQR = c(cv_summary$cv_area_q75 - cv_summary$cv_area_q25,
          cv_summary$cv_max_q75 - cv_summary$cv_max_q25,
          cv_summary$cv_fwhm_q75 - cv_summary$cv_fwhm_q25,
          cv_summary$cv_peakwidth_q75 - cv_summary$cv_peakwidth_q25,
          cv_summary$cv_snr_q75 - cv_summary$cv_snr_q25)
)

knitr::kable(cv_table,
             caption = "Inter-Lab CV for Consensus Peaks (median and interquartile range)",
             digits = 3)
```

### 5.4.2 CV Distribution Visualization

```{r cv-distribution}
#| fig-cap: "CV distribution across consensus peaks for each metric. Red dashed line = 30% CV threshold."
cv_long <- cv_by_consensus |>
 pivot_longer(
   cols = starts_with("cv_"),
   names_to = "metric",
   names_prefix = "cv_",
   values_to = "cv"
 ) |>
 mutate(metric = case_when(
   metric == "area" ~ "Area Under TIC",
   metric == "max" ~ "Max Intensity",
   metric == "fwhm" ~ "FWHM",
   metric == "peakwidth" ~ "Peak Width",
   metric == "snr" ~ "SNR"
 ))

ggplot(cv_long |> filter(!is.na(cv) & cv < 2), aes(x = metric, y = cv, fill = metric)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 geom_hline(yintercept = 0.3, linetype = "dashed", color = "red") +
 annotate("text", x = 5.3, y = 0.35, label = "CV = 30%", color = "red", size = 3) +
 labs(
   title = "Inter-Lab Variability for Consensus Peaks",
   subtitle = "Lower CV = better reproducibility across labs",
   x = "Metric",
   y = "Coefficient of Variation"
 ) +
 theme_minimal() +
 theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

Shape metrics (FWHM, peak width) show lower CV than intensity metrics
(area, max intensity), confirming that chromatographic behavior is more
reproducible across labs than ionization response. SNR shows the highest
variability, reflecting differences in baseline estimation across
instruments.

## 5.5 Peak Shape Similarity

Beyond intensity metrics, we examine the actual chromatographic peak
profiles to assess shape reproducibility across labs. For each consensus
peak, EIC profiles are interpolated to a common 0–1 normalized x-axis
and compared using Pearson correlation (shape agreement) and Tanimoto
similarity (intensity distribution overlap). Full details of the
computation are in `3a_peak_feature_calculation_detected.qmd`.

### 5.5.1 Shape Similarity Summary

```{r load-shape-results}
shape_results <- read.csv(file.path(base_path, "object/peak_shape_similarity.csv"))

shape_summary <- shape_results |>
  group_by(pair) |>
  summarise(
    n = n(),
    pearson_median = median(pearson, na.rm = TRUE),
    pearson_mean = mean(pearson, na.rm = TRUE),
    tanimoto_median = median(tanimoto, na.rm = TRUE),
    tanimoto_mean = mean(tanimoto, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(shape_summary,
             caption = "Peak Shape Similarity by Lab Pair",
             digits = 3)
```

Median Pearson correlations range from 0.59 (hmgu:icl) to 0.73
(afekta:hmgu). Tanimoto values are broadly similar across pairs
(0.59–0.64), indicating that absolute profile overlap is more uniform
than linear shape correlation.

### 5.5.2 Pearson Correlation Distribution

The 0.9 threshold is a standard cutoff for high-quality shape agreement
in chromatographic peak matching.

```{r pearson-distribution-global}
#| fig-cap: "Pearson correlation distribution by lab pair. Red dashed line = 0.9 threshold."
#| fig-width: 10
#| fig-height: 6
shape_pearson <- shape_results |> filter(!is.na(pearson))

pct_above <- shape_pearson |>
  group_by(pair) |>
  summarise(
    n_comparisons = n(),
    n_above_0.9 = sum(pearson >= 0.9),
    pct_above_0.9 = round(100 * mean(pearson >= 0.9), 1),
    .groups = "drop"
  )

knitr::kable(pct_above,
             caption = "Percentage of consensus peaks with Pearson >= 0.9",
             col.names = c("Lab Pair", "Total Comparisons",
                           "N above 0.9", "% above 0.9"))

ggplot(shape_pearson, aes(x = pearson, fill = pair)) +
  geom_histogram(bins = 50, alpha = 0.7, color = "white", linewidth = 0.2) +
  facet_wrap(~pair, scales = "free_y") +
  geom_vline(xintercept = 0.9, linetype = "dashed", color = "red") +
  labs(
    title = "Pearson Correlation Distribution by Lab Pair",
    subtitle = "Red dashed line = 0.9 (high-quality shape match)",
    x = "Pearson Correlation",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Only 11–20% of consensus peaks reach the 0.9 threshold depending on the
lab pair, with afekta:hmgu highest (~20%) and afekta:icl lowest (~11%).
The distributions are left-skewed with a long tail of poorly correlated
peaks.

---

# Summary {#summary}

This analysis assessed LC-MS reproducibility across 4 laboratories
analyzing 19 standard mixtures in positive and negative ionization modes.

1. **Peak detection improves reproducibility**: Detected signal ICC
   0.50–0.56 vs full signal ICC 0.27–0.36. Isolating compound signal
   from baseline noise is essential.

2. **Same-platform instruments agree better**: Agilent Q-TOF labs
   (Afekta, Cembio) show ICC 0.84/0.92 and Spearman ρ ≈ 0.76; different
   platforms show lower agreement (ρ ≈ 0.4–0.5).

3. **Normalization strategy matters**: Median normalization improves ICC
   from ~0.04 to ~0.53 — crossing the threshold from poor to moderate
   reproducibility. Autoscaling introduces unwanted residual patterns.

4. **RT-dependent biases**: Instrument sensitivity profiles vary non-linearly
   across the gradient (e.g., ICL loses sensitivity in late quartiles;
   Cembio shows lipid hyper-sensitivity). Local normalization by RT
   quartile may be needed.

5. **Chromatographic shape is more reproducible than ionization**: Shape
   metrics (FWHM, peak width) show lower inter-lab CV than intensity
   metrics (area, max intensity), confirming that chromatographic
   behavior is better conserved across instruments.

6. **Peak shape similarity**: Median Pearson correlations range from 0.59
   to 0.73 across lab pairs. Only 11–20% of consensus peaks exceed the
   0.9 high-quality threshold, indicating substantial room for
   improvement.

7. **Negative mode bias**: Consistent ~9% systematic bias between Agilent
   labs; needs further investigation.

---

# Future Directions

- Normalize peak intensities at the consensus level and recompute ICC
- Apply local (RT-quartile) normalization and assess improvement
- Investigate negative mode bias systematically
- Expand to Afekta–Cembio overlap analysis (50k+ peaks)
- Test in-house vs commercial reverse-phase columns (Afekta)
- Annotated data analysis once available for HE samples
- Expand analysis to other HUMAN chemical classes (FE, Others) beyond HE
  endosome mixtures
- Redo analysis on imputed data using `imputePeaksData(eics, method = "gaussian")`


