---
title: "Global Signal Analysis - Positive Mode (Simplified)"
format: html
editor_options:
  chunk_output_type: console
---

# Introduction

This simplified analysis computes key metrics from preprocessed LC-MS data to compare across different labs and instrument setups. The analysis focuses on:

- **Base Peak Chromatogram (BPC)**: Sum of peak apices
- **Total Ion Current (TIC)**: Sum of all peak areas
- **BPC/TIC Ratio**: Metric for peak purity/complexity
- **TIC Quantiles**: Distribution assessment (Q25, Q50, Q75)

For each metric, the analysis evaluates data distribution to determine if transformation is needed before statistical analysis, then fits linear models: `Metric ~ lab + mixture + setup`

The analysis is performed on three data levels:
1. **Full Data**: All detected signal (includes baseline noise)
2. **Detected Data**: Signal from peak picking (excludes baseline noise)
3. **Annotated Data**: Only annotated standard compounds

Setup information is derived from the LC and MS instruments in the lab metadata.

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggbeeswarm)
library(nortest)
library(car)
library(broom)
library(here)

base_path <- here("5_downstream_analysis")
```

```{r, cores-cluster}
# bpparam <- MulticoreParam(workers = parallel::detectCores() - 1)
```

## Setup Information

Load and prepare the setup metadata:

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"), skip = 1)

# Create a combined setup variable from LC and MS instruments
setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`) %>%
  distinct() %>%
  mutate(
    lc_short = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent1290",
      grepl("Acquity", `LC.Instrument`) ~ "Acquity",
      grepl("ExionLC", `LC.Instrument`) ~ "ExionLC",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_short = case_when(
      grepl("6540|6545", `MS.Instrument`) ~ "Agilent",
      grepl("Synapt", `MS.Instrument`) ~ "Waters",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    setup = paste(lc_short, ms_short, sep = "_")
  ) %>%
  select(lab, setup)

setup_simplified
```

# Full Data Analysis

Load the full preprocessed data:

```{r}
pol <- 1L  # positive polarity

load(file.path(base_path, "object/sp_full.RData"))
sp_full <- sp_full[sp_full$polarity == pol]

load(file.path(base_path, "object/bpc_full.RData"))
bpc_full <- bpc_full[bpc_full$polarity == pol]

load(file.path(base_path, "object/tic_full.RData"))
tic_full <- tic_full[tic_full$polarity == pol]
```

## Compute Metrics

```{r compute-full-metrics}
metrics_full <- data.frame(
  lab = bpc_full$lab,
  mixture = bpc_full$mixture,
  bpc = vapply(intensity(bpc_full), sum, numeric(1), na.rm = TRUE), ## sum or median, i'd say sum especially to compare instrument perf
  tic = vapply(intensity(tic_full), sum, numeric(1), na.rm = TRUE)
)

metrics_full$bpc_tic_ratio <- metrics_full$bpc / metrics_full$tic

# Add TIC quantiles
tic_quantiles <- lapply(intensity(tic_full), function(x) {
  quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
})
metrics_full$tic_q25 <- sapply(tic_quantiles, `[`, 1)
metrics_full$tic_q50 <- sapply(tic_quantiles, `[`, 2)
metrics_full$tic_q75 <- sapply(tic_quantiles, `[`, 3)

# Add setup information
metrics_full <- metrics_full %>%
  left_join(setup_simplified, by = "lab")

head(metrics_full)
```

## Distribution Assessment

For each metric, assess normality to determine if transformation is needed:

```{r dist-assessment-full}
metrics_to_assess <- c("bpc", "tic", "bpc_tic_ratio", "tic_q25", "tic_q50", "tic_q75")

distribution_summary_full <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in metrics_to_assess) {
  data_vec <- metrics_full[[m]]
  data_vec <- data_vec[!is.na(data_vec)]
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)  # Add 1 to handle potential zeros
  shapiro_log <- shapiro.test(log_data)
  
  # Recommend transform if log is more normal (p > 0.05)
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_full <- rbind(distribution_summary_full, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_full
```

## Distribution Plots - Full Data

Violin plots and histograms for all metrics:

```{r plots-full, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_full <- metrics_full %>%
  pivot_longer(
    cols = all_of(metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

# Violin plots by lab
ggplot(plot_data_full, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Full Data - Distribution of Metrics by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_full$lab)) {
  p <- ggplot(plot_data_full %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Full Data -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab (where applicable)
plot_data_full_log <- plot_data_full %>%
  filter(metric %in% c("bpc", "tic", "tic_q25", "tic_q50", "tic_q75"))

for (lab_name in unique(plot_data_full_log$lab)) {
  p <- ggplot(plot_data_full_log %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Full Data -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Full Data

```{r lm-full}
# BPC model
lm_bpc_full <- lm(log(bpc) ~ lab + setup, data = metrics_full)
cat("\n=== BPC Full Data ===\n")
summary(lm_bpc_full)

# TIC model
lm_tic_full <- lm(log(tic) ~ lab + setup, data = metrics_full)
cat("\n=== TIC Full Data ===\n")
summary(lm_tic_full)

# BPC/TIC ratio model
lm_ratio_full <- lm(bpc_tic_ratio ~ lab + setup, data = metrics_full)
cat("\n=== BPC/TIC Ratio Full Data ===\n")
summary(lm_ratio_full)

#TIC 25 model
lm_tic_q25_full <- lm(log(tic_q25) ~ lab + setup, data = metrics_full)
cat("\n=== TIC Q25 Full Data ===\n")
summary(lm_tic_q25_full)

# TIC Q50 model
lm_tic_q50_full <- lm(log(tic_q50) ~ lab + setup, data = metrics_full)
cat("\n=== TIC Q50 Full Data ===\n")
summary(lm_tic_q50_full)

# TIC Q75 model
lm_tic_q75_full <- lm(log(tic_q75) ~ lab + setup, data = metrics_full)
cat("\n=== TIC Q75 Full Data ===\n")
summary(lm_tic_q75_full)
```

## Interpretation of Full Data Linear Models

### Key Findings

**Laboratory Effects (Primary Finding)**
All metrics show highly significant laboratory effects (p < 2e-16), indicating substantial differences in signal intensity across labs:

- **BPC (Log scale)**: HMGU produces ~40× less signal (exp(-3.937)), cembio ~44% less (exp(-0.750)), and ICL ~37% less (exp(-0.980)) compared to afekta (baseline).
- **TIC (Log scale)**: Even stronger lab effects. HMGU shows ~50× reduction, cembio ~72% reduction, ICL ~61% reduction. This metric is highly discriminatory (R² = 0.9953).
- **BPC/TIC Ratio**: Only cembio differs significantly (p = 0.003), with 0.037 higher ratio. HMGU and ICL show no significant difference from afekta. This suggests different peak complexity patterns between labs.
- **TIC Quantiles (Q25, Q50, Q75)**: All show extremely strong lab effects (p < 2e-16) with consistent patterns—HMGU >> cembio > ICL differences. Effects are strongest at Q75 (narrower residuals: 0.0747 vs 0.1038 for BPC).

**Instrument Setup Effects**
All setup coefficients return NA (singularities) due to **rank deficiency**: afekta and cembio share the same setup (Agilent1290_Agilent), while icl and hmgu each use unique setups. The model cannot separately estimate setup effects from lab effects when some labs share setups and others don't. To estimate separate setup effects:
- Collect data from multiple labs using the same setup, or
- Remove setup from the model to focus on lab effects alone (which is the primary factor here).

**Model Quality**
- **Best models**: TIC Q75 (R² = 0.9972) and TIC Q25 (R² = 0.9959) explain 99.7% and 99.6% of variance
- **Moderate model**: BPC (R² = 0.943) explains 94.3% of variance
- **Weak model**: BPC/TIC Ratio (R² = 0.1442) explains only 14.4%—this ratio is not a good discriminator between labs

**Residual Diagnostics**
- BPC residuals show right skew and outliers (Max = 1.29, Min = -0.76), suggesting non-normality despite log transformation
- TIC metrics have much tighter, more symmetric residuals
- BPC/TIC has smallest residual range, indicating more consistent measurement

### Interpretation Summary

1. **Strong Lab Differences**: Across all metrics, HMGU consistently produces the weakest signal, while afekta (baseline) produces the strongest. This likely reflects instrument sensitivity differences or sample handling protocols.

2. **TIC is More Reliable**: TIC metrics (especially quantiles) show tighter models and less residual variation, making them more suitable for inter-lab comparisons than BPC alone.

3. **Peak Complexity Similar**: The BPC/TIC ratio shows minimal variation between labs (except cembio), suggesting all labs produce similarly complex peak distributions despite different signal magnitudes.

## Paired Analysis by Mixture

```{r paired-test, fig.width=12, fig.height=8}
# 1. VISUALIZATION (Paired Trends) ----------------------------------------
# "Spaghetti Plot" - Tracks how each lab performs across the *same* mixtures
# If lines cross significantly, there's a lab×mixture interaction.
# If they are parallel but separated, there's a systematic lab offset.

ggplot(
  metrics_full,
  aes(x = mixture, y = bpc_tic_ratio, color = lab, group = lab)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  theme_bw() +
  labs(
    title = "BPC/TIC Ratio Trends across Mixtures by Lab",
    subtitle = "Parallel lines = consistent lab effect; crossing lines = lab×mixture interaction",
    y = "BPC / TIC Ratio (Signal Quality)",
    x = "Mixture ID"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# 2. BOXPLOT - Lab effects aggregated across mixtures
ggplot(metrics_full, aes(x = lab, y = bpc_tic_ratio, fill = setup)) +
  geom_boxplot(alpha = 0.7) +
  theme_bw() +
  labs(
    title = "BPC/TIC Ratio Distribution by Lab",
    y = "BPC / TIC Ratio",
    x = "Laboratory"
  )

# 3. STATISTICS (Paired/Blocked) ------------------------------------------

# Friedman Test: Is there a difference between labs, controlling for mixture?
friedman_res <- friedman.test(
  bpc_tic_ratio ~ lab | mixture,
  data = metrics_full
)
cat("\n=== Friedman Test (Non-parametric ANOVA) ===\n")
print(friedman_res)

# Pairwise Wilcoxon Test: Compare each pair of labs (paired by mixture)
cat("\n=== Pairwise Wilcoxon Test (Bonferroni corrected) ===\n")
pairwise_res <- pairwise.wilcox.test(
  metrics_full$bpc_tic_ratio,
  metrics_full$lab,
  p.adjust.method = "bonferroni",
  paired = TRUE
)
print(pairwise_res)

# Focused comparison: Agilent labs (afekta vs cembio) - same setup, different protocols
agilent_data <- metrics_full %>% filter(setup == "Agilent1290_Agilent")
agilent_diff <- wilcox.test(bpc_tic_ratio ~ lab, data = agilent_data, paired = TRUE)
cat("\n=== Wilcoxon Test: Afekta vs Cembio (identical setup, paired by mixture) ===\n")
print(agilent_diff)
```

## Interpretation of Paired Analysis Results

### Why Use BPC/TIC Ratio for Paired Tests?

The paired analysis focuses on **BPC/TIC ratio** rather than absolute metrics (BPC, TIC, quantiles) for several reasons:

1. **Removes Instrument/Lab Sensitivity Effects**: The ratio normalizes for differences in instrument sensitivity. A 40× difference in absolute TIC between HMGU and afekta becomes irrelevant; we only care if the *peak complexity* differs.

2. **Tests Biological/Method Variance, Not Hardware**: The ratio captures whether labs process samples differently (e.g., different peak picking parameters, sample preparation), not just raw detector sensitivity.

3. **Appropriate for Paired Design**: Each mixture is measured across all labs. The ratio stays relatively constant within mixtures if labs are equivalent, making paired comparisons more powerful and interpretable.

4. **Better Statistical Properties**: The ratio is dimensionless and less likely to violate non-parametric test assumptions compared to raw signal intensities.

### Key Findings from Paired Tests

**Friedman Test (p = 0.00086)**
- Highly significant difference between labs when controlling for mixture
- Confirms that labs differ systematically across the same sample set
- Demonstrates the paired design effectively controls for sample variation

**Pairwise Comparisons (Bonferroni corrected)**
- **Afekta vs Cembio: p = 2.3e-05*** (strongest difference)
  - Despite using identical equipment, these two Agilent labs produce significantly different peak ratios
  - Suggests **human/protocol differences** (sample prep, calibration, operator technique) dominate over hardware
  
- **Cembio vs HMGU: p = 0.049** (borderline significant)
  - Different instruments (Agilent vs Sciex) show marginal difference in peak ratio
  - Suggests instrument type has less effect than lab-specific practices
  
- **Afekta vs HMGU: p = 1.000** (not significant)
- **Afekta vs ICL: p = 1.000** (not significant)
- **HMGU vs ICL: p = 1.000** (not significant)
  - Most lab pairs are equivalent in peak complexity

**Paired Wilcoxon: Afekta vs Cembio (p = 3.8e-06)**
- Confirms strongest difference is between the two "identical-setup" labs
- V = 190 indicates strong consistent direction of difference across mixtures
- Evidence that **lab practices/protocols matter more than hardware specifications**

### Recommendations

**Should you run paired tests on other metrics? NO, here's why:**

1. **BPC & TIC**: These show ~40-50× differences between labs (lab×hardware confounded). Paired tests would just re-confirm hardware differences we already know from the linear models (R² = 0.99). Low interpretive value.

2. **TIC Quantiles (Q25, Q50, Q75)**: Same issue—these are absolute signal measures showing strong hardware effects. Paired tests add no new insight beyond "HMGU is weaker."

3. **Ratio is the only metric that separates hardware from protocols**: It's the only metric where afekta and cembio can differ despite identical equipment.

**Consider removing/simplifying to:**
- Keep: **Friedman test** (global test controlling for mixture)
- Keep: **Pairwise Wilcoxon** (identifies which labs differ)
- **Optional removal**: The three non-significant pairwise comparisons (afekta vs HMGU/ICL, HMGU vs ICL) could be condensed with a note that "remaining pairs are not significant"
- Keep: **Agilent paired comparison** (illustrates protocol vs hardware effect)

This focused approach avoids redundancy and highlights the key insight: **lab protocols matter as much or more than instrument hardware**.





---

# Detected Signal Analysis

Load the detected signal data:

```{r load-detect}
load(file.path(base_path, "object/sp_full_detect.RData"))
sp_full_detect <- sp_full_detect[sp_full_detect$polarity == pol]

load(file.path(base_path, "object/bpc_detect.RData"))
bpc_detect <- bpc_detect[bpc_detect$polarity == pol]

load(file.path(base_path, "object/tic_detect.RData"))
tic_detect <- tic_detect[tic_detect$polarity == pol]
```

## Compute Metrics

```{r compute-detect-metrics}
metrics_detect <- data.frame(
  lab = bpc_detect$lab,
  mixture = bpc_detect$mixture,
  bpc = vapply(intensity(bpc_detect), sum, numeric(1), na.rm = TRUE),
  tic = vapply(intensity(tic_detect), sum, numeric(1), na.rm = TRUE)
)

metrics_detect$bpc_tic_ratio <- metrics_detect$bpc / metrics_detect$tic

# Add TIC quantiles
tic_quantiles <- lapply(intensity(tic_detect), function(x) {
  quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
})
metrics_detect$tic_q25 <- sapply(tic_quantiles, `[`, 1)
metrics_detect$tic_q50 <- sapply(tic_quantiles, `[`, 2)
metrics_detect$tic_q75 <- sapply(tic_quantiles, `[`, 3)

# Add setup information
metrics_detect <- metrics_detect %>%
  left_join(setup_simplified, by = "lab")

head(metrics_detect)
```

## Distribution Assessment

```{r dist-assessment-detect}
distribution_summary_detect <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in metrics_to_assess) {
  data_vec <- metrics_detect[[m]]
  data_vec <- data_vec[!is.na(data_vec)]
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)
  shapiro_log <- shapiro.test(log_data)
  
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_detect <- rbind(distribution_summary_detect, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_detect
```

## Distribution Plots - Detected Data

Violin plots and histograms for all metrics:

```{r plots-detect, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_detect <- metrics_detect %>%
  pivot_longer(
    cols = all_of(metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

# Violin plots by lab
ggplot(plot_data_detect, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Detected Data - Distribution of Metrics by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_detect$lab)) {
  p <- ggplot(plot_data_detect %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Detected Data -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab (where applicable)
plot_data_detect_log <- plot_data_detect %>%
  filter(metric %in% c("bpc", "tic", "tic_q25", "tic_q50", "tic_q75"))

for (lab_name in unique(plot_data_detect_log$lab)) {
  p <- ggplot(plot_data_detect_log %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Detected Data -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Detected Data

```{r lm-detect}
# BPC model
lm_bpc_detect <- lm(log(bpc) ~ lab + setup, data = metrics_detect)
cat("\n=== BPC Detected Data ===\n")
summary(lm_bpc_detect)

# TIC model
lm_tic_detect <- lm(log(tic) ~ lab + setup, data = metrics_detect)
cat("\n=== TIC Detected Data ===\n")
summary(lm_tic_detect)

# BPC/TIC ratio model
lm_ratio_detect <- lm(bpc_tic_ratio ~ lab + setup, data = metrics_detect)
cat("\n=== BPC/TIC Ratio Detected Data ===\n")
summary(lm_ratio_detect)

# TIC Q50 model
lm_tic_q50_detect <- lm(log(tic_q50) ~ lab + setup, data = metrics_detect)
cat("\n=== TIC Q50 Detected Data ===\n")
summary(lm_tic_q50_detect)
```

## Paired Analysis by Mixture - Detected Data

```{r paired-test-detect, fig.width=12, fig.height=8}
# Visualization: Spaghetti plot for detected data
ggplot(
  metrics_detect,
  aes(x = mixture, y = bpc_tic_ratio, color = lab, group = lab)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  theme_bw() +
  labs(
    title = "BPC/TIC Ratio Trends (Detected Signal)",
    subtitle = "Parallel lines = consistent lab effect; crossing lines = lab×mixture interaction",
    y = "BPC / TIC Ratio",
    x = "Mixture ID"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Statistics: Friedman Test
friedman_detect <- friedman.test(
  bpc_tic_ratio ~ lab | mixture,
  data = metrics_detect
)
cat("\n=== Friedman Test (Detected Data) ===\n")
print(friedman_detect)

# Pairwise comparisons
cat("\n=== Pairwise Wilcoxon Test (Detected Data) ===\n")
pairwise_detect <- pairwise.wilcox.test(
  metrics_detect$bpc_tic_ratio,
  metrics_detect$lab,
  p.adjust.method = "bonferroni",
  paired = TRUE
)
print(pairwise_detect)

# Agilent labs comparison
agilent_detect <- metrics_detect %>% filter(setup == "Agilent1290_Agilent")
agilent_diff_detect <- wilcox.test(bpc_tic_ratio ~ lab, data = agilent_detect, paired = TRUE)
cat("\n=== Wilcoxon Test: Afekta vs Cembio (Detected Data) ===\n")
print(agilent_diff_detect)
```

```{r clean-detect}
rm(tic_detect, bpc_detect, sp_full_detect)
```

---

# Annotated Signal Analysis

Load the annotated signal data:

```{r load-ann}
load(file.path(base_path, "object/sp_full_ann.RData"))
sp_full_ann <- sp_full_ann[sp_full_ann$polarity == pol]

load(file.path(base_path, "object/bpc_ann.RData"))
bpc_ann <- bpc_ann[bpc_ann$polarity == pol]

load(file.path(base_path, "object/tic_ann.RData"))
tic_ann <- tic_ann[tic_ann$polarity == pol]
```

## Compute Metrics

```{r compute-ann-metrics}
metrics_ann <- data.frame(
  lab = bpc_ann$lab,
  mixture = bpc_ann$mixture,
  bpc = vapply(intensity(bpc_ann), sum, numeric(1), na.rm = TRUE),
  tic = vapply(intensity(tic_ann), sum, numeric(1), na.rm = TRUE)
)

metrics_ann$bpc_tic_ratio <- metrics_ann$bpc / metrics_ann$tic

# Add TIC quantiles
tic_quantiles <- lapply(intensity(tic_ann), function(x) {
  quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
})
metrics_ann$tic_q25 <- sapply(tic_quantiles, `[`, 1)
metrics_ann$tic_q50 <- sapply(tic_quantiles, `[`, 2)
metrics_ann$tic_q75 <- sapply(tic_quantiles, `[`, 3)

# Add setup information
metrics_ann <- metrics_ann %>%
  left_join(setup_simplified, by = "lab")

head(metrics_ann)
```

## Distribution Assessment

```{r dist-assessment-ann}
distribution_summary_ann <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in metrics_to_assess) {
  data_vec <- metrics_ann[[m]]
  data_vec <- data_vec[!is.na(data_vec)]
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)
  shapiro_log <- shapiro.test(log_data)
  
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_ann <- rbind(distribution_summary_ann, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_ann
```

## Distribution Plots - Annotated Data

Violin plots and histograms for all metrics:

```{r plots-ann, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_ann <- metrics_ann %>%
  pivot_longer(
    cols = all_of(metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

# Violin plots by lab
ggplot(plot_data_ann, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Annotated Data - Distribution of Metrics by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_ann$lab)) {
  p <- ggplot(plot_data_ann %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Annotated Data -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab (where applicable)
plot_data_ann_log <- plot_data_ann %>%
  filter(metric %in% c("bpc", "tic", "tic_q25", "tic_q50", "tic_q75"))

for (lab_name in unique(plot_data_ann_log$lab)) {
  p <- ggplot(plot_data_ann_log %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Annotated Data -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Annotated Data

```{r lm-ann}
# BPC model
lm_bpc_ann <- lm(log(bpc) ~ lab + setup, data = metrics_ann)
cat("\n=== BPC Annotated Data ===\n")
summary(lm_bpc_ann)

# TIC model
lm_tic_ann <- lm(log(tic) ~ lab + setup, data = metrics_ann)
cat("\n=== TIC Annotated Data ===\n")
summary(lm_tic_ann)

# BPC/TIC ratio model
lm_ratio_ann <- lm(bpc_tic_ratio ~ lab + setup, data = metrics_ann)
cat("\n=== BPC/TIC Ratio Annotated Data ===\n")
summary(lm_ratio_ann)

# TIC Q50 model
lm_tic_q50_ann <- lm(log(tic_q50) ~ lab + setup, data = metrics_ann)
cat("\n=== TIC Q50 Annotated Data ===\n")
summary(lm_tic_q50_ann)
```

## Paired Analysis by Mixture - Annotated Data

```{r paired-test-ann, fig.width=12, fig.height=8}
# Visualization: Spaghetti plot for annotated data
ggplot(
  metrics_ann,
  aes(x = mixture, y = bpc_tic_ratio, color = lab, group = lab)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  theme_bw() +
  labs(
    title = "BPC/TIC Ratio Trends (Annotated Compounds Only)",
    subtitle = "Parallel lines = consistent lab effect; crossing lines = lab×mixture interaction",
    y = "BPC / TIC Ratio",
    x = "Mixture ID"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Statistics: Friedman Test
friedman_ann <- friedman.test(
  bpc_tic_ratio ~ lab | mixture,
  data = metrics_ann
)
cat("\n=== Friedman Test (Annotated Data) ===\n")
print(friedman_ann)

# Pairwise comparisons
cat("\n=== Pairwise Wilcoxon Test (Annotated Data) ===\n")
pairwise_ann <- pairwise.wilcox.test(
  metrics_ann$bpc_tic_ratio,
  metrics_ann$lab,
  p.adjust.method = "bonferroni",
  paired = TRUE
)
print(pairwise_ann)

# Agilent labs comparison
agilent_ann <- metrics_ann %>% filter(setup == "Agilent1290_Agilent")
agilent_diff_ann <- wilcox.test(bpc_tic_ratio ~ lab, data = agilent_ann, paired = TRUE)
cat("\n=== Wilcoxon Test: Afekta vs Cembio (Annotated Data) ===\n")
print(agilent_diff_ann)
```

```{r clean-ann}
rm(tic_ann, bpc_ann, sp_full_ann)
```

---

# Noise Analysis

Compute noise/background signal metrics:

```{r noise-analysis}
# Reload data for noise calculation
load("object/tic_full.RData")
tic_full <- tic_full[tic_full$polarity == pol]

load("object/tic_detect.RData")
tic_detect <- tic_detect[tic_detect$polarity == pol]

load("object/tic_ann.RData")
tic_ann <- tic_ann[tic_ann$polarity == pol]

# Sum TIC for each level
stic_full <- vapply(intensity(tic_full), sum, numeric(1), na.rm = TRUE)
stic_detect <- vapply(intensity(tic_detect), sum, numeric(1), na.rm = TRUE)
stic_ann <- vapply(intensity(tic_ann), sum, numeric(1), na.rm = TRUE)

# Compute noise metrics
noise_metrics <- data.frame(
  lab = tic_full$lab,
  mixture = tic_full$mixture,
  
  # Background noise (full - detected)
  noise_baseline = stic_full - stic_detect,
  
  # Other signal (detected - annotated)
  noise_unannotated = stic_detect - stic_ann,
  
  # All non-annotated signal
  noise_total_other = stic_full - stic_ann,
  
  # Ratios
  ratio_detected_to_background = stic_detect / (stic_full - stic_detect),
  ratio_annotated_to_other = stic_ann / (stic_detect - stic_ann),
  ratio_annotated_to_full = stic_ann / stic_full
)

# Add setup information
noise_metrics <- noise_metrics %>%
  left_join(setup_simplified, by = "lab")

head(noise_metrics)
```

## Distribution Assessment - Noise

```{r dist-assessment-noise}
noise_metrics_to_assess <- c("noise_baseline", "noise_unannotated", "noise_total_other",
                              "ratio_detected_to_background", "ratio_annotated_to_other", 
                              "ratio_annotated_to_full")

distribution_summary_noise <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in noise_metrics_to_assess) {
  data_vec <- noise_metrics[[m]]
  data_vec <- data_vec[!is.na(data_vec) & data_vec > 0]
  
  if (length(data_vec) < 3) next  # Skip if too few observations
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)
  shapiro_log <- shapiro.test(log_data)
  
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_noise <- rbind(distribution_summary_noise, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_noise
```

## Distribution Plots - Noise Metrics

Violin plots and histograms for all noise metrics:

```{r plots-noise, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_noise <- noise_metrics %>%
  pivot_longer(
    cols = all_of(noise_metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value) & is.finite(value) & value > 0)

# Violin plots by lab
ggplot(plot_data_noise, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Noise Metrics - Distribution by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_noise$lab)) {
  p <- ggplot(plot_data_noise %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Noise Metrics -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab
for (lab_name in unique(plot_data_noise$lab)) {
  p <- ggplot(plot_data_noise %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Noise Metrics -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Noise

```{r lm-noise}
# Remove rows with infinite or NA values in ratios
noise_metrics_clean <- noise_metrics %>%
  filter(is.finite(ratio_detected_to_background) & 
         is.finite(ratio_annotated_to_other) &
         noise_baseline > 0 & noise_unannotated > 0)

# Baseline noise model
lm_baseline <- lm(log(noise_baseline) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Baseline Noise (Full - Detected) ===\n")
summary(lm_baseline)

# Unannotated signal model
lm_unannotated <- lm(log(noise_unannotated) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Unannotated Signal (Detected - Annotated) ===\n")
summary(lm_unannotated)

# Ratio: detected to background
lm_ratio_det_bg <- lm(log(ratio_detected_to_background) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Ratio: Detected / Background ===\n")
summary(lm_ratio_det_bg)

# Ratio: annotated to other
lm_ratio_ann_other <- lm(log(ratio_annotated_to_other) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Ratio: Annotated / Other Detected ===\n")
summary(lm_ratio_ann_other)
```

```{r clean-noise}
rm(tic_full, tic_detect, tic_ann)
```
