---
title: "Global Signal Analysis - Positive Mode (Simplified)"
format: html
editor_options:
  chunk_output_type: console
---

# Introduction

This simplified analysis computes key metrics from preprocessed LC-MS data to compare across different labs and instrument setups. The analysis focuses on:

- **Base Peak Chromatogram (BPC)**: Sum of peak apices
- **Total Ion Current (TIC)**: Sum of all peak areas
- **BPC/TIC Ratio**: Metric for peak purity/complexity
- **TIC Quantiles**: Distribution assessment (Q25, Q50, Q75)

For each metric, the analysis evaluates data distribution to determine if transformation is needed before statistical analysis, then fits linear models: `Metric ~ lab + mixture + setup`

The analysis is performed on three data levels:
1. **Full Data**: All detected signal (includes baseline noise)
2. **Detected Data**: Signal from peak picking (excludes baseline noise)
3. **Annotated Data**: Only annotated standard compounds

Setup information is derived from the LC and MS instruments in the lab metadata.

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggbeeswarm)
library(nortest)
library(car)
library(broom)
library(here)

base_path <- here("5_downstream_analysis")
```

```{r, cores-cluster}
# bpparam <- MulticoreParam(workers = parallel::detectCores() - 1)
```

## Setup Information

Load and prepare the setup metadata:

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"), skip = 1)

# Create a combined setup variable from LC and MS instruments
setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`) %>%
  distinct() %>%
  mutate(
    lc_short = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent1290",
      grepl("Acquity", `LC.Instrument`) ~ "Acquity",
      grepl("ExionLC", `LC.Instrument`) ~ "ExionLC",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_short = case_when(
      grepl("6540|6545", `MS.Instrument`) ~ "Agilent",
      grepl("Synapt", `MS.Instrument`) ~ "Waters",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    setup = paste(lc_short, ms_short, sep = "_")
  ) %>%
  select(lab, setup)

setup_simplified
```

# Full Data Analysis

Load the full preprocessed data:

```{r}
pol <- 1L  # positive polarity

load(file.path(base_path, "object/sp_full.RData"))
sp_full <- sp_full[sp_full$polarity == pol]

load(file.path(base_path, "object/bpc_full.RData"))
bpc_full <- bpc_full[bpc_full$polarity == pol]

load(file.path(base_path, "object/tic_full.RData"))
tic_full <- tic_full[tic_full$polarity == pol]
```

## Compute Metrics

```{r compute-full-metrics}
metrics_full <- data.frame(
  lab = bpc_full$lab,
  mixture = bpc_full$mixture,
  bpc = vapply(intensity(bpc_full), sum, numeric(1), na.rm = TRUE), ## sum or median, i'd say sum especially to compare instrument perf
  tic = vapply(intensity(tic_full), sum, numeric(1), na.rm = TRUE)
)

metrics_full$bpc_tic_ratio <- metrics_full$bpc / metrics_full$tic

# Add TIC quantiles
tic_quantiles <- lapply(intensity(tic_full), function(x) {
  quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
})
metrics_full$tic_q25 <- sapply(tic_quantiles, `[`, 1)
metrics_full$tic_q50 <- sapply(tic_quantiles, `[`, 2)
metrics_full$tic_q75 <- sapply(tic_quantiles, `[`, 3)

# Add setup information
metrics_full <- metrics_full %>%
  left_join(setup_simplified, by = "lab")

head(metrics_full)
```

## Distribution Assessment

For each metric, assess normality to determine if transformation is needed:

```{r dist-assessment-full}
metrics_to_assess <- c("bpc", "tic", "bpc_tic_ratio", "tic_q25", "tic_q50", "tic_q75")

distribution_summary_full <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in metrics_to_assess) {
  data_vec <- metrics_full[[m]]
  data_vec <- data_vec[!is.na(data_vec)]
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)  # Add 1 to handle potential zeros
  shapiro_log <- shapiro.test(log_data)
  
  # Recommend transform if log is more normal (p > 0.05)
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_full <- rbind(distribution_summary_full, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_full
```

## Distribution Plots - Full Data

Violin plots and histograms for all metrics:

```{r plots-full, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_full <- metrics_full %>%
  pivot_longer(
    cols = all_of(metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

# Violin plots by lab
ggplot(plot_data_full, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Full Data - Distribution of Metrics by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_full$lab)) {
  p <- ggplot(plot_data_full %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Full Data -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab (where applicable)
plot_data_full_log <- plot_data_full %>%
  filter(metric %in% c("bpc", "tic", "tic_q25", "tic_q50", "tic_q75"))

for (lab_name in unique(plot_data_full_log$lab)) {
  p <- ggplot(plot_data_full_log %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Full Data -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Full Data

```{r lm-full}
# BPC model
lm_bpc_full <- lm(log(bpc) ~ lab + setup, data = metrics_full)
cat("\n=== BPC Full Data ===\n")
summary(lm_bpc_full)

# TIC model
lm_tic_full <- lm(log(tic) ~ lab + setup, data = metrics_full)
cat("\n=== TIC Full Data ===\n")
summary(lm_tic_full)

# BPC/TIC ratio model
lm_ratio_full <- lm(bpc_tic_ratio ~ lab + setup, data = metrics_full)
cat("\n=== BPC/TIC Ratio Full Data ===\n")
summary(lm_ratio_full)

# TIC Q50 model
lm_tic_q50_full <- lm(log(tic_q50) ~ lab + setup, data = metrics_full)
cat("\n=== TIC Q50 Full Data ===\n")
summary(lm_tic_q50_full)
```

```{r clean-full}
rm(tic_full, bpc_full, sp_full)
```

---

# Detected Signal Analysis

Load the detected signal data:

```{r load-detect}
load(file.path(base_path, "object/sp_full_detect.RData"))
sp_full_detect <- sp_full_detect[sp_full_detect$polarity == pol]

load(file.path(base_path, "object/bpc_detect.RData"))
bpc_detect <- bpc_detect[bpc_detect$polarity == pol]

load(file.path(base_path, "object/tic_detect.RData"))
tic_detect <- tic_detect[tic_detect$polarity == pol]
```

## Compute Metrics

```{r compute-detect-metrics}
metrics_detect <- data.frame(
  lab = bpc_detect$lab,
  mixture = bpc_detect$mixture,
  bpc = vapply(intensity(bpc_detect), sum, numeric(1), na.rm = TRUE),
  tic = vapply(intensity(tic_detect), sum, numeric(1), na.rm = TRUE)
)

metrics_detect$bpc_tic_ratio <- metrics_detect$bpc / metrics_detect$tic

# Add TIC quantiles
tic_quantiles <- lapply(intensity(tic_detect), function(x) {
  quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
})
metrics_detect$tic_q25 <- sapply(tic_quantiles, `[`, 1)
metrics_detect$tic_q50 <- sapply(tic_quantiles, `[`, 2)
metrics_detect$tic_q75 <- sapply(tic_quantiles, `[`, 3)

# Add setup information
metrics_detect <- metrics_detect %>%
  left_join(setup_simplified, by = "lab")

head(metrics_detect)
```

## Distribution Assessment

```{r dist-assessment-detect}
distribution_summary_detect <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in metrics_to_assess) {
  data_vec <- metrics_detect[[m]]
  data_vec <- data_vec[!is.na(data_vec)]
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)
  shapiro_log <- shapiro.test(log_data)
  
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_detect <- rbind(distribution_summary_detect, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_detect
```

## Distribution Plots - Detected Data

Violin plots and histograms for all metrics:

```{r plots-detect, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_detect <- metrics_detect %>%
  pivot_longer(
    cols = all_of(metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

# Violin plots by lab
ggplot(plot_data_detect, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Detected Data - Distribution of Metrics by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_detect$lab)) {
  p <- ggplot(plot_data_detect %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Detected Data -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab (where applicable)
plot_data_detect_log <- plot_data_detect %>%
  filter(metric %in% c("bpc", "tic", "tic_q25", "tic_q50", "tic_q75"))

for (lab_name in unique(plot_data_detect_log$lab)) {
  p <- ggplot(plot_data_detect_log %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Detected Data -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Detected Data

```{r lm-detect}
# BPC model
lm_bpc_detect <- lm(log(bpc) ~ lab + setup, data = metrics_detect)
cat("\n=== BPC Detected Data ===\n")
summary(lm_bpc_detect)

# TIC model
lm_tic_detect <- lm(log(tic) ~ lab + setup, data = metrics_detect)
cat("\n=== TIC Detected Data ===\n")
summary(lm_tic_detect)

# BPC/TIC ratio model
lm_ratio_detect <- lm(bpc_tic_ratio ~ lab + setup, data = metrics_detect)
cat("\n=== BPC/TIC Ratio Detected Data ===\n")
summary(lm_ratio_detect)

# TIC Q50 model
lm_tic_q50_detect <- lm(log(tic_q50) ~ lab + setup, data = metrics_detect)
cat("\n=== TIC Q50 Detected Data ===\n")
summary(lm_tic_q50_detect)
```

```{r clean-detect}
rm(tic_detect, bpc_detect, sp_full_detect)
```

---

# Annotated Signal Analysis

Load the annotated signal data:

```{r load-ann}
load(file.path(base_path, "object/sp_full_ann.RData"))
sp_full_ann <- sp_full_ann[sp_full_ann$polarity == pol]

load(file.path(base_path, "object/bpc_ann.RData"))
bpc_ann <- bpc_ann[bpc_ann$polarity == pol]

load(file.path(base_path, "object/tic_ann.RData"))
tic_ann <- tic_ann[tic_ann$polarity == pol]
```

## Compute Metrics

```{r compute-ann-metrics}
metrics_ann <- data.frame(
  lab = bpc_ann$lab,
  mixture = bpc_ann$mixture,
  bpc = vapply(intensity(bpc_ann), sum, numeric(1), na.rm = TRUE),
  tic = vapply(intensity(tic_ann), sum, numeric(1), na.rm = TRUE)
)

metrics_ann$bpc_tic_ratio <- metrics_ann$bpc / metrics_ann$tic

# Add TIC quantiles
tic_quantiles <- lapply(intensity(tic_ann), function(x) {
  quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
})
metrics_ann$tic_q25 <- sapply(tic_quantiles, `[`, 1)
metrics_ann$tic_q50 <- sapply(tic_quantiles, `[`, 2)
metrics_ann$tic_q75 <- sapply(tic_quantiles, `[`, 3)

# Add setup information
metrics_ann <- metrics_ann %>%
  left_join(setup_simplified, by = "lab")

head(metrics_ann)
```

## Distribution Assessment

```{r dist-assessment-ann}
distribution_summary_ann <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in metrics_to_assess) {
  data_vec <- metrics_ann[[m]]
  data_vec <- data_vec[!is.na(data_vec)]
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)
  shapiro_log <- shapiro.test(log_data)
  
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_ann <- rbind(distribution_summary_ann, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_ann
```

## Distribution Plots - Annotated Data

Violin plots and histograms for all metrics:

```{r plots-ann, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_ann <- metrics_ann %>%
  pivot_longer(
    cols = all_of(metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

# Violin plots by lab
ggplot(plot_data_ann, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Annotated Data - Distribution of Metrics by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_ann$lab)) {
  p <- ggplot(plot_data_ann %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Annotated Data -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab (where applicable)
plot_data_ann_log <- plot_data_ann %>%
  filter(metric %in% c("bpc", "tic", "tic_q25", "tic_q50", "tic_q75"))

for (lab_name in unique(plot_data_ann_log$lab)) {
  p <- ggplot(plot_data_ann_log %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Annotated Data -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Annotated Data

```{r lm-ann}
# BPC model
lm_bpc_ann <- lm(log(bpc) ~ lab + setup, data = metrics_ann)
cat("\n=== BPC Annotated Data ===\n")
summary(lm_bpc_ann)

# TIC model
lm_tic_ann <- lm(log(tic) ~ lab + setup, data = metrics_ann)
cat("\n=== TIC Annotated Data ===\n")
summary(lm_tic_ann)

# BPC/TIC ratio model
lm_ratio_ann <- lm(bpc_tic_ratio ~ lab + setup, data = metrics_ann)
cat("\n=== BPC/TIC Ratio Annotated Data ===\n")
summary(lm_ratio_ann)

# TIC Q50 model
lm_tic_q50_ann <- lm(log(tic_q50) ~ lab + setup, data = metrics_ann)
cat("\n=== TIC Q50 Annotated Data ===\n")
summary(lm_tic_q50_ann)
```

```{r clean-ann}
rm(tic_ann, bpc_ann, sp_full_ann)
```

---

# Noise Analysis

Compute noise/background signal metrics:

```{r noise-analysis}
# Reload data for noise calculation
load("object/tic_full.RData")
tic_full <- tic_full[tic_full$polarity == pol]

load("object/tic_detect.RData")
tic_detect <- tic_detect[tic_detect$polarity == pol]

load("object/tic_ann.RData")
tic_ann <- tic_ann[tic_ann$polarity == pol]

# Sum TIC for each level
stic_full <- vapply(intensity(tic_full), sum, numeric(1), na.rm = TRUE)
stic_detect <- vapply(intensity(tic_detect), sum, numeric(1), na.rm = TRUE)
stic_ann <- vapply(intensity(tic_ann), sum, numeric(1), na.rm = TRUE)

# Compute noise metrics
noise_metrics <- data.frame(
  lab = tic_full$lab,
  mixture = tic_full$mixture,
  
  # Background noise (full - detected)
  noise_baseline = stic_full - stic_detect,
  
  # Other signal (detected - annotated)
  noise_unannotated = stic_detect - stic_ann,
  
  # All non-annotated signal
  noise_total_other = stic_full - stic_ann,
  
  # Ratios
  ratio_detected_to_background = stic_detect / (stic_full - stic_detect),
  ratio_annotated_to_other = stic_ann / (stic_detect - stic_ann),
  ratio_annotated_to_full = stic_ann / stic_full
)

# Add setup information
noise_metrics <- noise_metrics %>%
  left_join(setup_simplified, by = "lab")

head(noise_metrics)
```

## Distribution Assessment - Noise

```{r dist-assessment-noise}
noise_metrics_to_assess <- c("noise_baseline", "noise_unannotated", "noise_total_other",
                              "ratio_detected_to_background", "ratio_annotated_to_other", 
                              "ratio_annotated_to_full")

distribution_summary_noise <- data.frame(
  metric = character(),
  n = numeric(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  shapiro_p = numeric(),
  log_shapiro_p = numeric(),
  transform_recommended = character(),
  stringsAsFactors = FALSE
)

for (m in noise_metrics_to_assess) {
  data_vec <- noise_metrics[[m]]
  data_vec <- data_vec[!is.na(data_vec) & data_vec > 0]
  
  if (length(data_vec) < 3) next  # Skip if too few observations
  
  skewness <- (mean(data_vec) - median(data_vec)) / sd(data_vec)
  shapiro_test <- shapiro.test(data_vec)
  log_data <- log(data_vec + 1)
  shapiro_log <- shapiro.test(log_data)
  
  transform_rec <- ifelse(shapiro_log$p.value > shapiro_test$p.value, "log", "none")
  
  distribution_summary_noise <- rbind(distribution_summary_noise, data.frame(
    metric = m,
    n = length(data_vec),
    mean = mean(data_vec),
    sd = sd(data_vec),
    skewness = skewness,
    shapiro_p = round(shapiro_test$p.value, 4),
    log_shapiro_p = round(shapiro_log$p.value, 4),
    transform_recommended = transform_rec
  ))
}

distribution_summary_noise
```

## Distribution Plots - Noise Metrics

Violin plots and histograms for all noise metrics:

```{r plots-noise, fig.width=14, fig.height=12}
# Prepare data for plotting
plot_data_noise <- noise_metrics %>%
  pivot_longer(
    cols = all_of(noise_metrics_to_assess),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value) & is.finite(value) & value > 0)

# Violin plots by lab
ggplot(plot_data_noise, aes(x = lab, y = value, fill = lab)) +
  geom_violin(alpha = 0.5) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    title = "Noise Metrics - Distribution by Laboratory",
    x = "Laboratory",
    y = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histograms by lab (raw scale) - one plot per lab
for (lab_name in unique(plot_data_noise$lab)) {
  p <- ggplot(plot_data_noise %>% filter(lab == lab_name), aes(x = value, fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Noise Metrics -", lab_name, "- Distribution Histograms (Raw Scale)"),
      x = "Value",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}

# Histograms on log scale by lab
for (lab_name in unique(plot_data_noise$lab)) {
  p <- ggplot(plot_data_noise %>% filter(lab == lab_name), aes(x = log(value + 1), fill = metric)) +
    geom_histogram(bins = 20, color = "black", alpha = 0.7) +
    facet_wrap(~metric, scales = "free", ncol = 3) +
    labs(
      title = paste("Noise Metrics -", lab_name, "- Distribution Histograms (Log Scale)"),
      x = "Log(Value + 1)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
}
```

## Linear Models - Noise

```{r lm-noise}
# Remove rows with infinite or NA values in ratios
noise_metrics_clean <- noise_metrics %>%
  filter(is.finite(ratio_detected_to_background) & 
         is.finite(ratio_annotated_to_other) &
         noise_baseline > 0 & noise_unannotated > 0)

# Baseline noise model
lm_baseline <- lm(log(noise_baseline) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Baseline Noise (Full - Detected) ===\n")
summary(lm_baseline)

# Unannotated signal model
lm_unannotated <- lm(log(noise_unannotated) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Unannotated Signal (Detected - Annotated) ===\n")
summary(lm_unannotated)

# Ratio: detected to background
lm_ratio_det_bg <- lm(log(ratio_detected_to_background) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Ratio: Detected / Background ===\n")
summary(lm_ratio_det_bg)

# Ratio: annotated to other
lm_ratio_ann_other <- lm(log(ratio_annotated_to_other) ~ lab + setup, data = noise_metrics_clean)
cat("\n=== Ratio: Annotated / Other Detected ===\n")
summary(lm_ratio_ann_other)
```

```{r clean-noise}
rm(tic_full, tic_detect, tic_ann)
```
