---
title: "Global analysis"
format: html
---

# Introduction

This analysis evaluates QC metrics across 4 laboratories using positive
and negative ionization modes. Instrumental setup, sample preparation, and
data processing all affect chromatographic metrics on both full and
detected signal data.

Data: 38 standard mixtures (HUMAN endosome, different compounds) from 4
labs with harmonized LC systems.

Limitations: No technical replicates; laboratory identity is confounded
with instrumental parameters. Observed differences reflect combined effects
of lab environment, operator expertise, and equipment configuration.


## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)

base_path <- here("5_downstream_analysis")
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`, `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) %>%
  distinct() %>%
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) %>%
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

Each lab's protocol is unique; lab identity is completely confounded with
instrumental setup. Notable variations: MS1 scan rate, injection volume, and
solvent brand.

# Metrics

Three main chromatographic metrics:
- **TIC**: Sum of all ion intensities
- **BPC**: Max intensity per scan
- **BPC/TIC Ratio**: Peak sharpness; naturally normalized

Two data levels:
- **Full Signal**: Raw profile including baseline
- **Detected Signal**: Peaks from centWave; isolates compound signal from
  baseline

```{r load-data-both-modes}
# Function to compute metrics for a given polarity and data level
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  # Load BPC and TIC data based on data level
  if (data_level == "full") {
    load(file.path(base_path, "object/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]

    load(file.path(base_path, "object/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "object/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]

    load(file.path(base_path, "object/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  # Compute base metrics
  metrics <- data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  )

  metrics$bpc_tic_ratio <- metrics$bpc / metrics$tic

  # Add setup information and data level
  metrics <- metrics %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = ifelse(data_level == "full", "Full Signal", "Detected Signal")) %>%
    group_by(lab, polarity) %>%
    mutate(
      # Calculate the median signal for this specific lab
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),

      # Normalize to median (center around 1.0)
      bpc_norm = bpc / median_bpc,
      tic_norm = tic / median_tic,
      bpc_tic_ratio_norm = bpc_tic_ratio
    ) %>%
    ungroup()

  return(metrics)
}

# Helper: Run function for both polarities and combine results
run_for_polarities <- function(fn, data_level = NULL) {
  pos_result <- fn(1L, "Positive", data_level)
  neg_result <- fn(0L, "Negative", data_level)
  bind_rows(pos_result, neg_result)
}

# Helper: Run function for both polarities with polarity name (no data_level)
run_for_polarities_names <- function(fn) {
  pos_result <- fn(1L, "Positive")
  neg_result <- fn(0L, "Negative")
  bind_rows(pos_result, neg_result)
}

# Helper: Run function for both polarities (single polarity argument)
run_for_polarities_simple <- function(fn) {
  pos_result <- fn(1L)
  neg_result <- fn(0L)
  bind_rows(pos_result, neg_result)
}

# Compute metrics for both data levels
metrics_full <- run_for_polarities(compute_metrics, "full")
metrics_detect <- run_for_polarities(compute_metrics, "detect")

# Combine all
metrics_combined <- bind_rows(metrics_full, metrics_detect) %>%
  group_by(lab, polarity, data_level) %>%
  mutate(
    # Median scaling (divide by median)
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),

    # Autoscaling (mean center, then divide by MAD)
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE),

    # Apply scaling to ratio as well
    bpc_tic_ratio_median_scaled = bpc_tic_ratio / median(bpc_tic_ratio, na.rm = TRUE),
    bpc_tic_ratio_autoscaled = (bpc_tic_ratio - mean(bpc_tic_ratio, na.rm = TRUE)) / mad(bpc_tic_ratio, na.rm = TRUE)
  ) %>%
  ungroup()
```


# Data distribution

I want to plot violin plots for the distribution of raw metrics.

```{r}
# Plotting function
plot_violin <- function(data, metric) {
  # Log-transform absolute intensity metrics
  if (metric %in% c("bpc", "tic")) {
    p <- ggplot(data, aes_string(x = "lab", y = metric, fill = "polarity")) +
      geom_violin(trim = FALSE, alpha = 0.6) +
      geom_jitter(aes(color = as.factor(mixture)), width = 0.2, size = 1.5, 
      alpha = 0.7) +
      scale_color_viridis_d(option = "turbo") +
      scale_y_continuous(trans = "log2") +
      facet_wrap(~ data_level) +
      theme_minimal() +
      labs(title = paste("Violin plot of", metric, "(log2 scale)"),
           x = "Laboratory",
           y = paste(metric, "(log2)"),
           color = "Mixture")
  } else {
    p <- ggplot(data, aes_string(x = "lab", y = metric, fill = "polarity")) +
      geom_violin(trim = FALSE, alpha = 0.6) +
      geom_jitter(aes(color = as.factor(mixture)), width = 0.2, size = 1.5, alpha = 0.7) +
      scale_color_viridis_d(option = "turbo") +
      facet_wrap(~ data_level) +
      theme_minimal() +
      labs(title = paste("Violin plot of", metric),
           x = "Laboratory",
           y = metric,
           color = "Mixture")
  }
  return(p)
}

# Generate and display plots for each metric separately
for (metric in c("bpc", "tic", "bpc_tic_ratio")) {
  print(plot_violin(metrics_combined, metric))
}
```

Detected signal recenters the negative and positive distributions. The bimodal shape is evident:

```{r}
ggplot(metrics_combined %>% filter(data_level == "Full Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_minimal()

ggplot(metrics_combined %>% filter(data_level == "Detected Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_minimal()
```

Full signal is more compressed than detected signal, likely because
each mixture has different numbers of compounds.

 ```{r}
plot_mixture_distribution <- function(data, metric) {
  if (metric %in% c("bpc", "tic")) {
    p <- ggplot(
      data,
      aes_string(x = "mixture", y = metric, color = "lab", group = "lab")
    ) +
      geom_line(alpha = 0.7) +
      scale_y_continuous(trans = "log2") +
      facet_grid(data_level ~ polarity) +
      theme_minimal() +
      labs(
        title = paste("Mixture distribution across labs -", metric, "(log2)")
      )
  } else {
    p <- ggplot(
      data,
      aes_string(x = "mixture", y = metric, color = "lab", group = "lab")
    ) +
      geom_line(alpha = 0.7) +
      facet_grid(data_level ~ polarity) +
      theme_minimal() +
      labs(title = paste("Mixture distribution across labs -", metric))
  }
  return(p)
}

for (metric in c("bpc", "tic", "bpc_tic_ratio")) {
  print(plot_mixture_distribution(metrics_combined, metric))
}
 ```

There is good agreement between Cembio and Afekta, and some with ICL.
However, HMGU is consistently lower intensity. Let's run linear models to
evaluate mixture effects and batch effects.

Compare distribution before and after normalization:

```{r}
## run simple linear model
testdata <- metrics_combined %>%
  filter(polarity == "Positive", data_level == "Full Signal")

model <- lm(log2(tic) ~ lab + mixture, data = testdata)

plot(model$coefficients) ## add some legend ?
plot(model$residuals) ## here need to chekc hte dsitribution is random and the R2 is high .
```

ANOVA aggregates coefficients into Sum of Squares (SS) to quantify
the relative power of each factor. Here, lab variance dominates while
mixture variance is much lower.

```{r}
an <- anova(model)

## make a lil pie chart
ss_total <- sum(an[, "Sum Sq"])
ss_lab <- (an["lab", "Sum Sq"] / ss_total) * 100
ss_mix <- (an["mixture", "Sum Sq"] / ss_total) * 100
ss_res <- (an["Residuals", "Sum Sq"] / ss_total) * 100
# Prepare data for pie chart
pie_data <- data.frame(
  Source = c("Lab Variance", "Mixture Variance", "Residual Variance"),
  Percentage = c((ss_lab / ss_total) * 100, (ss_mix / ss_total) * 100, (ss_res / ss_total) * 100)
)

# Plot pie chart
ggplot(pie_data, aes(x = "", y = Percentage, fill = Source)) +
  geom_col() +
  coord_polar(theta = "y") +
  labs(title = "Variance Contribution by Source") +
  theme_void()
```

Raw TIC shows massive lab variance, making it unsuitable for comparing
biology across labs.

```{r}
# Runs on TIC across all data levels and polarities
check_metric_quality <- function(data, metric_name) {
  # Run linear model with log2 transformation for intensity metrics
  if (metric_name %in% c("tic", "bpc")) {
    model <- lm(log2(get(metric_name)) ~ lab + mixture, data = data)
  } else {
    model <- lm(get(metric_name) ~ lab + mixture, data = data)
  }

  # Calculate R-squared
  r_squared <- summary(model)$r.squared

  # ANOVA for variance decomposition
  an <- anova(model)
  ss_total <- sum(an[, "Sum Sq"])

  # Extract variance components
  ss_lab <- as.numeric(an["lab", "Sum Sq"])
  ss_mix <- as.numeric(an["mixture", "Sum Sq"])
  ss_res <- as.numeric(an["Residuals", "Sum Sq"])

  # Percentages
  pct_lab <- (ss_lab / ss_total) * 100
  pct_mix <- (ss_mix / ss_total) * 100
  pct_res <- (ss_res / ss_total) * 100

  # Create variance data
  variance_data <- data.frame(
    Source = c("Lab Variance", "Mixture Variance", "Residual Variance"),
    Percentage = c(pct_lab, pct_mix, pct_res)
  )

  # Residual plot
  residual_plot <- data.frame(
    Fitted = fitted(model),
    Residuals = residuals(model)
  ) %>%
    ggplot(aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(
      title = paste("Residual Plot -", metric_name),
      x = "Fitted Values",
      y = "Residuals"
    ) +
    theme_minimal()

  # Return list with all outputs
  list(
    model = model,
    r_squared = r_squared,
    residual_plot = residual_plot,
    variance_table = variance_data
  )
}

# Run analysis for TIC across all data levels and polarities
results_table <- data.frame()
residual_plots <- list()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>%
      filter(polarity == pol, data_level == level)

    message("Processing: ", pol, " - ", level)
    result <- check_metric_quality(data_subset, "tic")

    # Store residual plots
    residual_plots[[paste(pol, level, sep = " - ")]] <- result$residual_plot

    # Store summary in table
    results_table <- rbind(results_table, data.frame(
      Polarity = pol,
      Data_Level = level,
      R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}

# Prepare data for stacked bar chart
variance_long <- results_table %>%
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) %>%
  pivot_longer(
    cols = ends_with("_Pct"),
    names_to = "Source",
    values_to = "Percentage"
  ) %>%
  mutate(
    Source = gsub("_Variance_Pct", "", Source),
    Source = gsub("_Pct", "", Source),
    Label = paste(Polarity, Data_Level, sep = "\n"),
    R2_label = paste("R² =", round(R_squared, 3))
  )

# Create text labels data (one per condition)
text_labels <- variance_long %>%
  distinct(Label, R2_label)

# Create single stacked bar chart with all 4 combinations
stacked_bar_plot <- ggplot(
  variance_long,
  aes(x = Label, y = Percentage, fill = Source)
) +
  geom_col(alpha = 0.8) +
  geom_text(
    aes(x = Label, label = R2_label, y = 105),
    data = text_labels,
    size = 3,
    inherit.aes = FALSE
  ) +
  labs(
    title = "Variance Decomposition: TIC across all conditions",
    x = "Polarity & Data Level",
    y = "Variance Contribution (%)",
    fill = "Source"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    legend.position = "top"
  ) +
  ylim(0, 110)

print(stacked_bar_plot)

# Create combined residual plots with all 4 conditions
residual_data <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>%
      filter(polarity == pol, data_level == level)

    result <- check_metric_quality(data_subset, "tic")

    # Extract residuals and add condition info, including lab
    residual_df <- data.frame(
      Fitted = fitted(result$model),
      Residuals = residuals(result$model),
      Polarity = pol,
      Data_Level = level,
      Lab = data_subset$lab
    )

    residual_data <- rbind(residual_data, residual_df)
  }
}

# Create faceted residual plots with lab legend
combined_residual_plot <- ggplot(
  residual_data,
  aes(x = Fitted, y = Residuals, shape = Lab)
) +
  geom_point(alpha = 0.6, size = 2.5, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ paste(Polarity, Data_Level, sep = "\n"), nrow = 2, ncol = 2) +
  labs(
    title = "Faceted Residual Plots: TIC across all conditions",
    x = "Fitted Values",
    y = "Residuals",
    shape = "Laboratory"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 10)
  )

print(combined_residual_plot)

# Display summary table
print(results_table)
```

The residual plot clearly shows separation between labs vs HMGU and
still lab separation in others. Lab separation remains unexplained by
mixture. We cannot compare labs directly.

Next: Apply normalization or standardization across labs to account for
these differences. Test methods and examine effects on linear models.

What could be tested:
- Median normalization: Adjusting each lab's data to have the same median could
 help account for systematic differences. In that case we take the median of 
 that measure by that lab to correct each mixture.
    - this is safe as the experimental design is balanced. every lab measures
     the same 38 mixture, heir intensity distributions should be identical in 
     shape, just shifted up or down.
    - this would be wrong if the mixture were different across labs
- Auto scaling: forces the lab to have the same dynamic change.
    - stronger scaling. will remove any instrument bias.

What i am not testing:
- max: ESI are subject to random electrical spike/detector saturation, 
       therefore max could be an outlier and crush other signal to 0
    - generally not used for entire run.

I can decided what would be better also based on violin plot shape (that i made
 above actually).

- if similar : median scaling should be fine
- if not : auto-scaling may be required

The violin plot and density plot above actually are a bit hard to be definitive
about. The positive data is fairly similar in shape, but the negative data
shows more variation. I would test both and see how the model are affected. 
and if we manage to drop the variance due to lab to acceptable levels.

Let's see the differnece in results between the different scaling methods.

```{r}
# Visualize distribution before and after scaling
# Create violin plots comparing raw vs median scaled TIC
plot_scaling_comparison <- function(data, metric_raw, metric_scaled, metric_label) {
  # Prepare data for plotting - combine raw and scaled into long format
  plot_data <- data %>%
    select(lab, polarity, data_level, mixture,
           raw = all_of(metric_raw),
           scaled = all_of(metric_scaled)) %>%
    pivot_longer(cols = c(raw, scaled), names_to = "Scaling",
     values_to = "Value") %>%
    mutate(Scaling = ifelse(Scaling == "raw", "Unscaled", "Median Scaled"))

  # Create plot with free y-axis scales
  p <- ggplot(plot_data, aes(x = lab, y = log2(Value), fill = polarity)) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    geom_jitter(aes(color = as.factor(mixture)), width = 0.2, size = 1.5,
     alpha = 0.7) +
    scale_color_viridis_d(option = "turbo") +
    facet_grid(Scaling ~ data_level, scales = "free_y") +
    theme_minimal() +
    labs(title = paste("Before vs After Median Scaling:", metric_label),
         subtitle = "Distribution shape should remain similar, but scale shifts",
         x = "Laboratory",
         y = paste(metric_label, "(log2)"),
         color = "Mixture") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  return(p)
}

# Generate comparison plots for TIC and BPC
print(plot_scaling_comparison(metrics_combined, "tic", "tic_median_scaled", "TIC"))
print(plot_scaling_comparison(metrics_combined, "bpc", "bpc_median_scaled", "BPC"))

# Compare the different scaling methods
test_data <- metrics_combined %>%
  filter(polarity == "Positive", data_level == "Detected Signal")

# Run check_metric_quality on each normalization
result_raw <- check_metric_quality(test_data, "tic")
result_median_scaled <- check_metric_quality(test_data, "tic_median_scaled")
result_autoscaled <- check_metric_quality(test_data, "tic_autoscaled")

# Compile results (extracting variance_table and r_squared from each result list)
normalization_results <- bind_rows(
  result_raw$variance_table %>% mutate(Normalization = "Raw TIC", 
  R_squared = result_raw$r_squared),
  result_median_scaled$variance_table %>% mutate(Normalization = "Median Scaled", 
  R_squared = result_median_scaled$r_squared),
  result_autoscaled$variance_table %>% mutate(Normalization = "Autoscaled (MAD)", 
  R_squared = result_autoscaled$r_squared)
)

print("=== TIC Variance Decomposition: Effect of Normalization ===")
print(normalization_results)

# Visualize
normalization_results %>%
  ggplot(aes(x = Normalization, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  labs(title = "Effect of Normalization on Variance Decomposition (TIC)",
       subtitle = "Positive Mode, Detected Signal",
       x = "Normalization Method",
       y = "Variance Contribution (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Noise likely reflects injection variability and ionization
fluctuation—factors we cannot isolate with this experimental design.


```{r}
# Re-run the models to extract Absolute Sum of Squares and create residual plots
# (Assuming your previous code setup)

# 1. Raw Model
mod_raw <- lm(log2(tic) ~ lab + factor(mixture), data = test_data)
ss_raw <- anova(mod_raw)$"Sum Sq"

# 2. Median Scaled Model
mod_median_scaled <- lm(log2(tic_median_scaled) ~ lab + factor(mixture), data = test_data)
ss_median_scaled <- anova(mod_median_scaled)$"Sum Sq"

# 3. Autoscaled Model
mod_autoscaled <- lm(tic_autoscaled ~ lab + factor(mixture), data = test_data) ## no log bc scaled
ss_autoscaled <- anova(mod_autoscaled)$"Sum Sq"

# 4. Compare the "Noise" (Residuals are the 3rd item in ANOVA)
print(paste("Raw Absolute Noise:", round(ss_raw[3], 2)))
print(paste("Median Scaled Absolute Noise:", round(ss_median_scaled[3], 2)))
print(paste("Autoscaled Absolute Noise:", round(ss_autoscaled[3], 2)))

print(paste("Raw Total Variance:", round(sum(ss_raw), 2)))
print(paste("Median Scaled Total Variance:", round(sum(ss_median_scaled), 2)))
print(paste("Autoscaled Total Variance:", round(sum(ss_autoscaled), 2)))

# Plot residuals vs fitted for each normalization method
residual_plots_comparison <- list(
  raw = ggplot(data.frame(Fitted = fitted(mod_raw), Residuals = residuals(mod_raw)),
               aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    labs(title = "Raw TIC (log2)", x = "Fitted Values", y = "Residuals") +
    theme_minimal(),

  median_scaled = ggplot(data.frame(Fitted = fitted(mod_median_scaled), Residuals = residuals(mod_median_scaled)),
                         aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    labs(title = "Median Scaled TIC", x = "Fitted Values", y = "Residuals") +
    theme_minimal(),

  autoscaled = ggplot(data.frame(Fitted = fitted(mod_autoscaled), Residuals = residuals(mod_autoscaled)),
                      aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    labs(title = "Autoscaled TIC (MAD)", x = "Fitted Values", y = "Residuals") +
    theme_minimal()
)

# Display side by side (requires gridExtra or patchwork)
library(patchwork)
residual_plots_comparison$raw / residual_plots_comparison$median_scaled / residual_plots_comparison$autoscaled +
  plot_layout(ncol = 1) +
  plot_annotation(title = "Residual Plots: Comparison of Normalization Methods")

print(paste("Median Scaled Total Variance:", round(sum(ss_median_scaled), 2)))
```

Noise magnitude matters less than residual variance stability. Biological
variance (mixture) changed as expected.

Conclusions:
- Use median scaling
- Noise level acceptable; does not significantly impact biological
  variance
- Focus on biological signals without technical variability obscuring
  them

To focus on instrument variability, we remove the mixture effect:

```{r}
# 1. Isolate Lab Effects
#    Center each mixture at 0 to reveal only lab differences
inst_focus_data <- metrics_combined %>%
  filter(data_level == "Detected Signal") %>%
  group_by(mixture, polarity) %>%
  mutate(
    # Log2 transform, then subtract mixture median to isolate instrument variation
    log_tic = log2(tic),
    tic_mix_centered = log_tic - median(log_tic, na.rm = TRUE)
  ) %>%
  ungroup()

# 2. Fit Model
model <- lm(tic_mix_centered ~ lab + factor(mixture), data = inst_focus_data)
an <- anova(model)
total_ss <- sum(an$`Sum Sq`)

# 3. Create the Result Table
result <- data.frame(
  Source = c("Instrument (Lab)", "Biology (Mixture)", "Noise (Residual)"),
  Percentage = c(
    (an["lab", "Sum Sq"] / total_ss) * 100,
    (an["factor(mixture)", "Sum Sq"] / total_ss) * 100,
    (an["Residuals", "Sum Sq"] / total_ss) * 100
  )
)

print(result)

# Visualize instrument effect (biological signal removed)
ggplot(inst_focus_data, aes(x = lab, y = tic_mix_centered, fill = lab)) +
  geom_boxplot() +
  facet_wrap(~polarity) +
  labs(title = "Pure Instrument Variance",
       subtitle = "Relative sensitivity of each lab on log2 scale.",
       y = "Relative Sensitivity (Log2 Scale)") +
  theme_bw() +
  theme(legend.position = "none")
```

Mixture-centered analysis isolates instrument effects. No single metric
reveals both instrument and biological variance; side-by-side comparison
is needed.

# Reproducibility: Full Signal vs Detected

Normalization addresses sensitivity differences across labs, but
lab-dependent noise remains. The intraclass correlation coefficient (ICC)
tests this. ICC repeats earlier variance analysis but provides scale-
invariant comparability across papers—unlike raw Sum of Squares, it's
more meaningful for visualization.

ICC = variance_between_mixtures / (variance_between_mixtures +
variance_within_mixtures)

High ICC confirms measurements reflect biological differences, not
technical noise.

## Normalization Effects

```{r metric-stability-all-labs}
compute_icc <- function(data, metric_name) {
  # ICC = variance_between_mixtures / (variance_between_mixtures + variance_within_mixtures)
  # Log-transform absolute metrics (TIC, BPC); ratios use natural scale
  if (grepl("ratio", metric_name, ignore.case = TRUE)) {
    data$value <- data[[metric_name]] # Keep ratios on natural 0-1 scale
  } else {
    data$value <- log2(data[[metric_name]]) # Log-transform absolute intensities
  }
  # Calculate variances
  between_mixture_var <- data %>%
    group_by(mixture) %>%
    summarise(mean_val = mean(value, na.rm = TRUE), .groups = "drop") %>%
    pull(mean_val) %>%
    var(na.rm = TRUE)
  within_mixture_var <- data %>%
    group_by(mixture) %>%
    summarise(var_val = var(value, na.rm = TRUE), .groups = "drop") %>%
    pull(var_val) %>%
    mean(na.rm = TRUE)
  icc <- between_mixture_var / (between_mixture_var + within_mixture_var)
  return(data.frame(
    Metric = metric_name,
    ICC = icc,
    Between_mixture_var = between_mixture_var,
    Within_mixture_var = within_mixture_var
  ))
}

# Compute ICC for BOTH absolute and normalized metrics in one loop
icc_all_results <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal")) {
    data_subset <- metrics_combined %>%
      filter(polarity == pol, data_level == level)

    # Absolute metrics
    icc_tic_abs <- compute_icc(data_subset, "tic") %>% mutate(Type = "Absolute")
    icc_bpc_abs <- compute_icc(data_subset, "bpc") %>% mutate(Type = "Absolute")
    icc_ratio_abs <- compute_icc(data_subset, "bpc_tic_ratio") %>%
      mutate(Type = "Absolute")

    # Normalized metrics
    icc_tic_norm <- compute_icc(data_subset, "tic_norm") %>%
      mutate(Type = "Normalized", Metric = "tic")
    icc_bpc_norm <- compute_icc(data_subset, "bpc_norm") %>%
      mutate(Type = "Normalized", Metric = "bpc")
    icc_ratio_norm <- compute_icc(data_subset, "bpc_tic_ratio_norm") %>%
      mutate(Type = "Normalized", Metric = "bpc_tic_ratio")

    results <- bind_rows(
      icc_tic_abs,
      icc_bpc_abs,
      icc_ratio_abs,
      icc_tic_norm,
      icc_bpc_norm,
      icc_ratio_norm
    ) %>%
      mutate(
        Polarity = pol,
        Data_Level = level,
        Metric_Label = case_when(
          Metric == "tic" ~ "TIC",
          Metric == "bpc" ~ "BPC",
          Metric == "bpc_tic_ratio" ~ "BPC/TIC Ratio",
          TRUE ~ Metric
        )
      )

    icc_all_results <- bind_rows(icc_all_results, results)
  }
}

print(
  "=== Metric Reproducibility: Absolute vs Normalized (ICC: 0=poor, 1=perfect) ==="
)
icc_all_results %>%
  select(Polarity, Data_Level, Metric_Label, Type, ICC) %>%
  arrange(desc(ICC)) %>%
  print()
```


```{r icc-comparison-plots}
#| code-fold: true
# Direct comparison of absolute vs normalized ICC
ggplot(icc_all_results, aes(x = Metric_Label, y = ICC, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  facet_grid(Data_Level ~ Polarity) +
  labs(
    title = "Reproducibility Comparison: Absolute vs Normalized Metrics",
    subtitle = "ICC by Polarity and Data Level"
  )
```

```{r variance-decomposition}
#| code-fold: true
## Plot 3A: Variance decomposition for TIC and BPC (log2 scale)
variance_plot_intensity <- icc_all_results %>%
  filter(Metric_Label != "BPC/TIC Ratio") %>%
  select(Polarity, Data_Level, Metric_Label, Type, Between_mixture_var, Within_mixture_var) %>%
  pivot_longer(cols = c(Between_mixture_var, Within_mixture_var),
               names_to = "Source", values_to = "Variance") %>%
  mutate(
    Source = case_when(
      Source == "Between_mixture_var" ~ "Between mixtures (signal)",
      Source == "Within_mixture_var" ~ "Within mixtures (lab noise)",
      TRUE ~ Source
    )
  )

ggplot(variance_plot_intensity, aes(x = Metric_Label, y = Variance, fill = Source)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.7) +
  facet_grid(Data_Level + Type ~ Polarity, scales = "free_y") +
  labs(
    title = "Variance Decomposition: TIC and BPC",
    subtitle = "Higher 'between mixtures' = more signal; Higher 'within mixtures' = more lab noise",
    x = "Metric",
    y = "Variance (log2 scale)",
    fill = "Source of Variation"
  ) +
  scale_fill_manual(values = c("Between mixtures (signal)" = "#619CFF",
                                "Within mixtures (lab noise)" = "#F8766D")) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))

# Plot 3B: Variance decomposition for BPC/TIC Ratio (natural scale)
variance_plot_ratio <- icc_all_results %>%
  filter(Metric_Label == "BPC/TIC Ratio") %>%
  select(Polarity, Data_Level, Metric_Label, Type, Between_mixture_var, Within_mixture_var) %>%
  pivot_longer(cols = c(Between_mixture_var, Within_mixture_var),
               names_to = "Source", values_to = "Variance") %>%
  mutate(
    Source = case_when(
      Source == "Between_mixture_var" ~ "Between mixtures (signal)",
      Source == "Within_mixture_var" ~ "Within mixtures (lab noise)",
      TRUE ~ Source
    )
  )

ggplot(variance_plot_ratio, aes(x = Type, y = Variance, fill = Source)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.7) +
  facet_grid(Data_Level ~ Polarity, scales = "free_y") +
  labs(
    title = "Variance Decomposition: BPC/TIC Ratio",
    subtitle = "Ratio is on natural scale (0-1), not log-transformed",
    x = "Normalization Type",
    y = "Variance (natural scale)",
    fill = "Source of Variation"
  ) +
  scale_fill_manual(values = c("Between mixtures (signal)" = "#619CFF",
                                "Within mixtures (lab noise)" = "#F8766D")) +
  theme_bw() +
  theme(legend.position = "top")
```

The overall goal is that we end up with Value = Biology.

Starting point: Value = (Biology × Sensitivity) + Background Noise

With detected signal and normalization: Value ≈ Biology

Findings: Normalized TIC shows highest ICC. Detected signal (ICC 0.5–0.6)
substantially beats full signal (ICC 0.3) because peak detection removes
vendor-specific solvent noise. Median normalization improved ICC from 0.05
to 0.55—crossing the threshold from poor to moderate reproducibility. The
remaining ~0.5 ICC gap likely reflects ionization preferences and
platform-specific sensitivities.

```{r peak-detection-validation}
# Compare Reproducibility: Full vs Detected Signal (using the best metric: Normalized TIC)

# 1. ICC Improvement Table
icc_comparison_tic <- icc_all_results %>%
  filter(Type == "Normalized", Metric == "tic") %>%
  select(Polarity, Data_Level, ICC) %>%
  pivot_wider(names_from = Data_Level, values_from = ICC) %>%
  mutate(
    # Calculate how much better Detected is compared to Full
    ICC_Improvement = `Detected Signal` - `Full Signal`,
    Percent_Gain = (`Detected Signal` - `Full Signal`) / `Full Signal` * 100
  )

print("=== Value of Peak Detection: ICC Improvement (Normalized TIC) ===")
print(icc_comparison_tic)
```


```{r}
#| code-fold: true
# 2. Visualization: The "Before & After" Plot
ggplot(icc_all_results %>% filter(Type == "Normalized", Metric == "tic"),
       aes(x = Polarity, y = ICC, fill = Data_Level)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(ICC, 2)),
            position = position_dodge(0.9), vjust = -0.5) +
  labs(
    title = "Impact of Peak Detection on Reproducibility",
    subtitle = "Comparison of Normalized TIC Stability (Full vs. Detected)",
    y = "Intraclass Correlation Coefficient (ICC)",
    x = "Polarity",
    fill = "Processing Level",
    caption = "Higher ICC = Better Agreement between labs.\nPeak detection doubles reproducibility in Negative Mode."
  ) +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```

Peak detection substantially improves reproducibility:

- Negative mode: ICC increased from 0.275 to 0.556 (102% improvement)
- Positive mode: ICC increased from 0.375 to 0.503 (34% improvement)

Conclusion: All subsequent analyses use detected signal exclusively.
Peak detection removes instrument-specific baseline noise obscuring
biological signal. Full signal contains too much vendor-specific noise
for meaningful inter-lab comparisons.

For comparison, two labs (CEMBIO and AFEKTA) share the same LC-MS
platform (Agilent Q-TOF). ICC of 0.83 (TIC) and 0.92 (BPC) for these labs
proves selectivity matches when vendor matches. The 0.5–0.8 gap reflects
instrument-specific and ionization differences.

```{r}
# Compute ICC for Agilent Labs Only
agilent_labs <- c("afekta", "cembio")

# Filter metrics_combined to only include Agilent labs
agilent_data <- metrics_combined %>%
  filter(lab %in% agilent_labs, polarity == "Positive", data_level == "Detected Signal")

# Compute ICC for Agilent labs only
icc_agilent_tic <- compute_icc(agilent_data, "tic_norm")
icc_agilent_bpc <- compute_icc(agilent_data, "bpc_norm")

print("=== Ceiling ICC Estimate (Agilent Labs Only, Positive Mode, Detected Signal) ===")
print(paste("ICC for Normalized TIC:", round(icc_agilent_tic$ICC, 3)))
print(paste("ICC for Normalized BPC:", round(icc_agilent_bpc$ICC, 3)))
```

```{r agilent-effect-size-detected}
# Effect size between Agilent labs using normalized TIC and BPC (Detected Signal)
agilent_effect_size <- metrics_combined %>%
  filter(lab %in% c("afekta", "cembio"), data_level == "Detected Signal") %>%
  select(lab, mixture, polarity, tic_norm, bpc_norm) %>%
  pivot_wider(
    names_from = lab,
    values_from = c(tic_norm, bpc_norm)
  ) %>%
  mutate(
    log_diff_tic = log2(tic_norm_cembio / tic_norm_afekta),
    log_diff_bpc = log2(bpc_norm_cembio / bpc_norm_afekta)
  )

effect_summary <- agilent_effect_size %>%
  group_by(polarity) %>%
  summarise(
    TIC_Mean_Log = mean(log_diff_tic, na.rm = TRUE),
    TIC_Median_Log = median(log_diff_tic, na.rm = TRUE),
    TIC_Percent_Diff = (2^abs(TIC_Mean_Log) - 1) * 100,
    BPC_Mean_Log = mean(log_diff_bpc, na.rm = TRUE),
    BPC_Median_Log = median(log_diff_bpc, na.rm = TRUE),
    BPC_Percent_Diff = (2^abs(BPC_Mean_Log) - 1) * 100,
    .groups = "drop"
  )

effect_summary
```

PROOF THAT MEDIAN DIFF MEAN IN OUR DATA

TIC_Median_Log (-0.014):

Effectively zero. Median normalization worked perfectly for the
"typical" mixture. Labs agree within ~1%.

TIC_Mean_Log (-0.093):

Not zero (approx. 6.6% difference). The average is pulled down by
outliers.


Negative mode: Mean (-0.12) and Median (-0.13) nearly identical but
non-zero. Why? Distributions differ in shape. Even after aligning centers
(global median), different shapes create point-by-point offsets. Negative
mode ionization creates ~9% bias that median normalization cannot erase.

Positive mode discrepancy (Mean -0.09 vs Median -0.01) indicates skewness
in dynamic range. Negative mode shows consistent ~9% bias after
normalization.

Conclusion: Agilent Q-TOF generations (6546 vs 6545) can be compared for
trends (high ICC) but need non-linear correction for absolute quantitation.

Afekta (6546) has flatter distribution (smaller median-max difference).
Cembio (6545) has spikier distribution (larger difference).

**6545 (Cembio)**: "Swarm Autotune" optics; optimized for resolution &
sensitivity to small molecules.

**6546 (Afekta)**: "Simultaneous Dynamic Range"; sees bright AND dim
signals without saturation.

Why higher TIC in Afekta? 6546 has massive dynamic range (5 orders).
It captures low-level ions ("grass") and high-level ions ("trees"), seeing
more total biological signal.

Why higher ratio in Cembio? Ratio = Height / Area. Afekta's huge TIC
(wide dynamic range) makes denominator large, shrinking ratio. Cembio's
cleaner baseline (less background sensitivity) has lower TIC, inflating
the ratio.

"The systematic difference between the two Agilent labs can be attributed
to the specific architectural goals of the instrument generations. The
Agilent 6546 (Afekta) prioritizes wide dynamic range, resulting in higher
integrated TIC (due to enhanced detection of low-abundance background
ions).

In contrast, the Agilent 6545 (Cembio), utilizing 'Swarm Autotune'
optics, produces a 'cleaner' spectral profile with a higher BPC/TIC
ratio—not necessarily because the peaks are sharper, but because the total
integrated current (TIC) is lower due to reduced dynamic range coverage
compared to the 6546."

Summary table:
| Metric   | Pattern           | Explanation |
|----------|-------------------|---------------------------------------------|
| TIC_norm | Afekta Higher     | 6546 superior dynamic range; integrates |
|          |                   | more low-abundance signals |
| BPC_norm | Cembio Higher     | 6545 distribution more top-heavy; peaks |
|          |                   | appear taller after normalization |
| Ratio    | Cembio Higher     | 6545 misses background; lower TIC |
|          |                   | denominator inflates ratio |

### Testing the ionization preference idea

For this we will run a spearman analysis, which will rank the compound 
intensities and see how well the ranks are preserved across labs.

```{r}
## detected data for each lab:
pt_afekta <- read.csv(file.path(
  base_path,
  "object/detected_peaks_afekta_HE.csv"
))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(
  base_path,
  "object/detected_peaks_cembio_HE.csv"
))b
lab_counts <- c(
  afekta = nrow(pt_afekta),
  hmgu = nrow(pt_hmgu),
  icl = nrow(pt_icl),
  cembio = nrow(pt_cembio)
)
barplot(
  lab_counts,
  main = "Number of detected peaks per lab",
  ylab = "Number of detected peaks",
  xlab = "Labs",
  col = "steelblue"
)
## load consensus table: peaks that are present in each 4 labs.
ct_filtered <- read.csv(file.path(base_path, "object/ct_filtered4.csv"))
## i need to intyensity values for each lab
#whic can be found for each peaks here:
## but what should the table looks like exaclty.
## It should have columns for consensus_id, sample, intensity for each lab, 
## and RT for each lab.
## Create a named vector for counts.
lab_counts <- c(
  afekta = nrow(pt_afekta),
  icl = nrow(pt_icl),
  hmgu = nrow(pt_hmgu), # Replace with nrow(pt_hmgu) when available
  cembio = nrow(pt_cembio) # Replace with nrow(pt_cembio) when available
)
# Plot
barplot(
  lab_counts,
  main = "Number of detected peaks per lab",
  ylab = "Number of detected peaks",
  xlab = "Labs",
  col = "steelblue",
  ylim = c(0, max(lab_counts) * 1.2) # Add some space at the top
)

# We join the consensus table with each lab's peak table.
# We match the specific 'chrom_peak_id' from the consensus table
# to the 'chrom_peak_id' in the lab's file to get the 'intb' (intensity) and 'rt' (retention time).
intensity_table <- ct_filtered %>%
  select(
    consensus_id,
    sample,
    idx_afekta,
    chrom_peak_id_afekta,
    chrom_peak_id_icl,
    chrom_peak_id_hmgu,
    chrom_peak_id_cembio
  ) %>%

  # Join Afekta Intensity and RT
  left_join(
    pt_afekta %>% select(chrom_peak_id, intb, rt),
    by = c("chrom_peak_id_afekta" = "chrom_peak_id")
  ) %>%
  rename(intensity_afekta = intb, rt_afekta = rt) %>%

  # Join ICL Intensity and RT
  left_join(
    pt_icl %>% select(chrom_peak_id, intb, rt),
    by = c("chrom_peak_id_icl" = "chrom_peak_id")
  ) %>%
  rename(intensity_icl = intb, rt_icl = rt) %>%

  # Join HMGU Intensity and RT
  left_join(
    pt_hmgu %>% select(chrom_peak_id, intb, rt),
    by = c("chrom_peak_id_hmgu" = "chrom_peak_id")
  ) %>%
  rename(intensity_hmgu = intb, rt_hmgu = rt) %>%

  # Join CEMBIO Intensity and RT
  left_join(
    pt_cembio %>% select(chrom_peak_id, intb, rt),
    by = c("chrom_peak_id_cembio" = "chrom_peak_id")
  ) %>%
  rename(intensity_cembio = intb, rt_cembio = rt)

# Preview the table structure you asked about
print("Intensity and RT Table Preview:")
print(head(
  intensity_table %>%
    select(consensus_id, sample, starts_with("intensity"), starts_with("rt"))
))
```

```{r}
# Select only the intensity columns for correlation
intensity_matrix <- intensity_table %>%
  select(starts_with("intensity"))

# Calculate Spearman correlation
# use="pairwise.complete.obs" handles missing values (NAs) by ignoring them
cor_results <- cor(
  intensity_matrix,
  method = "spearman",
  use = "pairwise.complete.obs"
)

print("Spearman Correlation Matrix:")
print(cor_results)
```

Check this using Spearman correlation between raw data of each lab.
Approach:
- Analyze mixture by mixture
- Use common peaks across all labs to assess ionization preference
- Use Spearman to check if rank order of intensities preserved
- Answer: Which labs prioritize certain ionization pathways?

This would explain some ICC differences observed above.

Different labs 

It seems that againcembio and afekta are the most similar, with a spearman
correlation of 0.77. and hmgu has the lowest correlation with all the labs



## Retention time analysis and biological differences

How do the labs agree?

First, check if raw data is appropriate for this analysis. Compare global
vs local normalization.

Median scaling using the overall median: What fraction of total metabolic
payload elutes in Q1 vs Q2? This preserves quartile relationships and tests
if labs agree on gradient "shape." This only works if metabolite
distribution is similar across gradient—check this below.

Raw data check: Look at instrument sensitivity profile.
- Flat line: global normalization fine
- Sloped line: local normalization needed

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Define helpers for polarity iteration
run_for_polarities <- function(fn, data_level = NULL) {
  pos_result <- fn(1L, "Positive", data_level)
  neg_result <- fn(0L, "Negative", data_level)
  bind_rows(pos_result, neg_result)
}

run_for_polarities_simple <- function(fn) {
  pos_result <- fn(1L)
  neg_result <- fn(0L)
  bind_rows(pos_result, neg_result)
}

compute_consortium_sensitivity <- function(polarity) {

  # 1. Load Data (All Labs)
  load(file.path(base_path, "object/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]

  # 2. Define Breaks
  all_rts <- unlist(rtime(tic_data))
  breaks <- quantile(all_rts, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)

  # 3. Compute Raw Sums per Quartile per Lab
  raw_stats <- t(vapply(seq_along(tic_data), function(i) {
      rt <- rtime(tic_data[i])[[1]]
      int <- intensity(tic_data[i])[[1]]

      c(sum(int[rt >= breaks[1] & rt < breaks[2]], na.rm=TRUE),
        sum(int[rt >= breaks[2] & rt < breaks[3]], na.rm=TRUE),
        sum(int[rt >= breaks[3] & rt < breaks[4]], na.rm=TRUE),
        sum(int[rt >= breaks[4] & rt <= breaks[5]], na.rm=TRUE))
  }, numeric(4)))

  df <- data.frame(lab = tic_data$lab,
                   Q1 = raw_stats[,1], Q2 = raw_stats[,2],
                   Q3 = raw_stats[,3], Q4 = raw_stats[,4])

  # 4. Calculate Deviation from Consortium Median
  # First, get the "Grand Median" for each quartile across ALL labs
  grand_medians <- df %>%
    summarise(across(starts_with("Q"), median, na.rm = TRUE))

  # Then normalize each lab to this grand median
  sensitivity_profile <- df %>%
    group_by(lab) %>%
    summarise(across(starts_with("Q"), median, na.rm = TRUE)) %>%
    pivot_longer(-lab, names_to = "Quartile", values_to = "Intensity") %>%
    mutate(
      # Log2 Ratio vs The Group Average
      Consensus = case_when(
        Quartile == "Q1" ~ grand_medians$Q1,
        Quartile == "Q2" ~ grand_medians$Q2,
        Quartile == "Q3" ~ grand_medians$Q3,
        Quartile == "Q4" ~ grand_medians$Q4
      ),
      Log2_Deviation = log2(Intensity / Consensus),
      Polarity = ifelse(polarity == 1, "Positive", "Negative")
    )

  return(sensitivity_profile)
}

# Compute sensitivity profiles for both polarities
all_profiles <- run_for_polarities_simple(compute_consortium_sensitivity)
```

```{r}
# 5. Plot
ggplot(all_profiles, aes(x = Quartile, y = Log2_Deviation, group = lab, color = lab)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Polarity) +
  labs(
    title = "Justification for Local Normalization",
    subtitle = "Lab sensitivity varies by retention time (Slopes = Non-linear Bias)",
    y = "Log2 Deviation from Consortium Median",
    x = NULL
  ) +
  theme_minimal()
```

Sensitivity slopes are non-linear: ICL shows progressive loss from Q1
(+0.23) to Q4 (-0.78); Agilents show inverse trend. Cembio exhibits
distinct hyper-sensitivity to lipids in Q3 (+1.55 log deviation). These
retention-time-specific biases cannot be corrected with global
normalization. Solution: Apply Local Median Normalization, computing
scaling factors independently for each RT quartile.

**Local Normalization Analysis**

Distribution of biological signal (TIC and BPC) across RT quartiles:

```{r}
# Compute quartiles for linear model analysis (Detected Signal - Biological)
compute_rt_quartiles_biological <- function(polarity, polarity_name) {

  # 1. Load Data (Use Detected Signal as it's the harmonized standard)
  load(file.path(base_path, "object/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]

  load(file.path(base_path, "object/bpc_detect.RData"))
  bpc_data <- bpc_detect[bpc_detect$polarity == polarity]

  # 2. Compute median TIC for each lab for normalization
  tic_medians <- data.frame(
    lab = tic_data$lab,
    tic_total = vapply(intensity(tic_data), sum, numeric(1), na.rm = TRUE)
  ) %>%
    group_by(lab) %>%
    summarise(median_tic = median(tic_total, na.rm = TRUE), .groups = "drop")

  # 3. Compute median BPC for each lab for normalization
  bpc_medians <- data.frame(
    lab = bpc_data$lab,
    bpc_total = vapply(intensity(bpc_data), sum, numeric(1), na.rm = TRUE)
  ) %>%
    group_by(lab) %>%
    summarise(median_bpc = median(bpc_total, na.rm = TRUE), .groups = "drop")

  # 4. Determine GLOBAL Fixed Breaks (Average RT across all labs)
  all_rts <- unlist(rtime(tic_data))
  if(length(all_rts) == 0) return(NULL)

  # Global quartiles: 0-25%, 25-50%, etc. of the total run time
  global_breaks <- quantile(all_rts, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)

  # Helper: Compute normalized quartile sums for a metric
  compute_quartile_sums <- function(intensity_vec, rt_vec, breaks, lab_median) {
    vapply(1:4, function(q) {
      lower <- breaks[q]
      upper <- breaks[q + 1]
      sum(intensity_vec[rt_vec >= lower & rt_vec < upper], na.rm = TRUE) / lab_median
    }, numeric(1))
  }

  # Helper to compute normalized fractions in fixed windows
  compute_window_stats <- function(tic_chr, bpc_chr, breaks, lab_name,
                                   tic_medians_df, bpc_medians_df) {
    # Extract RT and intensity
    rt <- rtime(tic_chr)[[1]]
    int_tic <- intensity(tic_chr)[[1]]
    int_bpc <- intensity(bpc_chr)[[1]]

    # Get lab-specific medians
    lab_tic_median <- tic_medians_df$median_tic[tic_medians_df$lab == lab_name]
    lab_bpc_median <- bpc_medians_df$median_bpc[bpc_medians_df$lab == lab_name]

    # Compute normalized quartile sums
    tic_quartiles <- compute_quartile_sums(int_tic, rt, breaks, lab_tic_median)
    bpc_quartiles <- compute_quartile_sums(int_bpc, rt, breaks, lab_bpc_median)

    return(c(tic_quartiles, bpc_quartiles))
  }

  # Apply to all samples
  stats <- t(vapply(seq_along(tic_data), function(i) {
    compute_window_stats(tic_data[i], bpc_data[i], global_breaks,
                        tic_data$lab[i], tic_medians, bpc_medians)
  }, numeric(8)))

  colnames(stats) <- c(
    paste0("TIC_Frac_Q", 1:4),
    paste0("BPC_Frac_Q", 1:4)
  )

  result <- data.frame(
    lab = tic_data$lab,
    mixture = tic_data$mixture,
    polarity = polarity_name,
    stats
  )

  return(result)
}

# Compute biological quartile distributions for both polarities
bio_combined_quartiles <- run_for_polarities_names(compute_rt_quartiles_biological)

# Function to get Variance stats from LM
get_variance <- function(data, value_col) {
  # Log2 transform to handle intensity scale
  f <- as.formula(paste0("log2(", value_col, ") ~ lab + mixture"))
  an <- anova(lm(f, data = data))

  ss_total <- sum(an[, "Sum Sq"])
  return(data.frame(
    Mixture_Var = (an["mixture", "Sum Sq"] / ss_total) * 100,
    Lab_Var     = (an["lab", "Sum Sq"] / ss_total) * 100
  ))
}

# Run linear models on quartile data
quartile_results <- data.frame()

for (pol in c("Positive", "Negative")) {
  data_subset <- bio_combined_quartiles %>% filter(polarity == pol)

  # Analyze each TIC quartile
  for (q in 1:4) {
    col_name <- paste0("TIC_Frac_Q", q)
    var_result <- get_variance(data_subset, col_name)

    quartile_results <- rbind(quartile_results, data.frame(
      Polarity = pol,
      Metric = col_name,
      Mixture_Var = var_result$Mixture_Var,
      Lab_Var = var_result$Lab_Var
    ))
  }

  # Analyze each BPC quartile
  for (q in 1:4) {
    col_name <- paste0("BPC_Frac_Q", q)
    var_result <- get_variance(data_subset, col_name)

    quartile_results <- rbind(quartile_results, data.frame(
      Polarity = pol,
      Metric = col_name,
      Mixture_Var = var_result$Mixture_Var,
      Lab_Var = var_result$Lab_Var
    ))
  }
}

quartile_results
```

```{r}
# Helper: Create and plot quartile variance distribution
plot_quartile_variance <- function(data, metric_type = "TIC") {
  quartile_map <- c(
    "Q1" = "Q1 (Polar)",
    "Q2" = "Q2 (Semi-polar)",
    "Q3" = "Q3 (Lipids)",
    "Q4" = "Q4 (Wash)"
  )

  plot_data <- data %>%
    mutate(Residual_Var = 100 - (Mixture_Var + Lab_Var)) %>%
    filter(grepl(metric_type, Metric)) %>%
    mutate(
      Quartile = case_when(
        grepl("Q1", Metric) ~ quartile_map["Q1"],
        grepl("Q2", Metric) ~ quartile_map["Q2"],
        grepl("Q3", Metric) ~ quartile_map["Q3"],
        grepl("Q4", Metric) ~ quartile_map["Q4"]
      )
    ) %>%
    pivot_longer(cols = c(Mixture_Var, Lab_Var, Residual_Var),
                 names_to = "Source", values_to = "Variance") %>%
    mutate(Source = factor(Source,
                          levels = c("Residual_Var", "Lab_Var", "Mixture_Var"),
                          labels = c("Unexplained (Noise)", "Laboratory Bias", "Mixture (Biology)")))

  ggplot(plot_data, aes(x = Quartile, y = Variance, fill = Source)) +
    geom_col(width = 0.7, color = "white") +
    facet_wrap(~Polarity) +
    scale_fill_manual(values = c("gray85", "#E7B800", "#00AFBB")) +
    labs(
      title = paste(metric_type, "Distribution Across Retention Time Quartiles"),
      y = "Variance Contribution (%)",
      x = NULL
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "top",
      panel.grid.major.x = element_blank()
    )
}

# Plot both metrics
print(plot_quartile_variance(quartile_results, "TIC"))
print(plot_quartile_variance(quartile_results, "BPC"))
```

Local normalization by RT quartile reveals instrument-specific biases
in Q3/Q4 (30–50% residual lab variance). Hydrophobic compounds show
irreducible differences in selectivity and dynamic range that median
scaling cannot eliminate. 

### ionization preference accross rt 

Now that we showed that quartile varied highly in lab variance, we can check if
the ionization preference is also varying across rt quartile.

I want to quartile to be defined on the same axis as before (so using the total 
data run), so I will use the same function as before to compute the quartile
 breaks, and pool the peaks (in table within these.

```{r}
## create quartile boundaries using the total data rt (TICobject) 

## pull the peaks from the intensity_table into these quartiles. there is a rt 
## for each peak so you can use that.😂

# Load TIC data to get global RT breaks
load(file.path(base_path, "object/tic_full.RData"))
tic_detected <- tic_full[tic_full$polarity == 1L]  # Start with positive mode

# Get all RT values from detected signal
all_rts <- unlist(rtime(tic_detected))
global_breaks <- quantile(all_rts, probs = c(0, 0.25, 0.5, 0.75, 1), 
    na.rm = TRUE)
# Assign quartiles to peaks based on their RT
intensity_table <- intensity_table %>%
  mutate(
    # Use mean of available RTs across labs to assign quartile
    mean_rt = rowMeans(select(., starts_with("rt_")), na.rm = TRUE),
    quartile = case_when(
      mean_rt >= global_breaks[1] & mean_rt < global_breaks[2] ~ "Q1",
      mean_rt >= global_breaks[2] & mean_rt < global_breaks[3] ~ "Q2",
      mean_rt >= global_breaks[3] & mean_rt < global_breaks[4] ~ "Q3",
      mean_rt >= global_breaks[4] & mean_rt <= global_breaks[5] ~ "Q4",
      TRUE ~ NA_character_
    )
  )

# Check quartile distribution
cat("Quartile Breaks (min, Q1, Q2, Q3, max):\n")
print(round(global_breaks, 2))
cat("\nPeaks per quartile:\n")
print(table(intensity_table$quartile, useNA = "ifany"))
```


```{r}
# Perform Spearman correlation analysis of intensity values across labs within each quartile
quartile_correlations <- intensity_table %>%
  filter(!is.na(quartile)) %>%
  select(consensus_id, quartile, starts_with("intensity_")) %>%
  group_by(quartile) %>%
  summarise(
    n_peaks = n(),
    spearman_cor_afekta_icl = cor(intensity_afekta, intensity_icl, 
     method = "spearman", use = "pairwise.complete.obs"),
    spearman_cor_afekta_hmgu = cor(intensity_afekta, intensity_hmgu,
     method = "spearman", use = "pairwise.complete.obs"),
    spearman_cor_afekta_cembio = cor(intensity_afekta, intensity_cembio, 
     method = "spearman", use = "pairwise.complete.obs"),
    spearman_cor_icl_hmgu = cor(intensity_icl, intensity_hmgu, 
     method = "spearman", use = "pairwise.complete.obs"),
    spearman_cor_icl_cembio = cor(intensity_icl, intensity_cembio, 
     method = "spearman", use = "pairwise.complete.obs"),
    spearman_cor_hmgu_cembio = cor(intensity_hmgu, intensity_cembio, 
     method = "spearman", use = "pairwise.complete.obs"),
    .groups = "drop"
  )

print(quartile_correlations)
```


Notes: i need to check because i see an NA in one of the row. i think i should 
recalculate the nlab myself in the other script.
also i think this tells me that i should normalize by quartile and redo the
global analysis.


## Summary
This analysis assessed LC-MS reproducibility across 4 laboratories
analyzing 38 standard mixtures. Key findings:

1. **Peak detection improves reproducibility**: Detected signal ICC >0.5
   vs full signal ICC <0.1. Isolating compound signal from baseline noise
   is essential.

2. **Same-platform instruments agree better**: Agilent Q-TOF labs
   (Afekta, Cembio) show ICC 0.83–0.92; different platforms show lower
   agreement. Suggests instrument design dominates variability.

3. **Normalization strategy matters**: Median normalization improves ICC
   from 0.05 to 0.55; volume normalization increases differences.
   Naturally normalized metrics (ratios, percentages) are more stable
   across labs.

4. **Negative mode bias**: ~9% systematic bias persists after
   normalization, indicating fundamental ionization differences.

5. **Local normalization**: RT-quartile-specific normalization handles
   non-linear instrument sensitivity, improving biological signal
   detection in hydrophobic regions (Q3).

## Future Directions

- Incorporate MsQuality metrics (peak count, baseline intensity, rtIqrChromatograms)
- Expand to additional chemical classes
- Test in-house vs. commercial reverse-phase columns within same lab
- Validate findings on new mixture sets

## Appendix: Technical Notes

Volume normalization increased differences; platform-independent
intensity scaling is ineffective. Median scaling is preferred. Naturally
normalized metrics (sensitivity, ratios) remain stable across labs.

```{r}
library(ggplot2)
library(dplyr)
library(patchwork) # For side-by-side plots

# 1. Prepare Data
check_data <- metrics_combined %>%
  filter(data_level == "Detected Signal") %>%
  select(lab, polarity, tic, bpc, bpc_tic_ratio) %>%
  mutate(
    log_tic = log2(tic),
    log_bpc = log2(bpc)
  )

# Test A: Ratio vs TIC (should be independent)
p_tic <- ggplot(check_data, aes(x = log_tic, y = bpc_tic_ratio)) +
  geom_point(aes(color = lab), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  facet_wrap(~polarity) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Test A: TIC vs Ratio",
    subtitle = "Is efficiency affected by total load?",
    x = "Log2 Total Intensity (TIC)",
    y = "BPC/TIC Ratio"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Test B: Ratio vs BPC (expected high correlation)
p_bpc <- ggplot(check_data, aes(x = log_bpc, y = bpc_tic_ratio)) +
  geom_point(aes(color = lab), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  facet_wrap(~polarity) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Test B: BPC vs Ratio",
    subtitle = "Is this just autocorrelation?",
    x = "Log2 Max Intensity (BPC)",
    y = "BPC/TIC Ratio"
  ) +
  theme_minimal()

# Display plots side-by-side
p_combined <- p_tic + p_bpc
print(p_combined)

# Compute correlations
cor_stats <- check_data %>%
  group_by(polarity, lab) %>%
  summarise(
    cor_with_TIC = cor(log_tic, bpc_tic_ratio, method = "spearman"),
    cor_with_BPC = cor(log_bpc, bpc_tic_ratio, method = "spearman"),
    .groups = "drop"
  )

print(cor_stats)
```

Lab-dependent correlations indicate the ratio cannot serve as a universal
metric for cross-lab comparisons.


