---
title: "NAPS chromExtract — re-extraction producer for 4_rt_alignment"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
---

# Goal

Re-extract NAPS anchor apex RTs by `chromExtract` at the **theoretical
m/z** (from compound formula) plus an RT-seeded window, with cross-lab
**propagation** for gaps. This replaces the xcms-based per-lab
`naps.csv` as the alignment input.

Why: the xcms-derived NAPS RTs in `1_preprocessing/{lab}/HE/naps.csv`
have systematic problems — `icl` is missing `C1`–`C3` entirely (xcms
didn't pick them) and `cembio-pos` has a few replicates with
catastrophic outliers (wrong-peak picks). chromExtract at theoretical
m/z lets us recover those because we're not relying on xcms to find a
feature first — we ask: "is there a peak at this m/z near this RT?".

Output: `5_downstream_analysis/object/naps_chrom_metrics.csv` keyed on
`(lab, polarity, anchor_n, replicate, source)` with columns
`apex_rt, xicFwhm, maxIntensity, areaUnderTic`.

Source semantics:

- `source = "own"`   — `rt_center` came from the lab's own `naps.csv`
  for that (anchor, replicate).
- `source = "consensus"` — `rt_center` came from cross-lab median
  (this replicate's slot was NA in the lab's CSV; propagated).

## Setup

```{r setup}
#| message: false
library(Spectra)
library(MsExperiment)
library(Chromatograms)
library(MsQuality)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(readxl)

labs           <- c("afekta", "cembio", "hmgu", "icl")
out_dir        <- here("5_downstream_analysis", "object")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Extraction windows
ppm_tol      <- 20          # m/z tolerance
pad_wide     <- 60          # symmetric ±60 s — gives real tail context for
                            # noise estimation AND tolerates cross-lab
                            # chromatographic drift (e.g. icl C10 pos is
                            # 11 s before consensus; the legacy ±10 s
                            # window missed it and the legacy two-pass
                            # retry skipped because peakBoundary returned
                            # a noise spike inside the tight window).
apex_search_pad <- 20       # how far from rt_center the picker looks
                            # for the apex. NAPS RTs can drift up to
                            # ~15-20 s lab-to-lab; ±20 s catches that
                            # safely without crossing into neighbouring
                            # anchor regions.
```

## 1. Load anchor metadata (theoretical m/z, RTI ladder)

Any lab's `naps.csv` works — `Name`, `mz_pos`, `mz_neg`, `RTI` are
shared (only the recorded per-replicate RT columns differ).

```{r anchor-metadata}
# Helper: drop CSV row-name column (first column has an empty header "" in
# every lab's naps.csv) — otherwise dplyr's mutate refuses the data frame.
.drop_empty_cols <- function(df) df[, nzchar(names(df)), drop = FALSE]

anchor_meta <- read.csv(here("1_preprocessing", "afekta", "HE", "naps.csv"),
                        stringsAsFactors = FALSE,
                        check.names = FALSE) |>
  .drop_empty_cols() |>
  mutate(anchor_n = suppressWarnings(as.integer(
           gsub("^C(\\d+).*", "\\1", Name)))) |>
  filter(!is.na(anchor_n)) |>
  select(anchor_n, RTI, mz_pos, mz_neg) |>
  arrange(anchor_n)

cat("anchor_meta rows:", nrow(anchor_meta), " (expected 20)\n")
head(anchor_meta)
```

## 2. Identify NAPS mzml files per `(lab, polarity, replicate)`

`seq_*.xlsx` is the canonical injection order. Filter to filenames
containing `NAPS`, sort by row order, assign `replicate = 1..N`.

```{r naps-files}
naps_files <- bind_rows(lapply(labs, function(lab) {
  study_path <- here("1_preprocessing", lab, "HE")
  mzml_dir   <- file.path(study_path, "mzml")
  bind_rows(lapply(c("pos", "neg"), function(pol) {
    seq <- read_xlsx(file.path(study_path, sprintf("seq_%s.xlsx", pol))) |>
      as.data.frame()
    seq$filename <- paste0(seq$`Data File`, ".mzML")
    seq <- seq[grepl("NAPS", seq$filename, fixed = TRUE) &
               seq$filename %in% list.files(mzml_dir), ]
    if (nrow(seq) == 0) return(NULL)
    data.frame(lab       = lab,
               polarity  = pol,
               replicate = seq_len(nrow(seq)),
               filename  = seq$filename,
               path      = file.path(mzml_dir, seq$filename),
               stringsAsFactors = FALSE)
  }))
}))

cat("NAPS files per (lab, polarity):\n")
naps_files |> count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable()
```

## 3. Build seed `rt_center` per `(lab, polarity, anchor, replicate)`

Read each lab's existing `naps.csv` to pull the per-replicate RT
column. Then compute cross-lab consensus per `(polarity, anchor)`
(median across labs of per-lab median RT) for the propagation
fallback.

```{r seed-rt}
read_naps_long <- function(lab) {
  df <- read.csv(here("1_preprocessing", lab, "HE", "naps.csv"),
                 stringsAsFactors = FALSE, check.names = FALSE) |>
    .drop_empty_cols()
  df$anchor_n <- suppressWarnings(as.integer(
    gsub("^C(\\d+).*", "\\1", df$Name)))
  pos_cols <- grep("^NAPS_POS_", names(df), value = TRUE)
  neg_cols <- grep("^NAPS_NEG_", names(df), value = TRUE)
  bind_rows(
    df |> select(anchor_n, all_of(pos_cols)) |>
      pivot_longer(all_of(pos_cols), names_to = "rep_col",
                   values_to = "rt_csv") |>
      mutate(polarity = "pos",
             replicate = as.integer(gsub("^NAPS_POS_", "", rep_col))),
    df |> select(anchor_n, all_of(neg_cols)) |>
      pivot_longer(all_of(neg_cols), names_to = "rep_col",
                   values_to = "rt_csv") |>
      mutate(polarity = "neg",
             replicate = as.integer(gsub("^NAPS_NEG_", "", rep_col)))
  ) |>
    transmute(lab = lab, polarity, anchor_n, replicate, rt_csv)
}

naps_csv_long <- bind_rows(lapply(labs, read_naps_long)) |>
  filter(!is.na(anchor_n))

# Cross-lab consensus per (polarity, anchor) = median of per-lab medians
consensus_rt <- naps_csv_long |>
  filter(!is.na(rt_csv)) |>
  group_by(lab, polarity, anchor_n) |>
  summarise(rt_lab_med = median(rt_csv), .groups = "drop") |>
  group_by(polarity, anchor_n) |>
  summarise(consensus_rt = median(rt_lab_med),
            n_labs       = n(),
            .groups = "drop")

# Seed table: own rt_csv if present, else consensus_rt
seed_table <- naps_csv_long |>
  left_join(consensus_rt, by = c("polarity", "anchor_n")) |>
  mutate(rt_center = coalesce(rt_csv, consensus_rt),
         source    = ifelse(is.na(rt_csv), "consensus", "own")) |>
  filter(!is.na(rt_center))  # drop rows with no consensus either

cat("Seed table: own =", sum(seed_table$source == "own"),
    " | consensus (propagated) =", sum(seed_table$source == "consensus"), "\n")
```

## 4. Build flat peak table

One row per `(lab, polarity, anchor_n, replicate)`. Join with
`naps_files` to get `dataOrigin` and with `anchor_meta` to get
theoretical m/z.

```{r peak-table}
sps_all <- Spectra(naps_files$path, backend = MsBackendMzR())
sps_all <- sps_all[msLevel(sps_all) == 1L]
canon   <- unique(dataOrigin(sps_all))
naps_files$dataOrigin <- canon[match(basename(naps_files$path),
                                     basename(canon))]
stopifnot(all(!is.na(naps_files$dataOrigin)))

peak_table <- seed_table |>
  inner_join(naps_files |> select(lab, polarity, replicate, dataOrigin),
             by = c("lab", "polarity", "replicate")) |>
  inner_join(anchor_meta |> select(anchor_n, mz_pos, mz_neg),
             by = "anchor_n") |>
  mutate(mz_center = ifelse(polarity == "pos", mz_pos, mz_neg)) |>
  filter(!is.na(mz_center)) |>
  transmute(
    mzMin     = mz_center * (1 - ppm_tol / 1e6),
    mzMax     = mz_center * (1 + ppm_tol / 1e6),
    rtMin     = pmax(0, rt_center - pad_wide),
    rtMax     = rt_center + pad_wide,
    msLevel   = 1L,
    dataOrigin, lab, polarity, anchor_n, replicate, source,
    mz_center, rt_center
  )

cat("peak_table rows:", nrow(peak_table),
    " (max possible 4 × 2 × 20 × 6 = 960)\n")
```

## 5. chromExtract — single wide pass (±60 s)

The legacy two-pass design (tight ±10 s pass 1, fallback ±60/+20 s
pass 2) was replaced after audit showed it systematically missed real
peaks for icl pos NAPS C10: a 5 × 10⁵-count peak sits 11 s before
the consensus RT; the pass-1 ±10 s window excluded it and the pass-2
retry never fired because `peakBoundary()` returned a 386-count noise
spike inside the tight window. A single wide extraction + SNR-aware
picker (§6) avoids that failure mode.

```{r chromextract}
#| message: false
chr_base <- Chromatograms(sps_all)
chr_all  <- chromExtract(chr_base, peak.table = peak_table,
                         by = c("msLevel", "dataOrigin"))
cat("chromatograms:", length(chr_all),
    " | non-empty:", length(filterEmptyChromatograms(chr_all)), "\n")
```

## 6. SNR-aware peak picking

We use the same picker as `sirius_annotation_all_ms2.qmd` §7.6 (with
the bottom-half-tail noise estimate). For NAPS the SNR/shape/absolute-
floor guards almost never trigger — NAPS standards are concentrated
and easily above detection — but having them present means the few
genuine NAPS extraction failures (e.g. an injection where the column
didn't elute the standard at all) get cleanly flagged rather than
fabricated.

```{r snr-picker}
extract_one <- function(rt, intensity, rt_prior,
                        apex_search_pad = 20,
                        tail_pad        = 25,
                        snr_threshold   = 5,
                        abs_min_counts  = 30) {
  ok <- !is.na(intensity); rt <- rt[ok]; intensity <- intensity[ok]
  if (length(rt) < 5L)
    return(list(detection = "no_data",
                apex_rt = NA_real_, maxIntensity = NA_real_,
                areaUnderTic = NA_real_, xicFwhm = NA_real_,
                snr = NA_real_, baseline = NA_real_, noise = NA_real_,
                n_scans_in_shape = 0L))
  sm <- as.numeric(stats::filter(intensity, rep(1, 3) / 3, sides = 2))
  sm[is.na(sm)] <- intensity[is.na(sm)]
  search_idx <- which(rt >= rt_prior - apex_search_pad &
                      rt <= rt_prior + apex_search_pad)
  if (!length(search_idx))
    return(list(detection = "no_data",
                apex_rt = NA_real_, maxIntensity = NA_real_,
                areaUnderTic = NA_real_, xicFwhm = NA_real_,
                snr = NA_real_, baseline = NA_real_, noise = NA_real_,
                n_scans_in_shape = 0L))
  apex_local <- search_idx[which.max(intensity[search_idx])]
  apex_rt    <- rt[apex_local]; apex_int <- intensity[apex_local]
  tail_idx <- which(abs(rt - apex_rt) > tail_pad)
  if (length(tail_idx) < 10L) {
    n <- length(rt); tail_idx <- c(seq_len(round(0.25 * n)),
                                    (n - round(0.25 * n) + 1L):n)
  }
  tail_int <- intensity[tail_idx]
  tail_low <- tail_int[tail_int <= quantile(tail_int, 0.5, na.rm = TRUE)]
  baseline <- median(tail_low, na.rm = TRUE)
  noise_mad <- mad(tail_low, na.rm = TRUE)
  q15 <- as.numeric(quantile(tail_low, 0.15, na.rm = TRUE))
  q85 <- as.numeric(quantile(tail_low, 0.85, na.rm = TRUE))
  raw_noise <- max(noise_mad, (q85 - q15) / 2, na.rm = TRUE)
  noise_at_floor <- !is.finite(raw_noise) || raw_noise <= 0
  noise <- max(raw_noise, 1, na.rm = TRUE)
  snr   <- (apex_int - baseline) / noise
  v_idx <- tryCatch(MsCoreUtils::valleys(sm), error = function(e) integer(0))
  if (length(v_idx) >= 2L) {
    left  <- suppressWarnings(max(v_idx[v_idx <= apex_local], na.rm = TRUE))
    right <- suppressWarnings(min(v_idx[v_idx >= apex_local], na.rm = TRUE))
    if (!is.finite(left))  left  <- 1L
    if (!is.finite(right)) right <- length(rt)
  } else {
    left <- apex_local; right <- apex_local
    while (left  > 1L         && sm[left  - 1L] > baseline + noise) left  <- left  - 1L
    while (right < length(rt) && sm[right + 1L] > baseline + noise) right <- right + 1L
  }
  pk_rt <- rt[left:right]; pk_int <- intensity[left:right]
  area  <- if (length(pk_rt) >= 2L)
             sum(diff(pk_rt) * (pk_int[-1] + pk_int[-length(pk_int)]) / 2)
           else NA_real_
  hm    <- (apex_int - baseline) / 2 + baseline
  above <- which(pk_int >= hm)
  fwhm  <- if (length(above) >= 2L) diff(range(pk_rt[above])) else NA_real_
  nb_lo <- max(1L, apex_local - 2L); nb_hi <- min(length(rt), apex_local + 2L)
  n_shape <- sum(intensity[nb_lo:nb_hi] >= 0.3 * apex_int, na.rm = TRUE)
  detection <- if (apex_int <= baseline) "below_detection"
               else if (snr < snr_threshold) "below_detection"
               else if (n_shape < 2L) "below_detection"
               else if (noise_at_floor && apex_int < abs_min_counts) "below_detection"
               else "detected"
  list(detection = detection, apex_rt = apex_rt,
       maxIntensity = if (detection == "detected") apex_int else NA_real_,
       areaUnderTic = if (detection == "detected") area     else NA_real_,
       xicFwhm     = if (detection == "detected") fwhm     else NA_real_,
       snr = snr, baseline = baseline, noise = noise,
       n_scans_in_shape = n_shape)
}
```

## 7. Apply the picker

```{r apply-picker}
#| message: false
cd_all <- as.data.frame(chromData(chr_all))
pkData <- peaksData(chr_all)

metrics_rows <- vector("list", nrow(cd_all))
for (i in seq_len(nrow(cd_all))) {
  pd <- pkData[[i]]
  if (is.null(pd) || !nrow(pd)) {
    metrics_rows[[i]] <- list(detection = "no_data",
                              apex_rt = NA_real_, maxIntensity = NA_real_,
                              areaUnderTic = NA_real_, xicFwhm = NA_real_,
                              snr = NA_real_, baseline = NA_real_,
                              noise = NA_real_, n_scans_in_shape = 0L)
    next
  }
  metrics_rows[[i]] <- extract_one(pd[, "rtime"], pd[, "intensity"],
                                   rt_prior        = cd_all$rt_center[i],
                                   apex_search_pad = apex_search_pad)
}

metrics_df <- bind_cols(
  cd_all |> select(lab, polarity, anchor_n, replicate, source),
  bind_rows(lapply(metrics_rows, as.data.frame))
)

cat("Picker done — ", nrow(metrics_df), " rows\n",
    "  detected:        ", sum(metrics_df$detection == "detected"), "\n",
    "  below_detection: ", sum(metrics_df$detection == "below_detection"), "\n",
    "  no_data:         ", sum(metrics_df$detection == "no_data"), "\n",
    sep = "")
```

## 8. Detection summary

```{r detection-summary}
metrics_df |>
  count(lab, polarity, detection) |>
  tidyr::pivot_wider(names_from = detection, values_from = n,
                     values_fill = 0L) |>
  knitr::kable(caption = "NAPS detection counts per (lab, polarity).")
```

## 9. Assemble final NAPS chrom_metrics

```{r assemble-output}
naps_chrom_metrics <- metrics_df |>
  filter(detection == "detected") |>
  select(lab, polarity, anchor_n, replicate, source,
         apex_rt, areaUnderTic, xicFwhm, maxIntensity,
         signalToNoiseRatio = snr)

write.csv(naps_chrom_metrics,
          file.path(out_dir, "naps_chrom_metrics.csv"),
          row.names = FALSE)

cat("naps_chrom_metrics.csv  --  ", nrow(naps_chrom_metrics), " rows kept",
    "  |  excluded (below_detection / no_data): ",
    nrow(metrics_df) - nrow(naps_chrom_metrics), "\n",
    "  path: ", file.path(out_dir, "naps_chrom_metrics.csv"), "\n", sep = "")
```

## 10. Diagnostic — sample below_detection EICs

Rows the picker classified `below_detection` (failed SNR, shape, or
absolute-floor guards from §6). NAPS standards are concentrated so
this set should be small; if it isn't, the picker thresholds need
revisiting. Each panel = the wide (±60 s) EIC for one excluded row.
Dotted line = the seed `rt_center`.

```{r diag-below-detection}
#| fig-height: 8
#| fig-cap: "Wide EICs (±60 s) for rows the picker flagged below_detection. Expected: flat or single-spike traces (rightly excluded). A visible peak here = guard threshold too strict."
bd_idx <- which(metrics_df$detection != "detected")
cat("Rows flagged below_detection or no_data:", length(bd_idx), "\n")

if (length(bd_idx) > 0L) {
  n_show <- min(12L, length(bd_idx))
  show_idx <- if (length(bd_idx) <= 12L) bd_idx else sample(bd_idx, n_show)
  bd_eic <- bind_rows(lapply(show_idx, function(j) {
    pd <- pkData[[j]]; if (is.null(pd) || !nrow(pd)) return(NULL)
    md <- metrics_df[j, ]
    tibble(rt        = pd[, "rtime"],
           intensity = pd[, "intensity"],
           rt_center = cd_all$rt_center[j],
           panel = sprintf("%s | %s | C%d r%d (%s)\ndetection=%s | snr=%.2f",
                           md$lab, md$polarity, md$anchor_n, md$replicate,
                           md$source, md$detection,
                           ifelse(is.finite(md$snr), md$snr, NA_real_)))
  }))
  ggplot(bd_eic, aes(x = rt, y = intensity)) +
    geom_line(colour = "#ff7f0e", linewidth = 0.5) +
    geom_point(size = 0.8, alpha = 0.7) +
    geom_vline(aes(xintercept = rt_center),
               linetype = "dotted", colour = "grey50") +
    facet_wrap(~ panel, scales = "free", ncol = 3) +
    labs(x = "RT (s)", y = "Intensity",
         title = "Below-detection EICs (sample)") +
    theme_bw() +
    theme(strip.text = element_text(size = 6.5))
} else {
  message("No below_detection / no_data rows.")
}
```

## 11. QC — chromExtract vs original `naps.csv`

Side-by-side: per `(lab, polarity, anchor_n, replicate)`, compare the
re-extracted `apex_rt` to the original CSV value. Should agree to
within ~1 s where xcms picked the right peak, recover obvious NAs,
and (we hope) replace the high-SD outlier replicates in cembio-pos /
icl.

```{r qc-old-vs-new}
qc <- naps_chrom_metrics |>
  left_join(naps_csv_long, by = c("lab", "polarity",
                                  "anchor_n", "replicate")) |>
  mutate(delta = apex_rt - rt_csv)

qc |> filter(!is.na(rt_csv)) |>
  group_by(lab, polarity) |>
  summarise(n               = n(),
            median_abs_delta = round(median(abs(delta), na.rm = TRUE), 3),
            p95_abs_delta   = round(quantile(abs(delta), 0.95, na.rm = TRUE), 3),
            n_disagree_2s   = sum(abs(delta) > 2, na.rm = TRUE),
            .groups = "drop") |>
  knitr::kable(caption = "chromExtract vs original CSV — agreement per (lab, polarity).")

cat("rows recovered (CSV NA → chromExtract found peak):",
    sum(is.na(qc$rt_csv) & !is.na(qc$apex_rt)), "\n")
cat("rows lost (CSV had value → chromExtract NA):",
    sum(!is.na(qc$rt_csv) & is.na(qc$apex_rt)), "\n")
```

```{r qc-within-lab-sd}
#| fig-height: 4
#| fig-cap: "Per-anchor within-lab replicate SD: original CSV (xcms) vs chromExtract. Lower = tighter."
sd_old <- naps_csv_long |>
  filter(!is.na(rt_csv)) |>
  group_by(lab, polarity, anchor_n) |>
  summarise(rt_sd = sd(rt_csv), n_rep = n(), .groups = "drop") |>
  filter(n_rep >= 2L, !is.na(rt_sd)) |>
  mutate(source_tag = "xcms (original CSV)")

sd_new <- naps_chrom_metrics |>
  filter(!is.na(apex_rt)) |>
  group_by(lab, polarity, anchor_n) |>
  summarise(rt_sd = sd(apex_rt), n_rep = n(), .groups = "drop") |>
  filter(n_rep >= 2L, !is.na(rt_sd)) |>
  mutate(source_tag = "chromExtract (new)")

bind_rows(sd_old, sd_new) |>
  ggplot(aes(x = lab, y = rt_sd, fill = source_tag)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7,
               position = position_dodge(0.8)) +
  geom_jitter(aes(group = source_tag),
              position = position_jitterdodge(jitter.width = 0.1,
                                              dodge.width = 0.8),
              alpha = 0.4, size = 0.7) +
  scale_y_log10() +
  facet_wrap(~ polarity) +
  scale_fill_manual(values = c(`xcms (original CSV)` = "grey60",
                               `chromExtract (new)`   = "#4DBBD5")) +
  labs(x = NULL, y = "Within-lab replicate SD (s, log)",
       fill = NULL,
       title = "NAPS within-lab precision: xcms vs chromExtract") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

# Notes

- `4_rt_alignment.qmd` should now read `naps_chrom_metrics.csv` in
  preference to the per-lab `naps.csv` files. The structure is
  long-form already (`(lab, polarity, anchor_n, replicate, source,
  apex_rt)`), so the existing `naps_long` build in §P2.1 collapses to
  a single read.
- `source` distinguishes own-seeded vs consensus-propagated rows.
  Downstream filtering on `source == "own"` reproduces the
  xcms-replacement view; including `consensus` rows is the
  propagation-augmented view.
- The QC scatter / SD plots above are the gate: if the new chromExtract
  SDs are not noticeably tighter than the original xcms ones for the
  problematic labs (cembio-pos, icl), revisit the seed propagation or
  the m/z tolerance before trusting the alignment downstream.
