---
title: "Peak Feature Calculation — Detected Peaks"
format:
  html:
    toc: true
execute:
  dir: project
---

```{r}
library(BiocParallel)
bp <- SnowParam(workers = 4)
register(bp)
```

# Introduction

This document performs peak feature calculation and analysis for the detected
detected ion chromatograms (EICs). It reuses the processed detected EICs, computes peak
metrics, and runs downstream comparisons and stability analyses.

- Use all detected peaks from the beginning, in the chromData have info
  whether they are annotated
- Only keep peaks overlapping the 4 labs to create a common feature set

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)
library(UpSetR)
library(MsQuality)

base_path <- here::here("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
source(here::here("1_preprocessing/setup.R"))
```


# Create Consensus Peak Table Across Labs

```{r}
pt_afekta <- read.csv(file.path(
  base_path,
  "object/detected_peaks_afekta_HE.csv"
))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(
  base_path,
  "object/detected_peaks_cembio_HE.csv"
))
lab_counts <- c(
  afekta = nrow(pt_afekta),
  hmgu = nrow(pt_hmgu),
  icl = nrow(pt_icl),
  cembio = nrow(pt_cembio)
)
barplot(
  lab_counts,
  main = "Number of Detected Peaks per Lab",
  ylab = "Number of detected peaks",
  xlab = "Labs",
  col = "steelblue"
)
```

```{r build-consensus-peaks}
# the table will be generated on the cluster. and loaded below
# the parameter used for mclosest are: ppm tol = 10 and rt tol = 10
# Load all consensus_all_sample*.csv files (excluding _edge.csv files)
consensus_dir <- file.path(base_path, "object/consensus_tables")
consensus_files <- list.files(consensus_dir,
 pattern = "^consensus_all_sample[0-9]+\\.csv$", full.names = TRUE)

ct_all_samples <- do.call(rbind, lapply(consensus_files, read.csv))
rownames(ct_all_samples) <- NULL

nrow(ct_all_samples)
```


```{r}
# Add detection flags to consensus_matches
ct_all_samples <- ct_all_samples |>
  mutate(
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  )

cat("Peak counts by number of labs detected:\n")
print(table(ct_all_samples$n_labs_matched))
```

```{r}
# Replace mixture column from 1.1 to 1_1
meta$mixture <- gsub("\\.", "_", meta$Mixture)
nc_mix <- table(meta$mixture)
# This shows the number of compounds per mixture.

# Create mapping of sample to mixture name using pt_* data
sample_to_mixture <- pt_afekta |>
  select(sample, mixture) |>
  distinct()

# Below we can summarize the number of peaks detected per sample with compound count
mixture_summary <- lapply(split(ct_all_samples, ct_all_samples$sample), function(df) {
  sample_id <- unique(df$sample)
  mixture_name <- sample_to_mixture$mixture[sample_to_mixture$sample == sample_id]

  res <- data.frame(
    sample = sample_id,
    mixture = mixture_name,
    detected_4labs = sum(df$n_labs_matched == 4),
    detected_3labs = sum(df$n_labs_matched == 3),
    detected_2labs = sum(df$n_labs_matched == 2),
    detected_1lab = sum(df$n_labs_matched == 1),
    total_peaks = nrow(df),
    n_compounds = as.integer(nc_mix[mixture_name])
  )
  res
})

mixture_summary_df <- do.call(rbind, mixture_summary)
rownames(mixture_summary_df) <- NULL

knitr::kable(mixture_summary_df, caption = "Peak Detection Summary by Mixture")
```


## Upset Plot - Lab Overlap

```{r upset-plot, message = FALSE, warning = FALSE}
# Prepare binary matrix for upset plot
upset_data <- ct_all_samples |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(
    afekta = as.integer(afekta_detected),
    hmgu = as.integer(hmgu_detected),
    icl = as.integer(icl_detected),
    cembio = as.integer(cembio_detected)
  ) |>
  select(afekta, hmgu, icl, cembio)

# Convert to data frame with rownames for upset function
upset_data_df <- as.data.frame(upset_data)
rownames(upset_data_df) <- paste0("peak_", seq_len(nrow(upset_data_df)))

# Only plot intersections with at least 1 peak (exclude empty combinations)
upset(upset_data_df,
      sets = c("afekta", "hmgu", "icl", "cembio"),
      main.bar.color = "steelblue",
      sets.bar.color = "darkred",
      text.scale = c(2, 2, 2, 1.5, 2, 2),
      point.size = 3,
      line.size = 1.2)
```

## Pairwise Lab Comparisons

```{r pairwise-overlap}
# Create pairwise overlap summary
labs <- c("afekta", "hmgu", "icl", "cembio")
pairwise_summary <- data.frame()

for (i in seq_along(labs)) {
  for (j in seq_along(labs)) {
    if (i < j) {
      lab1 <- labs[i]
      lab2 <- labs[j]

      col1 <- paste0(lab1, "_detected")
      col2 <- paste0(lab2, "_detected")

      both <- sum(ct_all_samples[[col1]] & ct_all_samples[[col2]])
      only_lab1 <- sum(ct_all_samples[[col1]] & !ct_all_samples[[col2]])
      only_lab2 <- sum(!ct_all_samples[[col1]] & ct_all_samples[[col2]])

      total_lab1 <- sum(ct_all_samples[[col1]])
      total_lab2 <- sum(ct_all_samples[[col2]])

      pairwise_summary <- rbind(pairwise_summary, data.frame(
        lab1 = lab1,
        lab2 = lab2,
        both_detected = both,
        only_lab1 = only_lab1,
        only_lab2 = only_lab2,
        total_lab1 = total_lab1,
        total_lab2 = total_lab2,
        jaccard_index = round(100 * both / (both + only_lab1 + only_lab2), 1),
        overlap_percent_lab1 = round(100 * both / total_lab1, 1),
        overlap_percent_lab2 = round(100 * both / total_lab2, 1)
      ))
    }
  }
}

knitr::kable(pairwise_summary, caption = "Pairwise Lab Overlap - Detailed Metrics")
```

As expected, afekta and cembio have the highest overlap.

```{r}
ct_filtered <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Consensus peaks retained:", nrow(ct_filtered), "\n")
cat(
  "Percentage of detected peaks:",
  round(100 * nrow(ct_filtered) / nrow(ct_all_samples), 1),
  "%\n"
)
```

Build the final consensus peak table with chrom_peak_id mapping:
Save the consensus table for downstream analysis:

```{r save-consensus}
write.csv(
  ct_all_samples,
  file = file.path(base_path, "object/ct_all_samples.csv"),
  row.names = FALSE
)
write.csv(
  ct_filtered,
  file = file.path(base_path, "object/ct_filtered4.csv"),
  row.names = FALSE
)
```

## Summary Tables

### Overlap: Cembio vs Afekta

```{r cembio-afekta-overlap}
cembio_afekta_summary <- data.frame(
  category = c("Both detected", "Only afekta", "Only cembio", "Neither"),
  count = c(
    sum(ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected)
  )
)

cembio_afekta_summary <- cembio_afekta_summary |>
  mutate(percent = round(100 * count / sum(count), 1))

knitr::kable(cembio_afekta_summary,
             caption = "Peak Overlap: Cembio vs Afekta")
```

so we could look at 2526 detected peaks ( for 2 mixture), that's really nice.

### Overlap: All 4 Labs

```{r all-labs-overlap}
all_labs_summary <- data.frame(
  detection_pattern = character(),
  n_peaks = integer(),
  percent = numeric()
)

for (combo in 0:15) {
  afekta <- as.logical(bitwAnd(combo, 1))
  hmgu <- as.logical(bitwAnd(combo, 2))
  icl <- as.logical(bitwAnd(combo, 4))
  cembio <- as.logical(bitwAnd(combo, 8))

  n <- sum(
    ct_all_samples$afekta_detected == afekta &
    ct_all_samples$hmgu_detected == hmgu &
    ct_all_samples$icl_detected == icl &
    ct_all_samples$cembio_detected == cembio
  )

  if (n > 0) {
    pattern <- paste0(
      if_else(afekta, "A", "-"),
      if_else(hmgu, "H", "-"),
      if_else(icl, "I", "-"),
      if_else(cembio, "C", "-")
    )

    all_labs_summary <- rbind(all_labs_summary, data.frame(
      detection_pattern = pattern,
      n_peaks = n,
      percent = round(100 * n / nrow(ct_all_samples), 1)
    ))
  }
}

all_labs_summary <- all_labs_summary |>
  arrange(desc(n_peaks))

knitr::kable(all_labs_summary,
             caption = "Peak Detection Patterns: All 4 Labs (A=afekta, H=hmgu, I=icl, C=cembio)")
```



# Peak Feature Summary by Lab (Detected)

## Prepare subset for downstream analysis

```{r prepare-subsets}
# Peaks present in ALL 4 labs (main analysis subset)
all_four_labs <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Peaks in all 4 labs:", nrow(all_four_labs), "\n")
```

# Creating eics
## 4 labs

```{r}
pt <- all_four_labs |>
  mutate(consensus_id = row_number()) |>
  select(consensus_id, sample, chrom_peak_id_afekta, chrom_peak_id_hmgu,
  chrom_peak_id_icl, chrom_peak_id_cembio) |>
  pivot_longer(
    cols = c(chrom_peak_id_afekta, chrom_peak_id_hmgu, chrom_peak_id_icl,
    chrom_peak_id_cembio),
    names_to = "lab",
    names_prefix = "chrom_peak_id_",
    values_to = "chrom_peak_id"
  ) |>
  filter(!is.na(chrom_peak_id))
```

```{r}
# Combine all lab peak tables with lab identifier for simple merge
pt_all <- rbind(
  transform(pt_afekta, lab = "afekta"),
  transform(pt_hmgu, lab = "hmgu"),
  transform(pt_icl, lab = "icl"),
  transform(pt_cembio, lab = "cembio")
)

# Join peak info using merge instead of case_when
join_cols <- c("chrom_peak_id", "lab", "mzmin", "mzmax", "rtmin", "rtmax",
               "mz", "polarity", "mixture", "into", "intb", "maxo")
pt <- merge(pt, pt_all[, join_cols], by = c("chrom_peak_id", "lab"), all.x = TRUE)

# Rename and adjust boundaries
names(pt)[names(pt) == "mzmin"] <- "mzMin"
names(pt)[names(pt) == "mzmax"] <- "mzMax"
names(pt)[names(pt) == "rtmin"] <- "rtMin"
names(pt)[names(pt) == "rtmax"] <- "rtMax"

pt$rtMin <- pt$rtMin - 2
pt$rtMax <- pt$rtMax + 2
pt$mzMin <- pt$mzMin - 0.02
pt$mzMax <- pt$mzMax + 0.02
pt$polarity <- ifelse(pt$sample %in% 1:19, 1L, 0L)

write.csv(
  pt,
  file = file.path(base_path, "object/peak_table_detected_all4labs.csv"),
  row.names = FALSE
)
```

```{r}
library(Chromatograms)

if (!file.exists(file.path(base_path, "object/eics_detected_all4labs.RData"))) {
  load(file.path(base_path, "object/sp_full.RData"))
  tic_full <- Chromatograms(
    sp_full,
    summarize.method = "sum",
    factorize.by = "dataOrigin",
    spectraVariables = c("mixture", "lab")
  )
  eics <- chromExtract(
    tic_full,
    peak.table = pt,
    by = c("polarity", "mixture", "lab")
  )
  nrow(pt) == length(eics)
  eics <- setBackend(eics, ChromBackendMemory())
  save(eics, file = file.path(base_path, "object/eics_detected_all4labs.RData"))
} else {
  load(file.path(base_path, "object/eics_detected_all4labs.RData"))
}

eics
```

## Compute metrics


```{r compute-metrics}
# Step 1: peak boundaries — now in Chromatograms, no longer a calculateMetrics metric
cat("Computing peak boundaries...\n")
pb <- Chromatograms::peakBoundary(eics)
cat("peakBoundary columns:", paste(names(pb), collapse = ", "), "\n")
chromData(eics) <- cbind(chromData(eics), pb)

# Step 2: quality metrics used for filtering and downstream doc-3 analysis
# Removed: peakBoundary (above), gaussianSimilarity (computed on narrow EICs below)
# Removed: f = seq_len() — not a valid parameter in current MsQuality
pk <- calculateMetrics(
  eics,
  metrics = c(
    "peakCount", "xicFwhm", "maxIntensity",
    "baselineIntensity", "signalToNoiseRatio",
    "peakWidth", "peakProminence"
  ),
  na.rm = TRUE,
  filterEmptyObject = FALSE,
  probs = 0.05
)

chromData(eics) <- cbind(chromData(eics), pk)
cat("Metrics computed for", length(eics), "EICs\n")
```

## Filter bad peaks

Remove consensus peaks with poor quality metrics in a single step:

```{r filter-bad-peaks}
# Consolidate all filtering criteria
bad_ids <- unique(c(
  eics$consensus_id[is.na(eics$xicFwhm)],           # NA FWHM
  eics$consensus_id[eics$peakProminence < 1.5],     # Low prominence
  eics$consensus_id[eics$peakWidth < 1],            # Too narrow
  eics$consensus_id[eics$peakWidth > 30]            # Too wide
))

cat("Removing", length(bad_ids), "consensus peaks with poor metrics\n")
eics <- eics[!eics$consensus_id %in% bad_ids]
cat("Remaining EICs:", length(eics), "\n")
cat("Unique consensus IDs:", length(unique(eics$consensus_id)), "\n")
```

## Re-extract EICs using peak boundaries

```{r re-extract}
# NOTE: delete the cached RData below if you want to force re-extraction
# (necessary after changes to boundary computation or imputation).
rdata_path <- file.path(base_path, "object/eics_filtered_detected_all4labs.RData")

if (!file.exists(rdata_path)) {
  # Trim each EIC to its peak boundaries + small margin, then impute edge NAs.
  # peakBoundary column names are printed above — adjust if different.
  final_eics <- lapply(seq_along(eics), function(i) {
    eic  <- eics[i]
    rt1  <- eic$left_boundary
    rt2  <- eic$right_boundary
    if (is.na(rt1) || is.na(rt2)) return(eic)
    eic  <- filterPeaksData(eic, variables = "rtime",
                            ranges = c(rt1 - 0.5, rt2 + 0.5))
    applyProcessing(eic)
  })
  final_eics <- concatenateChromatograms(final_eics)
  # Impute residual NAs (boundary edge artefacts from scan-grid misalignment)
  final_eics <- imputePeaksData(final_eics, method = "linear")
  save(final_eics, file = rdata_path)
} else {
  load(rdata_path)
}

stopifnot(all(final_eics$consensus_id == eics$consensus_id))
cat("Re-extracted EICs:", length(final_eics), "\n")

# Patch remaining edge NAs — linear imputation cannot fill leading/trailing NAs
# because there is no boundary value to interpolate toward.  At peak boundaries
# the signal approaches baseline, so 0 is the appropriate fill value.
# This block also runs on a loaded cache, so stale objects self-heal.
n_na <- sum(is.na(unlist(intensity(final_eics))))
cat("NAs after linear imputation:", n_na, "\n")
if (n_na > 0) {
  cat("  -> filling", n_na, "edge NAs with 0 (signal at peak boundary ≈ baseline)...\n")
  pd <- peaksData(final_eics)
  pd <- lapply(pd, function(df) {
    df$intensity[is.na(df$intensity)] <- 0
    df
  })
  peaksData(final_eics) <- pd
  cat("NAs after 0-fill:", sum(is.na(unlist(intensity(final_eics)))), "\n")
  save(final_eics, file = rdata_path)  # patch the cache so next load is already clean
} else {
  cat("  -> no edge NAs found.\n")
}
```

```{r compute-metrics-final}
# Second metrics round — computed on boundary-trimmed, imputed EICs.
# Only the three metrics needed by 4_mitigating_inter_lab_variance.qmd.
# (areaUnderTic and xicFwhm on narrow EICs are more accurate than on wide EICs.)
# Guard: skip if metrics are already stored in the loaded RData.
if (!("areaUnderTic" %in% names(chromData(final_eics)))) {
  cat("Computing tight-EIC metrics (areaUnderTic, xicFwhm, gaussianSimilarity)...\n")
  pk2 <- calculateMetrics(
    final_eics,
    metrics = c("areaUnderTic", "xicFwhm", "gaussianSimilarity"),
    na.rm = TRUE,
    filterEmptyObject = FALSE
  )
  # Store metrics in chromData(final_eics) so doc 4 can access them after
  # loading the same RData object it needs for compareChromatograms anyway.
  chromData(final_eics) <- cbind(chromData(final_eics), pk2)
  save(final_eics, file = rdata_path)
  cat("Metrics stored in chromData and RData re-saved.\n")
} else {
  cat("Tight-EIC metrics already in chromData — skipping calculateMetrics.\n")
  pk2 <- as.data.frame(chromData(final_eics))[,
    intersect(c("areaUnderTic", "xicFwhm",
                "gaussianSimilarity.gaussian_similarity",
                "gaussianSimilarity.gaussian_residuals"),
              names(chromData(final_eics)))
  ]
}

# Build the full analysis table for doc-3 analysis: tight-EIC metrics +
# QC metrics from the first round.
# left_boundary/right_boundary column names must match what peakBoundary() added.
pk_filtered <- cbind(
  chromData(eics)[, c(
    "lab", "consensus_id", "mixture", "polarity", "mz", "chrom_peak_id",
    "peakCount", "maxIntensity", "baselineIntensity", "signalToNoiseRatio",
    "peakWidth", "peakProminence", "left_boundary", "right_boundary"
  )],
  pk2
)

cat("pk_filtered columns:", paste(names(pk_filtered), collapse = ", "), "\n")
cat("Rows:", nrow(pk_filtered), "\n")

write.csv(
  pk_filtered,
  file = file.path(base_path, "object/peak_metrics_detected_all4labs.csv"),
  row.names = FALSE
)
```

# Peak Feature analysis

This section analyzes peak metrics to assess inter-laboratory reproducibility
at the individual peak level. We examine whether peaks with the same
consensus ID show consistent metrics across labs.

## Metric Summary Statistics by Lab

```{r metric-summary}
metric_summary <- pk_filtered |>
  group_by(lab) |>
  summarise(
    n_peaks = n(),
    mean_area = mean(areaUnderTic, na.rm = TRUE),
    median_area = median(areaUnderTic, na.rm = TRUE),
    mean_fwhm = mean(xicFwhm, na.rm = TRUE),
    median_fwhm = median(xicFwhm, na.rm = TRUE),
    mean_snr = mean(signalToNoiseRatio, na.rm = TRUE),
    median_snr = median(signalToNoiseRatio, na.rm = TRUE),
    mean_peakwidth = mean(peakWidth, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(metric_summary,
             caption = "Peak Metric Summary by Lab",
             digits = 2)
```

## Inter-Lab Coefficient of Variation

Compute CV for each consensus peak across all 4 labs:

```{r cv-analysis}
cv_by_peak <- pk_filtered |>
  group_by(consensus_id) |>
  summarise(
    n_labs = n(),
    cv_area = sd(areaUnderTic, na.rm = TRUE) / mean(areaUnderTic, na.rm = TRUE),
    cv_max = sd(maxIntensity, na.rm = TRUE) / mean(maxIntensity, na.rm = TRUE),
    cv_fwhm = sd(xicFwhm, na.rm = TRUE) / mean(xicFwhm, na.rm = TRUE),
    cv_peakwidth = sd(peakWidth, na.rm = TRUE) / mean(peakWidth, na.rm = TRUE),
    cv_snr = sd(signalToNoiseRatio, na.rm = TRUE) / mean(signalToNoiseRatio,
    na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_labs == 4)

cat("Consensus peaks with all 4 labs:", nrow(cv_by_peak), "\n\n")

cv_summary <- cv_by_peak |>
  summarise(
    across(starts_with("cv_"), list(
      median = ~median(.x, na.rm = TRUE),
      mean = ~mean(.x, na.rm = TRUE),
      q25 = ~quantile(.x, 0.25, na.rm = TRUE),
      q75 = ~quantile(.x, 0.75, na.rm = TRUE)
    ))
  )

# Reshape for display
cv_table <- data.frame(
  Metric = c("Area Under TIC", "Max Intensity", "FWHM", "Peak Width", "SNR"),
  Median_CV = c(cv_summary$cv_area_median, cv_summary$cv_max_median,
                cv_summary$cv_fwhm_median, cv_summary$cv_peakwidth_median,
                cv_summary$cv_snr_median),
  Mean_CV = c(cv_summary$cv_area_mean, cv_summary$cv_max_mean,
              cv_summary$cv_fwhm_mean, cv_summary$cv_peakwidth_mean,
              cv_summary$cv_snr_mean),
  IQR = c(cv_summary$cv_area_q75 - cv_summary$cv_area_q25,
          cv_summary$cv_max_q75 - cv_summary$cv_max_q25,
          cv_summary$cv_fwhm_q75 - cv_summary$cv_fwhm_q25,
          cv_summary$cv_peakwidth_q75 - cv_summary$cv_peakwidth_q25,
          cv_summary$cv_snr_q75 - cv_summary$cv_snr_q25)
)

knitr::kable(cv_table,
             caption = "Inter-Lab CV for Consensus Peaks",
             digits = 3)
```

## CV Distribution Visualization

```{r cv-distribution-plot, fig.width=10, fig.height=6}
cv_long <- cv_by_peak |>
  pivot_longer(
    cols = starts_with("cv_"),
    names_to = "metric",
    names_prefix = "cv_",
    values_to = "cv"
  ) |>
  mutate(metric = case_when(
    metric == "area" ~ "Area",
    metric == "max" ~ "Max Intensity",
    metric == "fwhm" ~ "FWHM",
    metric == "peakwidth" ~ "Peak Width",
    metric == "snr" ~ "SNR"
  ))

ggplot(cv_long |> filter(!is.na(cv) & cv < 2),
       aes(x = metric, y = cv, fill = metric)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
  geom_hline(yintercept = 0.3, linetype = "dashed", color = "red") +
  labs(
    title = "Inter-Lab Coefficient of Variation",
    subtitle = "For consensus peaks detected in all 4 labs",
    x = "Metric",
    y = "CV"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Shape metrics (FWHM, peak width) show lower CV than intensity metrics,
suggesting chromatographic behavior is more reproducible than ionization.

Note that CV (standard deviation / mean) is inherently scale-normalized:
it measures relative spread regardless of absolute intensity differences
between labs. Normalizing AUC or max intensity beforehand would not
change the CV values.

## Spearman Correlation with Recalculated Intensities

We use Spearman (rank-based) correlation rather than Pearson because
absolute intensity values differ substantially across labs due to
instrument sensitivity, ionization efficiency, and injection volume.
Spearman tests whether peaks that are relatively intense in one lab are
also relatively intense in another — i.e., whether the **rank order** of
peak intensities is preserved. This makes it robust to the non-linear
scale differences across labs without requiring explicit normalization.

```{r spearman-recalculated}
# Pivot to wide format
intensity_wide <- pk_filtered |>
  select(consensus_id, lab, areaUnderTic, maxIntensity) |>
  pivot_wider(
    id_cols = consensus_id,
    names_from = lab,
    values_from = c(areaUnderTic, maxIntensity)
  )

# Spearman correlation
area_cols <- paste0("areaUnderTic_", c("afekta", "hmgu", "icl", "cembio"))
area_matrix <- intensity_wide[, area_cols]
names(area_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_area <- cor(area_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("Spearman Correlation - Area Under TIC (Recalculated):\n")
print(round(cor_area, 3))
```

Afekta and Cembio show the strongest rank correlation (ρ ≈ 0.76),
consistent with their shared instrument platform. All other pairs are
moderate (ρ ≈ 0.4–0.5), indicating that relative peak intensity rankings
are partially conserved across labs but with substantial instrument-
specific variation.

# Peak shape analysis

This section examines the actual chromatographic peak profiles to assess
shape similarity across laboratories using both Pearson correlation and
Tanimoto similarity (from the `philentropy` package).

**Similarity metrics:**

- **Pearson correlation**: Measures linear relationship between intensity
  profiles. Sensitive to shape but invariant to scale.
- **Tanimoto similarity**: Originally for binary data, but can be applied
  to continuous vectors. More sensitive to differences in peak
  presence/absence.

## Peak Profile Visualization

```{r shape-example}
# Plot example consensus peak across all 4 labs
example_id <- unique(final_eics$consensus_id)[1]
eic_example <- final_eics[final_eics$consensus_id == example_id]

plotChromatograms(eic_example,
                  main = paste("Consensus ID:", example_id,
                               "- Labs:", paste(eic_example$lab,
                               collapse = ", ")))
```

## Peak Shape Similarity Function (Pearson + Tanimoto)

Pearson correlation and Tanimoto similarity both require vectors of equal
length. Peaks from different labs often have different numbers of data
points (due to scan rate or peak boundary differences). To handle this,
we interpolate all profiles within a consensus ID to match the length of
the profile with the most data points. Interpolation uses a normalized
0–1 x-axis (ignoring absolute RT), so we compare shape only, not
position.

```{r shape-similarity-function}
library(philentropy)

compute_shape_similarity <- function(eics_subset) {
  if (length(eics_subset) < 2) return(NULL)

  labs_present <- eics_subset$lab
  int_list <- intensity(eics_subset)
  n_points <- sapply(int_list, length)

  # Use the highest peakCount as the reference length
  target_n <- max(n_points)

  # Interpolate all profiles to target_n points (0-1 normalized x-axis)
  interp_matrix <- sapply(seq_along(eics_subset), function(i) {
    y <- int_list[[i]]
    y[is.na(y)] <- 0
    n <- length(y)
    if (n < 2) return(rep(0, target_n))
    if (n == target_n) return(y)
    approx(seq(0, 1, length.out = n), y,
           xout = seq(0, 1, length.out = target_n))$y
  })
  colnames(interp_matrix) <- labs_present

  # Normalize for Tanimoto
  interp_norm <- apply(interp_matrix, 2, function(x) {
    x[is.na(x) | x < 0] <- 0
    s <- sum(x)
    if (is.na(s) || s == 0) x else x / s
  })

  pairs <- combn(labs_present, 2, simplify = FALSE)

  results <- lapply(pairs, function(p) {
    v1 <- interp_matrix[, p[1]]
    v2 <- interp_matrix[, p[2]]

    # Pearson (NA if flat or contains NAs)
    sd1 <- sd(v1, na.rm = TRUE)
    sd2 <- sd(v2, na.rm = TRUE)
    pearson <- if (is.na(sd1) || is.na(sd2) || sd1 == 0 || sd2 == 0)
      NA_real_ else cor(v1, v2, method = "pearson", use = "complete.obs")

    # Tanimoto
    tanimoto <- tryCatch(
      1 - distance(rbind(interp_norm[, p[1]], interp_norm[, p[2]]),
                   method = "tanimoto", mute.message = TRUE),
      error = function(e) NA_real_
    )

    data.frame(
      pair = paste(p, collapse = ":"),
      pearson = pearson,
      tanimoto = as.numeric(tanimoto)
    )
  })

  do.call(rbind, results)
}
```

## Compute Shape Similarities (All Consensus Peaks)

```{r compute-shape-similarities, message=FALSE, warning=FALSE}
# Compute for all consensus peaks
all_consensus_ids <- unique(final_eics$consensus_id)
cat("Computing shape similarities for", length(all_consensus_ids),
    "consensus peaks...\n")

shape_results <- lapply(all_consensus_ids, function(cid) {
  eics_subset <- final_eics[final_eics$consensus_id == cid]
  res <- compute_shape_similarity(eics_subset)
  if (!is.null(res)) {
    res$consensus_id <- cid
  }
  res
}) |>
  do.call(rbind, args = _)

cat("Successfully computed:", nrow(shape_results), "pairwise comparisons\n")
cat("NA Pearson values:", sum(is.na(shape_results$pearson)), "\n")
cat("NA Tanimoto values:", sum(is.na(shape_results$tanimoto)), "\n")
```

## Shape Similarity Summary

```{r shape-summary-table}
if (!is.null(shape_results) && nrow(shape_results) > 0) {
  shape_summary <- shape_results |>
    group_by(pair) |>
    summarise(
      n = n(),
      pearson_median = median(pearson, na.rm = TRUE),
      pearson_mean = mean(pearson, na.rm = TRUE),
      pearson_sd = sd(pearson, na.rm = TRUE),
      tanimoto_median = median(tanimoto, na.rm = TRUE),
      tanimoto_mean = mean(tanimoto, na.rm = TRUE),
      tanimoto_sd = sd(tanimoto, na.rm = TRUE),
      .groups = "drop"
    )

  knitr::kable(shape_summary,
               caption = "Peak Shape Similarity by Lab Pair (Pearson & Tanimoto)",
               digits = 3)
}
```

## Peak Count Analysis

Do peaks with more data points produce better quality metrics and higher
inter-lab reproducibility? We examine two relationships:

1. **Peak count vs quality metrics** — Does higher `peakCount` correlate
   with better `peakBeta1` (Gaussian fit, 0–1) and `peakBeta2`
   (signal-to-noise of the beta fit)?
2. **Peak count vs inter-lab shape agreement** — Do consensus peaks with
   more data points show higher Pearson/Tanimoto similarity across labs?

### Peak Count vs Quality Metrics

```{r peak-count-quality, fig.width=10, fig.height=5}
# Per-EIC relationship: peakCount vs gaussianSimilarity metrics
# (peakBeta1/peakBeta2 renamed to gaussianSimilarity.* in new MsQuality)
gs_sim <- "gaussianSimilarity.gaussian_similarity"
gs_res <- "gaussianSimilarity.gaussian_residuals"

pq <- pk_filtered |>
  filter(!is.na(.data[[gs_sim]]), !is.na(.data[[gs_res]]))

cat("Spearman cor (peakCount vs gaussian_similarity):",
    round(cor(pq$peakCount, pq[[gs_sim]],
              method = "spearman", use = "complete.obs"), 3), "\n")
cat("Spearman cor (peakCount vs gaussian_residuals):",
    round(cor(pq$peakCount, pq[[gs_res]],
              method = "spearman", use = "complete.obs"), 3), "\n")

p1 <- ggplot(pq, aes(x = peakCount, y = .data[[gs_sim]])) +
  geom_bin2d(bins = 50) +
  scale_fill_viridis_c(option = "inferno", trans = "log10") +
  geom_smooth(method = "loess", se = FALSE, color = "red",
              linewidth = 0.8) +
  labs(
    title = "Peak Count vs Gaussian Similarity",
    x = "Peak Count (data points)",
    y = "Gaussian similarity (0\u20131)"
  ) +
  theme_minimal()

p2 <- ggplot(pq, aes(x = peakCount, y = .data[[gs_res]])) +
  geom_bin2d(bins = 50) +
  scale_fill_viridis_c(option = "inferno", trans = "log10") +
  geom_smooth(method = "loess", se = FALSE, color = "red",
              linewidth = 0.8) +
  labs(
    title = "Peak Count vs Gaussian Residuals",
    x = "Peak Count (data points)",
    y = "Gaussian residuals"
  ) +
  theme_minimal()

ggarrange(p1, p2, ncol = 2)
```

The negative Spearman correlation between `peakCount` and `peakBeta1`
(ρ ≈ −0.40) indicates that peaks with more data points tend to have
*lower* Gaussian-fit scores. This is expected: broader peaks (more scans)
deviate more from an ideal Gaussian shape. In contrast, `peakBeta2` shows
a weak positive correlation (ρ ≈ 0.20), suggesting that more data points
modestly improve the signal-to-noise of the fit.

### Peak Count vs Inter-Lab Shape Agreement

```{r peak-count-shape, fig.width=10, fig.height=5}
# Mean peakCount per consensus peak (across labs)
pc_summary <- pk_filtered |>
  group_by(consensus_id) |>
  summarise(
    mean_peakCount = mean(peakCount, na.rm = TRUE),
    min_peakCount = min(peakCount, na.rm = TRUE),
    .groups = "drop"
  )

# Median similarity per consensus peak
sim_summary <- shape_results |>
  group_by(consensus_id) |>
  summarise(
    median_pearson = median(pearson, na.rm = TRUE),
    median_tanimoto = median(tanimoto, na.rm = TRUE),
    .groups = "drop"
  )

pc_sim <- inner_join(pc_summary, sim_summary, by = "consensus_id") |>
  filter(!is.na(median_pearson), !is.na(median_tanimoto))

cat("Spearman cor (peakCount vs Pearson):",
    round(cor(pc_sim$mean_peakCount, pc_sim$median_pearson,
              method = "spearman", use = "complete.obs"), 3), "\n")
cat("Spearman cor (peakCount vs Tanimoto):",
    round(cor(pc_sim$mean_peakCount, pc_sim$median_tanimoto,
              method = "spearman", use = "complete.obs"), 3), "\n")

p1 <- ggplot(pc_sim, aes(x = mean_peakCount, y = median_pearson)) +
  geom_bin2d(bins = 50) +
  scale_fill_viridis_c(option = "inferno", trans = "log10") +
  geom_smooth(method = "loess", se = FALSE, color = "red",
              linewidth = 0.8) +
  labs(
    title = "Peak Count vs Pearson Correlation",
    x = "Mean Peak Count (across labs)",
    y = "Median Pearson (per consensus peak)"
  ) +
  theme_minimal()

p2 <- ggplot(pc_sim, aes(x = mean_peakCount, y = median_tanimoto)) +
  geom_bin2d(bins = 50) +
  scale_fill_viridis_c(option = "inferno", trans = "log10") +
  geom_smooth(method = "loess", se = FALSE, color = "red",
              linewidth = 0.8) +
  labs(
    title = "Peak Count vs Tanimoto Similarity",
    x = "Mean Peak Count (across labs)",
    y = "Median Tanimoto (per consensus peak)"
  ) +
  theme_minimal()

ggarrange(p1, p2, ncol = 2)
```

Peak count shows a weak negative correlation with Pearson similarity
(ρ ≈ −0.19) and essentially no relationship with Tanimoto (ρ ≈ 0.00).
More data points do not improve — and may slightly worsen — inter-lab
shape agreement, likely because broader peaks introduce more
opportunities for shape divergence between instruments.

## Pearson Correlation Distribution

The 0.9 threshold is a commonly used cutoff in chromatographic peak
matching to distinguish high-quality shape agreement from moderate or
poor matches.

The table below shows, for each lab pair, the percentage of consensus
peaks where the Pearson correlation exceeds 0.9 — i.e., the fraction of
peaks with high-quality shape agreement between the two labs.

```{r pearson-distribution, fig.width=10, fig.height=6}
if (!is.null(shape_results) && nrow(shape_results) > 0) {
  shape_pearson <- shape_results |> filter(!is.na(pearson))

  pct_above <- shape_pearson |>
    group_by(pair) |>
    summarise(
      n_comparisons = n(),
      n_above_0.9 = sum(pearson >= 0.9),
      pct_above_0.9 = round(100 * mean(pearson >= 0.9), 1),
      .groups = "drop"
    )
  knitr::kable(pct_above,
               caption = "Percentage of consensus peaks with Pearson >= 0.9",
               col.names = c("Lab Pair", "Total Comparisons",
                             "N above 0.9", "% above 0.9"))

  ggplot(shape_pearson, aes(x = pearson, fill = pair)) +
    geom_histogram(bins = 50, alpha = 0.7, color = "white",
                   linewidth = 0.2) +
    facet_wrap(~pair, scales = "free_y") +
    geom_vline(xintercept = 0.9, linetype = "dashed", color = "red") +
    labs(
      title = "Pearson Correlation Distribution by Lab Pair",
      subtitle = "Red dashed line = 0.9 (high-quality shape match)",
      x = "Pearson Correlation",
      y = "Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
}
```

Only 11–20% of consensus peaks reach the 0.9 Pearson threshold depending
on the lab pair, with afekta:hmgu highest (~20%) and afekta:icl lowest
(~11%). The distributions are left-skewed with a long tail of poorly
correlated peaks, suggesting that while many peaks have good shape
agreement, a substantial fraction diverges across labs.

## Tanimoto Similarity Distribution

```{r tanimoto-distribution, fig.width=10, fig.height=6}
if (!is.null(shape_results) && nrow(shape_results) > 0) {
  shape_tanimoto <- shape_results |> filter(!is.na(tanimoto))

  ggplot(shape_tanimoto, aes(x = tanimoto, fill = pair)) +
    geom_histogram(bins = 50, alpha = 0.7, color = "white", linewidth = 0.2) +
    facet_wrap(~pair, scales = "free_y") +
    geom_vline(xintercept = 0.9, linetype = "dashed", color = "red") +
    labs(
      title = "Tanimoto Similarity Distribution by Lab Pair",
      subtitle = "Red dashed line = 0.9 threshold",
      x = "Tanimoto Similarity",
      y = "Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
}
```

Tanimoto distributions are broadly similar across lab pairs, with most
values clustering between 0.4 and 0.8. The distributions are more
symmetric than Pearson's, reflecting Tanimoto's sensitivity to absolute
profile differences rather than just linear correlation.

## Pearson vs Tanimoto Comparison

```{r pearson-vs-tanimoto, fig.width=10, fig.height=6}
if (!is.null(shape_results) && nrow(shape_results) > 0) {
  # Filter out rows with NA in either metric
  shape_complete <- shape_results |>
    filter(!is.na(pearson), !is.na(tanimoto))

  cat("Complete cases:", nrow(shape_complete), "of", nrow(shape_results),
      "total comparisons\n")
  cat("Removed:", nrow(shape_results) - nrow(shape_complete),
      "rows with NA values\n")

  ggplot(shape_complete, aes(x = pearson, y = tanimoto)) +
    geom_bin2d(bins = 40) +
    scale_fill_viridis_c(option = "inferno", trans = "log10",
                         name = "Count") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "white", linewidth = 0.5) +
    facet_wrap(~pair) +
    labs(
      title = "Pearson vs Tanimoto Similarity",
      subtitle = "Density heatmap; diagonal = perfect agreement",
      x = "Pearson Correlation",
      y = "Tanimoto Similarity"
    ) +
    theme_minimal() +
    coord_cartesian(xlim = c(-0.5, 1), ylim = c(0, 1))
}
```

Pearson and Tanimoto generally agree but are not redundant: peaks with
high Pearson (similar shape) can have moderate Tanimoto (different
intensity distribution), and vice versa. The two metrics are
complementary — Pearson captures shape correlation while Tanimoto is
more sensitive to absolute profile overlap.

## Identify Poorly Matched Peaks

Peaks with low similarity scores may indicate:

- Mismatched consensus assignments
- Different peak shapes due to co-elution or matrix effects
- Instrument-specific artifacts

```{r low-similarity-peaks}
if (!is.null(shape_results) && nrow(shape_results) > 0) {
  # Find peaks with low Pearson OR low Tanimoto
  low_similarity <- shape_results |>
    filter(pearson < 0.7 | tanimoto < 0.7) |>
    arrange(pearson)

  cat("Peaks with low similarity (Pearson < 0.7 or Tanimoto < 0.7):",
      nrow(low_similarity), "\n")
  cat("Unique consensus IDs affected:",
      length(unique(low_similarity$consensus_id)), "\n\n")

  if (nrow(low_similarity) > 0) {
    knitr::kable(head(low_similarity, 10),
                 caption = "Top 10 Lowest Similarity Peaks",
                 digits = 3)
  }
}
```

## Visualize Low Similarity Example

```{r plot-low-similarity}
if (!is.null(shape_results) && nrow(shape_results) > 0) {
  low_sim_ids <- shape_results |>
    filter(pearson < 0.7) |>
    pull(consensus_id) |>
    unique()

  if (length(low_sim_ids) > 0) {
    # Plot first low similarity example
    example_low <- final_eics[final_eics$consensus_id == low_sim_ids[1]]
    plotChromatograms(example_low,
                      main = paste("Low Similarity Example - Consensus ID:",
                                   low_sim_ids[1]))
  }
}
```

## Save Shape Similarity Results

```{r save-shape-results}
if (!is.null(shape_results) && nrow(shape_results) > 0) {
  write.csv(
    shape_results,
    file = file.path(base_path, "object/peak_shape_similarity.csv"),
    row.names = FALSE
  )
  cat("Shape similarity results saved to peak_shape_similarity.csv\n")
}
```

## Summary

Key findings from the peak feature and shape analysis:

1. **CV analysis**: Shape metrics (FWHM, peak width) show lower inter-lab
   CV than intensity metrics, confirming that chromatographic behavior is
   more reproducible than ionization response.
2. **Spearman intensity correlation**: Afekta–Cembio show the strongest
   rank correlation (ρ ≈ 0.76), consistent with their shared platform.
   Other pairs are moderate (ρ ≈ 0.4–0.5).
3. **Peak count**: More data points do not improve shape agreement across
   labs. Broader peaks actually show slightly lower Gaussian-fit scores
   (`gaussianSimilarity.gaussian_similarity`), though the residuals
   (`gaussianSimilarity.gaussian_residuals`) improve modestly.
4. **Shape similarity**: Median Pearson correlations range from 0.59
   (hmgu:icl) to 0.73 (afekta:hmgu). Only 11–20% of peaks exceed the
   0.9 threshold, indicating room for improvement.
5. **Pearson vs Tanimoto**: The two metrics are complementary — Pearson
   captures shape correlation while Tanimoto is more sensitive to
   absolute profile overlap.

# Next Steps

1. Peak metrics saved to `peak_metrics_detected_all4labs.csv`
2. Filtered EICs saved to `eics_filtered_detected_all4labs.RData`
3. Run Phase 5 in `2_global_analysis.qmd` for full integration
4. Apply peak-level normalization and recompute ICC
5. Expand to Afekta–Cembio overlap (50k peaks)
6. Redo analysis on imputed data using
   `imputePeaksData(eics, method = "gaussian")`



