---
title: "Peak Feature Calculation — Detected Peaks"
format:
  html:
    toc: true
execute:
  dir: project
---

```{r}
library(BiocParallel)
bp <- SnowParam(workers = 4)
register(bp)
```

# Introduction

This document performs peak feature calculation and analysis for the detected
detected ion chromatograms (EICs). It reuses the processed detected EICs, computes peak
metrics, and runs downstream comparisons and stability analyses.

- Use all detected peaks from the beginning, in the chromData have info
  whether they are annotated
- Only keep peaks overlapping the 4 labs to create a common feature set

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)
library(UpSetR)
library(MsQuality)
library(igraph)

base_path <- here::here("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
source(here::here("1_preprocessing/setup.R"))
```


# Create Consensus Peak Table Across Labs

```{r}
pt_afekta <- read.csv(file.path(
  base_path,
  "object/detected_peaks_afekta_HE.csv"
))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(
  base_path,
  "object/detected_peaks_cembio_HE.csv"
))
lab_counts <- c(
  afekta = nrow(pt_afekta),
  hmgu = nrow(pt_hmgu),
  icl = nrow(pt_icl),
  cembio = nrow(pt_cembio)
)
barplot(
  lab_counts,
  main = "Number of Detected Peaks per Lab",
  ylab = "Number of detected peaks",
  xlab = "Labs",
  col = "steelblue"
)
```

## Method

Correspondence is computed per sample (one mixture × polarity), on
**all detected peaks** from the four labs (no annotation filter). For
each sample:

1. Peaks from the four labs are pooled and sorted by m/z.
2. A graph is built where peaks are nodes and an edge connects two peaks
   whenever |Δmz| ≤ `PPM_TOL` × m/z × 10⁻⁶ **and** |Δrt| ≤ `RT_TOL` s.
   The edge list is constructed by a two-pointer sweep over the
   m/z-sorted peaks.
3. Connected components of the graph are the consensus peaks; every
   peak in the same component shares a candidate match with at least one
   other peak in the component.
4. When a single lab contributes more than one peak to a component, the
   peak closest to the centroid (anchored on labs that contribute
   exactly one peak) is kept, where distance is mz+rt distance
   normalised by `PPM_TOL` and `RT_TOL`.

Only m/z and RT are used to define correspondence — no FWHM or other
shape-based information.

```{r build-consensus-peaks}
PPM_TOL <- 20    # m/z tolerance (parts per million)
RT_TOL  <- 15    # RT tolerance (seconds)
LABS    <- c("afekta", "hmgu", "icl", "cembio")

# Pool peaks from the four labs, tag with lab id, drop NAs in intb.
pt_all <- rbind(
  transform(pt_afekta, lab = "afekta"),
  transform(pt_hmgu,   lab = "hmgu"),
  transform(pt_icl,    lab = "icl"),
  transform(pt_cembio, lab = "cembio")
)
pt_all <- pt_all[!is.na(pt_all$intb), ]

# Per-sample mz+rt graph grouping (two-pointer sweep + connected components)
group_sample <- function(pt_all, sample_id,
                         ppm = PPM_TOL, rt_tol = RT_TOL) {
  sub <- pt_all[pt_all$sample == sample_id, ]
  sub <- sub[order(sub$mz), ]
  sub$peak_idx <- seq_len(nrow(sub))

  n      <- nrow(sub)
  mz_vec <- sub$mz
  rt_vec <- sub$rt

  cap  <- 1024L # what is this cap ? 
  from <- integer(cap); to <- integer(cap); k <- 0L

  for (i in seq_len(n - 1)) {
    tol_i <- mz_vec[i] * ppm * 1e-6
    j     <- i + 1L
    while (j <= n && mz_vec[j] - mz_vec[i] <= tol_i) {
      if (abs(rt_vec[j] - rt_vec[i]) <= rt_tol) {
        k <- k + 1L
        if (k > cap) {
          cap <- cap * 2L
          length(from) <- cap; length(to) <- cap
        }
        from[k] <- i; to[k] <- j
      }
      j <- j + 1L
    }
  }
  edges <- data.frame(from = from[seq_len(k)], to = to[seq_len(k)])

  g <- graph_from_data_frame(edges, directed = FALSE,
                             vertices = data.frame(name = sub$peak_idx))
  sub$consensus_id <- components(g)$membership
  sub
}

corr_cache <- file.path(base_path, "object/correspondence_grouped.rds")
if (file.exists(corr_cache)) {
  grouped <- readRDS(corr_cache)
} else {
  samples <- sort(unique(pt_all$sample))
  grouped <- do.call(rbind, lapply(samples, function(s) {
    g <- group_sample(pt_all, s)
    g$consensus_id <- paste(s, g$consensus_id, sep = "_")
    g
  }))
  grouped$consensus_id <- as.integer(factor(grouped$consensus_id))
  saveRDS(grouped, corr_cache)
}
cat("Pooled peaks:", nrow(grouped),
    "in", length(unique(grouped$consensus_id)), "consensus groups\n")
```

```{r disambiguate-and-build-ct}
# Disambiguate: when one lab contributes >1 peak to a group, pick the peak
# closest to the centroid defined by the anchor labs (those contributing
# exactly one peak).
disambig <- grouped |>
  group_by(consensus_id) |>
  mutate(
    n_per_lab = ave(seq_along(lab), lab, FUN = length),
    anchor    = n_per_lab == 1L,
    ref_mz    = if (any(anchor)) median(mz[anchor]) else median(mz),
    ref_rt    = if (any(anchor)) median(rt[anchor]) else median(rt),
    score     = abs(mz - ref_mz) / (ref_mz * PPM_TOL * 1e-6) +
                abs(rt - ref_rt) / RT_TOL
  ) |>
  group_by(consensus_id, lab) |>
  slice_min(score, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(-n_per_lab, -anchor, -ref_mz, -ref_rt, -score, -peak_idx)

# Pivot to wide consensus-table format expected downstream:
# one row per (sample, consensus_id), one column per lab.
ct_all_samples <- disambig |>
  select(sample, consensus_id, lab, chrom_peak_id) |>
  pivot_wider(
    id_cols     = c(sample, consensus_id),
    names_from  = lab,
    values_from = chrom_peak_id,
    names_prefix = "chrom_peak_id_"
  ) |>
  mutate(
    idx_afekta = chrom_peak_id_afekta,
    idx_hmgu   = chrom_peak_id_hmgu,
    idx_icl    = chrom_peak_id_icl,
    idx_cembio = chrom_peak_id_cembio,
    n_labs_matched = (!is.na(chrom_peak_id_afekta)) +
                    (!is.na(chrom_peak_id_hmgu))   +
                    (!is.na(chrom_peak_id_icl))    +
                    (!is.na(chrom_peak_id_cembio))
  ) |>
  as.data.frame()

cat("Consensus peaks (any lab count):", nrow(ct_all_samples), "\n")
cat("Consensus peaks present in all 4 labs:",
    sum(ct_all_samples$n_labs_matched == 4), "\n")
```


```{r}
# Add detection flags to consensus_matches
ct_all_samples <- ct_all_samples |>
  mutate(
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  )

cat("Peak counts by number of labs detected:\n")
print(table(ct_all_samples$n_labs_matched))
```

```{r}
# Replace mixture column from 1.1 to 1_1
meta$mixture <- gsub("\\.", "_", meta$Mixture)
nc_mix <- table(meta$mixture)
# This shows the number of compounds per mixture.

# Create mapping of sample to mixture name using pt_* data
sample_to_mixture <- pt_afekta |>
  select(sample, mixture) |>
  distinct()

# Below we can summarize the number of peaks detected per sample with compound count
mixture_summary <- lapply(split(ct_all_samples, ct_all_samples$sample), function(df) {
  sample_id <- unique(df$sample)
  mixture_name <- sample_to_mixture$mixture[sample_to_mixture$sample == sample_id]

  res <- data.frame(
    sample = sample_id,
    mixture = mixture_name,
    detected_4labs = sum(df$n_labs_matched == 4),
    detected_3labs = sum(df$n_labs_matched == 3),
    detected_2labs = sum(df$n_labs_matched == 2),
    detected_1lab = sum(df$n_labs_matched == 1),
    total_peaks = nrow(df),
    n_compounds = as.integer(nc_mix[mixture_name])
  )
  res
})

mixture_summary_df <- do.call(rbind, mixture_summary)
rownames(mixture_summary_df) <- NULL

knitr::kable(mixture_summary_df, caption = "Peak Detection Summary by Mixture")
```


## Upset Plot - Lab Overlap

```{r upset-plot, message = FALSE, warning = FALSE}
# Prepare binary matrix for upset plot
upset_data <- ct_all_samples |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(
    afekta = as.integer(afekta_detected),
    hmgu = as.integer(hmgu_detected),
    icl = as.integer(icl_detected),
    cembio = as.integer(cembio_detected)
  ) |>
  select(afekta, hmgu, icl, cembio)

# Convert to data frame with rownames for upset function
upset_data_df <- as.data.frame(upset_data)
rownames(upset_data_df) <- paste0("peak_", seq_len(nrow(upset_data_df)))

# Only plot intersections with at least 1 peak (exclude empty combinations)
upset(upset_data_df,
      sets = c("afekta", "hmgu", "icl", "cembio"),
      main.bar.color = "steelblue",
      sets.bar.color = "darkred",
      text.scale = c(2, 2, 2, 1.5, 2, 2),
      point.size = 3,
      line.size = 1.2)
```

## Pairwise Lab Comparisons

```{r pairwise-overlap}
# Create pairwise overlap summary
labs <- c("afekta", "hmgu", "icl", "cembio")
pairwise_summary <- data.frame()

for (i in seq_along(labs)) {
  for (j in seq_along(labs)) {
    if (i < j) {
      lab1 <- labs[i]
      lab2 <- labs[j]

      col1 <- paste0(lab1, "_detected")
      col2 <- paste0(lab2, "_detected")

      both <- sum(ct_all_samples[[col1]] & ct_all_samples[[col2]])
      only_lab1 <- sum(ct_all_samples[[col1]] & !ct_all_samples[[col2]])
      only_lab2 <- sum(!ct_all_samples[[col1]] & ct_all_samples[[col2]])

      total_lab1 <- sum(ct_all_samples[[col1]])
      total_lab2 <- sum(ct_all_samples[[col2]])

      pairwise_summary <- rbind(pairwise_summary, data.frame(
        lab1 = lab1,
        lab2 = lab2,
        both_detected = both,
        only_lab1 = only_lab1,
        only_lab2 = only_lab2,
        total_lab1 = total_lab1,
        total_lab2 = total_lab2,
        jaccard_index = round(100 * both / (both + only_lab1 + only_lab2), 1),
        overlap_percent_lab1 = round(100 * both / total_lab1, 1),
        overlap_percent_lab2 = round(100 * both / total_lab2, 1)
      ))
    }
  }
}

knitr::kable(pairwise_summary, caption = "Pairwise Lab Overlap - Detailed Metrics")
```

As expected, afekta and cembio have the highest overlap.

```{r}
ct_filtered <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Consensus peaks retained:", nrow(ct_filtered), "\n")
cat(
  "Percentage of detected peaks:",
  round(100 * nrow(ct_filtered) / nrow(ct_all_samples), 1),
  "%\n"
)
```

Build the final consensus peak table with chrom_peak_id mapping:
Save the consensus table for downstream analysis:

```{r save-consensus}
write.csv(
  ct_all_samples,
  file = file.path(base_path, "object/ct_all_samples.csv"),
  row.names = FALSE
)
write.csv(
  ct_filtered,
  file = file.path(base_path, "object/ct_filtered4.csv"),
  row.names = FALSE
)
```

## Summary Tables

### Overlap: Cembio vs Afekta

```{r cembio-afekta-overlap}
cembio_afekta_summary <- data.frame(
  category = c("Both detected", "Only afekta", "Only cembio", "Neither"),
  count = c(
    sum(ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected)
  )
)

cembio_afekta_summary <- cembio_afekta_summary |>
  mutate(percent = round(100 * count / sum(count), 1))

knitr::kable(cembio_afekta_summary,
             caption = "Peak Overlap: Cembio vs Afekta")
```

so we could look at 2526 detected peaks ( for 2 mixture), that's really nice.

### Overlap: All 4 Labs

```{r all-labs-overlap}
all_labs_summary <- data.frame(
  detection_pattern = character(),
  n_peaks = integer(),
  percent = numeric()
)

for (combo in 0:15) {
  afekta <- as.logical(bitwAnd(combo, 1))
  hmgu <- as.logical(bitwAnd(combo, 2))
  icl <- as.logical(bitwAnd(combo, 4))
  cembio <- as.logical(bitwAnd(combo, 8))

  n <- sum(
    ct_all_samples$afekta_detected == afekta &
    ct_all_samples$hmgu_detected == hmgu &
    ct_all_samples$icl_detected == icl &
    ct_all_samples$cembio_detected == cembio
  )

  if (n > 0) {
    pattern <- paste0(
      if_else(afekta, "A", "-"),
      if_else(hmgu, "H", "-"),
      if_else(icl, "I", "-"),
      if_else(cembio, "C", "-")
    )

    all_labs_summary <- rbind(all_labs_summary, data.frame(
      detection_pattern = pattern,
      n_peaks = n,
      percent = round(100 * n / nrow(ct_all_samples), 1)
    ))
  }
}

all_labs_summary <- all_labs_summary |>
  arrange(desc(n_peaks))

knitr::kable(all_labs_summary,
             caption = "Peak Detection Patterns: All 4 Labs (A=afekta, H=hmgu, I=icl, C=cembio)")
```



# Peak Feature Summary by Lab (Detected)

## Prepare subset for downstream analysis

```{r prepare-subsets}
# Peaks present in ALL 4 labs (main analysis subset)
all_four_labs <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Peaks in all 4 labs:", nrow(all_four_labs), "\n")
```

# Creating eics
## 4 labs

```{r}
pt <- all_four_labs |>
  mutate(consensus_id = row_number()) |>
  select(consensus_id, sample, chrom_peak_id_afekta, chrom_peak_id_hmgu,
  chrom_peak_id_icl, chrom_peak_id_cembio) |>
  pivot_longer(
    cols = c(chrom_peak_id_afekta, chrom_peak_id_hmgu, chrom_peak_id_icl,
    chrom_peak_id_cembio),
    names_to = "lab",
    names_prefix = "chrom_peak_id_",
    values_to = "chrom_peak_id"
  ) |>
  filter(!is.na(chrom_peak_id))
```

```{r}
# Combine all lab peak tables with lab identifier for simple merge
pt_all <- rbind(
  transform(pt_afekta, lab = "afekta"),
  transform(pt_hmgu, lab = "hmgu"),
  transform(pt_icl, lab = "icl"),
  transform(pt_cembio, lab = "cembio")
)

# Join peak info using merge instead of case_when
join_cols <- c("chrom_peak_id", "lab", "mzmin", "mzmax", "rtmin", "rtmax",
               "mz", "polarity", "mixture", "into", "intb", "maxo")
pt <- merge(pt, pt_all[, join_cols], by = c("chrom_peak_id", "lab"), all.x = TRUE)

# Rename and adjust boundaries
names(pt)[names(pt) == "mzmin"] <- "mzMin"
names(pt)[names(pt) == "mzmax"] <- "mzMax"
names(pt)[names(pt) == "rtmin"] <- "rtMin"
names(pt)[names(pt) == "rtmax"] <- "rtMax"

pt$rtMin <- pt$rtMin - 2
pt$rtMax <- pt$rtMax + 2
pt$mzMin <- pt$mzMin - 0.02
pt$mzMax <- pt$mzMax + 0.02
pt$polarity <- ifelse(pt$sample %in% 1:19, 1L, 0L)

write.csv(
  pt,
  file = file.path(base_path, "object/peak_table_detected_all4labs.csv"),
  row.names = FALSE
)
```

```{r}
library(Chromatograms)

if (!file.exists(file.path(base_path, "object/eics_detected_all4labs.RData"))) {
  load(file.path(base_path, "object/sp_full.RData"))
  tic_full <- Chromatograms(
    sp_full,
    summarize.method = "sum",
    factorize.by = "dataOrigin",
    spectraVariables = c("mixture", "lab")
  )
  eics <- chromExtract(
    tic_full,
    peak.table = pt,
    by = c("polarity", "mixture", "lab")
  )
  nrow(pt) == length(eics)
  eics <- setBackend(eics, ChromBackendMemory())
  save(eics, file = file.path(base_path, "object/eics_detected_all4labs.RData"))
} else {
  load(file.path(base_path, "object/eics_detected_all4labs.RData"))
}

eics
```

## Compute metrics


```{r compute-metrics}
# Step 1: peak boundaries — now in Chromatograms, no longer a calculateMetrics metric
cat("Computing peak boundaries...\n")
pb <- Chromatograms::peakBoundary(eics)
cat("peakBoundary columns:", paste(names(pb), collapse = ", "), "\n")
chromData(eics) <- cbind(chromData(eics), pb)

# Step 2: quality metrics used for filtering and downstream doc-3 analysis
# Removed: peakBoundary (above), gaussianSimilarity (computed on narrow EICs below)
# Removed: f = seq_len() — not a valid parameter in current MsQuality
pk <- calculateMetrics(
  eics,
  metrics = c(
    "peakCount", "xicFwhm", "maxIntensity",
    "baselineIntensity", "signalToNoiseRatio",
    "peakWidth", "peakProminence"
  ),
  na.rm = TRUE,
  filterEmptyObject = FALSE,
  probs = 0.05
)

chromData(eics) <- cbind(chromData(eics), pk)
cat("Metrics computed for", length(eics), "EICs\n")
```

## Filter bad peaks

Remove consensus peaks with poor quality metrics in a single step.

```{r filter-bad-peaks}
# Consolidate all filtering criteria
bad_ids <- unique(c(
  eics$consensus_id[is.na(eics$xicFwhm)],           # NA FWHM
  eics$consensus_id[eics$peakProminence < 1.5],     # Low prominence
  eics$consensus_id[eics$peakWidth < 1],            # Too narrow
  eics$consensus_id[eics$peakWidth > 30]            # Too wide
))

cat("Removing", length(bad_ids), "consensus peaks with poor metrics\n")
eics <- eics[!eics$consensus_id %in% bad_ids]
cat("Remaining EICs:", length(eics), "\n")
cat("Unique consensus IDs:", length(unique(eics$consensus_id)), "\n")
```

### Concordant peaks per mixture

```{r concordant-per-mixture, fig.width=10, fig.height=4}
#| fig-cap: "Number of consensus peaks detected in all 4 labs per mixture (post-filter)."
mix_counts <- as.data.frame(chromData(eics)) |>
  distinct(consensus_id, mixture, polarity) |>
  count(mixture, polarity, name = "n_peaks") |>
  mutate(polarity = ifelse(polarity == 1L, "Positive", "Negative"))

ggplot(mix_counts, aes(x = mixture, y = n_peaks)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  geom_text(aes(label = n_peaks), vjust = -0.3, size = 3) +
  facet_wrap(~polarity, ncol = 1, scales = "free_y") +
  labs(x = "Mixture", y = "Concordant peaks (all 4 labs)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Re-extract EICs using peak boundaries

```{r re-extract}
# NOTE: delete the cached RData below if you want to force re-extraction
# (necessary after changes to boundary computation or imputation).
rdata_path <- file.path(base_path, "object/eics_filtered_detected_all4labs.RData")

if (!file.exists(rdata_path)) {
  # Trim each EIC to its peak boundaries + small margin, then impute edge NAs.
  # peakBoundary column names are printed above — adjust if different.
  final_eics <- lapply(seq_along(eics), function(i) {
    eic  <- eics[i]
    rt1  <- eic$left_boundary
    rt2  <- eic$right_boundary
    if (is.na(rt1) || is.na(rt2)) return(eic)
    eic  <- filterPeaksData(eic, variables = "rtime",
                            ranges = c(rt1 - 0.5, rt2 + 0.5))
    applyProcessing(eic)
  })
  final_eics <- concatenateChromatograms(final_eics)
  # Impute residual NAs (boundary edge artefacts from scan-grid misalignment)
  final_eics <- imputePeaksData(final_eics, method = "linear")
  save(final_eics, file = rdata_path)
} else {
  load(rdata_path)
}

stopifnot(all(final_eics$consensus_id == eics$consensus_id))
cat("Re-extracted EICs:", length(final_eics), "\n")

# Patch remaining edge NAs — linear imputation cannot fill leading/trailing NAs
# because there is no boundary value to interpolate toward.  At peak boundaries
# the signal approaches baseline, so 0 is the appropriate fill value.
# This block also runs on a loaded cache, so stale objects self-heal.
n_na <- sum(is.na(unlist(intensity(final_eics))))
cat("NAs after linear imputation:", n_na, "\n")
if (n_na > 0) {
  cat("  -> filling", n_na, "edge NAs with 0 (signal at peak boundary ≈ baseline)...\n")
  pd <- peaksData(final_eics)
  pd <- lapply(pd, function(df) {
    df$intensity[is.na(df$intensity)] <- 0
    df
  })
  peaksData(final_eics) <- pd
  cat("NAs after 0-fill:", sum(is.na(unlist(intensity(final_eics)))), "\n")
  save(final_eics, file = rdata_path)  # patch the cache so next load is already clean
} else {
  cat("  -> no edge NAs found.\n")
}
```

```{r compute-metrics-final}
# Second metrics round — computed on boundary-trimmed, imputed EICs.
# Only the three metrics needed by 4_mitigating_inter_lab_variance.qmd.
# (areaUnderTic and xicFwhm on narrow EICs are more accurate than on wide EICs.)
# Guard: skip if metrics are already stored in the loaded RData.
if (!("areaUnderTic" %in% names(chromData(final_eics)))) {
  cat("Computing tight-EIC metrics (areaUnderTic, xicFwhm, gaussianSimilarity)...\n")
  pk2 <- calculateMetrics(
    final_eics,
    metrics = c("areaUnderTic", "xicFwhm", "gaussianSimilarity"),
    na.rm = TRUE,
    filterEmptyObject = FALSE
  )
  # Store metrics in chromData(final_eics) so doc 4 can access them after
  # loading the same RData object it needs for compareChromatograms anyway.
  chromData(final_eics) <- cbind(chromData(final_eics), pk2)
  save(final_eics, file = rdata_path)
  cat("Metrics stored in chromData and RData re-saved.\n")
} else {
  cat("Tight-EIC metrics already in chromData — skipping calculateMetrics.\n")
  pk2 <- as.data.frame(chromData(final_eics))[,
    intersect(c("areaUnderTic", "xicFwhm",
                "gaussianSimilarity.gaussian_similarity",
                "gaussianSimilarity.gaussian_residuals"),
              names(chromData(final_eics)))
  ]
}

# Build the full analysis table for doc-3 analysis: tight-EIC metrics +
# QC metrics from the first round.
# left_boundary/right_boundary column names must match what peakBoundary() added.
pk_filtered <- cbind(
  chromData(eics)[, c(
    "lab", "consensus_id", "mixture", "polarity", "mz", "chrom_peak_id",
    "peakCount", "maxIntensity", "baselineIntensity", "signalToNoiseRatio",
    "peakWidth", "peakProminence", "left_boundary", "right_boundary"
  )],
  pk2
)

cat("pk_filtered columns:", paste(names(pk_filtered), collapse = ", "), "\n")
cat("Rows:", nrow(pk_filtered), "\n")

write.csv(
  pk_filtered,
  file = file.path(base_path, "object/peak_metrics_detected_all4labs.csv"),
  row.names = FALSE
)
```

## Create Spectra - Consensus Signal (post-filter)

Spectra filtered to the m/z × RT rectangles of the **filtered +
boundary-trimmed** consensus peaks, per (lab, mixture, polarity)
sample. Same pattern as `detect_signal()` in
`rt_analysis_function.R` — `filterPeaksRanges()` keeps only mass peaks
inside any of that sample's consensus windows.

```{r build-sp-consensus}
library(Spectra)

if (!file.exists(file.path(base_path, "object/sp_consensus.RData"))) {
  load(file.path(base_path, "object/sp_full.RData"))

  # Per-(lab, mixture, polarity) consensus peak windows, boundary-trimmed
  cd <- as.data.frame(chromData(final_eics))
  cd <- cd[!is.na(cd$left_boundary) & !is.na(cd$right_boundary), ]

  # Match spectra ↔ consensus windows on (lab, mixture, polarity)
  sp_key <- paste(sp_full$lab, sp_full$mixture, polarity(sp_full), sep = "_")
  cd_key <- paste(cd$lab, cd$mixture, cd$polarity, sep = "_")

  # Per-sample: filter mass peaks to consensus rectangles, then apply
  # the lazy processing queue so concatenateSpectra() will accept the
  # result (mirrors detect_signal() at rt_analysis_function.R:104-117).
  filtered_list <- lapply(unique(sp_key), function(key) {
    s <- sp_full[sp_key == key]
    w <- cd[cd_key == key, , drop = FALSE]
    if (nrow(w) == 0) {
      return(Spectra::applyProcessing(s[FALSE]))
    }
    s <- Spectra::filterPeaksRanges(
      s,
      mz = as.matrix(w[, c("mzMin", "mzMax")]),
      rtime = as.matrix(w[, c("left_boundary", "right_boundary")]),
      keep = TRUE
    )
    Spectra::applyProcessing(s)
  })
  sp_consensus <- lapply(filtered_list, applyProcessing)

  sp_consensus <- concatenateSpectra(sp_consensus)
  sp_consensus <- filterEmptySpectra(sp_consensus)
  save(sp_consensus, file = file.path(base_path, "object/sp_consensus.RData"))
} else {
  load(file.path(base_path, "object/sp_consensus.RData"))
}
```

## Create Total Ion Current (TIC) - Consensus Signal

Same `Chromatograms()` pattern used for `tic_full` / `tic_detect` in
`1_object_creation.qmd`, applied to `sp_consensus`.

```{r build-tic-consensus}
if (!file.exists(file.path(base_path, "object/tic_consensus.RData"))) {
  tic_consensus <- Chromatograms(sp_consensus,
                                 summarize.method = "sum",
                                 factorize.by     = "dataOrigin",
                                 spectraVariables = c("mixture", "lab"))
  tic_consensus <- setBackend(tic_consensus, ChromBackendMemory())
  save(tic_consensus,
       file = file.path(base_path, "object/tic_consensus.RData"))
} else {
  load(file.path(base_path, "object/tic_consensus.RData"))
}

cat("tic_consensus:", length(tic_consensus), "chromatograms\n")
```

# Peak Quality Diagnostics

## Distributions per Lab

Per-lab density of one intensity metric (`areaUnderTic`, log2-scaled)
and four shape/quality metrics (`peakWidth`, `signalToNoiseRatio`,
`xicFwhm`, `gaussianSimilarity`), shown separately for positive and
negative polarity. Each panel has its own independent x and y range
so distributions are not squashed by metrics on a wider scale.

```{r quality-distributions, fig.width=12, fig.height=6}
#| fig-cap: "Peak-quality metric distributions per lab, by polarity. Each panel has its own x/y range. Filtered consensus peaks only."
keep_metrics <- c("log2(areaUnderTic)", "peakWidth",
                  "signalToNoiseRatio", "xicFwhm", "gaussianSimilarity")

quality_long <- pk_filtered |>
  mutate(polarity = ifelse(polarity == 1L, "Positive", "Negative")) |>
  transmute(
    lab, polarity,
    `log2(areaUnderTic)` = log2(areaUnderTic),
    peakWidth, signalToNoiseRatio, xicFwhm,
    gaussianSimilarity   = `gaussianSimilarity.gaussian_similarity`
  ) |>
  pivot_longer(-c(lab, polarity),
               names_to = "metric", values_to = "value") |>
  filter(!is.na(value), is.finite(value)) |>
  mutate(metric = factor(metric, levels = keep_metrics))

ggplot(quality_long, aes(x = value, fill = lab, colour = lab)) +
  geom_density(alpha = 0.25, linewidth = 0.5) +
  facet_wrap(vars(polarity, metric), scales = "free", ncol = 5) +
  scale_fill_viridis_d(option = "turbo") +
  scale_colour_viridis_d(option = "turbo") +
  labs(title = "Peak-quality distributions per lab",
       subtitle = paste0("Consensus peaks retained after filter ",
                         "(n = ", n_distinct(pk_filtered$consensus_id), ")"),
       x = NULL, y = "Density") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Inter-Lab Relative Abundance Correlation

Absolute `areaUnderTic` is not directly comparable across labs —
instrument sensitivity, ionisation efficiency and injection volume
shift the whole intensity scale. To make labs comparable, for each
peak we compute a log2 fold change of its AUC against the lab's
own median AUC, within each polarity:

`rel_abund[peak, lab] = log2( AUC[peak, lab] / median(AUC[, lab]) )`

The lab median becomes 0, a peak twice as intense as the lab median is
+1, half-median is −1, regardless of the lab's absolute scale.

For each consensus peak we then have a 4-vector
`(rel_abund_afekta, rel_abund_cembio, rel_abund_hmgu, rel_abund_icl)`.
Stacking those vectors across all consensus peaks gives, per lab pair,
two columns of relative abundances over the same set of peaks.
**Pearson r between those two columns is the inter-lab relative
abundance correlation** — it asks whether peaks that are above-median
in lab A are also above-median in lab B.

Polarities are kept separate because positive and negative AUCs are
on different scales and would otherwise need their own normalisation.

```{r abundance-correlation}
LABS <- c("afekta", "cembio", "hmgu", "icl")

abund_wide <- pk_filtered |>
  mutate(polarity = ifelse(polarity == 1L, "Positive", "Negative")) |>
  filter(!is.na(areaUnderTic), areaUnderTic > 0) |>
  group_by(lab, polarity) |>
  mutate(rel_abund = log2(areaUnderTic / median(areaUnderTic))) |>
  ungroup() |>
  select(consensus_id, polarity, lab, rel_abund) |>
  group_by(consensus_id, polarity) |>
  filter(n() == 4) |>
  ungroup() |>
  pivot_wider(names_from = lab, values_from = rel_abund) |>
  drop_na(all_of(LABS))

lab_pairs <- combn(LABS, 2, simplify = FALSE)

pair_corrs <- do.call(rbind, lapply(c("Positive", "Negative"), function(pol) {
  sub <- abund_wide[abund_wide$polarity == pol, ]
  do.call(rbind, lapply(lab_pairs, function(p) {
    v <- sub[, p]
    data.frame(
      polarity      = pol,
      pair          = paste(p, collapse = ":"),
      same_platform = identical(sort(p), c("afekta", "cembio")),
      r             = cor(v[[1]], v[[2]]),
      n             = nrow(v)
    )
  }))
}))
pair_corrs <- pair_corrs[order(pair_corrs$polarity, -pair_corrs$r), ]

knitr::kable(transform(pair_corrs, r = round(r, 3)),
  caption = "Inter-lab Pearson r on log2(AUC / lab median), by polarity. Consensus peaks present in all 4 labs.")
```

```{r abundance-scatter, fig.width=11, fig.height=8}
#| fig-cap: "Inter-lab log2(AUC / lab median) per lab pair. Red dashed = identity; black = OLS fit. Same-platform afekta:cembio highlighted."
abund_pairs <- do.call(rbind, lapply(lab_pairs, function(p) {
  data.frame(
    consensus_id  = abund_wide$consensus_id,
    polarity      = abund_wide$polarity,
    x             = abund_wide[[p[1]]],
    y             = abund_wide[[p[2]]],
    pair          = paste(p, collapse = ":"),
    same_platform = identical(sort(p), c("afekta", "cembio"))
  )
}))

ggplot(abund_pairs, aes(x = x, y = y, colour = same_platform)) +
  geom_abline(slope = 1, intercept = 0,
              colour = "red", linetype = "dashed") +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "black", linewidth = 0.5) +
  scale_colour_manual(values = c(`FALSE` = "grey40",
                                 `TRUE`  = "#E64B35"),
                      labels = c("Cross-platform", "Same-platform"),
                      name = NULL) +
  facet_grid(polarity ~ pair, scales = "free") +
  labs(title = "Inter-lab relative abundance (log2 AUC / lab median)",
       subtitle = "Consensus peaks present in all 4 labs",
       x = "Lab 1 (log2 AUC / lab median)",
       y = "Lab 2 (log2 AUC / lab median)") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Note on EIC profile correlation

We do **not** compute inter-lab EIC profile correlation (e.g.
apex-aligned Pearson on the full chromatographic shape) in this
document. Earlier exploration showed that profile correlation between
matched consensus peaks is high in absolute terms, but it is **equally
high against random peaks** drawn from the same sample (AUROC ≈ 0.5
against an RT-gap-controlled null). A score that does not separate
matched from non-matched peaks cannot be used for correspondence,
quality assessment, or any downstream filter — so the metric is
omitted here.

# Next Steps

1. Peak metrics saved to `object/peak_metrics_detected_all4labs.csv`.
2. Filtered EICs saved to `object/eics_filtered_detected_all4labs.RData`.
3. Post-filter consensus Spectra saved to `object/sp_consensus.RData`.
4. Post-filter consensus TIC saved to `object/tic_consensus.RData`.
3. Downstream variance and normalisation analysis continues in
   `4_mitigating_inter_lab_variance.qmd`.
