---
title: "RT alignment: per-lab NAPS spline (Phase 2)"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
        embed-resources: true
---

# Goal

Build a per-`(lab, polarity)` **monotonic spline** that maps each
lab's local apex RT onto a cross-lab consensus RT in seconds, using
the NAPS anchor ladder. The output is a reusable RT → aligned-RT
mapping consumed downstream:

- **P3** uses it to bin RT for RT-local normalization factors.
- **P5** uses it to compute cross-lab RT concordance for the
  confidence tier.

Per the project plan: **per-lab local alignment, not a single global
RTI map**.

The phase deliverables are:

1. Within-lab NAPS precision (P2.2) — the precision floor against
   which the post-alignment cross-lab residual is benchmarked.
2. Per-lab monotonic spline (P2.3) and optional standards-as-anchor
   extension (P2.4).
3. Cross-lab apex-RT residual distribution **before vs after**
   alignment (P2.5), compared to the within-lab floor.
4. Early-RT region caveat (P2.6) — NAPS anchor spacing is sparse for
   C1–C5, so a wider residual band is expected there.
5. Reusable mapping (P2.7) — `object/rt_alignment_table.csv` for
   linear interpolation and `object/rt_alignment_splines.RData` for
   direct spline application.

This file also runs the **annotated-EIC pipeline** on the SIRIUS
outputs: §P2.9 runs MS1 anchoring (Stage 1 candidates + Stage 2
similarity-rule consensus + Case-B isobaric resolution); §P2.10 runs
Stage 3 — inverse-aligns the per-(mixture, compound) consensus RT to
each lab's local time and extracts an SNR-picked EIC per compound,
writing `object/annotation_chrom_metrics.csv` for P3 onward.

## Setup

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(purrr)

# Force SERIAL BiocParallel: the installed mzR is built against a
# different Rcpp than is loaded, and parallel workers reading mzML
# concurrently intermittently segfault. Serial is slower but reliable.
BiocParallel::register(BiocParallel::SerialParam())

labs           <- c("afekta", "cembio", "hmgu", "icl")
producer_dir   <- here("5_downstream_analysis", "object")
naps_xcms_path <- function(lab) here("1_preprocessing", lab, "HE", "naps.csv")
naps_ce_path   <- here("5_downstream_analysis", "object",
                       "naps_chrom_metrics.csv")
out_dir        <- here("5_downstream_analysis", "object")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# NAPS source: "chromextract" reads naps_chrom_metrics.csv from
# 2_naps_extraction.qmd (preferred — recovers icl C1–C3 and replaces
# cembio-pos outliers). "xcms" reads the per-lab naps.csv (legacy path,
# kept for side-by-side comparison).
naps_source <- "chromextract"

# When using chromextract, "own" keeps only seed=own rows (xcms had a
# value); "all" includes consensus-propagated rows too.
naps_chromextract_use <- "all"   # "all" or "own"

# Standards-as-anchor selection.
#
# Two modes:
# - "raw_sd": qualify a compound when ≥N labs annotate it and raw cross-lab
#   apex-RT SD ≤ standards_anchor_max_raw_sd.
# - "iterative": fit splines from NAPS only, align candidates through them,
#   qualify by *aligned* SD ≤ standards_anchor_max_aligned_sd. Picks up
#   many more compounds in the metabolite-dense region by accounting for
#   the systematic per-lab shift before judging cross-lab agreement.
standards_anchor_mode            <- "iterative"
standards_anchor_min_labs        <- 3
standards_anchor_max_raw_sd      <- 5   # raw_sd mode threshold (s)
# Threshold sweep (P13.1, all values from rendered production runs):
#   threshold  n_std  median_aligned_sd  p95_aligned_sd  pos_main_sd  pos_A_tier
#       3 s      2          7.80 s         23.80 s         9.71 s        21
#       5 s      3          7.75 s         26.09 s        11.73 s        21
#       7 s      5          7.15 s         23.35 s         9.87 s        27   ← chosen
# 7 s is the sweet spot. At 5 s, two standards holding the pos-main
# region spline together drop out → pos-main aligned SD jumps to
# 11.73 s (worse than baseline 9.71) and all 6 A-tier promotions
# revert. At 7 s, 5 standards qualify, pos-early 6.49 → 5.58 s, +6
# A-tier promotions (mostly cembio pos). Beyond 10 s the picture
# degraded again (low-quality standards dominate).
standards_anchor_max_aligned_sd  <- 7   # iterative mode threshold (s)
```

# P2.1 Load NAPS data

Long form: one row per `(lab, polarity, anchor_n, rep, rt_sec)`. RTI =
N × 100 (deterministic ladder); seconds is the working unit because
the alignment target is in seconds.

Two sources, switched by `naps_source` in setup:

- `"chromextract"` (default) reads `naps_chrom_metrics.csv` from
  `2_naps_extraction.qmd`. Recovers anchors xcms missed (icl C1–C3)
  and replaces the wrong-peak outliers in cembio-pos. Carries a
  `source` column (`own` / `consensus`) that survives downstream so
  you can filter on it.
- `"xcms"` reads the legacy `1_preprocessing/{lab}/HE/naps.csv` files
  (xcms-derived). Useful for direct before-vs-after-chromExtract
  comparison.

```{r load-naps}
read_naps_lab <- function(lab) {
  df <- read.csv(naps_xcms_path(lab), stringsAsFactors = FALSE,
                 check.names = FALSE)
  df$anchor_n <- suppressWarnings(
    as.integer(gsub("^C(\\d+).*", "\\1", df$Name)))
  pos_cols <- grep("^NAPS_POS_", names(df), value = TRUE)
  neg_cols <- grep("^NAPS_NEG_", names(df), value = TRUE)
  bind_rows(
    df |> select(anchor_n, RTI, all_of(pos_cols)) |>
      pivot_longer(all_of(pos_cols), names_to = "rep_col",
                   values_to = "rt_sec") |>
      mutate(polarity = "pos",
             rep = as.integer(gsub("^NAPS_POS_", "", rep_col))),
    df |> select(anchor_n, RTI, all_of(neg_cols)) |>
      pivot_longer(all_of(neg_cols), names_to = "rep_col",
                   values_to = "rt_sec") |>
      mutate(polarity = "neg",
             rep = as.integer(gsub("^NAPS_NEG_", "", rep_col)))
  ) |>
    select(anchor_n, RTI, polarity, rep, rt_sec) |>
    mutate(lab = lab, source = "own", .before = 1)
}

read_naps_chromextract <- function() {
  # The NA-FWHM peak-integrity gate now lives in the producer
  # (2_naps_extraction.qmd §9), so the CSV here is already clean.
  df <- read.csv(naps_ce_path, stringsAsFactors = FALSE)
  if (naps_chromextract_use == "own") {
    df <- df |> filter(source == "own")
  }
  df |>
    mutate(RTI = anchor_n * 100L) |>
    transmute(lab, source, anchor_n, RTI, polarity,
              rep = replicate, rt_sec = apex_rt)
}

naps_long <- if (naps_source == "chromextract" && file.exists(naps_ce_path)) {
  cat("NAPS source: chromextract  (", naps_ce_path, ")\n")
  read_naps_chromextract()
} else {
  if (naps_source == "chromextract") {
    warning("naps_chrom_metrics.csv not found; falling back to xcms naps.csv.",
            call. = FALSE)
  }
  cat("NAPS source: xcms (per-lab naps.csv)\n")
  bind_rows(lapply(labs, read_naps_lab))
}
naps_long <- naps_long |>
  filter(!is.na(rt_sec), !is.na(anchor_n))

cat("NAPS rows (lab × polarity × anchor × rep):", nrow(naps_long),
    " | per-lab non-NA counts:\n")
naps_long |> count(lab, polarity) |>
  tidyr::pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable()
```

# P2.2 Within-lab NAPS precision (the floor)

Per `(lab, polarity, anchor_n)`, SD of `rt_sec` across the 6 replicates
is the per-anchor within-lab precision. The median across anchors is
the lab's typical precision. This is the fair benchmark — cross-lab
residual after alignment should approach (not surpass) this floor.

```{r within-lab-sd}
within_lab_sd <- naps_long |>
  group_by(lab, polarity, anchor_n) |>
  summarise(n_rep   = n(),
            rt_mean = mean(rt_sec, na.rm = TRUE),
            rt_sd   = sd(rt_sec, na.rm = TRUE),
            .groups = "drop") |>
  filter(n_rep >= 2L, !is.na(rt_sd))

within_lab_sd |>
  group_by(lab, polarity) |>
  summarise(n_anchors = n(),
            median_sd = round(median(rt_sd, na.rm = TRUE), 3),
            mean_sd   = round(mean(rt_sd,   na.rm = TRUE), 3),
            p95_sd    = round(quantile(rt_sd, 0.95, na.rm = TRUE), 3),
            .groups = "drop") |>
  knitr::kable(caption = "Within-lab NAPS replicate SD (seconds).")
```

```{r within-lab-sd-plot}
#| fig-height: 4
#| fig-cap: "Per-anchor within-lab replicate SD (seconds). Median per (lab, polarity) = the precision floor."
# Drop SD == 0 (single-replicate anchors): undefined on a log scale.
ggplot(within_lab_sd |> filter(rt_sd > 0), aes(x = lab, y = rt_sd, fill = polarity)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(aes(group = polarity),
              position = position_jitterdodge(jitter.width = 0.1,
                                              dodge.width = 0.8),
              alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  labs(x = NULL, y = "Within-lab replicate SD (s, log scale)",
       fill = "Polarity",
       title = "Within-lab NAPS replicate precision") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

# P2.3 Build cross-lab consensus RT

**What "consensus RT" means here.** For each NAPS standard (C1-APS,
C2-APS, …) we first collapse each lab's 6 NAPS replicates to its **per-lab
median apex RT** at that anchor. The **consensus RT** is then the **median
across labs** of those per-lab medians, computed per `(polarity,
anchor_n)`. So an anchor's consensus is the central RT that the four labs
collectively put it at — the cross-lab "where does this standard elute"
target. Each lab is later compared against this consensus to measure
its drift. Anchors visible in fewer than `min_labs_for_consensus` labs
are dropped (they have no defensible consensus).

```{r consensus-rt}
min_labs_for_consensus <- 2

per_lab_median <- naps_long |>
  group_by(lab, polarity, anchor_n) |>
  summarise(rt_med = median(rt_sec, na.rm = TRUE),
            n_rep  = sum(!is.na(rt_sec)),
            .groups = "drop") |>
  filter(n_rep >= 1L)

consensus_rt <- per_lab_median |>
  group_by(polarity, anchor_n) |>
  summarise(consensus_rt = median(rt_med, na.rm = TRUE),
            n_labs       = n_distinct(lab),
            .groups = "drop") |>
  filter(n_labs >= min_labs_for_consensus)

cat("Consensus anchors per polarity:\n")
consensus_rt |> count(polarity) |>
  knitr::kable(col.names = c("polarity", "n_anchors"))
```

```{r consensus-plot}
#| fig-height: 4
#| fig-cap: "Each point = one (lab, NAPS anchor) — y is that lab's median apex RT (over 6 replicates), x is the cross-lab median for the same anchor (the consensus). On the dashed diagonal = the lab elutes that anchor exactly where the consortium does; systematic offset above/below the line = per-lab gradient drift. Per-(lab, polarity) Hyman monotonic splines (§P2.4) map y → x to remove that drift."
per_lab_median |>
  inner_join(consensus_rt, by = c("polarity", "anchor_n")) |>
  ggplot(aes(x = consensus_rt, y = rt_med, colour = lab)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50") +
  geom_point(size = 2, alpha = 0.85) +
  facet_wrap(~ polarity, scales = "free") +
  scale_color_hue() +
  labs(x = "Consensus RT (s)", y = "Lab-local median RT (s)",
       colour = "Lab", title = "Per-lab NAPS RT vs consensus") +
  theme_bw() +
  theme(legend.position = "top")
```

# P2.4 Per-lab monotonic spline alignment

For each `(lab, polarity)` we fit `splinefun(method = "hyman")`. Hyman
guarantees monotonicity, so the alignment never inverts the RT order —
a real LC peak that elutes before another in the lab cannot end up
elution-reordered after alignment.

**Why this model.** Every simpler model is ruled out by what we know
about the data:

- A **single shift** (`rt + b`) assumes uniform drift across the
  gradient. NAPS shows per-lab offsets ~0.5 s at C1 (RT 25 s) but
  ~5 s at C20 (RT 560 s) — drift grows with RT.
- A **linear fit** (`a · rt + b`) handles proportional drift but the
  consensus_rt vs rt_local plot has clear curvature in the [30, 60] s
  region — early-gradient solvation physics differs from
  late-gradient.
- A **polynomial** fits curvature but oscillates at the ends with
  only ~20 anchors (Runge's phenomenon) and can't accommodate
  region-specific drift.
- **Piecewise linear** has kinks at every anchor; downstream the
  LOESS RT-normalization surface ([`5_normalization.qmd`](5_normalization.qmd))
  propagates those kinks as artifacts.
- An **unconstrained cubic spline** (or LOWESS) is smooth but can be
  non-monotonic. Non-monotonic RT alignment would scramble peak
  order, breaking cross-lab feature matching and the decile binning
  in P3.
- **Hyman** is the cheapest model that simultaneously gives
  per-region nonlinear fit, smoothness (continuous first derivative),
  and a hard monotonicity guarantee. With slope-limited cubic
  segments it's stable with the ~20 anchors we have.

**Why per-lab** rather than one global function: each LC system has
its own dwell volume / gradient delay / column equilibration. The
drift profiles are different *shapes*, not just different magnitudes,
so a global function would wash out the corrections we want to make.

**Why a cross-lab consensus target** rather than a reference lab:
picking one lab as the anchor target silently exports its drift to
everyone else. The cross-lab median is robust — one bad anchor
contributes ≤ 25 % of the target.

**Why the PAV pre-step**: the rare cases where a single anchor's
local rank disagrees with its consensus rank (we saw 2 such anchors
in icl_neg C2/C3) would otherwise break Hyman's monotonic input
contract. `isoreg` collapses these to their pooled value before the
spline sees them.

**Why not obiwarp / DTW / COW**: those are designed for un-anchored
non-targeted alignment where you discover correspondences from full
MS1 chromatograms. We have ~20 known-identity NAPS anchors with
sub-second within-lab replicate SD — anchor-based alignment is
strictly better-conditioned for this data structure.

Two paths to qualify spiked standards as anchors (toggle via
`standards_anchor_mode` in setup):

- **`raw_sd`** — a compound is added when annotated in ≥
  `standards_anchor_min_labs` labs **and** raw cross-lab apex-RT SD ≤
  `standards_anchor_max_raw_sd`. Strict on raw seconds — usually only
  a handful of compounds qualify.
- **`iterative`** (default) — round 1 fits splines from NAPS only;
  round 2 aligns every candidate's apex_rt through the round-1 spline
  and qualifies on the **aligned** SD ≤
  `standards_anchor_max_aligned_sd`. Removes the systematic per-lab
  offset before judging agreement, so many more compounds make it in
  — especially in the metabolite-dense early-RT region where APS
  spacing is sparse.

```{r load-annotation}
identity_df <- read.csv(file.path(producer_dir,
                                  "annotation_identity.csv"),
                        stringsAsFactors = FALSE,
                        colClasses = c(mixture = "character"))
identity_df$mixture <- gsub("\\.", "_", identity_df$mixture)
chrom_df    <- read.csv(file.path(producer_dir,
                                  "annotation_chrom_metrics.csv"),
                        stringsAsFactors = FALSE,
                        colClasses = c(mixture = "character"))
chrom_df$mixture <- gsub("\\.", "_", chrom_df$mixture)

# Per (mixture, lab, compound, polarity), dominant peak.
chrom_dom <- chrom_df |>
  filter(!is.na(apex_rt), !is.na(maxIntensity)) |>
  group_by(mixture, lab, compound, polarity) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup()

# Keep only compounds with a SIRIUS identity in that (lab, polarity).
chrom_with_id <- chrom_dom |>
  semi_join(identity_df |>
              distinct(mixture, lab, compound, polarity),
            by = c("mixture", "lab", "compound", "polarity"))
```

```{r build-naps-anchor-table}
naps_anchor_long <- per_lab_median |>
  inner_join(consensus_rt |> select(polarity, anchor_n, consensus_rt),
             by = c("polarity", "anchor_n")) |>
  transmute(lab, polarity,
            rt_local     = rt_med,
            consensus_rt = consensus_rt,
            anchor_label = paste0("C", anchor_n, "-APS"))
```

```{r fit-one-spline}
# Fit monotonic spline per (lab, polarity).
#
# splinefun(method = "hyman") requires the input (rt_local, consensus_rt)
# to be monotonic in BOTH x and y. Standards anchors can violate this
# (a compound's local apex RT and the cross-lab consensus disagree on
# the rank order). Steps:
#   1. dedupe x (median y per unique rt_local)
#   2. enforce monotonic y via isoreg (PAV — drops y inversions)
#   3. dedupe y (collapse flat runs by averaging their x)
#   4. fit hyman → monoH.FC → linear approxfun as triple-tier fallback
fit_one_spline <- function(df) {
  df <- df |> filter(is.finite(rt_local), is.finite(consensus_rt))
  if (nrow(df) < 4L) return(NULL)
  df <- df |>
    group_by(rt_local) |>
    summarise(consensus_rt = median(consensus_rt), .groups = "drop") |>
    arrange(rt_local)
  # PAV monotonicity. isoreg returns a non-decreasing fitted vector `yf`.
  df$consensus_rt <- isoreg(df$rt_local, df$consensus_rt)$yf
  df <- df |>
    group_by(consensus_rt) |>
    summarise(rt_local = mean(rt_local), .groups = "drop") |>
    arrange(rt_local)
  if (nrow(df) < 4L) return(NULL)
  tryCatch(
    splinefun(df$rt_local, df$consensus_rt, method = "hyman"),
    error = function(e1) {
      message("  hyman failed: ", conditionMessage(e1),
              " — falling back to monoH.FC.")
      tryCatch(
        splinefun(df$rt_local, df$consensus_rt, method = "monoH.FC"),
        error = function(e2) {
          message("  monoH.FC failed; using linear approxfun.")
          approxfun(df$rt_local, df$consensus_rt,
                    method = "linear", rule = 2)
        })
    })
}

# Fit a `splines` list keyed by "lab_polarity" from any anchor table.
fit_splines_for <- function(table) {
  grps        <- table |> group_by(lab, polarity) |> group_split()
  names(grps) <- vapply(grps,
    function(d) paste(d$lab[1], d$polarity[1], sep = "_"),
    character(1))
  lapply(grps, fit_one_spline)
}

# Apply a `splines` map to a data frame's apex_rt, clamped to per-key
# anchor range. Used twice: once for round-1 candidate qualification,
# once for the final P2.5 alignment.
apply_splines <- function(df, splines_map, anchors_for_range) {
  df$key        <- paste(df$lab, df$polarity, sep = "_")
  df$aligned_rt <- NA_real_
  for (key_i in unique(df$key)) {
    f <- splines_map[[key_i]]
    if (is.null(f)) next
    anchor_range <- anchors_for_range |>
      filter(paste(lab, polarity, sep = "_") == key_i) |>
      pull(rt_local) |> range(na.rm = TRUE)
    idx       <- which(df$key == key_i)
    rt_lab    <- df$apex_rt[idx]
    in_range  <- !is.na(rt_lab) &
                 rt_lab >= anchor_range[1] &
                 rt_lab <= anchor_range[2]
    df$aligned_rt[idx[in_range]] <- as.numeric(f(rt_lab[in_range]))
  }
  df$key <- NULL
  df
}
```

```{r qualify-standards}
if (standards_anchor_mode == "raw_sd") {
  qualifying <- chrom_with_id |>
    group_by(mixture, compound, polarity) |>
    summarise(n_labs = n_distinct(lab),
              rt_sd  = sd(apex_rt, na.rm = TRUE),
              .groups = "drop") |>
    filter(n_labs >= standards_anchor_min_labs,
           !is.na(rt_sd),
           rt_sd <= standards_anchor_max_raw_sd) |>
    select(mixture, compound, polarity)

  cat("Standards qualifying (raw_sd, ≥", standards_anchor_min_labs,
      " labs, raw SD ≤ ", standards_anchor_max_raw_sd, " s): ",
      nrow(qualifying), "\n", sep = "")

} else if (standards_anchor_mode == "iterative") {
  splines_r1 <- fit_splines_for(naps_anchor_long)

  cand <- chrom_with_id |>
    group_by(mixture, compound, polarity) |>
    filter(n_distinct(lab) >= standards_anchor_min_labs) |>
    ungroup() |>
    apply_splines(splines_r1, naps_anchor_long) |>
    rename(aligned_rt_r1 = aligned_rt)

  qualifying <- cand |>
    filter(!is.na(aligned_rt_r1)) |>
    group_by(mixture, compound, polarity) |>
    summarise(n_labs_aligned = n(),
              aligned_sd     = sd(aligned_rt_r1),
              .groups = "drop") |>
    filter(n_labs_aligned >= standards_anchor_min_labs,
           !is.na(aligned_sd),
           aligned_sd <= standards_anchor_max_aligned_sd) |>
    select(mixture, compound, polarity)

  cat("Standards qualifying (iterative, ≥", standards_anchor_min_labs,
      " labs, aligned SD ≤ ", standards_anchor_max_aligned_sd, " s): ",
      nrow(qualifying), "\n", sep = "")
} else {
  stop("Unknown standards_anchor_mode: ", standards_anchor_mode)
}

standards_anchor_long <- chrom_with_id |>
  inner_join(qualifying,
             by = c("mixture", "compound", "polarity")) |>
  group_by(mixture, compound, polarity) |>
  mutate(consensus_rt = median(apex_rt, na.rm = TRUE)) |>
  ungroup() |>
  select(lab, polarity,
         rt_local     = apex_rt,
         consensus_rt,
         anchor_label = compound)
```

```{r build-splines}
# Toggle to compare NAPS-only vs NAPS+standards downstream.
# NAPS-only (FALSE) avoids circularity: the same SIRIUS spike-in panel
# would otherwise be used both as anchors AND as the compounds we score
# cross-lab agreement on. NAPS-only also avoids the L-phe [M+Na]+ adduct
# leakage where the dominant-peak (slice_max) at the cembio L-phe pos row
# picks a 43 s isobar over the real 58 s peak (see §P2.9.2).
use_standards_anchors <- FALSE
anchor_table <- if (use_standards_anchors) {
  bind_rows(naps_anchor_long, standards_anchor_long)
} else {
  naps_anchor_long
}

cat("Final anchor table — rows per (lab, polarity):\n")
anchor_table |> count(lab, polarity) |>
  tidyr::pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable()

splines <- fit_splines_for(anchor_table)

cat("Splines fitted (lab_polarity):\n  ",
    paste(names(splines), collapse = ", "), "\n")
```

```{r spline-curves-plot}
#| fig-height: 6
#| fig-cap: "Per-lab spline alignment curves. x = local RT (s); y = consensus RT (s). The diagonal = identity (no alignment needed)."
plot_grid <- bind_rows(lapply(names(splines), function(key) {
  parts <- strsplit(key, "_")[[1]]
  lab_i <- parts[1]; pol_i <- parts[2]
  rt_local_range <- anchor_table |>
    filter(lab == lab_i, polarity == pol_i) |>
    pull(rt_local) |> range(na.rm = TRUE)
  if (any(!is.finite(rt_local_range))) return(NULL)
  rt_seq <- seq(rt_local_range[1], rt_local_range[2], length.out = 400)
  data.frame(lab = lab_i, polarity = pol_i,
             rt_local = rt_seq,
             rt_aligned = splines[[key]](rt_seq))
}))

ggplot(plot_grid, aes(x = rt_local, y = rt_aligned, colour = lab)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.9) +
  geom_point(data = anchor_table,
             aes(x = rt_local, y = consensus_rt, colour = lab),
             size = 1.5, alpha = 0.6, inherit.aes = FALSE) +
  facet_wrap(~ polarity, scales = "free") +
  scale_color_hue() +
  labs(x = "Local apex RT (s)", y = "Consensus / aligned RT (s)",
       colour = "Lab",
       title = "Per-lab monotonic spline alignment",
       subtitle = "Lines = spline; points = anchors (NAPS + qualifying standards)") +
  theme_bw() +
  theme(legend.position = "top")
```

# P2.5 Cross-lab apex-RT residual: before vs after alignment

For each `(mixture × compound × polarity)` where at least 2 labs detected
the compound, take the labs' apex_rt values for that compound and compute
their **standard deviation** — this is "how much the labs disagree on
where this compound elutes". SD = 0 = perfect agreement; SD = 10 s = a
10-second spread of estimates across the detecting labs.

Each compound gives **two** SD values:

- **Before** = SD of the raw `apex_rt` values (each lab's local clock,
  no alignment).
- **After** = SD of the same compound's `aligned_rt` values (each lab's
  RT first passed through that lab's NAPS spline, putting all labs on a
  common consensus clock).

Reference: the **within-lab floor** (red lines below) is the median SD
we get even when *one* lab measures the same NAPS standard 6 times. It
is the noise of the chromatography itself — the best cross-lab agreement
that is theoretically achievable. After alignment, cross-lab SD should
approach this floor.

```{r apply-alignment}
apply_alignment <- function(df) {
  # Apply final (NAPS + qualified standards) splines, clamped to the
  # per-(lab, polarity) anchor range. Out-of-range rows → aligned_rt = NA
  # so they don't poison cross-lab residual stats downstream.
  apply_splines(df, splines, anchor_table)
}

chrom_aligned <- chrom_dom |>
  apply_alignment()

residual_df <- chrom_aligned |>
  filter(!is.na(apex_rt), !is.na(aligned_rt)) |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs        = n_distinct(lab),
            sd_raw        = sd(apex_rt,    na.rm = TRUE),
            sd_aligned    = sd(aligned_rt, na.rm = TRUE),
            range_raw     = max(apex_rt)     - min(apex_rt),
            range_aligned = max(aligned_rt) - min(aligned_rt),
            .groups = "drop") |>
  filter(n_labs >= 2L, !is.na(sd_raw), !is.na(sd_aligned))

residual_df |>
  summarise(n = n(),
            median_sd_raw     = round(median(sd_raw),     2),
            median_sd_aligned = round(median(sd_aligned), 2),
            p95_sd_raw        = round(quantile(sd_raw,     0.95), 2),
            p95_sd_aligned    = round(quantile(sd_aligned, 0.95), 2)) |>
  knitr::kable(caption = "Cross-lab apex-RT SD (seconds) before vs after alignment.")
```

**Plot 1 — histogram of cross-lab SDs.** *x-axis (log s)*: SD value.
*y-axis*: number of compounds at that SD. *Grey bars* = the *Before*
distribution (cross-lab SD on raw RTs). *Blue bars* = the *After*
distribution (same compounds, cross-lab SD on aligned RTs). *Red dashed
line* = the within-lab floor. Two facets = negative / positive polarity.
**How to read it**: alignment is working when the blue distribution sits
clearly to the left of the grey one (compounds shifted to smaller SD).
The closer the blue mass gets to the red line, the closer cross-lab
agreement is to the theoretical floor.

```{r residual-vs-floor}
#| fig-height: 5
#| fig-cap: "Distribution of cross-lab apex-RT SD before (grey) vs after (blue) per-lab spline alignment. Red line = median within-lab NAPS floor."
floor_sd <- median(within_lab_sd$rt_sd, na.rm = TRUE)

residual_long <- residual_df |>
  select(mixture, compound, polarity, sd_raw, sd_aligned) |>
  pivot_longer(c(sd_raw, sd_aligned), names_to = "stage",
               values_to = "sd_s") |>
  mutate(stage = recode(stage, sd_raw = "Before", sd_aligned = "After"),
         stage = factor(stage, levels = c("Before", "After")))

# Drop SD == 0 (perfect cross-lab agreement): undefined on a log scale.
ggplot(residual_long |> filter(sd_s > 0), aes(x = sd_s, fill = stage)) +
  geom_histogram(bins = 60, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = floor_sd, colour = "red", linetype = "dashed") +
  annotate("text", x = floor_sd, y = Inf,
           label = sprintf("within-lab floor = %.2f s", floor_sd),
           hjust = -0.05, vjust = 1.4, colour = "red", size = 3) +
  scale_fill_manual(values = c(Before = "grey60", After = "#4DBBD5")) +
  scale_x_log10(labels = scales::label_number(accuracy = 0.1)) +
  facet_wrap(~ polarity) +
  labs(x = "Cross-lab apex-RT SD (s, log)", y = "Count",
       fill = NULL,
       title = "Cross-lab RT residual before vs after per-lab alignment") +
  theme_bw() +
  theme(legend.position = "top")
```

**Plot 2 — same SDs, but plotted *against* where the compound elutes.**
Each point = one compound. *x-axis*: where on the consensus gradient that
compound elutes (its consensus RT). *y-axis (log s)*: the same cross-lab
SD value as in Plot 1. *Grey points and grey smoothed line* = Before
alignment. *Blue points and blue smoothed line* = After alignment. *Red
dashed line* = within-lab floor. **How to read it**: where the blue line
drops below the grey line, alignment helped at that RT region. Where the
blue line stays high or rises (typically the early-RT polar region), the
spline can't fix it because there are no NAPS anchors there (the §P2.6
caveat).

```{r residual-by-rt-region}
#| fig-height: 4
#| fig-cap: "Cross-lab SD vs consensus apex RT, before (grey) and after (blue). x-axis = where the peak elutes on the consensus axis. Wider band at early RT reflects the NAPS-sparse C1–C5 region (see P2.6)."
chrom_aligned |>
  filter(!is.na(aligned_rt)) |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs        = n_distinct(lab),
            rt_consensus  = median(aligned_rt, na.rm = TRUE),
            sd_raw        = sd(apex_rt,    na.rm = TRUE),
            sd_aligned    = sd(aligned_rt, na.rm = TRUE),
            .groups = "drop") |>
  filter(n_labs >= 2L) |>
  pivot_longer(c(sd_raw, sd_aligned), names_to = "stage", values_to = "sd_s") |>
  mutate(stage = recode(stage, sd_raw = "Before", sd_aligned = "After"),
         stage = factor(stage, levels = c("Before", "After"))) |>
  filter(sd_s > 0) |>   # SD == 0 undefined on the log y-axis
  ggplot(aes(x = rt_consensus, y = sd_s, colour = stage)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linewidth = 0.8) +
  geom_hline(yintercept = floor_sd, colour = "red", linetype = "dashed") +
  scale_colour_manual(values = c(Before = "grey50", After = "#4DBBD5")) +
  scale_y_log10() +
  facet_wrap(~ polarity) +
  labs(x = "Consensus apex RT (s)", y = "Cross-lab SD (s, log)",
       colour = NULL,
       title = "Residual SD vs consensus RT") +
  theme_bw() +
  theme(legend.position = "top")
```

# P2.6 Early-RT region caveat

The APS ladder spaces anchors by ~1 unit of carbon count = ~1 s in
the early (most polar) region. C1 (~27 s) and C2 (~28 s) are nearly
co-eluting; C2 → C3 → C4 spans ~6 s on most labs. **A 1-second RT
delta in this region equals ~100 RTI units.** Metabolites cluster in
this 23–45 s window, so:

- Spline slopes are steep but the per-anchor SD is comparable to
  later anchors in absolute seconds → the spline can over-correct or
  under-correct local distortion.
- Standards-as-anchors (P2.4) could in principle help here, but we
  switched to **NAPS-only** because: (a) spike-in standards are
  themselves the compounds we score cross-lab agreement on (using them
  as anchors is circular), and (b) the dominant-peak per-(lab, compound)
  selection can land the anchor on an isobar instead of the real peak
  (the cembio L-phe [M+Na]+ case, §P2.9.2). NAPS-only is honest at the
  cost of leaving the early-RT region under-anchored.
- We tested **scam** (monotone-constrained GAM) as an alternative
  interpolant in this gap. scam gives a gentler slope through the gap
  than Hyman, but **neither matches the truth**:
  the §P2.5 outlier compounds elute ~10 s later in cembio pos than in
  the other labs, but no spline through (rt_local 43, consensus 48)
  and (rt_local 130, consensus 127) can put rt_local 65 at consensus
  55 — the gap is structural, not algorithmic. Denser polar NAPS
  standards in C1–C5 is the only real fix.

Practically: report alignment uncertainty separately for the
< 60 s region. Expect the post-alignment residual band to remain
wider here even after alignment helps in absolute terms.

```{r early-rt-residual}
chrom_aligned |>
  filter(!is.na(aligned_rt)) |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs        = n_distinct(lab),
            rt_consensus  = median(aligned_rt, na.rm = TRUE),
            sd_raw        = sd(apex_rt,    na.rm = TRUE),
            sd_aligned    = sd(aligned_rt, na.rm = TRUE),
            .groups = "drop") |>
  filter(n_labs >= 2L) |>
  mutate(rt_region = ifelse(rt_consensus < 60, "early (< 60 s)",
                            "main (≥ 60 s)")) |>
  group_by(polarity, rt_region) |>
  summarise(n                 = n(),
            median_sd_raw     = round(median(sd_raw), 2),
            median_sd_aligned = round(median(sd_aligned), 2),
            p95_sd_aligned    = round(quantile(sd_aligned, 0.95), 2),
            .groups = "drop") |>
  knitr::kable(caption = "Residual SD split by RT region.")
```

### Per-lab NAPS coverage

Which NAPS anchors did each lab actually detect? Anchors a lab is
missing have **no spline knot for that region**, so peaks in that region
are aligned by extrapolation from neighbouring knots (or fall outside
the spline's anchor range entirely and get `aligned_rt = NA`). A lab
missing C1–C5 has **no anchor at all** in the polar region; early-RT
peaks for that lab are functionally un-aligned.

```{r naps-coverage}
#| message: false
naps_ce <- read.csv(here::here("5_downstream_analysis", "object",
                               "naps_chrom_metrics.csv"),
                    stringsAsFactors = FALSE)
all_anchors <- sort(unique(naps_ce$anchor_n))
naps_cov <- expand.grid(lab = c("afekta", "cembio", "hmgu", "icl"),
                        polarity = c("pos", "neg"),
                        anchor_n = all_anchors,
                        stringsAsFactors = FALSE) |>
  dplyr::left_join(naps_ce |>
                     dplyr::group_by(lab, polarity, anchor_n) |>
                     dplyr::summarise(n_rep = dplyr::n(), .groups = "drop"),
                   by = c("lab", "polarity", "anchor_n")) |>
  dplyr::mutate(present = ifelse(is.na(n_rep), "·", "✓"))

naps_cov |>
  dplyr::mutate(lp = paste(lab, polarity, sep = " ")) |>
  tidyr::pivot_wider(id_cols = lp, names_from = anchor_n,
                     values_from = present, values_fill = "·") |>
  knitr::kable(caption = "NAPS anchor detection per (lab, polarity). ✓ = detected; · = MISSING. Missing anchors leave the spline unanchored in that region.")

cat("\nMissing-anchor summary:\n")
naps_cov |>
  dplyr::filter(present == "·") |>
  dplyr::group_by(lab, polarity) |>
  dplyr::summarise(missing_n = paste(sort(anchor_n), collapse = ","),
                   .groups = "drop") |>
  dplyr::transmute(lp = paste(lab, polarity), missing_anchors = missing_n) |>
  knitr::kable(caption = "Labs × polarities with at least one missing NAPS anchor.")
```

**What this means in practice.** All four labs share the natural C4→C5
gap of the NAPS panel design (no carbon-count-between-4-and-5 standard
exists), so every lab's spline has the same wide unanchored region in
the metabolite-dense early-RT zone — the §P2.5 outlier story doesn't go
away by anchor density alone. **But** `icl neg` is missing the entire
C1–C5 block, meaning icl-neg's spline has no anchor knots below ~190 s.
icl-neg early-RT peaks fall outside the spline's anchor range and are
treated as un-alignable (`aligned_rt = NA`) — they're dropped under
Strict and use the global fallback under Inclusive. That's a meaningful
caveat for icl's negative-mode reproducibility numbers.

# P2.7 Export reusable RT → aligned mapping

Two outputs at `5_downstream_analysis/object/`:

- `rt_alignment_table.csv` — sampled lookup of `(lab, polarity,
  rt_local, rt_aligned)` so downstream R / Python code can apply via
  linear interpolation without needing the splines.
- `rt_alignment_splines.RData` — the `splines` list of `splinefun`
  objects, for direct use from R sessions that load this file.

P3 and P5 should source `rt_alignment_table.csv` (vendor-neutral) by
default and fall back to the RData for sub-second precision when
needed.

```{r export-mapping}
sample_grid <- bind_rows(lapply(names(splines), function(key) {
  parts <- strsplit(key, "_")[[1]]
  lab_i <- parts[1]; pol_i <- parts[2]
  rt_range <- anchor_table |>
    filter(lab == lab_i, polarity == pol_i) |>
    pull(rt_local) |> range(na.rm = TRUE)
  if (any(!is.finite(rt_range))) return(NULL)
  rt_seq <- seq(rt_range[1], rt_range[2], length.out = 1000)
  data.frame(lab        = lab_i,
             polarity   = pol_i,
             rt_local   = rt_seq,
             rt_aligned = splines[[key]](rt_seq))
}))

write.csv(sample_grid,
          file.path(out_dir, "rt_alignment_table.csv"),
          row.names = FALSE)

# Also write the per-(lab, polarity, anchor) input table so P5 can
# trace which anchors drove the fit.
write.csv(anchor_table,
          file.path(out_dir, "rt_alignment_anchors.csv"),
          row.names = FALSE)

save(splines, file = file.path(out_dir, "rt_alignment_splines.RData"))

cat("Exports written:\n",
    " ", file.path(out_dir, "rt_alignment_table.csv"),    "(", nrow(sample_grid), " rows )\n",
    " ", file.path(out_dir, "rt_alignment_anchors.csv"),  "(", nrow(anchor_table), " rows )\n",
    " ", file.path(out_dir, "rt_alignment_splines.RData"), "\n", sep = "")
```

# P2.8 RTI reproducibility across labs

The Hyman spline above maps each lab's local apex RT to the cross-lab
**consensus RT in seconds**. The classical alternative is the
**retention time index (RTI)**: assign each NAPS anchor a fixed
reference index (C1 = 100, C2 = 200, ..., C20 = 2000) and convert any
compound's `apex_rt` to RTI by **linear interpolation between the two
NAPS anchors bracketing it** (the van den Dool / Kratz formula for
temperature-programmed / gradient runs):

$$
\text{RTI}(x) \;=\; \text{RTI}_n \;+\; (\text{RTI}_{n+1} - \text{RTI}_n) \cdot \frac{rt_x - rt_n}{rt_{n+1} - rt_n}
$$

where `n` is the NAPS anchor eluting **just before** compound `x` and
`n+1` is the next one **just after**, with `rt_n`, `rt_{n+1}` their
per-lab median apex RTs and `RTI_n`, `RTI_{n+1}` the carbon-number
indices `100 × n_carbon` (so consecutive C-anchors are 100 RTI apart).

The computation uses the **standard Bioconductor helper
`MetaboCoreUtils::indexRtime(x, y, FUN = rtiLinear)`** — `rtiLinear`
is `approx(y$rtime, y$rindex, x)`, i.e. exactly the linear
interpolation written above. Out-of-range values (compound elutes
before the first NAPS anchor or after the last) return `NA`, matching
the classical definition (no extrapolation).

RTI carries two advantages over the consensus-rt approach:

- **Per-lab independence**: each lab's RTI scale is calibrated against
  ITS OWN NAPS injections, so a noisy NAPS replicate in one lab does
  not contaminate other labs' alignments (unlike consensus_rt where
  the cross-lab median absorbs the outlier).
- **Reportable as a standardized cross-lab vocabulary**: RTI is a
  pre-existing convention in the LC-MS community.

This section reports **how reproducible the per-lab RTI is across
labs** for SIRIUS-confirmed compounds — not a comparison to the
spline-aligned RT (the two are essentially the same here); just the
honest cross-lab SD distribution in RTI units.

```{r rti-classical}
#| message: false
# Build per-lab NAPS anchor tables in the column layout
# MetaboCoreUtils::indexRtime() expects: `rtime` (lab-local apex
# seconds) and `rindex` (target RTI = anchor_n × 100 so consecutive
# C-anchors are 100 RTI apart).
suppressPackageStartupMessages(library(MetaboCoreUtils))
naps_rti_long <- per_lab_median |>
  mutate(rtime  = rt_med,
         rindex = anchor_n * 100L) |>
  arrange(lab, polarity, rtime)

# Convert each compound's apex_rt to RTI via the standard
# classical-Kovats helper. `indexRtime(x, y, FUN = rtiLinear)` linearly
# interpolates between the two anchors that bracket x; out-of-range
# values become NA (yleft = yright = NA), matching the classical
# definition (no extrapolation).
chrom_rti <- chrom_with_id
chrom_rti$rti <- NA_real_
for (key in unique(paste(chrom_rti$lab, chrom_rti$polarity, sep = "/"))) {
  parts <- strsplit(key, "/", fixed = TRUE)[[1]]
  sub_anchors <- naps_rti_long |>
    filter(lab == parts[1], polarity == parts[2]) |>
    select(rtime, rindex) |>
    arrange(rtime)
  if (nrow(sub_anchors) < 2L) next
  idx <- which(chrom_rti$lab == parts[1] &
               chrom_rti$polarity == parts[2] &
               !is.na(chrom_rti$apex_rt))
  if (!length(idx)) next
  # Use the default FUN (rtiLinear, internal to MetaboCoreUtils) — the
  # default lookup resolves it within the package namespace. yleft/yright
  # default to NA already, so out-of-range values become NA as desired.
  chrom_rti$rti[idx] <- indexRtime(chrom_rti$apex_rt[idx],
                                    as.data.frame(sub_anchors))
}

# Per (mixture, compound, polarity): cross-lab SD of the RTI values.
rti_repro <- chrom_rti |>
  filter(!is.na(rti)) |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs       = dplyr::n_distinct(lab),
            rti_med      = median(rti, na.rm = TRUE),
            rti_sd       = sd(rti,     na.rm = TRUE),
            rt_consensus = median(apex_rt, na.rm = TRUE),
            .groups      = "drop") |>
  filter(n_labs >= 2L, !is.na(rti_sd))

cat(sprintf("Compounds with cross-lab RTI computable (≥ 2 labs, inside anchor range): %d\n",
            nrow(rti_repro)))
cat(sprintf("Cross-lab RTI SD:  median = %.1f | p75 = %.1f | p95 = %.1f  (RTI units)\n",
            median(rti_repro$rti_sd),
            quantile(rti_repro$rti_sd, 0.75),
            quantile(rti_repro$rti_sd, 0.95)))
```

**Plot 1 — distribution of cross-lab RTI SD per compound.**
*x-axis*: cross-lab SD of the labs' RTI values for one compound (RTI
units, log scale). *y-axis*: number of compounds at that SD. *Red
dashed line* = median across compounds. Two facets = neg / pos
polarity. If RTI is reproducible, the whole distribution sits at low
RTI units — i.e. tens, not hundreds.

```{r rti-cross-lab-sd}
#| fig-cap: "Distribution of cross-lab RTI SD per (mixture × compound × polarity), for SIRIUS-confirmed compounds detected in ≥ 2 labs and inside every lab's NAPS anchor range. Red dashed line = median per polarity."
#| fig-height: 4
med_per_pol <- rti_repro |>
  group_by(polarity) |>
  summarise(med = median(rti_sd, na.rm = TRUE), .groups = "drop")
ggplot(rti_repro |> filter(rti_sd > 0),
       aes(x = rti_sd, fill = polarity)) +
  geom_histogram(bins = 50, alpha = 0.75) +
  geom_vline(data = med_per_pol, aes(xintercept = med),
             colour = "red", linetype = "dashed") +
  geom_text(data = med_per_pol,
            aes(x = med, y = Inf,
                label = sprintf("median = %.1f RTI", med)),
            colour = "red", hjust = -0.1, vjust = 1.4, size = 3) +
  scale_x_log10(labels = scales::label_number(accuracy = 0.1)) +
  scale_fill_manual(values = c(neg = "#3B7C92", pos = "#E64B35")) +
  facet_wrap(~ polarity, ncol = 2) +
  labs(x = "Cross-lab RTI SD (RTI units, log)",
       y = "Number of compounds",
       fill = NULL,
       title = "Cross-lab RTI reproducibility per compound") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r rti-per-lab-bias}
#| fig-cap: "Per-lab RTI residual vs the cross-lab compound consensus RTI. Systematic offsets (median ≠ 0) reveal lab-specific RTI biases — e.g. between-anchor spline curvature differences that the strict NAPS-only spline can't capture."
#| fig-height: 3.5
lab_resid <- chrom_rti |>
  filter(!is.na(rti)) |>
  group_by(mixture, compound, polarity) |>
  mutate(rti_med = median(rti, na.rm = TRUE),
         residual_rti = rti - rti_med) |>
  ungroup() |>
  filter(!is.na(residual_rti))

ggplot(lab_resid, aes(x = lab, y = residual_rti, colour = lab)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ polarity) +
  scale_color_hue() +
  labs(x = NULL, y = "RTI residual vs cross-lab compound consensus",
       title = "Per-lab RTI bias") +
  theme_bw() + theme(legend.position = "none")
```

**Interpretation**. The histogram (Plot 1) shows whether RTI is
reproducible across labs in absolute terms — low cross-lab SD means
"a compound's RTI is roughly the same number in every lab", which is
what would make RTI useful as a transferable cross-lab vocabulary.
The per-lab bias boxplot (Plot 2) shows whether any one lab carries a
**systematic** offset (median ≠ 0) — those are candidates for a
single-offset correction. A lab with a wide box has noisy RTI overall,
typically because its NAPS replicates are noisy and the interpolation
between adjacent anchors inherits that noise.

# P2.9 MS1-anchor every lab's MS2-trigger RTs

The SIRIUS producer writes **`feat_rts.csv`** with per-feature `rt_sec`
values taken from the MS2-trigger times. For every lab there is some
offset between the trigger RT and the chromatographic peak in the
**MS1 file we use for EIC extraction**:

- **hmgu** — single combined MS1+MS2 injection. Trigger RT ≈ peak in
  the same file. Anchoring produces a ~0 s shift but is run for
  symmetry and to surface any per-cycle jitter.
- **afekta, icl** (DDA, two injections) — trigger RT is from the
  `_MS2` injection but the EIC comes from a separate MS1-only
  injection. **Inter-injection drift** typically ~1-5 s.
- **cembio** (inclusion-list, two injections) — same inter-injection
  drift PLUS the MS2 fires on a pre-built schedule rather than at the
  apex, so the trigger RT can be ~10-60 s off the real peak (median
  ~-18 s on mixes 1.1-1.6).

Anchoring is uniform across all four labs: extract the MS1 EIC at
each precursor m/z (chromExtract over all unique windows in one
pass), run the SNR-aware peak picker, then rewrite each row's
`rt_sec` to the strongest peak within the lab's `search_window` of
the trigger RT. Logic lives in
`5_downstream_analysis/ms1_anchor.R` (`anchor_rts()`).

```{r ms1-anchor}
#| message: true
source(here::here("5_downstream_analysis", "ms1_anchor.R"))

feat_rts_in   <- here::here("5_downstream_analysis", "object",
                            "feat_rts.csv")
feat_rts_out  <- here::here("5_downstream_analysis", "object",
                            "feat_rts_ms1_anchored.csv")
anchor_helper <- here::here("5_downstream_analysis", "ms1_anchor.R")
pipe_helper   <- here::here("5_downstream_analysis",
                            "annotated_eic_pipeline.R")
diag_dir      <- here::here("5_downstream_analysis", "object",
                            "ms1_anchor_diagnostics")

# Cache: skip the (slow) anchoring run if the anchored CSV is newer
# than feat_rts.csv, ms1_anchor.R, and annotated_eic_pipeline.R
# (the latter is sourced by ms1_anchor.R for extract_one + spline
# helpers — changes invalidate the candidate-guard logic).
cache_valid <- file.exists(feat_rts_out) &&
  file.exists(file.path(diag_dir, "per_lab_shifts.csv")) &&
  file.info(feat_rts_out)$mtime > file.info(feat_rts_in)$mtime &&
  file.info(feat_rts_out)$mtime > file.info(anchor_helper)$mtime &&
  file.info(feat_rts_out)$mtime > file.info(pipe_helper)$mtime

if (cache_valid) {
  message("[anchor] CACHE HIT — anchored CSV newer than inputs, skipping.")
  message("[anchor]   to force a re-run, `touch` ", basename(feat_rts_in),
          " or delete ", basename(feat_rts_out))
  diag <- list(
    per_lab_shifts = read.csv(file.path(diag_dir, "per_lab_shifts.csv"),
                              stringsAsFactors = FALSE),
    per_lab_status = read.csv(file.path(diag_dir, "per_lab_status.csv"),
                              stringsAsFactors = FALSE),
    overall_status = read.csv(file.path(diag_dir, "overall_status.csv"),
                              stringsAsFactors = FALSE))
} else {
  feat <- read.csv(feat_rts_in, stringsAsFactors = FALSE,
                   colClasses = c(mixture = "character"))
  message(sprintf("[anchor] input %d rows | per-lab: %s",
                  nrow(feat),
                  paste(sprintf("%s=%d", names(table(feat$lab)),
                                as.integer(table(feat$lab))),
                        collapse = ", ")))

  feat_anc <- anchor_rts(feat,
                         splines          = splines,
                         anchor_table     = anchor_table,
                         labs             = c("afekta", "cembio",
                                              "hmgu", "icl"),
                         ppm_tol          = 20,
                         search_window    = 60,
                         snr_threshold    = 5,
                         agreement_window = 8,
                         isobar_rt_window = 5)
  write.csv(feat_anc, feat_rts_out, row.names = FALSE)
  message(sprintf("[anchor] wrote %s (%d rows)",
                  basename(feat_rts_out), nrow(feat_anc)))

  diag <- attr(feat_anc, "diagnostics")
  dir.create(diag_dir, recursive = TRUE, showWarnings = FALSE)
  write.csv(diag$overall_status,
            file.path(diag_dir, "overall_status.csv"), row.names = FALSE)
  write.csv(diag$per_lab_status,
            file.path(diag_dir, "per_lab_status.csv"), row.names = FALSE)
  write.csv(diag$per_lab_shifts,
            file.path(diag_dir, "per_lab_shifts.csv"), row.names = FALSE)
  if (nrow(diag$isobar_summary))
    write.csv(diag$isobar_summary,
              file.path(diag_dir, "isobar_dropped.csv"), row.names = FALSE)
}

cat("Per-lab RT shift summary:\n")
print(diag$per_lab_shifts)
cat("\nPer-lab anchor status:\n")
print(diag$per_lab_status)
```

## P2.9.1 Worked example — why MS1 anchoring matters

The aggregate shift statistics are dominated by **rescue cases**: compounds
where SIRIUS identified the same compound at many different RTs, only one
of which is the real chromatographic peak. The anchoring pulls all the
spurious hits back to the only real peak. **N,N-dimethyl-β-alanine**
(small polar betaine, m/z 118.086, M+H+) in mix 2.2 pos is a textbook
example:

- Three labs (afekta DDA, cembio inclusion-list, hmgu DDA) acquired
  MS2 spectra at this precursor m/z **many times across the gradient**.
- Each lab's MS2 file confirms triggers are real (precursor m/z 118 at the
  reported RTs, distance ≤ 0.5 s from claimed values).
- The MS1 EICs show **one peak at ~25 s**, intensity > 1 × 10⁶ counts.
  Everywhere else there is just flat background of 30–250 k counts — the
  in-source 118 ion that drifts up and down with the matrix elution. No
  chromatographic peak shape, SNR ≪ 5 against the void peak's baseline.
- **Every one of the ~22 MS2 triggers shown in the figure was
  SIRIUS-annotated as N,N-dimethyl-β-alanine** — each orange tick is a
  row in `feat_rts.csv` that SIRIUS scored as a match to this compound.
  SIRIUS sees the matrix-118 fragment spectra and is willing to call them
  dimethyl-β-alanine on fragment match alone (it doesn't know about
  chromatography). So we get 6 afekta hits, ~10 cembio hits, ~6 hmgu hits
  scattered through RT 24–771 s, each one a "successful" SIRIUS
  identification but most of them on background signal — the chemistry
  of identification has no way to reject in-source fragments without an
  external chromatographic constraint.
- The MS1 anchor step extracts the real EIC per lab, runs the SNR picker,
  and snaps every trigger to the only peak that exists — ~25 s. Now the
  cross-lab cluster algorithm sees three labs in tight agreement instead
  of ~20 scattered triggers.

This is the mechanism behind both cembio's −18 s median schedule offset
and the long-tailed |shift| distribution for every DDA lab.

```{r anchor-case-study}
#| fig-cap: "MS1 anchoring on N,N-dimethyl-β-alanine (m/z 118.086, [M+H]+) in mix 2.2 pos for the three labs that annotated it. Blue: MS1 EIC. Orange vertical bars: **every MS2 trigger RT that SIRIUS annotated as dimethyl-β-alanine** (one row per trigger in `feat_rts.csv`). Red points: peaks detected by find_all_peaks above SNR 5. Green dashed line(s): the per-feature anchored RT — where each MS2 hit gets pulled to. The compound has exactly one chromatographic peak; everything else is in-source matrix at m/z 118 that SIRIUS still identified as the target on fragment-pattern match alone."
#| fig-height: 5.5
#| fig-width: 9
#| message: false
source(here::here("5_downstream_analysis", "ms1_anchor.R"))  # ms1_path_for + find_all_peaks

CASE_COMPOUND <- "N,N-dimethyl-beta-alanine"
CASE_MIX <- "2.2"; CASE_POL <- "pos"; CASE_PPM <- 20
LABS_CASE <- c("afekta", "cembio", "hmgu")

# 1. Pull trigger RTs (pre-anchoring) + anchored RTs per feature
feat_pre  <- read.csv(here::here("5_downstream_analysis", "object",
                                  "feat_rts.csv"), stringsAsFactors = FALSE)
feat_post <- read.csv(here::here("5_downstream_analysis", "object",
                                  "feat_rts_ms1_anchored.csv"),
                       stringsAsFactors = FALSE)
case_pre_all <- feat_pre |>
  filter(mixture == CASE_MIX, polarity == CASE_POL,
         compound == CASE_COMPOUND, lab %in% LABS_CASE)
case_post_all <- feat_post |>
  filter(mixture == CASE_MIX, polarity == CASE_POL,
         compound == CASE_COMPOUND, lab %in% LABS_CASE)

# Pick a single m/z window for the displayed EICs. Then RESTRICT the
# displayed triggers + anchored values to features whose ionMass falls
# inside this window — otherwise hmgu would display anchored RTs that
# came from a slightly different m/z's EIC (different EIC = different
# peaks), which is confusing in a single-panel plot.
mz_center <- median(case_pre_all$ionMass)
mz_lo <- mz_center * (1 - CASE_PPM / 1e6)
mz_hi <- mz_center * (1 + CASE_PPM / 1e6)
case_pre  <- case_pre_all  |> filter(ionMass >= mz_lo, ionMass <= mz_hi)
case_post <- case_post_all |> filter(ionMass >= mz_lo, ionMass <= mz_hi)

# Per-lab text summary table — show BOTH inside-window and excluded
# rows so the reader can see the m/z heterogeneity.
case_summary <- case_pre_all |>
  group_by(lab) |>
  summarise(n_triggers = n(),
            n_in_window = sum(ionMass >= mz_lo & ionMass <= mz_hi),
            trigger_rts = paste(round(rt_sec, 0), collapse = ", "),
            .groups = "drop") |>
  left_join(case_post_all |>
              group_by(lab) |>
              summarise(anchored_rts_all = paste(unique(round(rt_sec, 1)),
                                                  collapse = ", "),
                        .groups = "drop"),
            by = "lab")

# 2. Resolve MS1 paths and run ONE chromExtract over all three labs.
paths_df <- tibble::tibble(lab = LABS_CASE) |>
  mutate(path = vapply(lab,
                       \(l) ms1_path_for(l, CASE_MIX, CASE_POL),
                       character(1))) |>
  filter(!is.na(path), file.exists(path))

# Cache the case-study EIC (chromData + peaksData). chromExtract reads
# mzML via mzR which segfaults intermittently here; once this succeeds
# the cache lets every re-render skip it. Delete case_study_eic.rds (or
# touch feat_rts_ms1_anchored.csv) to force a re-extract.
case_eic_rds <- file.path(producer_dir, "case_study_eic.rds")
if (file.exists(case_eic_rds) &&
    file.mtime(case_eic_rds) > file.mtime(feat_rts_out)) {
  .ce <- readRDS(case_eic_rds); cd <- .ce$cd; pk <- .ce$pk
} else {
  sps <- Spectra::Spectra(paths_df$path, backend = Spectra::MsBackendMzR())
  sps <- sps[Spectra::msLevel(sps) == 1L]
  canon <- unique(Spectra::dataOrigin(sps))
  paths_df$dataOrigin <- canon[match(basename(paths_df$path),
                                      basename(canon))]
  pt <- paths_df |>
    transmute(mzMin = mz_lo, mzMax = mz_hi,
              rtMin = 0, rtMax = 1000, msLevel = 1L,
              dataOrigin, lab)
  chr <- Chromatograms::chromExtract(
    Chromatograms::Chromatograms(sps),
    peak.table = pt,
    by = c("msLevel", "dataOrigin"))
  cd <- as.data.frame(Chromatograms::chromData(chr))
  pk <- Chromatograms::peaksData(chr)
  saveRDS(list(cd = cd, pk = pk), case_eic_rds)
}

# 3. Build per-lab tidy data: EIC + detected peaks
eic_df <- purrr::map_dfr(seq_len(nrow(cd)), \(i) {
  pd <- pk[[i]]
  if (is.null(pd) || !nrow(pd)) return(NULL)
  tibble::tibble(lab = cd$lab[i],
                 rt = pd[, "rtime"], intensity = pd[, "intensity"])
})
peaks_df <- purrr::map_dfr(seq_len(nrow(cd)), \(i) {
  pd <- pk[[i]]
  if (is.null(pd) || !nrow(pd)) return(NULL)
  ec <- data.frame(rt = pd[, "rtime"], intensity = pd[, "intensity"])
  pks <- find_all_peaks(ec, snr_threshold = 5)
  if (!nrow(pks)) return(NULL)
  # Display filter: drop peaks below 50% of the lab's max-intensity peak.
  # SNR alone passes baseline noise scrapes for very weak EICs (e.g.
  # hmgu's m/z 118 at ~120 counts, where noise of 30 counts still
  # crosses SNR 5). These never affect the anchoring (the strongest
  # peak still wins) but they clutter the figure.
  pks <- pks[pks$maxIntensity >= 0.5 * max(pks$maxIntensity), , drop = FALSE]
  tibble::as_tibble(pks) |> mutate(lab = cd$lab[i])
})

# 4. Per-lab y-max for sizing trigger bars + plotting anchored values.
y_max_per_lab <- eic_df |>
  group_by(lab) |>
  summarise(y_max = max(intensity, na.rm = TRUE), .groups = "drop")

trigger_df <- case_pre |>
  left_join(y_max_per_lab, by = "lab") |>
  mutate(y_top = 0.12 * y_max)

anchored_df <- case_post |>
  dplyr::distinct(lab, rt_sec) |>
  dplyr::left_join(y_max_per_lab, by = "lab") |>
  dplyr::rename(anchored_rt = rt_sec)

# 5. Plot
lab_levels <- intersect(LABS_CASE, unique(eic_df$lab))
eic_df$lab        <- factor(eic_df$lab, levels = lab_levels)
peaks_df$lab      <- factor(peaks_df$lab, levels = lab_levels)
trigger_df$lab    <- factor(trigger_df$lab, levels = lab_levels)
anchored_df$lab   <- factor(anchored_df$lab, levels = lab_levels)
case_summary$lab  <- factor(case_summary$lab, levels = lab_levels)

knitr::kable(case_summary,
             col.names = c("lab", "n triggers", "n in m/z window",
                            "MS2 trigger RTs (s)",
                            "anchored RT(s) (s) — all features"),
             caption = "Per-lab MS2 trigger counts and the distinct anchored RTs for N,N-dimethyl-β-alanine, mix 2.2 pos. `n in m/z window` counts how many features fall inside the displayed ±20 ppm window of m/z 118.086; outside-window features still anchor (using their own m/z's EIC) but aren't drawn on the figure. hmgu has multiple distinct anchored RTs because its per-feature `ionMass` values vary by ~26 ppm; the outlier at 252 s comes from an ionMass-118.0832 feature whose m/z window catches a different background ridge.")

# Drop empty-scan NA intensities (chromExtract pads the RT grid with NA
# where no signal); geom_line would otherwise warn about removed rows.
ggplot(eic_df |> filter(!is.na(intensity)), aes(x = rt, y = intensity)) +
  geom_line(colour = "#1F77B4", linewidth = 0.4) +
  # MS2 trigger bars: visible vertical lines at bottom
  geom_segment(data = trigger_df,
               aes(x = rt_sec, xend = rt_sec, y = 0, yend = y_top),
               colour = "darkorange", linewidth = 1.0,
               inherit.aes = FALSE) +
  # Anchored RT lines — one per distinct value per lab
  geom_vline(data = anchored_df,
             aes(xintercept = anchored_rt),
             colour = "darkgreen", linewidth = 0.7,
             linetype = "dashed") +
  # Labels for anchored RT lines
  geom_text(data = anchored_df,
            aes(x = anchored_rt, y = y_max * 0.9,
                label = sprintf("→ %.0f s", anchored_rt)),
            colour = "darkgreen", size = 3, hjust = -0.05,
            inherit.aes = FALSE) +
  geom_point(data = peaks_df,
             aes(x = apex_rt, y = maxIntensity),
             colour = "red", size = 2, inherit.aes = FALSE) +
  facet_wrap(~ lab, ncol = 1, scales = "free_y") +
  # coord_cartesian zooms without dropping data (scale_*_continuous(limits=)
  # silently removes out-of-range rows and warns; coord_cartesian does not).
  coord_cartesian(xlim = c(0, 1000)) +
  scale_y_continuous(labels = scales::label_scientific(digits = 1)) +
  labs(x = "RT (s)", y = "MS1 intensity",
       title = sprintf("%s — m/z %.4f ± %d ppm — mix %s %s",
                       CASE_COMPOUND, mz_center, CASE_PPM,
                       CASE_MIX, CASE_POL),
       subtitle = "orange bars = MS2 trigger RTs   |   red dots = MS1 peaks above SNR 5   |   green dashed = anchored RT for each feature (label shows value)") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

## P2.9.2 Worked example — candidate + consensus resolves a multi-peak m/z window

The dominant-peak example above assumes the strongest peak at the *m/z*
**is** the compound. That fails when a second peak of comparable height
sits at the same *m/z* in a different RT region. cembio mix 2.1 pos at
the L-phenylalanine *m/z* (166.086) has **two** clean peaks of nearly
equal height — one at ~43 s (5.7 M) and one at ~58 s (5.3 M).

Both peaks are physically real (different compounds at the same *m/z*,
chromatographically resolved by the LC). To assign which one is L-phe in
cembio, the candidate + consensus rule pools every ≥ 50 % candidate from
**every (lab, polarity, adduct)** for the compound and decides in
**aligned consensus space** — the time axis the per-lab splines map onto.
For L-phe the aligned consensus sits at **48.5 s** (afekta pos + hmgu neg
+ both cembio polarities' 43 s candidates all agree there;
cembio-pos-only candidates at aligned ~63 disagree). Inverse-aligning
48.5 s consensus to cembio-pos local time lands at **~43 s** — the
cembio peak whose MS2 trigger evidence is shown by the orange dashes
clustered around it.

This worked example also exposes a real limitation: NAPS-only alignment
interpolates over an 86 s gap between C4 (lab-local 43.6 s) and C5
(lab-local 130 s), and that interpolation isn't perfect — cembio's
aligned-space L-phe residual (43 s candidates → consensus 48.5 vs 58 s
candidates → consensus 63) is much larger than the dataset-wide
post-alignment median SD (~2 s). The figure encodes the principled
choice (aligned-space consensus → 43 s), but the assignment for
compounds in NAPS-sparse regions should be reviewed when alignment is
improved (e.g. by adding a few unambiguous mid-RT helpers).

```{r anchor-case-lphe}
#| fig-cap: "cembio mix 2.1 pos EIC at the L-phenylalanine m/z (166.086). Two real, chromatographically-separated peaks at this m/z. The cross-lab aligned-space consensus (48.5 s on the consensus axis) inverse-aligns to **~43 s on cembio's local axis** (green line) — the same RT where MS2 was triggered (orange) and where other labs' L-phe lands after alignment (grey dotted = each non-cembio lab's L-phe consensus value inverse-aligned to cembio's local time). The 58 s peak (red) is a co-eluting interferent at the same m/z."
#| fig-height: 4
#| message: false
source(here::here("5_downstream_analysis", "annotated_eic_pipeline.R"))  # build_inverse_splines, apply_inv

LP_MZ <- 166.0862; LP_PPM <- 20
lp_lo <- LP_MZ * (1 - LP_PPM / 1e6); lp_hi <- LP_MZ * (1 + LP_PPM / 1e6)

# Aligned-space consensus RT for L-phe — read from the post-anchor csv
# (where Stage 2 stores it as the `consensus_rt_aligned` column).
LP_CONS_ALIGNED <- feat_post |>
  dplyr::filter(compound == "L-phenylalanine", mixture == "2_1") |>
  dplyr::pull(consensus_rt_aligned) |>
  stats::median(na.rm = TRUE)

inv_splines <- build_inverse_splines(splines, anchor_table)
lp_cembio_local <- apply_inv(LP_CONS_ALIGNED, "cembio", "pos",
                             inv_splines, anchor_table)

# cembio EIC at the L-phe m/z (cached; openMSfile reads the mzML directly).
lphe_rds <- file.path(producer_dir, "lphe_case_eic.rds")
if (file.exists(lphe_rds)) {
  lp_eic <- readRDS(lphe_rds)
} else {
  lp_path <- ms1_path_for("cembio", "2.1", "pos")
  lp_m <- mzR::openMSfile(lp_path, backend = "pwiz")
  lp_h <- mzR::header(lp_m); lp_idx <- which(lp_h$msLevel == 1L)
  lp_eic <- data.frame(
    rt = lp_h$retentionTime[lp_idx],
    intensity = vapply(lp_idx, function(j) {
      sp <- mzR::peaks(lp_m, j); if (!nrow(sp)) return(0)
      k <- sp[, 1] >= lp_lo & sp[, 1] <= lp_hi
      if (any(k)) sum(sp[k, 2]) else 0 }, numeric(1)))
  mzR::close(lp_m); saveRDS(lp_eic, lphe_rds)
}

# Cembio's MS2 trigger RTs at m/z 166 — read from cembio's separate
# _MS2 mzML (ms1_path_for() returns the MS1-only file which has zero
# MS2 scans). Capture EVERY MS2 firing whose precursor m/z lands in
# the L-phe window, not just the ones SIRIUS annotated as L-phe.
lp_ms2_rds <- file.path(producer_dir, "lphe_case_ms2_triggers_v2.rds")
if (file.exists(lp_ms2_rds)) {
  lp_ms2_rts <- readRDS(lp_ms2_rds)
} else {
  lp_ms2_path <- list.files(
    here::here("1_preprocessing", "cembio", "HE", "mzml"),
    pattern = "_pos_MS2_.*2\\.1.*\\.mzML$",
    full.names = TRUE)[1]
  if (is.na(lp_ms2_path)) {
    lp_ms2_rts <- numeric(0)
  } else {
    lp_m <- mzR::openMSfile(lp_ms2_path, backend = "pwiz")
    lp_h <- mzR::header(lp_m)
    ms2_idx <- which(lp_h$msLevel == 2L &
                     !is.na(lp_h$precursorMZ) &
                     lp_h$precursorMZ >= lp_lo &
                     lp_h$precursorMZ <= lp_hi)
    lp_ms2_rts <- lp_h$retentionTime[ms2_idx]
    mzR::close(lp_m)
  }
  saveRDS(lp_ms2_rts, lp_ms2_rds)
}

# Other labs' L-phe consensus inverse-aligned into cembio's local time:
# i.e. where each non-cembio lab "sees" L-phe when projected onto cembio's
# chromatographic axis. Demonstrates that alignment brings them all to
# the same region (~43 s) as cembio's chosen peak.
other_labs_lphe <- feat_post |>
  dplyr::filter(compound == "L-phenylalanine", mixture == "2_1",
                lab != "cembio") |>
  dplyr::distinct(lab, polarity, rt_sec)
other_aligned <- mapply(apply_fwd, other_labs_lphe$rt_sec,
                        other_labs_lphe$lab, other_labs_lphe$polarity,
                        MoreArgs = list(splines = splines,
                                         anchor_table = anchor_table))
other_in_cembio <- apply_inv(other_aligned, "cembio", "pos",
                              inv_splines, anchor_table)
support_df <- tibble::tibble(
  lab_pol = paste(other_labs_lphe$lab, other_labs_lphe$polarity),
  rt_in_cembio = other_in_cembio) |> dplyr::filter(!is.na(rt_in_cembio))

lp_pks  <- find_all_peaks(lp_eic, snr_threshold = 5)
lp_cand <- lp_pks[lp_pks$maxIntensity >= 0.5 * max(lp_pks$maxIntensity), , drop = FALSE]
lp_isobar <- lp_cand$apex_rt[which.max(abs(lp_cand$apex_rt - lp_cembio_local))]

ms2_df <- tibble::tibble(rt = lp_ms2_rts,
                          y_top = 0.18 * max(lp_eic$intensity, na.rm = TRUE))

ggplot(lp_eic |> dplyr::filter(rt <= 130, !is.na(intensity)),
       aes(rt, intensity)) +
  geom_line(colour = "#1F77B4", linewidth = 0.5) +
  geom_segment(data = ms2_df,
               aes(x = rt, xend = rt, y = 0, yend = y_top),
               colour = "darkorange", linewidth = 0.9, alpha = 0.85) +
  geom_vline(data = support_df, aes(xintercept = rt_in_cembio),
             colour = "grey55", linetype = "dotted", linewidth = 0.45) +
  geom_vline(xintercept = lp_cembio_local, colour = "darkgreen",
             linewidth = 0.9) +
  geom_point(data = lp_cand, aes(apex_rt, maxIntensity),
             colour = "grey30", size = 1.6) +
  annotate("point", x = lp_isobar,
           y = lp_eic$intensity[which.min(abs(lp_eic$rt - lp_isobar))],
           colour = "red", size = 3) +
  annotate("text", x = lp_isobar, y = max(lp_eic$intensity) * 0.95,
           label = sprintf("co-eluting peak\n%.0f s (rejected)", lp_isobar),
           colour = "red", size = 3, hjust = 1.1) +
  annotate("text", x = lp_cembio_local, y = max(lp_eic$intensity) * 0.6,
           label = sprintf("consensus → cembio\n%.1f s", lp_cembio_local),
           colour = "darkgreen", size = 3, hjust = -0.1) +
  scale_y_continuous(labels = scales::label_scientific(digits = 1)) +
  labs(x = "cembio RT (s)", y = "MS1 intensity",
       title = "L-phenylalanine in cembio mix 2.1 pos — aligned-space consensus",
       subtitle = paste(
         sprintf("green = aligned consensus (%.1f s) → cembio local",
                  LP_CONS_ALIGNED),
         "orange = MS2 triggers at m/z 166",
         "grey dotted = other labs' L-phe brought into cembio's time",
         "red = rejected co-eluting peak",
         sep = "  |  ")) +
  theme_bw()
```

# P2.10 Stage 3 — per-lab abundance measurement

This section consumes the **MS1-anchored** `feat_rts` (from §P2.9,
where Stage 1 + Stage 2 have already produced the per-(mixture,
compound) consensus RT) and the **`annotation_identity.csv`**
(per-row SIRIUS identifications), inverse-aligns the consensus RT to
each lab's local time via the per-lab inverse splines, extracts the
EIC from each lab's mzML at the precursor *m/z*, runs the rule-based
peak picker with the same six guards used in Stage 1 (§8 of the
presentation outline), and writes
**`5_downstream_analysis/object/annotation_chrom_metrics.csv`**.

All algorithmic code lives in `annotated_eic_pipeline.R` to keep this
QMD readable. The cross-lab consensus and isobar resolution happen
**upstream** in `ms1_anchor.R::anchor_rts()` (§P2.9). The downstream
pipeline still runs a duplicate consensus-clustering pass for
historical reasons, but on the new inputs it is essentially a no-op
— every (mixture, compound) already converges to a single cluster
because Stage 2 snapped every row to the same consensus-aligned
candidate.

**EIC imputation (interior NA-fill).** The SNR-aware picker
(`extract_one`) linearly fills *interior* NA gaps in each EIC before
peak detection (`rule = 1`: leading/trailing NAs are left as NA so the
tail-noise estimate is not contaminated by flat edge extrapolation).
Sparse MS1 scan grids — notably icl, and some late-RT hmgu *m/z* — can
under-sample a real peak to a single non-NA apex scan, which the
`n_scans_in_shape >= 2` guard would otherwise reject as a lone spike.

A controlled before/after test (`impute = FALSE` vs `TRUE` on the same
extracted EICs, current picker with all guards) shows imputation
**rescues 13 below_detection rows to detected — every one with
SNR ≥ 6.9** (median 368), so no noise is fabricated. The *net* gain is
small (**+3**: +13 rescued, −10 lost) because the real-scan-support
(`n_real_near_apex ≥ 3`) and signal-concentration (`≥ 0.45`) guards now
do most of the quality filtering, and they correctly reject 10 marginal
rows that interpolation alone would have passed:

| lab | net Δ | rescued | lost |
|-----|------:|--------:|-----:|
| afekta | −2 | 0 | 2 |
| cembio | −1 | 2 | 3 |
| hmgu   | +1 | 4 | 3 |
| icl    | +5 | 7 | 2 |

Imputation is kept because the 13 rescues are genuine high-SNR
under-sampled peaks; with the stricter guards in place it is a modest
refinement, not the headline it was before the guards existed.

```{r run-annotated-eic-pipeline}
#| message: true
source(here::here("5_downstream_analysis", "annotated_eic_pipeline.R"))

# anchor_table is built earlier in §P2.4 (NAPS + qualifying standards).
# splines is the list of Hyman fits built in §P2.4.
mixtures_all <- c("1_1","1_2","1_3","1_4","1_5","1_6",
                  "2_1","2_2","2_3","2_4","2_5","2_6",
                  "2_7","2_8","2_9","2_10",
                  "3_1","3_2","3_3")

obj_dir   <- here::here("5_downstream_analysis", "object")
out_csv   <- file.path(obj_dir, "annotation_chrom_metrics.csv")
out_rds   <- file.path(obj_dir, "pipeline_out.rds")
pipe_src  <- here::here("5_downstream_analysis", "annotated_eic_pipeline.R")
ident_csv <- file.path(obj_dir, "annotation_identity.csv")

# Cache: skip the (slow) chromExtract + picker if pipeline_out.rds is newer
# than every input (anchored feat_rts, identities, and the algorithm file).
# To force a re-run, delete pipeline_out.rds or touch annotated_eic_pipeline.R.
inputs_mtime <- max(file.mtime(feat_rts_out), file.mtime(ident_csv),
                    file.mtime(pipe_src))
if (file.exists(out_rds) && file.exists(out_csv) &&
    file.mtime(out_rds) > inputs_mtime) {
  message("[pipeline] CACHE HIT — pipeline_out.rds newer than inputs, skipping.")
  pipeline_out <- readRDS(out_rds)
} else {
  pipeline_out <- run_annotated_eic_pipeline(
    feat_rts_path = feat_rts_out,   # MS1-anchored from §P2.9
    identity_path = ident_csv,
    splines       = splines,
    anchor_table  = anchor_table,
    output_path   = out_csv,
    mixtures      = mixtures_all,
    eic_dump_path = file.path(obj_dir, "eic_dump.rds"))
  saveRDS(pipeline_out, out_rds)
}

winning_cluster <- pipeline_out$winning_cluster
cluster_scores  <- pipeline_out$cluster_scores
ann_chrom       <- pipeline_out$annotation_chrom_metrics
isobar_diag     <- pipeline_out$isobar_diagnostics
```

### Isobaric collision resolution — audit

Cross-compound isobaric resolution is part of **Stage 2 in
`ms1_anchor.R`** now (see §P2.9): per mixture, pairs of compounds
whose consensus RTs are within 5 s and whose adducts share any *m/z*
within 20 ppm collapse to the higher-scoring compound
(`n_labs × 10 + n_polarities × 5 + Σ COSMIC`), and the lower-scoring
side's rows are dropped before they reach this pipeline.

The check below is a **post-hoc audit**: `annotated_eic_pipeline.R`
still runs the same collision check on its inputs, so a non-zero count
here would mean Stage 2 missed a case worth investigating. The
expected outcome is **0** — anything detected upstream has already
been resolved.

```{r isobar-diagnostics}
cat(sprintf("Audit — collisions still present at the downstream step: %d (expected 0)\n",
            isobar_diag$n_collisions_resolved))
if (nrow(isobar_diag$dropped) > 0L) {
  knitr::kable(isobar_diag$dropped |>
                 dplyr::transmute(mixture, compound,
                                  consensus_rt = round(consensus_rt, 1),
                                  consensus_score = round(consensus_score, 2)),
               caption = "Collisions Stage 2 missed (would indicate a bug in the upstream rule).")
} else {
  cat("(audit OK — Stage 2 resolved everything upstream)\n")
}

# Compounds Stage 2 actually resolved (read from the ms1_anchor diagnostics).
isobar_log <- file.path(diag_dir, "isobar_dropped.csv")
if (file.exists(isobar_log)) {
  ms1_drops <- read.csv(isobar_log, stringsAsFactors = FALSE)
  cat(sprintf("\nStage 2 (ms1_anchor.R) dropped %d compounds as lower-scoring isobars:\n",
              nrow(ms1_drops)))
  if (nrow(ms1_drops))
    knitr::kable(ms1_drops,
                 caption = "Compounds resolved upstream in Stage 2.")
}
```

**How to read this.** For each `(mixture, compound)`, the Stage 2
similarity rule (§P2.9) pools every Stage-1 candidate across labs,
polarities, and adducts and computes a consensus RT supported by some
number of `(lab, polarity)` pairs. The heatmap is a 2-D histogram of
that support — *the best cross-lab evidence for that compound's RT*,
not a requirement that all labs agree. A compound annotated by only 1
or 2 labs has consensus support of 1–2 `(lab, polarity)` pairs; that
is expected, not a failure (most spike-ins are not annotated by all
four labs — see the detection summary in QMD 8).

- **x = number of labs** in the consensus support set (1–4). Higher =
  stronger cross-lab agreement on the RT.
- **y = number of polarities** in the consensus support set: `1` = the
  compound's consensus came from a single polarity, `2` = corroborated
  in **both** pos and neg.
- **cell number / colour = how many compounds** fall in that
  `(n_labs, n_polarities)` bin. Colour just re-encodes that count
  (bright = many compounds); the printed number is the same value.

Bottom-left (1 lab, 1 polarity) = weakly-supported compounds;
top-right (4 labs, 2 polarities) = strongest-consensus compounds.

```{r consensus-support-diagnostics}
#| fig-cap: "Distribution of cross-lab consensus support per compound. Each cell = how many compounds had that combination of (n labs, n polarities) supporting the Stage-2 consensus RT."
#| fig-height: 4
winning_cluster |>
  dplyr::count(consensus_n_labs, consensus_n_pol) |>
  dplyr::mutate(pol_lab = factor(consensus_n_pol, levels = c(1, 2),
                                 labels = c("1 (single)", "2 (pos+neg)"))) |>
  ggplot2::ggplot(ggplot2::aes(x = factor(consensus_n_labs),
                                y = pol_lab,
                                fill = n)) +
  ggplot2::geom_tile(colour = "white") +
  ggplot2::geom_text(ggplot2::aes(label = n), size = 4, colour = "grey15") +
  ggplot2::scale_fill_viridis_c(name = "n compounds") +
  ggplot2::labs(x = "Number of labs supporting the consensus RT",
                y = "Number of polarities supporting the consensus RT",
                title = "Cross-lab consensus support (one cell per compound)") +
  ggplot2::theme_bw()
```

```{r detection-summary}
ann_chrom |>
  dplyr::count(lab, polarity, detection) |>
  tidyr::pivot_wider(names_from = detection, values_from = n,
                     values_fill = 0L) |>
  dplyr::mutate(total = detected + below_detection + no_data,
                pct_detected = round(100 * detected / total, 1)) |>
  knitr::kable(caption = "Detection counts per (lab, polarity) — the per-lab sensitivity for the annotated spike-in panel.")
```

# Summary

- Within-lab NAPS replicate SD is the precision floor; numbers in
  P2.2.
- Per-lab Hyman monotonic splines map local apex RT → consensus RT
  per polarity (P2.3). Standards extension active by default
  (`use_standards_anchors = TRUE`).
- Cross-lab apex-RT SD distribution before vs after alignment is in
  P2.5. The post-alignment median should approach the within-lab
  floor everywhere except the C1–C5 early-RT region (P2.6).
- Outputs at `5_downstream_analysis/object/`:
  - `rt_alignment_table.csv` — interpolation lookup
  - `rt_alignment_anchors.csv` — anchor table used for fitting
  - `rt_alignment_splines.RData` — `splines` list of `splinefun`
    objects
- P3 and P5 should load `rt_alignment_table.csv` (or RData for
  sub-second precision) and apply per `(lab, polarity)`.
