---
title: "Peak Feature Calculation — Detected Peaks"
format:
  html:
    toc: true
execute:
  dir: project
---

# Introduction

This document performs peak feature calculation and analysis for the detected
ion chromatograms (EICs). It reuses the processed detected EICs, computes peak
metrics, and runs downstream comparisons and stability analyses.

- Use all detected peaks form the beginning, in the chromData have info wether they are annotated
  - only keep peaks overlapping the 4 labs to create a common feature set

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)
library(UpSetR)

base_path <- file.path("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
```

# Load Detected EICs

We are not computing the eic at the moment because that is justy TOO many peaks

```{r load-eics-detected, warning = FALSE, eval = FALSE}
load(file.path(base_path, "object/eics_detected.RData"))
# Impute and apply processing
eics <- imputePeaksData(eics, method = "linear")
eics <- applyProcessing(eics)
```

Before anything let's load all the detected peaks table 

```{r}
pt_afekta <- read.csv(file.path(base_path, "object/detected_peaks_afekta_HE.csv"))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(base_path, "object/detected_peaks_cembio_HE.csv"))
```

# Create Consensus Peak Table Across Labs

We use a two-stage approach to efficiently match peaks across labs:

1. **Coarse grouping** by m/z bins (window of 1 m/z) to reduce search space
2. **Fine matching** within each bin using `mclosest` on RT and m/z
3. **Map chrom_peak_id** from each lab for EIC extraction

This avoids comparing all ~1M combinations and instead groups by approximate m/z.

```{r build-consensus-peaks}
library(MetaboCoreUtils)

# Define peaks data list with lab names
peaks_list <- list(
  afekta = pt_afekta,
  hmgu = pt_hmgu,
  icl = pt_icl,
  cembio = pt_cembio
)
peak_cols <- 1:4

# Strategy: Group peaks by m/z windows to reduce comparisons
# 1. Create m/z bins (window = 1 m/z)
# 2. For each bin, find peaks from all labs
# 3. Use mclosest within each bin group

mz_window <- 1.0  # 1 m/z window for coarse grouping

# Get global m/z range
all_mz_min <- min(sapply(peaks_list, function(x) min(x$mzmin)))
all_mz_max <- max(sapply(peaks_list, function(x) max(x$mzmax)))

mz_breaks <- seq(floor(all_mz_min), ceiling(all_mz_max), by = mz_window)

# Store matches: for each afekta peak, find closest peak in other labs
consensus_matches <- data.frame(
  idx_afekta = integer(),
  idx_hmgu = integer(),
  idx_icl = integer(),
  idx_cembio = integer(),
  n_labs_matched = integer()
)

## Symmetric per-bin clustering (captures peaks present in any lab)
# Parameters (tunable)
ppm_tol <- 10  # ppm tolerance for m/z
rt_tol  <- 10  # RT tolerance in seconds

compute_norm_dist <- function(mz_vec, rt_vec, ppm_tol, rt_tol) {
  n <- length(mz_vec)
  if (n == 1) return(matrix(0, 1, 1))
  D <- matrix(0, n, n)
  for (ii in seq_len(n - 1)) {
    for (jj in (ii + 1):n) {
      mz_mean <- (mz_vec[ii] + mz_vec[jj]) / 2
      mz_ppm <- abs(mz_vec[ii] - mz_vec[jj]) * 1e6 / mz_mean
      mz_norm <- mz_ppm / ppm_tol
      rt_norm <- abs(rt_vec[ii] - rt_vec[jj]) / rt_tol
      D[ii, jj] <- D[jj, ii] <- sqrt(mz_norm^2 + rt_norm^2)
    }
  }
  D
}

collect_bin_rows <- function(bin_peaks, pt_afekta, pt_hmgu, pt_icl, pt_cembio) {
  rows <- list()
  if (length(bin_peaks$afekta) > 0) {
    idx <- bin_peaks$afekta
    rows[[length(rows) + 1]] <- data.frame(lab = "afekta",
                                           idx = idx,
                                           mz = (pt_afekta$mzmin[idx] + pt_afekta$mzmax[idx]) / 2,
                                           rt = (pt_afekta$rtmin[idx] + pt_afekta$rtmax[idx]) / 2,
                                           stringsAsFactors = FALSE)
  }
  if (length(bin_peaks$hmgu) > 0) {
    idx <- bin_peaks$hmgu
    rows[[length(rows) + 1]] <- data.frame(lab = "hmgu",
                                           idx = idx,
                                           mz = (pt_hmgu$mzmin[idx] + pt_hmgu$mzmax[idx]) / 2,
                                           rt = (pt_hmgu$rtmin[idx] + pt_hmgu$rtmax[idx]) / 2,
                                           stringsAsFactors = FALSE)
  }
  if (length(bin_peaks$icl) > 0) {
    idx <- bin_peaks$icl
    rows[[length(rows) + 1]] <- data.frame(lab = "icl",
                                           idx = idx,
                                           mz = (pt_icl$mzmin[idx] + pt_icl$mzmax[idx]) / 2,
                                           rt = (pt_icl$rtmin[idx] + pt_icl$rtmax[idx]) / 2,
                                           stringsAsFactors = FALSE)
  }
  if (length(bin_peaks$cembio) > 0) {
    idx <- bin_peaks$cembio
    rows[[length(rows) + 1]] <- data.frame(lab = "cembio",
                                           idx = idx,
                                           mz = (pt_cembio$mzmin[idx] + pt_cembio$mzmax[idx]) / 2,
                                           rt = (pt_cembio$rtmin[idx] + pt_cembio$rtmax[idx]) / 2,
                                           stringsAsFactors = FALSE)
  }
  if (length(rows) == 0) return(NULL)
  do.call(rbind, rows)
}

# Process by m/z bin using clustering
for (i in seq_along(mz_breaks[-length(mz_breaks)])) {
  mz_low <- mz_breaks[i]
  mz_high <- mz_breaks[i + 1]

  bin_peaks <- lapply(peaks_list, function(x) {
    which(x$mzmin >= mz_low & x$mzmax <= mz_high)
  })

  if (all(lengths(bin_peaks) == 0)) next

  combined <- collect_bin_rows(bin_peaks, pt_afekta, pt_hmgu, pt_icl, pt_cembio)
  if (is.null(combined)) next

  D <- compute_norm_dist(combined$mz, combined$rt, ppm_tol, rt_tol)

  if (nrow(D) == 1) {
    cluster_ids <- 1L
  } else {
    hc <- hclust(as.dist(D), method = "complete")
    cluster_ids <- cutree(hc, h = 1)
  }

  clusters <- split(seq_len(nrow(combined)), cluster_ids)
  for (cl in clusters) {
    members <- combined[cl, , drop = FALSE]
    idx_afekta <- if ("afekta" %in% members$lab) members$idx[members$lab == "afekta"][1] else NA_integer_
    idx_hmgu   <- if ("hmgu"   %in% members$lab) members$idx[members$lab == "hmgu"][1]   else NA_integer_
    idx_icl    <- if ("icl"    %in% members$lab) members$idx[members$lab == "icl"][1]    else NA_integer_
    idx_cembio <- if ("cembio" %in% members$lab) members$idx[members$lab == "cembio"][1] else NA_integer_

    n_labs_matched <- sum(!is.na(c(idx_afekta, idx_hmgu, idx_icl, idx_cembio)))

    bin_results <- data.frame(
      idx_afekta = idx_afekta,
      idx_hmgu = idx_hmgu,
      idx_icl = idx_icl,
      idx_cembio = idx_cembio,
      n_labs_matched = n_labs_matched
    )
    consensus_matches <- rbind(consensus_matches, bin_results)
  }
}

# Analyze lab overlap before filtering

## Summary statistics
cat("=== LAB OVERLAP ANALYSIS ===\n")
cat("Total peaks in afekta:", nrow(pt_afekta), "\n")
cat("Total peaks in hmgu:", nrow(pt_hmgu), "\n")
cat("Total peaks in icl:", nrow(pt_icl), "\n")
cat("Total peaks in cembio:", nrow(pt_cembio), "\n\n")

# Add detection flags to consensus_matches
consensus_matches <- consensus_matches |>
  mutate(
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  )

cat("Peaks matched in consensus (before filtering):", nrow(consensus_matches), "\n\n")
cat("Peak counts by number of labs detected:\n")
print(table(consensus_matches$n_labs_matched))
```

## Upset Plot - Lab Overlap

```{r upset-plot, message = FALSE, warning = FALSE}
library(UpSetR)

# Prepare binary matrix for upset plot
upset_data <- consensus_matches |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(
    afekta = as.integer(afekta_detected),
    hmgu = as.integer(hmgu_detected),
    icl = as.integer(icl_detected),
    cembio = as.integer(cembio_detected)
  ) |>
  select(afekta, hmgu, icl, cembio)

upset(as.data.frame(upset_data), 
      sets = c("afekta", "hmgu", "icl", "cembio"),
      order.by = "freq",
      main.bar.color = "steelblue",
      sets.bar.color = "darkred",
      text.scale = c(1.3, 1.3, 1, 1, 1.5, 1),
      point.size = 3,
      line.size = 2)
```

## Pairwise Lab Comparisons

```{r pairwise-overlap}
# Create pairwise overlap summary
labs <- c("afekta", "hmgu", "icl", "cembio")
pairwise_summary <- data.frame()

for (i in seq_along(labs)) {
  for (j in seq_along(labs)) {
    if (i < j) {
      lab1 <- labs[i]
      lab2 <- labs[j]
      
      col1 <- paste0(lab1, "_detected")
      col2 <- paste0(lab2, "_detected")
      
      both <- sum(consensus_matches[[col1]] & consensus_matches[[col2]])
      only_lab1 <- sum(consensus_matches[[col1]] & !consensus_matches[[col2]])
      only_lab2 <- sum(!consensus_matches[[col1]] & consensus_matches[[col2]])
      
      pairwise_summary <- rbind(pairwise_summary, data.frame(
        lab1 = lab1,
        lab2 = lab2,
        both_detected = both,
        only_lab1 = only_lab1,
        only_lab2 = only_lab2,
        overlap_percent = round(100 * both / max(
          sum(consensus_matches[[col1]]),
          sum(consensus_matches[[col2]])
        ), 1)
      ))
    }
  }
}

knitr::kable(pairwise_summary, caption = "Pairwise Lab Overlap")
```

## Summary by Detection Pattern

```{r detection-patterns}
# Count peaks by detection pattern
detection_patterns <- consensus_matches |>
  group_by(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  summarize(n_peaks = n(), .groups = "drop") |>
  arrange(desc(n_peaks)) |>
  mutate(
    pattern = paste0(
      if_else(afekta_detected, "A", "-"),
      if_else(hmgu_detected, "H", "-"),
      if_else(icl_detected, "I", "-"),
      if_else(cembio_detected, "C", "-")
    )
  ) |>
  select(pattern, n_peaks, everything())

knitr::kable(detection_patterns, 
             caption = "Peak Detection Patterns (A=afekta, H=hmgu, I=icl, C=cembio)")
```

Now filter to peaks present in at least 2 labs:

```{r filter-consensus}
# Filter: keep only peaks matched in at least 2 labs
consensus_matches_filtered <- consensus_matches |>
  filter(n_labs_matched >= 2)

cat("\n=== AFTER FILTERING (≥2 labs) ===\n")
cat("Consensus peaks retained:", nrow(consensus_matches_filtered), "\n")
cat("Percentage of detected peaks:", 
    round(100 * nrow(consensus_matches_filtered) / nrow(consensus_matches), 1), "%\n")
```

Build the final consensus peak table with chrom_peak_id mapping:

```{r build-consensus-table}
# Create consensus table: one row per unique peak across labs
consensus_peaks <- consensus_matches_filtered |>
  rowwise() |>
  mutate(
    # Consensus m/z and RT from all matched peaks
    consensus_mz = mean(c(
      pt_afekta$mzmin[idx_afekta],
      if (!is.na(idx_hmgu)) pt_hmgu$mzmin[idx_hmgu] else NA,
      if (!is.na(idx_icl)) pt_icl$mzmin[idx_icl] else NA,
      if (!is.na(idx_cembio)) pt_cembio$mzmin[idx_cembio] else NA
    ), na.rm = TRUE),
    consensus_rt = mean(c(
      pt_afekta$rtmin[idx_afekta],
      if (!is.na(idx_hmgu)) pt_hmgu$rtmin[idx_hmgu] else NA,
      if (!is.na(idx_icl)) pt_icl$rtmin[idx_icl] else NA,
      if (!is.na(idx_cembio)) pt_cembio$rtmin[idx_cembio] else NA
    ), na.rm = TRUE),
    # Map chrom_peak_id from each lab
    chrom_peak_id_afekta = if (!is.na(idx_afekta)) pt_afekta$chrom_peak_id[idx_afekta] else NA_character_,
    chrom_peak_id_hmgu = if (!is.na(idx_hmgu)) pt_hmgu$chrom_peak_id[idx_hmgu] else NA_character_,
    chrom_peak_id_icl = if (!is.na(idx_icl)) pt_icl$chrom_peak_id[idx_icl] else NA_character_,
    chrom_peak_id_cembio = if (!is.na(idx_cembio)) pt_cembio$chrom_peak_id[idx_cembio] else NA_character_
  ) |>
  ungroup() |>
  mutate(
    consensus_id = seq_len(n()),
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  ) |>
  select(consensus_id, consensus_mz, consensus_rt, n_labs_matched,
         afekta_detected, chrom_peak_id_afekta,
         hmgu_detected, chrom_peak_id_hmgu,
         icl_detected, chrom_peak_id_icl,
         cembio_detected, chrom_peak_id_cembio,
         everything())

head(consensus_peaks)
```

Save the consensus table for downstream analysis:

```{r save-consensus}
save(consensus_peaks, consensus_matches_filtered,
     file = file.path(base_path, "object/consensus_peaks.RData"))
write.csv(consensus_peaks,
          file = file.path(base_path, "object/consensus_peaks.csv"),
          row.names = FALSE)
```

## Summary Tables

### Overlap: Cembio vs Afekta

```{r cembio-afekta-overlap}
cembio_afekta_summary <- data.frame(
  category = c("Both detected", "Only afekta", "Only cembio", "Neither"),
  count = c(
    sum(consensus_matches$afekta_detected & consensus_matches$cembio_detected),
    sum(consensus_matches$afekta_detected & !consensus_matches$cembio_detected),
    sum(!consensus_matches$afekta_detected & consensus_matches$cembio_detected),
    sum(!consensus_matches$afekta_detected & !consensus_matches$cembio_detected)
  )
)

cembio_afekta_summary <- cembio_afekta_summary |>
  mutate(percent = round(100 * count / sum(count), 1))

knitr::kable(cembio_afekta_summary, 
             caption = "Peak Overlap: Cembio vs Afekta")
```

### Overlap: All 4 Labs

```{r all-labs-overlap}
all_labs_summary <- data.frame(
  detection_pattern = character(),
  n_peaks = integer(),
  percent = numeric()
)

for (combo in 0:15) {
  afekta <- as.logical(bitwAnd(combo, 1))
  hmgu <- as.logical(bitwAnd(combo, 2))
  icl <- as.logical(bitwAnd(combo, 4))
  cembio <- as.logical(bitwAnd(combo, 8))
  
  n <- sum(
    consensus_matches$afekta_detected == afekta &
    consensus_matches$hmgu_detected == hmgu &
    consensus_matches$icl_detected == icl &
    consensus_matches$cembio_detected == cembio
  )
  
  if (n > 0) {
    pattern <- paste0(
      if_else(afekta, "A", "-"),
      if_else(hmgu, "H", "-"),
      if_else(icl, "I", "-"),
      if_else(cembio, "C", "-")
    )
    
    all_labs_summary <- rbind(all_labs_summary, data.frame(
      detection_pattern = pattern,
      n_peaks = n,
      percent = round(100 * n / nrow(consensus_matches), 1)
    ))
  }
}

all_labs_summary <- all_labs_summary |>
  arrange(desc(n_peaks))

knitr::kable(all_labs_summary,
             caption = "Peak Detection Patterns: All 4 Labs (A=afekta, H=hmgu, I=icl, C=cembio)")
```

# Peak Feature Summary by Lab (Detected)

## Prepare subsets for downstream analysis

```{r prepare-subsets}
# Subset 1: Peaks present in BOTH cembio AND afekta
cembio_afekta_both <- consensus_matches |>
  filter(afekta_detected & cembio_detected)

cat("Peaks in both cembio & afekta:", nrow(cembio_afekta_both), "\n")

# Subset 2: Peaks present in ALL 4 labs
all_four_labs <- consensus_matches |>
  filter(afekta_detected & hmgu_detected & icl_detected & cembio_detected)

cat("Peaks in all 4 labs:", nrow(all_four_labs), "\n")
```

## Entropy

## Entropy

```{r entropy-summary}
split(full_peak_report_detected$entropy, full_peak_report_detected$lab) |>
  lapply(summary)
```

## FWHM

```{r fwhm-summary}
split(full_peak_report_detected$fwhm, full_peak_report_detected$lab) |>
  lapply(summary)
```

## Gaussian Similarity

```{r gaussian-summary}
split(full_peak_report_detected$gaussian_similarity, full_peak_report_detected$lab) |>
  lapply(summary)
```

## Tailing Factor

```{r tailing-summary}
split(full_peak_report_detected$tailing_factor, full_peak_report_detected$lab) |>
  lapply(summary)
```

# Downstream Peak Feature Analysis (Detected)

## RT Stability - Cembio vs Afekta (Both Labs)

```{r rt-stability-cembio-afekta}
# Filter full_peak_report_detected to only cembio & afekta peaks that are in both labs
cembio_afekta_peak_ids <- c(
  cembio_afekta_both$chrom_peak_id_afekta,
  cembio_afekta_both$chrom_peak_id_cembio
)

peak_report_cembio_afekta <- full_peak_report_detected |>
  filter(chrom_peak_id %in% cembio_afekta_peak_ids)

rt_report_ca <- peak_report_cembio_afekta |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report_ca, 10)

ggplot(rt_report_ca, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (Cembio & Afekta)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency - Cembio vs Afekta (Both Labs)

```{r intensity-consistency-cembio-afekta}
intensity_report_ca <- peak_report_cembio_afekta |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_ca, 10)

ggplot(intensity_report_ca, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (Cembio & Afekta)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## RT Stability - All 4 Labs

```{r rt-stability-all-four}
# Filter full_peak_report_detected to only peaks present in all 4 labs
all_four_peak_ids <- c(
  all_four_labs$chrom_peak_id_afekta,
  all_four_labs$chrom_peak_id_hmgu,
  all_four_labs$chrom_peak_id_icl,
  all_four_labs$chrom_peak_id_cembio
)

peak_report_all_four <- full_peak_report_detected |>
  filter(chrom_peak_id %in% all_four_peak_ids)

rt_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report_all4, 10)

ggplot(rt_report_all4, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (All 4 Labs)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency - All 4 Labs

```{r intensity-consistency-all-four}
intensity_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_all4, 10)

ggplot(intensity_report_all4, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (All 4 Labs)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Inter-Lab Statistical Comparison

```{r fwhm-analysis}
p1 <- ggviolin(full_peak_report_detected_meta, x = "lab", y = "fwhm", fill = "lab",
         add = "boxplot", add.params = list(fill = "white")) +
  geom_hline(yintercept = median(full_peak_report_detected_meta$fwhm), linetype = 2) +
  labs(title = "Peak Width (FWHM) by Lab (Detected)", y = "FWHM (s)", x = "Laboratory") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p1

kw_res <- full_peak_report_detected_meta %>% kruskal_test(fwhm ~ lab)
kw_res

eff_res <- full_peak_report_detected_meta %>% kruskal_effsize(fwhm ~ lab)
eff_res
```

## Hardware Impact

```{r lc-impact}
plot_data <- full_peak_report_detected_meta %>%
  filter(!is.na(lc_instrument_clean))

ggplot(plot_data, aes(x = lc_instrument_clean, y = fwhm, fill = lc_instrument_clean)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Impact of LC Instrument on Peak Width (Detected)", x = "LC System", y = "FWHM (s)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")
```

## PCA of Peak Features (Detected)

```{r pca-analysis}
pca_cols <- c("fwhm", "tailing_factor", "asymmetry", "entropy", "auc", "apex_intensity")
pca_input <- full_peak_report_detected_meta[, pca_cols]

pca_input$auc <- log10(pca_input$auc + 1)
pca_input$apex_intensity <- log10(pca_input$apex_intensity + 1)
pca_input <- na.omit(pca_input)

pca_res <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_scores <- as.data.frame(pca_res$x)
pca_scores$lab <- full_peak_report_detected_meta$lab[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$lc_type <- full_peak_report_detected_meta$lc_instrument_clean[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$mixture <- full_peak_report_detected_meta$mixture[rownames(pca_input) %in% rownames(pca_scores)]

var_expl <- summary(pca_res)$importance[2, 1:2] * 100

g1 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by Lab (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2])) +
  theme_bw()

g2 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lc_type)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by LC System (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2]),
       color = "LC System") +
  theme_bw()

ggarrange(g1, g2, ncol = 2, common.legend = FALSE)
```

## Conclusions (Detected)

Detected peak feature analysis complete. Results saved to `object/full_peak_report_detected.*`. 