---
title: "Peak Feature Calculation — Detected Peaks"
format:
  html:
    toc: true
execute:
  dir: project
---

# Introduction

This document performs peak feature calculation and analysis for the detected
ion chromatograms (EICs). It reuses the processed detected EICs, computes peak
metrics, and runs downstream comparisons and stability analyses.

- Use all detected peaks form the beginning, in the chromData have info wether they are annotated
  - only keep peaks overlapping the 4 labs to create a common feature set

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)
library(UpSetR)

base_path <- file.path("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
```

# Load Detected EICs

We are not computing the eic at the moment because that is justy TOO many peaks

Before anything let's load all the detected peaks table. 

The code of this part and the next one could be moved to the global analysis.
could use peak count in some part of the global analysis.
and moved the consensus table to do a spearman analysis.

```{r}
```

# Create Consensus Peak Table Across Labs

Below we will do a test analysis on mixture 1_1 (e.g. sample 1).

```{r build-consensus-peaks}
# the table will be generated on the cluster. and loaded below
# the parameter used for mclosest are: ppm tol = 10 and rt tol = 10
# 

# Load all consensus_all_sample*.csv files (excluding _edge.csv files)
consensus_dir <- file.path(base_path, "object/consensus_tables")
consensus_files <- list.files(consensus_dir, pattern = "^consensus_all_sample[0-9]+\\.csv$", full.names = TRUE)

ct_all_samples <- do.call(rbind, lapply(consensus_files, read.csv))
rownames(ct_all_samples) <- NULL

nrow(ct_all_samples) 
```

```{r}
## split per lab 
# match the total using the sample and chromPeak and add columns "into", "intb", "maxo" from the pt* of the respective lab
```


```{r}
# Add detection flags to consensus_matches
ct_all_samples <- ct_all_samples |>
  mutate(
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  )

cat(
  "Peaks matched in consensus (before filtering):",
  nrow(ct_all_samples),
  "\n\n"
)
cat("Peak counts by number of labs detected:\n")
print(table(ct_all_samples$n_labs_matched))
```

```{r}
lapply(split(ct_all_samples, ct_all_samples$sample), function(df) {
  print(paste("Sample:", unique(df$sample)))
  table(df$n_labs_matched)
}) ## would be nice to plot this properly. 
```

Look slike the nlab is fixed.

## Upset Plot - Lab Overlap

```{r upset-plot, message = FALSE, warning = FALSE}
# Prepare binary matrix for upset plot
upset_data <- ct_all_samples |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(
    afekta = as.integer(afekta_detected),
    hmgu = as.integer(hmgu_detected),
    icl = as.integer(icl_detected),
    cembio = as.integer(cembio_detected)
  ) |>
  select(afekta, hmgu, icl, cembio)

# Convert to data frame with rownames for upset function
upset_data_df <- as.data.frame(upset_data)
rownames(upset_data_df) <- paste0("peak_", seq_len(nrow(upset_data_df)))

# Only plot intersections with at least 1 peak (exclude empty combinations)
upset(upset_data_df,
      sets = c("afekta", "hmgu", "icl", "cembio"),
      main.bar.color = "steelblue",
      sets.bar.color = "darkred",
      text.scale = c(3,3, 3, 3, 3, 3), # reduce size of font for set size and number above bars
      point.size = 4,
      line.size = 1.2) ## need to fix set size 
```

## Pairwise Lab Comparisons

```{r pairwise-overlap}
# Create pairwise overlap summary
labs <- c("afekta", "hmgu", "icl", "cembio")
pairwise_summary <- data.frame()

for (i in seq_along(labs)) {
  for (j in seq_along(labs)) {
    if (i < j) {
      lab1 <- labs[i]
      lab2 <- labs[j]

      col1 <- paste0(lab1, "_detected")
      col2 <- paste0(lab2, "_detected")

      both <- sum(ct_all_samples[[col1]] & ct_all_samples[[col2]])
      only_lab1 <- sum(ct_all_samples[[col1]] & !ct_all_samples[[col2]])
      only_lab2 <- sum(!ct_all_samples[[col1]] & ct_all_samples[[col2]])

      total_lab1 <- sum(ct_all_samples[[col1]])
      total_lab2 <- sum(ct_all_samples[[col2]])
      
      pairwise_summary <- rbind(pairwise_summary, data.frame(
        lab1 = lab1,
        lab2 = lab2,
        both_detected = both,
        only_lab1 = only_lab1,
        only_lab2 = only_lab2,
        total_lab1 = total_lab1,
        total_lab2 = total_lab2,
        jaccard_index = round(100 * both / (both + only_lab1 + only_lab2), 1),
        overlap_percent_lab1 = round(100 * both / total_lab1, 1),
        overlap_percent_lab2 = round(100 * both / total_lab2, 1)
      ))
    }
  }
}

knitr::kable(pairwise_summary, caption = "Pairwise Lab Overlap - Detailed Metrics")
```

## Pairwise Overlap Visualization

```{r pairwise-plot}
# Reshape for visualization
pairwise_long <- pairwise_summary |>
  select(lab1, lab2, overlap_percent_lab1, overlap_percent_lab2) |>
  pivot_longer(
    cols = starts_with("overlap"),
    names_to = "metric",
    names_prefix = "overlap_percent_",
    values_to = "percent"
  ) |>
  mutate(
    pair = paste0(lab1, " ↔ ", lab2),
    direction = case_when(
      metric == "lab1" ~ paste0(lab1, " → ", lab2),
      metric == "lab2" ~ paste0(lab2, " → ", lab1)
    )
  )

# Heatmap-style plot
ggplot(pairwise_long, aes(x = pair, y = direction, fill = percent)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = paste0(percent, "%")), size = 3, color = "black") +
  scale_fill_gradient(low = "#fee5d9", high = "#a1d99b", limits = c(0, 100), name = "Overlap %") +
  labs(
    title = "Pairwise Peak Overlap Between Labs",
    subtitle = "Shows % of peaks from each lab found in the other lab",
    x = "Lab Pair",
    y = "Direction"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

# Bar plot comparing overlaps
ggplot(pairwise_long, aes(x = direction, y = percent, fill = metric)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap(~pair, scales = "free_x", nrow = 2) +
  labs(
    title = "Asymmetric Pairwise Overlaps",
    subtitle = "Shows directional recovery: % of one lab's peaks detected in the other",
    x = "Direction",
    y = "Overlap %",
    fill = "Reference Lab"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  ylim(0, 100)
```

As expected, afekta and cembio have the highest overlap.


Could remove column 3:6 in table above
Below decide on the threshold to study a peak. 
we could go with 4

```{r}
thres <- 4

ct_filtered <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Consensus peaks retained:", nrow(ct_filtered), "\n")
cat(
  "Percentage of detected peaks:",
  round(100 * nrow(ct_filtered) / nrow(ct_all_samples), 1),
  "%\n"
)
```

Build the final consensus peak table with chrom_peak_id mapping:
Save the consensus table for downstream analysis:

```{r save-consensus}
write.csv(
  ct_all_samples,
  file = file.path(base_path, "object/ct_all_samples.csv"),
  row.names = FALSE
)
write.csv(
  ct_filtered,
  file = file.path(base_path, "object/ct_filtered4.csv"),
  row.names = FALSE
)
```

## Summary Tables

### Overlap: Cembio vs Afekta

```{r cembio-afekta-overlap}
cembio_afekta_summary <- data.frame(
  category = c("Both detected", "Only afekta", "Only cembio", "Neither"),
  count = c(
    sum(ct_2samples$afekta_detected & ct_2samples$cembio_detected),
    sum(ct_2samples$afekta_detected & !ct_2samples$cembio_detected),
    sum(!ct_2samples$afekta_detected & ct_2samples$cembio_detected),
    sum(!ct_2samples$afekta_detected & !ct_2samples$cembio_detected)
  )
)

cembio_afekta_summary <- cembio_afekta_summary |>
  mutate(percent = round(100 * count / sum(count), 1))

knitr::kable(cembio_afekta_summary,
             caption = "Peak Overlap: Cembio vs Afekta")
```

so we could look at 2526 detected peaks ( for 2 mixture), that's really nice. 

### Overlap: All 4 Labs

```{r all-labs-overlap}
all_labs_summary <- data.frame(
  detection_pattern = character(),
  n_peaks = integer(),
  percent = numeric()
)

for (combo in 0:15) {
  afekta <- as.logical(bitwAnd(combo, 1))
  hmgu <- as.logical(bitwAnd(combo, 2))
  icl <- as.logical(bitwAnd(combo, 4))
  cembio <- as.logical(bitwAnd(combo, 8))

  n <- sum(
    ct_2samples$afekta_detected == afekta &
    ct_2samples$hmgu_detected == hmgu &
    ct_2samples$icl_detected == icl &
    ct_2samples$cembio_detected == cembio
  )

  if (n > 0) {
    pattern <- paste0(
      if_else(afekta, "A", "-"),
      if_else(hmgu, "H", "-"),
      if_else(icl, "I", "-"),
      if_else(cembio, "C", "-")
    )

    all_labs_summary <- rbind(all_labs_summary, data.frame(
      detection_pattern = pattern,
      n_peaks = n,
      percent = round(100 * n / nrow(consensus_matches_1), 1)
    ))
  }
}

all_labs_summary <- all_labs_summary |>
  arrange(desc(n_peaks))

knitr::kable(all_labs_summary,
             caption = "Peak Detection Patterns: All 4 Labs (A=afekta, H=hmgu, I=icl, C=cembio)")
```

i think the code above could replace 2 code chunk above it. 

# Peak Feature Summary by Lab (Detected)

## Prepare subsets for downstream analysis

```{r prepare-subsets}
# Subset 1: Peaks present in BOTH cembio AND afekta
cembio_afekta_both <- ct_all_samples |>
  filter(afekta_detected & cembio_detected) ## keep the id for cembio and afekta

cat("Peaks in both cembio & afekta:", nrow(cembio_afekta_both), "\n")

# Subset 2: Peaks present in ALL 4 labs
all_four_labs <- ct_all_samples |>
  filter(afekta_detected & hmgu_detected & icl_detected & cembio_detected)

cat("Peaks in all 4 labs:", nrow(all_four_labs), "\n")
```

All the part above could be cleaned up for redundancy.
the afekta/cembio overlap is insane. 51000 peaks to analyze...

# Creating eics
## 4 labs

when i have the full table i will need to add the "sample" info here. 
```{r}
pt <- all_four_labs |>
  mutate(consensus_id = row_number()) |>
  select(consensus_id, sample, chrom_peak_id_afekta, chrom_peak_id_hmgu, 
  chrom_peak_id_icl, chrom_peak_id_cembio) |>
  pivot_longer(
    cols = c(chrom_peak_id_afekta, chrom_peak_id_hmgu, chrom_peak_id_icl, 
    chrom_peak_id_cembio),
    names_to = "lab",
    names_prefix = "chrom_peak_id_",
    values_to = "chrom_peak_id"
  ) |>
  filter(!is.na(chrom_peak_id))
```

Get mzmin, mzmax, rtmin, rtmax info from pt_* of each lab USING THE chrom_peak_id column of the peak.table and the lab column  T between the 2 table. probably need to do it sequnetially for eahc lab

```{r}
# Add mzmin, mzmax, rtmin, rtmax to peak.table by joining with each lab's pt_* table
pt <- pt |>
  mutate(
    mzMin = case_when(
      lab == "afekta" ~ pt_afekta$mzmin[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mzmin[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mzmin[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mzmin[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    mzMax = case_when(
      lab == "afekta" ~ pt_afekta$mzmax[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mzmax[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mzmax[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mzmax[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    rtMin = case_when(
      lab == "afekta" ~ pt_afekta$rtmin[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$rtmin[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$rtmin[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$rtmin[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    rtMax = case_when(
      lab == "afekta" ~ pt_afekta$rtmax[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$rtmax[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$rtmax[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$rtmax[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    mz = case_when(
      lab == "afekta" ~ pt_afekta$mz[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mz[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mz[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mz[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    polarity = case_when(
      lab == "afekta" ~ pt_afekta$polarity[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$polarity[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$polarity[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$polarity[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ), 
    mixture = case_when(
      lab == "afekta" ~ pt_afekta$mixture[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mixture[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mixture[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mixture[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    )
    into = case_when(
      lab == "afekta" ~ pt_afekta$into[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$into[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$into[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$into[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )],
    intb = case_when(
      lab == "afekta" ~ pt_afekta$intb[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$intb[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$intb[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$intb[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )],
    maxo = case_when(
      lab == "afekta" ~ pt_afekta$maxo[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$maxo[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$maxo[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$maxo[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    )
  )

write.csv(
  pt,
  file = file.path(base_path, "object/peak_table_detected_all4labs.csv"),
  row.names = FALSE
#rename rtmin ot rtMin, .. and add/remove 5s to each rt size and 0.001 to each mz side

pt <- pt |>
  mutate(
    rtMin = rtMin - 5,
    rtMax = rtMax + 5,
    mzMin = mzMin - 0.001,
    mzMax = mzMax + 0.001
  )

pt$polarity[pt$sample %in% 1:19] <- 1L
pt$polarity[pt$sample %in% 20:38] <- 0L
pt 
```




```{r}
library(Chromatograms)

if (!exists("object/eics_detected_all4labs.RData")) {
    load(file.path(base_path, "object/sp_full.RData"))
    tic_full <- Chromatograms(sp_full, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab"))
    eics <- chromExtract(tic_full, peak.table = pt, by = c("polarity", "mixture", "lab"))
    nrow(pt) == length(eics)
    eics <- setBackend(eics, ChromBackendMemory())
    save(eics, file = file.path(base_path, "object/eics_detected_all4labs.RData"))
} else {
    load(file.path(base_path, "object/eics_detected_all4labs.RData"))
}
```


```{r}
plotChromatograms(eics[eics$consensus_id == 10],
main= eics$lab[eics$consensus_id == 10]) #nb of datapoint change is quite high. 


plotChromatograms(eics[eics$consensus_id == 1204],
main= eics$lab[eics$consensus_id == 1204]) #a good example

## ok there is something weird with the id 1000 
peaksData(eics[eics$consensus_id == 1000]) ## ok something is weird. I need to 
# check if maybe this was there before i move the data to mem. or its a 
# splitting issue, anyway that is very weird.
```

Ok next do that on all the table and build a big object table. 
but continue testing the metrics calculation just with 2 samples.


# Downstream Peak Feature Analysis (Detected)

## RT Stability - Cembio vs Afekta (Both Labs)

```{r rt - stability - cembio - afekta}
# Filter full_peak_report_detected to only cembio & afekta peaks that are in both labs

cembio_afekta_peak_ids <- c(
  cembio_afekta_both$afekta,
  cembio_afekta_both$cembio
)

peak_report_cembio_afekta <- full_peak_report_detected |>
  filter(chrom_peak_id %in% cembio_afekta_peak_ids)

rt_report_ca <- peak_report_cembio_afekta |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report_ca, 10)

ggplot(rt_report_ca, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (Cembio & Afekta)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency - Cembio vs Afekta (Both Labs)

```{r intensity-consistency-cembio-afekta}
intensity_report_ca <- peak_report_cembio_afekta |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_ca, 10)

ggplot(intensity_report_ca, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (Cembio & Afekta)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## RT Stability - All 4 Labs

```{r rt-stability-all-four}
# Filter full_peak_report_detected to only peaks present in all 4 labs

all_four_peak_ids <- c(
  all_four_labs$afekta,
  all_four_labs$hmgu,
  all_four_labs$icl,
  all_four_labs$cembio
)

peak_report_all_four <- full_peak_report_detected |>
  filter(chrom_peak_id %in% all_four_peak_ids)

rt_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report_all4, 10)

ggplot(rt_report_all4, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (All 4 Labs)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency - All 4 Labs

```{r intensity-consistency-all-four}
intensity_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_all4, 10)

ggplot(intensity_report_all4, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (All 4 Labs)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Inter-Lab Statistical Comparison

```{r fwhm-analysis}
p1 <- ggviolin(full_peak_report_detected_meta, x = "lab", y = "fwhm", fill = "lab",
         add = "boxplot", add.params = list(fill = "white")) +
  geom_hline(yintercept = median(full_peak_report_detected_meta$fwhm), linetype = 2) +
  labs(title = "Peak Width (FWHM) by Lab (Detected)", y = "FWHM (s)", x = "Laboratory") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p1

kw_res <- full_peak_report_detected_meta %>% kruskal_test(fwhm ~ lab)
kw_res

eff_res <- full_peak_report_detected_meta %>% kruskal_effsize(fwhm ~ lab)
eff_res
```

## Hardware Impact

```{r lc-impact}
plot_data <- full_peak_report_detected_meta %>%
  filter(!is.na(lc_instrument_clean))

ggplot(plot_data, aes(x = lc_instrument_clean, y = fwhm, fill = lc_instrument_clean)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Impact of LC Instrument on Peak Width (Detected)", x = "LC System", y = "FWHM (s)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")
```

## PCA of Peak Features (Detected)

```{r pca-analysis}
pca_cols <- c("fwhm", "tailing_factor", "asymmetry", "entropy", "auc", "apex_intensity")
pca_input <- full_peak_report_detected_meta[, pca_cols]

pca_input$auc <- log10(pca_input$auc + 1)
pca_input$apex_intensity <- log10(pca_input$apex_intensity + 1)
pca_input <- na.omit(pca_input)

pca_res <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_scores <- as.data.frame(pca_res$x)
pca_scores$lab <- full_peak_report_detected_meta$lab[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$lc_type <- full_peak_report_detected_meta$lc_instrument_clean[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$mixture <- full_peak_report_detected_meta$mixture[rownames(pca_input) %in% rownames(pca_scores)]

var_expl <- summary(pca_res)$importance[2, 1:2] * 100

g1 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by Lab (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2])) +
  theme_bw()

g2 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lc_type)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by LC System (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2]),
       color = "LC System") +
  theme_bw()

ggarrange(g1, g2, ncol = 2, common.legend = FALSE)
```

## Conclusions (Detected)

Detected peak feature analysis complete. Results saved to `object/full_peak_report_detected.*`.
