---
title: "Peak Feature Calculation — Detected Peaks"
format:
  html:
    toc: true
execute:
  dir: project
---

# Introduction

This document performs peak feature calculation and analysis for the detected
ion chromatograms (EICs). It reuses the processed detected EICs, computes peak
metrics, and runs downstream comparisons and stability analyses.

- Use all detected peaks form the beginning, in the chromData have info wether they are annotated
  - only keep peaks overlapping the 4 labs to create a common feature set

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)
library(UpSetR)

base_path <- file.path("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
source(file.path("1_preprocessing/setup.R"))
```


# Create Consensus Peak Table Across Labs

```{r}
pt_afekta <- read.csv(file.path(
  base_path,
  "object/detected_peaks_afekta_HE.csv"
))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(
  base_path,
  "object/detected_peaks_cembio_HE.csv"
))
lab_counts <- c(
  afekta = nrow(pt_afekta),
  hmgu = nrow(pt_hmgu),
  icl = nrow(pt_icl),
  cembio = nrow(pt_cembio)
)
barplot(
  lab_counts,
  main = "Number of Detected Peaks per Lab",
  ylab = "Number of detected peaks",
  xlab = "Labs",
  col = "steelblue"
)
```

```{r build-consensus-peaks}
# the table will be generated on the cluster. and loaded below
# the parameter used for mclosest are: ppm tol = 10 and rt tol = 10
# Load all consensus_all_sample*.csv files (excluding _edge.csv files)
consensus_dir <- file.path(base_path, "object/consensus_tables")
consensus_files <- list.files(consensus_dir,
 pattern = "^consensus_all_sample[0-9]+\\.csv$", full.names = TRUE)

ct_all_samples <- do.call(rbind, lapply(consensus_files, read.csv))
rownames(ct_all_samples) <- NULL

nrow(ct_all_samples) 
```


```{r}
# Add detection flags to consensus_matches
ct_all_samples <- ct_all_samples |>
  mutate(
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  )

cat("Peak counts by number of labs detected:\n")
print(table(ct_all_samples$n_labs_matched))
```

```{r}
# Replace mixture column from 1.1 to 1_1
meta$mixture <- gsub("\\.", "_", meta$Mixture)
nc_mix <- table(meta$mixture)
# This shows the number of compounds per mixture.

# Create mapping of sample to mixture name using pt_* data
sample_to_mixture <- pt_afekta |>
  select(sample, mixture) |>
  distinct()

# Below we can summarize the number of peaks detected per sample with compound count
mixture_summary <- lapply(split(ct_all_samples, ct_all_samples$sample), function(df) {
  sample_id <- unique(df$sample)
  mixture_name <- sample_to_mixture$mixture[sample_to_mixture$sample == sample_id]
  
  res <- data.frame(
    sample = sample_id,
    mixture = mixture_name,
    detected_4labs = sum(df$n_labs_matched == 4),
    detected_3labs = sum(df$n_labs_matched == 3),
    detected_2labs = sum(df$n_labs_matched == 2),
    detected_1lab = sum(df$n_labs_matched == 1),
    total_peaks = nrow(df),
    n_compounds = as.integer(nc_mix[mixture_name])
  )
  res
})

mixture_summary_df <- do.call(rbind, mixture_summary)
rownames(mixture_summary_df) <- NULL

knitr::kable(mixture_summary_df, caption = "Peak Detection Summary by Mixture")
```


## Upset Plot - Lab Overlap

```{r upset-plot, message = FALSE, warning = FALSE}
# Prepare binary matrix for upset plot
upset_data <- ct_all_samples |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(
    afekta = as.integer(afekta_detected),
    hmgu = as.integer(hmgu_detected),
    icl = as.integer(icl_detected),
    cembio = as.integer(cembio_detected)
  ) |>
  select(afekta, hmgu, icl, cembio)

# Convert to data frame with rownames for upset function
upset_data_df <- as.data.frame(upset_data)
rownames(upset_data_df) <- paste0("peak_", seq_len(nrow(upset_data_df)))

# Only plot intersections with at least 1 peak (exclude empty combinations)
upset(upset_data_df,
      sets = c("afekta", "hmgu", "icl", "cembio"),
      main.bar.color = "steelblue",
      sets.bar.color = "darkred",
      text.scale = c(2, 2, 2, 1.5, 2, 2),
      point.size = 3,
      line.size = 1.2) 
```

## Pairwise Lab Comparisons

```{r pairwise-overlap}
# Create pairwise overlap summary
labs <- c("afekta", "hmgu", "icl", "cembio")
pairwise_summary <- data.frame()

for (i in seq_along(labs)) {
  for (j in seq_along(labs)) {
    if (i < j) {
      lab1 <- labs[i]
      lab2 <- labs[j]

      col1 <- paste0(lab1, "_detected")
      col2 <- paste0(lab2, "_detected")

      both <- sum(ct_all_samples[[col1]] & ct_all_samples[[col2]])
      only_lab1 <- sum(ct_all_samples[[col1]] & !ct_all_samples[[col2]])
      only_lab2 <- sum(!ct_all_samples[[col1]] & ct_all_samples[[col2]])

      total_lab1 <- sum(ct_all_samples[[col1]])
      total_lab2 <- sum(ct_all_samples[[col2]])
      
      pairwise_summary <- rbind(pairwise_summary, data.frame(
        lab1 = lab1,
        lab2 = lab2,
        both_detected = both,
        only_lab1 = only_lab1,
        only_lab2 = only_lab2,
        total_lab1 = total_lab1,
        total_lab2 = total_lab2,
        jaccard_index = round(100 * both / (both + only_lab1 + only_lab2), 1),
        overlap_percent_lab1 = round(100 * both / total_lab1, 1),
        overlap_percent_lab2 = round(100 * both / total_lab2, 1)
      ))
    }
  }
}

knitr::kable(pairwise_summary, caption = "Pairwise Lab Overlap - Detailed Metrics")
```

## Pairwise Overlap Visualization

```{r pairwise-plot}
# Reshape for visualization
pairwise_long <- pairwise_summary |>
  select(lab1, lab2, overlap_percent_lab1, overlap_percent_lab2) |>
  pivot_longer(
    cols = starts_with("overlap"),
    names_to = "metric",
    names_prefix = "overlap_percent_",
    values_to = "percent"
  ) |>
  mutate(
    pair = paste0(lab1, " ↔ ", lab2),
    direction = case_when(
      metric == "lab1" ~ paste0(lab1, " → ", lab2),
      metric == "lab2" ~ paste0(lab2, " → ", lab1)
    )
  )

# Bar plot comparing overlaps
ggplot(pairwise_long, aes(x = direction, y = percent, fill = metric)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap(~pair, scales = "free_x", nrow = 2) +
  labs(
    title = "Asymmetric Pairwise Overlaps",
    subtitle = "Shows directional recovery: % of one lab's peaks detected in the other",
    x = "Direction",
    y = "Overlap %",
    fill = "Reference Lab"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  ylim(0, 100)
```

As expected, afekta and cembio have the highest overlap.

```{r}
thres <- 4

ct_filtered <- ct_all_samples |>
  filter(n_labs_matched == 4)

cat("Consensus peaks retained:", nrow(ct_filtered), "\n")
cat(
  "Percentage of detected peaks:",
  round(100 * nrow(ct_filtered) / nrow(ct_all_samples), 1),
  "%\n"
)
```

Build the final consensus peak table with chrom_peak_id mapping:
Save the consensus table for downstream analysis:

```{r save-consensus}
write.csv(
  ct_all_samples,
  file = file.path(base_path, "object/ct_all_samples.csv"),
  row.names = FALSE
)
write.csv(
  ct_filtered,
  file = file.path(base_path, "object/ct_filtered4.csv"),
  row.names = FALSE
)
```

## Summary Tables

### Overlap: Cembio vs Afekta

```{r cembio-afekta-overlap}
cembio_afekta_summary <- data.frame(
  category = c("Both detected", "Only afekta", "Only cembio", "Neither"),
  count = c(
    sum(ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & ct_all_samples$cembio_detected),
    sum(!ct_all_samples$afekta_detected & !ct_all_samples$cembio_detected)
  )
)

cembio_afekta_summary <- cembio_afekta_summary |>
  mutate(percent = round(100 * count / sum(count), 1))

knitr::kable(cembio_afekta_summary,
             caption = "Peak Overlap: Cembio vs Afekta")
```

so we could look at 2526 detected peaks ( for 2 mixture), that's really nice. 

### Overlap: All 4 Labs

```{r all-labs-overlap}
all_labs_summary <- data.frame(
  detection_pattern = character(),
  n_peaks = integer(),
  percent = numeric()
)

for (combo in 0:15) {
  afekta <- as.logical(bitwAnd(combo, 1))
  hmgu <- as.logical(bitwAnd(combo, 2))
  icl <- as.logical(bitwAnd(combo, 4))
  cembio <- as.logical(bitwAnd(combo, 8))

  n <- sum(
    ct_all_samples$afekta_detected == afekta &
    ct_all_samples$hmgu_detected == hmgu &
    ct_all_samples$icl_detected == icl &
    ct_all_samples$cembio_detected == cembio
  )

  if (n > 0) {
    pattern <- paste0(
      if_else(afekta, "A", "-"),
      if_else(hmgu, "H", "-"),
      if_else(icl, "I", "-"),
      if_else(cembio, "C", "-")
    )

    all_labs_summary <- rbind(all_labs_summary, data.frame(
      detection_pattern = pattern,
      n_peaks = n,
      percent = round(100 * n / nrow(ct_all_samples), 1)
    ))
  }
}

all_labs_summary <- all_labs_summary |>
  arrange(desc(n_peaks))

knitr::kable(all_labs_summary,
             caption = "Peak Detection Patterns: All 4 Labs (A=afekta, H=hmgu, I=icl, C=cembio)")
```



# Peak Feature Summary by Lab (Detected)

## Prepare subsets for downstream analysis

```{r prepare-subsets}
# Subset 1: Peaks present in BOTH cembio AND afekta
cembio_afekta_both <- ct_all_samples |>
  filter(afekta_detected & cembio_detected) ## keep the id for cembio and afekta

cat("Peaks in both cembio & afekta:", nrow(cembio_afekta_both), "\n")

# Subset 2: Peaks present in ALL 4 labs
all_four_labs <- ct_all_samples |>
  filter(afekta_detected & hmgu_detected & icl_detected & cembio_detected)

cat("Peaks in all 4 labs:", nrow(all_four_labs), "\n")
```

All the part above could be cleaned up for redundancy.
the afekta/cembio overlap is insane. 51000 peaks to analyze...

# Creating eics
## 4 labs
 
```{r}
pt <- all_four_labs |>
  mutate(consensus_id = row_number()) |>
  select(consensus_id, sample, chrom_peak_id_afekta, chrom_peak_id_hmgu, 
  chrom_peak_id_icl, chrom_peak_id_cembio) |>
  pivot_longer(
    cols = c(chrom_peak_id_afekta, chrom_peak_id_hmgu, chrom_peak_id_icl, 
    chrom_peak_id_cembio),
    names_to = "lab",
    names_prefix = "chrom_peak_id_",
    values_to = "chrom_peak_id"
  ) |>
  filter(!is.na(chrom_peak_id))
```

Get mzmin, mzmax, rtmin, rtmax info from pt_* of each lab USING THE chrom_peak_id column of the peak.table and the lab column  T between the 2 table. probably need to do it sequnetially for eahc lab

```{r}
# Add mzmin, mzmax, rtmin, rtmax to peak.table by joining with each lab's pt_* table
pt <- pt |>
  mutate(
    mzMin = case_when(
      lab == "afekta" ~ pt_afekta$mzmin[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mzmin[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mzmin[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mzmin[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    mzMax = case_when(
      lab == "afekta" ~ pt_afekta$mzmax[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mzmax[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mzmax[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mzmax[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    rtMin = case_when(
      lab == "afekta" ~ pt_afekta$rtmin[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$rtmin[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$rtmin[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$rtmin[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    rtMax = case_when(
      lab == "afekta" ~ pt_afekta$rtmax[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$rtmax[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$rtmax[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$rtmax[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    mz = case_when(
      lab == "afekta" ~ pt_afekta$mz[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mz[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mz[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mz[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    polarity = case_when(
      lab == "afekta" ~ pt_afekta$polarity[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$polarity[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$polarity[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$polarity[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ), 
    mixture = case_when(
      lab == "afekta" ~ pt_afekta$mixture[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mixture[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mixture[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mixture[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    into = case_when(
      lab == "afekta" ~ pt_afekta$into[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$into[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$into[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$into[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    intb = case_when(
      lab == "afekta" ~ pt_afekta$intb[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$intb[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$intb[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$intb[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    maxo = case_when(
      lab == "afekta" ~ pt_afekta$maxo[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$maxo[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$maxo[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$maxo[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    )
  )

write.csv(
  pt,
  file = file.path(base_path, "object/peak_table_detected_all4labs.csv"),
  row.names = FALSE
)

# Rename rtmin to rtMin, rtmax to rtMax and adjust boundaries
# add/remove 5s to each rt size and 0.001 to each mz side

pt <- pt |>
  mutate(
    rtMin = rtMin - 3,
    rtMax = rtMax + 3,
    mzMin = mzMin - 0.001,
    mzMax = mzMax + 0.001
  )

pt$polarity[pt$sample %in% 1:19] <- 1L
pt$polarity[pt$sample %in% 20:38] <- 0L
pt 
```




```{r}
library(Chromatograms)

if (!file.exists(file.path(base_path, "object/eics_detected_all4labs.RData"))) {
    load(file.path(base_path, "object/sp_full.RData"))
    tic_full <- Chromatograms(sp_full, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab"))
    eics <- chromExtract(tic_full, peak.table = pt, by = c("polarity", "mixture", "lab"))
    nrow(pt) == length(eics)
    eics <- setBackend(eics, ChromBackendMemory())
    save(eics, file = file.path(base_path, "object/eics_detected_all4labs.RData"))
} else {
    load(file.path(base_path, "object/eics_detected_all4labs.RData"))
}

eics
```


```{r}
plotChromatograms(eics[eics$consensus_id == 10],
main= eics$lab[eics$consensus_id == 10]) #nb of datapoint change is quite high. 

plotChromatograms(eics[eics$consensus_id == 1204],
main= eics$lab[eics$consensus_id == 1204]) #a good example

## ok there is something weird with the id 1000 
peaksData(eics[eics$consensus_id == 1000]) ## ok something is weird. I need to 
# check if maybe this was there before i move the data to mem. or its a 
# splitting issue, anyway that is very weird.
```

Ok next do that on all the table and build a big object table. 
but continue testing the metrics calculation just with 2 samples.

TODO check the issue with the NA values. 

## Compute metrics

```{r}
eic1 <- eics[eics$consensus_id == 1204]

## for this we will use my curretn implementation in MsQuality
#BiocManager::install("philouail/MsQuality", ref = "Chromatograms")
library(MsQuality)
metric <- c("xicFwhm") ## change this for classic fwhm calculation.
## Maybe also do duration above threshold, number of datapoints, etc.
## could have a generic function in chr that gets the peak boundaries, does fwhm, etc.
## while also reusing the MSQuality things as much as possible.
## peak 

test <- calculateMetrics(eic1, metrics = metric)
rownames(test) <- eic1$lab
qualityMetrics(eics)
```

Ok so we have one metric for one peak.
Todo, check what is this metric giving us. why is it quartile. i feel like i 
should just get one fwhm value for a peak
and I also want to know how is the peak detected to get this metric.

# Downstream Peak Feature Analysis (Detected)


## Intensity Consistency - Cembio vs Afekta (Both Labs)

```{r intensity-consistency-cembio-afekta}
intensity_report_ca <- peak_report_cembio_afekta |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_ca, 10)

ggplot(intensity_report_ca, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (Cembio & Afekta)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## RT Stability - All 4 Labs

```{r rt-stability-all-four}
# Filter full_peak_report_detected to only peaks present in all 4 labs

all_four_peak_ids <- c(
  all_four_labs$afekta,
  all_four_labs$hmgu,
  all_four_labs$icl,
  all_four_labs$cembio
)

peak_report_all_four <- full_peak_report_detected |>
  filter(chrom_peak_id %in% all_four_peak_ids)

rt_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report_all4, 10)

ggplot(rt_report_all4, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (All 4 Labs)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency - All 4 Labs

```{r intensity-consistency-all-four}
intensity_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_all4, 10)

ggplot(intensity_report_all4, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (All 4 Labs)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Inter-Lab Statistical Comparison

```{r fwhm-analysis}
p1 <- ggviolin(full_peak_report_detected_meta, x = "lab", y = "fwhm", fill = "lab",
         add = "boxplot", add.params = list(fill = "white")) +
  geom_hline(yintercept = median(full_peak_report_detected_meta$fwhm), linetype = 2) +
  labs(title = "Peak Width (FWHM) by Lab (Detected)", y = "FWHM (s)", x = "Laboratory") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p1

kw_res <- full_peak_report_detected_meta %>% kruskal_test(fwhm ~ lab)
kw_res

eff_res <- full_peak_report_detected_meta %>% kruskal_effsize(fwhm ~ lab)
eff_res
```

## Hardware Impact

```{r lc-impact}
plot_data <- full_peak_report_detected_meta %>%
  filter(!is.na(lc_instrument_clean))

ggplot(plot_data, aes(x = lc_instrument_clean, y = fwhm, fill = lc_instrument_clean)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Impact of LC Instrument on Peak Width (Detected)", x = "LC System", y = "FWHM (s)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")
```

## PCA of Peak Features (Detected)

```{r pca-analysis}
pca_cols <- c("fwhm", "tailing_factor", "asymmetry", "entropy", "auc", "apex_intensity")
pca_input <- full_peak_report_detected_meta[, pca_cols]

pca_input$auc <- log10(pca_input$auc + 1)
pca_input$apex_intensity <- log10(pca_input$apex_intensity + 1)
pca_input <- na.omit(pca_input)

pca_res <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_scores <- as.data.frame(pca_res$x)
pca_scores$lab <- full_peak_report_detected_meta$lab[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$lc_type <- full_peak_report_detected_meta$lc_instrument_clean[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$mixture <- full_peak_report_detected_meta$mixture[rownames(pca_input) %in% rownames(pca_scores)]

var_expl <- summary(pca_res)$importance[2, 1:2] * 100

g1 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by Lab (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2])) +
  theme_bw()

g2 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lc_type)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by LC System (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2]),
       color = "LC System") +
  theme_bw()

ggarrange(g1, g2, ncol = 2, common.legend = FALSE)
```

## Conclusions (Detected)

Detected peak feature analysis complete. Results saved to `object/full_peak_report_detected.*`.
