---
title: "Peak Feature Calculation — Detected Peaks"
format:
  html:
    toc: true
execute:
  dir: project
---

# Introduction

This document performs peak feature calculation and analysis for the detected
ion chromatograms (EICs). It reuses the processed detected EICs, computes peak
metrics, and runs downstream comparisons and stability analyses.

- Use all detected peaks form the beginning, in the chromData have info wether they are annotated
  - only keep peaks overlapping the 4 labs to create a common feature set

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)
library(UpSetR)

base_path <- file.path("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
```

# Load Detected EICs

We are not computing the eic at the moment because that is justy TOO many peaks

Before anything let's load all the detected peaks table

```{r}
pt_afekta <- read.csv(file.path(
  base_path,
  "object/detected_peaks_afekta_HE.csv"
))
pt_hmgu <- read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv"))
pt_icl <- read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv"))
pt_cembio <- read.csv(file.path(
  base_path,
  "object/detected_peaks_cembio_HE.csv"
))

# Plot number of detected peaks per lab
lab_counts <- c(
  afekta = nrow(pt_afekta),
  hmgu = nrow(pt_hmgu),
  icl = nrow(pt_icl),
  cembio = nrow(pt_cembio)
)

barplot(
  lab_counts,
  main = "Number of detected peaks per lab",
  ylab = "Number of detected peaks",
  xlab = "Labs",
  col = "steelblue"
)
```

# Create Consensus Peak Table Across Labs

Below we will do a test analysis on mixture 1_1 (e.g. sample 1).

```{r build-consensus-peaks}
# the table will be generated on the cluster. and loaded below
# the parameter used for mclosest are: ppm tol = 10 and rt tol = 10
# 

# in the meantime i need an example of the table structure to build the rest
# of the code.

# Example structure of the consensus_matches table, with nas too
# consensus_matches <- data.frame(
#   idx_afekta = c(12, 34, 56, 78, 91, 101, 202, 303, 404, 501, 502, 503, 504, NA, NA, 601, 602, 603, 604, 605),
#   idx_hmgu   = c(23, 35, 57, 79, 92, 110, 210, 304, 405, 511, NA, NA, NA, 89, 90, 611, 612, NA, NA, NA),
#   idx_icl    = c(34, 36, 58, 80, 93, 120, 220, 305, 406, 521, 522, NA, NA, 99, NA, 621, NA, 623, 624, NA),
#   idx_cembio = c(56, 37, 59, 81, 94, 130, 230, 306, 407, 531, 532, 533, NA, 109, 110, 641, 642, 643, NA, NA)
# )

consensus_matches_1 <- read.csv(file.path(base_path, "object/consensus_tables/consensus_all_sample1.csv"))

head(consensus_matches_1)

nrow(consensus_matches_1) 
```

The chrompeak_id are already there, as well as the n_lab_matched

```{r}
# Analyze lab overlap before filtering

## Summary statistics

# Add detection flags to consensus_matches
consensus_matches_1 <- consensus_matches_1 |>
  mutate(
    afekta_detected = !is.na(idx_afekta),
    hmgu_detected = !is.na(idx_hmgu),
    icl_detected = !is.na(idx_icl),
    cembio_detected = !is.na(idx_cembio)
  )

cat(
  "Peaks matched in consensus (before filtering):",
  nrow(consensus_matches_1),
  "\n\n"
)
cat("Peak counts by number of labs detected:\n")
print(table(consensus_matches_1$n_labs_matched))
```

mm very little overlap for all 4 labs. CCould consider increasing the mz and rt ranges.
should I increase the ppm tol ? Discuss this


## Upset Plot - Lab Overlap

```{r upset-plot, message = FALSE, warning = FALSE}
library(UpSetR)

# Prepare binary matrix for upset plot
upset_data <- consensus_matches_1 |>
  select(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  mutate(
    afekta = as.integer(afekta_detected),
    hmgu = as.integer(hmgu_detected),
    icl = as.integer(icl_detected),
    cembio = as.integer(cembio_detected)
  ) |>
  select(afekta, hmgu, icl, cembio)

# Convert to data frame with rownames for upset function
upset_data_df <- as.data.frame(upset_data)
rownames(upset_data_df) <- paste0("peak_", seq_len(nrow(upset_data_df)))

# Only plot intersections with at least 1 peak (exclude empty combinations)
upset(upset_data_df,
      sets = c("afekta", "hmgu", "icl", "cembio"),
      order.by = "freq",
      empty.intersections = "on",
      main.bar.color = "steelblue",
      sets.bar.color = "darkred",
      text.scale = c(1.5, 1.5, 1.5, 1, 1.5, 1),
      point.size = 4,
      line.size = 1.2)
```

## Pairwise Lab Comparisons

```{r pairwise-overlap}
# Create pairwise overlap summary
labs <- c("afekta", "hmgu", "icl", "cembio")
pairwise_summary <- data.frame()

for (i in seq_along(labs)) {
  for (j in seq_along(labs)) {
    if (i < j) {
      lab1 <- labs[i]
      lab2 <- labs[j]

      col1 <- paste0(lab1, "_detected")
      col2 <- paste0(lab2, "_detected")

      both <- sum(consensus_matches_1[[col1]] & consensus_matches_1[[col2]])
      only_lab1 <- sum(consensus_matches_1[[col1]] & !consensus_matches_1[[col2]])
      only_lab2 <- sum(!consensus_matches_1[[col1]] & consensus_matches_1[[col2]])

      pairwise_summary <- rbind(pairwise_summary, data.frame(
        lab1 = lab1,
        lab2 = lab2,
        both_detected = both,
        only_lab1 = only_lab1,
        only_lab2 = only_lab2,
        overlap_percent = round(100 * both / max(
          sum(consensus_matches_1[[col1]]),
          sum(consensus_matches_1[[col2]])
        ), 1)
      ))
    }
  }
}

knitr::kable(pairwise_summary, caption = "Pairwise Lab Overlap")
```

## Summary by Detection Pattern

```{r detection-patterns}
# Count peaks by detection pattern
detection_patterns <- consensus_matches_1 |>
  group_by(afekta_detected, hmgu_detected, icl_detected, cembio_detected) |>
  summarize(n_peaks = n(), .groups = "drop") |>
  arrange(desc(n_peaks)) |>
  mutate(
    pattern = paste0(
      if_else(afekta_detected, "A", "-"),
      if_else(hmgu_detected, "H", "-"),
      if_else(icl_detected, "I", "-"),
      if_else(cembio_detected, "C", "-")
    )
  ) |>
  select(pattern, n_peaks, everything())

knitr::kable(detection_patterns,
             caption = "Peak Detection Patterns (A=afekta, H=hmgu, I=icl, C=cembio)")
```

Could remove column 3:6 in table above
Below decide on the threshold to study a peak. Would start with 4 at the beginning.

```{r}
thres <- 4

consensus_matches_1_filtered <- consensus_matches_1 |>
  filter(n_labs_matched == 4)

cat("Consensus peaks retained:", nrow(consensus_matches_1_filtered), "\n")
cat(
  "Percentage of detected peaks:",
  round(100 * nrow(consensus_matches_1_filtered) / nrow(consensus_matches_1), 1),
  "%\n"
)
```


Build the final consensus peak table with chrom_peak_id mapping:

Save the consensus table for downstream analysis:

```{r save-consensus}
write.csv(consensus_peaks,
          file = file.path(base_path, "object/consensus_peaks.csv"),
          row.names = FALSE)
write.csv(consensus_matches_1_filtered,
          file = file.path(base_path, "object/consensus_matches_1_filtered.csv"),
          row.names = FALSE)
```

## Summary Tables

### Overlap: Cembio vs Afekta

```{r cembio-afekta-overlap}
cembio_afekta_summary <- data.frame(
  category = c("Both detected", "Only afekta", "Only cembio", "Neither"),
  count = c(
    sum(consensus_matches_1$afekta_detected & consensus_matches_1$cembio_detected),
    sum(consensus_matches_1$afekta_detected & !consensus_matches_1$cembio_detected),
    sum(!consensus_matches_1$afekta_detected & consensus_matches_1$cembio_detected),
    sum(!consensus_matches_1$afekta_detected & !consensus_matches_1$cembio_detected)
  )
)

cembio_afekta_summary <- cembio_afekta_summary |>
  mutate(percent = round(100 * count / sum(count), 1))

knitr::kable(cembio_afekta_summary,
             caption = "Peak Overlap: Cembio vs Afekta")
```

### Overlap: All 4 Labs

```{r all-labs-overlap}
all_labs_summary <- data.frame(
  detection_pattern = character(),
  n_peaks = integer(),
  percent = numeric()
)

for (combo in 0:15) {
  afekta <- as.logical(bitwAnd(combo, 1))
  hmgu <- as.logical(bitwAnd(combo, 2))
  icl <- as.logical(bitwAnd(combo, 4))
  cembio <- as.logical(bitwAnd(combo, 8))

  n <- sum(
    consensus_matches_1$afekta_detected == afekta &
    consensus_matches_1$hmgu_detected == hmgu &
    consensus_matches_1$icl_detected == icl &
    consensus_matches_1$cembio_detected == cembio
  )

  if (n > 0) {
    pattern <- paste0(
      if_else(afekta, "A", "-"),
      if_else(hmgu, "H", "-"),
      if_else(icl, "I", "-"),
      if_else(cembio, "C", "-")
    )

    all_labs_summary <- rbind(all_labs_summary, data.frame(
      detection_pattern = pattern,
      n_peaks = n,
      percent = round(100 * n / nrow(consensus_matches_1), 1)
    ))
  }
}

all_labs_summary <- all_labs_summary |>
  arrange(desc(n_peaks))

knitr::kable(all_labs_summary,
             caption = "Peak Detection Patterns: All 4 Labs (A=afekta, H=hmgu, I=icl, C=cembio)")
```

# Peak Feature Summary by Lab (Detected)

## Prepare subsets for downstream analysis

```{r prepare-subsets}
# Subset 1: Peaks present in BOTH cembio AND afekta
cembio_afekta_both <- consensus_matches_1 |>
  filter(afekta_detected & cembio_detected) ## keep the id for cembio and afekta

cat("Peaks in both cembio & afekta:", nrow(cembio_afekta_both), "\n")

# Subset 2: Peaks present in ALL 4 labs
all_four_labs <- consensus_matches_1 |>
  filter(afekta_detected & hmgu_detected & icl_detected & cembio_detected)

cat("Peaks in all 4 labs:", nrow(all_four_labs), "\n")
```


# Creating eics

## 4 labs

maybe we can map the id from the begining if the consensus table give back indices.
```{r}
peak.table <- all_four_labs |>
  mutate(consensus_id = row_number()) |>
  select(consensus_id, afekta, hmgu, icl, cembio) |>
  pivot_longer(
    cols = c(afekta, hmgu, icl, cembio),
    names_to = "lab",
    values_to = "chrom_peak_id"
  ) |>
  filter(!is.na(chrom_peak_id))
```

Get mzmin, mzmax, rtmin, rtmax info from pt_* of each lab USING THE chrom_peak_id column of the peak.table and the lab column  T between the 2 table. probably need to do it sequnetially for eahc lab

```{r}
# Add mzmin, mzmax, rtmin, rtmax to peak.table by joining with each lab's pt_* table
peak.table <- peak.table |>
  mutate(
    mzmin = case_when(
      lab == "afekta" ~ pt_afekta$mzmin[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mzmin[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mzmin[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mzmin[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    mzmax = case_when(
      lab == "afekta" ~ pt_afekta$mzmax[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$mzmax[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$mzmax[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$mzmax[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    rtmin = case_when(
      lab == "afekta" ~ pt_afekta$rtmin[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$rtmin[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$rtmin[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$rtmin[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    ),
    rtmax = case_when(
      lab == "afekta" ~ pt_afekta$rtmax[match(
        chrom_peak_id,
        pt_afekta$chrom_peak_id
      )],
      lab == "hmgu" ~ pt_hmgu$rtmax[match(
        chrom_peak_id,
        pt_hmgu$chrom_peak_id
      )],
      lab == "icl" ~ pt_icl$rtmax[match(chrom_peak_id, pt_icl$chrom_peak_id)],
      lab == "cembio" ~ pt_cembio$rtmax[match(
        chrom_peak_id,
        pt_cembio$chrom_peak_id
      )]
    )
  )

#rename rtmin ot rtMin, .. and add/remove 5s to each rt size and 0.001 to each mz side

peak.table <- peak.table |>
  rename(rtMin = rtmin, rtMax = rtmax) |>
  mutate(
    rtMin = rtMin - 5,
    rtMax = rtMax + 5,
    mzmin = mzmin - 0.001,
    mzmax = mzmax + 0.001
  )

## mm i think i am missing the sample info.
```

well right now the peaks are wrong so there is no overlap but anyway


```{r}
library(Chromatograms)
load(file.path(base_path, "object/sp_full.RData"))
eics_4 <- chromExtract()
```



# Downstream Peak Feature Analysis (Detected)

## RT Stability - Cembio vs Afekta (Both Labs)

```{r rt - stability - cembio - afekta}
# Filter full_peak_report_detected to only cembio & afekta peaks that are in both labs

cembio_afekta_peak_ids <- c(
  cembio_afekta_both$afekta,
  cembio_afekta_both$cembio
)

peak_report_cembio_afekta <- full_peak_report_detected |>
  filter(chrom_peak_id %in% cembio_afekta_peak_ids)

rt_report_ca <- peak_report_cembio_afekta |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report_ca, 10)

ggplot(rt_report_ca, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (Cembio & Afekta)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency - Cembio vs Afekta (Both Labs)

```{r intensity-consistency-cembio-afekta}
intensity_report_ca <- peak_report_cembio_afekta |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_ca, 10)

ggplot(intensity_report_ca, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (Cembio & Afekta)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## RT Stability - All 4 Labs

```{r rt-stability-all-four}
# Filter full_peak_report_detected to only peaks present in all 4 labs

all_four_peak_ids <- c(
  all_four_labs$afekta,
  all_four_labs$hmgu,
  all_four_labs$icl,
  all_four_labs$cembio
)

peak_report_all_four <- full_peak_report_detected |>
  filter(chrom_peak_id %in% all_four_peak_ids)

rt_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report_all4, 10)

ggplot(rt_report_all4, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (All 4 Labs)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency - All 4 Labs

```{r intensity-consistency-all-four}
intensity_report_all4 <- peak_report_all_four |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report_all4, 10)

ggplot(intensity_report_all4, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (All 4 Labs)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Inter-Lab Statistical Comparison

```{r fwhm-analysis}
p1 <- ggviolin(full_peak_report_detected_meta, x = "lab", y = "fwhm", fill = "lab",
         add = "boxplot", add.params = list(fill = "white")) +
  geom_hline(yintercept = median(full_peak_report_detected_meta$fwhm), linetype = 2) +
  labs(title = "Peak Width (FWHM) by Lab (Detected)", y = "FWHM (s)", x = "Laboratory") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p1

kw_res <- full_peak_report_detected_meta %>% kruskal_test(fwhm ~ lab)
kw_res

eff_res <- full_peak_report_detected_meta %>% kruskal_effsize(fwhm ~ lab)
eff_res
```

## Hardware Impact

```{r lc-impact}
plot_data <- full_peak_report_detected_meta %>%
  filter(!is.na(lc_instrument_clean))

ggplot(plot_data, aes(x = lc_instrument_clean, y = fwhm, fill = lc_instrument_clean)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Impact of LC Instrument on Peak Width (Detected)", x = "LC System", y = "FWHM (s)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")
```

## PCA of Peak Features (Detected)

```{r pca-analysis}
pca_cols <- c("fwhm", "tailing_factor", "asymmetry", "entropy", "auc", "apex_intensity")
pca_input <- full_peak_report_detected_meta[, pca_cols]

pca_input$auc <- log10(pca_input$auc + 1)
pca_input$apex_intensity <- log10(pca_input$apex_intensity + 1)
pca_input <- na.omit(pca_input)

pca_res <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_scores <- as.data.frame(pca_res$x)
pca_scores$lab <- full_peak_report_detected_meta$lab[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$lc_type <- full_peak_report_detected_meta$lc_instrument_clean[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$mixture <- full_peak_report_detected_meta$mixture[rownames(pca_input) %in% rownames(pca_scores)]

var_expl <- summary(pca_res)$importance[2, 1:2] * 100

g1 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by Lab (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2])) +
  theme_bw()

g2 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lc_type)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by LC System (Detected)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2]),
       color = "LC System") +
  theme_bw()

ggarrange(g1, g2, ncol = 2, common.legend = FALSE)
```

## Conclusions (Detected)

Detected peak feature analysis complete. Results saved to `object/full_peak_report_detected.*`.
