---
title: "Peak Feature Calculation — Annotated Peaks"
format:
  html:
    toc: true
execute:
  dir: project
---

# Introduction

This document performs peak feature analysis restricted to annotated peaks.
We load detected peak metrics and filter to the subset of peaks present in each
lab’s annotation library.

## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(stringr)
library(janitor)
library(pheatmap)

base_path <- file.path("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
```

# Load Detected Peak Report

```{r load-detected-report}
# Reuse metrics computed in the detected analysis
load(file.path(base_path, "object/full_peak_report_detected.RData"))
```

# Load Annotation Libraries and Build Peak Table

```{r load-annotations}
study_group <- "HE"
res_afekta <- read.csv(file.path("4_library_generation", "afekta",
                                 study_group, "ring_trial_library_HE.csv"))
res_hmgu <- read.csv(file.path("4_library_generation", "hmgu",
                               study_group, "ring_trial_library_HE.csv"))
res_icl <- read.csv(file.path("4_library_generation", "icl",
                               study_group, "ring_trial_library_HE.csv"))
res_cembio <- read.csv(file.path("4_library_generation", "cembio",
                                 study_group, "ring_trial_library_HE.csv"))

res_afekta$lab <- "afekta"
res_hmgu$lab <- "hmgu"
res_icl$lab <- "icl"
res_cembio$lab <- "cembio"

peak_table <- rbind(res_hmgu, res_afekta, res_icl, res_cembio)

# Standardize column names used downstream
colnames(peak_table)[3:6] <- c("mzMin", "mzMax", "rtMin", "rtMax")
peak_table$polarity <- ifelse(peak_table$polarity == "pos", 1, 0)

# Basic summaries
table(peak_table$lab)
split(peak_table$compound_name, peak_table$lab) |>
  lapply(unique) |>
  lapply(length)
```

# Filter Detected Report to Annotated Peaks

We define annotated peaks as those whose `compound_name` appears in the lab’s
annotation library.

```{r filter-annotated}
# Use compound_name and lab to define annotation membership
annot_keys <- peak_table |>
  distinct(lab, compound_name)

full_peak_report_annotated <- full_peak_report_detected |>
  inner_join(annot_keys, by = c("lab", "compound_name")) |>
  mutate(is_annotated = TRUE)

# Optional intensity filter (same as detected)
idx_low_int <- which(full_peak_report_annotated$apex_intensity < 500)
if (length(idx_low_int) > 0) {
  full_peak_report_annotated <- full_peak_report_annotated[-idx_low_int, ]
}

head(full_peak_report_annotated)
```

Save the annotated peak report:

```{r save-peak-report}
save(full_peak_report_annotated, file = file.path(base_path, "object/full_peak_report_annotated.RData"))
write.csv(full_peak_report_annotated, file = file.path(base_path, "object/full_peak_report_annotated.csv"), row.names = FALSE)
```

## Load and Clean Lab Setup Metadata

```{r load-metadata}
lab_meta_raw <- read.csv(
  file.path(base_path, "lab_setup_phase1.csv"),
  header = FALSE,
  stringsAsFactors = FALSE
)

colnames(lab_meta_raw) <- lab_meta_raw[2, ]
lab_meta <- lab_meta_raw[-c(1, 2), ] %>%
  clean_names() %>%
  rename(lab_id = 1) %>%
  filter(lab_id != "") %>%
  mutate(
    lc_instrument_clean = str_extract(lc_instrument, "^[A-Za-z]+"),
    ms_instrument_clean = str_extract(ms_instrument, "^[A-Za-z]+"),
    column_brand = str_extract(column, "^[A-Za-z]+")
  ) %>%
  filter(details == "harmonized")

full_peak_report_annotated_meta <- full_peak_report_annotated %>%
  left_join(lab_meta, by = c("lab" = "lab_id"))

head(full_peak_report_annotated_meta)
```

# Peak Feature Summary by Lab (Annotated)

## Entropy

```{r entropy-summary}
split(full_peak_report_annotated$entropy, full_peak_report_annotated$lab) |>
  lapply(summary)
```

## FWHM

```{r fwhm-summary}
split(full_peak_report_annotated$fwhm, full_peak_report_annotated$lab) |>
  lapply(summary)
```

## Gaussian Similarity

```{r gaussian-summary}
split(full_peak_report_annotated$gaussian_similarity, full_peak_report_annotated$lab) |>
  lapply(summary)
```

## Tailing Factor

```{r tailing-summary}
split(full_peak_report_annotated$tailing_factor, full_peak_report_annotated$lab) |>
  lapply(summary)
```

# Downstream Peak Feature Analysis (Annotated)

## RT Stability

```{r rt-stability}
rt_report <- full_peak_report_annotated |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_rt))

head(rt_report, 10)

ggplot(rt_report, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Retention Time CVs (Annotated)",
    x = "CV of RT (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Intensity Consistency

```{r intensity-consistency}
intensity_report <- full_peak_report_annotated |>
  group_by(compound_name, mixture) |>
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n(),
    .groups = "drop"
  ) |>
  arrange(desc(cv_auc))

head(intensity_report, 10)

ggplot(intensity_report, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(
    title = "Distribution of Peak Area (AUC) CVs (Annotated)",
    x = "CV of AUC (per compound)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Inter-Lab Statistical Comparison

```{r fwhm-analysis}
p1 <- ggviolin(full_peak_report_annotated_meta, x = "lab", y = "fwhm", fill = "lab",
         add = "boxplot", add.params = list(fill = "white")) +
  geom_hline(yintercept = median(full_peak_report_annotated_meta$fwhm), linetype = 2) +
  labs(title = "Peak Width (FWHM) by Lab (Annotated)", y = "FWHM (s)", x = "Laboratory") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p1

kw_res <- full_peak_report_annotated_meta %>% kruskal_test(fwhm ~ lab)
kw_res

eff_res <- full_peak_report_annotated_meta %>% kruskal_effsize(fwhm ~ lab)
eff_res
```

## Hardware Impact

```{r lc-impact}
plot_data <- full_peak_report_annotated_meta %>%
  filter(!is.na(lc_instrument_clean))

ggplot(plot_data, aes(x = lc_instrument_clean, y = fwhm, fill = lc_instrument_clean)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Impact of LC Instrument on Peak Width (Annotated)", x = "LC System", y = "FWHM (s)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")
```

## PCA of Peak Features (Annotated)

```{r pca-analysis}
pca_cols <- c("fwhm", "tailing_factor", "asymmetry", "entropy", "auc", "apex_intensity")
pca_input <- full_peak_report_annotated_meta[, pca_cols]

pca_input$auc <- log10(pca_input$auc + 1)
pca_input$apex_intensity <- log10(pca_input$apex_intensity + 1)
pca_input <- na.omit(pca_input)

pca_res <- prcomp(pca_input, center = TRUE, scale. = TRUE)

pca_scores <- as.data.frame(pca_res$x)
pca_scores$lab <- full_peak_report_annotated_meta$lab[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$lc_type <- full_peak_report_annotated_meta$lc_instrument_clean[rownames(pca_input) %in% rownames(pca_scores)]
pca_scores$mixture <- full_peak_report_annotated_meta$mixture[rownames(pca_input) %in% rownames(pca_scores)]

var_expl <- summary(pca_res)$importance[2, 1:2] * 100

g1 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by Lab (Annotated)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2])) +
  theme_bw()

g2 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = lc_type)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Colored by LC System (Annotated)",
       x = sprintf("PC1 (%.1f%%)", var_expl[1]),
       y = sprintf("PC2 (%.1f%%)", var_expl[2]),
       color = "LC System") +
  theme_bw()

ggarrange(g1, g2, ncol = 2, common.legend = FALSE)
```

## Conclusions (Annotated)

Annotated peak feature analysis complete. Results saved to `object/full_peak_report_annotated.*`. 