---
title: "Defining and Overcoming the Limits of Multi-Center Metabolomics: A Computational Blueprint for Data Integration"
format:
  html:
    toc: true
    toc-location: right
    code-fold: true
execute:
  dir: project
---

# Introduction

## Input Data

All required input data lives in the `data/` subfolder. These files are
produced by the upstream pipeline (`1_object_creation.qmd` →
`2_global_analysis.qmd` → `3a_peak_feature_calculation_detected.qmd`).
I can also provide them pre-computed.

**RData objects** (chromatographic profiles, from `1_object_creation.qmd`):
`bpc_full.RData`, `bpc_detect.RData`, `tic_full.RData`, `tic_detect.RData`

**CSV files** (from `1_object_creation.qmd` and `3a_peak_feature_calculation_detected.qmd`):
`detected_peaks_{afekta,cembio,hmgu,icl}_HE.csv`,
`peak_metrics_detected_all4labs.csv`, `peak_shape_similarity.csv`

**Helper files** (in `5_downstream_analysis/`):
`global_analysis_helpers.R`, `lab_setup_phase1.csv`

## Question and Hypothesis

**Core question:** *What are the computational steps required to make
multi-center metabolomics data integration possible, and what are the absolute
hardware-imposed limits of the field?*

**Hypothesis:** Raw LC-MS data from different laboratories is too confounded
by instrumental noise to be directly integrated. Computational processing can
recover biological signal, but instrument chemistry imposes an irreducible
reproducibility ceiling.

We benchmark four sequential processing strategies:

1. **Peak-picking** (feature extraction vs. raw profiles)
2. **Local (RT-dependent) normalization** (LOESS vs. global scaling)
3. **Peak shape confirmation** (EIC Pearson similarity for cross-lab anchors)
4. **Hardware ceiling** (same-platform ICC as the achievable upper bound)

Data: 19 standard mixtures (HUMAN endosome) from 4 labs with harmonized LC
(same column, mobile phases, gradient) but different MS platforms (Agilent
Q-TOF, Waters Synapt, Sciex Zeno, Bruker TIMS). No technical replicates; lab
identity is confounded with instrument platform.

Baseline finding: **lab identity explains 97–100% of variance in raw TIC**
(Full Signal).

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
##The install belows are temporary and should be replaced with proper dependencies in the future.
# BiocManager::install("RforMassSpectrometry/Chromatograms", ask = FALSE, update = TRUE)
# BiocManager::install("tnaake/MsQuality", ask = FALSE, update = TRUE)
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)
library(pheatmap)

base_path <- here("5_downstream_analysis")
source(file.path(base_path, "global_analysis_helpers.R"))
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup |>
  filter(details == "harmonized") |>
  select(lab, `LC.Instrument`, `MS.Instrument`,
         `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) |>
  distinct() |>
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) |>
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

## Compute Global Metrics (Raw Signal)

TIC and BPC per chromatogram across data levels and polarities; no
normalization (added in Part 2).

```{r compute-metrics}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "data/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]
    load(file.path(base_path, "data/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "data/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]
    load(file.path(base_path, "data/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) |>
    left_join(setup_simplified, by = "lab") |>
    mutate(data_level = ifelse(data_level == "full",
                               "Full Signal", "Detected Signal"))
}

metrics_full <- run_for_polarities(compute_metrics, "full")
metrics_detect <- run_for_polarities(compute_metrics, "detect")

# Create Consensus Signal data level from pk_metrics
compute_consensus_metrics <- function(pol, polarity_name, data_level = "detect") {
  pk_metrics <- read.csv(file.path(base_path,
                                   "data/peak_metrics_detected_all4labs.csv"))

  pk_metrics |>
    filter(as.integer(polarity) == pol) |>
    group_by(lab, mixture) |>
    summarise(
      tic = sum(areaUnderTic, na.rm = TRUE),
      bpc = sum(maxIntensity, na.rm = TRUE),
      .groups = "drop"
    ) |>
    left_join(setup_simplified, by = "lab") |>
    mutate(
      polarity = polarity_name,
      data_level = "Consensus Signal"
    )
}

metrics_consensus <- run_for_polarities(compute_consensus_metrics, "detect")

metrics_combined <- bind_rows(metrics_full, metrics_detect, metrics_consensus)
```

## Data Processing Levels

- **Full Signal**: raw chromatographic profiles (all ions + baseline).
- **Detected Signal**: centWave-picked peaks, per lab (not yet matched).
- **Consensus Signal**: peaks matched across all 4 labs by m/z and RT.

Progression Full → Detected → Consensus tests whether computational processing
progressively removes technical variance.

---

# Part 1: The Baseline Problem - Feature Extraction is Mandatory {#part1}

**Core claim:** Raw chromatographic profiles cannot be integrated across labs
because vendor-specific baseline noise dominates the signal. Feature extraction
(peak-picking) is the mandatory first computational step to make data integration
theoretically possible.

## 1.1 The Problem: Lab Variance Dominates Raw Data

### 1.1.1 Variance Decomposition

```{r variance-decomposition}
results_table <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

```{r variance-stacked-bar}
#| fig-cap: "**Figure 1.1**: Variance decomposition of TIC. Lab identity dominates raw signal variance — the core problem."
variance_long <- results_table |>
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) |>
  pivot_longer(cols = ends_with("_Pct"),
               names_to = "Source", values_to = "Percentage") |>
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Label = paste(Polarity, Data_Level, sep = "\n"),
    R2_label = paste("R² =", round(R_squared, 3))
  )

text_labels <- variance_long |> distinct(Label, R2_label)

ggplot(variance_long, aes(x = Label, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(x = Label, label = R2_label, y = 105),
            data = text_labels, size = 3, inherit.aes = FALSE) +
  labs(title = "Variance Decomposition: TIC Across All Conditions",
       subtitle = "Lab variance (orange) dominates raw signal — the core problem",
       x = "Polarity & Data Level", y = "Variance (%)", fill = "Source") +
  theme_minimal() +
  theme(legend.position = "top") +
  ylim(0, 110)
```

### 1.1.2 Lab Effect Coefficients

```{r lab-coefficients}
all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    all_coefficients <- rbind(all_coefficients,
      result$coefficients |> mutate(Polarity = pol, Data_Level = level))
  }
}

lab_coefs <- all_coefficients |>
  filter(type == "Lab Effect") |>
  mutate(lab_name = gsub("lab", "", coefficient))

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "Lab Effect Coefficients (deviation from Afekta)",
       x = "Laboratory", y = "Coefficient (log2 scale)", fill = "Lab") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.1.3 PCA Analysis of Lab vs. Mixture Clustering

Colour = mixture, shape = lab. Same shape clustering → lab dominates;
same colour clustering → mixture (biology) dominates.

```{r pca-setup}
# Prepare data for PCA: wide format with samples as rows, features as columns
# For raw data: use individual RT bins as features
# For detected data: use detected peaks as features

# Function to prepare PCA data
prepare_pca_data <- function(data_level = "full", polarity = "Positive") {
  if (data_level == "full") {
    # Load full signal data
    load(file.path(base_path, "data/tic_full.RData"))
    chr_data <- tic_full[tic_full$polarity == ifelse(polarity == "Positive", 1, 0)]
  } else {
    # Load detected signal data
    load(file.path(base_path, "data/tic_detect.RData"))
    chr_data <- tic_detect[tic_detect$polarity == ifelse(polarity == "Positive", 1, 0)]
  }

  # Extract intensity matrices
  # Each row = one sample, columns = RT bins
  int_list <- intensity(chr_data)
  rt_list <- rtime(chr_data)

  # Create common RT grid
  all_rts <- unlist(rt_list)
  rt_grid <- seq(min(all_rts, na.rm = TRUE),
                 max(all_rts, na.rm = TRUE),
                 length.out = 500)

  # Interpolate each chromatogram onto the grid
  int_matrix <- t(sapply(seq_along(int_list), function(i) {
    approx(rt_list[[i]], int_list[[i]], xout = rt_grid,
           method = "linear", rule = 2)$y
  }))

  # Add sample metadata
  rownames(int_matrix) <- paste0(chr_data$lab, "_", chr_data$mixture)
  colnames(int_matrix) <- paste0("RT_", round(rt_grid, 1))

  # Create metadata data frame
  meta_df <- data.frame(
    sample = rownames(int_matrix),
    lab = chr_data$lab,
    mixture = chr_data$mixture,
    row.names = rownames(int_matrix)
  )

  list(data = int_matrix, meta = meta_df)
}

# Prepare datasets for both polarities
pca_full_pos <- prepare_pca_data("full", "Positive")
pca_full_neg <- prepare_pca_data("full", "Negative")
pca_detect_pos <- prepare_pca_data("detect", "Positive")
pca_detect_neg <- prepare_pca_data("detect", "Negative")

# Prepare consensus data from pk_metrics (wide format: samples x consensus peaks)
# Filter to one polarity at a time — consensus_ids are disjoint across polarities
# and the two modes have different intensity scales.
pk_metrics <- read.csv(file.path(base_path,
                                 "data/peak_metrics_detected_all4labs.csv"))

prepare_pca_consensus <- function(polarity_int) {
  wide <- pk_metrics |>
    filter(as.integer(polarity) == polarity_int) |>
    mutate(sample = paste0(lab, "_", mixture)) |>
    select(sample, consensus_id, areaUnderTic, lab, mixture) |>
    pivot_wider(names_from = consensus_id, values_from = areaUnderTic,
                values_fill = 0, names_prefix = "peak_") |>
    arrange(sample)

  data_mat <- wide |> select(starts_with("peak_")) |> as.matrix()
  rownames(data_mat) <- wide$sample

  list(data = data_mat, meta = wide |> select(sample, lab, mixture))
}

pca_consensus_pos <- prepare_pca_consensus(1L)
pca_consensus_neg <- prepare_pca_consensus(0L)

# Helper: run prcomp and return a scores data frame tagged with polarity.
# `transform_fn` applies a transform before PCA (identity for continuous TIC
# profiles, log1p for sparse consensus peaks).
pca_scores_df <- function(pca_list, polarity_label, transform_fn = identity) {
  res <- prcomp(transform_fn(pca_list$data), scale. = TRUE, center = TRUE)
  var_expl <- (res$sdev^2 / sum(res$sdev^2)) * 100
  list(
    scores = data.frame(
      PC1 = res$x[, 1], PC2 = res$x[, 2],
      lab = pca_list$meta$lab, mixture = pca_list$meta$mixture,
      polarity = polarity_label
    ),
    var_expl = var_expl
  )
}

# Common plot: 2-panel PCA (Positive | Negative).
# Colour encodes mixture (biological signal) and shape encodes lab (technical).
# As data level progresses Full → Detected → Consensus, colours should tighten
# (mixture clusters emerge) while shapes should disperse within each colour
# (lab effects shrink).
plot_pca_by_polarity <- function(pos, neg, title, subtitle) {
  scores <- bind_rows(pos$scores, neg$scores) |>
    mutate(polarity = factor(polarity, levels = c("Positive", "Negative")))
  panel_labels <- c(
    Positive = sprintf("Positive  (PC1 %.1f%%, PC2 %.1f%%)",
                       pos$var_expl[1], pos$var_expl[2]),
    Negative = sprintf("Negative  (PC1 %.1f%%, PC2 %.1f%%)",
                       neg$var_expl[1], neg$var_expl[2])
  )
  ggplot(scores, aes(x = PC1, y = PC2, color = mixture, shape = lab)) +
    geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    facet_wrap(~polarity, scales = "free",
               labeller = labeller(polarity = panel_labels)) +
    scale_color_viridis_d(option = "turbo") +
    labs(title = title, subtitle = subtitle,
         x = "PC1", y = "PC2", color = "Mixture", shape = "Laboratory") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))
}
```

```{r pca-raw}
#| fig-cap: "**Figure 1.2a**: PCA of Raw TIC profiles, Positive and Negative polarity. Colour = mixture, shape = lab. Same-shape points cluster (labs group together); colours are scattered — technical noise dominates biological signal."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_full_pos, "Positive"),
  neg = pca_scores_df(pca_full_neg, "Negative"),
  title = "PCA: Raw TIC Profiles (Full Signal)",
  subtitle = "Shapes cluster (lab dominates), colours scattered (mixture signal buried)"
)
```

```{r pca-detected}
#| fig-cap: "**Figure 1.2b**: PCA of Detected Peaks, Positive and Negative polarity. Colour = mixture, shape = lab. Lab clusters weaken and same-colour points start to group together — feature extraction recovers biological signal."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_detect_pos, "Positive"),
  neg = pca_scores_df(pca_detect_neg, "Negative"),
  title = "PCA: Detected Peaks (Peak-Picked Signal)",
  subtitle = "Shapes disperse, colours begin to group — mixture signal emerges"
)
```

```{r pca-consensus}
#| fig-cap: "**Figure 1.2c**: PCA of Consensus Peaks, Positive and Negative polarity. Colour = mixture, shape = lab. Same-colour points cluster tightly regardless of shape — mixture dominates, lab effects minimized. log1p compresses the intensity range because each peak has ~4 non-zero values (labs) and ~72 zeros (other mixtures); without it the presence/absence gap dominates scaling."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_consensus_pos, "Positive", transform_fn = log1p),
  neg = pca_scores_df(pca_consensus_neg, "Negative", transform_fn = log1p),
  title = "PCA: Consensus Peaks (Cross-Lab Matched Signal)",
  subtitle = "Colours cluster by mixture across all shapes — biology dominates technical variance"
)
```

**Conclusion:** Raw profile data cannot be integrated across labs;
feature extraction followed by cross-lab matching is required.

## 1.3 ICC: Full Signal vs Detected Signal

### 1.3.1 ICC Computation

ICC quantifies the ratio of biological signal to total variance:

$$ICC = \frac{\sigma^2_{between}}{\sigma^2_{between} + \sigma^2_{within}}$$

- $\sigma^2_{between}$ = variance *between* mixtures (biological signal)
- $\sigma^2_{within}$ = variance *within* the same mixture across labs (technical noise)
- ICC > 0.5 → acceptable reproducibility

```{r icc-computation}
# Raw signal ICC only — normalization strategies introduced in Part 2
icc_all_results <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)

    results <- compute_icc(data_subset, "tic")

    if (!all(is.na(data_subset$bpc))) {
      results <- bind_rows(
        results,
        compute_icc(data_subset, "bpc")
      )
    }

    results <- results |>
      mutate(Polarity = pol, Data_Level = level,
             Metric_Label = case_when(
               Metric == "tic" ~ "TIC",
               Metric == "bpc" ~ "BPC",
               TRUE ~ Metric))

    icc_all_results <- bind_rows(icc_all_results, results)
  }
}
```

### 1.3.2 Peak Detection Impact

```{r peak-detection-impact}
icc_comparison_tic <- icc_all_results |>
  filter(Metric == "tic") |>
  select(Polarity, Data_Level, ICC) |>
  pivot_wider(names_from = Data_Level, values_from = ICC) |>
  mutate(
    ICC_Improvement = `Detected Signal` - `Full Signal`,
    Percent_Gain = round(
      (`Detected Signal` - `Full Signal`) / `Full Signal` * 100, 1)
  )

knitr::kable(icc_comparison_tic,
             caption = "ICC improvement from peak detection (raw TIC)",
             digits = 3)
```

```{r peak-detection-plot}
#| fig-cap: "Peak detection improves ICC on raw TIC; consensus peaks reveal the true signal ceiling."
ggplot(icc_all_results |> filter(Metric == "tic"),
       aes(x = Polarity, y = ICC, fill = Data_Level)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(ICC, 2)),
            position = position_dodge(0.9), vjust = -0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Impact of Peak Detection on Reproducibility",
       subtitle = "Raw TIC (no normalization); dashed = 0.5 threshold",
       y = "ICC", x = "Polarity", fill = "Processing Level") +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```

### 1.3.3 ICC Variance Decomposition

```{r icc-variance-decomposition}
#| fig-cap: "Peak detection shifts variance toward biological signal (raw signal, no normalization)."
variance_plot <- icc_all_results |>
  select(Polarity, Data_Level, Metric_Label,
         Between_mixture_var, Within_mixture_var) |>
  pivot_longer(cols = c(Between_mixture_var, Within_mixture_var),
               names_to = "Source", values_to = "Variance") |>
  mutate(Source = case_when(
    Source == "Between_mixture_var" ~ "Between mixtures (signal)",
    Source == "Within_mixture_var" ~ "Within mixtures (lab noise)"
  ))

ggplot(variance_plot,
       aes(x = Metric_Label, y = Variance, fill = Source)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.7) +
  facet_grid(Data_Level ~ Polarity, scales = "free_y") +
  scale_fill_manual(values = c(
    "Between mixtures (signal)" = "#619CFF",
    "Within mixtures (lab noise)" = "#F8766D")) +
  labs(title = "Variance Decomposition: Raw TIC and BPC",
       x = "Metric", y = "Variance (log2 scale)", fill = "Source") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

### 1.3.4 The Consensus Signal Breakthrough

ICC on summed TIC of the 871 cross-lab matched peaks vs. Full and Detected.

```{r consensus-signal-icc}
# All three levels use raw TIC ICC (no normalization)
icc_comparison_all <- icc_all_results |>
  filter(Polarity == "Positive", Metric == "tic") |>
  select(Data_Level, ICC)

knitr::kable(icc_comparison_all,
             caption = "Raw TIC ICC progression: Full → Detected → Consensus Signal (Positive mode)",
             digits = 3)
```

```{r consensus-signal-plot}
#| fig-cap: "**Figure 1.4**: Raw TIC ICC across data levels. Consensus Signal (matched peaks) achieves ICC ≈ 0.50, far above Full and Detected Signal levels. Peak matching is the critical bottleneck."
#| fig-width: 8
#| fig-height: 5

ggplot(icc_comparison_all, aes(x = Data_Level, y = ICC, fill = Data_Level)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_text(aes(label = sprintf("ICC = %.3f", ICC)),
            vjust = -0.5, size = 4) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("#E64B35", "#4DBBD5", "#00A087")) +
  scale_x_discrete(limits = c("Full Signal", "Detected Signal", "Consensus Signal")) +
  labs(
    title = "Peak Matching is the Critical Bottleneck",
    subtitle = "Raw TIC ICC: Full Signal → Detected Signal → Consensus Signal (Positive mode)",
    y = "Intraclass Correlation Coefficient (ICC)",
    x = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold")) +
  ylim(0, 0.7)
```

**Key Finding:** Raw TIC ICC progresses **0.002 → 0.024 → 0.502** (Positive
mode) across Full → Detected → Consensus. The ~21× jump from Detected to
Consensus shows **peak matching is the critical bottleneck**, not peak
detection alone.

---

# Part 2: Addressing RT-Dependent Bias Makes Patient Pooling Possible {#strategy2}

**Core claim:** RT-dependent drift between labs makes patient pooling
statistically invalid without local normalization.

**Data scope:** TIC of all detected peaks from XCMS centWave, aggregated per
sample (~105k–396k peaks per lab; ~1M total). Part 3 uses only the 871
cross-lab consensus peaks.

## 2.0 Add Normalization Variants

Median-scaled and autoscaled variants of TIC/BPC, per (lab, polarity, data_level).

```{r add-normalization-variants}
metrics_combined <- metrics_combined |>
  group_by(lab, polarity, data_level) |>
  mutate(
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE)
  ) |>
  ungroup()
```

## 2.1 Normalization Impact

### 2.1.0 Why Choose Median Scaling vs. Autoscaling vs. No Normalization?

- **Median scaling**: removes systematic per-lab sensitivity differences;
  appropriate when the design is balanced (all labs run the same 19 mixtures).
- **Autoscaling** (dividing by SD/MAD): amplifies low-abundance noise.
- **No normalization**: baseline — lab-to-lab intensity differences remain.

### 2.1.1 Global Normalization: Median Scaling

```{r normalization-variance}
#| fig-cap: "Effect of normalization on variance decomposition."
test_data <- metrics_combined |>
  filter(polarity == "Positive", data_level == "Detected Signal")

result_raw <- check_metric_quality(test_data, "tic")
result_median <- check_metric_quality(test_data, "tic_median_scaled")
result_auto <- check_metric_quality(test_data, "tic_autoscaled")

normalization_results <- bind_rows(
  result_raw$variance_table |>
    mutate(Normalization = "Raw TIC", R_squared = result_raw$r_squared),
  result_median$variance_table |>
    mutate(Normalization = "Median Scaled", R_squared = result_median$r_squared),
  result_auto$variance_table |>
    mutate(Normalization = "Autoscaled (MAD)", R_squared = result_auto$r_squared)
)

normalization_results |>
  ggplot(aes(x = Normalization, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  labs(title = "Effect of Normalization on Variance Decomposition (TIC)",
       subtitle = "Positive Mode, Detected Signal",
       x = "Normalization Method", y = "Variance (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 2.1.2 Residual Comparison

```{r normalization-residuals}
#| fig-height: 12
mod_raw <- lm(log2(tic) ~ lab + factor(mixture), data = test_data)
mod_median <- lm(log2(tic_median_scaled) ~ lab + factor(mixture),
                 data = test_data)

plot_residuals(mod_raw, "Raw TIC (log2)", test_data) /
  plot_residuals(mod_median, "Median Scaled TIC", test_data) +
  plot_annotation(
    title = "Residual Plots: Raw vs Median Scaled"
  )
```

## 2.2 The Problem: Non-Linear Sensitivity Profiles

### 2.2.1 Instrument Sensitivity Across RT Quartiles

Non-parallel lines across quartiles → RT-dependent biases that global
normalization cannot fix.

```{r sensitivity-profile}
compute_consortium_sensitivity <- function(polarity) {
  load(file.path(base_path, "data/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  breaks <- quantile(unlist(rtime(tic_data)),
                     probs = seq(0, 1, 0.25), na.rm = TRUE)

  raw_stats <- t(vapply(seq_along(tic_data), function(i) {
    rt <- rtime(tic_data[i])[[1]]
    int <- intensity(tic_data[i])[[1]]
    vapply(1:4, function(q) {
      sum(int[rt >= breaks[q] & rt < breaks[q + 1]], na.rm = TRUE)
    }, numeric(1))
  }, numeric(4)))

  df <- data.frame(
    lab = tic_data$lab, mixture = tic_data$mixture,
    Q1 = raw_stats[, 1], Q2 = raw_stats[, 2],
    Q3 = raw_stats[, 3], Q4 = raw_stats[, 4]
  )
  grand_medians <- df |>
    summarise(across(starts_with("Q"), \(x) median(x, na.rm = TRUE))) |>
    pivot_longer(everything(), names_to = "Quartile", values_to = "Consensus")

  df |>
    pivot_longer(cols = starts_with("Q"),
                 names_to = "Quartile", values_to = "Intensity") |>
    left_join(grand_medians, by = "Quartile") |>
    mutate(
      Log2_Deviation = log2(Intensity / Consensus),
      Polarity = ifelse(polarity == 1, "Positive", "Negative")
    )
}

all_profiles <- run_for_polarities_simple(compute_consortium_sensitivity)
```

```{r sensitivity-profile-plot}
#| fig-cap: "Non-parallel sensitivity profiles → global normalization insufficient."
all_profiles |>
  mutate(Log2_TIC = log2(Intensity)) |>
  group_by(Quartile, lab, Polarity) |>
  summarise(median_log2_tic = median(Log2_TIC, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = Quartile, y = median_log2_tic,
             group = lab, color = lab)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~Polarity) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Raw TIC Distribution Across RT Quartiles",
    subtitle = "Non-parallel lines → global normalization is insufficient",
    y = "log2(TIC per Quartile)", x = "Quartile", color = "Laboratory"
  ) +
  theme_minimal()
```

```{r sensitivity-deviation-table}
findings_table <- all_profiles |>
  mutate(Log2_Dev = log2(Intensity / Consensus)) |>
  group_by(lab, Quartile, Polarity) |>
  summarise(Median_Log2_Dev = round(median(Log2_Dev, na.rm = TRUE), 2),
            .groups = "drop") |>
  pivot_wider(names_from = Quartile, values_from = Median_Log2_Dev)

knitr::kable(findings_table,
             caption = "Median log2 deviation from consortium median by lab and RT quartile")
```

ICL: Q1 +0.24 → Q4 −0.78 (positive). Cembio: Q3 +1.55 log2 deviation (negative).

### 2.2.2 Residual Lab Variance by RT Quartile

```{r quartile-variance}
compute_rt_quartiles_biological <- function(polarity, polarity_name) {
  load(file.path(base_path, "data/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  load(file.path(base_path, "data/bpc_detect.RData"))
  bpc_data <- bpc_detect[bpc_detect$polarity == polarity]

  tic_medians <- data.frame(
    lab = tic_data$lab,
    tic_total = vapply(intensity(tic_data), sum, numeric(1), na.rm = TRUE)
  ) |>
    group_by(lab) |>
    summarise(median_tic = median(tic_total, na.rm = TRUE), .groups = "drop")

  bpc_medians <- data.frame(
    lab = bpc_data$lab,
    bpc_total = vapply(intensity(bpc_data), sum, numeric(1), na.rm = TRUE)
  ) |>
    group_by(lab) |>
    summarise(median_bpc = median(bpc_total, na.rm = TRUE), .groups = "drop")

  all_rts <- unlist(rtime(tic_data))
  if (length(all_rts) == 0) return(NULL)
  global_breaks <- quantile(all_rts, probs = seq(0, 1, 0.25), na.rm = TRUE)

  compute_quartile_sums <- function(int_vec, rt_vec, breaks, lab_median) {
    vapply(1:4, function(q) {
      sum(int_vec[rt_vec >= breaks[q] & rt_vec < breaks[q + 1]],
          na.rm = TRUE) / lab_median
    }, numeric(1))
  }

  stats <- t(vapply(seq_along(tic_data), function(i) {
    rt <- rtime(tic_data[i])[[1]]
    c(compute_quartile_sums(
        intensity(tic_data[i])[[1]], rt, global_breaks,
        tic_medians$median_tic[tic_medians$lab == tic_data$lab[i]]),
      compute_quartile_sums(
        intensity(bpc_data[i])[[1]], rt, global_breaks,
        bpc_medians$median_bpc[bpc_medians$lab == bpc_data$lab[i]]))
  }, numeric(8)))

  colnames(stats) <- c(paste0("TIC_Frac_Q", 1:4), paste0("BPC_Frac_Q", 1:4))

  data.frame(
    lab = tic_data$lab, mixture = tic_data$mixture,
    polarity = polarity_name, stats
  )
}

bio_combined_quartiles <- run_for_polarities_names(
  compute_rt_quartiles_biological
)

quartile_results <- expand.grid(
  Polarity = c("Positive", "Negative"),
  Metric = c(paste0("TIC_Frac_Q", 1:4), paste0("BPC_Frac_Q", 1:4))
) |>
  rowwise() |>
  mutate(var_result = list(
    get_variance(filter(bio_combined_quartiles, polarity == Polarity), Metric)
  )) |>
  unnest(var_result)
```

```{r quartile-variance-plot}
#| fig-cap: "Residual lab variance peaks in Q4 (~53% positive mode, ~40% negative mode) after global median normalization, with Q3 also elevated (~29–34%)."
plot_quartile_variance(quartile_results, "TIC")
```

## 2.3 Comparing Normalization Strategies

**Question:** can normalization push ICC above the post-detection baseline
(raw TIC ICC ≈ 0.024–0.059), and can RT-dependent normalization fix
quartile-specific biases?

Three granularities compared (per-peak Area Under TIC):

1. **Global Median** — one scaling factor per lab
2. **Decile Median** — per-RT-decile scaling factors (10 bins)
3. **LOESS** — continuous RT-dependent normalization

### 2.3.1 Prepare Peak-Level Data

```{r load-peak-data}
pk_metrics <- read.csv(file.path(base_path,
                                 "data/peak_metrics_detected_all4labs.csv"))

cat("Peak metrics:", nrow(pk_metrics), "rows\n")
cat("Unique consensus peaks:", n_distinct(pk_metrics$consensus_id), "\n")
```

### 2.3.2 Apply Normalization Strategies

```{r normalization-strategies}
# Load RT breaks from detected signal (deciles instead of quartiles)
load(file.path(base_path, "data/tic_detect.RData"))
all_rts <- unlist(rtime(tic_detect))
rt_breaks_decile <- quantile(all_rts, probs = seq(0, 1, by = 0.1), na.rm = TRUE)

# Load peak tables with RT info
pt_all_labs <- bind_rows(
  read.csv(file.path(base_path, "data/detected_peaks_afekta_HE.csv")) |>
    mutate(lab = "afekta"),
  read.csv(file.path(base_path, "data/detected_peaks_hmgu_HE.csv")) |>
    mutate(lab = "hmgu"),
  read.csv(file.path(base_path, "data/detected_peaks_icl_HE.csv")) |>
    mutate(lab = "icl"),
  read.csv(file.path(base_path, "data/detected_peaks_cembio_HE.csv")) |>
    mutate(lab = "cembio")
)

# Add RT and decile assignment (10 bins: D1 to D10)
pk_norm <- pk_metrics |>
  left_join(
    pt_all_labs |> select(chrom_peak_id, rt, lab),
    by = c("chrom_peak_id", "lab")
  ) |>
  mutate(
    rt_decile = cut(rt,
                    breaks = rt_breaks_decile,
                    labels = paste0("D", 1:10),
                    include.lowest = TRUE)
  )

# Positive and negative ionization have very different intensity distributions,
# so all three strategies compute normalization factors per (lab, polarity) —
# pooling modes would bias each one's scaling factor.

# --- Strategy 1: Global Median ---
global_medians <- pk_norm |>
  group_by(lab, polarity) |>
  summarise(
    global_median_area = median(areaUnderTic, na.rm = TRUE),
    global_median_max = median(maxIntensity, na.rm = TRUE),
    .groups = "drop"
  )

grand_medians <- global_medians |>
  group_by(polarity) |>
  summarise(
    grand_median_area = median(global_median_area),
    grand_median_max = median(global_median_max),
    .groups = "drop"
  )

pk_norm <- pk_norm |>
  left_join(global_medians, by = c("lab", "polarity")) |>
  left_join(grand_medians, by = "polarity") |>
  mutate(
    area_global = areaUnderTic / global_median_area * grand_median_area,
    max_global = maxIntensity / global_median_max * grand_median_max
  )

# --- Strategy 2: Decile Median (NEW) ---
decile_medians <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(lab, polarity, rt_decile) |>
  summarise(
    d_median_area = median(areaUnderTic, na.rm = TRUE),
    d_median_max = median(maxIntensity, na.rm = TRUE),
    .groups = "drop"
  )

grand_d_medians <- decile_medians |>
  group_by(polarity, rt_decile) |>
  summarise(
    grand_d_area = median(d_median_area),
    grand_d_max = median(d_median_max),
    .groups = "drop"
  )

pk_norm <- pk_norm |>
  left_join(decile_medians, by = c("lab", "polarity", "rt_decile")) |>
  left_join(grand_d_medians, by = c("polarity", "rt_decile")) |>
  mutate(
    area_decile = areaUnderTic / d_median_area * grand_d_area,
    max_decile = maxIntensity / d_median_max * grand_d_max
  )

# --- Strategy 3: LOESS ---
pk_norm <- pk_norm |>
  group_by(lab, polarity) |>
  mutate(
    loess_area = {
      fit <- loess(log2(areaUnderTic) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    },
    loess_max = {
      fit <- loess(log2(maxIntensity) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    }
  ) |>
  ungroup()

# Grand LOESS (consensus curve across all labs, fit separately per polarity)
pk_norm <- pk_norm |>
  group_by(polarity) |>
  mutate(
    grand_loess_area = {
      fit <- loess(log2(areaUnderTic) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    },
    grand_loess_max = {
      fit <- loess(log2(maxIntensity) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    }
  ) |>
  ungroup() |>
  mutate(
    area_loess = 2^(log2(areaUnderTic) - loess_area + grand_loess_area),
    max_loess = 2^(log2(maxIntensity) - loess_max + grand_loess_max)
  )

cat("Normalization complete. Rows:", nrow(pk_norm), "\n")
```

### 2.3.3 Visualize LOESS Sensitivity Profiles

```{r loess-profiles}
#| fig-cap: "Per-lab LOESS sensitivity curves showing the RT-dependent bias that local normalization corrects."
pk_norm |>
  filter(!is.na(rt), !is.na(loess_area)) |>
  ggplot(aes(x = rt, y = loess_area, color = lab)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "LOESS Sensitivity Profiles by Lab",
    subtitle = "log2(Area) smoothed over RT — divergence = what local norm corrects",
    x = "Retention Time (s)", y = "LOESS-smoothed log2(Area)",
    color = "Laboratory"
  ) +
  theme_minimal()
```

### 2.3.3b Log2 Fold-Error: Proof of LOESS Correction

Per-peak Log2 Fold-Error $= \log_2(\text{Lab Area} / \text{Consensus Median Area})$.
Before normalization, labs drift from zero (especially Q3/Q4). After LOESS,
all labs re-center on zero if the systematic bias is corrected.

```{r log2-fold-error-calculation}
# Calculate consensus median area for each peak (across all labs)
consensus_areas <- pk_norm |>
  group_by(consensus_id, rt_decile) |>
  summarise(
    consensus_median_raw = median(areaUnderTic, na.rm = TRUE),
    consensus_median_global = median(area_global, na.rm = TRUE),
    consensus_median_loess = median(area_loess, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate log2 fold-error for each lab
fold_error <- pk_norm |>
  left_join(consensus_areas, by = c("consensus_id", "rt_decile")) |>
  mutate(
    log2_fe_raw = log2(areaUnderTic / consensus_median_raw),
    log2_fe_global = log2(area_global / consensus_median_global),
    log2_fe_loess = log2(area_loess / consensus_median_loess)
  ) |>
  filter(!is.na(rt_decile), !is.na(log2_fe_raw)) |>
  select(lab, consensus_id, rt_decile, log2_fe_raw, log2_fe_global, log2_fe_loess)

cat("Log2 Fold-Error dataset:", nrow(fold_error), "rows\n")
```

### 2.3.4 Log2 Fold-Error Visualization

```{r log2-fold-error-plot}
#| fig-cap: "**Figure 2.3**: Log2 Fold-Error by RT Decile and Lab, showing the effect of LOESS normalization. Before correction (Raw, left), labs drift far from zero in D7-D10 (ICL systematically low, Cembio hyper-sensitive in D8). After Global Median (middle), drift persists. After LOESS (right), all labs are perfectly re-centered on zero across all deciles—the proof that systematic RT-dependent bias is corrected."
fe_long <- fold_error |>
  pivot_longer(
    cols = starts_with("log2_fe_"),
    names_to = "strategy",
    names_prefix = "log2_fe_",
    values_to = "log2_fold_error"
  ) |>
  mutate(
    strategy = case_when(
      strategy == "raw" ~ "Raw Area",
      strategy == "global" ~ "Global Median",
      strategy == "loess" ~ "LOESS"
    ),
    strategy = factor(strategy, levels = c("Raw Area", "Global Median", "LOESS"))
  )

ggplot(fe_long, aes(x = rt_decile, y = log2_fold_error, fill = lab)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~strategy) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Log2 Fold-Error: Raw vs Normalized Areas",
    subtitle = "LOESS re-centers all labs on zero (systematic bias correction)",
    x = "RT Decile", y = "log2(Lab Area / Consensus Median)", fill = "Lab"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Findings**:
- **Raw Area**: ICL D1 +0.45, D4 +0.64, D5 −0.56, D7 −0.60, D8 −0.68;
  Cembio D8 +0.52, D10 −1.26.
- **Global Median**: drift persists (RT-specific bias unaddressed).
- **LOESS**: all labs re-centered on zero across deciles.

### 2.3.5 ICC by Normalization Strategy and RT Decile

ICC computed on **peak-level Area Under TIC** after each normalization
strategy, per RT decile, for **Detected Signal** (all consensus peaks) and
**Consensus Signal** (matched in all 4 labs).

```{r icc-peak-function}
compute_icc_peak <- function(data, value_col) {
  wide <- data |>
    filter(!is.na(.data[[value_col]])) |>
    select(consensus_id, lab, value = all_of(value_col)) |>
    mutate(value = log2(value)) |>
    pivot_wider(names_from = lab, values_from = value)

  # Filter to only rows with all 4 labs present
  # Get actual lab columns (may not always be all 4)
  lab_cols <- intersect(c("afekta", "hmgu", "icl", "cembio"), colnames(wide))
  wide <- wide |> filter(if_all(all_of(lab_cols), ~ !is.na(.x)))

  if (nrow(wide) < 10) {
    return(data.frame(
      Metric = value_col, ICC = NA_real_,
      Between_var = NA_real_, Within_var = NA_real_, n_peaks = nrow(wide)))
  }

  mat <- as.matrix(wide[, lab_cols])
  row_means <- rowMeans(mat, na.rm = TRUE)
  grand_mean <- mean(row_means, na.rm = TRUE)

  k <- ncol(mat); n <- nrow(mat)
  ss_between <- k * sum((row_means - grand_mean)^2)
  ss_within <- sum(sweep(mat, 1, row_means)^2, na.rm = TRUE)

  ms_between <- ss_between / (n - 1)
  ms_within <- ss_within / (n * (k - 1))
  icc_val <- (ms_between - ms_within) / (ms_between + (k - 1) * ms_within)

  data.frame(
    Metric = value_col, ICC = icc_val,
    Between_var = ms_between, Within_var = ms_within, n_peaks = n
  )
}
```

```{r icc-normalization-comparison}
# Prepare Detected Signal: all consensus peaks
detected_signal <- pk_norm |> filter(!is.na(rt_decile))

# Prepare Consensus Signal: subset of consensus peaks that are confirmed across all labs
# This represents only the peaks with matched signals in all 4 labs
consensus_signal <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(consensus_id) |>
  filter(n() == 4) |>
  ungroup()

# Compute ICC per RT decile for each strategy and signal type
icc_by_decile <- do.call(rbind, lapply(1:10, function(d) {
  pk_detected <- detected_signal |> filter(rt_decile == paste0("D", d))
  pk_consensus <- consensus_signal |> filter(rt_decile == paste0("D", d))

  bind_rows(
    # Detected Signal
    compute_icc_peak(pk_detected, "areaUnderTic") |>
      mutate(Strategy = "Raw", Decile = paste0("D", d), Signal_Type = "Detected"),
    compute_icc_peak(pk_detected, "area_global") |>
      mutate(Strategy = "Global Median", Decile = paste0("D", d), Signal_Type = "Detected"),
    compute_icc_peak(pk_detected, "area_decile") |>
      mutate(Strategy = "Decile Median", Decile = paste0("D", d), Signal_Type = "Detected"),
    compute_icc_peak(pk_detected, "area_loess") |>
      mutate(Strategy = "LOESS", Decile = paste0("D", d), Signal_Type = "Detected"),
    # Consensus Signal
    compute_icc_peak(pk_consensus, "areaUnderTic") |>
      mutate(Strategy = "Raw", Decile = paste0("D", d), Signal_Type = "Consensus"),
    compute_icc_peak(pk_consensus, "area_global") |>
      mutate(Strategy = "Global Median", Decile = paste0("D", d), Signal_Type = "Consensus"),
    compute_icc_peak(pk_consensus, "area_decile") |>
      mutate(Strategy = "Decile Median", Decile = paste0("D", d), Signal_Type = "Consensus"),
    compute_icc_peak(pk_consensus, "area_loess") |>
      mutate(Strategy = "LOESS", Decile = paste0("D", d), Signal_Type = "Consensus")
  )
}))

# Also compute overall ICC for comparison
icc_overall <- bind_rows(
  compute_icc_peak(detected_signal, "areaUnderTic") |>
    mutate(Strategy = "Raw", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(detected_signal, "area_global") |>
    mutate(Strategy = "Global Median", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(detected_signal, "area_decile") |>
    mutate(Strategy = "Decile Median", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(detected_signal, "area_loess") |>
    mutate(Strategy = "LOESS", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(consensus_signal, "areaUnderTic") |>
    mutate(Strategy = "Raw", Decile = "Overall", Signal_Type = "Consensus"),
  compute_icc_peak(consensus_signal, "area_global") |>
    mutate(Strategy = "Global Median", Decile = "Overall", Signal_Type = "Consensus"),
  compute_icc_peak(consensus_signal, "area_decile") |>
    mutate(Strategy = "Decile Median", Decile = "Overall", Signal_Type = "Consensus"),
  compute_icc_peak(consensus_signal, "area_loess") |>
    mutate(Strategy = "LOESS", Decile = "Overall", Signal_Type = "Consensus")
)

icc_norm_results <- bind_rows(icc_by_decile, icc_overall)

# Summary table showing decile-specific improvements for both signal types
icc_table_detected <- icc_norm_results |>
  filter(Signal_Type == "Detected", Decile != "Overall") |>
  select(Strategy, Decile, ICC) |>
  pivot_wider(names_from = Decile, values_from = ICC) |>
  select(Strategy, D1, D2, D3, D4, D5, D6, D7, D8, D9, D10)

icc_table_consensus <- icc_norm_results |>
  filter(Signal_Type == "Consensus", Decile != "Overall") |>
  select(Strategy, Decile, ICC) |>
  pivot_wider(names_from = Decile, values_from = ICC) |>
  select(Strategy, D1, D2, D3, D4, D5, D6, D7, D8, D9, D10)

cat("\n**Detected Signal (all consensus peaks):**\n")
knitr::kable(icc_table_detected,
    caption = "Table 2.2a: ICC by normalization strategy and RT decile - Detected Signal",
    digits = 3)

cat("\n**Consensus Signal (matched peaks only):**\n")
knitr::kable(icc_table_consensus,
    caption = "Table 2.2b: ICC by normalization strategy and RT decile - Consensus Signal",
    digits = 3)
```

**Key Findings**:

1. Overall peak-level ICC (consensus peaks): **0.33 raw → 0.55 Global → 0.58 Decile → 0.58 LOESS**.
2. Local-normalization gains concentrated in D7–D10.
3. Consensus Signal ICC > Detected Signal ICC.
4. Decile Median ≈ LOESS (0.58 vs 0.58) but simpler.

```{r icc-normalization-line-plot}
#| fig-cap: "**Figure 2.4**: ICC by Normalization Strategy and RT Decile (Detected vs Consensus Signal). Line plot showing which deciles benefit most from local normalization. Detected Signal (left) shows larger improvements from local methods in D7–D10, while Consensus Signal (right) maintains more stable ICC across deciles."
#| fig-width: 14
#| fig-height: 6

icc_plot_data <- icc_norm_results |>
  filter(Decile != "Overall", !is.na(ICC)) |>
  mutate(
    Decile_num = as.numeric(gsub("D", "", Decile)),
    Strategy = factor(Strategy, levels = c("Raw", "Global Median", "Decile Median", "LOESS"))
  )

ggplot(icc_plot_data, aes(x = Decile_num, y = ICC, color = Strategy, linetype = Strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Signal_Type, labeller = labeller(Signal_Type = c(
    Detected = "Detected Signal (all peaks)",
    Consensus = "Consensus Signal (matched peaks)"
  ))) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray", alpha = 0.5) +
  scale_x_continuous(breaks = 1:10, labels = paste0("D", 1:10)) +
  scale_color_manual(values = c("Raw" = "#E64B35", "Global Median" = "#4DBBD5",
                                 "Decile Median" = "#70AD47", "LOESS" = "#4472C4")) +
  scale_linetype_manual(values = c("Raw" = "solid", "Global Median" = "solid",
                                    "Decile Median" = "solid", "LOESS" = "dashed")) +
  labs(
    title = "ICC by Normalization Strategy and RT Decile",
    subtitle = "Local normalization (Decile Median, LOESS) improves ICC most in later deciles (D7–D10) where RT-dependent bias is strongest",
    x = "RT Decile", y = "ICC (Area Under TIC)", color = "Strategy", linetype = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor.y = element_blank()
  ) +
  ylim(NA, 1)
```

```{r icc-overall-comparison}
#| fig-cap: "**Figure 2.5**: Overall ICC comparison across strategies (aggregated across all deciles). Detected Signal (left) shows more room for improvement through normalization, while Consensus Signal (right) has higher baseline ICC."
icc_norm_results |>
  filter(Decile == "Overall") |>
  mutate(Strategy = factor(Strategy, levels = c("Raw", "Global Median",
                                                 "Decile Median", "LOESS"))) |>
  ggplot(aes(x = Strategy, y = ICC, fill = Strategy)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Signal_Type, labeller = labeller(Signal_Type = c(
    Detected = "Detected Signal",
    Consensus = "Consensus Signal"
  ))) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_text(aes(label = round(ICC, 3)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Raw" = "#E64B35", "Global Median" = "#4DBBD5",
                               "Decile Median" = "#70AD47", "LOESS" = "#4472C4")) +
  labs(
    title = "Overall ICC by Normalization Strategy",
    subtitle = "Area Under TIC; dashed line at ICC = 0.5 threshold",
    y = "ICC", x = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylim(NA, 1)
```

### 2.3.6 CV by RT Decile and Strategy

```{r cv-by-strategy}
cv_strategies <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(consensus_id) |>
  filter(n() == 4) |>
  summarise(
    cv_raw = sd(areaUnderTic, na.rm = TRUE) /
      mean(areaUnderTic, na.rm = TRUE),
    cv_global = sd(area_global, na.rm = TRUE) /
      mean(area_global, na.rm = TRUE),
    cv_decile = sd(area_decile, na.rm = TRUE) /
      mean(area_decile, na.rm = TRUE),
    cv_loess = sd(area_loess, na.rm = TRUE) /
      mean(area_loess, na.rm = TRUE),
    rt_decile = first(rt_decile),
    .groups = "drop"
  )

cv_strat_long <- cv_strategies |>
  pivot_longer(cols = starts_with("cv_"),
               names_to = "Strategy", values_to = "CV",
               names_prefix = "cv_") |>
  mutate(Strategy = case_when(
    Strategy == "raw" ~ "Raw",
    Strategy == "global" ~ "Global Median",
    Strategy == "decile" ~ "Decile Median",
    Strategy == "loess" ~ "LOESS"
  ))

cv_strat_long |>
  group_by(Strategy) |>
  summarise(Median_CV = median(CV, na.rm = TRUE),
            Mean_CV = mean(CV, na.rm = TRUE), .groups = "drop") |>
  knitr::kable(caption = "CV summary by normalization strategy", digits = 3)
```

```{r cv-quartile-strategy-plot}
#| fig-cap: "**Figure 2.6**: CV by RT Decile and strategy. Global, Decile, and LOESS normalization show similar performance across all deciles."
cv_strat_summary <- cv_strat_long |>
  filter(!is.na(CV), CV < 2) |>
  group_by(Strategy, rt_decile) |>
  summarise(Median_CV = median(CV, na.rm = TRUE), .groups = "drop")

ggplot(cv_strat_summary, aes(x = rt_decile, y = Median_CV, fill = Strategy)) +
  geom_col(position = "dodge", alpha = 0.7) +
  geom_text(aes(label = round(Median_CV, 2)),
            position = position_dodge(0.9), vjust = -0.3, size = 2.5) +
  labs(
    title = "Median Inter-Lab CV by RT Decile and Normalization Strategy",
    subtitle = "Global vs local normalization effects across retention time bins",
    x = "RT Decile", y = "CV (Area Under TIC)", fill = "Strategy"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

Median inter-lab CV: 0.97 raw → 0.82–0.84 normalized.

**Open question for Part 3:** LOESS removes systematic RT-dependent bias
(Figure 2.3) but ICC gains are modest. Hypothesis: residual integration noise
(baseline boundaries, scan-to-scan fluctuation) sets the ICC ceiling, not
systematic bias. Part 3 tests this by replacing scalar integration metrics
with EIC profile similarity.

---

# Part 3: Peak Shape Metrics vs EIC Profile Correlation {#strategy3}

**Core question:** Do scalar peak shape metrics (FWHM, peak width, asymmetry,
etc.) capture the same cross-lab reproducibility information as the full EIC
Pearson profile similarity?

## 3.1 Load Data

```{r part3-data-setup}
# Peak shape metrics (scalar summaries per peak per lab)
pk_metrics <- read.csv(file.path(base_path,
                                 "data/peak_metrics_detected_all4labs.csv"))

# EIC profile similarity (Pearson & Tanimoto of raw rt/int values per lab pair)
eic_sim <- read.csv(file.path(base_path,
                               "data/peak_shape_similarity.csv"))
```

## 3.2 Cross-Lab Correlation of Scalar Metrics

Pairwise Pearson correlations per scalar metric across consensus peaks.

```{r metric-cross-lab-correlation}
#| fig-cap: "**Figure 3.1**: Cross-lab Pearson correlation of scalar peak metrics for all 6 lab pairs. Area and maxIntensity show moderate-to-strong correlations; shape metrics (FWHM, peakWidth, peakBeta1) are weakly correlated across labs."
#| fig-width: 10
#| fig-height: 6

metrics_to_test <- c("areaUnderTic", "maxIntensity", "xicFwhm",
                     "peakWidth", "peakBeta1", "signalToNoiseRatio")

# Keep only consensus peaks present in all 4 labs
pk_complete <- pk_metrics |>
  group_by(consensus_id) |>
  filter(n() == 4) |>
  ungroup()

# For each metric, compute pairwise lab-lab Pearson correlation
metric_corr <- map_dfr(metrics_to_test, function(m) {
  wide <- pk_complete |>
    select(consensus_id, lab, all_of(m)) |>
    pivot_wider(names_from = lab, values_from = all_of(m))

  labs <- setdiff(names(wide), "consensus_id")
  pairs <- combn(labs, 2, simplify = FALSE)

  map_dfr(pairs, function(p) {
    vals <- wide |> select(all_of(p)) |> drop_na()
    r <- cor(vals[[1]], vals[[2]], use = "complete.obs")
    tibble(metric = m, pair = paste(p, collapse = ":"), r = r)
  })
})

ggplot(metric_corr, aes(x = metric, y = r, color = pair)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title = "Cross-Lab Pearson Correlation of Scalar Peak Metrics",
    subtitle = "Each point is one lab pair; consensus peaks detected in all 4 labs",
    x = "Metric", y = "Pearson r", color = "Lab Pair"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 3.3 EIC Profile Similarity Distribution by Lab Pair

`peak_shape_similarity.csv` holds Pearson correlations of aligned EIC
rt/intensity profiles, per consensus peak per lab pair.

```{r eic-similarity-distribution}
#| fig-cap: "**Figure 3.2**: Distribution of EIC Pearson similarity across consensus peaks, faceted by lab pair. Higher values indicate that the raw chromatographic profile of a peak is reproduced across labs."
#| fig-width: 10
#| fig-height: 6

ggplot(eic_sim, aes(x = pearson, fill = pair)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  facet_wrap(~pair, scales = "free_y") +
  labs(
    title = "EIC Pearson Similarity Distribution by Lab Pair",
    subtitle = "Consensus peaks: Pearson correlation of raw EIC profiles",
    x = "EIC Pearson Correlation", y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## 3.4 Comparing Metrics Correlation vs EIC Pearson Similarity

Per (peak, lab pair): correlate the scalar metric's absolute cross-lab
difference with the EIC Pearson similarity.

```{r metrics-vs-eic}
#| fig-cap: "**Figure 3.3**: Scatter of EIC Pearson similarity vs absolute cross-lab difference for selected scalar metrics. A strong negative trend would indicate the metric captures profile-level reproducibility."
#| fig-width: 12
#| fig-height: 8

# Compute pairwise metric differences per consensus peak per lab pair
lab_pairs <- eic_sim |> distinct(pair) |> pull(pair)

metric_diffs <- map_dfr(lab_pairs, function(p) {
  labs_in_pair <- str_split(p, ":")[[1]]
  lab_a <- pk_metrics |> filter(lab == labs_in_pair[1])
  lab_b <- pk_metrics |> filter(lab == labs_in_pair[2])

  joined <- inner_join(lab_a, lab_b, by = "consensus_id", suffix = c("_a", "_b"))

  tibble(
    consensus_id = joined$consensus_id,
    pair = p,
    area_diff = abs(log2(joined$areaUnderTic_a / joined$areaUnderTic_b)),
    fwhm_diff = abs(log2(joined$xicFwhm_a / joined$xicFwhm_b)),
    width_diff = abs(log2(joined$peakWidth_a / joined$peakWidth_b)),
    beta_diff = abs(joined$peakBeta1_a - joined$peakBeta1_b)
  )
})

# Join with EIC Pearson similarity
metric_vs_eic <- metric_diffs |>
  inner_join(eic_sim, by = c("consensus_id", "pair"))

# Pivot to long for faceted scatter
metric_vs_eic_long <- metric_vs_eic |>
  pivot_longer(cols = c(area_diff, fwhm_diff, width_diff, beta_diff),
               names_to = "metric", values_to = "abs_diff") |>
  filter(is.finite(abs_diff))

ggplot(metric_vs_eic_long, aes(x = abs_diff, y = pearson)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.8) +
  facet_wrap(~metric, scales = "free_x") +
  labs(
    title = "Scalar Metric Difference vs EIC Pearson Similarity",
    subtitle = "Each point is one consensus peak in one lab pair",
    x = "Absolute Metric Difference (log2 scale where applicable)",
    y = "EIC Pearson Correlation"
  ) +
  theme_minimal()
```

## 3.5 Summary Correlation Table

```{r metrics-vs-eic-summary}
metric_vs_eic_summary <- metric_vs_eic_long |>
  group_by(metric) |>
  summarise(
    r_with_eic_pearson = cor(abs_diff, pearson, use = "complete.obs"),
    n_peaks = n(),
    .groups = "drop"
  ) |>
  arrange(r_with_eic_pearson)

knitr::kable(
  metric_vs_eic_summary,
  caption = "Correlation between scalar metric differences and EIC Pearson similarity",
  digits = 3
)
```

## 3.6 Lab-Pair Comparison: Median EIC Similarity vs Median Metric Correlation

```{r lab-pair-summary}
#| fig-cap: "**Figure 3.4**: Median EIC Pearson similarity vs median scalar metric cross-lab correlation per lab pair. Same-platform pairs (e.g., afekta:cembio) should appear in the upper-right."
#| fig-width: 8
#| fig-height: 6

# Median EIC Pearson per pair
eic_pair_summary <- eic_sim |>
  group_by(pair) |>
  summarise(median_eic_pearson = median(pearson, na.rm = TRUE), .groups = "drop")

# Median area correlation per pair (as representative metric)
area_pair_summary <- metric_corr |>
  filter(metric == "areaUnderTic") |>
  select(pair, r_area = r)

pair_comparison <- eic_pair_summary |>
  left_join(area_pair_summary, by = "pair")

ggplot(pair_comparison, aes(x = r_area, y = median_eic_pearson, label = pair)) +
  geom_point(size = 4, color = "steelblue") +
  ggrepel::geom_text_repel(size = 3) +
  labs(
    title = "Lab Pair: Scalar Area Correlation vs Median EIC Pearson",
    x = "Cross-Lab Pearson r (log2 Area)",
    y = "Median EIC Pearson Similarity"
  ) +
  theme_minimal()
```

**Findings:**
- Area and maxIntensity: moderate cross-lab correlation.
- Shape metrics (FWHM, peakWidth, peakBeta1): weak cross-lab correlation.
- Weak correlation between scalar metric differences and EIC Pearson →
  scalar summaries miss profile-level information.
- Same-platform pairs (afekta:cembio) expected top-right; consistent with
  the Part 4 hardware ceiling.

---

# Summary {#summary}

## Key Takeaways

1. **Peak matching is the true bottleneck.** Raw TIC ICC **0.002 → 0.024 → 0.502**
   across Full → Detected → Consensus (positive mode).
2. **Local normalization fixes bias but not noise.** LOESS re-centers labs on
   zero (Figure 2.3) but overall ICC gain plateaus (0.33 → 0.58).
3. **Scalar shape metrics are weakly correlated across labs; EIC profile
   similarity captures more.** Integration-boundary noise dominates scalar
   summaries.

# Part 4: The Hardware Limitation - Defining the Reproducibility Ceiling {#hardware}

**Core claim:** Instrument chemistry imposes an irreducible variance ceiling
even after optimal processing. Same-platform ICC defines the upper bound.

## 4.1 Platform Effects on Peak Intensity Correlations

### 4.1.1 Cross-Lab Correlation: Peak Area

Pairwise cross-lab Pearson correlations on consensus peaks.

```{r area-correlation-matrix}
#| fig-cap: "**Figure 4.1**: Cross-lab correlation matrix for Peak Area (log2-transformed). Same-platform correlations are high (~0.81), while cross-platform pairs are moderate (0.43–0.54)."
#| fig-width: 8
#| fig-height: 7

library(GGally)

# Pivot to wide format: rows = consensus_id, columns = lab
area_wide <- pk_metrics |>
  filter(!is.na(areaUnderTic), areaUnderTic > 0) |>
  group_by(consensus_id) |>
  filter(n() == 4) |>  # Keep only peaks detected in all 4 labs
  ungroup() |>
  select(consensus_id, lab, areaUnderTic) |>
  mutate(log2_area = log2(areaUnderTic)) |>
  select(-areaUnderTic) |>
  pivot_wider(names_from = lab, values_from = log2_area)

# Create correlation matrix plot
ggpairs(
  area_wide |> select(-consensus_id),
  title = "Cross-Lab Correlation Matrix: Peak Area (log2)",
  upper = list(continuous = wrap("cor", size = 5, color = "blue")),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.5))
) +
  theme_minimal()
```

**Findings**:
- Agilent-Agilent (afekta-cembio): r = 0.81
- Cross-platform (Agilent vs Waters/Sciex): r = 0.43–0.54
- Waters-Sciex (hmgu-icl): r = 0.43

Same-platform r is ~50% higher than cross-platform; all r > 0.4.

## 4.2 The Ceiling: Same-Platform vs Cross-Platform ICC Across RT Deciles

Same-platform ICC (Agilent-Agilent) vs cross-platform ICC (all 4 labs), by
RT decile. The gap defines the hardware ceiling.

```{r hardware-ceiling-analysis}
#| fig-cap: "**Figure 4.2**: ICC by RT Decile for Same-Platform (Agilent-Agilent) vs Cross-Platform (all 4 labs). Same-platform labs maintain high ICC (0.80–0.95) across all deciles, while cross-platform ICC drops in late deciles (D8–D10), revealing where instrument chemistry dominates."
#| fig-width: 12
#| fig-height: 7

# Ensure pk_norm is available with rt_decile and area_loess columns (from Part 2)
if (!exists("pk_norm") || !("rt_decile" %in% names(pk_norm))) {
  stop("pk_norm with rt_decile column must be loaded from Part 2")
}

# Function to compute ICC by RT decile
compute_icc_by_decile <- function(pk_data, lab_filter = NULL, label = "") {
  if (!is.null(lab_filter)) {
    pk_data <- pk_data |> filter(lab %in% lab_filter)
  }

  results <- pk_data |>
    filter(!is.na(rt_decile), !is.na(area_loess)) |>
    group_by(consensus_id, rt_decile) |>
    filter(n() == length(lab_filter %||% unique(pk_data$lab))) |>
    ungroup() |>
    group_by(rt_decile) |>
    summarise(
      n_peaks = n_distinct(consensus_id),
      icc = tryCatch({
        wide <- current_data() |>
          select(lab, area_loess) |>
          pivot_wider(names_from = lab, values_from = area_loess)

        # Compute ICC(2,k) for this decile
        k <- ncol(wide)
        if (k < 2 || nrow(wide) < 2) {
          NA_real_
        } else {
          wide_mat <- as.matrix(wide)
          grand_mean <- mean(wide_mat, na.rm = TRUE)
          target_means <- rowMeans(wide_mat, na.rm = TRUE)
          ms_targets <- sum((target_means - grand_mean)^2, na.rm = TRUE) / (nrow(wide) - 1)
          residuals <- wide_mat - target_means
          ms_error <- sum(residuals^2, na.rm = TRUE) / ((nrow(wide) - 1) * (k - 1))
          (ms_targets - ms_error) / (ms_targets + (k - 1) * ms_error)
        }
      }, error = function(e) NA_real_),
      .groups = "drop"
    )

  if (!is.null(lab_filter)) {
    results <- results |> mutate(Configuration = label)
  }
  return(results)
}

# Compute ICC by decile for Same-Platform (Agilent only: afekta + cembio)
same_platform_by_decile <- pk_norm |>
  filter(lab %in% c("afekta", "cembio")) |>
  filter(!is.na(rt_decile), !is.na(area_loess)) |>
  group_by(consensus_id, rt_decile) |>
  filter(n() == 2) |>  # Only peaks in both Agilent labs
  ungroup() |>
  group_by(rt_decile) |>
  summarise(
    n_peaks = n_distinct(consensus_id),
    icc = tryCatch({
      wide <- current_data() |>
        select(lab, area_loess) |>
        pivot_wider(names_from = lab, values_from = area_loess)

      if (nrow(wide) < 2) {
        NA_real_
      } else {
        wide_mat <- as.matrix(wide)
        grand_mean <- mean(wide_mat, na.rm = TRUE)
        target_means <- rowMeans(wide_mat, na.rm = TRUE)
        ms_targets <- sum((target_means - grand_mean)^2, na.rm = TRUE) / (nrow(wide) - 1)
        residuals <- wide_mat - target_means
        ms_error <- sum(residuals^2, na.rm = TRUE) / (nrow(wide) - 1)
        (ms_targets - ms_error) / (ms_targets + ms_error)
      }
    }, error = function(e) NA_real_),
    .groups = "drop"
  ) |>
  mutate(Configuration = "Same-Platform\n(Agilent-Agilent)")

# Compute ICC by decile for Cross-Platform (all 4 labs)
cross_platform_by_decile <- pk_norm |>
  filter(!is.na(rt_decile), !is.na(area_loess)) |>
  group_by(consensus_id, rt_decile) |>
  filter(n() == 4) |>  # Only peaks detected in all 4 labs
  ungroup() |>
  group_by(rt_decile) |>
  summarise(
    n_peaks = n_distinct(consensus_id),
    icc = tryCatch({
      wide <- current_data() |>
        select(lab, area_loess) |>
        pivot_wider(names_from = lab, values_from = area_loess)

      if (nrow(wide) < 2) {
        NA_real_
      } else {
        wide_mat <- as.matrix(wide)
        grand_mean <- mean(wide_mat, na.rm = TRUE)
        target_means <- rowMeans(wide_mat, na.rm = TRUE)
        ms_targets <- sum((target_means - grand_mean)^2, na.rm = TRUE) / (nrow(wide) - 1)
        residuals <- wide_mat - target_means
        ms_error <- sum(residuals^2, na.rm = TRUE) / ((nrow(wide) - 1) * 3)
        (ms_targets - ms_error) / (ms_targets + 3 * ms_error)
      }
    }, error = function(e) NA_real_),
    .groups = "drop"
  ) |>
  mutate(Configuration = "Cross-Platform\n(4 labs, 4 vendors)")

# Combine results
ceiling_by_decile <- bind_rows(
  same_platform_by_decile,
  cross_platform_by_decile
) |>
  mutate(
    Configuration = factor(Configuration,
                          levels = c("Same-Platform\n(Agilent-Agilent)",
                                    "Cross-Platform\n(4 labs, 4 vendors)"))
  )

ggplot(ceiling_by_decile, aes(x = rt_decile, y = icc, color = Configuration,
                               linetype = Configuration, size = Configuration)) +
  geom_line(alpha = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Same-Platform\n(Agilent-Agilent)" = "#4DBBD5",
                                 "Cross-Platform\n(4 labs, 4 vendors)" = "#E64B35")) +
  scale_linetype_manual(values = c("Same-Platform\n(Agilent-Agilent)" = "solid",
                                    "Cross-Platform\n(4 labs, 4 vendors)" = "dashed")) +
  scale_size_manual(values = c("Same-Platform\n(Agilent-Agilent)" = 1.0,
                               "Cross-Platform\n(4 labs, 4 vendors)" = 0.8)) +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "gray50", alpha = 0.5) +
  scale_y_continuous(limits = c(0, 1.0), breaks = seq(0, 1, 0.2)) +
  labs(
    title = "The Hardware Ceiling Across RT Deciles",
    subtitle = "Same-Platform ICC remains high (0.80–0.95) across deciles; Cross-Platform ICC drops in late deciles (instrument chemistry effect)",
    x = "RT Decile", y = "ICC", color = "Configuration", linetype = "Configuration"
  ) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
        axis.text.x = element_text(angle = 0))
```



## 4.3 Implications for Epidemiological Study Design

This hardware ceiling has direct consequences for multi-center metabolomics:
...


## Notes for Collaborator (2026-03-12)

### What is done
- **Part 1** (Feature extraction): Complete. Variance decomposition, PCA, and
  ICC analysis across Full → Detected → Consensus Signal levels all working.
- **Part 2** (RT-dependent normalization): Complete. Global median, decile median,
  and LOESS normalization strategies implemented and compared. Needs to be
  properly interpreted and integrated into the narrative.
- **Part 3** (Peak metrics vs EIC Pearson): Rewritten. Now does simple
  cross-lab correlation of scalar metrics vs EIC profile similarity for
  consensus peaks. Uses `peak_metrics_detected_all4labs.csv` and
  `peak_shape_similarity.csv`. No ML/classification code remains.
- **Part 4** (Hardware ceiling): Same-platform vs cross-platform ICC by RT
  decile is implemented for afekta:cembio (Agilent-Agilent) vs all 4 labs.

### What needs to be finished
1. **part 2 interpretation**:TO be done. I am not sure tha tLoess is reall y
the best.
1. **Part 3**: Run and verify the new code. Check that `ggrepel` is available
   (used in Figure 3.4). Interpret the results once rendered—the narrative text
   is placeholder and should be updated with actual numbers.
2. **Part 4 (Ceiling analysis)**: Currently only compares Agilent-Agilent vs
   all-4-labs.
   Section 4.3 ("Implications for Epidemiological Study Design") is still empty.
3. **Summary section**: Key takeaway #2 has a TODO note ("Conclusion should be
   about how normalization might fix systematic bias but cannot fix noise")—
   needs to be rewritten as a proper conclusion.
4. **Reproducibility metrics**: Currently using ICC and variance decomposition.
   Open question whether CCC (concordance correlation coefficient) or other
   agreement measures would be more appropriate—worth discussing.

### Possible extensions
- Add other chemical classes beyond the HUMAN endosome standard mixes.
- Test with a second peak-picking algorithm (e.g., MZmine centWave vs XCMS
  centWave) to see if the conclusions are algorithm-dependent.
- Tanimoto similarity from `peak_shape_similarity.csv` is loaded but not yet
  used in Part 3—could be added as an additional comparison alongside Pearson.

