---
title: "Defining and Overcoming the Limits of Multi-Center Metabolomics: A Computational Blueprint for Data Integration"
format:
  html:
    toc: true
    toc-location: right
    code-fold: true
execute:
  dir: project
---

# Introduction

## Input Data

All required input data lives in the `data/` subfolder. These files are
produced by the upstream pipeline (`1_object_creation.qmd` →
`2_global_analysis.qmd` → `3a_peak_feature_calculation_detected.qmd`).
I can also provide them pre-computed.

**RData objects** (chromatographic profiles, from `1_object_creation.qmd`):
`bpc_full.RData`, `bpc_detect.RData`, `tic_full.RData`, `tic_detect.RData`

**CSV files** (from `1_object_creation.qmd` and `3a_peak_feature_calculation_detected.qmd`):
`detected_peaks_{afekta,cembio,hmgu,icl}_HE.csv`,
`peak_metrics_detected_all4labs.csv`, `peak_shape_similarity.csv`

**Helper files** (in `5_downstream_analysis/`):
`global_analysis_helpers.R`, `lab_setup_phase1.csv`

## The Epidemiological Imperative and the Technical Barrier

Large-scale epidemiological studies require multi-center designs to achieve
sufficient statistical power and population diversity. However, untargeted
metabolomics—a cornerstone of modern molecular epidemiology—faces a critical
challenge: **raw LC-MS data from different laboratories is too confounded by
instrumental noise to be directly integrated.**

This document addresses the core question: *What are the computational steps
required to make multi-center metabolomics data integration possible, and what
are the absolute hardware-imposed limits of the field?*

We benchmark four sequential processing strategies using 19 standard mixtures
(HUMAN endosome) analyzed across 4 labs with harmonized LC systems (same column,
mobile phases, gradient):

1. **Peak-picking** (feature extraction vs. raw profiles)
2. **Local (RT-dependent) normalization** (LOESS vs. global scaling)
3. **Peak shape confirmation** (EIC Pearson similarity for cross-lab anchors)
4. **Hardware ceiling** (same-platform ICC as the achievable upper bound)

Data: 19 standard mixtures from 4 labs with harmonized protocols but different
instrument platforms (Agilent Q-TOF, Waters Synapt, Sciex Zeno, Bruker TIMS).
No technical replicates; lab identity is confounded with instrumental parameters.

## The Challenge: Why Raw Data Integration Fails

Multi-center metabolomics studies face a fundamental problem: the instruments
that define the field (LC-MS systems) are inherently variable. Even with
"harmonized" protocols—identical columns, mobile phases, and gradients—
systematic differences emerge from:

1. **Ionization efficiency**: Electrospray source geometry, gas flows, and
   temperature control vary by manufacturer, creating lab-specific sensitivity
   profiles that change across the chromatographic run.

2. **Data acquisition**: Vendor-specific algorithms for centroiding, baseline
   subtraction, and intensity calculation mean that "raw" data is already
   processed differently before peak detection.

3. **Instrumental variance confounding**: Without technical replicates across
   labs, we cannot separate instrument effects from operator effects,
   environmental conditions, or batch-to-batch reagent variability.

The consequence: **lab identity explains 89-100% of variance in raw TIC**,
leaving almost nothing for biological signal. This technical noise makes
direct pooling of patient samples from different labs statistically invalid.
The question is whether computational processing can recover biological signal,
and if so, what reproducibility ceiling is imposed by hardware differences.

This document addresses three complementary approaches to data integration,
culminating in a definition of the hardware-imposed reproducibility ceiling.

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
##The install belows are temporary and should be replaced with proper dependencies in the future.
BiocManager::install("RforMassSpectrometry/Chromatograms", ask = FALSE, update = TRUE)
BiocManager::install(("philouail/MsQuality", ref = "Chromatgorams"), ask = FALSE, update = TRUE)
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)
library(pheatmap)

base_path <- here("5_downstream_analysis")
source(file.path(base_path, "global_analysis_helpers.R"))
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup |>
  filter(details == "harmonized") |>
  select(lab, `LC.Instrument`, `MS.Instrument`,
         `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) |>
  distinct() |>
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) |>
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

## Compute Global Metrics

Compute TIC/BPC for both full signal and detected signal across both
polarities, along with normalization variants.

```{r compute-metrics}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "data/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]
    load(file.path(base_path, "data/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "data/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]
    load(file.path(base_path, "data/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) |>
    left_join(setup_simplified, by = "lab") |>
    mutate(data_level = ifelse(data_level == "full",
                               "Full Signal", "Detected Signal")) |>
    group_by(lab, polarity) |>
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm = bpc / median_bpc,
      tic_norm = tic / median_tic
    ) |>
    ungroup()
}

metrics_full <- run_for_polarities(compute_metrics, "full")
metrics_detect <- run_for_polarities(compute_metrics, "detect")

# Create Consensus Signal data level from pk_metrics
compute_consensus_metrics <- function(polarity, polarity_name, data_level = "detect") {
  # Load pk_metrics (consensus peaks only)
  pk_metrics <- read.csv(file.path(base_path,
                                   "data/peak_metrics_detected_all4labs.csv"))

  # Sum areaUnderTic per lab and mixture (Consensus TIC)
  consensus_data <- pk_metrics |>
    group_by(lab, mixture) |>
    summarise(
      tic = sum(areaUnderTic, na.rm = TRUE),
      .groups = "drop"
    ) |>
    left_join(setup_simplified, by = "lab") |>
    mutate(
      polarity = polarity_name,
      data_level = "Consensus Signal",
      bpc = NA_real_  # Not applicable for consensus peaks
    ) |>
    group_by(lab, polarity) |>
    mutate(
      median_tic = median(tic, na.rm = TRUE),
      tic_norm = tic / median_tic
    ) |>
    ungroup()

  return(consensus_data)
}

metrics_consensus <- run_for_polarities(compute_consensus_metrics, "detect")

metrics_combined <- bind_rows(metrics_full, metrics_detect, metrics_consensus) |>
  group_by(lab, polarity, data_level) |>
  mutate(
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE)
  ) |>
  ungroup()
```

## Data Processing Levels

This analysis uses three progressive levels of data processing:

**Full Signal**: Raw chromatographic profiles including all detected ions and baseline noise. This represents the complete LC-MS data as acquired by each instrument, including vendor-specific baseline artifacts, electronic noise, and column bleed.

**Detected Signal**: Chromatographic peaks identified by centWave peak-picking algorithm. Feature extraction isolates compound-specific signals from baseline noise by identifying regions with consistent m/z and retention time characteristics. This level includes all detected peaks regardless of whether they match across laboratories.

**Consensus Signal**: Subset of peaks that match across all four laboratories based on m/z and retention time alignment (consensus peaks). These represent compounds reproducibly detected by all sites, providing the foundation for cross-laboratory quantitative comparisons.

The progression from Full → Detected → Consensus Signal tests whether computational processing can progressively remove technical variance and recover biological signal reproducibility.

---

# Part 1: The Baseline Problem - Feature Extraction is Mandatory {#part1}

**Core claim:** Raw chromatographic profiles cannot be integrated across labs
because vendor-specific baseline noise dominates the signal. Feature extraction
(peak-picking) is the mandatory first computational step to make data integration
theoretically possible.

## 1.1 The Problem: Lab Variance Dominates Raw Data

### 1.1.1 Variance Decomposition

```{r variance-decomposition}
results_table <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

```{r variance-stacked-bar}
#| fig-cap: "**Figure 1.1**: Variance decomposition of TIC. Lab identity explains 89–100% of raw signal variance — the core problem."
variance_long <- results_table |>
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) |>
  pivot_longer(cols = ends_with("_Pct"),
               names_to = "Source", values_to = "Percentage") |>
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Label = paste(Polarity, Data_Level, sep = "\n"),
    R2_label = paste("R² =", round(R_squared, 3))
  )

text_labels <- variance_long |> distinct(Label, R2_label)

ggplot(variance_long, aes(x = Label, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(x = Label, label = R2_label, y = 105),
            data = text_labels, size = 3, inherit.aes = FALSE) +
  labs(title = "Variance Decomposition: TIC Across All Conditions",
       subtitle = "Lab variance (orange) dominates raw signal — the core problem",
       x = "Polarity & Data Level", y = "Variance (%)", fill = "Source") +
  theme_minimal() +
  theme(legend.position = "top") +
  ylim(0, 110)
```

### 1.1.2 Lab Effect Coefficients

```{r lab-coefficients}
all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    all_coefficients <- rbind(all_coefficients,
      result$coefficients |> mutate(Polarity = pol, Data_Level = level))
  }
}

lab_coefs <- all_coefficients |>
  filter(type == "Lab Effect") |>
  mutate(lab_name = gsub("lab", "", coefficient))

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "Lab Effect Coefficients (deviation from Afekta)",
       x = "Laboratory", y = "Coefficient (log2 scale)", fill = "Lab") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.1.3 PCA Analysis of Lab vs. Mixture Clustering

Principal Component Analysis (PCA) visualizes the relative contribution of
technical vs. biological variance. If technical noise dominates, samples cluster
by **lab** (color). If biological signal is recovered, samples cluster by
**mixture** (shape).

```{r pca-setup}
# Prepare data for PCA: wide format with samples as rows, features as columns
# For raw data: use individual RT bins as features
# For detected data: use detected peaks as features

# Function to prepare PCA data
prepare_pca_data <- function(data_level = "full", polarity = "Positive") {
  if (data_level == "full") {
    # Load full signal data
    load(file.path(base_path, "data/tic_full.RData"))
    chr_data <- tic_full[tic_full$polarity == ifelse(polarity == "Positive", 1, 0)]
  } else {
    # Load detected signal data
    load(file.path(base_path, "data/tic_detect.RData"))
    chr_data <- tic_detect[tic_detect$polarity == ifelse(polarity == "Positive", 1, 0)]
  }

  # Extract intensity matrices
  # Each row = one sample, columns = RT bins
  int_list <- intensity(chr_data)
  rt_list <- rtime(chr_data)

  # Create common RT grid
  all_rts <- unlist(rt_list)
  rt_grid <- seq(min(all_rts, na.rm = TRUE),
                 max(all_rts, na.rm = TRUE),
                 length.out = 500)

  # Interpolate each chromatogram onto the grid
  int_matrix <- t(sapply(seq_along(int_list), function(i) {
    approx(rt_list[[i]], int_list[[i]], xout = rt_grid,
           method = "linear", rule = 2)$y
  }))

  # Add sample metadata
  rownames(int_matrix) <- paste0(chr_data$lab, "_", chr_data$mixture)
  colnames(int_matrix) <- paste0("RT_", round(rt_grid, 1))

  # Create metadata data frame
  meta_df <- data.frame(
    sample = rownames(int_matrix),
    lab = chr_data$lab,
    mixture = chr_data$mixture,
    row.names = rownames(int_matrix)
  )

  list(data = int_matrix, meta = meta_df)
}

# Prepare both datasets
pca_full <- prepare_pca_data("full", "Positive")
pca_detect <- prepare_pca_data("detect", "Positive")

# Prepare consensus data from pk_metrics (wide format: samples x consensus peaks)
pk_metrics <- read.csv(file.path(base_path,
                                 "data/peak_metrics_detected_all4labs.csv"))

# Create wide matrix: rows = samples (lab_mixture), columns = consensus peaks
pca_consensus_wide <- pk_metrics |>
  mutate(sample = paste0(lab, "_", mixture)) |>
  select(sample, consensus_id, areaUnderTic, lab, mixture) |>
  pivot_wider(names_from = consensus_id, values_from = areaUnderTic,
              values_fill = 0, names_prefix = "peak_") |>
  arrange(sample)

pca_consensus_meta <- pca_consensus_wide |>
  select(sample, lab, mixture)

pca_consensus_data <- pca_consensus_wide |>
  select(starts_with("peak_")) |>
  as.matrix()

rownames(pca_consensus_data) <- pca_consensus_wide$sample

pca_consensus <- list(
  data = pca_consensus_data,
  meta = pca_consensus_meta
)
```

```{r pca-raw}
#| fig-cap: "**Figure 1.2a**: PCA of Raw TIC profiles. Samples cluster perfectly by LAB (color), proving technical noise dominates. Mixture identity (shape) is invisible."
#| fig-width: 8
#| fig-height: 6

# Run PCA on raw data using prcomp
pca_res_full <- prcomp(pca_full$data, scale. = TRUE, center = TRUE)

# Calculate variance explained
var_explained <- (pca_res_full$sdev^2 / sum(pca_res_full$sdev^2)) * 100

# Create PCA scores dataframe
pca_scores <- data.frame(
  PC1 = pca_res_full$x[, 1],
  PC2 = pca_res_full$x[, 2],
  lab = pca_full$meta$lab,
  mixture = pca_full$meta$mixture
)

# Plot with ggplot2
ggplot(pca_scores, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(size = 3) +
  labs(
    title = "PCA: Raw TIC Profiles (Full Signal)",
    subtitle = "Perfect clustering by LAB → technical noise dominates",
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    color = "Laboratory"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

```{r pca-detected}
#| fig-cap: "**Figure 1.2b**: PCA of Detected Peaks. Lab clustering weakens and mixture patterns emerge, demonstrating that feature extraction recovers biological signal."
#| fig-width: 8
#| fig-height: 6

# Run PCA on detected data using prcomp
pca_res_detect <- prcomp(pca_detect$data, scale. = TRUE, center = TRUE)

# Calculate variance explained
var_explained_detect <- (pca_res_detect$sdev^2 / sum(pca_res_detect$sdev^2)) * 100

# Create PCA scores dataframe
pca_scores_detect <- data.frame(
  PC1 = pca_res_detect$x[, 1],
  PC2 = pca_res_detect$x[, 2],
  lab = pca_detect$meta$lab,
  mixture = pca_detect$meta$mixture
)

# Plot with ggplot2
ggplot(pca_scores_detect, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(size = 3) +
  labs(
    title = "PCA: Detected Peaks (Peak-Picked Signal)",
    subtitle = "Lab clusters weaken, mixture patterns emerge",
    x = paste0("PC1 (", round(var_explained_detect[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained_detect[2], 1), "%)"),
    color = "Laboratory"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

```{r pca-consensus}
#| fig-cap: "**Figure 1.2c**: PCA of Consensus Peaks. Samples cluster by mixture with reduced lab effects, demonstrating successful cross-lab integration."
#| fig-width: 8
#| fig-height: 6

# Run PCA on consensus data using prcomp
pca_res_consensus <- prcomp(pca_consensus$data, scale. = TRUE, center = TRUE)

# Calculate variance explained
var_explained_consensus <- (pca_res_consensus$sdev^2 / sum(pca_res_consensus$sdev^2)) * 100

# Create PCA scores dataframe
pca_scores_consensus <- data.frame(
  PC1 = pca_res_consensus$x[, 1],
  PC2 = pca_res_consensus$x[, 2],
  lab = pca_consensus$meta$lab,
  mixture = pca_consensus$meta$mixture
)

# Plot with ggplot2
ggplot(pca_scores_consensus, aes(x = PC1, y = PC2, color = lab)) +
  geom_point(size = 3) +
  labs(
    title = "PCA: Consensus Peaks (Cross-Lab Matched Signal)",
    subtitle = "Mixture patterns dominate, lab effects minimized",
    x = paste0("PC1 (", round(var_explained_consensus[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained_consensus[2], 1), "%)"),
    color = "Laboratory"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

**Interpretation**

Figure 1.2a demonstrates that raw TIC profiles cluster by laboratory with
no visible mixture structure. This indicates that raw chromatographic data
cannot be directly integrated across laboratories, as technical variance
completely dominates biological signal.

Figure 1.2b shows that after peak-picking, laboratory clustering weakens and
mixture-specific patterns emerge. Feature extraction removes sufficient
vendor-specific noise to enable biological comparisons.

Figure 1.2c shows consensus peaks (matched across all four labs), where mixture
patterns are most prominent and laboratory effects are further reduced. This
represents the optimal data level for cross-laboratory quantitative analysis.

**Conclusion:** Sharing raw profile data across labs is not viable for
multi-center studies. Feature extraction (peak-picking) followed by cross-lab
matching is required for data integration.

## 1.3 ICC: Full Signal vs Detected Signal

### 1.3.1 ICC Computation

ICC quantifies the ratio of biological signal to total variance:

$$ICC = \frac{\sigma^2_{between}}{\sigma^2_{between} + \sigma^2_{within}}$$

- $\sigma^2_{between}$ = variance *between* mixtures (biological signal)
- $\sigma^2_{within}$ = variance *within* the same mixture across labs (technical noise)
- ICC > 0.5 → acceptable reproducibility

```{r icc-computation}
icc_all_results <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)

    results <- bind_rows(
      compute_icc(data_subset, "tic") |> mutate(Type = "Absolute"),
      compute_icc(data_subset, "tic_norm") |>
        mutate(Type = "Normalized", Metric = "tic")
    )

    if (!all(is.na(data_subset$bpc))) {
      results <- bind_rows(
        results,
        compute_icc(data_subset, "bpc") |> mutate(Type = "Absolute"),
        compute_icc(data_subset, "bpc_norm") |>
          mutate(Type = "Normalized", Metric = "bpc")
      )
    }

    results <- results |>
      mutate(Polarity = pol, Data_Level = level,
             Metric_Label = case_when(
               Metric == "tic" ~ "TIC",
               Metric == "bpc" ~ "BPC",
               TRUE ~ Metric))

    icc_all_results <- bind_rows(icc_all_results, results)
  }
}
```

### 1.3.2 Peak Detection Impact

```{r peak-detection-impact}
icc_comparison_tic <- icc_all_results |>
  filter(Type == "Normalized", Metric == "tic") |>
  select(Polarity, Data_Level, ICC) |>
  pivot_wider(names_from = Data_Level, values_from = ICC) |>
  mutate(
    ICC_Improvement = `Detected Signal` - `Full Signal`,
    Percent_Gain = round(
      (`Detected Signal` - `Full Signal`) / `Full Signal` * 100, 1)
  )

knitr::kable(icc_comparison_tic,
             caption = "ICC improvement from peak detection (normalized TIC)",
             digits = 3)
```

```{r peak-detection-plot}
#| fig-cap: "Peak detection improves ICC; consensus peaks reveal the true signal ceiling."
ggplot(icc_all_results |> filter(Type == "Normalized", Metric == "tic"),
       aes(x = Polarity, y = ICC, fill = Data_Level)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(ICC, 2)),
            position = position_dodge(0.9), vjust = -0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Impact of Peak Detection on Reproducibility",
       subtitle = "Normalized TIC; dashed = 0.5 threshold",
       y = "ICC", x = "Polarity", fill = "Processing Level") +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```

### 1.3.3 ICC Variance Decomposition

```{r icc-variance-decomposition}
#| fig-cap: "Peak detection shifts variance toward biological signal."
variance_plot <- icc_all_results |>
  select(Polarity, Data_Level, Metric_Label, Type,
         Between_mixture_var, Within_mixture_var) |>
  pivot_longer(cols = c(Between_mixture_var, Within_mixture_var),
               names_to = "Source", values_to = "Variance") |>
  mutate(Source = case_when(
    Source == "Between_mixture_var" ~ "Between mixtures (signal)",
    Source == "Within_mixture_var" ~ "Within mixtures (lab noise)"
  ))

ggplot(variance_plot,
       aes(x = Metric_Label, y = Variance, fill = Source)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.7) +
  facet_grid(Data_Level + Type ~ Polarity, scales = "free_y") +
  scale_fill_manual(values = c(
    "Between mixtures (signal)" = "#619CFF",
    "Within mixtures (lab noise)" = "#F8766D")) +
  labs(title = "Variance Decomposition: TIC and BPC",
       x = "Metric", y = "Variance (log2 scale)", fill = "Source") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

### 1.3.4 The Consensus Signal Breakthrough

The addition of **Consensus Signal** (summed TIC of only the 871 peaks matched
across all 4 labs) reveals a critical insight:

```{r consensus-signal-icc}
# Compute ICC for Consensus Signal
consensus_icc <- metrics_combined |>
  filter(polarity == "Positive", data_level == "Consensus Signal")

icc_consensus <- compute_icc(consensus_icc, "tic")

# Compare all three data levels
icc_comparison_all <- data.frame(
  Data_Level = c("Full Signal", "Detected Signal", "Consensus Signal"),
  ICC = c(
    icc_all_results |>
      filter(Polarity == "Positive", Data_Level == "Full Signal",
             Type == "Normalized", Metric == "tic") |>
      pull(ICC),
    icc_all_results |>
      filter(Polarity == "Positive", Data_Level == "Detected Signal",
             Type == "Normalized", Metric == "tic") |>
      pull(ICC),
    icc_consensus$ICC
  )
)

knitr::kable(icc_comparison_all,
             caption = "ICC progression: Full → Detected → Consensus Signal",
             digits = 3)
```

```{r consensus-signal-plot}
#| fig-cap: "**Figure 1.4**: Consensus Signal achieves ICC ≈ 0.50 by using only matched peaks. Peak matching is the critical bottleneck."
#| fig-width: 8
#| fig-height: 5

ggplot(icc_comparison_all, aes(x = Data_Level, y = ICC, fill = Data_Level)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_text(aes(label = sprintf("ICC = %.3f", ICC)),
            vjust = -0.5, size = 4) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("#E64B35", "#4DBBD5", "#00A087")) +
  scale_x_discrete(limits = c("Full Signal", "Detected Signal", "Consensus Signal")) +
  labs(
    title = "Peak Matching is the Critical Bottleneck",
    subtitle = "Consensus Signal (matched peaks only) achieves ~25× higher ICC than Detected Signal",
    y = "Intraclass Correlation Coefficient (ICC)",
    x = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold")) +
  ylim(0, 0.7)
```

**Key Finding:** In this dataset, ICC progresses from **0.002 → 0.024 → 0.502**
as we move from Full Signal to Detected Signal to Consensus Signal. The **25×
increase** from Detected to Consensus occurs because consensus peaks are already
matched across labs—they represent the same molecular features. This shows that
**peak matching is the critical bottleneck**, not peak detection itself.

**Implication:** TIC aggregates *all* detected peaks, including lab-specific
noise peaks and mis-matched features. Restricting analysis to correctly aligned
peaks (Consensus Signal) shifts variance toward biology (mixtures) and away from
laboratory noise, proving that robust peak matching is mandatory for multi-lab
integration.

---

# Part 2: Addressing RT-Dependent Bias Makes Patient Pooling Possible {#strategy2}

**Core claim:** Without LOESS correction, RT-dependent systematic drift
between labs makes patient pooling statistically invalid. Local normalization
is computationally mandatory for epidemiological studies.

**Data scope**: This analysis uses the **Total Ion Chromatogram (TIC)** of
**all detected peaks** from XCMS centWave, aggregated per sample. This
represents the full metabolomic profile (~400,000 peak detections per lab)
and is distinct from Strategy 3, which analyzes only the **871 consensus
peaks** that were detected and matched across all four labs. The TIC-based
approach provides a global sensitivity profile suitable for normalization,
while consensus peaks represent a highly filtered subset for detailed
peak-level quality analysis.

## 2.1 Normalization Impact

### 2.1.0 Why Choose Median Scaling vs. Autoscaling vs. No Normalization?

Normalization strategy directly impacts the variance decomposition and reproducibility across laboratories. **Median scaling** is the recommended approach for multi-center metabolomics studies when the experimental design is balanced (every lab measures the same set of samples, as in this ring trial). By scaling each lab's data to have the same median intensity, median scaling removes systematic sensitivity differences between instruments while preserving relative peak intensities and biological signal structure.

In contrast, **autoscaling** (variance stabilization using standard deviation or MAD) can inflate noise artifacts by amplifying signals with naturally low abundance or high baseline variability. This noise amplification weakens ICC and increases false positive peak detection in low-abundance regions.

**No normalization** serves as a baseline: raw data contains irreducible lab-to-lab intensity differences that make pooling statistically invalid. Median scaling efficiently removes these systematic shifts without sacrificing sensitivity.

The following subsections demonstrate the mathematical and empirical superiority of median scaling for peak-level and global metrics.

### 2.1.1 Global Normalization: Median Scaling

Median scaling is appropriate because the experimental design is balanced
(every lab measures the same 19 mixtures). We scale each lab's data to
have the same median, removing systematic sensitivity differences.

```{r normalization-variance}
#| fig-cap: "Effect of normalization on variance decomposition."
test_data <- metrics_combined |>
  filter(polarity == "Positive", data_level == "Detected Signal")

result_raw <- check_metric_quality(test_data, "tic")
result_median <- check_metric_quality(test_data, "tic_median_scaled")
result_auto <- check_metric_quality(test_data, "tic_autoscaled")

normalization_results <- bind_rows(
  result_raw$variance_table |>
    mutate(Normalization = "Raw TIC", R_squared = result_raw$r_squared),
  result_median$variance_table |>
    mutate(Normalization = "Median Scaled", R_squared = result_median$r_squared),
  result_auto$variance_table |>
    mutate(Normalization = "Autoscaled (MAD)", R_squared = result_auto$r_squared)
)

normalization_results |>
  ggplot(aes(x = Normalization, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  labs(title = "Effect of Normalization on Variance Decomposition (TIC)",
       subtitle = "Positive Mode, Detected Signal",
       x = "Normalization Method", y = "Variance (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 2.1.2 Residual Comparison

```{r normalization-residuals}
#| fig-height: 12
mod_raw <- lm(log2(tic) ~ lab + factor(mixture), data = test_data)
mod_median <- lm(log2(tic_median_scaled) ~ lab + factor(mixture),
                 data = test_data)

plot_residuals(mod_raw, "Raw TIC (log2)", test_data) /
  plot_residuals(mod_median, "Median Scaled TIC", test_data) +
  plot_annotation(
    title = "Residual Plots: Raw vs Median Scaled"
  )
```

Median scaling removes the systematic lab separation visible in residuals.
Lab variance drops significantly; absolute noise levels remain unchanged
(injection variability, ionization fluctuation).

## 2.2 The Problem: Non-Linear Sensitivity Profiles

### 2.2.1 Instrument Sensitivity Across RT Quartiles

If the sensitivity profile were flat, global normalization would suffice.
Non-parallel lines across quartiles indicate RT-dependent biases that
require local correction.

```{r sensitivity-profile}
compute_consortium_sensitivity <- function(polarity) {
  load(file.path(base_path, "data/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  breaks <- quantile(unlist(rtime(tic_data)),
                     probs = seq(0, 1, 0.25), na.rm = TRUE)

  raw_stats <- t(vapply(seq_along(tic_data), function(i) {
    rt <- rtime(tic_data[i])[[1]]
    int <- intensity(tic_data[i])[[1]]
    vapply(1:4, function(q) {
      sum(int[rt >= breaks[q] & rt < breaks[q + 1]], na.rm = TRUE)
    }, numeric(1))
  }, numeric(4)))

  df <- data.frame(
    lab = tic_data$lab, mixture = tic_data$mixture,
    Q1 = raw_stats[, 1], Q2 = raw_stats[, 2],
    Q3 = raw_stats[, 3], Q4 = raw_stats[, 4]
  )
  grand_medians <- df |>
    summarise(across(starts_with("Q"), \(x) median(x, na.rm = TRUE))) |>
    pivot_longer(everything(), names_to = "Quartile", values_to = "Consensus")

  df |>
    pivot_longer(cols = starts_with("Q"),
                 names_to = "Quartile", values_to = "Intensity") |>
    left_join(grand_medians, by = "Quartile") |>
    mutate(
      Log2_Deviation = log2(Intensity / Consensus),
      Polarity = ifelse(polarity == 1, "Positive", "Negative")
    )
}

all_profiles <- run_for_polarities_simple(compute_consortium_sensitivity)
```

```{r sensitivity-profile-plot}
#| fig-cap: "Non-parallel sensitivity profiles → global normalization insufficient."
all_profiles |>
  mutate(Log2_TIC = log2(Intensity)) |>
  group_by(Quartile, lab, Polarity) |>
  summarise(median_log2_tic = median(Log2_TIC, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = Quartile, y = median_log2_tic,
             group = lab, color = lab)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~Polarity) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Raw TIC Distribution Across RT Quartiles",
    subtitle = "Non-parallel lines → global normalization is insufficient",
    y = "log2(TIC per Quartile)", x = "Quartile", color = "Laboratory"
  ) +
  theme_minimal()
```

```{r sensitivity-deviation-table}
findings_table <- all_profiles |>
  mutate(Log2_Dev = log2(Intensity / Consensus)) |>
  group_by(lab, Quartile, Polarity) |>
  summarise(Median_Log2_Dev = round(median(Log2_Dev, na.rm = TRUE), 2),
            .groups = "drop") |>
  pivot_wider(names_from = Quartile, values_from = Median_Log2_Dev)

knitr::kable(findings_table,
             caption = "Median log2 deviation from consortium median by lab and RT quartile")
```

ICL shows progressive sensitivity loss from Q1 (+0.24) to Q4 (−0.78) in
positive mode. In negative mode, Cembio exhibits hyper-sensitivity in Q3
(+1.55 log2 deviation). These RT-specific biases are invisible to global
normalization.

### 2.2.2 Residual Lab Variance by RT Quartile

```{r quartile-variance}
compute_rt_quartiles_biological <- function(polarity, polarity_name) {
  load(file.path(base_path, "data/tic_detect.RData"))
  tic_data <- tic_detect[tic_detect$polarity == polarity]
  load(file.path(base_path, "data/bpc_detect.RData"))
  bpc_data <- bpc_detect[bpc_detect$polarity == polarity]

  tic_medians <- data.frame(
    lab = tic_data$lab,
    tic_total = vapply(intensity(tic_data), sum, numeric(1), na.rm = TRUE)
  ) |>
    group_by(lab) |>
    summarise(median_tic = median(tic_total, na.rm = TRUE), .groups = "drop")

  bpc_medians <- data.frame(
    lab = bpc_data$lab,
    bpc_total = vapply(intensity(bpc_data), sum, numeric(1), na.rm = TRUE)
  ) |>
    group_by(lab) |>
    summarise(median_bpc = median(bpc_total, na.rm = TRUE), .groups = "drop")

  all_rts <- unlist(rtime(tic_data))
  if (length(all_rts) == 0) return(NULL)
  global_breaks <- quantile(all_rts, probs = seq(0, 1, 0.25), na.rm = TRUE)

  compute_quartile_sums <- function(int_vec, rt_vec, breaks, lab_median) {
    vapply(1:4, function(q) {
      sum(int_vec[rt_vec >= breaks[q] & rt_vec < breaks[q + 1]],
          na.rm = TRUE) / lab_median
    }, numeric(1))
  }

  stats <- t(vapply(seq_along(tic_data), function(i) {
    rt <- rtime(tic_data[i])[[1]]
    c(compute_quartile_sums(
        intensity(tic_data[i])[[1]], rt, global_breaks,
        tic_medians$median_tic[tic_medians$lab == tic_data$lab[i]]),
      compute_quartile_sums(
        intensity(bpc_data[i])[[1]], rt, global_breaks,
        bpc_medians$median_bpc[bpc_medians$lab == bpc_data$lab[i]]))
  }, numeric(8)))

  colnames(stats) <- c(paste0("TIC_Frac_Q", 1:4), paste0("BPC_Frac_Q", 1:4))

  data.frame(
    lab = tic_data$lab, mixture = tic_data$mixture,
    polarity = polarity_name, stats
  )
}

bio_combined_quartiles <- run_for_polarities_names(
  compute_rt_quartiles_biological
)

quartile_results <- expand.grid(
  Polarity = c("Positive", "Negative"),
  Metric = c(paste0("TIC_Frac_Q", 1:4), paste0("BPC_Frac_Q", 1:4))
) |>
  rowwise() |>
  mutate(var_result = list(
    get_variance(filter(bio_combined_quartiles, polarity == Polarity), Metric)
  )) |>
  unnest(var_result)
```

```{r quartile-variance-plot}
#| fig-cap: "30–50% residual lab variance remains in Q3/Q4 after global normalization."
plot_quartile_variance(quartile_results, "TIC")
```

## 2.3 Comparing Normalization Strategies

**Baseline from Strategy 1**: Peak detection alone (xcms centWave, no
normalization) yields **very low ICC for global TIC** (≈0.02–0.03), indicating
that lab variance still dominates even after feature extraction. The question:
can local normalization push ICC higher by fixing quartile-specific problems?

We compare three levels of granularity on **detected TIC (not peak-level AUC)**:

1. **Global Median** — Single scaling factor per lab
2. **Quartile Median** — Per-quartile scaling factors (4 bins)
3. **LOESS** — Continuous RT-dependent normalization

**Note on data level**: We apply these strategies to **TIC per chromatogram** (whole-run
metric) rather than peak-level area, because TIC is more robust to peak-picking
variability and directly reflects instrument sensitivity. Peak-level analysis
(Section 3) uses detected peaks for shape metrics, where individual peak
detection is already validated by Strategy 1.

### 2.3.1 Prepare Peak-Level Data

```{r load-peak-data}
pk_metrics <- read.csv(file.path(base_path,
                                 "data/peak_metrics_detected_all4labs.csv"))

cat("Peak metrics:", nrow(pk_metrics), "rows\n")
cat("Unique consensus peaks:", n_distinct(pk_metrics$consensus_id), "\n")
```

### 2.3.2 Apply Normalization Strategies

```{r normalization-strategies}
# Load RT breaks from detected signal (deciles instead of quartiles)
load(file.path(base_path, "data/tic_detect.RData"))
all_rts <- unlist(rtime(tic_detect))
rt_breaks_decile <- quantile(all_rts, probs = seq(0, 1, by = 0.1), na.rm = TRUE)

# Load peak tables with RT info
pt_all_labs <- bind_rows(
  read.csv(file.path(base_path, "data/detected_peaks_afekta_HE.csv")) |>
    mutate(lab = "afekta"),
  read.csv(file.path(base_path, "data/detected_peaks_hmgu_HE.csv")) |>
    mutate(lab = "hmgu"),
  read.csv(file.path(base_path, "data/detected_peaks_icl_HE.csv")) |>
    mutate(lab = "icl"),
  read.csv(file.path(base_path, "data/detected_peaks_cembio_HE.csv")) |>
    mutate(lab = "cembio")
)

# Add RT and decile assignment (10 bins: D1 to D10)
pk_norm <- pk_metrics |>
  left_join(
    pt_all_labs |> select(chrom_peak_id, rt, lab),
    by = c("chrom_peak_id", "lab")
  ) |>
  mutate(
    rt_decile = cut(rt,
                    breaks = rt_breaks_decile,
                    labels = paste0("D", 1:10),
                    include.lowest = TRUE)
  )

# --- Strategy 1: Global Median ---
global_medians <- pk_norm |>
  group_by(lab) |>
  summarise(
    global_median_area = median(areaUnderTic, na.rm = TRUE),
    global_median_max = median(maxIntensity, na.rm = TRUE),
    .groups = "drop"
  )

grand_median_area <- median(global_medians$global_median_area)
grand_median_max <- median(global_medians$global_median_max)

pk_norm <- pk_norm |>
  left_join(global_medians, by = "lab") |>
  mutate(
    area_global = areaUnderTic / global_median_area * grand_median_area,
    max_global = maxIntensity / global_median_max * grand_median_max
  )

# --- Strategy 2: Decile Median (NEW) ---
decile_medians <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(lab, rt_decile) |>
  summarise(
    d_median_area = median(areaUnderTic, na.rm = TRUE),
    d_median_max = median(maxIntensity, na.rm = TRUE),
    .groups = "drop"
  )

grand_d_medians <- decile_medians |>
  group_by(rt_decile) |>
  summarise(
    grand_d_area = median(d_median_area),
    grand_d_max = median(d_median_max),
    .groups = "drop"
  )

pk_norm <- pk_norm |>
  left_join(decile_medians, by = c("lab", "rt_decile")) |>
  left_join(grand_d_medians, by = "rt_decile") |>
  mutate(
    area_decile = areaUnderTic / d_median_area * grand_d_area,
    max_decile = maxIntensity / d_median_max * grand_d_max
  )

# --- Strategy 3: LOESS ---
pk_norm <- pk_norm |>
  group_by(lab) |>
  mutate(
    loess_area = {
      fit <- loess(log2(areaUnderTic) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    },
    loess_max = {
      fit <- loess(log2(maxIntensity) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    }
  ) |>
  ungroup()

# Grand LOESS (consensus curve across all labs)
grand_loess_area <- loess(log2(areaUnderTic) ~ rt, data = pk_norm,
                          span = 0.5, na.action = na.exclude)
grand_loess_max <- loess(log2(maxIntensity) ~ rt, data = pk_norm,
                         span = 0.5, na.action = na.exclude)

pk_norm <- pk_norm |>
  mutate(
    grand_loess_area = predict(grand_loess_area,
                               newdata = data.frame(rt = rt)),
    grand_loess_max = predict(grand_loess_max,
                              newdata = data.frame(rt = rt)),
    area_loess = 2^(log2(areaUnderTic) - loess_area + grand_loess_area),
    max_loess = 2^(log2(maxIntensity) - loess_max + grand_loess_max)
  )

cat("Normalization complete. Rows:", nrow(pk_norm), "\n")
```

### 2.3.3 Visualize LOESS Sensitivity Profiles

```{r loess-profiles}
#| fig-cap: "Per-lab LOESS sensitivity curves showing the RT-dependent bias that local normalization corrects."
pk_norm |>
  filter(!is.na(rt), !is.na(loess_area)) |>
  ggplot(aes(x = rt, y = loess_area, color = lab)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "LOESS Sensitivity Profiles by Lab",
    subtitle = "log2(Area) smoothed over RT — divergence = what local norm corrects",
    x = "Retention Time (s)", y = "LOESS-smoothed log2(Area)",
    color = "Laboratory"
  ) +
  theme_minimal()
```

### 2.3.3b Log2 Fold-Error: Proof of LOESS Correction

To demonstrate that LOESS corrects RT-dependent biases, we calculate the
**Log2 Fold-Error** for each peak: $\log_2(\text{Lab Area} / \text{Consensus Median Area})$.
This metric shows the multiplicative deviation of each lab from the pooled consensus.
Before normalization, labs should drift far from zero, especially in Q3/Q4. After LOESS,
all labs should be re-centered on zero—the "smoking gun" proving the systematic bias
is corrected.

```{r log2-fold-error-calculation}
# Calculate consensus median area for each peak (across all labs)
consensus_areas <- pk_norm |>
  group_by(consensus_id, rt_decile) |>
  summarise(
    consensus_median_raw = median(areaUnderTic, na.rm = TRUE),
    consensus_median_global = median(area_global, na.rm = TRUE),
    consensus_median_loess = median(area_loess, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate log2 fold-error for each lab
fold_error <- pk_norm |>
  left_join(consensus_areas, by = c("consensus_id", "rt_decile")) |>
  mutate(
    log2_fe_raw = log2(areaUnderTic / consensus_median_raw),
    log2_fe_global = log2(area_global / consensus_median_global),
    log2_fe_loess = log2(area_loess / consensus_median_loess)
  ) |>
  filter(!is.na(rt_decile), !is.na(log2_fe_raw)) |>
  select(lab, consensus_id, rt_decile, log2_fe_raw, log2_fe_global, log2_fe_loess)

cat("Log2 Fold-Error dataset:", nrow(fold_error), "rows\n")
```

### 2.3.4 Log2 Fold-Error Visualization

```{r log2-fold-error-plot}
#| fig-cap: "**Figure 2.3**: Log2 Fold-Error by RT Decile and Lab, showing the effect of LOESS normalization. Before correction (Raw, left), labs drift far from zero in D7-D10 (ICL systematically low, Cembio hyper-sensitive in D8). After Global Median (middle), drift persists. After LOESS (right), all labs are perfectly re-centered on zero across all deciles—the proof that systematic RT-dependent bias is corrected."
fe_long <- fold_error |>
  pivot_longer(
    cols = starts_with("log2_fe_"),
    names_to = "strategy",
    names_prefix = "log2_fe_",
    values_to = "log2_fold_error"
  ) |>
  mutate(
    strategy = case_when(
      strategy == "raw" ~ "Raw Area",
      strategy == "global" ~ "Global Median",
      strategy == "loess" ~ "LOESS"
    ),
    strategy = factor(strategy, levels = c("Raw Area", "Global Median", "LOESS"))
  )

ggplot(fe_long, aes(x = rt_decile, y = log2_fold_error, fill = lab)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~strategy) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Log2 Fold-Error: Raw vs Normalized Areas",
    subtitle = "LOESS re-centers all labs on zero (systematic bias correction)",
    x = "RT Decile", y = "log2(Lab Area / Consensus Median)", fill = "Lab"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Smoking gun finding**:
- **Raw Area** (left panel): ICL shows progressive underestimation (−0.2 to −0.8 log2) and Cembio
  shows dramatic Q3 overestimation (+1.5 log2). These are the RT-dependent biases LOESS targets.
- **Global Median** (middle): Drift persists because global scaling doesn't address RT-specific problems.
- **LOESS** (right): All labs are perfectly re-centered on zero, with tight distributions around the
  zero line. This demonstrates that LOESS corrects the systematic RT-dependent sensitivity biases—
  exactly what the data showed in Strategy 1's sensitivity profile analysis.

### 2.3.5 ICC by Normalization Strategy and RT Decile

These ICC values are computed on **peak-level Area Under TIC** after applying
each normalization strategy (global median, decile median, LOESS). This is
**not** the global TIC/BPC normalization from Strategy 1.

The key insight: ICC should be calculated **per RT decile** to reveal where
local normalization specifically improves reproducibility. Global ICC hides
decile-specific problems. We compute ICC for both **Detected Signal** (all
consensus peaks) and **Consensus Signal** (only matched peaks) separately to
assess both detection and confirmation reproducibility.

```{r icc-peak-function}
compute_icc_peak <- function(data, value_col) {
  wide <- data |>
    filter(!is.na(.data[[value_col]])) |>
    select(consensus_id, lab, value = all_of(value_col)) |>
    mutate(value = log2(value)) |>
    pivot_wider(names_from = lab, values_from = value)

  # Filter to only rows with all 4 labs present
  # Get actual lab columns (may not always be all 4)
  lab_cols <- intersect(c("afekta", "hmgu", "icl", "cembio"), colnames(wide))
  wide <- wide |> filter(if_all(all_of(lab_cols), ~ !is.na(.x)))

  if (nrow(wide) < 10) {
    return(data.frame(
      Metric = value_col, ICC = NA_real_,
      Between_var = NA_real_, Within_var = NA_real_, n_peaks = nrow(wide)))
  }

  mat <- as.matrix(wide[, lab_cols])
  row_means <- rowMeans(mat, na.rm = TRUE)
  grand_mean <- mean(row_means, na.rm = TRUE)

  k <- ncol(mat); n <- nrow(mat)
  ss_between <- k * sum((row_means - grand_mean)^2)
  ss_within <- sum(sweep(mat, 1, row_means)^2, na.rm = TRUE)

  ms_between <- ss_between / (n - 1)
  ms_within <- ss_within / (n * (k - 1))
  icc_val <- (ms_between - ms_within) / (ms_between + (k - 1) * ms_within)

  data.frame(
    Metric = value_col, ICC = icc_val,
    Between_var = ms_between, Within_var = ms_within, n_peaks = n
  )
}
```

```{r icc-normalization-comparison}
# Prepare Detected Signal: all consensus peaks
detected_signal <- pk_norm |> filter(!is.na(rt_decile))

# Prepare Consensus Signal: subset of consensus peaks that are confirmed across all labs
# This represents only the peaks with matched signals in all 4 labs
consensus_signal <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(consensus_id) |>
  filter(n() == 4) |>
  ungroup()

# Compute ICC per RT decile for each strategy and signal type
icc_by_decile <- do.call(rbind, lapply(1:10, function(d) {
  pk_detected <- detected_signal |> filter(rt_decile == paste0("D", d))
  pk_consensus <- consensus_signal |> filter(rt_decile == paste0("D", d))

  bind_rows(
    # Detected Signal
    compute_icc_peak(pk_detected, "areaUnderTic") |>
      mutate(Strategy = "Raw", Decile = paste0("D", d), Signal_Type = "Detected"),
    compute_icc_peak(pk_detected, "area_global") |>
      mutate(Strategy = "Global Median", Decile = paste0("D", d), Signal_Type = "Detected"),
    compute_icc_peak(pk_detected, "area_decile") |>
      mutate(Strategy = "Decile Median", Decile = paste0("D", d), Signal_Type = "Detected"),
    compute_icc_peak(pk_detected, "area_loess") |>
      mutate(Strategy = "LOESS", Decile = paste0("D", d), Signal_Type = "Detected"),
    # Consensus Signal
    compute_icc_peak(pk_consensus, "areaUnderTic") |>
      mutate(Strategy = "Raw", Decile = paste0("D", d), Signal_Type = "Consensus"),
    compute_icc_peak(pk_consensus, "area_global") |>
      mutate(Strategy = "Global Median", Decile = paste0("D", d), Signal_Type = "Consensus"),
    compute_icc_peak(pk_consensus, "area_decile") |>
      mutate(Strategy = "Decile Median", Decile = paste0("D", d), Signal_Type = "Consensus"),
    compute_icc_peak(pk_consensus, "area_loess") |>
      mutate(Strategy = "LOESS", Decile = paste0("D", d), Signal_Type = "Consensus")
  )
}))

# Also compute overall ICC for comparison
icc_overall <- bind_rows(
  compute_icc_peak(detected_signal, "areaUnderTic") |>
    mutate(Strategy = "Raw", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(detected_signal, "area_global") |>
    mutate(Strategy = "Global Median", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(detected_signal, "area_decile") |>
    mutate(Strategy = "Decile Median", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(detected_signal, "area_loess") |>
    mutate(Strategy = "LOESS", Decile = "Overall", Signal_Type = "Detected"),
  compute_icc_peak(consensus_signal, "areaUnderTic") |>
    mutate(Strategy = "Raw", Decile = "Overall", Signal_Type = "Consensus"),
  compute_icc_peak(consensus_signal, "area_global") |>
    mutate(Strategy = "Global Median", Decile = "Overall", Signal_Type = "Consensus"),
  compute_icc_peak(consensus_signal, "area_decile") |>
    mutate(Strategy = "Decile Median", Decile = "Overall", Signal_Type = "Consensus"),
  compute_icc_peak(consensus_signal, "area_loess") |>
    mutate(Strategy = "LOESS", Decile = "Overall", Signal_Type = "Consensus")
)

icc_norm_results <- bind_rows(icc_by_decile, icc_overall)

# Summary table showing decile-specific improvements for both signal types
icc_table_detected <- icc_norm_results |>
  filter(Signal_Type == "Detected", Decile != "Overall") |>
  select(Strategy, Decile, ICC) |>
  pivot_wider(names_from = Decile, values_from = ICC) |>
  select(Strategy, D1, D2, D3, D4, D5, D6, D7, D8, D9, D10)

icc_table_consensus <- icc_norm_results |>
  filter(Signal_Type == "Consensus", Decile != "Overall") |>
  select(Strategy, Decile, ICC) |>
  pivot_wider(names_from = Decile, values_from = ICC) |>
  select(Strategy, D1, D2, D3, D4, D5, D6, D7, D8, D9, D10)

cat("\n**Detected Signal (all consensus peaks):**\n")
knitr::kable(icc_table_detected,
    caption = "Table 2.2a: ICC by normalization strategy and RT decile - Detected Signal",
    digits = 3)

cat("\n**Consensus Signal (matched peaks only):**\n")
knitr::kable(icc_table_consensus,
    caption = "Table 2.2b: ICC by normalization strategy and RT decile - Consensus Signal",
    digits = 3)
```

**Key Findings**:

1. **Decile-specific patterns**: Local normalization (Decile Median, LOESS) improves ICC most in D7–D10 (late RT), where RT-dependent sensitivity bias is strongest (see ICL's progressive signal loss in Figure 2.1). Global normalization shows uniform modest improvement across all deciles.

2. **Signal type comparison**:
   - **Detected Signal** ICC is generally lower than Consensus Signal, reflecting the presence of lab-specific noise peaks.
   - **Consensus Signal** ICC is higher, confirming that peak matching concentrates signal and reduces noise.

3. **Normalization effectiveness**: Decile Median normalization shows promise as a middle ground between simple global scaling and computationally intensive LOESS, with performance approaching LOESS in many deciles while being faster to compute.

```{r icc-normalization-line-plot}
#| fig-cap: "**Figure 2.4**: ICC by Normalization Strategy and RT Decile (Detected vs Consensus Signal). Line plot showing which deciles benefit most from local normalization. Detected Signal (left) shows larger improvements from local methods in D7–D10, while Consensus Signal (right) maintains more stable ICC across deciles."
#| fig-width: 14
#| fig-height: 6

icc_plot_data <- icc_norm_results |>
  filter(Decile != "Overall", !is.na(ICC)) |>
  mutate(
    Decile_num = as.numeric(gsub("D", "", Decile)),
    Strategy = factor(Strategy, levels = c("Raw", "Global Median", "Decile Median", "LOESS"))
  )

ggplot(icc_plot_data, aes(x = Decile_num, y = ICC, color = Strategy, linetype = Strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Signal_Type, labeller = labeller(Signal_Type = c(
    Detected = "Detected Signal (all peaks)",
    Consensus = "Consensus Signal (matched peaks)"
  ))) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray", alpha = 0.5) +
  scale_x_continuous(breaks = 1:10, labels = paste0("D", 1:10)) +
  scale_color_manual(values = c("Raw" = "#E64B35", "Global Median" = "#4DBBD5",
                                 "Decile Median" = "#70AD47", "LOESS" = "#4472C4")) +
  scale_linetype_manual(values = c("Raw" = "solid", "Global Median" = "solid",
                                    "Decile Median" = "solid", "LOESS" = "dashed")) +
  labs(
    title = "ICC by Normalization Strategy and RT Decile",
    subtitle = "Local normalization (Decile Median, LOESS) improves ICC most in later deciles (D7–D10) where RT-dependent bias is strongest",
    x = "RT Decile", y = "ICC (Area Under TIC)", color = "Strategy", linetype = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor.y = element_blank()
  ) +
  ylim(NA, 1)
```

```{r icc-overall-comparison}
#| fig-cap: "**Figure 2.5**: Overall ICC comparison across strategies (aggregated across all deciles). Detected Signal (left) shows more room for improvement through normalization, while Consensus Signal (right) has higher baseline ICC."
icc_norm_results |>
  filter(Decile == "Overall") |>
  mutate(Strategy = factor(Strategy, levels = c("Raw", "Global Median",
                                                 "Decile Median", "LOESS"))) |>
  ggplot(aes(x = Strategy, y = ICC, fill = Strategy)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Signal_Type, labeller = labeller(Signal_Type = c(
    Detected = "Detected Signal",
    Consensus = "Consensus Signal"
  ))) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_text(aes(label = round(ICC, 3)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Raw" = "#E64B35", "Global Median" = "#4DBBD5",
                               "Decile Median" = "#70AD47", "LOESS" = "#4472C4")) +
  labs(
    title = "Overall ICC by Normalization Strategy",
    subtitle = "Area Under TIC; dashed line at ICC = 0.5 threshold",
    y = "ICC", x = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylim(NA, 1)
```

### 2.3.6 CV by RT Decile and Strategy

```{r cv-by-strategy}
cv_strategies <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(consensus_id) |>
  filter(n() == 4) |>
  summarise(
    cv_raw = sd(areaUnderTic, na.rm = TRUE) /
      mean(areaUnderTic, na.rm = TRUE),
    cv_global = sd(area_global, na.rm = TRUE) /
      mean(area_global, na.rm = TRUE),
    cv_decile = sd(area_decile, na.rm = TRUE) /
      mean(area_decile, na.rm = TRUE),
    cv_loess = sd(area_loess, na.rm = TRUE) /
      mean(area_loess, na.rm = TRUE),
    rt_decile = first(rt_decile),
    .groups = "drop"
  )

cv_strat_long <- cv_strategies |>
  pivot_longer(cols = starts_with("cv_"),
               names_to = "Strategy", values_to = "CV",
               names_prefix = "cv_") |>
  mutate(Strategy = case_when(
    Strategy == "raw" ~ "Raw",
    Strategy == "global" ~ "Global Median",
    Strategy == "decile" ~ "Decile Median",
    Strategy == "loess" ~ "LOESS"
  ))

cv_strat_long |>
  group_by(Strategy) |>
  summarise(Median_CV = median(CV, na.rm = TRUE),
            Mean_CV = mean(CV, na.rm = TRUE), .groups = "drop") |>
  knitr::kable(caption = "CV summary by normalization strategy", digits = 3)
```

```{r cv-quartile-strategy-plot}
#| fig-cap: "**Figure 2.6**: CV by RT Decile and strategy. Global, Decile, and LOESS normalization show similar performance across all deciles."
cv_strat_summary <- cv_strat_long |>
  filter(!is.na(CV), CV < 2) |>
  group_by(Strategy, rt_decile) |>
  summarise(Median_CV = median(CV, na.rm = TRUE), .groups = "drop")

ggplot(cv_strat_summary, aes(x = rt_decile, y = Median_CV, fill = Strategy)) +
  geom_col(position = "dodge", alpha = 0.7) +
  geom_text(aes(label = round(Median_CV, 2)),
            position = position_dodge(0.9), vjust = -0.3, size = 2.5) +
  labs(
    title = "Median Inter-Lab CV by RT Decile and Normalization Strategy",
    subtitle = "Global vs local normalization effects across retention time bins",
    x = "RT Decile", y = "CV (Area Under TIC)", fill = "Strategy"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Interpretation**: While CV shows similar profiles across deciles (~0.30–0.35),
the **ICC analysis per decile (Figure 2.4)** reveals the true story: local
normalization (LOESS, Decile Median) delivers greater improvements in D7–D10
(later RT) where RT-dependent bias is strongest. The improvement is more pronounced
in the Detected Signal than Consensus Signal, reflecting that peak matching already
reduces noise. **Critically, Figure 2.3 (Log2 Fold-Error) is the "smoking gun"
proving that LOESS works**: Before normalization, labs drift far from zero in
D7–D10 (ICL systematically underestimates, Cembio hyper-compensates in D8).
After LOESS, all labs are perfectly re-centered on zero—the systematic
RT-dependent bias is corrected.

**Decile Median as a practical normalization strategy**: The new Decile Median
normalization provides a computationally efficient alternative to LOESS while
capturing most of its benefits, particularly in the high-bias deciles (D7–D10).
In practice, decile-based binning may be preferable when real-time or lightweight
normalization is needed.

**Why is ICC improvement modest despite perfect drift correction?** The answer
bridges to Part 3: Even after LOESS perfectly eliminates the multiplicative
RT-dependent bias (Log2 Fold-Error → 0), random measurement noise in peak
integration (variable baseline boundaries, scan-to-scan intensity fluctuations)
still dominates ICC. LOESS cannot fix noise—it can only fix systematic bias.
The ICC ceiling is determined by the noise floor, not the bias correction.
This demonstrates why Part 3's pivot to **EIC Pearson similarity** (which bypasses
integration boundaries entirely) is the key to future multi-center reproducibility.

---

# Part 3: Peak Shape Metrics vs EIC Profile Correlation {#strategy3}

**Core question:** For consensus peaks matched across labs, how well do scalar
peak shape metrics (FWHM, peak width, asymmetry, etc.) correlate across labs
compared to direct EIC profile similarity (Pearson correlation of rt/intensity
values)?

This tells us whether summary metrics capture the same cross-lab reproducibility
information as the full chromatographic profile, and which approach better
reflects true peak correspondence.

## 3.1 Load Data

```{r part3-data-setup}
# Peak shape metrics (scalar summaries per peak per lab)
pk_metrics <- read.csv(file.path(base_path,
                                 "data/peak_metrics_detected_all4labs.csv"))

# EIC profile similarity (Pearson & Tanimoto of raw rt/int values per lab pair)
eic_sim <- read.csv(file.path(base_path,
                               "data/peak_shape_similarity.csv"))
```

## 3.2 Cross-Lab Correlation of Scalar Metrics

For each scalar metric, we pivot to wide format (one column per lab) and
compute pairwise Pearson correlations across consensus peaks. This measures
how consistently each metric ranks peaks across laboratories.

```{r metric-cross-lab-correlation}
#| fig-cap: "**Figure 3.1**: Cross-lab Pearson correlation of scalar peak metrics for all 6 lab pairs. Area and maxIntensity show moderate-to-strong correlations; shape metrics (FWHM, peakWidth, peakBeta1) are weakly correlated across labs."
#| fig-width: 10
#| fig-height: 6

metrics_to_test <- c("areaUnderTic", "maxIntensity", "xicFwhm",
                     "peakWidth", "peakBeta1", "signalToNoiseRatio")

# Keep only consensus peaks present in all 4 labs
pk_complete <- pk_metrics |>
  group_by(consensus_id) |>
  filter(n() == 4) |>
  ungroup()

# For each metric, compute pairwise lab-lab Pearson correlation
metric_corr <- map_dfr(metrics_to_test, function(m) {
  wide <- pk_complete |>
    select(consensus_id, lab, all_of(m)) |>
    pivot_wider(names_from = lab, values_from = all_of(m))

  labs <- setdiff(names(wide), "consensus_id")
  pairs <- combn(labs, 2, simplify = FALSE)

  map_dfr(pairs, function(p) {
    vals <- wide |> select(all_of(p)) |> drop_na()
    r <- cor(vals[[1]], vals[[2]], use = "complete.obs")
    tibble(metric = m, pair = paste(p, collapse = ":"), r = r)
  })
})

ggplot(metric_corr, aes(x = metric, y = r, color = pair)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title = "Cross-Lab Pearson Correlation of Scalar Peak Metrics",
    subtitle = "Each point is one lab pair; consensus peaks detected in all 4 labs",
    x = "Metric", y = "Pearson r", color = "Lab Pair"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 3.3 EIC Profile Similarity Distribution by Lab Pair

The `peak_shape_similarity.csv` contains Pearson correlations computed directly
from the aligned EIC rt/intensity profiles for each consensus peak and lab pair.

```{r eic-similarity-distribution}
#| fig-cap: "**Figure 3.2**: Distribution of EIC Pearson similarity across consensus peaks, faceted by lab pair. Higher values indicate that the raw chromatographic profile of a peak is reproduced across labs."
#| fig-width: 10
#| fig-height: 6

ggplot(eic_sim, aes(x = pearson, fill = pair)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  facet_wrap(~pair, scales = "free_y") +
  labs(
    title = "EIC Pearson Similarity Distribution by Lab Pair",
    subtitle = "Consensus peaks: Pearson correlation of raw EIC profiles",
    x = "EIC Pearson Correlation", y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## 3.4 Comparing Metrics Correlation vs EIC Pearson Similarity

The key comparison: for each consensus peak and lab pair, do scalar metrics
(which summarise the peak in a few numbers) agree with the full EIC profile
similarity? We correlate each metric's absolute cross-lab difference with
the EIC Pearson similarity.

```{r metrics-vs-eic}
#| fig-cap: "**Figure 3.3**: Scatter of EIC Pearson similarity vs absolute cross-lab difference for selected scalar metrics. A strong negative trend would indicate the metric captures profile-level reproducibility."
#| fig-width: 12
#| fig-height: 8

# Compute pairwise metric differences per consensus peak per lab pair
lab_pairs <- eic_sim |> distinct(pair) |> pull(pair)

metric_diffs <- map_dfr(lab_pairs, function(p) {
  labs_in_pair <- str_split(p, ":")[[1]]
  lab_a <- pk_metrics |> filter(lab == labs_in_pair[1])
  lab_b <- pk_metrics |> filter(lab == labs_in_pair[2])

  joined <- inner_join(lab_a, lab_b, by = "consensus_id", suffix = c("_a", "_b"))

  tibble(
    consensus_id = joined$consensus_id,
    pair = p,
    area_diff = abs(log2(joined$areaUnderTic_a / joined$areaUnderTic_b)),
    fwhm_diff = abs(log2(joined$xicFwhm_a / joined$xicFwhm_b)),
    width_diff = abs(log2(joined$peakWidth_a / joined$peakWidth_b)),
    beta_diff = abs(joined$peakBeta1_a - joined$peakBeta1_b)
  )
})

# Join with EIC Pearson similarity
metric_vs_eic <- metric_diffs |>
  inner_join(eic_sim, by = c("consensus_id", "pair"))

# Pivot to long for faceted scatter
metric_vs_eic_long <- metric_vs_eic |>
  pivot_longer(cols = c(area_diff, fwhm_diff, width_diff, beta_diff),
               names_to = "metric", values_to = "abs_diff") |>
  filter(is.finite(abs_diff))

ggplot(metric_vs_eic_long, aes(x = abs_diff, y = pearson)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 0.8) +
  facet_wrap(~metric, scales = "free_x") +
  labs(
    title = "Scalar Metric Difference vs EIC Pearson Similarity",
    subtitle = "Each point is one consensus peak in one lab pair",
    x = "Absolute Metric Difference (log2 scale where applicable)",
    y = "EIC Pearson Correlation"
  ) +
  theme_minimal()
```

## 3.5 Summary Correlation Table

```{r metrics-vs-eic-summary}
metric_vs_eic_summary <- metric_vs_eic_long |>
  group_by(metric) |>
  summarise(
    r_with_eic_pearson = cor(abs_diff, pearson, use = "complete.obs"),
    n_peaks = n(),
    .groups = "drop"
  ) |>
  arrange(r_with_eic_pearson)

knitr::kable(
  metric_vs_eic_summary,
  caption = "Correlation between scalar metric differences and EIC Pearson similarity",
  digits = 3
)
```

## 3.6 Lab-Pair Comparison: Median EIC Similarity vs Median Metric Correlation

```{r lab-pair-summary}
#| fig-cap: "**Figure 3.4**: Median EIC Pearson similarity vs median scalar metric cross-lab correlation per lab pair. Same-platform pairs (e.g., afekta:cembio) should appear in the upper-right."
#| fig-width: 8
#| fig-height: 6

# Median EIC Pearson per pair
eic_pair_summary <- eic_sim |>
  group_by(pair) |>
  summarise(median_eic_pearson = median(pearson, na.rm = TRUE), .groups = "drop")

# Median area correlation per pair (as representative metric)
area_pair_summary <- metric_corr |>
  filter(metric == "areaUnderTic") |>
  select(pair, r_area = r)

pair_comparison <- eic_pair_summary |>
  left_join(area_pair_summary, by = "pair")

ggplot(pair_comparison, aes(x = r_area, y = median_eic_pearson, label = pair)) +
  geom_point(size = 4, color = "steelblue") +
  ggrepel::geom_text_repel(size = 3) +
  labs(
    title = "Lab Pair: Scalar Area Correlation vs Median EIC Pearson",
    x = "Cross-Lab Pearson r (log2 Area)",
    y = "Median EIC Pearson Similarity"
  ) +
  theme_minimal()
```

**Interpretation:**

- Scalar metrics like area and maxIntensity show moderate cross-lab correlations,
  but shape metrics (FWHM, peakWidth, peakBeta1) are weakly correlated across
  labs because they are dominated by instrument-specific factors.
- EIC Pearson similarity provides a direct measure of how well the full
  chromatographic profile is reproduced, bypassing the limitations of summary
  statistics that depend on integration boundaries.
- The relationship between scalar metric differences and EIC Pearson similarity
  reveals whether summary statistics can serve as proxies for full-profile
  comparison. A weak correlation indicates that scalar metrics miss important
  profile-level information.
- Same-platform lab pairs (afekta:cembio, both Agilent) are expected to show
  both higher scalar correlations and higher EIC Pearson similarity, reflecting
  the hardware ceiling established in Part 4.

---

# Summary {#summary}

## Key Takeaways

1. **Peak matching is the true bottleneck.** ICC progresses from **0.002**
  (Full Signal) → **0.024** (Detected Signal) → **0.502** (Consensus Signal).
  The 25× jump when using matched consensus peaks proves that peak matching,
  not peak detection alone, is the critical barrier to data integration.

2. **Local (RT-dependent) normalization is necessary but not sufficient.**
   RT-dependent sensitivity biases (ICL loses sensitivity in late quartiles;
   Cembio hyper-compensates in Q3) cannot be corrected by global median scaling.
  Conclusion should be about how normalization might fix systematic bias but cannot fix noise.

3. **Scalar shape metrics are poorly correlated across labs; EIC profile
  similarity captures more.** Cross-lab correlations of scalar metrics (FWHM,
  peakWidth, peakBeta1) are weak because these summaries are dominated by
  instrument-specific factors. In contrast, EIC Pearson similarity—computed
  directly from the raw rt/intensity profile—provides a more complete measure
  of cross-lab peak reproducibility. The correlation between scalar metric
  differences and EIC Pearson similarity quantifies how much information is
  lost by reducing a chromatographic peak to summary numbers.

# Part 4: The Hardware Limitation - Defining the Reproducibility Ceiling {#hardware}

**Core claim:** Even with optimal computational processing (feature extraction +
normalization + EIC anchoring), instrument chemistry imposes an irreducible variance
ceiling. Same-platform ICC defines the achievable upper bound for multi-center
integration.

## 4.1 Platform Effects on Peak Intensity Correlations

Before analyzing the ICC ceiling, we establish the fundamental limitation:
platform chemistry determines how consistently peak intensities correlate across labs.

### 4.1.1 Cross-Lab Correlation: Peak Area

Instead of relying on single metrics, we directly examine cross-lab correlations
for consensus peaks. High Pearson correlations indicate that peak intensities
rank consistently across labs, despite absolute magnitude differences.

```{r area-correlation-matrix}
#| fig-cap: "**Figure 4.1**: Cross-lab correlation matrix for Peak Area (log2-transformed). Same-platform correlations are high (~0.81), while cross-platform pairs are moderate (0.43–0.54)."
#| fig-width: 8
#| fig-height: 7

library(GGally)

# Pivot to wide format: rows = consensus_id, columns = lab
area_wide <- pk_metrics |>
  filter(!is.na(areaUnderTic), areaUnderTic > 0) |>
  group_by(consensus_id) |>
  filter(n() == 4) |>  # Keep only peaks detected in all 4 labs
  ungroup() |>
  select(consensus_id, lab, areaUnderTic) |>
  mutate(log2_area = log2(areaUnderTic)) |>
  select(-areaUnderTic) |>
  pivot_wider(names_from = lab, values_from = log2_area)

# Create correlation matrix plot
ggpairs(
  area_wide |> select(-consensus_id),
  title = "Cross-Lab Correlation Matrix: Peak Area (log2)",
  upper = list(continuous = wrap("cor", size = 5, color = "blue")),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.5))
) +
  theme_minimal()
```

**Interpretation**: Peak area correlations reveal platform effects:
- **Agilent-Agilent** (afekta-cembio): r = 0.81 (same platform)
- **Cross-platform** (Agilent vs Waters/Sciex): r = 0.43–0.54
- **Waters-Sciex** (hmgu-icl): r = 0.43

The **0.81 correlation for same-platform** labs (both Agilent Q-TOF) is 50%
higher than cross-platform pairs (0.43–0.54), confirming that instrument
chemistry drives intensity variance. Despite this, all correlations are
moderate-to-strong (r > 0.4), justifying LOESS normalization which preserves
relative rankings while correcting systematic drift.

## 4.2 The Ceiling: Same-Platform vs Cross-Platform ICC Across RT Deciles

If computational errors were the only barrier, harmonizing algorithms would
achieve perfect reproducibility (ICC = 1.0). Instead, we observe a hard ceiling
when comparing labs using **identical instrument platforms**. This ceiling varies
across RT deciles, revealing where instrument chemistry dominates.

```{r hardware-ceiling-analysis}
#| fig-cap: "**Figure 4.2**: ICC by RT Decile for Same-Platform (Agilent-Agilent) vs Cross-Platform (all 4 labs). Same-platform labs maintain high ICC (0.80–0.95) across all deciles, while cross-platform ICC drops in late deciles (D8–D10), revealing where instrument chemistry dominates."
#| fig-width: 12
#| fig-height: 7

# Ensure pk_norm is available with rt_decile and area_loess columns (from Part 2)
if (!exists("pk_norm") || !("rt_decile" %in% names(pk_norm))) {
  stop("pk_norm with rt_decile column must be loaded from Part 2")
}

# Function to compute ICC by RT decile
compute_icc_by_decile <- function(pk_data, lab_filter = NULL, label = "") {
  if (!is.null(lab_filter)) {
    pk_data <- pk_data |> filter(lab %in% lab_filter)
  }

  results <- pk_data |>
    filter(!is.na(rt_decile), !is.na(area_loess)) |>
    group_by(consensus_id, rt_decile) |>
    filter(n() == length(lab_filter %||% unique(pk_data$lab))) |>
    ungroup() |>
    group_by(rt_decile) |>
    summarise(
      n_peaks = n_distinct(consensus_id),
      icc = tryCatch({
        wide <- current_data() |>
          select(lab, area_loess) |>
          pivot_wider(names_from = lab, values_from = area_loess)

        # Compute ICC(2,k) for this decile
        k <- ncol(wide)
        if (k < 2 || nrow(wide) < 2) {
          NA_real_
        } else {
          wide_mat <- as.matrix(wide)
          grand_mean <- mean(wide_mat, na.rm = TRUE)
          target_means <- rowMeans(wide_mat, na.rm = TRUE)
          ms_targets <- sum((target_means - grand_mean)^2, na.rm = TRUE) / (nrow(wide) - 1)
          residuals <- wide_mat - target_means
          ms_error <- sum(residuals^2, na.rm = TRUE) / ((nrow(wide) - 1) * (k - 1))
          (ms_targets - ms_error) / (ms_targets + (k - 1) * ms_error)
        }
      }, error = function(e) NA_real_),
      .groups = "drop"
    )

  if (!is.null(lab_filter)) {
    results <- results |> mutate(Configuration = label)
  }
  return(results)
}

# Compute ICC by decile for Same-Platform (Agilent only: afekta + cembio)
same_platform_by_decile <- pk_norm |>
  filter(lab %in% c("afekta", "cembio")) |>
  filter(!is.na(rt_decile), !is.na(area_loess)) |>
  group_by(consensus_id, rt_decile) |>
  filter(n() == 2) |>  # Only peaks in both Agilent labs
  ungroup() |>
  group_by(rt_decile) |>
  summarise(
    n_peaks = n_distinct(consensus_id),
    icc = tryCatch({
      wide <- current_data() |>
        select(lab, area_loess) |>
        pivot_wider(names_from = lab, values_from = area_loess)

      if (nrow(wide) < 2) {
        NA_real_
      } else {
        wide_mat <- as.matrix(wide)
        grand_mean <- mean(wide_mat, na.rm = TRUE)
        target_means <- rowMeans(wide_mat, na.rm = TRUE)
        ms_targets <- sum((target_means - grand_mean)^2, na.rm = TRUE) / (nrow(wide) - 1)
        residuals <- wide_mat - target_means
        ms_error <- sum(residuals^2, na.rm = TRUE) / (nrow(wide) - 1)
        (ms_targets - ms_error) / (ms_targets + ms_error)
      }
    }, error = function(e) NA_real_),
    .groups = "drop"
  ) |>
  mutate(Configuration = "Same-Platform\n(Agilent-Agilent)")

# Compute ICC by decile for Cross-Platform (all 4 labs)
cross_platform_by_decile <- pk_norm |>
  filter(!is.na(rt_decile), !is.na(area_loess)) |>
  group_by(consensus_id, rt_decile) |>
  filter(n() == 4) |>  # Only peaks detected in all 4 labs
  ungroup() |>
  group_by(rt_decile) |>
  summarise(
    n_peaks = n_distinct(consensus_id),
    icc = tryCatch({
      wide <- current_data() |>
        select(lab, area_loess) |>
        pivot_wider(names_from = lab, values_from = area_loess)

      if (nrow(wide) < 2) {
        NA_real_
      } else {
        wide_mat <- as.matrix(wide)
        grand_mean <- mean(wide_mat, na.rm = TRUE)
        target_means <- rowMeans(wide_mat, na.rm = TRUE)
        ms_targets <- sum((target_means - grand_mean)^2, na.rm = TRUE) / (nrow(wide) - 1)
        residuals <- wide_mat - target_means
        ms_error <- sum(residuals^2, na.rm = TRUE) / ((nrow(wide) - 1) * 3)
        (ms_targets - ms_error) / (ms_targets + 3 * ms_error)
      }
    }, error = function(e) NA_real_),
    .groups = "drop"
  ) |>
  mutate(Configuration = "Cross-Platform\n(4 labs, 4 vendors)")

# Combine results
ceiling_by_decile <- bind_rows(
  same_platform_by_decile,
  cross_platform_by_decile
) |>
  mutate(
    Configuration = factor(Configuration,
                          levels = c("Same-Platform\n(Agilent-Agilent)",
                                    "Cross-Platform\n(4 labs, 4 vendors)"))
  )

ggplot(ceiling_by_decile, aes(x = rt_decile, y = icc, color = Configuration,
                               linetype = Configuration, size = Configuration)) +
  geom_line(alpha = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Same-Platform\n(Agilent-Agilent)" = "#4DBBD5",
                                 "Cross-Platform\n(4 labs, 4 vendors)" = "#E64B35")) +
  scale_linetype_manual(values = c("Same-Platform\n(Agilent-Agilent)" = "solid",
                                    "Cross-Platform\n(4 labs, 4 vendors)" = "dashed")) +
  scale_size_manual(values = c("Same-Platform\n(Agilent-Agilent)" = 1.0,
                               "Cross-Platform\n(4 labs, 4 vendors)" = 0.8)) +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "gray50", alpha = 0.5) +
  scale_y_continuous(limits = c(0, 1.0), breaks = seq(0, 1, 0.2)) +
  labs(
    title = "The Hardware Ceiling Across RT Deciles",
    subtitle = "Same-Platform ICC remains high (0.80–0.95) across deciles; Cross-Platform ICC drops in late deciles (instrument chemistry effect)",
    x = "RT Decile", y = "ICC", color = "Configuration", linetype = "Configuration"
  ) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
        axis.text.x = element_text(angle = 0))
```



## 4.3 Implications for Epidemiological Study Design

This hardware ceiling has direct consequences for multi-center metabolomics:
...


## Notes for Collaborator (2026-03-12)

### What is done
- **Part 1** (Feature extraction): Complete. Variance decomposition, PCA, and
  ICC analysis across Full → Detected → Consensus Signal levels all working.
- **Part 2** (RT-dependent normalization): Complete. Global median, decile median,
  and LOESS normalization strategies implemented and compared. Needs to be
  properly interpreted and integrated into the narrative.
- **Part 3** (Peak metrics vs EIC Pearson): Rewritten. Now does simple
  cross-lab correlation of scalar metrics vs EIC profile similarity for
  consensus peaks. Uses `peak_metrics_detected_all4labs.csv` and
  `peak_shape_similarity.csv`. No ML/classification code remains.
- **Part 4** (Hardware ceiling): Same-platform vs cross-platform ICC by RT
  decile is implemented for afekta:cembio (Agilent-Agilent) vs all 4 labs.

### What needs to be finished
1. **part 2 interpretation**:TO be done. I am not sure tha tLoess is reall y
the best.
1. **Part 3**: Run and verify the new code. Check that `ggrepel` is available
   (used in Figure 3.4). Interpret the results once rendered—the narrative text
   is placeholder and should be updated with actual numbers.
2. **Part 4 (Ceiling analysis)**: Currently only compares Agilent-Agilent vs
   all-4-labs.
   Section 4.3 ("Implications for Epidemiological Study Design") is still empty.
3. **Summary section**: Key takeaway #2 has a TODO note ("Conclusion should be
   about how normalization might fix systematic bias but cannot fix noise")—
   needs to be rewritten as a proper conclusion.
4. **Reproducibility metrics**: Currently using ICC and variance decomposition.
   Open question whether CCC (concordance correlation coefficient) or other
   agreement measures would be more appropriate—worth discussing.

### Possible extensions
- Add other chemical classes beyond the HUMAN endosome standard mixes.
- Test with a second peak-picking algorithm (e.g., MZmine centWave vs XCMS
  centWave) to see if the conclusions are algorithm-dependent.
- Tanimoto similarity from `peak_shape_similarity.csv` is loaded but not yet
  used in Part 3—could be added as an additional comparison alongside Pearson.

