---
title: "Defining and Overcoming the Limits of Multi-Center Metabolomics: A Computational Blueprint for Data Integration"
format:
  html:
    toc: true
    toc-location: right
    code-fold: true
execute:
  dir: project
---

# Introduction

## Input Data

All required input data lives in the `object/` subfolder. These files are
produced by the upstream pipeline (`1_object_creation.qmd` →
`2_global_analysis.qmd` → `3_peak_feature_calculation_detected.qmd`).

I can also provide them pre-computed.

**RData objects** (chromatographic profiles, from `1_object_creation.qmd`):
`tic_full.RData`, `tic_detect.RData`


**CSV files** (from `1_object_creation.qmd` and `3_peak_feature_calculation_detected.qmd`):
`detected_peaks_{afekta,cembio,hmgu,icl}_HE.csv`,
`peak_metrics_detected_all4labs.csv`


**Helper files** (in `5_downstream_analysis/`):
`global_analysis_helpers.R`, `lab_setup_phase1.csv`

## Question and Hypothesis

**Core question:** *What are the computational steps required to make multi-center metabolomics data integration possible, and what are the absolute hardware-imposed limits of the field?*

**Hypothesis:** Raw LC-MS data from different laboratories is too confounded by instrumental noise to be directly integrated. Computational processing can recover biological signal, but instrument chemistry imposes an irreducible reproducibility ceiling.

We benchmark four sequential processing strategies:

1. **Peak-picking** (feature extraction vs. raw profiles)
2. **Local (RT-dependent) normalization** (LOESS vs. global scaling)
3. **Peak shape confirmation** (EIC Pearson similarity for cross-lab anchors)
4. **Hardware ceiling** (same-platform ICC as the achievable upper bound)

Data: 19 standard mixtures (HUMAN Endosome) from 4 labs with harmonized LC (same column, mobile phases, gradient) but different MS platforms (Agilent 6546 Q-ToF, Waters Synapt G2S Q-ToF, Sciex Zeno TOF 7600, Agilent 6545 Q-ToF). No technical replicates; lab identity is confounded with instrument platform.

Baseline finding: **lab identity explains 97–100% of variance in raw TIC**
(Full Signal).

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
## this analysisis dependent on R 4.6.0 and Bioc 3.23
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)
library(pheatmap)
library(readxl)

base_path <- here("5_downstream_analysis")
source(file.path(base_path, "global_analysis_helpers.R"))
```

## Setup Information

```{r load-setup}
#| fig-cap: "**Table 1.0**: Laboratory settings."

setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup |>
  filter(details == "harmonized") |>
  select(lab, `LC.Instrument`, `MS.Instrument`,
         `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) |>
  distinct() |>
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) |>
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

## Compute Global Metrics (Raw Signal)

TIC per chromatogram across data levels and polarities; no normalization (added in Part 2).

```{r compute-metrics}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "object/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "object/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = tic$lab,
    mixture = tic$mixture,
    polarity = polarity_name,
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) |>
    left_join(setup_simplified, by = "lab") |>
    mutate(data_level = ifelse(data_level == "full",
                               "Full Signal", "Detected Signal"))
}

metrics_full <- run_for_polarities(compute_metrics, "full")
metrics_detect <- run_for_polarities(compute_metrics, "detect")

# Create Consensus Signal data level from pk_metrics
compute_consensus_metrics <- function(pol, polarity_name, data_level = "detect") {
  pk_metrics <- read.csv(file.path(base_path,
                                   "object/peak_metrics_detected_all4labs.csv"))

  pk_metrics |>
    filter(as.integer(polarity) == pol) |>
    group_by(lab, mixture) |>
    summarise(
      tic = sum(areaUnderTic, na.rm = TRUE),
      .groups = "drop"
    ) |>
    left_join(setup_simplified, by = "lab") |>
    mutate(
      polarity = polarity_name,
      data_level = "Consensus Signal"
    )
}

metrics_consensus <- run_for_polarities(compute_consensus_metrics, "detect")

metrics_combined <- bind_rows(metrics_full, metrics_detect, metrics_consensus)
```

## Data Processing Levels

- **Full Signal**: raw chromatographic profiles (all ions + baseline).
- **Detected Signal**: centWave-picked peaks, per lab (not yet matched).
- **Consensus Signal**: peaks matched across all 4 labs by m/z and RT.

Progression Full → Detected → Consensus tests whether computational processing
progressively removes technical variance.

 
```{r, echo=FALSE, out.width="50%", eval = FALSE}
#| fig-cap: "**Graphical Abstract**: Progressive refinement of the TIC from Full Signal to Detected Signal to Consensus Signal."
knitr::include_graphics(
  file.path(base_path, "FullSignal_DetectedSignal_ConsensusSignal.svg")
)
```
---

# Part 1: The Baseline Problem - Feature Extraction is Mandatory {#part1}

**Core claim:** Raw chromatographic profiles cannot be integrated across labs because vendor-specific baseline noise dominates the signal. Feature extraction (peak-picking) is the mandatory first computational step to make data integration theoretically possible.

## 1.1 The Problem: Lab Variance Dominates Raw Data

### 1.1.1 Variance Decomposition

```{r variance-decomposition}
results_table <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

```{r variance-stacked-bar}
#| fig-cap: "**Figure 1.1.1**: Variance decomposition of TIC across all conditions."

variance_long <- results_table |>
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) |>
  pivot_longer(cols = ends_with("_Pct"),
               names_to = "Source", values_to = "Percentage") |>
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Data_Level = factor(Data_Level,
                        levels = c("Full Signal", "Detected Signal", "Consensus Signal")),
    Label = paste(Polarity, Data_Level, sep = "\n"),
    Label = factor(Label, levels = unique(paste(
      rep(c("Positive", "Negative"), each = 3),
      rep(c("Full Signal", "Detected Signal", "Consensus Signal"), times = 2),
      sep = "\n"
    ))),
    R2_label = paste("R² =", sprintf("%.3f", R_squared))
  )

text_labels <- variance_long |> distinct(Label, R2_label)

ggplot(variance_long, aes(x = Label, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(x = Label, label = R2_label, y = 105),
            data = text_labels, size = 3, inherit.aes = FALSE) +
  labs(title = "Variance Decomposition: TIC Across All Conditions",
       subtitle = "Lab variance (orange) dominates raw signal — the core problem",
       x = "Polarity & Data Level", y = "Variance (%)", fill = "Source") +
  theme_minimal() +
  theme(legend.position = "top") +
  ylim(0, 110)
```
### 1.1.2 Lab Effect Coefficients

```{r lab-coefficients}
#| fig-cap: "**Figure 1.1.2**: Comparison of Laboratory Effect Coefficients against Afekta (TIC)."

all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    all_coefficients <- rbind(all_coefficients,
      result$coefficients |> mutate(Polarity = pol, Data_Level = level))
  }
}

lab_coefs <- all_coefficients |>
  filter(type == "Lab Effect") |>
  mutate(lab_name = gsub("lab", "", coefficient))

lab_coefs <- lab_coefs |>
  mutate(
    Polarity = factor(Polarity, levels = c("Positive", "Negative"))
  )

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "Lab Effect Coefficients (deviation from Afekta)",
       x = "Polarity & Laboratory", y = "Coefficient (log2 scale)", fill = "Lab") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### 1.1.3 PCA Analysis of Lab vs. Mixture Clustering

Colour = mixture, shape = lab.
Same shape clustering → lab dominates; same colour clustering → mixture (biology) dominates.

```{r pca-setup-tic}
# Prepare data for PCA: wide format with samples as rows, features as columns
# For raw data: use individual RT bins as features
# For detected data: use detected peaks as features

# Function to prepare PCA data
prepare_pca_data <- function(data_level = "full", polarity = "Positive") {
  if (data_level == "full") {
    # Load full signal data
    load(file.path(base_path, "object/tic_full.RData"))
    chr_data <- tic_full[tic_full$polarity == ifelse(polarity == "Positive", 1, 0)]
  } else {
    # Load detected signal data
    load(file.path(base_path, "object/tic_detect.RData"))
    chr_data <- tic_detect[tic_detect$polarity == ifelse(polarity == "Positive", 1, 0)]
  }

  # Extract intensity matrices
  # Each row = one sample, columns = RT bins
  int_list <- intensity(chr_data)
  rt_list <- rtime(chr_data)

  # Create common RT grid
  all_rts <- unlist(rt_list)
  rt_grid <- seq(min(all_rts, na.rm = TRUE),
                 max(all_rts, na.rm = TRUE),
                 length.out = 500)

  # Interpolate each chromatogram onto the grid
  int_matrix <- t(sapply(seq_along(int_list), function(i) {
    approx(rt_list[[i]], int_list[[i]], xout = rt_grid,
           method = "linear", rule = 2)$y
  }))

  # Add sample metadata
  rownames(int_matrix) <- paste0(chr_data$lab, "_", chr_data$mixture)
  colnames(int_matrix) <- paste0("RT_", round(rt_grid, 1))

  # Create metadata data frame
  meta_df <- data.frame(
    sample = rownames(int_matrix),
    lab = chr_data$lab,
    mixture = chr_data$mixture,
    row.names = rownames(int_matrix)
  )

  list(data = int_matrix, meta = meta_df)
}

# Prepare datasets for both polarities
pca_full_pos <- prepare_pca_data("full", "Positive")
pca_full_neg <- prepare_pca_data("full", "Negative")
pca_detect_pos <- prepare_pca_data("detect", "Positive")
pca_detect_neg <- prepare_pca_data("detect", "Negative")

# Prepare consensus data from pk_metrics (wide format: samples x consensus peaks)
# Filter to one polarity at a time — consensus_ids are disjoint across polarities
# and the two modes have different intensity scales.
pk_metrics <- read.csv(file.path(base_path,
                                 "object/peak_metrics_detected_all4labs.csv"))

prepare_pca_consensus <- function(polarity_int) {
  wide <- pk_metrics |>
    filter(as.integer(polarity) == polarity_int) |>
    mutate(sample = paste0(lab, "_", mixture)) |>
    select(sample, consensus_id, areaUnderTic, lab, mixture) |>
    pivot_wider(names_from = consensus_id, values_from = areaUnderTic,
                values_fill = 0, names_prefix = "peak_") |>
    arrange(sample)

  data_mat <- wide |> select(starts_with("peak_")) |> as.matrix()
  rownames(data_mat) <- wide$sample

  list(data = data_mat, meta = wide |> select(sample, lab, mixture))
}

pca_consensus_pos <- prepare_pca_consensus(1L)
pca_consensus_neg <- prepare_pca_consensus(0L)

# Helper: run prcomp and return a scores data frame tagged with polarity.
# `transform_fn` applies a transform before PCA (identity for continuous TIC
# profiles, log1p for sparse consensus peaks).
pca_scores_df <- function(pca_list, polarity_label, transform_fn = identity) {
  res <- prcomp(transform_fn(pca_list$data), scale. = TRUE, center = TRUE)
  var_expl <- (res$sdev^2 / sum(res$sdev^2)) * 100
  list(
    scores = data.frame(
      PC1 = res$x[, 1], PC2 = res$x[, 2],
      lab = pca_list$meta$lab, mixture = pca_list$meta$mixture,
      polarity = polarity_label
    ),
    var_expl = var_expl
  )
}

# Common plot: 2-panel PCA (Positive | Negative).
# Colour encodes mixture (biological signal) and shape encodes lab (technical).
# As data level progresses Full → Detected → Consensus, colours should tighten
# (mixture clusters emerge) while shapes should disperse within each colour
# (lab effects shrink).
plot_pca_by_polarity <- function(pos, neg, title, subtitle) {
  scores <- bind_rows(pos$scores, neg$scores) |>
    mutate(polarity = factor(polarity, levels = c("Positive", "Negative")))
  panel_labels <- c(
    Positive = sprintf("Positive  (PC1 %.1f%%, PC2 %.1f%%)",
                       pos$var_expl[1], pos$var_expl[2]),
    Negative = sprintf("Negative  (PC1 %.1f%%, PC2 %.1f%%)",
                       neg$var_expl[1], neg$var_expl[2])
  )
  ggplot(scores, aes(x = PC1, y = PC2, color = mixture, shape = lab)) +
    geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    facet_wrap(~polarity, scales = "free",
               labeller = labeller(polarity = panel_labels)) +
    scale_color_viridis_d(option = "turbo") +
    labs(title = title, subtitle = subtitle,
         x = "PC1", y = "PC2", color = "Mixture", shape = "Laboratory") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))
}
```

```{r pca-raw-tic}
#| fig-cap: "**Figure 1.1.3.1**: PCA of Raw TIC profiles, Positive and Negative polarity. Colour = mixture, shape = lab. Same-shape points cluster (labs group together); colours are scattered — technical noise dominates biological signal."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_full_pos, "Positive"),
  neg = pca_scores_df(pca_full_neg, "Negative"),
  title = "PCA: Raw TIC Profiles (Full Signal)",
  subtitle = "Shapes cluster (lab dominates), colours scattered (mixture signal buried)"
)
```

```{r pca-detected-tic}
#| fig-cap: "**Figure 1.1.3.2**: PCA of Detected Peaks, Positive and Negative polarity. Colour = mixture, shape = lab. Lab clusters weaken and same-colour points start to group together — feature extraction recovers biological signal."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_detect_pos, "Positive"),
  neg = pca_scores_df(pca_detect_neg, "Negative"),
  title = "PCA: Raw TIC Detected Peaks (Peak-Picked Signal)",
  subtitle = "Shapes disperse, colours begin to group — mixture signal emerges"
)
```

```{r pca-consensus-tic}
#| fig-cap: "**Figure 1.1.3.3**: PCA of Consensus Peaks, Positive and Negative polarity. Colour = mixture, shape = lab. Same-colour points cluster tightly regardless of shape — mixture dominates, lab effects minimized. log1p compresses the intensity range because each peak has ~4 non-zero values (labs) and ~72 zeros (other mixtures); without it the presence/absence gap dominates scaling."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_consensus_pos, "Positive", transform_fn = log1p),
  neg = pca_scores_df(pca_consensus_neg, "Negative", transform_fn = log1p),
  title = "PCA: Consensus Peaks (Cross-Lab Matched Signal)",
  subtitle = "Colours cluster by mixture across all shapes — biology dominates technical variance"
)
```

```{r consensus-matrix-example}
#| fig-cap: "**Table 1.1.3.1**: Example structure of the consensus peak matrix."
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

# data_pca_consensus_pos <- pca_consensus_neg[["data"]]
# data_pca_consensus_pos <- as.data.frame(data_pca_consensus_pos)

library(tibble)
library(knitr)

# Example structure of the consensus peak matrix
# Rows = samples (lab + mixture)
# Columns = consensus peaks
# Most values are zeros because each consensus peak
# only exists in one biological mixture.

example_consensus_matrix <- tribble(
  ~sample,        ~peak_1, ~peak_2, ~peak_3,
  "LabA_mix1",      10000,        0,        0,
  "LabB_mix1",       9000,        0,        0,
  "LabC_mix1",      11000,        0,        0,
  "LabD_mix1",      12000,        0,        0,
  "LabA_mix2",          0,     7000,        0,
  "LabB_mix2",          0,     6800,        0,
  "LabC_mix2",          0,     7200,        0,
  "LabD_mix2",          0,     7100,        0
)

kable(
  example_consensus_matrix,
  caption = paste0(
    "<span style='white-space: nowrap;'><b>Example structure of the consensus peak matrix.</b></span>",
    "<br>",
    "<span style='white-space: nowrap;'>",
    "Each consensus peak is only present within one mixture ",
    "(across laboratories), producing many zeros.",
    "</span>",
    "<br><br>"
  ),
  align = "lccc",
  escape = FALSE
)
```

# Uniform Manifold Approximation and Projection (UMAP)
UMAP works much better with sparse data, local structures, partial presence/absence and non-linear relationships. It also tends to reveal much clearer biological clusters.

```{r umap-consensus-tic}
#| fig-cap: "**Figure 1.1.3.1.4**: UMAP projection of Consensus Peaks. Colour = mixture, shape = lab. UMAP preserves local similarity structure in sparse consensus matrices better than PCA, revealing tight mixture-specific clusters across laboratories."
#| fig-width: 10
#| fig-height: 6

library(uwot)
library(dplyr)
library(ggplot2)

run_umap <- function(pca_list, polarity_label) {

  um <- umap(
    log1p(pca_list$data),
    n_neighbors = 10,
    min_dist = 0.2,
    metric = "cosine",
    scale = TRUE
  )

  data.frame(
    UMAP1 = um[,1],
    UMAP2 = um[,2],
    lab = pca_list$meta$lab,
    mixture = pca_list$meta$mixture,
    polarity = polarity_label
  )
}

umap_pos <- run_umap(pca_consensus_pos, "Positive")
umap_neg <- run_umap(pca_consensus_neg, "Negative")

umap_scores <- bind_rows(umap_pos, umap_neg) |>
  mutate(
    polarity = factor(polarity,
                      levels = c("Positive", "Negative"))
  )

ggplot(
  umap_scores,
  aes(x = UMAP1, y = UMAP2,
      color = mixture,
      shape = lab)
) +
  geom_point(size = 3, alpha = 0.9) +
  facet_wrap(~polarity) +
  scale_color_viridis_d(option = "turbo") +
  guides(color = guide_legend(ncol = 2)) +
  labs(
    title = "UMAP: Consensus Peaks",
    subtitle = "UMAP handles sparse consensus matrices better than PCA",
    x = "UMAP1",
    y = "UMAP2",
    color = "Mixture",
    shape = "Laboratory"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

# Presence/absence Heatmap
This option is convenient for visually illustrating the conceptual problem.
Here, it is possible to immediately understand that each mixture has its own cluster of peaks and the rest are zeros, which is why PCA fails

```{r heatmap-consensus-presence}
#| fig-cap: "**Figure 1.1.3.1.5**: Presence/absence heatmap of Consensus Peaks. Each mixture contains a distinct subset of consensus peaks shared across laboratories, producing a highly sparse block-structured matrix."
#| fig-width: 12
#| fig-height: 8

library(pheatmap)
library(dplyr)

# Positive mode example
consensus_binary <- ifelse(pca_consensus_pos$data > 0, 1, 0)

# Sample annotations
annotation_df <- data.frame(
  Mixture = pca_consensus_pos$meta$mixture,
  Lab = pca_consensus_pos$meta$lab
)

rownames(annotation_df) <- rownames(consensus_binary)

# Optional: keep only variable peaks
keep_cols <- colSums(consensus_binary) > 0
consensus_binary <- consensus_binary[, keep_cols]

ann_colors <- list(
  Lab = c(
    afekta = "#E57373",
    hmgu   = "#64B5F6",
    icl    = "#81C784",
    cembio = "#FFD54F"
  )
)

# Plot
pheatmap(
  consensus_binary,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  show_colnames = FALSE,
  annotation_row = annotation_df,
  annotation_colors = ann_colors,
  main = "Consensus Peak Presence/Absence Matrix\nRows = Samples | Columns = Consensus Peaks",
  border_color = NA
)
```

Consensus peak matrices are highly sparse because each consensus feature is only present within a single biological mixture across laboratories. Under this structure, Euclidean PCA becomes dominated by absence patterns and zero inflation. Therefore, non-Euclidean similarity methods better suited for sparse compositional data (UMAP with cosine distance and PCoA using Jaccard distance) were additionally employed.

#### 1.1.3.2 BPC

```{r pca-setup-bpc}
# Prepare data for PCA: wide format with samples as rows, features as columns
# For raw data: use individual RT bins as features
# For detected data: use detected peaks as features

# Function to prepare PCA data
prepare_pca_data <- function(data_level = "full", polarity = "Positive") {
  if (data_level == "full") {
    # Load full signal data
    load(file.path(base_path, "object/bpc_full.RData"))
    chr_data <- bpc_full[bpc_full$polarity == ifelse(polarity == "Positive", 1, 0)]
  } else {
    # Load detected signal data
    load(file.path(base_path, "object/bpc_detect.RData"))
    chr_data <- bpc_detect[bpc_detect$polarity == ifelse(polarity == "Positive", 1, 0)]
  }

  # Extract intensity matrices
  # Each row = one sample, columns = RT bins
  int_list <- intensity(chr_data)
  rt_list <- rtime(chr_data)

  # Create common RT grid
  all_rts <- unlist(rt_list)
  rt_grid <- seq(min(all_rts, na.rm = TRUE),
                 max(all_rts, na.rm = TRUE),
                 length.out = 500)

  # Interpolate each chromatogram onto the grid
  int_matrix <- t(sapply(seq_along(int_list), function(i) {
    approx(rt_list[[i]], int_list[[i]], xout = rt_grid,
           method = "linear", rule = 2)$y
  }))

  # Add sample metadata
  rownames(int_matrix) <- paste0(chr_data$lab, "_", chr_data$mixture)
  colnames(int_matrix) <- paste0("RT_", round(rt_grid, 1))

  # Create metadata data frame
  meta_df <- data.frame(
    sample = rownames(int_matrix),
    lab = chr_data$lab,
    mixture = chr_data$mixture,
    row.names = rownames(int_matrix)
  )

  list(data = int_matrix, meta = meta_df)
}

# Prepare datasets for both polarities
pca_full_pos <- prepare_pca_data("full", "Positive")
pca_full_neg <- prepare_pca_data("full", "Negative")
pca_detect_pos <- prepare_pca_data("detect", "Positive")
pca_detect_neg <- prepare_pca_data("detect", "Negative")

# Prepare consensus data from pk_metrics (wide format: samples x consensus peaks)
# Filter to one polarity at a time — consensus_ids are disjoint across polarities
# and the two modes have different intensity scales.
pk_metrics <- read.csv(file.path(base_path,
                                 "object/peak_metrics_detected_all4labs.csv"))

prepare_pca_consensus <- function(polarity_int) {
  wide <- pk_metrics |>
    filter(as.integer(polarity) == polarity_int) |>
    mutate(sample = paste0(lab, "_", mixture)) |>
    select(sample, consensus_id, areaUnderTic, lab, mixture) |>
    pivot_wider(names_from = consensus_id, values_from = areaUnderTic,
                values_fill = 0, names_prefix = "peak_") |>
    arrange(sample)

  data_mat <- wide |> select(starts_with("peak_")) |> as.matrix()
  rownames(data_mat) <- wide$sample

  list(data = data_mat, meta = wide |> select(sample, lab, mixture))
}

pca_consensus_pos <- prepare_pca_consensus(1L)
pca_consensus_neg <- prepare_pca_consensus(0L)

# Helper: run prcomp and return a scores data frame tagged with polarity.
# `transform_fn` applies a transform before PCA (identity for continuous TIC
# profiles, log1p for sparse consensus peaks).
pca_scores_df <- function(pca_list, polarity_label, transform_fn = identity) {
  res <- prcomp(transform_fn(pca_list$data), scale. = TRUE, center = TRUE)
  var_expl <- (res$sdev^2 / sum(res$sdev^2)) * 100
  list(
    scores = data.frame(
      PC1 = res$x[, 1], PC2 = res$x[, 2],
      lab = pca_list$meta$lab, mixture = pca_list$meta$mixture,
      polarity = polarity_label
    ),
    var_expl = var_expl
  )
}

# Common plot: 2-panel PCA (Positive | Negative).
# Colour encodes mixture (biological signal) and shape encodes lab (technical).
# As data level progresses Full → Detected → Consensus, colours should tighten
# (mixture clusters emerge) while shapes should disperse within each colour
# (lab effects shrink).
plot_pca_by_polarity <- function(pos, neg, title, subtitle) {
  scores <- bind_rows(pos$scores, neg$scores) |>
    mutate(polarity = factor(polarity, levels = c("Positive", "Negative")))
  panel_labels <- c(
    Positive = sprintf("Positive  (PC1 %.1f%%, PC2 %.1f%%)",
                       pos$var_expl[1], pos$var_expl[2]),
    Negative = sprintf("Negative  (PC1 %.1f%%, PC2 %.1f%%)",
                       neg$var_expl[1], neg$var_expl[2])
  )
  ggplot(scores, aes(x = PC1, y = PC2, color = mixture, shape = lab)) +
    geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    facet_wrap(~polarity, scales = "free",
               labeller = labeller(polarity = panel_labels)) +
    scale_color_viridis_d(option = "turbo") +
    labs(title = title, subtitle = subtitle,
         x = "PC1", y = "PC2", color = "Mixture", shape = "Laboratory") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))
}
```

```{r pca-raw-bpc}
#| fig-cap: "**Figure 1.1.3.2.1**: PCA of BPC profiles, Positive and Negative polarity. Colour = mixture, shape = lab. Same-shape points cluster (labs group together); colours are scattered — technical noise dominates biological signal."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_full_pos, "Positive"),
  neg = pca_scores_df(pca_full_neg, "Negative"),
  title = "PCA: BPC Profiles (Full Signal)",
  subtitle = "Shapes cluster (lab dominates), colours scattered (mixture signal buried)"
)
```

```{r pca-detected-bpc}
#| fig-cap: "**Figure 1.1.3.2.2**: PCA of BPC Detected Peaks, Positive and Negative polarity. Colour = mixture, shape = lab. Lab clusters weaken and same-colour points start to group together — feature extraction recovers biological signal."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_detect_pos, "Positive"),
  neg = pca_scores_df(pca_detect_neg, "Negative"),
  title = "PCA: BPC Detected Peaks (Peak-Picked Signal)",
  subtitle = "Shapes disperse, colours begin to group — mixture signal emerges"
)
```

```{r pca-consensus-bpc}
#| fig-cap: "**Figure 1.1.3.2.3**: PCA of Consensus Peaks, Positive and Negative polarity. Colour = mixture, shape = lab. Same-colour points cluster tightly regardless of shape — mixture dominates, lab effects minimized. log1p compresses the intensity range because each peak has ~4 non-zero values (labs) and ~72 zeros (other mixtures); without it the presence/absence gap dominates scaling."
#| fig-width: 12
#| fig-height: 6

plot_pca_by_polarity(
  pos = pca_scores_df(pca_consensus_pos, "Positive", transform_fn = log1p),
  neg = pca_scores_df(pca_consensus_neg, "Negative", transform_fn = log1p),
  title = "PCA: Consensus Peaks (Cross-Lab Matched Signal)",
  subtitle = "Colours cluster by mixture across all shapes — biology dominates technical variance"
)

# The PCA of Consensus Peaks in the BPC section is the same as the PCA of Consensus Peaks in the Raw TIC section, given that pca_consensus_pos and pca_consensus_neg have only been able to be constructed from pk_metrics using areaUnderTic (areaUnderBpc was not found)

```

**Conclusion:** Raw profile data cannot be integrated across labs;
feature extraction followed by cross-lab matching is required.

## 1.2 ICC: Full Signal vs Detected Signal

### 1.2.1 ICC Computation

ICC quantifies the ratio of biological signal to total variance:

$$ICC = \frac{\sigma^2_{between}}{\sigma^2_{between} + \sigma^2_{within}} \tag{Equation 1}$$

- $\sigma^2_{between}$ = variance *between* mixtures (biological signal)
- $\sigma^2_{within}$ = variance *within* the same mixture across labs (technical noise)
- ICC > 0.5 → acceptable reproducibility

```{r icc-computation}
# Raw signal ICC only — normalization strategies introduced in Part 2
icc_all_results <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level)

    results <- compute_icc(data_subset, "tic") |>
      mutate(Polarity = pol, Data_Level = level, Metric_Label = "TIC")

    icc_all_results <- bind_rows(icc_all_results, results)
  }
}
```

```{r icc-computation-afekta-cembio}
# Raw signal ICC only — Afekta vs Cembio comparison

icc_all_results_afekta_cembio <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {

    data_subset <- metrics_combined |>
      filter(
        polarity == pol,
        data_level == level,
        lab %in% c("afekta", "cembio")
      )

    results <- compute_icc(data_subset, "tic")

    if (!all(is.na(data_subset$bpc))) {
      results <- bind_rows(
        results,
        compute_icc(data_subset, "bpc")
      )
    }

    results <- results |>
      mutate(
        Polarity = pol,
        Data_Level = level,
        Metric_Label = case_when(
          Metric == "tic" ~ "TIC",
          Metric == "bpc" ~ "BPC",
          TRUE ~ Metric
        )
      )

    icc_all_results_afekta_cembio <- bind_rows(
      icc_all_results_afekta_cembio,
      results
    )
  }
}
```

### 1.2.2 Peak Detection Impact

```{r peak-detection-impact-tic}
#| fig-cap: "**Table 1.2.2**: ICC improvement from peak detection (raw TIC) in both positive and negative mode."

icc_comparison_tic <- icc_all_results |>
  filter(Metric == "tic") |>
  select(Polarity, Data_Level, ICC) |>
  pivot_wider(names_from = Data_Level, values_from = ICC)

knitr::kable(icc_comparison_tic,
             caption = "ICC improvement from peak detection (raw TIC)",
             digits = 3)
```

```{r peak-detection-plot-tic}

#| fig-cap: "**Figure 1.2.2**: Raw TIC ICC across data levels. 
default_cols <- scales::hue_pal()(3)

custom_cols <- c(
  "Full Signal" = default_cols[3],
  "Detected Signal" = default_cols[2],
  "Consensus Signal" = default_cols[1]
)

ggplot(icc_all_results |>
         filter(Metric == "tic") |>
         mutate(Data_Level = factor(Data_Level,
                                    levels = c("Full Signal", "Detected Signal", "Consensus Signal")),
                Polarity = factor(Polarity,
                                  levels = c("Positive", "Negative"))
                ),
       aes(x = Polarity, y = ICC, fill = Data_Level)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.3f", ICC)),
            position = position_dodge(0.9), vjust = -0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_fill_manual(values = custom_cols) +
  labs(title = "Impact of Peak Detection on Reproducibility",
       subtitle = "Raw TIC (no normalization); dashed = 0.5 threshold",
       y = "ICC", x = "Polarity", fill = "Processing Level") +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```

##### 1.2.2.1.2 Afekta vs Cembio

```{r peak-detection-impact-tic-afekta-cembio}
#| fig-cap: "**Table 1.2.2.1.2**: ICC improvement from peak detection (raw TIC) comparing only Afekta and Cembio."

icc_comparison_tic_afekta_cembio <- icc_all_results_afekta_cembio |>
  filter(Metric == "tic") |>
  select(Polarity, Data_Level, ICC) |>
  pivot_wider(names_from = Data_Level, values_from = ICC)

knitr::kable(
  icc_comparison_tic_afekta_cembio,
  caption = "ICC improvement from peak detection (raw TIC) — Afekta vs Cembio",
  digits = 3
)
```

```{r}
#| fig-cap: "**Figure 1.2.2.1.2**: Raw TIC ICC across data levels comparing Afekta vs Cembio only. Consensus Signal achieves substantially higher ICC than Full or Detected Signal, demonstrating that cross-lab matched peaks recover the underlying biological structure."

default_cols <- scales::hue_pal()(3)

custom_cols <- c(
  "Full Signal" = default_cols[3],
  "Detected Signal" = default_cols[2],
  "Consensus Signal" = default_cols[1]
)

ggplot(
  icc_all_results_afekta_cembio |>
    filter(Metric == "tic") |>
    mutate(
      Data_Level = factor(
        Data_Level,
        levels = c("Full Signal", "Detected Signal", "Consensus Signal")
      ),
      Polarity = factor(
        Polarity,
        levels = c("Positive", "Negative")
      )
    ),
  aes(x = Polarity, y = ICC, fill = Data_Level)
) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(
    aes(label = sprintf("%.3f", ICC)),
    position = position_dodge(0.9),
    vjust = -0.5
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_fill_manual(values = custom_cols) +
  labs(
    title = "Impact of Peak Detection on Reproducibility",
    subtitle = "Afekta vs Cembio only; Raw TIC (no normalization)",
    y = "ICC",
    x = "Polarity",
    fill = "Processing Level"
  ) +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```



### 1.2.3 ICC Variance Decomposition

```{r icc-variance-decomposition}
#| fig-cap: "**Figure 1.2.3**: Peak detection shifts variance toward biological signal (raw signal, no normalization)."

variance_plot <- icc_all_results |>
  select(Polarity, Data_Level, Metric_Label,
         Between_mixture_var, Within_mixture_var) |>
  pivot_longer(cols = c(Between_mixture_var, Within_mixture_var),
               names_to = "Source", values_to = "Variance") |>
  mutate(Source = case_when(
    Source == "Between_mixture_var" ~ "Between mixtures (signal)",
    Source == "Within_mixture_var" ~ "Within mixtures (lab noise)"
  ))

variance_plot <- variance_plot |>
  mutate(
    Polarity = factor(Polarity, levels = c("Positive", "Negative"))
  )

ggplot(variance_plot,
       aes(x = Metric_Label, y = Variance, fill = Source)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.7) +
  facet_grid(Data_Level ~ Polarity, scales = "free_y") +
  scale_fill_manual(values = c(
    "Between mixtures (signal)" = "#619CFF",
    "Within mixtures (lab noise)" = "#F8766D")) +
  labs(title = "Variance Decomposition: Raw TIC",
       x = "Polarity & Metric", y = "Variance (log2 scale)", fill = "Source") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

### 1.2.4 Same-Platform Ceiling: Afekta vs Cembio

Afekta and Cembio share the same MS platform family (Agilent Q-TOF).
Restricting ICC to this pair gives the **practical ceiling** for
cross-lab reproducibility in this trial: what agreement looks like
when instrument chemistry is held constant. Same data levels, same
metric — only the lab set differs.

```{r icc-pair-computation}
icc_pair_levels <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in c("Full Signal", "Detected Signal", "Consensus Signal")) {
    data_subset <- metrics_combined |>
      filter(polarity == pol, data_level == level,
             lab %in% c("afekta", "cembio"))

    results <- compute_icc(data_subset, "tic") |>
      mutate(Polarity = pol, Data_Level = level, Metric_Label = "TIC")

    icc_pair_levels <- bind_rows(icc_pair_levels, results)
  }
}
```

```{r icc-pair-table}
#| fig-cap: "**Table 1.2.4**: Same-platform pairwise ICC (Afekta + Cembio) across data levels."

icc_pair_levels |>
  select(Polarity, Data_Level, ICC) |>
  pivot_wider(names_from = Data_Level, values_from = ICC) |>
  knitr::kable(
    caption = "Pairwise ICC (Afekta vs Cembio) across data levels",
    digits = 3
  )
```

```{r icc-pair-plot}
#| fig-cap: "**Figure 1.2.4**: Same-platform pairwise ICC (Afekta + Cembio) across data levels — the practical ceiling for cross-lab reproducibility when MS chemistry is held constant."

ggplot(icc_pair_levels |>
         mutate(Data_Level = factor(Data_Level,
                                    levels = c("Full Signal",
                                               "Detected Signal",
                                               "Consensus Signal")),
                Polarity = factor(Polarity,
                                  levels = c("Positive", "Negative"))),
       aes(x = Polarity, y = ICC, fill = Data_Level)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.3f", ICC)),
            position = position_dodge(0.9), vjust = -0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_fill_manual(values = custom_cols) +
  labs(title = "Same-Platform Ceiling: Afekta vs Cembio",
       subtitle = "Pairwise ICC across data levels; dashed = 0.5 threshold",
       y = "ICC", x = "Polarity", fill = "Processing Level") +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 1)
```
---

# Part 2: Addressing RT-Dependent Bias Makes Patient Pooling Possible {#strategy2}

**Core claim:** RT-dependent drift between labs makes patient pooling
statistically invalid without local normalization.

**Data scope:** Consensus Signal — cross-lab matched peaks (peaks present
in all 4 labs), at per-peak Area Under TIC granularity.

## 2.0 Add Normalization Variants

Median-scaled TIC, per (lab, polarity, data_level).

```{r add-normalization-variants}
metrics_combined <- metrics_combined |>
  group_by(lab, polarity, data_level) |>
  mutate(
    tic_median_scaled = tic / median(tic, na.rm = TRUE)
  ) |>
  ungroup()
```

## 2.1 Normalization Impact

### 2.1.0 Why Median Scaling vs. No Normalization?

- **Median scaling**: removes systematic per-lab sensitivity differences;
  appropriate when the design is balanced (all labs run the same 19 mixtures).
- **No normalization**: baseline — lab-to-lab intensity differences remain.

### 2.1.1 Global Normalization: Median Scaling

```{r normalization-variance-tic}
#| fig-cap: "**Figure 2.1.1**: Effect of normalization on variance decomposition (TIC), per polarity (Consensus Signal)."
#| fig-width: 10
#| fig-height: 5

normalization_results <- bind_rows(lapply(
  c("Positive", "Negative"), function(pol) {
    test_data <- metrics_combined |>
      filter(polarity == pol, data_level == "Consensus Signal")
    result_raw <- check_metric_quality(test_data, "tic")
    result_median <- check_metric_quality(test_data, "tic_median_scaled")
    bind_rows(
      result_raw$variance_table |>
        mutate(Normalization = "Raw TIC",
               R_squared = result_raw$r_squared),
      result_median$variance_table |>
        mutate(Normalization = "Median Scaled",
               R_squared = result_median$r_squared)
    ) |>
      mutate(Polarity = pol)
  }
))

normalization_results |>
  mutate(
    Normalization = factor(Normalization,
                           levels = c("Raw TIC", "Median Scaled")),
    Polarity = factor(Polarity, levels = c("Positive", "Negative"))
  ) |>
  ggplot(aes(x = Normalization, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Polarity) +
  labs(
    title = "Effect of Normalization on Variance Decomposition (TIC)",
    subtitle = "Consensus Signal, per polarity",
    x = "Normalization Method", y = "Variance (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 2.1.2 Residual Comparison

```{r normalization-residuals-tic}
#| fig-cap: "**Figure 2.1.2**: Effect of normalization on residual distribution (TIC), per polarity (Consensus Signal)."
#| fig-height: 14
#| fig-width: 12

residual_plots <- lapply(c("Positive", "Negative"), function(pol) {
  test_data <- metrics_combined |>
    filter(polarity == pol, data_level == "Consensus Signal")
  mod_raw <- lm(log2(tic) ~ lab + mixture, data = test_data)
  mod_median <- lm(log2(tic_median_scaled) ~ lab + mixture,
                   data = test_data)
  (
    plot_residuals_new(mod_raw,
                       paste0(pol, " — Raw TIC (log2)"), test_data) /
    plot_residuals_new(mod_median,
                       paste0(pol, " — Median Scaled TIC"), test_data)
  )
})

(residual_plots[[1]] | residual_plots[[2]]) +
  plot_annotation(
    title = "Residual Plots: Raw vs Median Scaled (per polarity)",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )
```

## 2.2 The Problem: Non-Linear Sensitivity Profiles

Both §2.2.1 and §2.2.2 are computed on the **Consensus Signal**:
per-peak `areaUnderTic` from `peak_metrics_detected_all4labs.csv` joined
with each peak's apex RT from the per-lab `detected_peaks_*_HE.csv`
tables, then aggregated to per-(sample × RT quartile) sums.

```{r consensus-quartile-data-prep}
# Per-lab peak tables to get apex RT for each peak
pt_all_labs <- bind_rows(
  read.csv(file.path(base_path,
                     "object/detected_peaks_afekta_HE.csv")) |>
    mutate(lab = "afekta"),
  read.csv(file.path(base_path,
                     "object/detected_peaks_hmgu_HE.csv")) |>
    mutate(lab = "hmgu"),
  read.csv(file.path(base_path,
                     "object/detected_peaks_icl_HE.csv")) |>
    mutate(lab = "icl"),
  read.csv(file.path(base_path,
                     "object/detected_peaks_cembio_HE.csv")) |>
    mutate(lab = "cembio")
)

# Consensus peaks + apex RT per peak
pk_metrics_rt <- read.csv(file.path(base_path,
                                    "object/peak_metrics_detected_all4labs.csv")) |>
  left_join(pt_all_labs |> select(chrom_peak_id, lab, rt),
            by = c("chrom_peak_id", "lab")) |>
  mutate(polarity_label = ifelse(as.integer(polarity) == 1,
                                 "Positive", "Negative"))

# Global RT quartile breaks (across all consensus peaks)
rt_breaks_q <- quantile(pk_metrics_rt$rt,
                        probs = seq(0, 1, 0.25), na.rm = TRUE)
pk_metrics_rt$rt_quartile <- cut(
  pk_metrics_rt$rt, breaks = rt_breaks_q,
  labels = paste0("Q", 1:4), include.lowest = TRUE
)

# Per-sample × quartile sums of areaUnderTic (Consensus Signal TIC by quartile)
quartile_sums <- pk_metrics_rt |>
  filter(!is.na(rt_quartile)) |>
  group_by(lab, mixture, polarity_label, rt_quartile) |>
  summarise(quartile_tic = sum(areaUnderTic, na.rm = TRUE),
            .groups = "drop") |>
  rename(Polarity = polarity_label, Quartile = rt_quartile)
```

### 2.2.1 Instrument Sensitivity Across RT Quartiles

Non-parallel lines across quartiles → RT-dependent biases that global
normalization cannot fix.

```{r sensitivity-profile-tic}
# Consortium median per (polarity, quartile)
consortium_q_medians <- quartile_sums |>
  group_by(Polarity, Quartile) |>
  summarise(Consensus = median(quartile_tic, na.rm = TRUE),
            .groups = "drop")

all_profiles <- quartile_sums |>
  left_join(consortium_q_medians, by = c("Polarity", "Quartile")) |>
  mutate(Log2_Deviation = log2(quartile_tic / Consensus))
```

```{r sensitivity-profile-plot-tic}
#| fig-cap: "**Figure 2.2.1**: Non-parallel sensitivity profiles (Consensus Signal TIC) → global normalization insufficient."

all_profiles |>
  mutate(Log2_TIC = log2(quartile_tic)) |>
  group_by(Quartile, lab, Polarity) |>
  summarise(median_log2_tic = median(Log2_TIC, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = Quartile, y = median_log2_tic,
             group = lab, color = lab)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~Polarity) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Consensus TIC Distribution Across RT Quartiles",
    subtitle = "Non-parallel lines → global normalization is insufficient",
    y = "log2(Consensus TIC per Quartile)", x = "Quartile",
    color = "Laboratory"
  ) +
  theme_minimal()
```

```{r sensitivity-deviation-table-tic}
#| fig-cap: "**Table 2.2.1**: Consortium median comparison by lab and RT quartile (Consensus TIC)"

findings_table <- all_profiles |>
  group_by(lab, Quartile, Polarity) |>
  summarise(Median_Log2_Dev = round(
              median(Log2_Deviation, na.rm = TRUE), 2),
            .groups = "drop") |>
  pivot_wider(names_from = Quartile, values_from = Median_Log2_Dev)

knitr::kable(findings_table,
             caption = "Median log2 deviation from consortium median by lab and RT quartile (Consensus Signal)")
```

### 2.2.2 Residual Lab Variance by RT Quartile

```{r quartile-variance}
# Per-(lab, polarity) median of per-sample total Consensus TIC
sample_totals <- pk_metrics_rt |>
  group_by(lab, mixture, polarity_label) |>
  summarise(total_tic = sum(areaUnderTic, na.rm = TRUE),
            .groups = "drop")

lab_medians_q <- sample_totals |>
  group_by(lab, polarity_label) |>
  summarise(median_tic = median(total_tic, na.rm = TRUE),
            .groups = "drop") |>
  rename(Polarity = polarity_label)

# Fraction of lab-typical Consensus TIC per (sample, quartile)
quartile_fracs <- quartile_sums |>
  left_join(lab_medians_q, by = c("lab", "Polarity")) |>
  mutate(TIC_Frac = quartile_tic / median_tic) |>
  select(lab, mixture, Polarity, Quartile, TIC_Frac) |>
  pivot_wider(names_from = Quartile, values_from = TIC_Frac,
              names_prefix = "TIC_Frac_") |>
  rename(polarity = Polarity)

quartile_results <- expand.grid(
  Polarity = c("Positive", "Negative"),
  Metric = paste0("TIC_Frac_Q", 1:4)
) |>
  rowwise() |>
  mutate(var_result = list(
    get_variance(filter(quartile_fracs, polarity == Polarity), Metric)
  )) |>
  unnest(var_result)
```

```{r quartile-variance-plot-tic}
#| fig-cap: "**Figure 2.2.2**: Residual lab variance of Consensus TIC across RT quartiles, after per-lab median scaling."

plot_quartile_variance(quartile_results, "TIC")
```

## 2.3 Comparing Normalization Strategies

**Question:** can normalization push ICC above the post-detection baseline
(raw TIC ICC ≈ 0.024–0.059), and can RT-dependent normalization fix
quartile-specific biases?

Three granularities compared (per-peak Area Under TIC):

1. **Global Median** — one scaling factor per lab
2. **Decile Median** — per-RT-decile scaling factors (10 bins)
3. **LOESS** — continuous RT-dependent normalization

### 2.3.1 Prepare Peak-Level Data

```{r load-peak-data}
pk_metrics <- read.csv(file.path(base_path,
                                 "object/peak_metrics_detected_all4labs.csv"))

cat("Peak metrics:", nrow(pk_metrics), "rows\n")
cat("Unique consensus peaks:", n_distinct(pk_metrics$consensus_id), "\n")
```

### 2.3.2 Apply Normalization Strategies

```{r normalization-strategies}
# Load RT breaks from detected signal (deciles instead of quartiles)
load(file.path(base_path, "object/tic_detect.RData"))
all_rts <- unlist(rtime(tic_detect))
rt_breaks_decile <- quantile(all_rts, probs = seq(0, 1, by = 0.1), na.rm = TRUE)

# Load peak tables with RT info
pt_all_labs <- bind_rows(
  read.csv(file.path(base_path, "object/detected_peaks_afekta_HE.csv")) |>
    mutate(lab = "afekta"),
  read.csv(file.path(base_path, "object/detected_peaks_hmgu_HE.csv")) |>
    mutate(lab = "hmgu"),
  read.csv(file.path(base_path, "object/detected_peaks_icl_HE.csv")) |>
    mutate(lab = "icl"),
  read.csv(file.path(base_path, "object/detected_peaks_cembio_HE.csv")) |>
    mutate(lab = "cembio")
)

# Add RT and decile assignment (10 bins: D1 to D10)
pk_norm <- pk_metrics |>
  left_join(
    pt_all_labs |> select(chrom_peak_id, rt, lab),
    by = c("chrom_peak_id", "lab")
  ) |>
  mutate(
    rt_decile = cut(rt,
                    breaks = rt_breaks_decile,
                    labels = paste0("D", 1:10),
                    include.lowest = TRUE)
  )

# Positive and negative ionization have very different intensity distributions,
# so all three strategies compute normalization factors per (lab, polarity) —
# pooling modes would bias each one's scaling factor.

# --- Strategy 1: Global Median ---
global_medians <- pk_norm |>
  group_by(lab, polarity) |>
  summarise(
    global_median_area = median(areaUnderTic, na.rm = TRUE),
    global_median_max = median(maxIntensity, na.rm = TRUE),
    .groups = "drop"
  )

grand_medians <- global_medians |>
  group_by(polarity) |>
  summarise(
    grand_median_area = median(global_median_area),
    grand_median_max = median(global_median_max),
    .groups = "drop"
  )

pk_norm <- pk_norm |>
  left_join(global_medians, by = c("lab", "polarity")) |>
  left_join(grand_medians, by = "polarity") |>
  mutate(
    area_global = areaUnderTic / global_median_area * grand_median_area,
    max_global = maxIntensity / global_median_max * grand_median_max
  )

# --- Strategy 2: Decile Median (NEW) ---
decile_medians <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(lab, polarity, rt_decile) |>
  summarise(
    d_median_area = median(areaUnderTic, na.rm = TRUE),
    d_median_max = median(maxIntensity, na.rm = TRUE),
    .groups = "drop"
  )

grand_d_medians <- decile_medians |>
  group_by(polarity, rt_decile) |>
  summarise(
    grand_d_area = median(d_median_area),
    grand_d_max = median(d_median_max),
    .groups = "drop"
  )

pk_norm <- pk_norm |>
  left_join(decile_medians, by = c("lab", "polarity", "rt_decile")) |>
  left_join(grand_d_medians, by = c("polarity", "rt_decile")) |>
  mutate(
    area_decile = areaUnderTic / d_median_area * grand_d_area,
    max_decile = maxIntensity / d_median_max * grand_d_max
  )

# --- Strategy 3: LOESS ---
pk_norm <- pk_norm |>
  group_by(lab, polarity) |>
  mutate(
    loess_area = {
      fit <- loess(log2(areaUnderTic) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    },
    loess_max = {
      fit <- loess(log2(maxIntensity) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    }
  ) |>
  ungroup()

# Grand LOESS (consensus curve across all labs, fit separately per polarity)
pk_norm <- pk_norm |>
  group_by(polarity) |>
  mutate(
    grand_loess_area = {
      fit <- loess(log2(areaUnderTic) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    },
    grand_loess_max = {
      fit <- loess(log2(maxIntensity) ~ rt, span = 0.5,
                   na.action = na.exclude)
      predict(fit)
    }
  ) |>
  ungroup() |>
  mutate(
    area_loess = 2^(log2(areaUnderTic) - loess_area + grand_loess_area),
    max_loess = 2^(log2(maxIntensity) - loess_max + grand_loess_max)
  )

cat("Normalization complete. Rows:", nrow(pk_norm), "\n")
```

### 2.3.3 LOESS

#### 2.3.3.1 Visualize LOESS Sensitivity Profiles

```{r loess-profiles}
#| fig-cap: "**Figure 2.3.3.1**: Per-lab LOESS sensitivity curves showing the RT-dependent bias that local normalization corrects."

pk_norm |>
  filter(!is.na(rt), !is.na(loess_area)) |>
  ggplot(aes(x = rt, y = loess_area, color = lab)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "LOESS Sensitivity Profiles by Lab",
    subtitle = "log2(Area) smoothed over RT — divergence = what local norm corrects",
    x = "Retention Time (s)", y = "LOESS-smoothed log2(Area)",
    color = "Laboratory"
  ) +
  theme_minimal()
```

#### 2.3.3.2 Log2 Fold-Error: Proof of LOESS Correction

Per-peak Log2 Fold-Error $= \log_2(\text{Lab Area} / \text{Consensus Median Area})$.
Before normalization, labs drift from zero (especially Q3/Q4). After LOESS,
all labs re-center on zero if the systematic bias is corrected.

```{r log2-fold-error-calculation}
# Calculate consensus median area for each peak (across all labs)
consensus_areas <- pk_norm |>
  group_by(consensus_id, rt_decile) |>
  summarise(
    consensus_median_raw = median(areaUnderTic, na.rm = TRUE),
    consensus_median_global = median(area_global, na.rm = TRUE),
    consensus_median_decile = median(area_decile, na.rm = TRUE),
    consensus_median_loess = median(area_loess, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate log2 fold-error for each lab
fold_error <- pk_norm |>
  left_join(consensus_areas, by = c("consensus_id", "rt_decile")) |>
  mutate(
    log2_fe_raw = log2(areaUnderTic / consensus_median_raw),
    log2_fe_global = log2(area_global / consensus_median_global),
    log2_fe_decile = log2(area_decile / consensus_median_decile),
    log2_fe_loess = log2(area_loess / consensus_median_loess),
    Polarity = ifelse(as.integer(polarity) == 1, "Positive", "Negative")
  ) |>
  filter(!is.na(rt_decile), !is.na(log2_fe_raw)) |>
  select(lab, Polarity, consensus_id, rt_decile,
         log2_fe_raw, log2_fe_global, log2_fe_decile, log2_fe_loess)

cat("Log2 Fold-Error dataset:", nrow(fold_error), "rows\n")
```

### 2.3.4 Log2 Fold-Error Visualization

```{r log2-fold-error-plot}
#| fig-cap: "**Figure 2.3.4**: Log2 Fold-Error by RT Decile, Lab, and Polarity. Before correction (Raw), labs drift from zero in late deciles. After LOESS, all labs re-center on zero across deciles in both polarities."
#| fig-width: 12
#| fig-height: 8

fe_long <- fold_error |>
  pivot_longer(
    cols = starts_with("log2_fe_"),
    names_to = "strategy",
    names_prefix = "log2_fe_",
    values_to = "log2_fold_error"
  ) |>
  mutate(
    strategy = case_when(
      strategy == "raw" ~ "Raw Area",
      strategy == "global" ~ "Global Median",
      strategy == "decile" ~ "Decile Median",
      strategy == "loess" ~ "LOESS"
    ),
    strategy = factor(strategy, levels = c("Raw Area", "Global Median",
                                            "Decile Median", "LOESS")),
    Polarity = factor(Polarity, levels = c("Positive", "Negative"))
  )

ggplot(fe_long, aes(x = rt_decile, y = log2_fold_error, fill = lab)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_grid(Polarity ~ strategy) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Log2 Fold-Error: Raw vs Normalized Areas",
    subtitle = "LOESS re-centers all labs on zero (systematic bias correction), per polarity",
    x = "RT Decile", y = "log2(Lab Area / Consensus Median)", fill = "Lab"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Findings**:
- **Raw Area**: ICL D1 +0.45, D4 +0.64, D5 −0.56, D7 −0.60, D8 −0.68;
  Cembio D8 +0.52, D10 −1.26.
- **Global Median**: drift persists (RT-specific bias unaddressed).
- **LOESS**: all labs re-centered on zero across deciles.

### 2.3.5 ICC by Normalization Strategy

ICC computed on **peak-level Area Under TIC** after each normalization
strategy, per RT decile, on the **Consensus Signal** (peaks matched in
all 4 labs).

#### 2.3.5.1 By RT Decile

```{r icc-peak-function}
compute_icc_peak <- function(data, value_col) {
  wide <- data |>
    filter(!is.na(.data[[value_col]])) |>
    select(consensus_id, lab, value = all_of(value_col)) |>
    mutate(value = log2(value)) |>
    pivot_wider(names_from = lab, values_from = value)

  # Filter to only rows with all 4 labs present
  # Get actual lab columns (may not always be all 4)
  lab_cols <- intersect(c("afekta", "hmgu", "icl", "cembio"), colnames(wide))
  wide <- wide |> filter(if_all(all_of(lab_cols), ~ !is.na(.x)))

  if (nrow(wide) < 5) {
    return(data.frame(
      Metric = value_col, ICC = NA_real_,
      Between_var = NA_real_, Within_var = NA_real_, n_peaks = nrow(wide)))
  }

  mat <- as.matrix(wide[, lab_cols])
  row_means <- rowMeans(mat, na.rm = TRUE)
  grand_mean <- mean(row_means, na.rm = TRUE)

  k <- ncol(mat); n <- nrow(mat)
  ss_between <- k * sum((row_means - grand_mean)^2)
  ss_within <- sum(sweep(mat, 1, row_means)^2, na.rm = TRUE)

  ms_between <- ss_between / (n - 1)
  ms_within <- ss_within / (n * (k - 1))
  icc_val <- (ms_between - ms_within) / (ms_between + (k - 1) * ms_within)

  data.frame(
    Metric = value_col, ICC = icc_val,
    Between_var = ms_between, Within_var = ms_within, n_peaks = n
  )
}
```

```{r icc-normalization-comparison}
# Consensus Signal: peaks matched across all 4 labs
consensus_signal <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  group_by(consensus_id) |>
  filter(n() == 4) |>
  ungroup() |>
  mutate(Polarity = ifelse(as.integer(polarity) == 1,
                           "Positive", "Negative"))

strat_cols <- c(areaUnderTic = "Raw",
                area_global  = "Global Median",
                area_decile  = "Decile Median",
                area_loess   = "LOESS")

icc_norm_results <- do.call(rbind, lapply(c("Positive", "Negative"),
  function(pol_label) {
    pol_data <- consensus_signal |> filter(Polarity == pol_label)

    by_decile <- do.call(rbind, lapply(paste0("D", 1:10), function(dec) {
      sub <- pol_data |> filter(rt_decile == dec)
      do.call(rbind, lapply(names(strat_cols), function(col) {
        compute_icc_peak(sub, col) |>
          mutate(Strategy = strat_cols[[col]], Decile = dec,
                 Polarity = pol_label)
      }))
    }))

    overall <- do.call(rbind, lapply(names(strat_cols), function(col) {
      compute_icc_peak(pol_data, col) |>
        mutate(Strategy = strat_cols[[col]], Decile = "Overall",
               Polarity = pol_label)
    }))

    bind_rows(by_decile, overall)
  }))

# Per-polarity decile tables (ICC + n_peaks)
for (pol in c("Positive", "Negative")) {
  pol_results <- icc_norm_results |>
    filter(Polarity == pol, Decile != "Overall") |>
    mutate(Strategy = factor(Strategy,
              levels = c("Raw", "Global Median", "Decile Median", "LOESS")))

  icc_tbl <- pol_results |>
    select(Strategy, Decile, ICC) |>
    pivot_wider(names_from = Decile, values_from = ICC) |>
    select(Strategy, D1, D2, D3, D4, D5, D6, D7, D8, D9, D10)
  print(knitr::kable(
    icc_tbl,
    caption = sprintf("Table 2.3.5.1 (%s): ICC by normalization strategy and RT decile", pol),
    digits = 3))

  # Consensus peak count per decile (taken from Raw — same set across strategies).
  # By construction n is the number of peaks complete in all 4 labs, identical
  # across labs.
  n_tbl <- pol_results |>
    filter(Strategy == "Raw") |>
    select(Decile, n_peaks) |>
    pivot_wider(names_from = Decile, values_from = n_peaks) |>
    select(D1, D2, D3, D4, D5, D6, D7, D8, D9, D10)
  print(knitr::kable(
    n_tbl,
    caption = sprintf("Table 2.3.5.1 (%s, n_peaks): consensus peaks per decile (complete across all 4 labs)", pol)))
}
```

```{r icc-normalization-line-plot}
#| fig-cap: "**Figure 2.3.5.1**: ICC by Normalization Strategy and RT Decile, per polarity (Consensus Signal). Local normalization (Decile Median, LOESS) improves ICC most in late deciles (D7–D10) where RT-dependent bias is strongest."
#| fig-width: 12
#| fig-height: 5

icc_plot_data <- icc_norm_results |>
  filter(Decile != "Overall", !is.na(ICC)) |>
  mutate(
    Decile_num = as.numeric(gsub("D", "", Decile)),
    Strategy = factor(Strategy, levels = c("Raw", "Global Median",
                                            "Decile Median", "LOESS")),
    Polarity = factor(Polarity, levels = c("Positive", "Negative"))
  )

ggplot(icc_plot_data, aes(x = Decile_num, y = ICC,
                          color = Strategy, linetype = Strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Polarity) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray", alpha = 0.5) +
  scale_x_continuous(breaks = 1:10, labels = paste0("D", 1:10)) +
  scale_color_manual(values = c("Raw" = "#E64B35", "Global Median" = "#4DBBD5",
                                 "Decile Median" = "#70AD47", "LOESS" = "#4472C4")) +
  scale_linetype_manual(values = c("Raw" = "solid", "Global Median" = "solid",
                                    "Decile Median" = "solid", "LOESS" = "dashed")) +
  labs(
    title = "ICC by Normalization Strategy and RT Decile",
    subtitle = "Per polarity. Local normalization helps most in later deciles",
    x = "RT Decile", y = "ICC (Area Under TIC)",
    color = "Strategy", linetype = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor.y = element_blank()
  ) +
  ylim(NA, 1)
```

#### 2.3.5.2 Global ICC across all RT

```{r icc-overall-comparison}
#| fig-cap: "**Figure 2.3.5.2**: Overall ICC across strategies (Consensus Signal, aggregated across all deciles), per polarity."
#| fig-width: 10
#| fig-height: 5

icc_norm_results |>
  filter(Decile == "Overall") |>
  mutate(Strategy = factor(Strategy, levels = c("Raw", "Global Median",
                                                 "Decile Median", "LOESS")),
         Polarity = factor(Polarity, levels = c("Positive", "Negative"))) |>
  ggplot(aes(x = Strategy, y = ICC, fill = Strategy)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~Polarity) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_text(aes(label = sprintf("%.3f", ICC)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Raw" = "#E64B35", "Global Median" = "#4DBBD5",
                               "Decile Median" = "#70AD47", "LOESS" = "#4472C4")) +
  labs(
    title = "Overall ICC by Normalization Strategy",
    subtitle = "Consensus Signal — Area Under TIC, per polarity; dashed line at ICC = 0.5",
    y = "ICC", x = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylim(NA, 1)
```

```{r}
#| fig-cap: "**Table 2.3.5.2**: Global ICC across all RT per strategy and polarity"

icc_norm_results |>
  filter(Decile == "Overall") |>
  mutate(Strategy = factor(Strategy,
            levels = c("Raw", "Global Median", "Decile Median", "LOESS"))) |>
  select(Polarity, Strategy, ICC, Between_var, Within_var, n_peaks) |>
  arrange(Polarity, Strategy) |>
  knitr::kable(
    caption = "Table 2.3.5.2: Global ICC across all RT values, per polarity",
    digits = 3
  )
```

#### 2.3.5.3 Pairwise ICC: Afekta vs Cembio

This analysis evaluates reproducibility between only two laboratories (Afekta and Cembio), allowing direct assessment of pairwise agreement after normalization.

```{r icc-afekta-cembio}
compute_icc_peak_pair <- function(data, value_col,
                                  labs_keep = c("afekta", "cembio")) {

  na_result <- function(n) data.frame(
    Metric = value_col, ICC = NA_real_,
    Between_var = NA_real_, Within_var = NA_real_, n_peaks = n
  )

  wide <- data |>
    filter(lab %in% labs_keep) |>
    filter(!is.na(.data[[value_col]])) |>
    select(consensus_id, lab, value = all_of(value_col)) |>
    mutate(value = log2(value)) |>
    pivot_wider(names_from = lab, values_from = value)

  # After pivot, some labs may be missing (decile has no peaks for them).
  # Need both labs present as columns to compute pairwise ICC.
  lab_cols <- intersect(labs_keep, colnames(wide))
  if (length(lab_cols) < 2) return(na_result(0))

  # Keep only peaks present in both labs
  wide <- wide |> filter(if_all(all_of(lab_cols), ~ !is.na(.x)))

  if (nrow(wide) < 5) return(na_result(nrow(wide)))

  mat <- as.matrix(wide[, lab_cols])

  row_means <- rowMeans(mat, na.rm = TRUE)
  grand_mean <- mean(row_means, na.rm = TRUE)

  k <- ncol(mat)
  n <- nrow(mat)

  ss_between <- k * sum((row_means - grand_mean)^2)
  ss_within <- sum(sweep(mat, 1, row_means)^2, na.rm = TRUE)

  ms_between <- ss_between / (n - 1)
  ms_within <- ss_within / (n * (k - 1))

  icc_val <- (ms_between - ms_within) /
    (ms_between + (k - 1) * ms_within)

  data.frame(
    Metric = value_col,
    ICC = icc_val,
    Between_var = ms_between,
    Within_var = ms_within,
    n_peaks = n
  )
}

# Prepare data — restricted to Afekta + Cembio, polarity column carried through
pair_signal <- pk_norm |>
  filter(!is.na(rt_decile)) |>
  filter(lab %in% c("afekta", "cembio")) |>
  mutate(Polarity = ifelse(as.integer(polarity) == 1,
                           "Positive", "Negative"))

icc_pair_results <- do.call(rbind, lapply(c("Positive", "Negative"),
  function(pol_label) {
    pol_data <- pair_signal |> filter(Polarity == pol_label)

    by_decile <- do.call(rbind, lapply(paste0("D", 1:10), function(dec) {
      sub <- pol_data |> filter(rt_decile == dec)
      do.call(rbind, lapply(names(strat_cols), function(col) {
        compute_icc_peak_pair(sub, col) |>
          mutate(Strategy = strat_cols[[col]], Decile = dec,
                 Polarity = pol_label)
      }))
    }))

    overall <- do.call(rbind, lapply(names(strat_cols), function(col) {
      compute_icc_peak_pair(pol_data, col) |>
        mutate(Strategy = strat_cols[[col]], Decile = "Overall",
               Polarity = pol_label)
    }))

    bind_rows(by_decile, overall)
  }))
```

```{r}
#| fig-cap: "**Figure 2.3.5.3.1**: Pairwise ICC between Afekta and Cembio across RT deciles, per polarity."
#| fig-width: 12
#| fig-height: 5

icc_pair_plot <- icc_pair_results |>
  filter(Decile != "Overall") |>
  mutate(
    Decile_num = as.numeric(gsub("D", "", Decile)),
    Strategy = factor(Strategy,
      levels = c("Raw", "Global Median", "Decile Median", "LOESS")),
    Polarity = factor(Polarity, levels = c("Positive", "Negative"))
  )

ggplot(icc_pair_plot,
       aes(x = Decile_num, y = ICC,
           color = Strategy, linetype = Strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Polarity) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_x_continuous(breaks = 1:10, labels = paste0("D", 1:10)) +
  scale_color_manual(values = c("Raw" = "#E64B35",
                                "Global Median" = "#4DBBD5",
                                "Decile Median" = "#70AD47",
                                "LOESS" = "#4472C4")) +
  scale_linetype_manual(values = c("Raw" = "solid",
                                   "Global Median" = "solid",
                                   "Decile Median" = "solid",
                                   "LOESS" = "dashed")) +
  labs(
    title = "Pairwise ICC: Afekta vs Cembio",
    subtitle = "Per polarity. Local normalization helps most in later RT regions",
    x = "RT Decile", y = "ICC"
  ) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.minor = element_blank()) +
  ylim(NA, 1)
```

```{r}
#| fig-cap: "**Figure 2.3.5.3.2**: Overall pairwise ICC between Afekta and Cembio across normalization strategies, per polarity."
#| fig-width: 10
#| fig-height: 5

icc_pair_results |>
  filter(Decile == "Overall") |>
  mutate(Strategy = factor(Strategy,
            levels = c("Raw", "Global Median", "Decile Median", "LOESS")),
         Polarity = factor(Polarity, levels = c("Positive", "Negative"))) |>
  ggplot(aes(x = Strategy, y = ICC, fill = Strategy)) +
  geom_col(alpha = 0.85) +
  facet_wrap(~Polarity) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_text(aes(label = sprintf("%.3f", ICC)),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Raw" = "#E64B35",
                               "Global Median" = "#4DBBD5",
                               "Decile Median" = "#70AD47",
                               "LOESS" = "#4472C4")) +
  labs(
    title = "Overall Pairwise ICC: Afekta vs Cembio",
    subtitle = "ICC aggregated across all RT values, per polarity",
    x = NULL, y = "ICC"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylim(0, 1)
```

```{r}
#| fig-cap: "**Table 2.3.5.3.1**: Global pairwise ICC between Afekta and Cembio, per polarity."

icc_pair_results |>
  filter(Decile == "Overall") |>
  mutate(Strategy = factor(Strategy,
            levels = c("Raw", "Global Median", "Decile Median", "LOESS"))) |>
  select(Polarity, Strategy, ICC, Between_var, Within_var, n_peaks) |>
  arrange(Polarity, Strategy) |>
  knitr::kable(
    caption = "Table 2.3.5.3.1: Pairwise ICC across all RT values, per polarity",
    digits = 3
  )
```

**Open question for Part 3:** LOESS removes systematic RT-dependent bias
(Figure 2.3) but ICC gains are modest. Hypothesis: residual integration noise
(baseline boundaries, scan-to-scan fluctuation) sets the ICC ceiling, not
systematic bias. Part 3 tests this by replacing scalar integration metrics
with EIC profile similarity.

