---
title: "Downstream Analysis: Adduct Grouping & Peak Shape Similarity"
format: html
editor: source
toc: true
---

## 1. Setup and Data Loading

Mian goal is to explore method to group adduct of same compound in an unsupervised way

We begin by loading the annotated peak data and the raw EICs. The goal is to
establish methods to group peaks derived from the same compound
(adducts, in-source fragments) using both scalar features and raw data shapes.

In the futur i want to load the full detected eics (whihc will be names eics_detected)
an the annotated peakonly eics will be named  eics_ann. for this is need to run
teh new object creation code ont he cluster but it's not possible at the moment
therefore i will just set up the code as much as I can and correct it when i have the
updated objects.

essentially: 
- we have ground truth on the annotated peaks
- we want to test different methods to group peaks based on shape similarity
- we use all detected peaks for the unsupervised grouping as it is more realistic
- the peak shape metrics in the first part are derrived from the eics that we are calcualting similarity of peak shape below. also want to know which is best to group adduct between the full rzaw eics and the peak shape metrics.


```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(xcms)
library(MsCoreUtils)
library(MsFeatures)
library(mclust)
library(pROC)
library(Chromatograms)

base_path <- here::here("5_downstream_analysis")

# Load data
load(file = file.path(base_path, "object/full_peak_report.RData"))
load(file = file.path(base_path, "object/final_eics.RData"))

df_peaks <- as.data.frame(full_peak_report)
```

## 2. Peak Shape Similarity based on Features (Scalar Metrics)

**Hypothesis:** Adducts of the same compound elute at the same time and share physical peak shape properties (FWHM, Tailing, Asymmetry) driven by the column interaction.

### 2.1. Define Ground Truth (For Validation)

We use ALL detected peaks for clustering, but evaluate performance only on the 
subset with annotations (our ground truth).

```{r}
# Mark which peaks have annotations
df_peaks$is_annotated <- !is.na(df_peaks$compound_name)
df_peaks$true_group_id <- ifelse(
  df_peaks$is_annotated,
  paste0(df_peaks$compound_name, "_", df_peaks$lab, "_", df_peaks$mixture),
  NA  # Unknown peaks have no ground truth
)

cat("Total peaks:", nrow(df_peaks), "\n")
cat("Annotated peaks:", sum(df_peaks$is_annotated), 
    "(", round(100 * mean(df_peaks$is_annotated), 1), "%)\n")
cat("Unique compounds:", length(unique(na.omit(df_peaks$compound_name))), "\n")
```

### 2.2. Feature Consistency

Test which peak shape features are most consistent within true compound groups.
This uses only annotated peaks to establish which features are promising.

```{r}
features_to_test <- c("fwhm", "tailing_factor", "gaussian_similarity",
                      "rt_width", "entropy")

# Calculate CV for each feature within true groups (annotated only)
df_annotated <- df_peaks[df_peaks$is_annotated, ]

results_features <- map_dfr(features_to_test, function(feat) {
  df_annotated %>%
    filter(!is.na(.data[[feat]])) %>%
    group_by(true_group_id) %>%
    filter(n() > 1) %>%  # Need at least 2 peaks per group
    summarise(cv = sd(.data[[feat]]) / mean(.data[[feat]]) * 100, .groups = "drop") %>%
    summarise(median_cv = median(cv, na.rm = TRUE)) %>%
    mutate(feature = feat)
}) %>%
  arrange(median_cv) %>%
  as.data.frame()

print(results_features)

ggplot(results_features, aes(x = reorder(feature, median_cv), y = median_cv)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Feature Consistency (lower CV = better)", 
       subtitle = "Based on annotated peaks only",
       x = "Feature", y = "Median CV (%)")
```

### 2.3 Unsupervised Clustering by Features

Cluster ALL peaks, then evaluate using only the annotated subset.

#### Pre-clustering by Retention Time

```{r}
# Use ALL peaks
df_peaks$lab_mixture <- paste(df_peaks$lab, df_peaks$mixture, sep = "_")
inj_list <- split(df_peaks, df_peaks$lab_mixture)

# Group co-eluting peaks (within 10 seconds)
group_l <- lapply(inj_list, function(dt) {
  groupFeatures(dt$rt_apex, SimilarRtimeParam(diffRt = 10))
})

cat("Lab/mixture combinations:", length(inj_list), "\n")
cat("Total peaks being clustered:", nrow(df_peaks), "\n")
```

#### Clustering and Evaluation

```{r}
# Simple clustering function
cluster_feature <- function(values, k = 2) {
  if (length(unique(na.omit(values))) <= 1) return(rep(1, length(values)))
  valid <- !is.na(values)
  clusters <- rep(NA, length(values))
  if (sum(valid) > 1) {
    hc <- hclust(dist(values[valid]), method = "ward.D2")
    clusters[valid] <- cutree(hc, k = min(k, sum(valid)))
  }
  clusters
}

# Test each feature - cluster ALL peaks, evaluate on ANNOTATED only
ari_results <- map_dfr(features_to_test, function(feat) {
  combined <- map_dfr(names(inj_list), function(lab_mix) {
    dt <- inj_list[[lab_mix]]
    dt$rt_group <- group_l[[lab_mix]]
    dt$cluster <- ave(dt[[feat]], dt$rt_group, FUN = cluster_feature)
    dt
  })
  
  # Evaluate only on annotated peaks
  eval_subset <- combined[combined$is_annotated & !is.na(combined$cluster), ]
  
  data.frame(
    feature = feat,
    n_evaluated = nrow(eval_subset),
    ARI = adjustedRandIndex(eval_subset$true_group_id, eval_subset$cluster)
  )
}) %>%
  arrange(-ARI)

print(ari_results)

ggplot(ari_results, aes(x = reorder(feature, ARI), y = ARI)) +
  geom_col(fill = "darkgreen") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Clustering Performance by Feature (ARI)", 
       subtitle = "Clustering all peaks, evaluating on annotated subset",
       x = "", y = "ARI")
```

### 2.4 Multi-Metric Clustering (PCA-based)

Combine multiple features using PCA, then cluster on principal components.

```{r}
# Scale and combine features
features_multi <- c("fwhm", "tailing_factor", "gaussian_similarity", "rt_width", "entropy")

# Function to cluster using multiple features via PCA
cluster_pca <- function(feature_matrix, k = 2, n_pcs = 2) {
  n <- nrow(feature_matrix)
  if (n < 3) return(rep(1, n))
  
  # Remove rows with any NA
  complete_rows <- complete.cases(feature_matrix)
  if (sum(complete_rows) < 3) return(rep(NA, n))
  
  mat <- feature_matrix[complete_rows, , drop = FALSE]
  
  # Remove constant columns (zero variance)
  col_vars <- apply(mat, 2, var, na.rm = TRUE)
  valid_cols <- !is.na(col_vars) & col_vars > 1e-10
  
  if (sum(valid_cols) == 0) {
    # All features constant - assign all to same cluster
    clusters <- rep(NA, n)
    clusters[complete_rows] <- 1
    return(clusters)
  }
  
  mat <- mat[, valid_cols, drop = FALSE]
  
  # Scale and run PCA
  scaled <- scale(mat)
  scaled[is.na(scaled)] <- 0
  
  # Use tryCatch for safety
  pca <- tryCatch(
    prcomp(scaled, center = FALSE, scale. = FALSE),
    error = function(e) NULL
  )
  
  if (is.null(pca)) {
    clusters <- rep(NA, n)
    clusters[complete_rows] <- 1
    return(clusters)
  }
  
  pcs <- pca$x[, 1:min(n_pcs, ncol(pca$x)), drop = FALSE]
  
  # Cluster
  hc <- hclust(dist(pcs), method = "ward.D2")
  clusters <- rep(NA, n)
  clusters[complete_rows] <- cutree(hc, k = min(k, sum(complete_rows)))
  clusters
}

# Test multi-metric clustering
ari_multi <- map_dfr(names(inj_list), function(lab_mix) {
  dt <- inj_list[[lab_mix]]
  dt$rt_group <- group_l[[lab_mix]]
  
  # Cluster each RT group using all features
  dt$cluster_multi <- ave(
    seq_len(nrow(dt)), dt$rt_group,
    FUN = function(idx) {
      feat_mat <- as.matrix(dt[idx, features_multi])
      cluster_pca(feat_mat, k = 2, n_pcs = 2)
    }
  )
  dt
}) |>
  filter(is_annotated, !is.na(cluster_multi))

ari_multi_result <- adjustedRandIndex(ari_multi$true_group_id, ari_multi$cluster_multi)

cat("Multi-metric (PCA) clustering ARI:", round(ari_multi_result, 3), "\n")

# Add to results for comparison
ari_results <- rbind(
  ari_results,
  data.frame(feature = "MULTI_PCA", n_evaluated = nrow(ari_multi), ARI = ari_multi_result)
) |>
  arrange(-ARI)

ggplot(ari_results, aes(x = reorder(feature, ARI), y = ARI)) +
  geom_col(aes(fill = feature == "MULTI_PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("steelblue", "darkgreen"), guide = "none") +
  coord_flip() +
  labs(title = "Clustering Performance: Single vs Multi-Metric",
       x = "", y = "ARI")
```

## 3. Peak Shape Similarity based on Raw Data (EICs)

**Hypothesis:** EICs of adducts are perfectly correlated 
(Pearson r \> 0.9) because they are slices of the same chromatographic band.

### 3.1. Data Preparation

```{r}
ints <- Chromatograms::intensity(final_eics)
rts <- rtime(final_eics)
names(ints) <- names(rts) <- chromData(final_eics)$chrom_peak_id

cat("Total EICs:", length(ints), "\n")
```

### 3.2. Calculate EIC Correlation (Pearson)

```{r}
# Interpolate EICs to common RT axis
interpolate_eics <- function(rt_list, int_list, n = 100) {
  rt_min <- max(sapply(rt_list, min))
  rt_max <- min(sapply(rt_list, max))
  if (rt_max <= rt_min) return(NULL)
  
  rt_common <- seq(rt_min, rt_max, length.out = n)
  lapply(seq_along(rt_list), function(i) {
    approx(rt_list[[i]], int_list[[i]], xout = rt_common)$y
  })
}

# Calculate Pearson correlation within RT groups
eic_cors <- map_dfr(names(inj_list), function(lab_mix) {
  dt <- inj_list[[lab_mix]]
  dt$rt_group <- group_l[[lab_mix]]
  
  map_dfr(unique(dt$rt_group), function(rg) {
    grp <- dt[dt$rt_group == rg, ]
    if (nrow(grp) < 2) return(NULL)
    
    ids <- as.character(grp$chrom_peak_id)
    ids <- ids[ids %in% names(ints)]
    if (length(ids) < 2) return(NULL)
    
    grp <- grp[as.character(grp$chrom_peak_id) %in% ids, ]
    interp <- interpolate_eics(rts[ids], ints[ids])
    if (is.null(interp)) return(NULL)
    
    # Pairwise correlations
    n <- length(ids)
    expand.grid(i = 1:n, j = 1:n) |>
      filter(i < j) |>
      rowwise() |>
      mutate(
        peak_1 = ids[i], peak_2 = ids[j],
        pearson = cor(interp[[i]], interp[[j]], use = "complete.obs"),
        both_annotated = grp$is_annotated[i] & grp$is_annotated[j],
        same_compound = if (both_annotated) {
          grp$true_group_id[i] == grp$true_group_id[j]
        } else {
          NA
        }
      ) |>
      ungroup() |>
      select(peak_1, peak_2, pearson, both_annotated, same_compound)
  })
})

cat("Total EIC pairs:", nrow(eic_cors), "\n")
cat("Both annotated:", sum(eic_cors$both_annotated), "\n")
cat("  - Same compound:", sum(eic_cors$same_compound, na.rm = TRUE), "\n")
cat("  - Different compound:", sum(!eic_cors$same_compound, na.rm = TRUE), "\n")
```

### 3.3. Evaluation

```{r}
# Filter to annotated pairs
eic_annotated <- eic_cors[eic_cors$both_annotated, ]

# Summary stats
eic_annotated |>
  group_by(same_compound) |>
  summarise(
    n = n(),
    mean = mean(pearson, na.rm = TRUE),
    median = median(pearson, na.rm = TRUE),
    sd = sd(pearson, na.rm = TRUE),
    .groups = "drop"
  )

# Distribution plot
ggplot(eic_annotated, aes(x = pearson, fill = same_compound)) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = 0.9, linetype = "dashed") +
  labs(title = "EIC Pearson Correlation: Same vs Different Compounds",
       x = "Pearson Correlation", fill = "Same Compound")

# ROC analysis
roc_eic <- roc(eic_annotated$same_compound, eic_annotated$pearson,
               levels = c(FALSE, TRUE), direction = "<", quiet = TRUE)
eic_auc <- as.numeric(auc(roc_eic))
best_thresh <- coords(roc_eic, "best")

cat("\nAUC:", round(eic_auc, 3), "\n")
cat("Optimal threshold:", round(best_thresh$threshold, 3),
    "(sens:", round(best_thresh$sensitivity, 2),
    ", spec:", round(best_thresh$specificity, 2), ")\n")

plot(roc_eic, main = paste("ROC Curve (AUC =", round(eic_auc, 3), ")"))
```

## 4. Cross-Lab Analysis

Since all labs used the same chromatographic column and mobile phases, peak shape 
should be consistent across labs. Only the MS detection differs. This allows us to:

1. Pool data across labs for more datapoints per compound
2. Define ground truth as `compound_name + mixture` (ignoring lab)
3. Evaluate if cross-lab consistency improves clustering

### 4.1 Cross-Lab Ground Truth

```{r}
# Redefine ground truth: compound + mixture (pooled across labs)
df_peaks$cross_lab_group <- ifelse(
  df_peaks$is_annotated,
  paste0(df_peaks$compound_name, "_", df_peaks$mixture),
  NA
)

cat("Cross-lab compound groups:", length(unique(na.omit(df_peaks$cross_lab_group))), "\n")

# How many peaks per cross-lab group?
df_peaks |>
  filter(is_annotated) |>
  group_by(cross_lab_group) |>
  summarise(n_peaks = n(), n_labs = n_distinct(lab), .groups = "drop") |>
  summarise(
    total_groups = n(),
    median_peaks = median(n_peaks),
    max_peaks = max(n_peaks),
    groups_multi_lab = sum(n_labs > 1)
  )
```

### 4.2 Cross-Lab Feature Consistency

Test if peak shape features are consistent for the same compound across different labs.

```{r}
# CV within cross-lab groups
cross_lab_cv <- map_dfr(features_to_test, function(feat) {
  df_peaks |>
    filter(is_annotated, !is.na(.data[[feat]])) |>
    group_by(cross_lab_group) |>
    filter(n() > 1, n_distinct(lab) > 1) |>  # Must have multiple labs
    summarise(
      cv = sd(.data[[feat]]) / mean(.data[[feat]]) * 100,
      n_labs = n_distinct(lab),
      .groups = "drop"
    ) |>
    summarise(
      median_cv = median(cv, na.rm = TRUE),
      n_groups = n()
    ) |>
    mutate(feature = feat)
}) |>
  arrange(median_cv)

print(cross_lab_cv)

# Compare within-lab vs cross-lab CV
cv_comparison <- results_features |>
  rename(within_lab_cv = median_cv) |>
  left_join(cross_lab_cv |> select(feature, cross_lab_cv = median_cv), by = "feature")

print(cv_comparison)

cv_comparison |>
  pivot_longer(cols = c(within_lab_cv, cross_lab_cv), names_to = "type", values_to = "cv") |>
  ggplot(aes(x = reorder(feature, cv), y = cv, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Feature Consistency: Within-Lab vs Cross-Lab",
       x = "", y = "Median CV (%)", fill = "")
```

### 4.3 Cross-Lab Clustering Evaluation

Cluster by mixture (pooling labs), evaluate using cross-lab ground truth.

```{r}
# Split by mixture only (pool labs)
inj_list_cross <- split(df_peaks, df_peaks$mixture)

# Group co-eluting peaks
group_l_cross <- lapply(inj_list_cross, function(dt) {
  groupFeatures(dt$rt_apex, SimilarRtimeParam(diffRt = 10))
})

# Cluster and evaluate
ari_cross <- map_dfr(features_to_test, function(feat) {
  combined <- map_dfr(names(inj_list_cross), function(mix) {
    dt <- inj_list_cross[[mix]]
    dt$rt_group <- group_l_cross[[mix]]
    dt$cluster <- ave(dt[[feat]], dt$rt_group, FUN = cluster_feature)
    dt
  })
  
  eval_subset <- combined[combined$is_annotated & !is.na(combined$cluster), ]
  
  data.frame(
    feature = feat,
    n_evaluated = nrow(eval_subset),
    ARI_cross = adjustedRandIndex(eval_subset$cross_lab_group, eval_subset$cluster)
  )
}) |>
  arrange(-ARI_cross)

# Compare within-lab vs cross-lab ARI
ari_comparison <- ari_results |>
  select(feature, ARI_within = ARI) |>
  left_join(ari_cross |> select(feature, ARI_cross), by = "feature")

print(ari_comparison)

ari_comparison |>
  pivot_longer(cols = c(ARI_within, ARI_cross), names_to = "type", values_to = "ARI") |>
  ggplot(aes(x = reorder(feature, ARI), y = ARI, fill = type)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Clustering Performance: Within-Lab vs Cross-Lab",
       x = "", y = "ARI", fill = "")
```

## 5. Summary

```{r}
best_feature <- ari_results$feature[1]
best_ari <- ari_results$ARI[1]
best_cross_feature <- ari_cross$feature[1]
best_cross_ari <- ari_cross$ARI_cross[1]

cat("=== ANALYSIS APPROACH ===\n\n")
cat("Clustering was performed on ALL detected peaks.\n")
cat("Evaluation metrics (ARI, AUC) use only the annotated subset as ground truth.\n")
cat("Cross-lab analysis pools data across labs (same column/mobile phases).\n\n")

cat("=== WITHIN-LAB CLUSTERING (ARI) ===\n\n")
for (i in seq_len(nrow(ari_results))) {
  cat(sprintf("  %s: %.3f\n", ari_results$feature[i], ari_results$ARI[i]))
}
cat("\nBest:", best_feature, "(ARI =", round(best_ari, 3), ")\n")
if (best_ari < 0.1) cat("  >> POOR\n")
cat("\n")

cat("=== CROSS-LAB CLUSTERING (ARI) ===\n\n")
for (i in seq_len(nrow(ari_cross))) {
  cat(sprintf("  %s: %.3f\n", ari_cross$feature[i], ari_cross$ARI_cross[i]))
}
cat("\nBest:", best_cross_feature, "(ARI =", round(best_cross_ari, 3), ")\n")
if (best_cross_ari > best_ari) {
  cat("  >> Cross-lab pooling IMPROVED clustering performance!\n")
} else {
  cat("  >> Cross-lab pooling did not improve performance.\n")
}
cat("\n")

cat("=== EIC CORRELATION (Pearson) ===\n\n")
cat("AUC:", round(eic_auc, 3), "\n")
cat("Optimal threshold:", round(best_thresh$threshold, 2), "\n")
if (eic_auc < 0.7) {
  cat("  >> WEAK: AUC < 0.7\n")
} else {
  cat("  >> ACCEPTABLE: AUC >= 0.7\n")
}
cat("\n")

cat("=== INTERPRETATION ===\n\n")
if (best_cross_ari > best_ari) {
  cat("Cross-lab analysis shows promise - peak shapes are consistent across labs.\n")
  cat("The shared chromatography (column + mobile phase) drives peak shape.\n")
} else {
  cat("Neither within-lab nor cross-lab clustering performs well.\n")
}
cat("\nPossible issues:\n")
cat("  1. RT grouping window (10s) may be too wide\n")
cat("  2. Peak shape metrics may be noisy\n")
cat("  3. Adducts may genuinely differ in peak shape\n\n")

cat("=== NEXT STEPS ===\n\n")
cat("1. Try tighter RT windows (5s, 3s)\n")
cat("2. Examine specific compounds visually\n")
cat("3. Use m/z relationships as primary grouping criterion\n")
cat("4. Use peak shape as a validation filter, not primary grouping\n")
``` 
