---
title: "Downstream Analysis: Adduct Grouping & Peak Shape Similarity"
format: html
editor: source
toc: true
---

## 1. Setup and Data Loading

We begin by loading the annotated peak data and the raw EICs. The goal is to
establish methods to group peaks derived from the same compound
(adducts, in-source fragments) using both scalar features and raw data shapes.

In the futur i want to load the full detected eics (whihc will be names eics_detected)
an the annotated peakonly eics will be named  eics_ann. for this is need to run
teh new object creation code ont he cluster but it's not possible at the moment
therefore i will just set up the code as much as I can and correct it when i have the
updated objects.

essentially: 
- we have ground truth on the annotated peaks
- we want to test different methods to group peaks based on shape similarity
- we use all detected peaks for the unsupervised grouping as it is more realistic

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(xcms) # if needed for raw data handling
#library(corrr) # useful for correlations
library(MsCoreUtils) ## rt grouping
library(MsFeatures)

base_path <- file.path("5_downstream_analysis")

# Load data
# Ensure these objects are available in your environment
load(file = file.path(base_path, "object/full_peak_report.RData"))
load(file = file.path(base_path, "object/final_eics.RData"))

# Convert to data.frame
df_peaks <- as.data.frame(full_peak_report)
```

## 2. Peak Shape Similarity based on Features (Scalar Metrics)

**Hypothesis:** Adducts of the same compound elute at the same time and share physical peak shape properties (FWHM, Tailing, Asymmetry) driven by the column interaction.

### 2.1. Define Ground Truth (For Validation)

Since we are developing the method on *annotated* data first, we define what a 
"correct group" looks like.

```{r}
df_peaks$true_group_id <- paste0(df_peaks$compound_name, "_", df_peaks$lab, "_", df_peaks$mixture)
```

### 2.3. Variation of each features

We iterate through specific metrics (`fwhm`, `tailing_factor`, `asymmetry`) to 
see which one has the lowest CV within `true_group_id`.

```{r}
features_to_test <- c("fwhm", "tailing_factor", "gaussian_similarity",
 "rt_width", "entropy")


# Function to evaluate a single feature
evaluate_feature <- function(data, feature_name) {
  data %>%
    filter(!is.na(.data[[feature_name]])) %>%
    group_by(true_group_id) %>%
    summarise(
      n_peaks = n(),
      # Calculate variability of the feature within the TRUE group
      feature_cv = sd(.data[[feature_name]]) / mean(.data[[feature_name]]) * 100,
      .groups = "drop"
    ) %>%
    filter(n_peaks > 1) %>%
    summarise(
      avg_cv = mean(feature_cv, na.rm = TRUE),
      median_cv = median(feature_cv, na.rm = TRUE)
    ) %>%
    mutate(feature = feature_name)
}

# Run evaluation
results_features <- map_dfr(features_to_test, ~evaluate_feature(df_peaks, .x))
results_features <- as.data.frame(results_features)
print(results_features)
```

### 2.4 Unsupervised grouping 

Here i want to try to cluster within the RT groups using each feature 
individually to see how well they separate the true groups. This is run 
using ALL detected peaks, not just annotated ones.

Pre-clustering by Retention Time
peaks must co-elute. We create a "Candidate Group" ID based on RT windows.
not sure how strici to be here.

```{r}
# Split by Lab/Mixture to avoid cross-hardware RT shifts
df_peaks$lab_mixture <- paste(df_peaks$lab, df_peaks$mixture, sep = "_")
inj_list <- split(df_peaks, df_peaks$lab_mixture)

group_l <- lapply(inj_list, function(dt) {
  param <- SimilarRtimeParam(diffRt = 10, groupFun = MsCoreUtils::group)
  rt_groups <- groupFeatures(dt$rt_apex, param)
  return(rt_groups)
})
```


below we loop through the group_l list to cluster within each RT group using each feature
and we need to calcualte the clustering quality by comparing to true_group_id

use ARI ? 

```{r}
library(mclust) # for ARI calculation
# Example for one feature
feature_name <- "fwhm" # Example feature
clustering_results <- lapply(names(inj_list), function(lab_mix) {
  dt <- inj_list[[lab_mix]]
  dt$rt_group_id <- group_l[[lab_mix]]
  
  # For each RT group, cluster using the feature
  # Define clustering function for ave()
  clustering_func <- function(feature_values) {
    if (length(unique(feature_values)) > 1 && !any(is.na(feature_values))) {
      dist_matrix <- dist(feature_values)
      hc <- hclust(dist_matrix)
      cutree(hc, k = 2) # Example: split into 2 clusters, it could be user defined. or not specified. 
    } else {
      rep(1, length(feature_values))
    }
  }
  
  # Apply clustering per group using base R ave()
  dt$cluster_id <- ave(dt[[feature_name]], dt$rt_group_id, FUN = clustering_func)
  
  return(dt)
})

```

Below we run that on all groups and using all features and calculate ARI for each feature

```{r}
feature_clustering_results <- lapply(features_to_test, function(feature_name) {
  clustering_results <- lapply(names(inj_list), function(lab_mix) {
    dt <- inj_list[[lab_mix]]
    dt$rt_group_id <- group_l[[lab_mix]]
    
    # For each RT group, cluster using the feature
    clustering_func <- function(feature_values) {
      if (length(unique(feature_values)) > 1 && !any(is.na(feature_values))) {
        dist_matrix <- dist(feature_values)
        hc <- hclust(dist_matrix)
        cutree(hc, k = 2) # Example: split into 2 clusters
      } else {
        rep(1, length(feature_values))
      }
    }
    
    dt$cluster_id <- ave(dt[[feature_name]], dt$rt_group_id, FUN = clustering_func)
    
    return(dt)
  })
  
  # Combine results and calculate ARI
  combined_dt <- do.call(rbind, clustering_results)
  ari_value <- adjustedRandIndex(combined_dt$true_group_id, combined_dt$cluster_id)
  
  return(data.frame(feature = feature_name, ARI = ari_value))
})
```

*ARI score* :  explain 

Should we do random forest instead ? how, and what could that mean?

### 2.4. Combined Feature Model

Test all possible group of features (>=2) to cluster within the RT windows. 
and see if the y improve the clustering quality.

test every possible combination of features (2 to all) 

```{r}
library(combinat)
feature_combinations <- unlist(lapply(2:length(features_to_test), 
                                      function(x) combn(features_to_test, x, simplify = FALSE)), 
                               recursive = FALSE)

combination_clustering_results <- lapply(feature_combinations, function(feature_set) {
  clustering_results <- lapply(names(inj_list), function(lab_mix) {
    dt <- inj_list[[lab_mix]]
    dt$rt_group_id <- group_l[[lab_mix]]
    
    # For each RT group, cluster using the combined features
    clustering_func <- function(...) {
      feature_values <- as.data.frame(list(...))
      if (nrow(unique(feature_values)) > 1 && !any(is.na(feature_values))) {
        dist_matrix <- dist(feature_values)
        hc <- hclust(dist_matrix)
        cutree(hc, k = 2) # Example: split into 2 clusters
      } else {
        rep(1, nrow(feature_values))
      }
    }
    
    dt$cluster_id <- do.call(ave, c(dt[feature_set], list(FUN = clustering_func)))
    
    return(dt)
  })

  # Combine results and calculate ARI
  combined_dt <- do.call(rbind, clustering_results)
  ari_value <- adjustedRandIndex(combined_dt$true_group_id, combined_dt$cluster_id)
  return(data.frame(features = paste(feature_set, collapse = "+"), ARI = ari_value))
})
combination_clustering_results_df <- do.call(rbind, combination_clustering_results)


```

## 3. Peak Shape Similarity based on Raw Data (EICs)

**Hypothesis:** EICs of adducts are perfectly correlated 
(Pearson r \> 0.9) because they are slices of the same chromatographic band.

### 3.1. Data Preparation (Interpolation)

Raw data points are often not aligned on the exact same scan times. 
We must interpolate EICs in a candidate group to a common time axis.

get intensities 

```{r}
ints <- intensity(final_eics)


```

### 3.2. Calculate Similarity (Pearson & Cosine)

We perform this analysis per Lab/Mix to avoid hardware offsets.

```{r}
# Wrapper to apply correlation analysis
calculate_eic_correlation <- function(eic_list) {
  # 1. Identify common RT range
  # 2. Bin/Interpolate intensities
  # 3. cor(matrix, method = "pearson")
}

# Apply to a subset for testing (e.g., one specific lab/mix)
# test_group <- df_peaks |> filter(lab == "SpecificLab", mixture == "Mix1", rt_group_id == X)
# eic_correlations <- calculate_eic_correlation(final_eics[test_group$chrom_peak_id])

```

### 3.3. Evaluation of Raw Data Similarity

We check if `true_group_id` members have higher correlations than non-members within the same RT window.

```{r}
# Compare "Within-Group" correlation vs "Between-Group" correlation
# Goal: Minimize overlap between these two distributions.

```

## 4. Cross-Lab Consistency (Downstream Application)

We do *not* group peaks across labs based on shape (as shapes vary by hardware). Instead, we assess if the **grouping logic** holds across labs.

-   **Question:** If Compound X has a "good" shape match with its adduct in Lab A, does it also have a "good" match in Lab B?
-   **Metric:** Compare the `feature_cv` (Coefficient of Variation) for specific compounds across labs.

```{r}
# Example: Is FWHM variability for [M+H] vs [M+Na] consistent across labs?
cross_lab_check <- df_peaks %>%
  group_by(lab, compound_name) %>%
  summarise(
    fwhm_cv = sd(fwhm)/mean(fwhm),
    .groups = "drop"
  ) %>%
  as.data.frame()

ggplot(cross_lab_check, aes(x = lab, y = fwhm_cv)) +
  geom_boxplot() +
  ggtitle("Consistency of Adduct FWHM Similarity Across Labs")

```
