---
title: "Adduct Grouping Method Development"
format:
  html:
    toc: true
    code-fold: true
execute:
  dir: project
---

# Introduction

This document explores unsupervised methods to group chromatographic peaks
that originate from the same compound (adducts, in-source fragments). We
start with a single lab (afekta) and a file-by-file approach.

**Strategy:**

1. Load all detected peaks for afekta (per file/sample)
2. Group peaks by RT (co-eluting peaks likely originate from the same compound)
3. Within each RT group, compute m/z differences and check against known adduct
   mass differences
4. Compute peak shape metrics and test whether they can discriminate
   same-compound vs different-compound peak pairs
5. Compute peak shape similarity (Pearson correlation on raw EIC data)
6. Evaluate all approaches against the annotated ground truth

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(xcms)
library(MsCoreUtils)
library(MetaboCoreUtils)
library(MsFeatures)
library(Chromatograms)
library(mclust)
library(pheatmap)
library(ggplot2)
library(MsQuality)

base_path <- here::here("5_downstream_analysis")
source(file.path(base_path, "rt_analysis_function.R"))
source(here::here("1_preprocessing/setup.R"))
```

# 1. Load Data

## 1.1 Detected Peaks (afekta)

```{r load-detected}
pt_afekta <- read.csv(
  file.path(base_path, "object/detected_peaks_afekta_HE.csv")
)

cat("Total detected peaks (afekta):", nrow(pt_afekta), "\n")
cat("Columns:", paste(colnames(pt_afekta), collapse = ", "), "\n")
cat("Samples:", length(unique(pt_afekta$sample)), "\n")
cat("Mixtures:", paste(sort(unique(pt_afekta$mixture)), collapse = ", "), "\n")
```

## 1.2 Annotations (ground truth)

```{r load-annotations}
res_afekta <- read.csv(here::here(
  "4_library_generation", "afekta", "HE", "ring_trial_library_HE.csv"
))

cat("Annotated peaks (afekta):", nrow(res_afekta), "\n")
cat("Unique compounds:", length(unique(res_afekta$compound_name)), "\n")
cat("Adducts used:", paste(unique(res_afekta$adduct), collapse = ", "), "\n")
```

## 1.3 Merge annotations into detected peaks

We match annotations to detected peaks by `chrom_peak_id` so every detected
peak knows whether it has been annotated and to which compound it belongs.

```{r merge-annotations}
ann_cols <- c("chrom_peak_id", "compound_name", "adduct", "formula",
              "InChIKey")

pt_afekta <- merge(
  pt_afekta,
  res_afekta[, ann_cols],
  by = "chrom_peak_id",
  all.x = TRUE
)

pt_afekta$is_annotated <- !is.na(pt_afekta$compound_name)

cat("Annotated peaks in detected table:",
    sum(pt_afekta$is_annotated), "of", nrow(pt_afekta),
    "(", round(100 * mean(pt_afekta$is_annotated), 1), "%)\n")
```

# 2. RT Grouping — Anchored on Annotated Compounds

**Why not `SimilarRtimeParam` / `MsCoreUtils::group()`?** These use
neighbor-chaining — with dense peak data (~10k peaks across 700s, max
consecutive gap < 2s), everything chains into a single group. Fixed-width
binning (`cut()`) works but creates arbitrary boundaries.

**Our approach:** Anchor groups on annotated compounds. For each annotated
compound, we collect all detected peaks within ±`diffRt_sec` of its RT.
This:

- Prevents chaining (each group has a fixed center)
- Guarantees every group has ground truth
- Produces biologically meaningful groups (centered on real compounds)

A detected peak can appear in multiple compound groups if it falls within
the RT window of several annotated peaks — this is expected and correct
(it could be an adduct of any of them).

```{r rt-grouping}
diffRt_sec <- 2 # seconds — half-width of the RT window around each compound

# Build compound-level anchors: one row per compound × sample × polarity
# If a compound has multiple annotated peaks (adducts), we use the median RT
ann_peaks <- pt_afekta[pt_afekta$is_annotated, ]
anchor_keys <- paste(ann_peaks$compound_name, ann_peaks$sample,
                     ann_peaks$polarity, ann_peaks$mixture, sep = "|")

anchors <- do.call(rbind, lapply(split(ann_peaks, anchor_keys), function(x) {
  data.frame(
    compound_name = x$compound_name[1],
    sample        = x$sample[1],
    polarity      = x$polarity[1],
    mixture       = x$mixture[1],
    anchor_rt     = median(x$rt),
    n_adducts     = nrow(x),
    adducts       = paste(unique(x$adduct), collapse = "; "),
    stringsAsFactors = FALSE
  )
}))
rownames(anchors) <- NULL
anchors$group_id <- paste0("s", anchors$sample, "_", anchors$compound_name)

cat("Compound anchors:", nrow(anchors), "\n")
cat("Unique compounds:", length(unique(anchors$compound_name)), "\n")
```

```{r assign-groups}
# For each anchor, pull in all detected peaks within the RT window
pt_grouped <- do.call(rbind, lapply(seq_len(nrow(anchors)), function(i) {
  a <- anchors[i, ]
  idx <- pt_afekta$sample == a$sample &
    pt_afekta$polarity == a$polarity &
    abs(pt_afekta$rt - a$anchor_rt) <= diffRt_sec

  if (sum(idx) == 0L) return(NULL)

  out <- pt_afekta[idx, ]
  out$group_id        <- a$group_id
  out$anchor_compound <- a$compound_name
  out$anchor_rt       <- a$anchor_rt
  out$rt_offset       <- out$rt - a$anchor_rt
  out
}))

cat("Total peak-group assignments:", nrow(pt_grouped), "\n")
cat("Unique groups:", length(unique(pt_grouped$group_id)), "\n")
cat("Unique peaks assigned:", length(unique(pt_grouped$chrom_peak_id)), "\n")

# How many peaks per group?
group_sizes <- as.data.frame(table(pt_grouped$group_id),
                             stringsAsFactors = FALSE)
colnames(group_sizes) <- c("group_id", "n_peaks")

cat("\nGroup size summary:\n")
summary(group_sizes$n_peaks)

ggplot(group_sizes, aes(x = n_peaks)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Peaks per Compound-Anchored RT Group (afekta)",
    subtitle = paste("RT window = ±", diffRt_sec, "seconds"),
    x = "Number of peaks per group",
    y = "Count"
  ) +
  theme_minimal()
```

## 2.1 Ground truth

For each group, the anchor compound defines the "true" group. Any annotated
peak in the group that matches the anchor compound is a true positive.
Annotated peaks from *different* compounds are known negatives (co-eluting
but different).

```{r ground-truth}
pt_grouped$is_same_compound <- pt_grouped$is_annotated &
  pt_grouped$compound_name == pt_grouped$anchor_compound

cat("Peaks matching anchor compound:", sum(pt_grouped$is_same_compound), "\n")
cat("Annotated peaks from different compounds:",
    sum(pt_grouped$is_annotated & !pt_grouped$is_same_compound), "\n")
cat("Unannotated peaks:", sum(!pt_grouped$is_annotated), "\n")

# How many groups have co-eluting different compounds?
ann_grouped <- pt_grouped[pt_grouped$is_annotated, ]
coeluters <- do.call(rbind, lapply(
  split(ann_grouped, ann_grouped$group_id), function(x) {
    data.frame(
      group_id    = x$group_id[1],
      n_compounds = length(unique(x$compound_name)),
      compounds   = paste(unique(x$compound_name), collapse = "; "),
      stringsAsFactors = FALSE
    )
  }
))
rownames(coeluters) <- NULL

cat("\nGroups with co-eluting different compounds:",
    sum(coeluters$n_compounds > 1), "of", nrow(coeluters), "\n")

ggplot(coeluters, aes(x = n_compounds)) +
  geom_bar(fill = "darkgreen") +
  labs(
    title = "Annotated Compounds per Group",
    subtitle = "Groups anchored on annotated compounds",
    x = "Number of distinct annotated compounds",
    y = "Count"
  ) +
  theme_minimal()
```

# 3. Targeted Adduct Matching

we compute the
*expected m/z* for every plausible adduct and search the group for matches.

We use `MetaboCoreUtils::adducts()` for the full adduct table, filtered to
singly-charged species (multiply-charged ions are rare for small molecules
in human metabolomics). We cast a wide net here — false positives from
unlikely adducts will be filtered downstream by EIC correlation.

**Important:** We only search among *unannotated* peaks. Peaks that already
have a confirmed annotation are not re-searched — they already have an
identity. We also skip target adducts that match the compound's
already-known adduct types. The goal is to discover *new* adduct
assignments for peaks that are currently unannotated.

## 3.1 Build expected adduct m/z for each anchor

```{r expected-adducts}
# Get all singly-charged adducts from MetaboCoreUtils (both polarities)
all_adducts_tbl <- rbind(adducts("positive"), adducts("negative"))
all_adducts_tbl <- all_adducts_tbl[abs(all_adducts_tbl$charge) == 1, ]

pos_adduct_names <- all_adducts_tbl$name[all_adducts_tbl$positive]
neg_adduct_names <- all_adducts_tbl$name[!all_adducts_tbl$positive]

# Remove dimer/trimer adducts ([2M…], [3M…]) — unlikely in standard LC-MS
pos_adduct_names <- pos_adduct_names[!grepl("^\\[2M|^\\[3M", pos_adduct_names)]
neg_adduct_names <- neg_adduct_names[!grepl("^\\[2M|^\\[3M", neg_adduct_names)]

# Remove adducts requiring reagents/solvents not in standard RP-LC-MS
# (water/ACN/formic acid mobile phase, aqueous sample matrix)
implausible <- c(
  "[M+Li]+",         # lithium not in mobile phase

"[M+2Li-H]+",      # lithium not in mobile phase
  "[M+Br]-",         # bromine not in solvents
  "[M+C2F3O2]-",     # TFA — not used as modifier here
  "[M+C2H6OS+H]+",   # DMSO — not used as solvent
  "[M+C4H6N2+H]+",   # 1-methylimidazole — exotic
  "[M+C3H8O+H]+"     # isopropanol — not in mobile phase
)
pos_adduct_names <- setdiff(pos_adduct_names, implausible)
neg_adduct_names <- setdiff(neg_adduct_names, implausible)

# Remove adducts already used during annotation — those matches are known
annotated_adducts <- unique(res_afekta$adduct)
pos_adduct_names <- setdiff(pos_adduct_names, annotated_adducts)
neg_adduct_names <- setdiff(neg_adduct_names, annotated_adducts)

cat("Charge-1 adducts available:\n")
cat("  Positive:", length(pos_adduct_names), "-",
    paste(pos_adduct_names, collapse = ", "), "\n")
cat("  Negative:", length(neg_adduct_names), "-",
    paste(neg_adduct_names, collapse = ", "), "\n")

# Compute exact neutral mass directly from molecular formula
compound_mass <- unique(res_afekta[, c("compound_name", "formula")])
compound_mass$neutral_mass <- calculateMass(compound_mass$formula)

anchor_expected <- merge(
  anchors[, c("compound_name", "sample", "polarity", "group_id",
              "anchor_rt", "adducts")],
  compound_mass,
  by = "compound_name"
)
colnames(anchor_expected)[colnames(anchor_expected) == "adducts"] <- "known_adducts"
rownames(anchor_expected) <- NULL

cat("\nAnchors with neutral mass:", nrow(anchor_expected), "\n")
cat("Unique compounds:", length(unique(anchor_expected$compound_name)), "\n")
```

## 3.2 Search for matching peaks

For each anchor × target adduct, find co-eluting *unannotated* peaks whose
m/z matches the expected value within tolerance. Already-annotated peaks are
excluded from the search — they have a confirmed identity.

We use `MetaboAnnotation::matchValues()` with `Mass2MzParam` to perform
the m/z matching. This converts neutral masses to expected adduct m/z
values and matches them against observed peak m/z using a ppm tolerance.

```{r adduct-search}
library(MetaboAnnotation)

# Split pt_grouped by group_id for fast lookup
grp_split <- split(pt_grouped, pt_grouped$group_id)

ppm_tol <- 5
param_pos <- Mass2MzParam(adducts = pos_adduct_names, tolerance = 0,
 ppm = ppm_tol)
param_neg <- Mass2MzParam(adducts = neg_adduct_names, tolerance = 0,
 ppm = ppm_tol)

# For each group: match unannotated peaks against anchor's neutral mass
adduct_hits <- do.call(rbind, lapply(
  unique(anchor_expected$group_id), function(gid) {
    grp <- grp_split[[gid]]
    if (is.null(grp)) return(NULL)

    # Only consider unannotated peaks as candidates
    grp_unann <- grp[!grp$is_annotated, ]
    if (nrow(grp_unann) == 0L) return(NULL)
    ae <- anchor_expected[anchor_expected$group_id == gid, ]
    if (nrow(ae) == 0L || is.na(ae$neutral_mass)) return(NULL)
    target_df <- data.frame(exactmass = ae$neutral_mass)
    query_df <- data.frame(mz = grp_unann$mz)
    param <- if (ae$polarity == "pos") param_pos else param_neg

    res <- matchValues(query_df, target_df, param)
    md <- matchedData(res)

    # Keep only matched rows (non-NA adduct)
    matched_rows <- which(!is.na(md$adduct))
    if (length(matched_rows) == 0L) return(NULL)
    md <- md[matched_rows, ]

    # Map back to original peak rows using query index
    out <- grp_unann[matched_rows, ]
    out$matched_adduct <- as.character(md$adduct)
    out$mz_error       <- as.numeric(md$score)
    out$ppm_error      <- as.numeric(md$ppm_error)
    out$neutral_mass   <- ae$neutral_mass
    out
  }
))
rownames(adduct_hits) <- NULL

# All hits are unannotated peaks → status is "maybe" (needs validation)
adduct_hits$adduct_status <- "maybe"

cat("Total adduct hits (all 'maybe'):", nrow(adduct_hits), "\n")
cat("Unique groups with hits:", length(unique(adduct_hits$group_id)), "\n")
cat("Unique peaks matched:", length(unique(adduct_hits$chrom_peak_id)), "\n")

# Summary by adduct type
adduct_freq <- as.data.frame(table(adduct_hits$matched_adduct),
                             stringsAsFactors = FALSE)
colnames(adduct_freq) <- c("matched_adduct", "n")
adduct_freq <- adduct_freq[order(-adduct_freq$n), ]
knitr::kable(adduct_freq, caption = "Hits per adduct type", row.names = FALSE)
```

## 3.3 Hit ambiguity

Since we only searched unannotated peaks, all hits are "maybe" by
definition. We check whether the same peak was matched to multiple
anchors (indicating ambiguity that downstream validation must resolve).

```{r adduct-eval}
# How many peaks are matched to multiple groups/compounds?
peak_group_counts <- as.data.frame(
  table(adduct_hits$chrom_peak_id), stringsAsFactors = FALSE
)
colnames(peak_group_counts) <- c("chrom_peak_id", "n_assignments")

# Maybe hits per group
maybe_per_group <- as.data.frame(
  table(adduct_hits$group_id), stringsAsFactors = FALSE
)
colnames(maybe_per_group) <- c("group_id", "n_maybe")
```

I get reasonable numbers.

```{r candidate-distribution}
ggplot(maybe_per_group, aes(x = n_maybe)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "'Maybe' Adduct Hits per Compound Group",
    subtitle = "Unannotated peaks matching expected adduct m/z — need validation",
    x = "Number of 'maybe' adducts",
    y = "Number of groups"
  ) +
  theme_minimal()
```

## 3.4 m/z error distribution

```{r mz-error-dist, fig.width=10, fig.height=4}
library(patchwork)

p1 <- ggplot(adduct_hits, aes(x = mz_error * 1000)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Absolute m/z Error", x = "m/z Error (mDa)", y = "Count") +
  theme_minimal()

p2 <- ggplot(adduct_hits, aes(x = ppm_error)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(-ppm_tol, ppm_tol), linetype = "dashed",
             color = "red") +
  labs(title = "Relative m/z Error", x = "ppm Error", y = "Count") +
  theme_minimal()

p1 + p2 + plot_annotation(
  title = "m/z Error Distribution for Adduct Matches",
  subtitle = paste("Tolerance:", ppm_tol, "ppm — all hits are 'maybe'")
)
```

# 4. Peak Metrics for Adduct Grouping

Peak metrics (FWHM, peak width, tailing, etc.) are computed for **all**
peaks in the RT-anchored groups. This serves two purposes:

1. Filter low-quality adduct candidates (section 4.4)
2. Test whether metrics can discriminate same-compound vs different-compound
   peaks via clustering (section 5)

## 4.1 Extract EICs for all grouped peaks

We extract EICs for all peaks assigned to anchor groups — this includes
annotated peaks, adduct candidates, and all other co-eluting peaks.

```{r extract-eics}
annotated_in_groups <- pt_grouped[pt_grouped$is_annotated, ]

# All unique peaks across all anchor groups
all_peaks_needed <- unique(pt_grouped$chrom_peak_id)

# Build peak table for EIC extraction
pk_rows <- pt_afekta[pt_afekta$chrom_peak_id %in% all_peaks_needed, ]
pk_rows <- pk_rows[!duplicated(pk_rows$chrom_peak_id), ]

pk_table <- data.frame(
  chrom_peak_id = pk_rows$chrom_peak_id,
  mzMin    = pk_rows$mzmin - 0.003,
  mzMax    = pk_rows$mzmax + 0.003,
  rtMin    = pk_rows$rtmin - 3,
  rtMax    = pk_rows$rtmax + 3,
  polarity = ifelse(pk_rows$polarity == "pos", 1L, 0L),
  mixture  = pk_rows$mixture,
  sample   = pk_rows$sample,
  stringsAsFactors = FALSE
)

eics_file <- file.path(base_path, "object/eics_afekta_grouped.RData")

if (!file.exists(eics_file)) {
  load(file.path(base_path, "object/tic_full_spectra.RData"))
  tic_afekta <- tic_full[tic_full$lab == "afekta"]

  eics_afekta <- chromExtract(
    tic_afekta,
    peak.table = pk_table,
    by = c("polarity", "mixture")
  )
  eics_afekta <- setBackend(eics_afekta, ChromBackendMemory())
  save(eics_afekta, file = eics_file)
  rm(tic_full, tic_afekta)
} else {
  load(eics_file)
}

cat("Afekta EICs extracted:", length(eics_afekta), "\n")
```

## 4.2 Compute peak metrics

```{r compute-metrics}
metrics <- c(
  "peakCount",
  "xicFwhm",
  "areaUnderTic",
  "maxIntensity",
  "baselineIntensity",
  "signalToNoiseRatio",
  "peakWidth",
  "peakBoundary",
  "peakProminence",
  "peakBeta"
)

pk_metrics <- calculateMetrics(
  eics_afekta,
  metrics = metrics,
  na.rm = TRUE,
  filterEmptyObject = FALSE,
  f = seq_len(length(eics_afekta)),
  probs = 0.05,
  method = "adaptive",
  minPeakWidth = 3
)

cd_afekta <- chromData(eics_afekta)
pk_metrics <- cbind(
  cd_afekta[, intersect(
    c("chrom_peak_id", "mixture", "polarity", "mz", "sample"),
    colnames(cd_afekta)
  )],
  pk_metrics
)

cat("Metrics computed for", nrow(pk_metrics), "peaks\n")
```

## 4.3 Merge metrics with adduct hits

```{r merge-metrics}
metric_cols <- c("chrom_peak_id",
                 "peakCount", "xicFwhm", "areaUnderTic", "maxIntensity",
                 "baselineIntensity", "signalToNoiseRatio", "peakWidth",
                 "peakBoundary1", "peakBoundary2", "peakProminence",
                 "peakBeta1", "peakBeta2")
metric_cols <- intersect(metric_cols, colnames(pk_metrics))

adduct_hits <- merge(adduct_hits, pk_metrics[, metric_cols],
                     by = "chrom_peak_id", all.x = TRUE)

cat("Adduct hits with metrics:",
    sum(!is.na(adduct_hits$xicFwhm)), "of", nrow(adduct_hits), "\n")
```

## 4.4 Filter bad peaks

Remove candidate adduct peaks with poor quality metrics — same criteria
as in the detected-peaks analysis (`3a_peak_feature_calculation_detected`).

```{r filter-bad-peaks}
n_before <- nrow(adduct_hits)

bad_peaks <- is.na(adduct_hits$xicFwhm) |
  (!is.na(adduct_hits$peakProminence) & adduct_hits$peakProminence < 1.5) |
  (!is.na(adduct_hits$peakWidth) & adduct_hits$peakWidth < 1) |
  (!is.na(adduct_hits$peakWidth) & adduct_hits$peakWidth > 30)

adduct_hits <- adduct_hits[!bad_peaks, ]
rownames(adduct_hits) <- NULL

cat("Removed", sum(bad_peaks), "peaks with poor quality metrics\n")
cat("Remaining adduct hits:", nrow(adduct_hits), "of", n_before, "\n")
cat("Unique groups:", length(unique(adduct_hits$group_id)), "\n")
cat("Unique peaks:", length(unique(adduct_hits$chrom_peak_id)), "\n")
```

# 5. Clustering Evaluation


```{r}
# ------------------------------------------------------------------
# 2. STEP 1: Anti-Chaining RT Clustering (Sample-by-Sample)
# ------------------------------------------------------------------
cat("\n--- Running Unsupervised RT Clustering (Per Sample) ---\n")

# Initialize an empty vector for our new RT cluster IDs
pt_clustering$rt_cluster <- NA
max_rt_width <- 4 # Maximum allowed seconds between any two peaks in a group

# Loop through each sample individually
for (samp in unique(pt_clustering$sample)) {

  # Get row indices for the current sample
  idx <- which(pt_clustering$sample == samp)

  # Skip if a sample is empty or has only 1 peak
  if (length(idx) <= 1) {
    if (length(idx) == 1) pt_clustering$rt_cluster[idx] <- paste0(samp, "_rt1")
    next
  }

  # Calculate distance matrix ONLY for this specific sample
  # A few thousand peaks per sample uses almost no memory
  rt_dist <- dist(pt_clustering$rt[idx])

  # Complete linkage: strictly prevents chaining.
  # No cluster will ever be wider than max_rt_width.
  rt_hc <- hclust(rt_dist, method = "complete")

  # Cut the tree
  samp_clusters <- cutree(rt_hc, h = max_rt_width)

  # Assign unique cluster IDs (e.g., "Sample1_rt1", "Sample1_rt2")
  pt_clustering$rt_cluster[idx] <- paste0(samp, "_rt", samp_clusters)
}

cat("Created", length(unique(pt_clustering$rt_cluster)), "RT clusters across all samples.\n")
```


```{r}
cat("\n--- Running Metric Sub-Clustering within RT Groups ---\n")

# Scale the metrics globally so big numbers (Area) don't crush small ones (FWHM)
scaled_metrics <- scale(pt_clustering[, metric_cols])

# Initialize the final cluster column
pt_clustering$final_cluster <- NA
cluster_counter <- 1

# Loop through each of the 420 RT buckets and split them up using peak shapes
for (rt_id in unique(pt_clustering$rt_cluster)) {

  idx <- which(pt_clustering$rt_cluster == rt_id)

  # If the RT bucket somehow only has 1 peak, it is its own cluster
  if (length(idx) == 1) {
    pt_clustering$final_cluster[idx] <- paste0("C", cluster_counter, "_1")
    cluster_counter <- cluster_counter + 1
    next
  }

  # Extract scaled fingerprints for these co-eluting peaks
  group_fingerprints <- scaled_metrics[idx, , drop = FALSE]

  # Calculate Euclidean distance between peak shapes
  metric_dist <- dist(group_fingerprints)

  # Cluster based on peak shape
  metric_hc <- hclust(metric_dist, method = "average")

  # Cut the tree based on peak shape similarity
  # (h = 1.5 is a starting guess for scaled data; lower means stricter shape matching)
  metric_threshold <- 1.5
  sub_clusters <- cutree(metric_hc, h = metric_threshold)

  # Assign the final refined compound IDs
  pt_clustering$final_cluster[idx] <- paste0("C", cluster_counter, "_", sub_clusters)
  cluster_counter <- cluster_counter + 1
}

cat("Created", length(unique(pt_clustering$final_cluster)), "final peak-shape resolved clusters!\n")
```



```{r}
cat("\n--- Evaluating Cluster Quality Against Ground Truth ---\n")

# Isolate only the peaks that have a confirmed annotation
ann_data <- pt_clustering[!is.na(pt_clustering$compound_name), ]

cat("Total annotated peaks to evaluate:", nrow(ann_data), "\n")

# ---------------------------------------------------------
# TEST 1: Cluster Purity (Are we mixing different compounds?)
# ---------------------------------------------------------
# Group by final_cluster and count unique compound names
cluster_purity <- aggregate(compound_name ~ final_cluster, data = ann_data,
                            FUN = function(x) length(unique(x)))
colnames(cluster_purity)[2] <- "n_compounds"

mixed_clusters <- sum(cluster_purity$n_compounds > 1)
pure_clusters <- sum(cluster_purity$n_compounds == 1)

cat("\n[1] CLUSTER PURITY:\n")
cat(" - Pure clusters (1 compound per cluster):", pure_clusters, "\n")
cat(" - Mixed clusters (>1 compound per cluster):", mixed_clusters, "\n")
if(nrow(cluster_purity) > 0) {
  cat(" - Purity Rate:", round((pure_clusters / nrow(cluster_purity)) * 100, 1), "%\n")
}

# ---------------------------------------------------------
# TEST 2: Compound Integrity (Are we splitting up adducts?)
# ---------------------------------------------------------
# To prevent mixing samples, create a unique ID for each compound-sample pair
ann_data$compound_sample <- paste(ann_data$compound_name, ann_data$sample, sep = "|")

# Group by compound_sample and count how many final_clusters they got split into
compound_integrity <- aggregate(final_cluster ~ compound_sample, data = ann_data,
                                FUN = function(x) length(unique(x)))
colnames(compound_integrity)[2] <- "n_clusters"

intact_compounds <- sum(compound_integrity$n_clusters == 1)
split_compounds <- sum(compound_integrity$n_clusters > 1)

cat("\n[2] COMPOUND INTEGRITY:\n")
cat(" - Intact compounds (all adducts stayed in 1 cluster):", intact_compounds, "\n")
cat(" - Split compounds (adducts scattered across multiple clusters):", split_compounds, "\n")
if(nrow(compound_integrity) > 0) {
  cat(" - Integrity Rate:", round((intact_compounds / nrow(compound_integrity)) * 100, 1), "%\n")
}
```


```{r}
cat("\n--- Optimizing Peak Shape Threshold (h) ---\n")

# Store the hclust objects so we don't have to recalculate distances every time
hc_list <- list()
idx_list <- list()

for (rt_id in unique(pt_clustering$rt_cluster)) {
  idx <- which(pt_clustering$rt_cluster == rt_id)
  if (length(idx) > 1) {
    # Extract fingerprints and compute hclust once per RT group
    group_fingerprints <- scaled_metrics[idx, , drop = FALSE]
    hc_list[[rt_id]] <- hclust(dist(group_fingerprints), method = "average")
    idx_list[[rt_id]] <- idx
  }
}

# Test different h values
test_h_vals <- c(0.5, 0.8, 1.0, 1.2, 1.5, 1.8, 2.0, 2.5)
results <- data.frame(h = numeric(), purity = numeric(), integrity = numeric())

for (h_val in test_h_vals) {

  temp_cluster <- rep(NA, nrow(pt_clustering))
  cluster_counter <- 1

  for (rt_id in unique(pt_clustering$rt_cluster)) {
    idx <- which(pt_clustering$rt_cluster == rt_id)
    if (length(idx) == 1) {
      temp_cluster[idx] <- paste0("C", cluster_counter, "_1")
    } else {
      sub_clusters <- cutree(hc_list[[rt_id]], h = h_val)
      temp_cluster[idx] <- paste0("C", cluster_counter, "_", sub_clusters)
    }
    cluster_counter <- cluster_counter + 1
  }

  # --- Evaluation Section ---
  temp_ann <- pt_clustering[!is.na(pt_clustering$compound_name), ]
  temp_ann$test_cluster <- temp_cluster[!is.na(pt_clustering$compound_name)]

  # FIX: Create the compound_sample column here!
  temp_ann$compound_sample <- paste(temp_ann$compound_name, temp_ann$sample, sep = "|")

  # 1. Purity (Are we mixing different compounds?)
  pur_agg <- aggregate(compound_name ~ test_cluster, data = temp_ann,
                       FUN = function(x) length(unique(x)))
  pur_rate <- (sum(pur_agg$compound_name == 1) / nrow(pur_agg)) * 100

  # 2. Integrity (Are we splitting up adducts?)
  int_agg <- aggregate(test_cluster ~ compound_sample, data = temp_ann,
                       FUN = function(x) length(unique(x)))
  int_rate <- (sum(int_agg$test_cluster == 1) / nrow(int_agg)) * 100

  # Store results
  results <- rbind(results, data.frame(h = h_val, purity = pur_rate, integrity = int_rate))
}

print(results)
```

we could plot this. the user could decide what they want to trade offf on. In
our case we would focus more on purity as splitting up adducts is less bad than
 mixing different compounds together.

Below we try to optimize our fingerprint using pure shape metrics
 (FWHM, peak width) and remove intensity-based metrics that may be more
 variable across adducts. We test several combinations of metrics to see
 which best discriminates same-compound vs different-compound peaks.


```{r}
cat("\n--- Testing Different Peak Metric Combinations ---\n")

# 1. Bring in all potential metrics, including the Beta (tailing/fronting) metrics
all_candidates <- c("xicFwhm", "peakWidth", "peakProminence", "signalToNoiseRatio",
                    "peakCount", "peakBeta1", "peakBeta2")
all_candidates <- intersect(all_candidates, colnames(pk_metrics))

# Merge into our base table and drop NAs so all tests are perfectly comparable
pt_test <- merge(pt_afekta, pk_metrics[, c("chrom_peak_id", all_candidates)],
                 by = "chrom_peak_id", all.x = TRUE)
pt_test <- pt_test[complete.cases(pt_test[, all_candidates]), ]

# We need to quickly rebuild the RT clusters for this freshly filtered data
pt_test <- pt_test[order(pt_test$rt), ]
pt_test$rt_cluster <- NA
for (samp in unique(pt_test$sample)) {
  idx <- which(pt_test$sample == samp)
  if (length(idx) <= 1) {
    if (length(idx) == 1) pt_test$rt_cluster[idx] <- paste0(samp, "_rt1")
    next
  }
  rt_hc <- hclust(dist(pt_test$rt[idx]), method = "complete")
  pt_test$rt_cluster[idx] <- paste0(samp, "_rt", cutree(rt_hc, h = 4))
}

# 2. Define the different "recipes" to test
 # 2. Define the updated "recipes" to test
recipes <- list(
  "Original" = c("xicFwhm", "peakWidth", "peakProminence", "signalToNoiseRatio", "peakCount"),
  "Intensity Heavy" = c("peakProminence", "signalToNoiseRatio"),
  "Pure Shape (Strict)" = c("xicFwhm", "peakWidth"),
  "Shape + Tailing" = c("xicFwhm", "peakWidth", "peakBeta1")
)

results_metrics <- data.frame(Recipe = character(), Purity = numeric(), Integrity = numeric(), stringsAsFactors = FALSE)

# Fixed strict threshold for the test
test_h <- 0.8

# 3. Loop through each recipe
for (recipe_name in names(recipes)) {

  current_metrics <- recipes[[recipe_name]]

  # Ensure the metrics actually exist in the data
  current_metrics <- intersect(current_metrics, colnames(pt_test))

  # Scale only the metrics in the current recipe
  scaled_m <- scale(pt_test[, current_metrics])

  temp_cluster <- rep(NA, nrow(pt_test))
  cluster_counter <- 1

  for (rt_id in unique(pt_test$rt_cluster)) {
    idx <- which(pt_test$rt_cluster == rt_id)
    if (length(idx) == 1) {
      temp_cluster[idx] <- paste0("C", cluster_counter, "_1")
    } else {
      group_fp <- scaled_m[idx, , drop = FALSE]
      metric_hc <- hclust(dist(group_fp), method = "average")
      sub_clusters <- cutree(metric_hc, h = test_h)
      temp_cluster[idx] <- paste0("C", cluster_counter, "_", sub_clusters)
    }
    cluster_counter <- cluster_counter + 1
  }

  # --- Evaluate ---
  temp_ann <- pt_test[!is.na(pt_test$compound_name), ]
  temp_ann$test_cluster <- temp_cluster[!is.na(pt_test$compound_name)]
  temp_ann$compound_sample <- paste(temp_ann$compound_name, temp_ann$sample, sep = "|")

  pur_agg <- aggregate(compound_name ~ test_cluster, data = temp_ann, FUN = function(x) length(unique(x)))
  pur_rate <- (sum(pur_agg$compound_name == 1) / nrow(pur_agg)) * 100

  int_agg <- aggregate(test_cluster ~ compound_sample, data = temp_ann, FUN = function(x) length(unique(x)))
  int_rate <- (sum(int_agg$test_cluster == 1) / nrow(int_agg)) * 100

  results_metrics <- rbind(results_metrics, data.frame(Recipe = recipe_name, Purity = pur_rate, Integrity = int_rate))
}

print(results_metrics)
```



in conclusoin:

Step 1: Complete-linkage RT clustering (per-sample) with h = 4 seconds.

Step 2: Average-linkage metric clustering using
c("xicFwhm", "peakWidth", "peakBeta1") with Z-score scaling and h = 0.8.




```{r}
library(mclust)

# Assuming temp_ann is your annotated subset from the winning Recipe 4 evaluation
# True Labels: The actual compound-sample identities
true_labels <- temp_ann$compound_sample

# Predicted Labels: The unsupervised clusters we generated
predicted_labels <- temp_ann$test_cluster

# Calculate Adjusted Rand Index
ari_score <- adjustedRandIndex(true_labels, predicted_labels)

cat("\n--- Formal Clustering Evaluation ---\n")
cat("Adjusted Rand Index (ARI):", round(ari_score, 3), "\n")
```

Gemini ha detto
Here is a perfectly formatted summary you can drop straight into your Quarto document. I wrote it to match the tone of your introduction, balancing the strict machine-learning theory with the practical reality of LC-MS metabolomics.

You can copy and paste this directly into a text section of your project:

Clustering Evaluation Strategy: ARI vs. Domain-Specific Metrics
To evaluate the performance of our unsupervised adduct clustering against the annotated ground truth, we utilize two complementary approaches: standard machine learning metrics and domain-specific chromatographic metrics.

1. Domain-Specific Metrics: Purity (Homogeneity) and Integrity (Completeness)
In untargeted LC-MS, clustering errors do not carry equal weight. We manually calculate two specific metrics to understand exactly how the algorithm fails:

Cluster Purity: The percentage of clusters that contain only a single annotated compound. Low purity means the algorithm is grouping different co-eluting compounds together. This is a "fatal" error in metabolomics because it mixes MS1/MS2 spectra, leading to false library annotations.

Compound Integrity: The percentage of true compounds whose adducts were perfectly kept together in a single cluster. Low integrity means the algorithm split a single compound's adducts into multiple clusters. While suboptimal, this is a "safe" error—it merely creates redundant features in the dataset without corrupting the underlying mass spectra.

Strategy: We actively optimize our distance thresholds (e.g., h=0.8) and metric fingerprints to maximize Purity (typically ~90%), even if it costs us some Integrity (typically capping around 80-85% due to the physical intensity differences between major and minor adducts).

2. Statistical Benchmark: Adjusted Rand Index (ARI)

While Purity and Integrity inform our method development, they are highly directional. To provide a rigorously standardized and peer-review-friendly benchmark, we also calculate the Adjusted Rand Index (ARI) using the mclust package.

ARI measures the overall similarity between our unsupervised clusters and the true compound groupings, mathematically adjusted for the baseline chance of random grouping.

While ARI is the gold standard in machine learning, it penalizes false positives (mixing compounds) and false negatives (splitting adducts) equally. Therefore, we expect a moderate ARI score (e.g., 0.65 - 0.80), which reflects our intentional bias toward over-clustering to preserve spectral purity.



```{r}
library(mclust)

cat("\n--- Corrected Formal Clustering Evaluation ---\n")

# 1. Count how many peaks each compound has in the ground truth
compound_counts <- table(temp_ann$compound_sample)

# 2. Identify compounds that actually have multiple adducts (pairs exist!)
multi_peak_compounds <- names(compound_counts[compound_counts > 1])

# 3. Filter the evaluation data to only those multi-peak compounds
eval_data <- temp_ann[temp_ann$compound_sample %in% multi_peak_compounds, ]

cat("Number of multi-peak compounds to evaluate:", length(unique(eval_data$compound_sample)), "\n")
cat("Total peaks in these compounds:", nrow(eval_data), "\n")

# 4. Recalculate Adjusted Rand Index on fair data
true_labels_filtered <- eval_data$compound_sample
predicted_labels_filtered <- eval_data$test_cluster

ari_score_corrected <- adjustedRandIndex(true_labels_filtered, predicted_labels_filtered)

cat("Corrected Adjusted Rand Index (ARI):", round(ari_score_corrected, 3), "\n")
```

is there really such a small amount of multipeak compounds ?


```{r}
cat("\n--- Formal Pairwise Evaluation (Precision / Recall / F1) ---\n")

# We use the eval_data (the 156 peaks from 78 multi-peak compounds)
# Generate all possible pairs of these 156 peaks
all_pairs <- combn(eval_data$chrom_peak_id, 2)

# Vectorized check: Are they in the same TRUE compound?
true_compound_1 <- eval_data$compound_sample[match(all_pairs[1,], eval_data$chrom_peak_id)]
true_compound_2 <- eval_data$compound_sample[match(all_pairs[2,], eval_data$chrom_peak_id)]
is_true_pair <- (true_compound_1 == true_compound_2)

# Vectorized check: Are they in the same PREDICTED cluster?
pred_cluster_1 <- eval_data$test_cluster[match(all_pairs[1,], eval_data$chrom_peak_id)]
pred_cluster_2 <- eval_data$test_cluster[match(all_pairs[2,], eval_data$chrom_peak_id)]
is_pred_pair <- (pred_cluster_1 == pred_cluster_2)

# Calculate confusion matrix components
true_positives <- sum(is_true_pair & is_pred_pair)
false_positives <- sum(!is_true_pair & is_pred_pair)
false_negatives <- sum(is_true_pair & !is_pred_pair)

# Calculate Metrics
precision <- true_positives / (true_positives + false_positives)
recall <- true_positives / (true_positives + false_negatives)
f1_score <- 2 * ((precision * recall) / (precision + recall))

cat("Total True Adduct Pairs: ", sum(is_true_pair), "\n")
cat("Total Predicted Pairs:   ", sum(is_pred_pair), "\n")
cat("True Positives (Caught): ", true_positives, "\n\n")

cat("Precision (Are predicted groups real?):", round(precision * 100, 1), "%\n")
cat("Recall (Did we catch all real pairs?): ", round(recall * 100, 1), "%\n")
cat("F1-Score (Overall Quality, 0 to 1):    ", round(f1_score, 3), "\n")

```
