---
title: "RT-local normalization — producer (Phase 3)"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
---

# Goal — producer

Estimate the per-`(lab, injection, RT-decile)` response profile from
the **Detected Signal** (xcms output: ~1M peaks across the 4 labs),
apply that response correction back to the **Annotated Signal**
(canonical `annotation_chrom_metrics.csv` from the SIRIUS producer),
write the normalized annotated table to disk for downstream analysis.

This file is the **producer**. The matching **consumer** is
`7_icc_reproducibility.qmd`, which reads the output and computes the ICC
ladders, pairwise comparisons, and per-lab diagnostics.

Pipeline:

1. Load detected peaks, apply the P2 alignment splines.
2. Build three factor maps: `factor_global` (per-injection global
   scale), `factor_decile` (per-injection × RT-decile), `factor_loess`
   (smooth f(aligned_rt) per injection).
3. Apply each factor to the annotated peaks; write the result.

Inputs:

- `5_downstream_analysis/object/detected_peaks_{lab}_HE.csv` — per-lab
  `(rtmin, rtmax, mzmin, mzmax, rt, mz, sample, into, intb, maxo,
  polarity, mixture, chrom_peak_id)`.
- `5_downstream_analysis/object/rt_alignment_splines.RData` — per
  `(lab, polarity)` splines from P2 (`4_rt_alignment.qmd`).
- `5_downstream_analysis/object/rt_alignment_anchors.csv` — anchor
  table (for clamping out-of-range peaks).
- `3_annotation_manual/HE/3_Sirius_curation/results_all_ms2/annotation_chrom_metrics.csv`
  — the canonical annotated peak set we normalize.

Outputs:

- `5_downstream_analysis/object/annotation_chrom_metrics_normalized.csv`
  — annotated rows + `aligned_rt, decile, factor_global,
  factor_decile, factor_loess, maxIntensity_{global,decile,loess},
  areaUnderTic_{global,decile,loess}`.
- `5_downstream_analysis/object/detected_peak_counts.csv` — small
  `(lab, polarity, sample, n_peaks)` summary used by the consumer's
  diagnostic.

## Setup

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

labs           <- c("afekta", "cembio", "hmgu", "icl")
producer_dir   <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                       "results_all_ms2")
object_dir     <- here("5_downstream_analysis", "object")

# Number of RT bins (deciles → 10). Computed per polarity on the
# pooled aligned_rt distribution so bins are cross-lab comparable.
n_rt_bins <- 10L

# Palette for strategy comparison.
# "Aligned" is added as a neutral grey to isolate the contribution of the
# alignment step alone (no intensity factor) from the normalization on top.
# "Lab-median" added per the §P3.5.b sanity check — it's 2_global's
# Annotated Normalized formulation (per-(lab, polarity) median of bulk),
# kept side by side with the per-injection-decile-based "Global" so the
# reader can see which target works better.
strategy_palette <- c(Raw          = "#E64B35",
                      Aligned      = "#999999",
                      `Lab-median` = "#3B7C92",
                      Global       = "#4DBBD5",
                      Decile       = "#70AD47",
                      LOESS        = "#4472C4")
```

# P3.1 Load detected peaks (per-lab, per-injection)

**Data level: Detected Signal** — only used here to estimate the
response profile (P3.3). It is *not* the level on which P3.5 ICC is
computed.

Read the four per-lab CSVs, add `lab`, concatenate. Each row is one
detected peak; the `(lab, sample, polarity)` triple identifies one
injection (`sample` indexes 1..19 within each polarity per lab).

```{r load-detected-peaks}
read_lab_peaks <- function(lab) {
  read.csv(file.path(object_dir, sprintf("detected_peaks_%s_HE.csv", lab)),
           stringsAsFactors = FALSE,
           colClasses = c(mixture = "character")) |>
    mutate(lab = lab, .before = 1L)
}

peaks <- bind_rows(lapply(labs, read_lab_peaks))

# Polarity column comes through as character ("pos" / "neg") — same as
# the chromExtract / annotation pipelines.
cat("Detected peaks loaded:\n")
peaks |> count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable(caption = "Peaks per (lab, polarity).")

cat("\nInjections (lab × polarity × sample) detected:\n")
peaks |> distinct(lab, polarity, sample) |>
  count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable(caption = "Distinct injections per (lab, polarity).")
```

# P3.2 Apply P2 alignment to detected peak `rt`

Load the per-`(lab, polarity)` Hyman splines from P2 and apply them to
each peak's `rt`, with the same anchor-range clamp used in
`4_rt_alignment.qmd`. Peaks outside the anchor range get
`aligned_rt = NA` and are dropped from the factor-fitting step (they'd
be extrapolated otherwise).

```{r apply-alignment}
load(file.path(object_dir, "rt_alignment_splines.RData"))   # `splines`
anchors_for_range <- read.csv(
  file.path(object_dir, "rt_alignment_anchors.csv"),
  stringsAsFactors = FALSE)

apply_splines <- function(df, splines_map, anchors_for_range, rt_col = "rt") {
  df$key        <- paste(df$lab, df$polarity, sep = "_")
  df$aligned_rt <- NA_real_
  for (key_i in unique(df$key)) {
    f <- splines_map[[key_i]]
    if (is.null(f)) next
    anchor_range <- anchors_for_range |>
      filter(paste(lab, polarity, sep = "_") == key_i) |>
      pull(rt_local) |> range(na.rm = TRUE)
    idx       <- which(df$key == key_i)
    rt_lab    <- df[[rt_col]][idx]
    in_range  <- !is.na(rt_lab) &
                 rt_lab >= anchor_range[1] &
                 rt_lab <= anchor_range[2]
    df$aligned_rt[idx[in_range]] <- as.numeric(f(rt_lab[in_range]))
  }
  df$key <- NULL
  df
}

peaks <- apply_splines(peaks, splines, anchors_for_range, rt_col = "rt")

cat("Peaks with aligned_rt (in-range):", sum(!is.na(peaks$aligned_rt)),
    " / ", nrow(peaks), "\n")
cat("Out-of-range (NA aligned_rt) per lab / polarity:\n")
peaks |> filter(is.na(aligned_rt)) |>
  count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n, values_fill = 0L) |>
  knitr::kable()
```

# P3.3 Compute normalization factors

We define **three factor maps** (Raw is implicit = 1 everywhere) that
each take a peak's `(lab, sample, polarity, aligned_rt)` and return a
multiplicative correction.

`Global` — per `(lab, polarity)`. Single number per lab × polarity =
median peak intensity / cross-lab consensus median intensity. Removes
overall sensitivity offset.

`Decile` — per `(lab, sample, polarity, decile)`. RT-binned median.
Removes both per-injection variability and RT-local sensitivity
biases.

`LOESS` — per `(lab, sample, polarity)`, smooth `f(aligned_rt)`. Same
spirit as `Decile` but continuous in RT — avoids bin-edge jumps.

We compute factors using `into` (peak area) as the response metric;
`maxo` is carried through unchanged for consistency with the existing
annotated-level bulk metric in `8_data_level_progression.qmd`.

```{r factor-decile-build}
# Per-polarity aligned_rt deciles. Computed on the pooled distribution
# across all labs / samples so the bin edges are comparable.
peak_rt <- peaks |> filter(!is.na(aligned_rt), is.finite(into), into > 0)

# Bin edges per polarity (named list).
decile_breaks <- lapply(split(peak_rt$aligned_rt, peak_rt$polarity),
                        quantile,
                        probs   = seq(0, 1, length.out = n_rt_bins + 1L),
                        na.rm   = TRUE)

# Vectorized decile assignment — cut() per polarity, indexed back to
# the parent table. ~1M peaks; rowwise() would be unusable.
peak_rt$decile <- NA_integer_
for (pol in names(decile_breaks)) {
  idx <- which(peak_rt$polarity == pol)
  peak_rt$decile[idx] <- as.integer(cut(peak_rt$aligned_rt[idx],
                                        breaks         = decile_breaks[[pol]],
                                        labels         = FALSE,
                                        include.lowest = TRUE))
}

# Per (lab, sample, polarity, decile) lab response
lab_decile_response <- peak_rt |>
  group_by(lab, sample, polarity, decile) |>
  summarise(lab_median_into = median(into, na.rm = TRUE),
            n_peaks         = n(),
            .groups = "drop")

# Cross-lab consensus per (sample, polarity, decile)
consensus_decile <- lab_decile_response |>
  group_by(sample, polarity, decile) |>
  summarise(consensus_into = median(lab_median_into, na.rm = TRUE),
            n_labs         = n_distinct(lab),
            .groups = "drop") |>
  filter(n_labs >= 2L)

# Decile factor = consensus / lab. Multiply lab values by factor to land on consensus.
factor_decile_tbl <- lab_decile_response |>
  inner_join(consensus_decile |> select(sample, polarity, decile, consensus_into),
             by = c("sample", "polarity", "decile")) |>
  mutate(factor_decile = consensus_into / lab_median_into)

cat("Decile factor table — rows per (lab, polarity):\n")
factor_decile_tbl |> count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable()
```

```{r factor-global-build}
# Global = collapse the decile factor by taking median across deciles
# per (lab, sample, polarity). This is the "one number per injection"
# version of the same correction.
factor_global_tbl <- factor_decile_tbl |>
  group_by(lab, sample, polarity) |>
  summarise(factor_global = median(factor_decile, na.rm = TRUE),
            .groups = "drop")
```

```{r factor-loess-build}
# LOESS — per (lab, sample, polarity), fit a smooth log10(factor) vs
# aligned_rt curve from the decile-midpoint factors. P3.4 evaluates
# this surface at each annotated peak's aligned_rt.
#
# Each polarity's decile midpoints are computed once from decile_breaks
# (list of break vectors, named by polarity).
decile_midpoints <- bind_rows(lapply(names(decile_breaks), function(pol) {
  br <- decile_breaks[[pol]]
  tibble(polarity   = pol,
         decile     = seq_len(n_rt_bins),
         aligned_rt = (br[-1] + br[-length(br)]) / 2)
}))

factor_decile_with_rt <- factor_decile_tbl |>
  left_join(decile_midpoints, by = c("polarity", "decile"))

fit_loess_one <- function(df) {
  if (nrow(df) < 5L) return(NULL)
  tryCatch(
    loess(log10(factor_decile) ~ aligned_rt,
          data       = df,
          span       = 0.75,
          degree     = 1,
          na.action  = na.exclude),
    error = function(e) NULL)
}

loess_fits <- factor_decile_with_rt |>
  group_by(lab, sample, polarity) |>
  group_modify(~ tibble(fit = list(fit_loess_one(.x)))) |>
  ungroup()

cat("LOESS fits successful:",
    sum(!vapply(loess_fits$fit, is.null, logical(1))),
    " /", nrow(loess_fits), "\n")
```

# P3.4 Apply factors to `annotation_chrom_metrics.csv`

**Data level: Annotated Signal.** From here on we operate on the
canonical annotated peak table (SIRIUS producer output). The
Detected-Signal factors built in P3.3 are looked up by
`(lab, sample, polarity, aligned-RT decile)` for each annotated peak.

Mixture format gotcha: `annotation_chrom_metrics.csv` uses `"1.1"` /
`"2.10"`; `detected_peaks_*_HE.csv` uses `"1_1"` / `"2_10"`. The join
to recover `sample` requires converting one to the other — done with
`gsub("\\.", "_", ...)`.

```{r load-annotation}
ann_chrom <- read.csv(
  file.path(producer_dir, "annotation_chrom_metrics.csv"),
  stringsAsFactors = FALSE,
  colClasses = c(mixture = "character"))

# Mixture-to-sample lookup from the detected peaks. Each (lab, polarity)
# has a deterministic mixture → sample mapping.
mixture_to_sample <- peaks |>
  distinct(lab, polarity, sample, mixture) |>
  rename(mixture_us = mixture)

ann_chrom <- ann_chrom |>
  mutate(mixture_us = gsub("\\.", "_", mixture)) |>
  left_join(mixture_to_sample,
            by = c("lab", "polarity", "mixture_us")) |>
  select(-mixture_us)

cat("annotation rows with sample resolved:",
    sum(!is.na(ann_chrom$sample)), " /", nrow(ann_chrom), "\n")
```

```{r align-annotation}
ann_chrom <- apply_splines(ann_chrom, splines, anchors_for_range,
                           rt_col = "apex_rt")

# Decile per polarity, using the same bin edges as the factor build.
ann_chrom$decile <- NA_integer_
for (pol in names(decile_breaks)) {
  idx <- which(ann_chrom$polarity == pol &
               !is.na(ann_chrom$aligned_rt))
  ann_chrom$decile[idx] <- as.integer(cut(ann_chrom$aligned_rt[idx],
                                          breaks         = decile_breaks[[pol]],
                                          labels         = FALSE,
                                          include.lowest = TRUE))
}

cat("annotation rows with aligned_rt:",
    sum(!is.na(ann_chrom$aligned_rt)), " /", nrow(ann_chrom), "\n")
cat("annotation rows with decile assigned:",
    sum(!is.na(ann_chrom$decile)),    " /", nrow(ann_chrom), "\n")
```

```{r join-factors}
ann_chrom <- ann_chrom |>
  left_join(factor_global_tbl,
            by = c("lab", "sample", "polarity")) |>
  left_join(factor_decile_tbl |>
              select(lab, sample, polarity, decile, factor_decile),
            by = c("lab", "sample", "polarity", "decile"))

# Evaluate LOESS factor at each annotated peak's aligned_rt.
ann_chrom$factor_loess <- NA_real_
for (i in seq_len(nrow(loess_fits))) {
  fit <- loess_fits$fit[[i]]
  if (is.null(fit)) next
  idx <- which(ann_chrom$lab      == loess_fits$lab[i]     &
               ann_chrom$sample   == loess_fits$sample[i]  &
               ann_chrom$polarity == loess_fits$polarity[i] &
               !is.na(ann_chrom$aligned_rt))
  if (!length(idx)) next
  pred <- suppressWarnings(
    predict(fit, newdata = data.frame(aligned_rt = ann_chrom$aligned_rt[idx])))
  # LOESS was fit on log10(factor) — convert back to linear.
  ann_chrom$factor_loess[idx] <- 10 ^ pred
}

cat("annotation rows with each factor:",
    "\n  global:", sum(!is.na(ann_chrom$factor_global)),
    "\n  decile:", sum(!is.na(ann_chrom$factor_decile)),
    "\n  loess :", sum(!is.na(ann_chrom$factor_loess)),
    " /", nrow(ann_chrom), "\n")
```

```{r apply-and-export}
# Scale both abundance metrics by each factor. We carry both forward
# because they answer different questions:
#   - `maxIntensity`   — scan-rate robust (good when scan rates differ ~15x
#                         across labs)
#   - `areaUnderTic`   — standard TIC-style integral (matches the bulk
#                         metric in 8_data_level_progression.qmd)
ann_chrom <- ann_chrom |>
  mutate(maxIntensity_global = maxIntensity * factor_global,
         maxIntensity_decile = maxIntensity * factor_decile,
         maxIntensity_loess  = maxIntensity * factor_loess,
         areaUnderTic_global = areaUnderTic * factor_global,
         areaUnderTic_decile = areaUnderTic * factor_decile,
         areaUnderTic_loess  = areaUnderTic * factor_loess)

write.csv(ann_chrom,
          file.path(object_dir, "annotation_chrom_metrics_normalized.csv"),
          row.names = FALSE)

cat("Wrote: ",
    file.path(object_dir, "annotation_chrom_metrics_normalized.csv"),
    "\n  rows: ", nrow(ann_chrom), "\n", sep = "")
```

```{r write-peak-counts}
# Small companion CSV used by `7_icc_reproducibility.qmd`'s detected-peak-count
# diagnostic. Keeps the consumer from having to re-read ~189 MB of
# detected-peak CSVs just to count rows per injection.
peak_counts <- peaks |>
  count(lab, polarity, sample, name = "n_peaks")

write.csv(peak_counts,
          file.path(object_dir, "detected_peak_counts.csv"),
          row.names = FALSE)

cat("Wrote: ",
    file.path(object_dir, "detected_peak_counts.csv"),
    "\n  rows: ", nrow(peak_counts), "\n", sep = "")
```

# Next step

The normalized annotated peaks are now in
`5_downstream_analysis/object/annotation_chrom_metrics_normalized.csv`.

Open `7_icc_reproducibility.qmd` for the bulk ICC ladder, pairwise
comparisons, per-lab diagnostics, and per-(compound, mixture) cross-lab
CV analysis — all of which read this file.
