---
title: "RT-local normalization — producer (Phase 3)"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
        embed-resources: true
---

# Goal — producer

Estimate the per-`(lab, injection, RT-decile)` response profile from
the **Detected Signal** (xcms output: ~1M peaks across the 4 labs),
apply that response correction back to the **Annotated Signal**
(canonical `annotation_chrom_metrics.csv` from the SIRIUS producer),
write the normalized annotated table to disk for downstream analysis.

This file is **both producer and evaluator**: §P3.1–§P3.3 build the
factor maps, §P3.4 picks the winning strategy per data level (writes
`normalization_winners.csv`), and §P3.5–§P3.6 are the full
strategy-comparison evidence (ICC ladder, variance decomposition,
pairwise ICC heatmap). The headline **reproducibility results** on
the chosen winner (Raw vs Lab-median pairwise + per-compound
Spearman) live in `7_reproducibility_results.qmd`.

Pipeline:

1. Load detected peaks, apply the P2 alignment splines.
2. Build three factor maps: `factor_global` (per-injection global
   scale), `factor_decile` (per-injection × RT-decile), `factor_loess`
   (smooth f(aligned_rt) per injection).
3. Apply each factor to the annotated peaks; write the result.

Inputs:

- `5_downstream_analysis/object/detected_peaks_{lab}_HE.csv` — per-lab
  `(rtmin, rtmax, mzmin, mzmax, rt, mz, sample, into, intb, maxo,
  polarity, mixture, chrom_peak_id)`.
- `5_downstream_analysis/object/rt_alignment_splines.RData` — per
  `(lab, polarity)` splines from P2 (`3_rt_alignment.qmd`).
- `5_downstream_analysis/object/rt_alignment_anchors.csv` — anchor
  table (for clamping out-of-range peaks).
- `5_downstream_analysis/object/annotation_chrom_metrics.csv`
  — the canonical annotated peak set we normalize.

Outputs:

- `5_downstream_analysis/object/annotation_chrom_metrics_normalized.csv`
  — annotated rows + `aligned_rt, decile, factor_global,
  factor_decile, factor_loess, maxIntensity_{global,decile,loess},
  areaUnderTic_{global,decile,loess}`.
- `5_downstream_analysis/object/detected_peak_counts.csv` — small
  `(lab, polarity, sample, n_peaks)` summary used by the consumer's
  diagnostic.

## Setup

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

labs           <- c("afekta", "cembio", "hmgu", "icl")
producer_dir   <- here("5_downstream_analysis", "object")
object_dir     <- here("5_downstream_analysis", "object")

# Number of RT bins (deciles → 10). Computed per polarity on the
# pooled aligned_rt distribution so bins are cross-lab comparable.
n_rt_bins <- 10L

# Palette for strategy comparison.
# "Aligned" is added as a neutral grey to isolate the contribution of the
# alignment step alone (no intensity factor) from the normalization on top.
# "Lab-median" is the global Annotated-Normalized formulation
# (per-(lab, polarity) median of the bulk distribution), kept side by
# side with the per-injection-decile-based "Global" so the reader can
# see which target works better.
strategy_palette <- c(Raw          = "#E64B35",
                      Aligned      = "#999999",
                      `Lab-median` = "#3B7C92",
                      Global       = "#4DBBD5",
                      Decile       = "#70AD47",
                      LOESS        = "#4472C4")
```

# P3.1 Load detected peaks (per-lab, per-injection)

**Data level: Detected Signal** — only used here to estimate the
response profile (P3.3). It is *not* the level on which P3.5 ICC is
computed.

Read the four per-lab CSVs, add `lab`, concatenate. Each row is one
detected peak; the `(lab, sample, polarity)` triple identifies one
injection (`sample` indexes 1..19 within each polarity per lab).

```{r load-detected-peaks}
read_lab_peaks <- function(lab) {
  read.csv(file.path(object_dir, sprintf("detected_peaks_%s_HE.csv", lab)),
           stringsAsFactors = FALSE,
           colClasses = c(mixture = "character")) |>
    mutate(lab = lab, .before = 1L,
           mixture = gsub("\\.", "_", mixture))
}

peaks <- bind_rows(lapply(labs, read_lab_peaks))

# Polarity column comes through as character ("pos" / "neg") — same as
# the chromExtract / annotation pipelines.
cat("Detected peaks loaded:\n")
peaks |> count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable(caption = "Peaks per (lab, polarity).")

cat("\nInjections (lab × polarity × sample) detected:\n")
peaks |> distinct(lab, polarity, sample) |>
  count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable(caption = "Distinct injections per (lab, polarity).")
```

# P3.2 Apply P2 alignment to detected peak `rt`

Load the per-`(lab, polarity)` Hyman splines from P2 and apply them to
each peak's `rt`, with the same anchor-range clamp used in
`3_rt_alignment.qmd`. Peaks outside the anchor range get
`aligned_rt = NA` and are dropped from the factor-fitting step (they'd
be extrapolated otherwise).

```{r apply-alignment}
load(file.path(object_dir, "rt_alignment_splines.RData"))   # `splines`
anchors_for_range <- read.csv(
  file.path(object_dir, "rt_alignment_anchors.csv"),
  stringsAsFactors = FALSE)

apply_splines <- function(df, splines_map, anchors_for_range, rt_col = "rt") {
  df$key        <- paste(df$lab, df$polarity, sep = "_")
  df$aligned_rt <- NA_real_
  for (key_i in unique(df$key)) {
    f <- splines_map[[key_i]]
    if (is.null(f)) next
    anchor_range <- anchors_for_range |>
      filter(paste(lab, polarity, sep = "_") == key_i) |>
      pull(rt_local) |> range(na.rm = TRUE)
    idx       <- which(df$key == key_i)
    rt_lab    <- df[[rt_col]][idx]
    in_range  <- !is.na(rt_lab) &
                 rt_lab >= anchor_range[1] &
                 rt_lab <= anchor_range[2]
    df$aligned_rt[idx[in_range]] <- as.numeric(f(rt_lab[in_range]))
  }
  df$key <- NULL
  df
}

peaks <- apply_splines(peaks, splines, anchors_for_range, rt_col = "rt")

cat("Peaks with aligned_rt (in-range):", sum(!is.na(peaks$aligned_rt)),
    " / ", nrow(peaks), "\n")
cat("Out-of-range (NA aligned_rt) per lab / polarity:\n")
peaks |> filter(is.na(aligned_rt)) |>
  count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n, values_fill = 0L) |>
  knitr::kable()
```

# P3.3 Compute normalization factors

We define **three factor maps** (Raw is implicit = 1 everywhere) that
each take a peak's `(lab, sample, polarity, aligned_rt)` and return a
multiplicative correction.

`Global` — per `(lab, polarity)`. Single number per lab × polarity =
median peak intensity / cross-lab consensus median intensity. Removes
overall sensitivity offset.

`Decile` — per `(lab, sample, polarity, decile)`. RT-binned median.
Removes both per-injection variability and RT-local sensitivity
biases.

`LOESS` — per `(lab, sample, polarity)`, smooth `f(aligned_rt)`. Same
spirit as `Decile` but continuous in RT — avoids bin-edge jumps.

We compute factors using `into` (peak area) as the response metric;
`maxo` is carried through unchanged for consistency with the existing
annotated-level bulk metric in `7_reproducibility_results.qmd`.

```{r factor-decile-build}
# Per-polarity aligned_rt deciles. Computed on the pooled distribution
# across all labs / samples so the bin edges are comparable.
peak_rt <- peaks |> filter(!is.na(aligned_rt), is.finite(into), into > 0)

# Bin edges per polarity (named list).
decile_breaks <- lapply(split(peak_rt$aligned_rt, peak_rt$polarity),
                        quantile,
                        probs   = seq(0, 1, length.out = n_rt_bins + 1L),
                        na.rm   = TRUE)

# Vectorized decile assignment — cut() per polarity, indexed back to
# the parent table. ~1M peaks; rowwise() would be unusable.
peak_rt$decile <- NA_integer_
for (pol in names(decile_breaks)) {
  idx <- which(peak_rt$polarity == pol)
  peak_rt$decile[idx] <- as.integer(cut(peak_rt$aligned_rt[idx],
                                        breaks         = decile_breaks[[pol]],
                                        labels         = FALSE,
                                        include.lowest = TRUE))
}

# Per (lab, sample, polarity, decile) lab response
lab_decile_response <- peak_rt |>
  group_by(lab, sample, polarity, decile) |>
  summarise(lab_median_into = median(into, na.rm = TRUE),
            n_peaks         = n(),
            .groups = "drop")

# Cross-lab consensus per (sample, polarity, decile)
consensus_decile <- lab_decile_response |>
  group_by(sample, polarity, decile) |>
  summarise(consensus_into = median(lab_median_into, na.rm = TRUE),
            n_labs         = n_distinct(lab),
            .groups = "drop") |>
  filter(n_labs >= 2L)

# Decile factor = consensus / lab. Multiply lab values by factor to land on consensus.
factor_decile_tbl <- lab_decile_response |>
  inner_join(consensus_decile |> select(sample, polarity, decile, consensus_into),
             by = c("sample", "polarity", "decile")) |>
  mutate(factor_decile = consensus_into / lab_median_into)

cat("Decile factor table — rows per (lab, polarity):\n")
factor_decile_tbl |> count(lab, polarity) |>
  pivot_wider(names_from = polarity, values_from = n) |>
  knitr::kable()
```

```{r factor-global-build}
# Global = collapse the decile factor by taking median across deciles
# per (lab, sample, polarity). This is the "one number per injection"
# version of the same correction.
factor_global_tbl <- factor_decile_tbl |>
  group_by(lab, sample, polarity) |>
  summarise(factor_global = median(factor_decile, na.rm = TRUE),
            .groups = "drop")
```

```{r factor-loess-build}
# LOESS — per (lab, sample, polarity), fit a smooth log10(factor) vs
# aligned_rt curve from the decile-midpoint factors. P3.4 evaluates
# this surface at each annotated peak's aligned_rt.
#
# Each polarity's decile midpoints are computed once from decile_breaks
# (list of break vectors, named by polarity).
decile_midpoints <- bind_rows(lapply(names(decile_breaks), function(pol) {
  br <- decile_breaks[[pol]]
  tibble(polarity   = pol,
         decile     = seq_len(n_rt_bins),
         aligned_rt = (br[-1] + br[-length(br)]) / 2)
}))

factor_decile_with_rt <- factor_decile_tbl |>
  left_join(decile_midpoints, by = c("polarity", "decile"))

fit_loess_one <- function(df) {
  if (nrow(df) < 5L) return(NULL)
  tryCatch(
    loess(log10(factor_decile) ~ aligned_rt,
          data       = df,
          span       = 0.75,
          degree     = 1,
          na.action  = na.exclude),
    error = function(e) NULL)
}

loess_fits <- factor_decile_with_rt |>
  group_by(lab, sample, polarity) |>
  group_modify(~ tibble(fit = list(fit_loess_one(.x)))) |>
  ungroup()

cat("LOESS fits successful:",
    sum(!vapply(loess_fits$fit, is.null, logical(1))),
    " /", nrow(loess_fits), "\n")
```

# P3.4 Apply factors to `annotation_chrom_metrics.csv`

**Data level: Annotated Signal.** From here on we operate on the
canonical annotated peak table (SIRIUS producer output). The
Detected-Signal factors built in P3.3 are looked up by
`(lab, sample, polarity, aligned-RT decile)` for each annotated peak.

All mixture IDs are now normalized to underscore form (`"1_1"`,
`"2_10"`) on read in every QMD, so the join to recover `sample` is a
plain `(lab, polarity, mixture)` match. The defensive `gsub` is kept
in case any input csv is still in dot form.

```{r load-annotation}
ann_chrom <- read.csv(
  file.path(producer_dir, "annotation_chrom_metrics.csv"),
  stringsAsFactors = FALSE,
  colClasses = c(mixture = "character"))
ann_chrom$mixture <- gsub("\\.", "_", ann_chrom$mixture)

# Mixture-to-sample lookup from the detected peaks. Each (lab, polarity)
# has a deterministic mixture → sample mapping.
mixture_to_sample <- peaks |>
  distinct(lab, polarity, sample, mixture) |>
  rename(mixture_us = mixture)

ann_chrom <- ann_chrom |>
  mutate(mixture_us = mixture) |>
  left_join(mixture_to_sample,
            by = c("lab", "polarity", "mixture_us")) |>
  select(-mixture_us)

cat("annotation rows with sample resolved:",
    sum(!is.na(ann_chrom$sample)), " /", nrow(ann_chrom), "\n")
```

```{r align-annotation}
ann_chrom <- apply_splines(ann_chrom, splines, anchors_for_range,
                           rt_col = "apex_rt")

# Decile per polarity, using the same bin edges as the factor build.
ann_chrom$decile <- NA_integer_
for (pol in names(decile_breaks)) {
  idx <- which(ann_chrom$polarity == pol &
               !is.na(ann_chrom$aligned_rt))
  ann_chrom$decile[idx] <- as.integer(cut(ann_chrom$aligned_rt[idx],
                                          breaks         = decile_breaks[[pol]],
                                          labels         = FALSE,
                                          include.lowest = TRUE))
}

cat("annotation rows with aligned_rt:",
    sum(!is.na(ann_chrom$aligned_rt)), " /", nrow(ann_chrom), "\n")
cat("annotation rows with decile assigned:",
    sum(!is.na(ann_chrom$decile)),    " /", nrow(ann_chrom), "\n")
```

```{r join-factors}
ann_chrom <- ann_chrom |>
  left_join(factor_global_tbl,
            by = c("lab", "sample", "polarity")) |>
  left_join(factor_decile_tbl |>
              select(lab, sample, polarity, decile, factor_decile),
            by = c("lab", "sample", "polarity", "decile"))

# Evaluate LOESS factor at each annotated peak's aligned_rt.
ann_chrom$factor_loess <- NA_real_
for (i in seq_len(nrow(loess_fits))) {
  fit <- loess_fits$fit[[i]]
  if (is.null(fit)) next
  idx <- which(ann_chrom$lab      == loess_fits$lab[i]     &
               ann_chrom$sample   == loess_fits$sample[i]  &
               ann_chrom$polarity == loess_fits$polarity[i] &
               !is.na(ann_chrom$aligned_rt))
  if (!length(idx)) next
  pred <- suppressWarnings(
    predict(fit, newdata = data.frame(aligned_rt = ann_chrom$aligned_rt[idx])))
  # LOESS was fit on log10(factor) — convert back to linear.
  ann_chrom$factor_loess[idx] <- 10 ^ pred
}

cat("annotation rows with each factor:",
    "\n  global:", sum(!is.na(ann_chrom$factor_global)),
    "\n  decile:", sum(!is.na(ann_chrom$factor_decile)),
    "\n  loess :", sum(!is.na(ann_chrom$factor_loess)),
    " /", nrow(ann_chrom), "\n")
```

```{r apply-and-export}
# Scale both abundance metrics by each factor. We carry both forward
# because they answer different questions:
#   - `maxIntensity`   — scan-rate robust (good when scan rates differ ~15x
#                         across labs)
#   - `areaUnderTic`   — standard TIC-style integral (matches the bulk
#                         metric in 7_reproducibility_results.qmd)
ann_chrom <- ann_chrom |>
  mutate(maxIntensity_global = maxIntensity * factor_global,
         maxIntensity_decile = maxIntensity * factor_decile,
         maxIntensity_loess  = maxIntensity * factor_loess,
         areaUnderTic_global = areaUnderTic * factor_global,
         areaUnderTic_decile = areaUnderTic * factor_decile,
         areaUnderTic_loess  = areaUnderTic * factor_loess)

write.csv(ann_chrom,
          file.path(object_dir, "annotation_chrom_metrics_normalized.csv"),
          row.names = FALSE)

cat("Wrote: ",
    file.path(object_dir, "annotation_chrom_metrics_normalized.csv"),
    "\n  rows: ", nrow(ann_chrom), "\n", sep = "")
```

```{r write-peak-counts}
# Small companion CSV used by `7_reproducibility_results.qmd`'s detected-peak-count
# diagnostic. Keeps the consumer from having to re-read ~189 MB of
# detected-peak CSVs just to count rows per injection.
peak_counts <- peaks |>
  count(lab, polarity, sample, name = "n_peaks")

write.csv(peak_counts,
          file.path(object_dir, "detected_peak_counts.csv"),
          row.names = FALSE)

cat("Wrote: ",
    file.path(object_dir, "detected_peak_counts.csv"),
    "\n  rows: ", nrow(peak_counts), "\n", sep = "")
```

# P3.4 Evaluate normalizations — pick a winner per data level

The downstream QMDs should use **one** normalization per data level, not
a buffet of options. This section evaluates each candidate strategy on
the **Annotated Signal** (where the bulk of downstream analyses sit)
and writes the chosen winner to `normalization_winners.csv`. Full /
Detected / Consensus stay raw — there's no per-injection global factor
worth applying at those levels (mixture-content differences dominate).

## P3.4.1 RT-bin variance diagnostic (independent of whether we use it)

Even if we end up not picking a per-bin normalization, the *spread of
per-injection responses across RT bins* tells us whether response
non-linearity along the gradient is large enough to matter. If
within-injection bin-to-bin spread is small (every bin agrees with the
injection-global factor), a global per-lab factor is sufficient.

```{r rt-bin-variance}
#| fig-cap: "Within-injection bin spread: for each (lab, sample, polarity), the SD of the per-decile log10 factor across the 10 RT bins. Large = response varies a lot along the gradient (a per-bin or LOESS factor would help); small = a global per-injection factor is sufficient."
#| fig-height: 4
bin_variance <- factor_decile_tbl |>
  dplyr::filter(is.finite(factor_decile), factor_decile > 0) |>
  group_by(lab, sample, polarity) |>
  summarise(log_factor_sd = sd(log10(factor_decile)),
            n_deciles_used = n(),
            .groups = "drop") |>
  dplyr::filter(n_deciles_used >= 3L)

ggplot(bin_variance, aes(x = lab, y = log_factor_sd, fill = lab)) +
  geom_violin(scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.size = 0.5, alpha = 0.9) +
  facet_wrap(~ polarity) +
  labs(title = "Within-injection RT-bin variance",
       subtitle = "log10 SD of per-decile factor across the 10 bins — how much does response vary along the gradient?",
       x = NULL, y = "SD(log10 decile factor)") +
  theme_bw() +
  theme(legend.position = "none")

# Median bin-SD per lab as a quick summary.
knitr::kable(
  bin_variance |>
    group_by(lab, polarity) |>
    summarise(median_bin_sd = round(median(log_factor_sd, na.rm = TRUE), 3),
              p75_bin_sd    = round(quantile(log_factor_sd, 0.75, na.rm = TRUE), 3),
              .groups       = "drop"),
  caption = "Median and p75 of within-injection bin-to-bin SD per lab. ≪0.1 means a global factor is enough; ≥0.3 means a per-bin factor is materially better.")
```

## P3.4.2 Strategy comparison — annotated bulk ICC per strategy

Compute the cross-lab bulk ICC (variance-component ratio defined in
QMD 7 §Definitions) for each `(strategy × polarity)` on the Annotated
Signal, using the same Strict compound set QMD 7 uses for its ICC
ladder. Declare the strategy with the highest ICC per polarity as the
winner.

```{r normalization-winner-eval}
#| fig-cap: "Bulk ICC of the Annotated Signal under each candidate normalization. Higher = labs agree more within each mixture relative to between-mixture spread. The Lab-median + Decile + LOESS candidates use the factor maps built above; Raw is `maxIntensity`; Aligned is Raw on the alignment-restricted compound set (isolating the alignment contribution from any intensity correction)."
#| fig-height: 4

# Lab-median strategy: divide each (lab, polarity)'s rows by that lab's
# bulk median (computed on the same compound set) — the canonical
# "Annotated-Normalized" strategy used by QMD 7.
compute_lab_median_factor <- function(df) {
  df |>
    group_by(lab, polarity) |>
    mutate(value_lab_median = median(maxIntensity, na.rm = TRUE),
           maxIntensity_labmed =
             maxIntensity / value_lab_median) |>
    ungroup()
}

# Strict compound set: present (well-detected) in all 4 labs for that polarity.
strict_set <- ann_chrom |>
  filter(!is.na(maxIntensity), maxIntensity > 0) |>
  distinct(mixture, compound, lab, polarity) |>
  count(mixture, compound, polarity) |>
  filter(n == 4L) |>
  select(mixture, compound, polarity)

strict_chrom <- ann_chrom |>
  semi_join(strict_set, by = c("mixture", "compound", "polarity")) |>
  filter(!is.na(maxIntensity), maxIntensity > 0)

strict_chrom <- compute_lab_median_factor(strict_chrom)

bulk_per_strat <- bind_rows(
  strict_chrom |> mutate(value = maxIntensity,            strategy = "Raw"),
  strict_chrom |> mutate(value = maxIntensity_labmed,     strategy = "Lab-median"),
  strict_chrom |> mutate(value = maxIntensity_global,     strategy = "Global"),
  strict_chrom |> mutate(value = maxIntensity_decile,     strategy = "Decile"),
  strict_chrom |> mutate(value = maxIntensity_loess,      strategy = "LOESS")) |>
  filter(!is.na(value), value > 0) |>
  group_by(lab, mixture, polarity, strategy) |>
  summarise(bulk = sum(value), .groups = "drop")

icc_one <- function(df) {
  df$logv  <- log2(df$bulk)
  between <- var(tapply(df$logv, df$mixture, mean, na.rm = TRUE), na.rm = TRUE)
  within  <- mean(tapply(df$logv, df$mixture, var, na.rm = TRUE), na.rm = TRUE)
  if (!is.finite(between + within) || (between + within) == 0) return(NA_real_)
  between / (between + within)
}
icc_strat <- bulk_per_strat |>
  group_by(polarity, strategy) |>
  group_modify(~ tibble(ICC = icc_one(.x))) |>
  ungroup() |>
  mutate(strategy = factor(strategy,
                            levels = c("Raw", "Lab-median",
                                       "Global", "Decile", "LOESS")))

ggplot(icc_strat, aes(x = strategy, y = ICC, fill = strategy)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f", ICC)), vjust = -0.4, size = 3.4) +
  scale_fill_manual(values = strategy_palette) +
  facet_wrap(~ polarity) +
  labs(title = "Strategy comparison on Annotated Signal (Strict set)",
       subtitle = "Higher = better cross-lab agreement after normalization",
       x = NULL, y = "Bulk ICC") +
  theme_bw() +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 1.05))

# Pick the winner per polarity and write to disk.
winners <- icc_strat |>
  group_by(polarity) |>
  slice_max(ICC, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(data_level = "Annotated Signal", polarity,
            winner_strategy = as.character(strategy),
            ICC = round(ICC, 3))

# Full / Detected / Consensus winners — by design "Raw" (no normalization
# applied at these levels; per-injection global factor is dominated by
# mixture-content differences, not lab response).
winners <- bind_rows(
  tidyr::expand_grid(data_level = c("Full Signal", "Detected Signal",
                                     "Consensus Signal"),
                     polarity   = c("pos", "neg")) |>
    mutate(winner_strategy = "Raw", ICC = NA_real_),
  winners)

write.csv(winners,
          file.path(object_dir, "normalization_winners.csv"),
          row.names = FALSE)
knitr::kable(winners, caption = "Chosen normalization per data level × polarity.")
```

# P3.5 Full ICC ladder — every strategy × metric × compound-set policy

The §P3.4 evaluation above looks at `maxIntensity` on the Strict set
only — enough to declare the winner. This section is the **full
ladder**: both abundance metrics (`maxIntensity` and `areaUnderTic`)
across two compound-set policies, with all five normalization
strategies. The plots below are the canonical reproducibility-by-
normalization evidence for the deck.

```{r p35-setup-subsets}
#| message: false
source(here("5_downstream_analysis", "helpers.R"))

strategy_palette <- c(Raw          = "#E64B35",
                      Aligned      = "#999999",
                      `Lab-median` = "#3B7C92",
                      Global       = "#4DBBD5",
                      Decile       = "#70AD47",
                      LOESS        = "#4472C4")
strategy_names <- c("Raw", "Aligned", "Lab-median",
                    "Global", "Decile", "LOESS")
metric_names   <- c("maxIntensity", "areaUnderTic")

# Well-detected set = (mixture, compound, polarity) tuples with a
# positive maxIntensity in all 4 labs (matches Annotated Signal in
# QMD 8 §1.2).
well_detected <- ann_chrom |>
  filter(!is.na(maxIntensity), maxIntensity > 0) |>
  distinct(mixture, lab, compound, polarity) |>
  group_by(mixture, compound, polarity) |>
  filter(n_distinct(lab) == 4L) |>
  ungroup() |>
  distinct(mixture, compound, polarity)

# Dominant peak per (lab, mixture, polarity, compound) — collapse
# adducts / sources to one value before summing.
ann_well <- ann_chrom |>
  inner_join(well_detected, by = c("mixture", "compound", "polarity")) |>
  group_by(lab, mixture, polarity, compound) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup()

# Strict subset = also require aligned_rt for all 4 labs.
well_aligned <- ann_well |>
  filter(!is.na(aligned_rt)) |>
  count(mixture, compound, polarity, name = "n_labs_aligned") |>
  filter(n_labs_aligned == 4L) |>
  select(mixture, compound, polarity)
ann_well_aligned <- ann_well |>
  inner_join(well_aligned, by = c("mixture", "compound", "polarity"))

cat("Well-detected (mixture, compound, polarity) tuples:",
    nrow(well_detected), "\n")
cat("Strict subset (all 4 labs aligned):", nrow(well_aligned), "\n")
```

```{r p35-bulk-icc-strategies}
make_bulk_for <- function(metric_col, mode = c("strict", "inclusive")) {
  mode       <- match.arg(mode)
  global_col <- paste0(metric_col, "_global")
  decile_col <- paste0(metric_col, "_decile")
  loess_col  <- paste0(metric_col, "_loess")

  factor_set <- if (mode == "strict") ann_well_aligned else ann_well
  set_label  <- if (mode == "strict")
                  sprintf("Strict (4/4 aligned, n = %d)", nrow(well_aligned))
                else
                  sprintf("Inclusive (all well-detected, n = %d)",
                          nrow(well_detected))

  bulk_raw <- factor_set |>
    group_by(lab, mixture, polarity) |>
    summarise(bulk = sum(.data[[metric_col]], na.rm = TRUE),
              .groups = "drop") |>
    mutate(strategy = "Raw")

  bulk_aligned <- if (mode == "strict") {
    ann_well_aligned |>
      group_by(lab, mixture, polarity) |>
      summarise(bulk = sum(.data[[metric_col]], na.rm = TRUE),
                .groups = "drop") |>
      mutate(strategy = "Aligned")
  } else NULL

  bulk_norm <- factor_set |>
    mutate(int_global = .data[[global_col]],
           int_decile = coalesce(.data[[decile_col]], .data[[global_col]]),
           int_loess  = coalesce(.data[[loess_col]],  .data[[global_col]])) |>
    group_by(lab, mixture, polarity) |>
    summarise(Global = sum(int_global, na.rm = TRUE),
              Decile = sum(int_decile, na.rm = TRUE),
              LOESS  = sum(int_loess,  na.rm = TRUE),
              .groups = "drop") |>
    pivot_longer(c(Global, Decile, LOESS),
                 names_to = "strategy", values_to = "bulk")

  bind_rows(bulk_raw, bulk_aligned, bulk_norm) |>
    mutate(metric = metric_col, alignment_set = set_label)
}

bulk_all <- bind_rows(
  make_bulk_for("maxIntensity", "strict"),
  make_bulk_for("areaUnderTic", "strict"),
  make_bulk_for("maxIntensity", "inclusive"),
  make_bulk_for("areaUnderTic", "inclusive"))

# Lab-median strategy = each (lab, polarity) bulk divided by that pair's
# median across mixtures. Computed from the Raw bulk distribution itself.
lab_med_per_set <- bulk_all |>
  filter(strategy == "Raw") |>
  group_by(alignment_set, metric, lab, polarity) |>
  summarise(med_bulk = median(bulk, na.rm = TRUE), .groups = "drop")

bulk_labmed <- bulk_all |>
  filter(strategy == "Raw") |>
  left_join(lab_med_per_set,
            by = c("alignment_set", "metric", "lab", "polarity")) |>
  mutate(bulk = bulk / med_bulk, strategy = "Lab-median") |>
  select(-med_bulk)
bulk_all <- bind_rows(bulk_all, bulk_labmed)

# ICC per (metric, polarity, strategy, alignment_set).
icc_all <- bind_rows(lapply(metric_names, function(m) {
  bind_rows(lapply(unique(bulk_all$alignment_set), function(aset) {
    bind_rows(lapply(c("pos", "neg"), function(pol_str) {
      pol_name <- if (pol_str == "pos") "Positive" else "Negative"
      bind_rows(lapply(strategy_names, function(s) {
        df <- bulk_all |> filter(metric == m, polarity == pol_str,
                                  alignment_set == aset, strategy == s)
        if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
        res <- compute_icc(df, "bulk")
        res$strategy      <- s
        res$polarity      <- pol_name
        res$metric        <- m
        res$alignment_set <- aset
        res
      }))
    }))
  }))
}))

icc_all |>
  transmute(alignment_set, metric, polarity, strategy,
            ICC = round(ICC, 3)) |>
  arrange(alignment_set, metric, polarity,
          factor(strategy, levels = strategy_names)) |>
  knitr::kable(caption = "Annotated-Signal bulk ICC per (alignment_set × metric × polarity × strategy).")
```

```{r p35-icc-bar-plot}
#| fig-cap: "Reproducibility (ICC) by normalization strategy. One panel per (compound-set policy × metric × polarity). Bars left-to-right show Raw → Aligned → Lab-median → Global → Decile → LOESS. Taller = more reproducible. Dashed red = ICC 0.5 (moderate-reproducibility landmark). Lab-median consistently wins."
#| fig-height: 12
icc_all |>
  mutate(strategy = factor(strategy, levels = strategy_names)) |>
  ggplot(aes(x = strategy, y = ICC, fill = strategy)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = sprintf("%.3f", ICC)),
            vjust = -0.4, size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  facet_grid(alignment_set + metric ~ polarity, switch = "y") +
  scale_fill_manual(values = strategy_palette) +
  ylim(NA, 1) +
  labs(x = NULL, y = "ICC (cross-lab reproducibility)",
       title = "Reproducibility (ICC) by normalization strategy",
       subtitle = "Higher = labs agree more within each mixture relative to between-mixture spread. Red = ICC 0.5.") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.y.left = element_text(size = 8))
```

```{r p35-variance-decomp-plot}
#| fig-cap: "Variance decomposition under each normalization strategy: how does total log2-bulk variance split into Between-mixture (the biological signal we want) vs Within-mixture (lab disagreement on the same mixture)? Bigger blue / smaller red = higher ICC."
#| fig-height: 12
icc_all |>
  mutate(strategy = factor(strategy, levels = strategy_names)) |>
  select(metric, polarity, strategy, alignment_set,
         Between_mixture_var, Within_mixture_var) |>
  pivot_longer(c(Between_mixture_var, Within_mixture_var),
               names_to = "source", values_to = "variance") |>
  mutate(source = recode(source,
                         Between_mixture_var = "Between mixtures (signal)",
                         Within_mixture_var  = "Within mixtures (lab noise)")) |>
  ggplot(aes(x = strategy, y = variance, fill = source)) +
  geom_col(position = "stack", alpha = 0.85) +
  facet_grid(alignment_set + metric ~ polarity,
             scales = "free_y", switch = "y") +
  scale_fill_manual(values = c(`Between mixtures (signal)`   = "#619CFF",
                               `Within mixtures (lab noise)` = "#F8766D")) +
  labs(x = NULL, y = "Variance (log2)", fill = NULL,
       title = "Variance decomposition — biology vs cross-lab disagreement",
       subtitle = "Between = mixture-to-mixture biology | Within = labs disagreeing on the same mixture") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.y.left = element_text(size = 8))
```

# P3.6 Pairwise ICC across all 6 lab pairs

The all-labs ICC pools 4 labs into one number. The pairwise version
scores **each pair on its own**: brighter blue = the pair agrees
within each mixture as tightly as the mixtures differ. The
**same-platform pair** (afekta + cembio, both Agilent, marked ★) is
the ceiling reference — what reproducibility is achievable when
platform is held constant. Cross-vendor pairs approaching the ★ tile =
harmonized LC + good normalization is closing the platform gap.

```{r p36-pairwise-icc}
strict_label <- sprintf("Strict (4/4 aligned, n = %d)", nrow(well_aligned))
lab_pairs <- list(
  c("afekta", "cembio"), c("afekta", "hmgu"), c("afekta", "icl"),
  c("cembio", "hmgu"),   c("cembio", "icl"),  c("hmgu",   "icl"))
agilent_labs <- c("afekta", "cembio")

icc_pairs <- bind_rows(lapply(lab_pairs, function(p) {
  pair_label <- paste(sort(p), collapse = " + ")
  bulk_pair  <- bulk_all |> filter(lab %in% p)
  bind_rows(lapply(metric_names, function(m) {
    bind_rows(lapply(unique(bulk_pair$alignment_set), function(aset) {
      bind_rows(lapply(c("pos", "neg"), function(pol_str) {
        pol_name <- if (pol_str == "pos") "Positive" else "Negative"
        bind_rows(lapply(strategy_names, function(s) {
          df <- bulk_pair |> filter(metric == m, polarity == pol_str,
                                    alignment_set == aset, strategy == s)
          if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
          res <- compute_icc(df, "bulk")
          res$strategy      <- s
          res$polarity      <- pol_name
          res$metric        <- m
          res$alignment_set <- aset
          res$pair          <- pair_label
          res$same_platform <- all(p %in% agilent_labs)
          res
        }))
      }))
    }))
  }))
}))
```

```{r p36-pairwise-heatmap}
#| fig-cap: "Pairwise ICC, all normalization strategies. ★ = same-platform Agilent pair (afekta + cembio), the ceiling reference. Brighter blue = the pair agrees within each mixture as tightly as the mixtures differ."
#| fig-height: 9
icc_pairs |>
  filter(alignment_set == strict_label) |>
  mutate(pair = factor(pair, levels = sapply(lab_pairs,
                                              \(p) paste(sort(p), collapse = " + "))),
         strategy = factor(strategy, levels = strategy_names),
         pair_label = ifelse(same_platform,
                             paste(as.character(pair), "★"),
                             as.character(pair))) |>
  ggplot(aes(x = strategy, y = reorder(pair_label, ICC), fill = ICC)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", ICC)), size = 2.7) +
  facet_grid(metric ~ polarity) +
  scale_fill_gradient2(low = "white", mid = "#FDC086", high = "#1F77B4",
                       midpoint = 0.5, limits = c(0, 1), name = "ICC") +
  labs(x = "Strategy", y = "Lab pair (★ = same platform)",
       title = "Pairwise ICC by normalization strategy",
       subtitle = sprintf("Strict subset (n = %d)", nrow(well_aligned))) +
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

# Next step

The normalized annotated peaks (with **every** candidate factor for
audit) live in `annotation_chrom_metrics_normalized.csv`; the **chosen
winner per data level × polarity** is in `normalization_winners.csv`.
Downstream QMDs should default to the winner and treat the other
strategies as comparison material only.

For the **headline reproducibility numbers on the chosen winner**
(Raw vs Lab-median pairwise ICC, per-compound Spearman ρ across lab
pairs, variance decomposition + PCA + UMAP at the data-level
progression), see `7_reproducibility_results.qmd`.
