---
title: "Adduct Grouping Method Development"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  dir: project
---

# Overview

This document develops an unsupervised method to group LC-MS detected peaks
originating from the same compound (adducts, in-source fragments, isotopes).

**Method Components:**

1. **Extra annotation**: Targeted adduct matching using RT-anchored peak groups
2. **RT clustering optimization**: Test different RT thresholds (2-6 sec)
3. **Metric-based clustering optimization**: Test metric combinations
4. **Evaluation**: Precision/Recall/F1 against annotated ground truth

**Output**: Best-performing method configuration for use in benchmarking.

---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(xcms)
library(MsCoreUtils)
library(MetaboCoreUtils)
library(MsFeatures)
library(Chromatograms)
library(philentropy)
library(mclust)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(MsQuality)
library(BiocParallel)
library(dplyr)

# Register parallel backend
workers <- as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", "4"))
if (is.na(workers) || workers < 1) workers <- 4
bp <- SnowParam(workers = workers)
register(bp)

base_path <- here::here("5_downstream_analysis")
cache_path <- file.path(base_path, "object", "method_dev_cache")
if (!dir.exists(cache_path)) dir.create(cache_path, recursive = TRUE)

log_ts <- function(msg) {
  cat("[", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "] ", msg, "\n", sep = "")
}

time_block <- function(label, expr) {
  log_ts(paste0("START: ", label))
  t0 <- Sys.time()
  on.exit({
    dt <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
    log_ts(paste0("END: ", label, " (", round(dt, 2), " sec)"))
  }, add = TRUE)
  force(expr)
}

log_ts(paste0("Registered workers: ", workers))
source(file.path(base_path, "rt_analysis_function.R"))
source(here::here("1_preprocessing/setup.R"))
```

---

# Section 1: Data Preparation

## 1.1 Load Detected Peaks

```{r load-detected}
pt_afekta <- read.csv(
  file.path(base_path, "object/detected_peaks_afekta_HE.csv")
)

cat("Total detected peaks (afekta):", nrow(pt_afekta), "\n")
cat("Samples:", length(unique(pt_afekta$sample)), "\n")
cat("Unique m/z × RT features:", length(unique(
  paste(pt_afekta$mz, pt_afekta$rt)
)), "\n")
```

## 1.2 Load Annotations (Ground Truth)

```{r load-annotations}
res_afekta <- read.csv(here::here(
  "4_library_generation", "afekta", "HE", "ring_trial_library_HE.csv"
))

cat("Annotated peaks:", nrow(res_afekta), "\n")
cat("Unique compounds:", length(unique(res_afekta$compound_name)), "\n")
cat("Unique adducts:", length(unique(res_afekta$adduct)), "\n")
```

## 1.3 Merge Annotations

```{r merge-annotations}
ann_cols <- c("chrom_peak_id", "compound_name", "adduct", "formula", "InChIKey")

pt_afekta <- merge(
  pt_afekta,
  res_afekta[, ann_cols],
  by = "chrom_peak_id",
  all.x = TRUE
)

pt_afekta$is_annotated <- !is.na(pt_afekta$compound_name)

cat("Annotated peaks in detected table:",
    sum(pt_afekta$is_annotated), "of", nrow(pt_afekta),
    "(", round(100 * mean(pt_afekta$is_annotated), 1), "%)\n")
```

---

# Section 2: Extra Annotation via Targeted Adduct Matching

**Goal**: Expand ground truth by finding likely adducts of known compounds
within RT clusters, improving evaluation coverage.

## 2.1 Initial RT Clustering (h = 4 sec)

Use complete-linkage hierarchical clustering to create RT groups:

```{r initial-rt-cluster}
pt_afekta <- pt_afekta[order(pt_afekta$rt), ]
pt_afekta$rt_cluster <- NA

for (samp in unique(pt_afekta$sample)) {
  idx <- which(pt_afekta$sample == samp)

  if (length(idx) <= 1) {
    if (length(idx) == 1) pt_afekta$rt_cluster[idx] <- paste0(samp, "_rt1")
    next
  }

  # Complete-linkage prevents chaining
  rt_hc <- hclust(dist(pt_afekta$rt[idx]), method = "complete")
  rt_clusters <- cutree(rt_hc, h = 4)
  pt_afekta$rt_cluster[idx] <- paste0(samp, "_rt", rt_clusters)
}

cat("RT clusters created:", length(unique(pt_afekta$rt_cluster)), "\n")
cat("Peaks per cluster - median:",
    median(table(pt_afekta$rt_cluster)), "\n")
```

## 2.2 Build Expected Adduct m/z

```{r expected-adducts}
library(MetaboAnnotation)

# Get all singly-charged adducts
all_adducts_tbl <- rbind(adducts("positive"), adducts("negative"))
all_adducts_tbl <- all_adducts_tbl[abs(all_adducts_tbl$charge) == 1, ]

pos_adduct_names <- all_adducts_tbl$name[all_adducts_tbl$positive]
neg_adduct_names <- all_adducts_tbl$name[!all_adducts_tbl$positive]

# Remove implausible adducts (multimers, rare reagent-based)
pos_adduct_names <- pos_adduct_names[!grepl("^\\[2M|^\\[3M", pos_adduct_names)]
neg_adduct_names <- neg_adduct_names[!grepl("^\\[2M|^\\[3M", neg_adduct_names)]

implausible <- c(
  "[M+Li]+", "[M+2Li-H]+", "[M+Br]-", "[M+C2F3O2]-",
  "[M+C2H6OS+H]+", "[M+C4H6N2+H]+", "[M+C3H8O+H]+"
)
pos_adduct_names <- setdiff(pos_adduct_names, implausible)
neg_adduct_names <- setdiff(neg_adduct_names, implausible)

# Remove adducts already used in annotation
annotated_adducts <- unique(res_afekta$adduct)
pos_adduct_names <- setdiff(pos_adduct_names, annotated_adducts)
neg_adduct_names <- setdiff(neg_adduct_names, annotated_adducts)

cat("Target adducts for new annotation:\n")
cat("  Positive:", length(pos_adduct_names), "\n")
cat("  Negative:", length(neg_adduct_names), "\n")

# Compute neutral mass for each compound
compound_mass <- unique(res_afekta[, c("compound_name", "formula")])
compound_mass$neutral_mass <- calculateMass(compound_mass$formula)

cat("Compounds with neutral mass:", nrow(compound_mass), "\n")
```

## 2.3 Search for Matching Unannotated Peaks

```{r adduct-search}
adduct_cache_file <- file.path(cache_path, "adduct_hits.RData")

if (file.exists(adduct_cache_file)) {
  log_ts("Loading cached adduct hits")
  load(adduct_cache_file)
} else {
  time_block("adduct-search", {
    # Group annotated peaks by RT cluster
    ann_by_cluster <- split(
      pt_afekta[pt_afekta$is_annotated, ],
      pt_afekta[pt_afekta$is_annotated, ]$rt_cluster
    )

    ppm_tol <- 5
    param_pos <- Mass2MzParam(adducts = pos_adduct_names, tolerance = 0, ppm = ppm_tol)
    param_neg <- Mass2MzParam(adducts = neg_adduct_names, tolerance = 0, ppm = ppm_tol)

    adduct_hits <- do.call(rbind, lapply(
      names(ann_by_cluster), function(cluster_id) {
        ann_in_cluster <- ann_by_cluster[[cluster_id]]
        all_in_cluster <- pt_afekta[pt_afekta$rt_cluster == cluster_id, ]

        # Only search unannotated peaks
        unannotated_idx <- !all_in_cluster$chrom_peak_id %in% ann_in_cluster$chrom_peak_id
        unannotated <- all_in_cluster[unannotated_idx, ]
        if (nrow(unannotated) == 0) return(NULL)

        # For each annotated compound, search for new adducts
        results <- do.call(rbind, lapply(
          unique(ann_in_cluster$compound_name), function(cmpd) {
            neutral_mass <- compound_mass$neutral_mass[
              compound_mass$compound_name == cmpd
            ]
            if (length(neutral_mass) == 0 || is.na(neutral_mass)) return(NULL)

            # Determine polarity
            pol <- unique(ann_in_cluster$polarity[ann_in_cluster$compound_name == cmpd])
            if (length(pol) == 0) return(NULL)

            # Match unannotated peaks
            target_df <- data.frame(exactmass = neutral_mass)
            query_df <- data.frame(mz = unannotated$mz)
            param <- if (pol == "pos") param_pos else param_neg

            res <- matchValues(query_df, target_df, param)
            md <- matchedData(res)

            # Keep only matched rows
            matched_idx <- which(!is.na(md$adduct))
            if (length(matched_idx) == 0) return(NULL)

            out <- unannotated[matched_idx, ]
            out$matched_compound <- cmpd
            out$matched_adduct <- as.character(md$adduct[matched_idx])
            out$mz_error_ppm <- as.numeric(md$score[matched_idx])
            out
          }
        ))

        results
      }
    ))
    if (!is.null(adduct_hits)) rownames(adduct_hits) <- NULL
    save(adduct_hits, file = adduct_cache_file)
  })
}

if (!is.null(adduct_hits) && nrow(adduct_hits) > 0) {
  cat("New adduct matches found:", nrow(adduct_hits), "\n")
  cat("Unique peaks newly matched:", length(unique(adduct_hits$chrom_peak_id)), "\n")

  # Merge back into main table
  pt_afekta <- merge(
    pt_afekta,
    adduct_hits[, c("chrom_peak_id", "matched_compound", "matched_adduct", "mz_error_ppm")],
    by = "chrom_peak_id",
    all.x = TRUE
  )

  # Fill compound_name for newly annotated peaks
  pt_afekta$compound_name <- ifelse(
    !is.na(pt_afekta$matched_compound),
    pt_afekta$matched_compound,
    pt_afekta$compound_name
  )

  # Update is_annotated flag
  pt_afekta$is_annotated <- pt_afekta$is_annotated | !is.na(pt_afekta$matched_adduct)

  cat("Total annotated peaks after expansion:",
      sum(pt_afekta$is_annotated), "\n")
} else {
  cat("No new adduct matches found\n")
}
```

---

# Section 3: Extract EICs and Compute Peak Metrics

## 3.1 EIC Extraction (2-pass strategy)

```{r extract-eics}
# Get unique peaks
pk_rows <- pt_afekta[!duplicated(pt_afekta$chrom_peak_id), ]

# m/z tolerance
mz_tolerance <- 0.01  # ~10 ppm at 1000 m/z

# Pass 1: Extract with 5-second buffers for initial metrics
pk_table <- data.frame(
  chrom_peak_id = pk_rows$chrom_peak_id,
  mzMin = pk_rows$mzmin - mz_tolerance,
  mzMax = pk_rows$mzmax + mz_tolerance,
  rtMin = pk_rows$rtmin - 5,
  rtMax = pk_rows$rtmax + 5,
  polarity = ifelse(pk_rows$polarity == "pos", 1L, 0L),
  mixture = pk_rows$mixture,
  sample = pk_rows$sample,
  stringsAsFactors = FALSE
)

eics_file <- file.path(base_path, "object/eics_afekta_method_dev.RData")

if (!file.exists(eics_file)) {
  load(file.path(base_path, "object/tic_full_spectra.RData"))
  tic_afekta <- tic_full[tic_full$lab == "afekta"]

  eics_afekta <- chromExtract(
    tic_afekta,
    peak.table = pk_table,
    by = c("polarity", "mixture")
  )
  eics_afekta <- setBackend(eics_afekta, ChromBackendMemory())
  save(eics_afekta, file = eics_file)
  rm(tic_full, tic_afekta)
} else {
  load(eics_file)
}

cat("EICs extracted (Pass 1):", length(eics_afekta), "\n")
```

## 3.2 Compute Pass 1 Metrics

```{r pass1-metrics}
metrics_pass1 <- c(
  "peakCount",
  "xicFwhm",
  "areaUnderTic",
  "maxIntensity",
  "baselineIntensity",
  "signalToNoiseRatio",
  "peakWidth",
  "peakBoundary",
  "peakProminence",
  "peakBeta"
)

pass1_cache_file <- file.path(cache_path, "pk_metrics_pass1.RData")
if (file.exists(pass1_cache_file)) {
  log_ts("Loading cached Pass 1 metrics")
  load(pass1_cache_file)
} else {
  time_block("pass1-metrics", {
    pk_metrics_pass1 <- calculateMetrics(
      eics_afekta,
      metrics = metrics_pass1,
      na.rm = TRUE,
      filterEmptyObject = FALSE,
      f = seq_len(length(eics_afekta)),
      probs = 0.05,
      method = "adaptive",
      minPeakWidth = 3
    )

    # Extract metadata
    cd <- chromData(eics_afekta)
    pk_metrics_pass1 <- cbind(
      cd[, intersect(c("chrom_peak_id", "mixture", "polarity", "mz", "sample"),
                     colnames(cd))],
      pk_metrics_pass1
    )
    save(pk_metrics_pass1, file = pass1_cache_file)
  })
}

cat("Pass 1 metrics computed:", nrow(pk_metrics_pass1), "peaks\n")
```

## 3.3 Quality Filtering

```{r quality-filter}
n_before <- nrow(pt_afekta)

metric_cols_pass1 <- setdiff(colnames(pk_metrics_pass1),
                             c("chrom_peak_id", "mixture", "polarity", "mz", "sample"))

# Merge metrics
pt_afekta <- merge(pt_afekta, pk_metrics_pass1[, c("chrom_peak_id", metric_cols_pass1)],
                   by = "chrom_peak_id", all.x = TRUE)

# Quality filters
bad <- is.na(pt_afekta$xicFwhm) |
  (!is.na(pt_afekta$peakProminence) & pt_afekta$peakProminence < 1.0) |
  (!is.na(pt_afekta$peakWidth) & (pt_afekta$peakWidth < 1 | pt_afekta$peakWidth > 30))

n_removed <- sum(bad)
pt_afekta <- pt_afekta[!bad, ]

cat("Quality filtering:\n")
cat("  Removed:", n_removed, "peaks\n")
cat("  Remaining:", nrow(pt_afekta), "peaks\n")
cat("  Annotated retained:", sum(pt_afekta$is_annotated), "\n")
```

## 3.4 Re-extract EICs with Refined Boundaries

```{r reextract-eics}
# Use peakBoundary from Pass 1 to refine EIC windows
peak_boundaries <- pk_metrics_pass1[, c("chrom_peak_id", "peakBoundary1", "peakBoundary2")]
peak_boundaries <- peak_boundaries[peak_boundaries$chrom_peak_id %in% pt_afekta$chrom_peak_id, ]

eics_afekta_refined <- lapply(seq_len(nrow(peak_boundaries)), function(i) {
  eic_id <- peak_boundaries$chrom_peak_id[i]
  boundary1 <- peak_boundaries$peakBoundary1[i]
  boundary2 <- peak_boundaries$peakBoundary2[i]

  idx <- which(eics_afekta$chrom_peak_id == eic_id)
  if (length(idx) == 0) return(NULL)

  eic <- eics_afekta[idx]
  eic <- filterPeaksData(
    eic,
    variables = "rtime",
    ranges = c(boundary1 - 0.05, boundary2 + 0.05)  # 3-sec margin
  )
  eic <- applyProcessing(eic)
  eic
})

eics_afekta_refined <- eics_afekta_refined[!sapply(eics_afekta_refined, is.null)]
eics_afekta_refined <- concatenateChromatograms(eics_afekta_refined)

cat("Re-extracted EICs:", length(eics_afekta_refined), "\n")
```

## 3.5 Compute Pass 2 Metrics (Refined)

```{r pass2-metrics}
metrics_pass2 <- c(
  "xicFwhm",
  "peakWidth",
  "peakBeta",
  "signalToNoiseRatio",
  "peakProminence"
)

pass2_cache_file <- file.path(cache_path, "pk_metrics_pass2.RData")
if (file.exists(pass2_cache_file)) {
  log_ts("Loading cached Pass 2 metrics")
  load(pass2_cache_file)
} else {
  time_block("pass2-metrics", {
    pk_metrics_pass2 <- calculateMetrics(
      eics_afekta_refined,
      metrics = metrics_pass2,
      na.rm = TRUE,
      filterEmptyObject = FALSE,
      f = seq_len(length(eics_afekta_refined)),
      probs = 0.05,
      method = "adaptive",
      minPeakWidth = 3
    )

    cd2 <- chromData(eics_afekta_refined)
    pk_metrics_pass2 <- cbind(
      cd2[, "chrom_peak_id", drop = FALSE],
      pk_metrics_pass2
    )
    save(pk_metrics_pass2, file = pass2_cache_file)
  })
}

# Rename refined metrics
colnames(pk_metrics_pass2) <- c("chrom_peak_id", paste0(metrics_pass2, "_refined"))

# Merge refined metrics
pt_afekta <- merge(pt_afekta, pk_metrics_pass2, by = "chrom_peak_id", all.x = TRUE)

cat("Pass 2 metrics computed:", nrow(pk_metrics_pass2), "peaks\n")
```

## 3.6 LOESS Intensity Correction

Remove intensity bias from shape metrics using LOESS regression:

```{r loess-correction}
cat("\n--- Correcting Shape Metrics for Intensity Bias ---\n")

pt_afekta$log_intensity <- log10(pt_afekta$maxIntensity)

# Correct FWHM
fwhm_model <- loess(xicFwhm_refined ~ log_intensity, data = pt_afekta,
                    span = 0.5, na.action = na.exclude)
pt_afekta$corrected_FWHM <- residuals(fwhm_model)

# Correct Width
width_model <- loess(peakWidth_refined ~ log_intensity, data = pt_afekta,
                     span = 0.5, na.action = na.exclude)
pt_afekta$corrected_Width <- residuals(width_model)

# Correct Beta1
beta_model <- loess(peakBeta_refined ~ log_intensity, data = pt_afekta,
                    span = 0.5, na.action = na.exclude)
pt_afekta$corrected_Beta1 <- residuals(beta_model)

cat("Corrected metrics created:\n")
cat("  corrected_FWHM, corrected_Width, corrected_Beta1\n")
```

---

# Section 4: Optimize RT Clustering Threshold

Test different RT thresholds (2, 3, 4, 5, 6 seconds) to find optimal:

```{r optimize-rt-threshold}
cat("\n=== OPTIMIZING RT CLUSTERING THRESHOLD ===\n\n")

# Evaluation helper function
evaluate_clustering <- function(eval_df, cluster_col = "final_cluster") {
  # Generate all pairwise comparisons
  pairs <- combn(seq_len(nrow(eval_df)), 2, simplify = FALSE)

  pair_results <- lapply(pairs, function(p) {
    i <- p[1]
    j <- p[2]

    true_compound_1 <- eval_df$compound_sample[i]
    true_compound_2 <- eval_df$compound_sample[j]
    pred_cluster_1 <- eval_df[[cluster_col]][i]
    pred_cluster_2 <- eval_df[[cluster_col]][j]

    is_true_pair <- (true_compound_1 == true_compound_2)
    is_pred_pair <- (pred_cluster_1 == pred_cluster_2)

    data.frame(
      is_true_pair = is_true_pair,
      is_pred_pair = is_pred_pair
    )
  })

  pair_df <- do.call(rbind, pair_results)

  tp <- sum(pair_df$is_true_pair & pair_df$is_pred_pair)
  fp <- sum(!pair_df$is_true_pair & pair_df$is_pred_pair)
  fn <- sum(pair_df$is_true_pair & !pair_df$is_pred_pair)
  tn <- sum(!pair_df$is_true_pair & !pair_df$is_pred_pair)

  precision <- if ((tp + fp) > 0) tp / (tp + fp) else 0
  recall <- if ((tp + fn) > 0) tp / (tp + fn) else 0
  f1 <- if ((precision + recall) > 0) 2 * ((precision * recall) / (precision + recall)) else 0

  list(
    precision = precision,
    recall = recall,
    f1 = f1,
    tp = tp,
    fp = fp,
    fn = fn,
    tn = tn
  )
}

# Prepare evaluation data (annotated multi-peak compounds only)
compound_counts <- table(paste(pt_afekta$compound_name, pt_afekta$sample))
multi_peak_compounds <- names(compound_counts)[compound_counts > 1]

eval_data <- pt_afekta[pt_afekta$is_annotated, ]
eval_data$compound_sample <- paste(eval_data$compound_name, eval_data$sample)
eval_data <- eval_data[eval_data$compound_sample %in% multi_peak_compounds, ]

cat("Evaluation dataset:\n")
cat("  Multi-peak compounds:", length(unique(eval_data$compound_sample)), "\n")
cat("  Total peaks:", nrow(eval_data), "\n\n")

rt_opt_cache_file <- file.path(cache_path, "rt_optimization.RData")
if (file.exists(rt_opt_cache_file)) {
  log_ts("Loading cached RT-threshold optimization")
  load(rt_opt_cache_file)
} else {
  time_block("optimize-rt-threshold", {
    # Test RT thresholds
    rt_thresholds <- c(2, 3, 4, 5, 6)
    rt_results <- list()

    for (h in rt_thresholds) {
      cat("Testing RT threshold h =", h, "sec...\n")

      # Re-cluster with this threshold
      pt_temp <- pt_afekta
      pt_temp$rt_cluster_test <- NA

      for (samp in unique(pt_temp$sample)) {
        idx <- which(pt_temp$sample == samp)
        if (length(idx) <= 1) {
          if (length(idx) == 1) pt_temp$rt_cluster_test[idx] <- paste0(samp, "_rt1")
          next
        }

        rt_hc <- hclust(dist(pt_temp$rt[idx]), method = "complete")
        rt_clusters <- cutree(rt_hc, h = h)
        pt_temp$rt_cluster_test[idx] <- paste0(samp, "_rt", rt_clusters)
      }

      # Evaluate on eval_data
      eval_temp <- eval_data
      eval_temp$rt_cluster_test <- pt_temp$rt_cluster_test[
        match(eval_temp$chrom_peak_id, pt_temp$chrom_peak_id)
      ]

      perf <- evaluate_clustering(eval_temp, cluster_col = "rt_cluster_test")

      rt_results[[paste0("h", h)]] <- data.frame(
        threshold = h,
        precision = perf$precision,
        recall = perf$recall,
        f1 = perf$f1,
        n_clusters = length(unique(pt_temp$rt_cluster_test))
      )

      cat("  Precision:", round(perf$precision * 100, 1), "%\n")
      cat("  Recall:", round(perf$recall * 100, 1), "%\n")
      cat("  F1:", round(perf$f1, 3), "\n")
      cat("  Clusters:", length(unique(pt_temp$rt_cluster_test)), "\n\n")
    }

    rt_results_df <- do.call(rbind, rt_results)
    save(rt_results_df, file = rt_opt_cache_file)
  })
}

best_rt_idx <- which.max(rt_results_df$f1)
best_rt_threshold <- rt_results_df$threshold[best_rt_idx]

cat("BEST RT THRESHOLD: h =", best_rt_threshold, "sec\n")
cat("F1-Score:", round(rt_results_df$f1[best_rt_idx], 3), "\n\n")

# Apply best threshold
pt_afekta$rt_cluster <- NA
for (samp in unique(pt_afekta$sample)) {
  idx <- which(pt_afekta$sample == samp)
  if (length(idx) <= 1) {
    if (length(idx) == 1) pt_afekta$rt_cluster[idx] <- paste0(samp, "_rt1")
    next
  }

  rt_hc <- hclust(dist(pt_afekta$rt[idx]), method = "complete")
  rt_clusters <- cutree(rt_hc, h = best_rt_threshold)
  pt_afekta$rt_cluster[idx] <- paste0(samp, "_rt", rt_clusters)
}

cat("Final RT clustering applied with h =", best_rt_threshold, "\n")
```

---

# Section 5: Optimize Metric-Based Clustering

Test different metric combinations to find best performer:

```{r optimize-metrics}
cat("\n=== OPTIMIZING METRIC-BASED CLUSTERING ===\n\n")

# Define metric combinations to test
metric_combos <- list(
  # Singles (raw)
  xicFwhm_raw = c("xicFwhm_refined"),
  peakWidth_raw = c("peakWidth_refined"),
  peakBeta_raw = c("peakBeta_refined"),

  # Singles (corrected)
  FWHM_corrected = c("corrected_FWHM"),
  Width_corrected = c("corrected_Width"),
  Beta_corrected = c("corrected_Beta1"),

  # Pairs (raw)
  FWHM_Width_raw = c("xicFwhm_refined", "peakWidth_refined"),
  FWHM_Beta_raw = c("xicFwhm_refined", "peakBeta_refined"),
  Width_Beta_raw = c("peakWidth_refined", "peakBeta_refined"),

  # Pairs (corrected)
  FWHM_Width_corrected = c("corrected_FWHM", "corrected_Width"),
  FWHM_Beta_corrected = c("corrected_FWHM", "corrected_Beta1"),
  Width_Beta_corrected = c("corrected_Width", "corrected_Beta1"),

  # Triples (raw)
  all_raw = c("xicFwhm_refined", "peakWidth_refined", "peakBeta_refined"),

  # Triples (corrected)
  all_corrected = c("corrected_FWHM", "corrected_Width", "corrected_Beta1"),

  # Hybrids
  FWHM_corrected_Width_Beta_raw = c("corrected_FWHM", "peakWidth_refined", "peakBeta_refined"),
  FWHM_Width_corrected_Beta_raw = c("xicFwhm_refined", "corrected_Width", "peakBeta_refined"),
  FWHM_Width_raw_Beta_corrected = c("xicFwhm_refined", "peakWidth_refined", "corrected_Beta1"),
  all_hybrid = c("corrected_FWHM", "corrected_Width", "peakBeta_refined")
)

cat("Testing", length(metric_combos), "metric combinations...\n\n")

metric_opt_cache_file <- file.path(cache_path, "metric_optimization.RData")
if (file.exists(metric_opt_cache_file)) {
  log_ts("Loading cached metric-combination optimization")
  load(metric_opt_cache_file)
} else {
  time_block("optimize-metrics", {
    combo_results <- list()

    for (combo_name in names(metric_combos)) {
      metrics <- metric_combos[[combo_name]]

      cat("Testing:", combo_name, "(", paste(metrics, collapse = ", "), ")...\n")

      # Cluster within each RT cluster using these metrics
      pt_temp <- pt_afekta
      pt_temp$metric_cluster <- NA

      for (rt_clust in unique(pt_temp$rt_cluster)) {
        idx <- which(pt_temp$rt_cluster == rt_clust)

        if (length(idx) <= 1) {
          if (length(idx) == 1) pt_temp$metric_cluster[idx] <- paste0(rt_clust, "_m1")
          next
        }

        # Extract metric values
        metric_matrix <- as.matrix(pt_temp[idx, metrics, drop = FALSE])

        # Remove rows with all NA
        valid_rows <- rowSums(!is.na(metric_matrix)) > 0
        if (sum(valid_rows) <= 1) {
          pt_temp$metric_cluster[idx[valid_rows]] <- paste0(rt_clust, "_m1")
          next
        }

        metric_matrix <- metric_matrix[valid_rows, , drop = FALSE]

        # Impute remaining NA with column median
        for (col in seq_len(ncol(metric_matrix))) {
          na_idx <- is.na(metric_matrix[, col])
          if (any(na_idx)) {
            metric_matrix[na_idx, col] <- median(metric_matrix[, col], na.rm = TRUE)
          }
        }

        # Scale metrics
        metric_scaled <- scale(metric_matrix)

        # Average-linkage clustering
        metric_hc <- hclust(dist(metric_scaled), method = "average")
        metric_clusters <- cutree(metric_hc, h = 0.8)

        pt_temp$metric_cluster[idx[valid_rows]] <- paste0(rt_clust, "_m", metric_clusters)
      }

      # Evaluate
      eval_temp <- eval_data
      eval_temp$metric_cluster <- pt_temp$metric_cluster[
        match(eval_temp$chrom_peak_id, pt_temp$chrom_peak_id)
      ]

      perf <- evaluate_clustering(eval_temp, cluster_col = "metric_cluster")

      combo_results[[combo_name]] <- data.frame(
        metric_combo = combo_name,
        precision = perf$precision,
        recall = perf$recall,
        f1 = perf$f1,
        metrics_used = paste(metrics, collapse = "; "),
        stringsAsFactors = FALSE
      )

      cat("  Precision:", round(perf$precision * 100, 1), "%\n")
      cat("  Recall:", round(perf$recall * 100, 1), "%\n")
      cat("  F1:", round(perf$f1, 3), "\n\n")
    }

    combo_results_df <- do.call(rbind, combo_results)
    combo_results_df <- combo_results_df[order(-combo_results_df$f1), ]
    save(combo_results_df, file = metric_opt_cache_file)
  })
}

cat("=== METRIC COMBINATION RESULTS (sorted by F1) ===\n")
print(combo_results_df, row.names = FALSE)

best_combo <- combo_results_df$metric_combo[1]
best_metrics <- metric_combos[[best_combo]]
best_f1 <- combo_results_df$f1[1]

cat("\nBEST METRIC COMBINATION:", best_combo, "\n")
cat("Metrics:", paste(best_metrics, collapse = ", "), "\n")
cat("F1-Score:", round(best_f1, 3), "\n")
```

---

# Section 6: Apply Best Configuration

Apply optimal RT threshold + metric combination to full dataset:

```{r apply-best-config}
cat("\n=== APPLYING BEST CONFIGURATION ===\n\n")

# RT clustering already applied above (best_rt_threshold)

# Metric clustering with best_metrics
pt_afekta$final_cluster <- NA

for (rt_clust in unique(pt_afekta$rt_cluster)) {
  idx <- which(pt_afekta$rt_cluster == rt_clust)

  if (length(idx) <= 1) {
    if (length(idx) == 1) pt_afekta$final_cluster[idx] <- paste0(rt_clust, "_m1")
    next
  }

  metric_matrix <- as.matrix(pt_afekta[idx, best_metrics, drop = FALSE])

  valid_rows <- rowSums(!is.na(metric_matrix)) > 0
  if (sum(valid_rows) <= 1) {
    pt_afekta$final_cluster[idx[valid_rows]] <- paste0(rt_clust, "_m1")
    next
  }

  metric_matrix <- metric_matrix[valid_rows, , drop = FALSE]

  for (col in seq_len(ncol(metric_matrix))) {
    na_idx <- is.na(metric_matrix[, col])
    if (any(na_idx)) {
      metric_matrix[na_idx, col] <- median(metric_matrix[, col], na.rm = TRUE)
    }
  }

  metric_scaled <- scale(metric_matrix)
  metric_hc <- hclust(dist(metric_scaled), method = "average")
  metric_clusters <- cutree(metric_hc, h = 0.8)

  pt_afekta$final_cluster[idx[valid_rows]] <- paste0(rt_clust, "_m", metric_clusters)
}

cat("Final clustering applied\n")
cat("Total clusters:", length(unique(pt_afekta$final_cluster)), "\n")

# Final evaluation
eval_data$final_cluster <- pt_afekta$final_cluster[
  match(eval_data$chrom_peak_id, pt_afekta$chrom_peak_id)
]

final_perf <- evaluate_clustering(eval_data, cluster_col = "final_cluster")

cat("\n=== FINAL PERFORMANCE ===\n")
cat("RT Threshold:", best_rt_threshold, "sec\n")
cat("Metric Combination:", best_combo, "\n")
cat("Metrics:", paste(best_metrics, collapse = ", "), "\n")
cat("Precision:", round(final_perf$precision * 100, 1), "%\n")
cat("Recall:", round(final_perf$recall * 100, 1), "%\n")
cat("F1-Score:", round(final_perf$f1, 3), "\n")
```

---

# Section 7: Save Results for Benchmarking

```{r save-results}
# Save optimized configuration
method_config <- list(
  rt_threshold = best_rt_threshold,
  metric_combination = best_combo,
  metrics = best_metrics,
  performance = final_perf,
  rt_results = rt_results_df,
  metric_results = combo_results_df
)

save(method_config, file = file.path(base_path, "object/optimized_method_config.RData"))

# Save clustered peak table
save(pt_afekta, file = file.path(base_path, "object/pt_afekta_optimized.RData"))

# Save EICs for benchmarking
save(eics_afekta_refined, file = file.path(base_path, "object/eics_afekta_refined.RData"))

cat("\n=== SAVED FOR BENCHMARKING ===\n")
cat("Config: object/optimized_method_config.RData\n")
cat("Peak table: object/pt_afekta_optimized.RData\n")
cat("EICs: object/eics_afekta_refined.RData\n")
```

---

# Summary

**Optimized Method Configuration:**

- **RT Clustering**: Complete-linkage, h = `r best_rt_threshold` sec
- **Metric Clustering**: Average-linkage, h = 0.8
- **Best Metrics**: `r paste(best_metrics, collapse = ", ")`
- **Performance**: Precision = `r round(final_perf$precision * 100, 1)`%, Recall = `r round(final_perf$recall * 100, 1)`%, F1 = `r round(final_perf$f1, 3)`

This configuration will be used in the benchmarking document to compare against:
1. RT clustering + EIC correlation
2. CAMERA

