---
title: "Annotation analysis across labs (SIRIUS-curated)"
format: html
---

Compare the manually curated SIRIUS annotations across labs. Each lab has a
curated table at `3_annotation_manual/HE/3_Sirius_curation/{lab}_lab_annotations_sirius_curated.xlsx`.
Compound-level metadata lives in `1_preprocessing/standards.xlsx`.

So this is using annotation that was done on the automatically annotated peaks. no RT restriction. No correspondence was done between the labs.

### Source table structure

Each curated `.xlsx` is one row per **(compound, lab)**. Within that row:

- `chrom_peak_id` and `adduct` are pipe-separated, length-matched lists
  (one entry per peak picked for that compound in that lab — typically
  one peak per adduct).
- `rtmed` is a **single value per (compound, lab)** — the median RT of
  the curated peaks for that compound in that lab. It does not vary
  per peak or per adduct.
- `mz_[adduct]` columns give the theoretical m/z for each adduct.
- `best_structurePerIdRank` is the SIRIUS rank of the curated structure
  in SIRIUS's candidate list (1 = top hit), one value per
  (compound, lab).
- `ms2_count` is the number of MS2 spectra acquired for the curated
  peaks of that (compound, lab). Carried only for diagnostics — see
  the NA-rank breakdown below.

This whole analysis is restricted to **rows where SIRIUS actually
annotated the compound**, i.e. `best_structurePerIdRank` is not NA
(SIRIUS returned the curated standard in its candidate list). Rows
where SIRIUS did not (no MS2, or MS2 present but the standard was not
in the returned candidates) are excluded — we cannot judge an
annotation that SIRIUS did not produce.

After `unnest_lab()` below, `long` is one row per
**(lab, compound, adduct, peak)** with `rtmed` and `ms2_count` repeated
across the peaks of the same (compound, lab).

We look at:

- RT deviation per compound between labs
- m/z deviation per (compound, adduct) between labs
- Which compound categories overlap most between labs
- Per-lab coverage of unique compounds
- Per-lab coverage of adducts
- Whether large RT shifts coincide with worse SIRIUS structure ranks
  (i.e. is the annotation still trustworthy when one lab's RT
  disagrees with the others?)

```{r}
#| label: setup
#| message: false
library(readxl)
library(ggplot2)
library(here)

labs <- c("afekta", "cembio", "hmgu", "icl")
curated_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation")
std_path    <- here("1_preprocessing", "standards.xlsx")
```

## Load curated annotations

Each curated row carries pipe-separated `chrom_peak_id` and `adduct`. We
unnest these so each row is one (lab, compound, adduct, peak) tuple and
attach the matching `mz_{adduct}` column. We also carry `ms2_count`
(MS2 spectra acquired for the curated peaks) — it lives at
(compound, lab) granularity in the source table, so it is repeated
across the peaks of the same (compound, lab) after unnesting.

```{r}
#| label: load-curated
read_lab <- function(lab) {
    f <- file.path(curated_dir,
                   paste0(lab, "_lab_annotations_sirius_curated.xlsx"))
    x <- as.data.frame(read_xlsx(f), check.names = FALSE)
    x$lab <- lab
    x
}

raw <- lapply(labs, read_lab)
names(raw) <- labs

unnest_lab <- function(x) {
    cps <- strsplit(as.character(x$chrom_peak_id), "|", fixed = TRUE)
    ads <- strsplit(as.character(x$adduct),        "|", fixed = TRUE)
    n   <- lengths(cps)
    stopifnot(all(n == lengths(ads)))

    idx <- rep(seq_len(nrow(x)), n)
    out <- data.frame(
        lab           = x$lab[idx],
        compound      = trimws(x$target_ChEBI.name[idx]),
        chrom_peak_id = trimws(unlist(cps, use.names = FALSE)),
        adduct        = trimws(unlist(ads, use.names = FALSE)),
        rtmed         = suppressWarnings(as.numeric(x$rtmed[idx])),
        ms2_count     = suppressWarnings(as.numeric(x$ms2_count[idx])),
        stringsAsFactors = FALSE
    )

    # Attach mz from the matching mz_{adduct} column
    out$mz <- NA_real_
    for (a in unique(out$adduct)) {
        col <- paste0("mz_", a)
        if (col %in% colnames(x)) {
            sel <- out$adduct == a
            row_idx <- idx[sel]
            out$mz[sel] <- suppressWarnings(as.numeric(x[[col]][row_idx]))
        }
    }
    out
}

long_all <- do.call(rbind, lapply(raw, unnest_lab))
long_all <- long_all[!is.na(long_all$compound) & long_all$compound != "", ]
rownames(long_all) <- NULL
head(long_all)
```

SIRIUS reports a **structure rank** for the true standard
(`best_structurePerIdRank`) — the position of the curated structure in
SIRIUS's structural candidate list (1 = top hit). One value per curated
row, so we keep it at (compound, lab) granularity rather than unnesting
per peak. The curated tables do not carry a separate formula rank, so we
work with the structure rank alone. NA = SIRIUS did not return the
curated standard among its structural predictions for that row.

```{r}
#| label: load-rank
rank_lab_all <- do.call(rbind, lapply(raw, function(x) {
    data.frame(
        lab            = x$lab,
        compound       = trimws(x$target_ChEBI.name),
        structure_rank = suppressWarnings(
            as.numeric(x$best_structurePerIdRank)),
        ms2_count      = suppressWarnings(as.numeric(x$ms2_count)),
        stringsAsFactors = FALSE
    )
}))
rank_lab_all <- rank_lab_all[
    !is.na(rank_lab_all$compound) & rank_lab_all$compound != "", ]
rownames(rank_lab_all) <- NULL
head(rank_lab_all)
```

### Where do NA structure ranks come from? SIRIUS filter

`best_structurePerIdRank` is NA when SIRIUS did not return the curated
standard in its candidate list. Two possible reasons: (a) no MS2 was
acquired for the peak, so SIRIUS could not run; (b) MS2 was acquired
and SIRIUS ran, but the curated standard was not in the returned
candidates. Cross-tab to see the split:

```{r}
#| label: na-rank-vs-ms2
rank_lab_all$has_ms2  <- !is.na(rank_lab_all$ms2_count) &
                        rank_lab_all$ms2_count > 0
rank_lab_all$has_rank <- !is.na(rank_lab_all$structure_rank)
addmargins(table(has_ms2  = rank_lab_all$has_ms2,
                 has_rank = rank_lab_all$has_rank))
```

Either way, both cases mean the row is **not actually a SIRIUS
annotation**. We restrict everything downstream to rows where SIRIUS
returned the curated structure (`has_rank = TRUE`).

```{r}
#| label: sirius-filter
# Compound-level: keep (compound, lab) rows with a SIRIUS rank.
keep_r <- !is.na(rank_lab_all$structure_rank)
rank_lab <- rank_lab_all[keep_r,
    c("lab", "compound", "structure_rank", "ms2_count")]
rownames(rank_lab) <- NULL

# Peak-level: keep rows whose (compound, lab) survives the rank filter.
keys_keep <- unique(rank_lab[, c("compound", "lab")])
long <- merge(long_all, keys_keep, by = c("compound", "lab"))
rownames(long) <- NULL

cat(sprintf(
    "Kept %d / %d (peak-level) and %d / %d (compound-level) rows where SIRIUS annotated the compound\n",
    nrow(long), nrow(long_all),
    nrow(rank_lab), nrow(rank_lab_all)))
```

## Load standards metadata

We try to detect a "category"-like column in `standards.xlsx`. If none is
present, we fall back to grouping by molecular formula.

```{r}
#| label: load-standards
std <- as.data.frame(
    read_xlsx(std_path, skip = 1, .name_repair = "minimal"),
    check.names = FALSE
)

# Pick a compound-name column
name_col <- intersect(c("ChEBI name", "Name", "compound", "compound_name"),
                      colnames(std))[1]
stopifnot(!is.na(name_col))
std$compound <- trimws(as.character(std[[name_col]]))

# Pick a category-like column if it exists
cat_candidates <- c("Class", "class", "Category", "category",
                    "ChEBI class", "compound_class", "Compound class",
                    "Family", "family", "Group", "group", "Subclass")
cat_col <- intersect(cat_candidates, colnames(std))[1]
if (is.na(cat_col)) {
    # Fallback: use molecular formula as a coarse grouping
    fcol <- intersect(c("formula", "Formula", "Molecular formula"),
                      colnames(std))[1]
    if (!is.na(fcol)) {
        std$category <- as.character(std[[fcol]])
        message("No category column found; using '", fcol,
                "' as fallback category.")
    } else {
        std$category <- NA_character_
        message("No category or formula column found in standards.xlsx.")
    }
} else {
    std$category <- as.character(std[[cat_col]])
    message("Using category column: '", cat_col, "'")
}

std_lookup <- unique(std[, c("compound", "category")])
long <- merge(long, std_lookup, by = "compound", all.x = TRUE)
head(long)
```

## RT deviation between labs

Per compound, take the lab-level `rtmed` (one value per lab) and compute
the spread (sd and range) across labs that detected it.

```{r}
#| label: rt-deviation
rt_lab <- aggregate(rtmed ~ compound + lab, data = long,
                    FUN = function(v) mean(v, na.rm = TRUE))

rt_stats <- aggregate(rtmed ~ compound, data = rt_lab,
                      FUN = function(v) c(n      = length(v),
                                          mean   = mean(v, na.rm = TRUE),
                                          sd     = sd(v,   na.rm = TRUE),
                                          range  = diff(range(v, na.rm = TRUE))))
rt_stats <- do.call(data.frame, rt_stats)
colnames(rt_stats) <- c("compound", "n_labs", "rt_mean", "rt_sd", "rt_range")
rt_stats <- rt_stats[order(-rt_stats$rt_sd), ]
head(rt_stats, 15)
```

```{r}
#| label: rt-deviation-plot
#| fig-width: 7
#| fig-height: 4.5
ggplot(rt_stats[rt_stats$n_labs > 1, ],
       aes(x = n_labs, y = rt_sd)) +
    geom_jitter(width = 0.15, alpha = 0.6) +
    labs(x = "Number of labs detecting compound",
         y = "RT sd across labs (s)",
         title = "Between-lab RT variability per compound") +
    theme_bw()
```

Top compounds with highest RT spread (showing per-lab RT for the worst 10).
The y-axis label includes `n_labs` so you can tell at a glance whether the
spread comes from a 2-, 3-, or 4-lab consensus.

```{r}
#| label: rt-top-plot
#| fig-width: 8
#| fig-height: 5
worst <- head(rt_stats[rt_stats$n_labs > 1, ], 10)
sub <- rt_lab[rt_lab$compound %in% worst$compound, ]
sub <- merge(sub, worst[, c("compound", "n_labs", "rt_range")],
             by = "compound")
sub$label <- sprintf("%s (n=%d, range=%.1fs)",
                     sub$compound, sub$n_labs, sub$rt_range)
# Order labels by rt_range
lvls <- unique(sub$label[order(sub$rt_range)])
sub$label <- factor(sub$label, levels = lvls)

ggplot(sub, aes(x = label, y = rtmed, colour = lab)) +
    geom_point(size = 2) +
    coord_flip() +
    labs(x = NULL, y = "RT (s)",
         title = "Top 10 compounds by RT spread (n = labs detecting)") +
    theme_bw()
```

## m/z deviation between labs

Per (compound, adduct), compute the spread across labs of the observed
`mz`. We convert spread to ppm against the across-lab mean.

```{r}
#| label: mz-deviation
mz_lab <- aggregate(mz ~ compound + adduct + lab, data = long,
                    FUN = function(v) mean(v, na.rm = TRUE))
mz_lab <- mz_lab[!is.na(mz_lab$mz), ]

mz_stats <- aggregate(mz ~ compound + adduct, data = mz_lab,
                      FUN = function(v) c(n     = length(v),
                                          mean  = mean(v, na.rm = TRUE),
                                          sd    = sd(v,   na.rm = TRUE),
                                          range = diff(range(v, na.rm = TRUE))))
mz_stats <- do.call(data.frame, mz_stats)
colnames(mz_stats) <- c("compound", "adduct", "n_labs",
                        "mz_mean", "mz_sd", "mz_range")
mz_stats$ppm <- 1e6 * mz_stats$mz_range / mz_stats$mz_mean
mz_stats <- mz_stats[order(-mz_stats$ppm), ]
head(mz_stats, 15)
```

```{r}
#| label: mz-deviation-plot
#| fig-width: 7
#| fig-height: 4.5
ggplot(mz_stats[mz_stats$n_labs > 1, ],
       aes(x = adduct, y = ppm)) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    labs(x = "Adduct", y = "m/z range across labs (ppm)",
         title = "Between-lab m/z variability per (compound, adduct)") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Compound-category overlap

For each category, count how many compounds are seen by 1, 2, 3 or 4 labs.

```{r}
#| label: cat-overlap
comp_labs <- aggregate(lab ~ compound + category, data = long,
                       FUN = function(v) length(unique(v)))
colnames(comp_labs)[3] <- "n_labs"

cat_overlap <- as.data.frame(
    table(category = comp_labs$category, n_labs = comp_labs$n_labs))
cat_overlap <- cat_overlap[cat_overlap$Freq > 0, ]
head(cat_overlap[order(-cat_overlap$Freq), ], 20)
```

```{r}
#| label: cat-overlap-plot
#| fig-width: 8
#| fig-height: 5
# Order categories by total compound count
cat_totals <- aggregate(Freq ~ category, data = cat_overlap, FUN = sum)
cat_overlap$category <- factor(cat_overlap$category,
    levels = cat_totals$category[order(cat_totals$Freq)])

ggplot(cat_overlap, aes(x = category, y = Freq, fill = n_labs)) +
    geom_col() +
    coord_flip() +
    labs(x = NULL, y = "Number of compounds",
         fill = "Labs detecting",
         title = "Per-category compound counts, split by lab overlap") +
    theme_bw()
```

## Lab coverage of unique compounds

```{r}
#| label: lab-coverage-compounds
comp_per_lab <- aggregate(compound ~ lab, data = long,
                          FUN = function(v) length(unique(v)))
colnames(comp_per_lab)[2] <- "n_compounds"
comp_per_lab
```

```{r}
#| label: lab-coverage-compounds-plot
#| fig-width: 5
#| fig-height: 3.5
ggplot(comp_per_lab, aes(x = lab, y = n_compounds)) +
    geom_col(fill = "steelblue") +
    geom_text(aes(label = n_compounds), vjust = -0.3) +
    labs(x = NULL, y = "Unique compounds",
         title = "Compounds annotated per lab") +
    theme_bw()
```

Presence/absence heatmap (compounds × labs), ordered by how many labs
detected each compound:

```{r}
#| label: presence-heatmap
#| fig-width: 6
#| fig-height: 8
pres <- as.data.frame(table(compound = long$compound, lab = long$lab))
pres$present <- pres$Freq > 0

# Order compounds by total labs
ord <- aggregate(present ~ compound, data = pres, FUN = sum)
ord <- ord[order(ord$present, decreasing = TRUE), ]
pres$compound <- factor(pres$compound, levels = rev(ord$compound))

ggplot(pres, aes(x = lab, y = compound, fill = present)) +
    geom_tile(colour = "white") +
    scale_fill_manual(values = c(`TRUE` = "steelblue", `FALSE` = "grey90")) +
    labs(x = NULL, y = NULL, fill = "Detected",
         title = "Compound × lab presence") +
    theme_bw() +
    theme(axis.text.y = element_text(size = 5))
```

## Lab coverage of adducts

```{r}
#| label: adduct-coverage
ad_lab <- as.data.frame(table(adduct = long$adduct, lab = long$lab))
ad_lab
```

```{r}
#| label: adduct-coverage-plot
#| fig-width: 7
#| fig-height: 4.5
ggplot(ad_lab, aes(x = adduct, y = Freq, fill = lab)) +
    geom_col(position = "dodge") +
    labs(x = "Adduct", y = "Number of (compound, peak) annotations",
         title = "Adduct usage per lab") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Adduct breadth per compound (how many distinct adducts per compound,
per lab):

```{r}
#| label: adduct-breadth-plot
#| fig-width: 6
#| fig-height: 4
ad_per_comp <- aggregate(adduct ~ compound + lab, data = long,
                         FUN = function(v) length(unique(v)))
colnames(ad_per_comp)[3] <- "n_adducts"

ggplot(ad_per_comp, aes(x = lab, y = n_adducts)) +
    geom_boxplot(outlier.shape = NA, fill = "grey90") +
    geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
    labs(x = NULL, y = "Adducts per compound",
         title = "Per-compound adduct breadth, by lab") +
    theme_bw()
```

## SIRIUS structure rank of the true standard

`best_structurePerIdRank` is the position of the curated structure in
SIRIUS's structural candidate list (1 = top hit). Lower is better.
Because we already filtered to rows with a non-NA rank above, every
row in `rank_lab` is a real SIRIUS annotation; the summary describes
the quality of those annotations.

```{r}
#| label: rank-summary
rank_summary <- aggregate(structure_rank ~ lab, data = rank_lab,
    FUN = function(v) c(n         = length(v),
                        median    = median(v, na.rm = TRUE),
                        pct_rank1 = mean(v == 1, na.rm = TRUE) * 100,
                        pct_top3  = mean(v <= 3, na.rm = TRUE) * 100))
rank_summary <- do.call(data.frame, rank_summary)
colnames(rank_summary) <- c("lab", "n", "median_rank",
                            "pct_rank1", "pct_top3")
rank_summary
```

Distribution of ranks per lab, binned (a log-scaled rank axis is
misleading because rank=1 is the modal value — log10(1)=0 squashes the
"good" cases against the floor while exaggerating the long tail).

```{r}
#| label: rank-histogram
#| fig-width: 7
#| fig-height: 4
# Bin ranks: 1 / 2-3 / 4-10 / 11-50 / >50
rank_binned <- rank_lab[!is.na(rank_lab$structure_rank), ]
rank_binned$bin <- cut(rank_binned$structure_rank,
                       breaks = c(0, 1, 3, 10, 50, Inf),
                       labels = c("1", "2-3", "4-10", "11-50", ">50"))

ggplot(rank_binned, aes(x = bin)) +
    geom_bar(fill = "steelblue") +
    facet_wrap(~ lab, scales = "free_y") +
    labs(x = "Structure rank bin",
         y = "Number of compounds",
         title = "Structure rank distribution per lab (binned)") +
    theme_bw()
```

## Structure rank vs RT shift

The question is: **when a lab's RT for a compound disagrees with the
other labs, is the SIRIUS structure rank for that lab still good?**
If high RT shifts come with bad ranks, the curated annotation is
probably wrong (likely a wrong peak picked); if high shifts come with
rank=1, the labs are picking the same compound but their chromatography
runs differently.

### Defining "RT shift"

For each (compound, lab) we use the **leave-one-out (LOO) shift**:

```
loo_shift(compound, lab) = rtmed(compound, lab)
                         - median{rtmed(compound, other labs)}
```

i.e. how far this lab's RT is from the consensus of the *other* labs
that detected the same compound. This is cleaner than
`|rt - mean(all labs)|` because the suspect lab is no longer included
in the reference value.

Caveat — interpretation depends on `n_labs` (how many of the 4 labs
detected the compound):

- `n_labs = 4`: 3-vs-1 — a large LOO shift means the focal lab is an
  outlier vs. a real 3-lab consensus.
- `n_labs = 3`: 2-vs-1 — a large LOO shift means the focal lab differs
  from 2 labs that agree.
- `n_labs = 2`: 1-vs-1 — both labs get the same |LOO shift| and we
  cannot tell which is the outlier. Reported but flagged.

```{r}
#| label: rt-loo-shift
n_labs_per_compound <- aggregate(lab ~ compound, data = rt_lab,
    FUN = function(v) length(unique(v)))
colnames(n_labs_per_compound)[2] <- "n_labs"

rt_shift <- merge(rt_lab, n_labs_per_compound, by = "compound")
rt_shift$loo_shift <- NA_real_
for (cp in unique(rt_shift$compound)) {
    rows <- which(rt_shift$compound == cp)
    rts  <- rt_shift$rtmed[rows]
    if (length(rts) < 2) next
    for (i in seq_along(rows)) {
        rt_shift$loo_shift[rows[i]] <-
            rt_shift$rtmed[rows[i]] - median(rts[-i], na.rm = TRUE)
    }
}
rt_shift$abs_loo_shift <- abs(rt_shift$loo_shift)

rank_rt <- merge(rank_lab,
    rt_shift[, c("compound", "lab", "n_labs",
                 "rtmed", "loo_shift", "abs_loo_shift")],
    by = c("compound", "lab"))
rank_rt <- rank_rt[!is.na(rank_rt$abs_loo_shift), ]
```

### How many labs are in the large-RT-shift cases?

Bin the shifts and break down by `n_labs`. This answers the
"2/4, 3/4, or 4/4 match?" question directly.

```{r}
#| label: rt-shift-by-nlabs
rank_rt$shift_bin <- cut(rank_rt$abs_loo_shift,
    breaks = c(-Inf, 2, 5, 10, Inf),
    labels = c("<=2s", "2-5s", "5-10s", ">10s"))

# (compound, lab) counts by shift bin x n_labs
shift_by_nlabs <- as.data.frame(table(
    n_labs    = rank_rt$n_labs,
    shift_bin = rank_rt$shift_bin))
shift_by_nlabs <- shift_by_nlabs[shift_by_nlabs$Freq > 0, ]
shift_by_nlabs
```

```{r}
#| label: rt-shift-by-nlabs-plot
#| fig-width: 7
#| fig-height: 3.8
ggplot(shift_by_nlabs,
       aes(x = shift_bin, y = Freq, fill = factor(n_labs))) +
    geom_col(position = "dodge") +
    geom_text(aes(label = Freq),
              position = position_dodge(width = 0.9),
              vjust = -0.3, size = 3) +
    labs(x = "|LOO RT shift|", y = "(compound, lab) cases",
         fill = "n_labs detecting",
         title = "How many labs back up each RT-shift bin") +
    theme_bw()
```

### Does a large RT shift mean a bad annotation?

Cross-tabulate shift bin against a coarse rank tier (1 / 2-10 / >10).
If large shifts are concentrated in the worse rank tiers, the
annotation is suspicious; if rank stays at 1 across shift bins, the
labs are calling the same compound on chromatographically different
peaks.

```{r}
#| label: shift-vs-rank
rank_rt$rank_tier <- ifelse(rank_rt$structure_rank == 1,   "1 (top)",
                     ifelse(rank_rt$structure_rank <= 10,  "2-10",
                                                           ">10"))
rank_rt$rank_tier <- factor(rank_rt$rank_tier,
    levels = c("1 (top)", "2-10", ">10"))

shift_rank <- as.data.frame(table(
    shift_bin = rank_rt$shift_bin,
    rank_tier = rank_rt$rank_tier))
shift_rank <- shift_rank[shift_rank$Freq > 0, ]

# As proportions within each shift bin (easier to read)
shift_totals <- aggregate(Freq ~ shift_bin, data = shift_rank, FUN = sum)
shift_rank <- merge(shift_rank, shift_totals, by = "shift_bin",
                    suffixes = c("", "_total"))
shift_rank$pct <- 100 * shift_rank$Freq / shift_rank$Freq_total
shift_rank
```

```{r}
#| label: shift-vs-rank-plot
#| fig-width: 7
#| fig-height: 4
ggplot(shift_rank,
       aes(x = shift_bin, y = pct, fill = rank_tier)) +
    geom_col() +
    geom_text(aes(label = ifelse(pct >= 5,
                                 sprintf("%.0f%%", pct), "")),
              position = position_stack(vjust = 0.5), size = 3) +
    labs(x = "|LOO RT shift|", y = "% of (compound, lab) cases",
         fill = "Structure rank",
         title = "Annotation quality (rank tier) by RT shift magnitude") +
    theme_bw()
```

### Continuous view: rank vs RT shift, weighted by n_labs

Same `rank_rt` data, but the (RT shift, rank) relationship as a scatter
rather than bins. Point size encodes how many labs detected the
compound — a 4-lab shift is a stronger signal than a 2-lab one
because the cross-lab consensus is bigger to disagree with.

```{r}
#| label: shift-vs-rank-scatter
#| fig-width: 8
#| fig-height: 5
ggplot(rank_rt,
       aes(x = abs_loo_shift, y = structure_rank,
           colour = lab, size = factor(n_labs))) +
    geom_jitter(width = 0, height = 0.15, alpha = 0.7) +
    scale_size_manual(values = c("2" = 1.5, "3" = 2.5, "4" = 3.5)) +
    labs(x = "|RT - cross-lab mean RT| (s)",
         y = "Structure rank",
         colour = "Lab",
         size = "n_labs detecting",
         title = "Structure rank vs RT deviation across labs",
         subtitle = "Point size = labs detecting the compound (2/3/4)") +
    theme_bw()
```

### Worst RT-shift cases — context per compound

For the 10 (compound, lab) tuples with the highest LOO shift, show the
RT of *every* lab that detected the compound. Triangle = the focal
(shifted) lab. The header includes `n_labs` and the focal lab's
structure rank so you can decide at a glance whether the shift is one
lab disagreeing with 2 or 3 others, and whether SIRIUS still thinks
the annotation is correct.

```{r}
#| label: worst-rtshift-context
#| fig-width: 9
#| fig-height: 5
top_shift <- head(rank_rt[order(-rank_rt$abs_loo_shift), ], 10)
top_shift$label <- with(top_shift, sprintf(
    "%s | focal=%s | n=%d | rank=%d | shift=%.1fs",
    compound, lab, n_labs, structure_rank, loo_shift))

# All lab RTs for those compounds, plus a flag for the focal lab.
panel <- merge(
    top_shift[, c("compound", "lab", "label")],
    rt_lab,
    by = "compound",
    suffixes = c("_focal", ""))
panel$is_focal <- panel$lab == panel$lab_focal
panel$label <- factor(panel$label, levels = rev(top_shift$label))

ggplot(panel, aes(x = rtmed, y = label,
                  colour = lab, shape = is_focal, size = is_focal)) +
    geom_point() +
    scale_shape_manual(values = c(`TRUE` = 17, `FALSE` = 16),
                       guide = "none") +
    scale_size_manual(values = c(`TRUE` = 3.5, `FALSE` = 2.5),
                      guide = "none") +
    labs(x = "RT (s)", y = NULL, colour = "lab",
         title = "Top 10 LOO-RT-shift cases: per-lab RT context",
         subtitle = "Triangle = focal (shifted) lab; header gives n_labs and SIRIUS rank") +
    theme_bw()
```

Read this plot as follows:

- If the triangle sits well apart from the round points (the consensus
  of the other labs) **and** the rank in the header is 1, the focal
  lab probably picked the right compound on a chromatographically
  shifted peak — annotation likely OK.
- If the triangle is apart **and** the rank is high (e.g. >10), the
  focal lab's annotation is suspicious — likely a mis-picked peak.
- If `n_labs = 2`, both labs are equidistant in LOO-space; the
  triangle and the round point will just be far apart and we cannot
  decide which is the outlier — use this only as a flag.

## Annotation confidence via the consensus peak table

The consensus peak table from `3_peak_feature_calculation_detected.qmd`
(`ct_all_samples.csv`) groups peaks across labs into consensus peaks per
sample, using m/z (20 ppm) and RT (15 s) tolerances on the unfiltered
detected-peak set. We use the **unfiltered** table — not the 4-lab subset
`ct_filtered4.csv` — so consensus peaks where only 2 or 3 labs converged
still count.

For each curated `(sample, compound, adduct)` tuple we ask:

- Does the annotation converge to a single consensus peak (labs picked
  the same peak), or is it **split** across multiple consensus peaks?
- For each consensus peak that received the annotation, how many labs
  annotated it identically, vs. how many detected the underlying peak?

This is a peak-level confidence layer that is independent of the SIRIUS
rank/confidence — it measures whether the peaks the labs *picked* for a
given (compound, adduct) are the same peak.

### Load and pivot the consensus table

```{r}
#| label: load-consensus
ct_path <- here("5_downstream_analysis", "object", "ct_all_samples.csv")
ct_wide <- read.csv(ct_path, stringsAsFactors = FALSE)

ct_long <- do.call(rbind, lapply(labs, function(l) {
    col  <- paste0("chrom_peak_id_", l)
    keep <- !is.na(ct_wide[[col]])
    data.frame(
        consensus_id   = ct_wide$consensus_id[keep],
        sample         = ct_wide$sample[keep],
        n_labs_matched = ct_wide$n_labs_matched[keep],
        lab            = l,
        chrom_peak_id  = as.character(ct_wide[[col]][keep]),
        stringsAsFactors = FALSE
    )
}))
head(ct_long)
```

### Join annotations to consensus peaks

```{r}
#| label: join-annotations-consensus
ann_ct <- merge(long, ct_long,
                by = c("lab", "chrom_peak_id"),
                all.x = TRUE)

n_total   <- nrow(ann_ct)
n_matched <- sum(!is.na(ann_ct$consensus_id))
cat(sprintf("Annotated peaks matched to a consensus group: %d / %d (%.1f%%)\n",
            n_matched, n_total, 100 * n_matched / n_total))
```

Annotated peaks with no consensus match either fell outside the 20 ppm ×
15 s tolerance from any other lab's peak, or come from a sample for which
only one lab detected the peak. A low match rate here means the rest of
the section can only speak to a fraction of the curated annotations — it
is the first thing to check.

### Annotation splitting per (compound, adduct)

For each `(sample, compound, adduct)` tuple, count the distinct consensus
peaks the annotation was assigned to. **1** = labs converged on the same
peak; **>1** = the annotation is split across peaks.

```{r}
#| label: annotation-splitting
ann_matched <- ann_ct[!is.na(ann_ct$consensus_id), ]

split_stats <- aggregate(consensus_id ~ sample + compound + adduct,
                         data = ann_matched,
                         FUN  = function(v) length(unique(v)))
colnames(split_stats)[4] <- "n_consensus_peaks"

labs_anno <- aggregate(lab ~ sample + compound + adduct,
                       data = ann_matched,
                       FUN  = function(v) length(unique(v)))
colnames(labs_anno)[4] <- "n_labs_annotating"

split_stats <- merge(split_stats, labs_anno,
                     by = c("sample", "compound", "adduct"))

table(n_labs_annotating = split_stats$n_labs_annotating,
      n_consensus_peaks = split_stats$n_consensus_peaks)
```

The first column (`n_consensus_peaks == 1`) is the best case — labs
converged. The diagonal (`n_labs_annotating == n_consensus_peaks`) is the
worst — every annotating lab picked a different peak.

```{r}
#| label: split-plot
#| fig-width: 6
#| fig-height: 4
split_summary <- as.data.frame(table(
    n_labs_annotating = split_stats$n_labs_annotating,
    n_consensus_peaks = split_stats$n_consensus_peaks))
split_summary <- split_summary[split_summary$Freq > 0, ]

ggplot(split_summary,
       aes(x = n_labs_annotating, y = Freq, fill = n_consensus_peaks)) +
    geom_col() +
    labs(x = "Labs annotating the (compound, adduct)",
         y = "Number of (sample, compound, adduct) tuples",
         fill = "Distinct consensus\npeaks assigned",
         title = "Annotation convergence vs. number of annotating labs") +
    theme_bw()
```

### Cross-lab agreement per consensus peak

For each consensus peak that received ≥1 annotation, count the labs that
annotated it with the same `(compound, adduct)`, and compare to how many
labs *detected* the peak (`n_labs_matched`). A peak detected by 4 labs
but annotated by 1 is much weaker evidence than a peak detected by 3
labs and annotated by all 3.

```{r}
#| label: consensus-agreement
peak_agree <- aggregate(lab ~ consensus_id + compound + adduct +
                              sample + n_labs_matched,
                        data = ann_matched,
                        FUN  = function(v) length(unique(v)))
colnames(peak_agree)[6] <- "n_labs_annotating"

table(detected  = peak_agree$n_labs_matched,
      annotated = peak_agree$n_labs_annotating)
```

### Confidence tiers

For each `(sample, compound, adduct)`, pick the consensus peak with the
highest annotation count and assign a tier.

```{r}
#| label: confidence-tiers
peak_agree_sorted <- peak_agree[
    order(peak_agree$sample, peak_agree$compound, peak_agree$adduct,
          -peak_agree$n_labs_annotating), ]
key <- with(peak_agree_sorted,
            paste(sample, compound, adduct, sep = "|"))
peak_agree_best <- peak_agree_sorted[!duplicated(key), ]

# Drop n_labs_annotating from split_stats (it counts labs per
# compound/adduct regardless of peak); keep the peak_agree_best one,
# which counts labs converging on the *same* consensus peak.
split_stats_for_merge <- split_stats[, setdiff(colnames(split_stats),
                                               "n_labs_annotating")]

tiered <- merge(split_stats_for_merge,
                peak_agree_best[, c("sample", "compound", "adduct",
                                    "n_labs_matched", "n_labs_annotating",
                                    "consensus_id")],
                by = c("sample", "compound", "adduct"))

tiered$tier <- with(tiered, ifelse(
    n_consensus_peaks >  1,                              "C: split across peaks",
    ifelse(n_labs_annotating >= 3,                       "A: converged, >=3 labs",
    ifelse(n_labs_annotating == 2,                       "B: converged, 2 labs",
                                                         "D: single lab"))))

as.data.frame(table(tier = tiered$tier))
```

```{r}
#| label: confidence-tiers-plot
#| fig-width: 6
#| fig-height: 3.5
tier_tab <- as.data.frame(table(tier = tiered$tier))
ggplot(tier_tab, aes(x = tier, y = Freq, fill = tier)) +
    geom_col() +
    geom_text(aes(label = Freq), vjust = -0.3) +
    labs(x = NULL, y = "Number of (sample, compound, adduct) tuples",
         title = "Annotation confidence tiers") +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 20, hjust = 1))
```

Tier A annotations (≥3 labs converging on the same consensus peak) are
the strongest evidence that the curated identification is correct.
Tier C splits indicate labs all chose to annotate a compound but
disagreed on which peak it corresponds to — likely co-elution, isomers,
or RT/m/z misalignment beyond the consensus tolerance.

### Compounds with the most confident annotations

```{r}
#| label: confident-compounds
tier_A <- tiered[tiered$tier == "A: converged, >=3 labs", ]
conf_per_compound <- aggregate(tier ~ compound, data = tier_A,
                               FUN = length)
colnames(conf_per_compound)[2] <- "n_tier_A_tuples"
conf_per_compound <- conf_per_compound[
    order(-conf_per_compound$n_tier_A_tuples), ]
head(conf_per_compound, 20)
```

### Compounds whose annotations are most often split

```{r}
#| label: split-compounds
tier_C <- tiered[tiered$tier == "C: split across peaks", ]
split_per_compound <- aggregate(n_consensus_peaks ~ compound, data = tier_C,
                                FUN = function(v) c(n = length(v),
                                                    max_split = max(v)))
split_per_compound <- do.call(data.frame, split_per_compound)
colnames(split_per_compound) <- c("compound", "n_split_tuples", "max_split")
split_per_compound <- split_per_compound[
    order(-split_per_compound$n_split_tuples), ]
head(split_per_compound, 20)
```
