---
title: "Confidence tiering — populate `annotation_identity.csv` (Phase 5)"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
        embed-resources: true
---

# Goal

Populate the `confidence_tier` column in `annotation_identity.csv`,
overwriting the P1 `"unfiltered"` placeholder with A / B / C / D tiers
based on:

1. **COSMIC score** (within-lab confidence): the producer's
   `confidenceExactMatch` from SIRIUS. Each lab assigns a score per
   annotated compound; we use the **max across labs** per
   `(mixture, compound, polarity)` so a compound flagged confidently
   by *any* lab gets credit.

2. **MS1-RT concordance** (between-lab confidence): SD of P2-aligned
   `apex_rt` across labs that have a measurable peak (either `own` =
   SIRIUS-annotated, or `consensus` = MS1-rescued from cross-lab
   median). Small SD → labs agree on chromatographic position →
   high cross-lab confidence in the identification.

This column is overwritten in place in `annotation_identity.csv`.
Downstream consumers (`8_per_compound_diagnostics.qmd`,
`7_reproducibility_results.qmd`) can then filter by `confidence_tier %in% c("A",
"B")` to restrict to confident identifications.

## Tier definition

- **A**: COSMIC ≥ 0.5 **and** aligned-RT SD ≤ 3 s. Both axes strong.
- **B**: COSMIC ≥ 0.5 **and** aligned-RT SD ≤ 10 s, or COSMIC ≥ 0.1
  **and** aligned-RT SD ≤ 3 s. One axis strong, one marginal.
- **C**: anything else with both axes computable (some weight on
  both). Treat as low confidence.
- **D**: missing one axis (NA COSMIC or undetermined RT SD because <2
  labs detect / aligned). Treat as not-defensible.

Thresholds are at the top of the setup chunk; the producer report
shows tier counts, so they can be tuned by editing and re-running.

## Setup

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

producer_dir <- here("5_downstream_analysis", "object")
object_dir   <- here("5_downstream_analysis", "object")

# Tier thresholds. Adjust here and re-run to retune.
cosmic_strong  <- 0.5
cosmic_marginal <- 0.1
rt_sd_strong   <- 3   # seconds (aligned)
rt_sd_marginal <- 10  # seconds (aligned)

# Minimum number of labs whose aligned_rt is used to compute the SD.
# Setting <2 makes RT SD undefined.
min_labs_for_rt_sd <- 2L
```

## 1. Load inputs

```{r load-inputs}
# Identity table (one row per (mixture, lab, compound, polarity), with COSMIC).
# Read mixture as character — the canonical "2.10" trap.
identity_df <- read.csv(
  file.path(producer_dir, "annotation_identity.csv"),
  stringsAsFactors = FALSE,
  colClasses = c(mixture = "character"))
identity_df$mixture <- gsub("\\.", "_", identity_df$mixture)

cat("identity_df rows:", nrow(identity_df), "\n")

# Aligned chrom metrics from the P3 producer. Has aligned_rt.
chrom_norm <- read.csv(
  file.path(object_dir, "annotation_chrom_metrics_normalized.csv"),
  stringsAsFactors = FALSE,
  colClasses = c(mixture = "character"))
chrom_norm$mixture <- gsub("\\.", "_", chrom_norm$mixture)

cat("chrom_norm rows:", nrow(chrom_norm),
    " | with aligned_rt:", sum(!is.na(chrom_norm$aligned_rt)), "\n")
```

## 2. COSMIC axis: max score per `(mixture, compound, polarity)`

```{r cosmic-axis}
cosmic_max <- identity_df |>
  group_by(mixture, compound, polarity) |>
  summarise(cosmic_max = suppressWarnings(
              max(confidenceExactMatch, na.rm = TRUE)),
            cosmic_n_labs = sum(!is.na(confidenceExactMatch)),
            .groups = "drop") |>
  mutate(cosmic_max = ifelse(is.infinite(cosmic_max), NA_real_, cosmic_max))

cat("Tuples with a COSMIC score: ",
    sum(!is.na(cosmic_max$cosmic_max)), " / ", nrow(cosmic_max), "\n",
    sep = "")
```

## 3. RT-concordance axis: SD of aligned_rt across labs

```{r rt-axis}
# Per (mixture, compound, polarity), collapse to one apex_rt per lab
# (max-maxIntensity peak across adducts/sources), then SD across labs.
rt_concordance <- chrom_norm |>
  filter(!is.na(aligned_rt), !is.na(maxIntensity), maxIntensity > 0) |>
  group_by(mixture, compound, polarity, lab) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup() |>
  group_by(mixture, compound, polarity) |>
  summarise(rt_n_labs = n_distinct(lab),
            rt_sd_aligned = if (n_distinct(lab) >= min_labs_for_rt_sd)
                               sd(aligned_rt, na.rm = TRUE)
                            else NA_real_,
            .groups = "drop")

cat("Tuples with computable RT SD (>=", min_labs_for_rt_sd, "labs aligned): ",
    sum(!is.na(rt_concordance$rt_sd_aligned)), " / ", nrow(rt_concordance),
    "\n", sep = "")
```

## 4. Assemble + apply tier logic

```{r tier-assign}
combined_axes <- cosmic_max |>
  full_join(rt_concordance,
            by = c("mixture", "compound", "polarity")) |>
  mutate(tier = case_when(
    is.na(cosmic_max) | is.na(rt_sd_aligned)                             ~ "D",
    cosmic_max >= cosmic_strong   & rt_sd_aligned <= rt_sd_strong        ~ "A",
    cosmic_max >= cosmic_strong   & rt_sd_aligned <= rt_sd_marginal      ~ "B",
    cosmic_max >= cosmic_marginal & rt_sd_aligned <= rt_sd_strong        ~ "B",
    TRUE                                                                  ~ "C"))

cat("Tier counts (per mixture × compound × polarity):\n")
combined_axes |>
  count(polarity, tier) |>
  pivot_wider(names_from = tier, values_from = n, values_fill = 0L,
              names_sort = TRUE) |>
  knitr::kable(caption = "Tier counts per polarity. A = both axes strong; B = one strong + one marginal; C = both marginal-to-weak but defined; D = at least one axis undefined.")
```

```{r tier-vs-axes}
#| fig-cap: "Tier assignment in COSMIC × aligned-RT-SD space. Vertical dashed lines = COSMIC thresholds; horizontal = RT SD thresholds. Coloured by tier."
#| fig-height: 5
# rt_sd_aligned == 0 (single-lab / perfect agreement) is undefined on the
# log y-axis and needs >= 2 labs to be meaningful — drop it here.
ggplot(combined_axes |> filter(!is.na(cosmic_max), rt_sd_aligned > 0),
       aes(x = cosmic_max, y = rt_sd_aligned, colour = tier)) +
  geom_point(size = 1.5, alpha = 0.6) +
  geom_vline(xintercept = c(cosmic_marginal, cosmic_strong),
             linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = c(rt_sd_strong, rt_sd_marginal),
             linetype = "dashed", colour = "grey50") +
  facet_wrap(~ polarity) +
  scale_y_log10() +
  scale_colour_manual(values = c(A = "#1F77B4", B = "#2CA02C",
                                  C = "#FF7F0E", D = "#888888")) +
  labs(x = "max COSMIC score across labs",
       y = "aligned-RT SD across labs (s, log)",
       colour = "Tier",
       title = "Confidence tier vs the two axes",
       subtitle = sprintf("Thresholds: COSMIC %.1f / %.1f; RT SD %d / %d s",
                          cosmic_strong, cosmic_marginal,
                          rt_sd_strong, rt_sd_marginal)) +
  theme_bw() +
  theme(legend.position = "top")
```

## 5. Overwrite `confidence_tier` in `annotation_identity.csv`

`confidence_tier` is updated in place in `annotation_identity.csv`.
The same `(mixture, compound, polarity)` tuple gets the same tier
across all lab rows.

```{r overwrite-confidence}
identity_df_tiered <- identity_df |>
  select(-confidence_tier) |>
  left_join(combined_axes |>
              select(mixture, compound, polarity, tier),
            by = c("mixture", "compound", "polarity")) |>
  rename(confidence_tier = tier) |>
  # Defensive: any join misses default back to D so the column has no NAs.
  mutate(confidence_tier = coalesce(confidence_tier, "D"))

stopifnot(all(c("confidence_tier") %in% names(identity_df_tiered)))
stopifnot(nrow(identity_df_tiered) == nrow(identity_df))

# Re-order columns to match the original identity-table schema.
identity_df_tiered <- identity_df_tiered |>
  select(mixture, lab, compound, inchikey_2d, polarity,
         best_formula_rank, best_structure_rank, adducts,
         molecularFormula, expected_formula, expected_inchikey,
         mz_observed, confidenceExactMatch, confidenceApproxMatch,
         confidence_tier)

write.csv(identity_df_tiered,
          file.path(producer_dir, "annotation_identity.csv"),
          row.names = FALSE)

cat("Overwrote: ",
    file.path(producer_dir, "annotation_identity.csv"),
    "\n  rows: ", nrow(identity_df_tiered), "\n", sep = "")
```

## 6. Report — per-lab tier coverage

```{r tier-by-lab}
identity_df_tiered |>
  count(lab, polarity, confidence_tier) |>
  pivot_wider(names_from = confidence_tier, values_from = n,
              values_fill = 0L, names_sort = TRUE) |>
  knitr::kable(caption = "Tier counts per (lab, polarity) on the identity table itself. Same tier value is replicated across all lab rows for the same (mixture, compound, polarity) tuple.")
```

# Notes

- Re-running this file is **idempotent** — it always recomputes tiers
  from the same axes and overwrites the column.
- If thresholds change (top of setup), the producer's tier counts in
  this report show whether you're in a sensible regime. Tier A should
  not be empty; if it is, thresholds are too strict.
- `confidence_tier = "D"` rows are compounds where one of the two axes
  could not be computed (e.g. < 2 labs detect → no RT SD; COSMIC is
  NA). These are not "low confidence" so much as "not-defensible";
  downstream code should treat them as "not-yet-graded" rather than
  filtering them out as bad.
- Consumers that previously read `annotation_identity.csv` with
  `filter(confidence_tier == "unfiltered")` will now see no rows
  (placeholder gone) — switch them to `filter(confidence_tier %in%
  c("A", "B"))` for confident-only, or remove the filter to keep all.
