---
title: "Annotated abundance ICC and cross-lab analysis (Phase 3 consumer)"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
---

# Goal

Consume `annotation_chrom_metrics_normalized.csv` produced by
`5_normalization.qmd` and compute the cross-lab
reproducibility analyses:

- **§P3.5** — Bulk ICC ladder under five strategies × two metrics ×  two compound-set policies (Strict / Inclusive).
- **§P3.5.c** — Side analysis on the quorum-3+/4 compound set.
- **§P3.6** — Pairwise ICC across all 6 lab pairs, including the
  Agilent same-platform pair as the ceiling reference.
- **§P3.6.b–g** — Per-lab diagnostic chunks: rank/log correlation
  vs the cross-lab consensus, leave-one-out consensus, per-compound
  hmgu deep-dive, 3-lab ICC excluding hmgu, and pairwise
  per-compound correlations.
- **§P3.7** — Per-(compound, mixture) cross-lab CV distribution.

The producer file builds the normalized annotated table. This file
reads it and asks the methodological + reproducibility questions.

## Setup

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
base_path    <- here("5_downstream_analysis")
object_dir   <- file.path(base_path, "object")
producer_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                     "results_all_ms2")

source(file.path(base_path, "global_analysis_helpers.R"))

# Strategy palette and names — must match the producer's apply_and_export
# columns. "Lab-median" is computed post-hoc below from the bulk
# distribution and added as a 6th strategy.
strategy_palette <- c(Raw          = "#E64B35",
                      Aligned      = "#999999",
                      `Lab-median` = "#3B7C92",
                      Global       = "#4DBBD5",
                      Decile       = "#70AD47",
                      LOESS        = "#4472C4")

strategy_names <- c("Raw", "Aligned", "Lab-median",
                    "Global", "Decile", "LOESS")
metric_names   <- c("maxIntensity", "areaUnderTic")
```

```{r load-normalized}
# Read the producer's normalized annotated table. mixture must be
# character — read.csv would otherwise coerce "2.10" → 2.1.
ann_chrom <- read.csv(
  file.path(object_dir, "annotation_chrom_metrics_normalized.csv"),
  stringsAsFactors = FALSE,
  colClasses = c(mixture = "character"))

cat("ann_chrom rows:", nrow(ann_chrom),
    " | columns:", ncol(ann_chrom), "\n")
```

# Build compound-set subsets

The five strategies in §P3.5 score the abundance metric over three
compound sets. We compute them here so the downstream chunks are tidy.

```{r build-subsets}
# Well-detected set, on raw maxIntensity (alignment-agnostic — matches
# the Annotated level definition in 8_data_level_progression.qmd §1.2).
well_detected <- ann_chrom |>
  filter(!is.na(maxIntensity), maxIntensity > 0) |>
  distinct(mixture, lab, compound, polarity) |>
  group_by(mixture, compound, polarity) |>
  filter(n_distinct(lab) == 4L) |>
  ungroup() |>
  distinct(mixture, compound, polarity)

# Per (lab, mixture, polarity, compound) dominant peak (max maxIntensity
# across adduct / source). Mirrors 2_global's collapse before sum.
ann_well <- ann_chrom |>
  inner_join(well_detected, by = c("mixture", "compound", "polarity")) |>
  group_by(lab, mixture, polarity, compound) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup()

# Alignment-restricted subset: require aligned_rt for all 4 labs.
well_aligned <- ann_well |>
  filter(!is.na(aligned_rt)) |>
  count(mixture, compound, polarity, name = "n_labs_aligned") |>
  filter(n_labs_aligned == 4L) |>
  select(mixture, compound, polarity)

ann_well_aligned <- ann_well |>
  inner_join(well_aligned, by = c("mixture", "compound", "polarity"))

cat("Well-detected (mixture, compound, polarity) tuples:",
    nrow(well_detected), "\n")
cat("In-range subset (all 4 labs aligned):", nrow(well_aligned), "\n")
cat("Compounds with at least one lab out of anchor range:",
    nrow(well_detected) - nrow(well_aligned), "\n")
```

```{r alignment-quorum-diag}
# Per (mixture, compound, polarity), how many of the 4 labs had a usable
# aligned_rt. The 1-lab and 2-lab rows are compounds we can't trust on
# the consensus axis. The 3-lab and 4-lab rows are candidates to keep in
# different ways (see Strict vs Inclusive below).
align_quorum <- ann_well |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs_aligned = sum(!is.na(aligned_rt)),
            .groups = "drop")

align_quorum |>
  count(polarity, n_labs_aligned) |>
  pivot_wider(names_from = n_labs_aligned, values_from = n,
              names_prefix = "n_aligned=", values_fill = 0L) |>
  knitr::kable(caption = "Per-compound alignment quorum.")
```

# P3.5 ICC ladder under alignment + normalization strategies

**Data level: Annotated Signal** (same as `8_data_level_progression.qmd`'s
Annotated row). Per-`(lab, mixture, polarity)` bulk metric =
sum of an abundance column over the **well-detected set**.

Two abundance metrics scored in parallel:

- `maxIntensity` — sum of peak heights. Scan-rate robust.
- `areaUnderTic` — sum of peak areas. Standard TIC bulk; directly
  comparable to 2_global's Annotated TIC ICC.

Two compound-set policies:

- **Strict** — compounds where **all 4 labs** are in the anchor range.
  Lab-balanced and trustworthy on every row; loses 39% of well-detected.
- **Inclusive** — all well-detected compounds. For the lab(s) where
  alignment failed, the row falls back to `factor_global`. Keeps the
  full set; the unaligned lab(s) get a weaker correction.

```{r build-bulk-icc-strategies}
make_bulk_for <- function(metric_col, mode = c("strict", "inclusive")) {
  mode       <- match.arg(mode)
  global_col <- paste0(metric_col, "_global")
  decile_col <- paste0(metric_col, "_decile")
  loess_col  <- paste0(metric_col, "_loess")

  factor_set <- if (mode == "strict") ann_well_aligned else ann_well
  set_label  <- if (mode == "strict")
                  "Strict (4/4 labs aligned, n = strict_n)"
                else
                  "Inclusive (all well-detected, n = incl_n, global fallback)"

  bulk_raw <- ann_well |>
    group_by(lab, mixture, polarity) |>
    summarise(bulk = sum(.data[[metric_col]], na.rm = TRUE),
              .groups = "drop") |>
    mutate(strategy = "Raw")

  bulk_aligned <- if (mode == "strict") {
    ann_well_aligned |>
      group_by(lab, mixture, polarity) |>
      summarise(bulk = sum(.data[[metric_col]], na.rm = TRUE),
                .groups = "drop") |>
      mutate(strategy = "Aligned")
  } else NULL

  bulk_norm <- factor_set |>
    mutate(int_global = .data[[global_col]],
           int_decile = coalesce(.data[[decile_col]], .data[[global_col]]),
           int_loess  = coalesce(.data[[loess_col]],  .data[[global_col]])) |>
    group_by(lab, mixture, polarity) |>
    summarise(Global = sum(int_global, na.rm = TRUE),
              Decile = sum(int_decile, na.rm = TRUE),
              LOESS  = sum(int_loess,  na.rm = TRUE),
              .groups = "drop") |>
    pivot_longer(c(Global, Decile, LOESS),
                 names_to = "strategy", values_to = "bulk")

  bind_rows(bulk_raw, bulk_aligned, bulk_norm) |>
    mutate(metric = metric_col, alignment_set = set_label)
}

bulk_all <- bind_rows(
  make_bulk_for("maxIntensity", "strict"),
  make_bulk_for("areaUnderTic", "strict"),
  make_bulk_for("maxIntensity", "inclusive"),
  make_bulk_for("areaUnderTic", "inclusive")) |>
  mutate(alignment_set = gsub("strict_n", as.character(nrow(well_aligned)),
                              alignment_set),
         alignment_set = gsub("incl_n",   as.character(nrow(well_detected)),
                              alignment_set))

# Constants used by downstream filters — recomputed from the actual n so
# the chunks below adapt to whatever the producer emitted (used to be
# hard-coded "n = 253" strings).
strict_label    <- sprintf("Strict (4/4 labs aligned, n = %d)",
                           nrow(well_aligned))
inclusive_label <- sprintf("Inclusive (all well-detected, n = %d, global fallback)",
                           nrow(well_detected))
cat("strict_label    =", strict_label,    "\n",
    "inclusive_label =", inclusive_label, "\n")

# Add "Lab-median" strategy on top — 2_global's per-lab-median formulation,
# computed from the bulk distribution itself.
lab_med_per_set <- bulk_all |>
  filter(strategy == "Raw") |>
  group_by(alignment_set, metric, lab, polarity) |>
  summarise(med_bulk = median(bulk, na.rm = TRUE), .groups = "drop")

bulk_labmed <- bulk_all |>
  filter(strategy == "Raw") |>
  left_join(lab_med_per_set,
            by = c("alignment_set", "metric", "lab", "polarity")) |>
  mutate(bulk     = bulk / med_bulk,
         strategy = "Lab-median") |>
  select(-med_bulk)

bulk_all <- bind_rows(bulk_all, bulk_labmed)

# ICC per (metric, polarity, strategy, alignment_set).
icc_all <- bind_rows(lapply(metric_names, function(m) {
  bind_rows(lapply(unique(bulk_all$alignment_set), function(aset) {
    bind_rows(lapply(c("pos", "neg"), function(pol_str) {
      pol_name <- if (pol_str == "pos") "Positive" else "Negative"
      bind_rows(lapply(strategy_names, function(s) {
        df <- bulk_all |>
          filter(metric == m, polarity == pol_str,
                 alignment_set == aset, strategy == s)
        if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
        res <- compute_icc(df, "bulk")
        res$strategy      <- s
        res$polarity      <- pol_name
        res$metric        <- m
        res$alignment_set <- aset
        res
      }))
    }))
  }))
}))

icc_all |>
  transmute(alignment_set, metric, polarity, strategy,
            ICC            = round(ICC, 3),
            `Var(between)` = round(Between_mixture_var, 2),
            `Var(within)`  = round(Within_mixture_var, 2)) |>
  arrange(alignment_set, metric, polarity,
          factor(strategy, levels = strategy_names)) |>
  knitr::kable(caption = "Annotated-Signal bulk ICC per alignment_set × metric × polarity × strategy.")
```

```{r icc-bar-plot}
#| fig-cap: "Annotated abundance ICC across the alignment+normalization ladder × both metrics × both compound-set policies."
#| fig-height: 12
icc_all |>
  mutate(strategy = factor(strategy, levels = strategy_names)) |>
  ggplot(aes(x = strategy, y = ICC, fill = strategy)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = sprintf("%.3f", ICC)),
            vjust = -0.4, size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  facet_grid(alignment_set + metric ~ polarity, switch = "y") +
  scale_fill_manual(values = strategy_palette) +
  ylim(NA, 1) +
  labs(x = NULL, y = "ICC (bulk over compound subset)",
       title = "ICC ladder — alignment + normalization × metric × policy") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.y.left = element_text(size = 8))
```

```{r icc-variance-plot}
#| fig-cap: "Variance decomposition per strategy × metric × alignment_set."
#| fig-height: 12
icc_all |>
  mutate(strategy = factor(strategy, levels = strategy_names)) |>
  select(metric, polarity, strategy, alignment_set,
         Between_mixture_var, Within_mixture_var) |>
  pivot_longer(c(Between_mixture_var, Within_mixture_var),
               names_to = "source", values_to = "variance") |>
  mutate(source = recode(source,
                         Between_mixture_var = "Between mixtures (signal)",
                         Within_mixture_var  = "Within mixtures (lab noise)")) |>
  ggplot(aes(x = strategy, y = variance, fill = source)) +
  geom_col(position = "stack", alpha = 0.85) +
  facet_grid(alignment_set + metric ~ polarity,
             scales = "free_y", switch = "y") +
  scale_fill_manual(values = c(`Between mixtures (signal)`     = "#619CFF",
                               `Within mixtures (lab noise)`   = "#F8766D")) +
  labs(x = NULL, y = "Variance (log2)", fill = NULL,
       title = "Variance decomposition") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.y.left = element_text(size = 8))
```

## P3.5.c Option B sidebar — quorum-3+/4 with global fallback

For each compound, require at least 3 of 4 labs aligned. The lab(s) that
didn't align get `factor_global` fallback.

```{r icc-quorum-3plus}
well_3plus <- ann_well |>
  group_by(mixture, compound, polarity) |>
  filter(sum(!is.na(aligned_rt)) >= 3L) |>
  ungroup() |>
  distinct(mixture, compound, polarity)

ann_well_3plus <- ann_well |>
  inner_join(well_3plus, by = c("mixture", "compound", "polarity"))

cat("Quorum ≥3/4 compounds:", nrow(well_3plus),
    " (Strict 253, Inclusive 418)\n")

make_bulk_3plus <- function(metric_col) {
  global_col <- paste0(metric_col, "_global")
  decile_col <- paste0(metric_col, "_decile")
  loess_col  <- paste0(metric_col, "_loess")

  bulk_raw <- ann_well_3plus |>
    group_by(lab, mixture, polarity) |>
    summarise(bulk = sum(.data[[metric_col]], na.rm = TRUE),
              .groups = "drop") |>
    mutate(strategy = "Raw")

  bulk_norm <- ann_well_3plus |>
    mutate(int_global = .data[[global_col]],
           int_decile = coalesce(.data[[decile_col]], .data[[global_col]]),
           int_loess  = coalesce(.data[[loess_col]],  .data[[global_col]])) |>
    group_by(lab, mixture, polarity) |>
    summarise(Global = sum(int_global, na.rm = TRUE),
              Decile = sum(int_decile, na.rm = TRUE),
              LOESS  = sum(int_loess,  na.rm = TRUE),
              .groups = "drop") |>
    pivot_longer(c(Global, Decile, LOESS),
                 names_to = "strategy", values_to = "bulk")

  bind_rows(bulk_raw, bulk_norm) |>
    mutate(metric = metric_col)
}

bulk_3plus <- bind_rows(make_bulk_3plus("maxIntensity"),
                        make_bulk_3plus("areaUnderTic"))

icc_3plus <- bind_rows(lapply(metric_names, function(m) {
  bind_rows(lapply(c("pos", "neg"), function(pol_str) {
    pol_name <- if (pol_str == "pos") "Positive" else "Negative"
    bind_rows(lapply(c("Raw", "Global", "Decile", "LOESS"), function(s) {
      df <- bulk_3plus |>
        filter(metric == m, polarity == pol_str, strategy == s)
      if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
      res <- compute_icc(df, "bulk")
      res$strategy <- s
      res$polarity <- pol_name
      res$metric   <- m
      res
    }))
  }))
}))

icc_3plus |>
  transmute(metric, polarity, strategy,
            ICC          = round(ICC, 3),
            `Var(between)` = round(Between_mixture_var, 2),
            `Var(within)`  = round(Within_mixture_var, 2)) |>
  arrange(metric, polarity, factor(strategy, levels = strategy_names)) |>
  knitr::kable(caption = sprintf("Option B 3+/4 quorum ICC (n = %d).",
                                  nrow(well_3plus)))
```

# P3.6 Pairwise ICC across all 6 lab pairs

Includes the same-platform Agilent pair (afekta + cembio) as the ceiling
reference. ICC with 2 raters is well-defined; the within-mixture
component is estimated from a single `(lab_a − lab_b)²` per mixture so
the point estimate is noisier than the 4-lab number.

```{r p36-pairwise-icc}
lab_pairs <- list(
  c("afekta", "cembio"),
  c("afekta", "hmgu"),
  c("afekta", "icl"),
  c("cembio", "hmgu"),
  c("cembio", "icl"),
  c("hmgu",   "icl"))

agilent_labs <- c("afekta", "cembio")

icc_pairs <- bind_rows(lapply(lab_pairs, function(p) {
  pair_label <- paste(sort(p), collapse = " + ")
  bulk_pair  <- bulk_all |> filter(lab %in% p)
  bind_rows(lapply(metric_names, function(m) {
    bind_rows(lapply(unique(bulk_pair$alignment_set), function(aset) {
      bind_rows(lapply(c("pos", "neg"), function(pol_str) {
        pol_name <- if (pol_str == "pos") "Positive" else "Negative"
        bind_rows(lapply(strategy_names, function(s) {
          df <- bulk_pair |> filter(metric == m, polarity == pol_str,
                                    alignment_set == aset, strategy == s)
          if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
          res <- compute_icc(df, "bulk")
          res$strategy      <- s
          res$polarity      <- pol_name
          res$metric        <- m
          res$alignment_set <- aset
          res$pair          <- pair_label
          res$same_platform <- all(p %in% agilent_labs)
          res
        }))
      }))
    }))
  }))
}))

icc_pairs |>
  filter(alignment_set == strict_label) |>
  transmute(metric, polarity, pair, strategy,
            ICC = round(ICC, 3),
            same_platform = ifelse(same_platform, "★", "")) |>
  arrange(metric, polarity, pair,
          factor(strategy, levels = strategy_names)) |>
  knitr::kable(caption = "Pairwise ICC (Strict). ★ = same-platform pair.")
```

```{r p36-pairwise-heatmap}
#| fig-cap: "Pairwise ICC heatmap (Strict subset)."
#| fig-height: 9
icc_pairs |>
  filter(alignment_set == strict_label) |>
  mutate(pair = factor(pair, levels = sapply(lab_pairs,
                                              \(p) paste(sort(p), collapse = " + "))),
         strategy = factor(strategy, levels = strategy_names),
         pair_label = ifelse(same_platform,
                             paste(as.character(pair), "★"),
                             as.character(pair))) |>
  ggplot(aes(x = strategy, y = reorder(pair_label, ICC), fill = ICC)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", ICC)), size = 2.7) +
  facet_grid(metric ~ polarity) +
  scale_fill_gradient2(low = "white", mid = "#FDC086", high = "#1F77B4",
                       midpoint = 0.5, limits = c(0, 1), name = "ICC") +
  labs(x = "Strategy", y = "Lab pair (★ = same platform)",
       title = "Pairwise ICC heatmap — Strict subset") +
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.y = element_text(size = 8))
```

```{r p36-icc-by-platform-class}
#| fig-cap: "Same-platform ceiling vs cross-platform distribution (maxIntensity, Strict)."
#| fig-height: 4
class_df <- icc_pairs |>
  filter(alignment_set == strict_label,
         metric == "maxIntensity") |>
  mutate(strategy = factor(strategy, levels = strategy_names),
         class    = ifelse(same_platform,
                           "Same platform (Agilent + Agilent)",
                           "Cross platform"))

ggplot(class_df |> filter(class == "Cross platform"),
       aes(x = strategy, y = ICC)) +
  geom_boxplot(fill = "grey85", outlier.shape = NA) +
  geom_jitter(alpha = 0.6, size = 1.5, width = 0.15) +
  geom_point(data = class_df |> filter(class == "Same platform (Agilent + Agilent)"),
             colour = "red", size = 3.5, shape = 18) +
  geom_text(data = class_df |> filter(class == "Same platform (Agilent + Agilent)"),
            aes(label = sprintf("%.2f", ICC)),
            colour = "red", size = 3, hjust = -0.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  facet_wrap(~ polarity) +
  ylim(NA, 1) +
  labs(x = NULL, y = "Pairwise ICC",
       title = "Same-platform vs cross-platform pairwise ICC",
       subtitle = "Strict subset, maxIntensity. Red diamond = afekta+cembio.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## P3.6.b Per-lab diagnostic: bulk vs consensus

```{r p36b-rank-correlation}
bulk_with_consensus <- bulk_all |>
  filter(alignment_set == strict_label) |>
  group_by(metric, polarity, strategy, mixture) |>
  mutate(consensus_bulk = median(bulk, na.rm = TRUE)) |>
  ungroup()

rank_cor <- bulk_with_consensus |>
  filter(is.finite(bulk), bulk > 0,
         is.finite(consensus_bulk), consensus_bulk > 0) |>
  group_by(metric, polarity, strategy, lab) |>
  summarise(spearman_rho = cor(bulk, consensus_bulk, method = "spearman"),
            pearson_log  = cor(log2(bulk), log2(consensus_bulk),
                               method = "pearson"),
            n_mixtures   = n(),
            .groups = "drop")

rank_cor |>
  filter(strategy %in% c("Raw", "Lab-median")) |>
  transmute(metric, polarity, strategy, lab,
            Spearman    = round(spearman_rho, 3),
            `Pearson (log2)` = round(pearson_log, 3),
            n_mixtures) |>
  arrange(metric, polarity, strategy, lab) |>
  knitr::kable(caption = "Per-lab rank + log correlation vs cross-lab consensus.")
```

```{r p36b-rank-plot}
#| fig-cap: "Per-lab Spearman rho vs consensus."
#| fig-height: 6
rank_cor |>
  filter(strategy %in% c("Raw", "Lab-median", "Decile", "LOESS")) |>
  mutate(strategy = factor(strategy, levels = strategy_names)) |>
  ggplot(aes(x = lab, y = spearman_rho, fill = lab)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f", spearman_rho)),
            vjust = -0.4, size = 2.6) +
  geom_hline(yintercept = c(0.5, 0.8), linetype = "dashed", colour = "grey50") +
  facet_grid(strategy + metric ~ polarity, switch = "y") +
  scale_fill_hue() +
  ylim(NA, 1.05) +
  labs(x = NULL, y = "Spearman rho vs consensus",
       title = "Per-lab Spearman rank correlation vs cross-lab consensus") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.y.left = element_text(size = 7))
```

```{r p36b-bulk-vs-consensus-scatter}
#| fig-cap: "Per-lab bulk vs cross-lab consensus per mixture (log-log)."
#| fig-height: 8
bulk_with_consensus |>
  filter(strategy %in% c("Raw", "Lab-median"),
         is.finite(bulk), bulk > 0,
         is.finite(consensus_bulk), consensus_bulk > 0) |>
  ggplot(aes(x = consensus_bulk, y = bulk, colour = lab)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50") +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  facet_grid(metric + polarity ~ strategy + lab, scales = "free") +
  scale_x_log10() + scale_y_log10() +
  scale_color_hue() +
  labs(x = "Consensus bulk", y = "Lab bulk",
       title = "Per-lab bulk vs cross-lab consensus per mixture") +
  theme_bw() +
  theme(legend.position = "none",
        strip.text = element_text(size = 7),
        axis.text  = element_text(size = 7))
```

```{r p36b-detected-peak-count}
#| fig-cap: "Detected-peak count per injection per (lab, polarity)."
#| fig-height: 4
peak_counts_path <- file.path(object_dir, "detected_peak_counts.csv")
if (file.exists(peak_counts_path)) {
  peak_counts <- read.csv(peak_counts_path, stringsAsFactors = FALSE)

  peak_counts |>
    ggplot(aes(x = lab, y = n_peaks, fill = lab)) +
    geom_boxplot(alpha = 0.7, outlier.size = 0.7) +
    geom_jitter(width = 0.15, alpha = 0.6, size = 1) +
    facet_wrap(~ polarity) +
    scale_y_log10() +
    scale_fill_hue() +
    labs(x = NULL, y = "Detected peaks per injection (log)",
         title = "Per-injection detected-peak count") +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  message("detected_peak_counts.csv not found — re-render ",
          "5_normalization.qmd to produce it.")
}
```

## P3.6.c Leave-one-out consensus

```{r p36c-loo-rank}
bulk_with_loo_consensus <- bulk_all |>
  filter(alignment_set == strict_label) |>
  group_by(metric, polarity, strategy, mixture) |>
  mutate(loo_consensus = vapply(seq_along(bulk),
    function(i) median(bulk[-i], na.rm = TRUE), numeric(1))) |>
  ungroup()

loo_rank_cor <- bulk_with_loo_consensus |>
  filter(is.finite(bulk), bulk > 0,
         is.finite(loo_consensus), loo_consensus > 0) |>
  group_by(metric, polarity, strategy, lab) |>
  summarise(spearman_loo = cor(bulk, loo_consensus, method = "spearman"),
            pearson_loo_log = cor(log2(bulk), log2(loo_consensus),
                                  method = "pearson"),
            n_mixtures   = n(),
            .groups = "drop")

loo_rank_cor |>
  filter(strategy %in% c("Raw", "Lab-median")) |>
  transmute(metric, polarity, strategy, lab,
            Spearman_LOO        = round(spearman_loo, 3),
            `Pearson(log2)_LOO` = round(pearson_loo_log, 3),
            n_mixtures) |>
  arrange(metric, polarity, strategy, lab) |>
  knitr::kable(caption = "Per-lab rank/log correlation vs LOO consensus.")
```

```{r p36c-loo-plot}
#| fig-cap: "Spearman rho vs leave-one-out consensus."
#| fig-height: 6
loo_rank_cor |>
  filter(strategy %in% c("Raw", "Lab-median", "Decile", "LOESS")) |>
  mutate(strategy = factor(strategy, levels = strategy_names)) |>
  ggplot(aes(x = lab, y = spearman_loo, fill = lab)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f", spearman_loo)),
            vjust = -0.4, size = 2.6) +
  geom_hline(yintercept = c(0.5, 0.8), linetype = "dashed", colour = "grey50") +
  facet_grid(strategy + metric ~ polarity, switch = "y") +
  scale_fill_hue() +
  ylim(-0.6, 1.05) +
  labs(x = NULL, y = "Spearman rho vs LOO consensus",
       title = "Per-lab rank correlation vs leave-one-out consensus") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.y.left = element_text(size = 7))
```

## P3.6.e Per-compound hmgu vs other-3 labs

```{r p36e-per-compound-hmgu}
per_compound_hmgu <- ann_well_aligned |>
  group_by(mixture, compound, polarity) |>
  mutate(other_lab_median = median(maxIntensity[lab != "hmgu"], na.rm = TRUE)) |>
  ungroup() |>
  filter(lab == "hmgu",
         is.finite(maxIntensity),    maxIntensity > 0,
         is.finite(other_lab_median), other_lab_median > 0)

fit_stats <- per_compound_hmgu |>
  group_by(polarity) |>
  summarise(n_pts        = n(),
            spearman     = round(cor(maxIntensity, other_lab_median,
                                     method = "spearman"), 3),
            pearson_log2 = round(cor(log2(maxIntensity), log2(other_lab_median),
                                     method = "pearson"), 3),
            .groups = "drop")

knitr::kable(fit_stats,
             caption = "hmgu maxIntensity vs median of other-3 labs.")
```

```{r p36e-per-compound-hmgu-plot}
#| fig-cap: "Per-(mixture, compound) hmgu vs other-3 median (log-log)."
#| fig-height: 5
ggplot(per_compound_hmgu, aes(x = other_lab_median, y = maxIntensity)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 1.5, alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, colour = "red", linewidth = 0.7) +
  facet_wrap(~ polarity, scales = "free") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Median maxIntensity of other 3 labs",
       y = "hmgu maxIntensity",
       title = "Per-compound: hmgu vs other-3 median") +
  theme_bw()
```

## P3.6.f 3-lab ICC ladder (afekta + cembio + icl, no hmgu)

```{r p36f-three-lab-icc}
three_lab_set <- c("afekta", "cembio", "icl")
bulk_3lab     <- bulk_all |> filter(lab %in% three_lab_set)

icc_3lab <- bind_rows(lapply(metric_names, function(m) {
  bind_rows(lapply(unique(bulk_3lab$alignment_set), function(aset) {
    bind_rows(lapply(c("pos", "neg"), function(pol_str) {
      pol_name <- if (pol_str == "pos") "Positive" else "Negative"
      bind_rows(lapply(strategy_names, function(s) {
        df <- bulk_3lab |>
          filter(metric == m, polarity == pol_str,
                 alignment_set == aset, strategy == s)
        if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
        res <- compute_icc(df, "bulk")
        res$strategy      <- s
        res$polarity      <- pol_name
        res$metric        <- m
        res$alignment_set <- aset
        res
      }))
    }))
  }))
}))

icc_3lab |>
  filter(alignment_set == strict_label) |>
  transmute(metric, polarity, strategy,
            ICC          = round(ICC, 3),
            `Var(between)` = round(Between_mixture_var, 2),
            `Var(within)`  = round(Within_mixture_var, 2)) |>
  arrange(metric, polarity, factor(strategy, levels = strategy_names)) |>
  knitr::kable(caption = "3-lab ICC (afekta + cembio + icl) — Strict subset.")
```

```{r p36f-three-vs-four-lab-plot}
#| fig-cap: "ICC under 4 / 3 / 2 lab subsets, Strict."
#| fig-height: 7
compare_lab_subsets <- bind_rows(
  icc_all     |> mutate(lab_subset = "4 labs (all)"),
  icc_3lab    |> mutate(lab_subset = "3 labs (no hmgu)"),
  icc_pairs   |> filter(pair == "afekta + cembio") |>
                  mutate(lab_subset = "2 labs (Agilent only)") |>
                  select(-pair, -same_platform)
) |>
  mutate(strategy = factor(strategy, levels = strategy_names),
         lab_subset = factor(lab_subset,
                             levels = c("4 labs (all)",
                                        "3 labs (no hmgu)",
                                        "2 labs (Agilent only)"))) |>
  filter(alignment_set == strict_label)

ggplot(compare_lab_subsets,
       aes(x = strategy, y = ICC, fill = lab_subset)) +
  geom_col(position = position_dodge(width = 0.85), alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f", ICC)),
            position = position_dodge(width = 0.85),
            vjust = -0.3, size = 2.4) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  facet_grid(metric ~ polarity) +
  scale_fill_manual(values = c("4 labs (all)"          = "#E64B35",
                               "3 labs (no hmgu)"      = "#70AD47",
                               "2 labs (Agilent only)" = "#4DBBD5")) +
  ylim(NA, 1) +
  labs(x = NULL, y = "ICC", fill = NULL,
       title = "ICC under 4 / 3 / 2 lab subsets") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

## P3.6.g Per-compound pairwise correlations

```{r p36g-pairwise-per-compound}
wide_per_compound <- ann_well_aligned |>
  select(mixture, compound, polarity, lab, maxIntensity, areaUnderTic) |>
  pivot_wider(names_from  = lab,
              values_from = c(maxIntensity, areaUnderTic))

pairs_for_per_compound <- list(
  c("hmgu",   "afekta"),
  c("hmgu",   "cembio"),
  c("hmgu",   "icl"),
  c("afekta", "cembio"),
  c("afekta", "icl"),
  c("cembio", "icl"))

per_compound_pair_cor <- bind_rows(lapply(pairs_for_per_compound, function(p) {
  bind_rows(lapply(c("pos", "neg"), function(pol_str) {
    pol_name <- if (pol_str == "pos") "Positive" else "Negative"
    df <- wide_per_compound |> filter(polarity == pol_str)
    bind_rows(lapply(c("maxIntensity", "areaUnderTic"), function(m) {
      a <- df[[paste0(m, "_", p[1])]]
      b <- df[[paste0(m, "_", p[2])]]
      ok <- is.finite(a) & a > 0 & is.finite(b) & b > 0
      if (sum(ok) < 5L) return(NULL)
      tibble(pair         = paste(sort(p), collapse = " + "),
             polarity     = pol_name,
             metric       = m,
             n            = sum(ok),
             spearman     = cor(a[ok], b[ok], method = "spearman"),
             pearson_log2 = cor(log2(a[ok]), log2(b[ok]), method = "pearson"),
             contains_hmgu = "hmgu" %in% p)
    }))
  }))
}))

per_compound_pair_cor |>
  transmute(metric, polarity, pair, contains_hmgu,
            Spearman      = round(spearman, 3),
            `Pearson(log2)` = round(pearson_log2, 3), n) |>
  arrange(metric, polarity, contains_hmgu, pair) |>
  knitr::kable(caption = "Per-compound pairwise correlations.")
```

```{r p36g-pair-heatmap}
#| fig-cap: "Per-compound pairwise Spearman rho."
#| fig-height: 5
per_compound_pair_cor |>
  mutate(pair_factor = factor(pair,
                              levels = c("afekta + cembio",
                                         "afekta + icl",
                                         "cembio + icl",
                                         "afekta + hmgu",
                                         "cembio + hmgu",
                                         "hmgu + icl"))) |>
  ggplot(aes(x = metric, y = pair_factor, fill = spearman)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", spearman)),
            colour = "black", size = 3) +
  facet_wrap(~ polarity) +
  scale_fill_gradient2(low = "#B40426", mid = "white",
                       high = "#3B4CC0", midpoint = 0,
                       limits = c(-1, 1), name = "Spearman") +
  labs(x = NULL, y = "Lab pair",
       title = "Per-compound pairwise Spearman rho") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# P3.7 Per-(compound, mixture) cross-lab CV distribution

```{r p37-per-observation-cv}
per_obs_cv <- ann_well_aligned |>
  group_by(compound, mixture, polarity) |>
  summarise(
    n_labs    = n(),
    cv_Raw    = sd(maxIntensity,        na.rm = TRUE) /
                abs(mean(maxIntensity,        na.rm = TRUE)),
    cv_Global = sd(maxIntensity_global, na.rm = TRUE) /
                abs(mean(maxIntensity_global, na.rm = TRUE)),
    cv_Decile = sd(maxIntensity_decile, na.rm = TRUE) /
                abs(mean(maxIntensity_decile, na.rm = TRUE)),
    cv_LOESS  = sd(maxIntensity_loess,  na.rm = TRUE) /
                abs(mean(maxIntensity_loess,  na.rm = TRUE)),
    .groups = "drop") |>
  filter(n_labs == 4L) |>
  pivot_longer(starts_with("cv_"), names_to = "strategy",
               values_to = "cv", names_prefix = "cv_") |>
  filter(!is.na(cv), is.finite(cv))

per_obs_cv |>
  group_by(polarity, strategy) |>
  summarise(n              = n(),
            median_cv      = round(median(cv), 3),
            mean_cv        = round(mean(cv),   3),
            pct_below_0_3  = round(100 * mean(cv < 0.3), 1),
            pct_below_0_5  = round(100 * mean(cv < 0.5), 1),
            .groups = "drop") |>
  arrange(polarity, factor(strategy,
                           levels = c("Raw", "Global", "Decile", "LOESS"))) |>
  knitr::kable(caption = "Per-(compound, mixture) cross-lab CV distribution.")
```

```{r p37-per-observation-cv-plot}
#| fig-cap: "Per-(compound, mixture) cross-lab CV under each strategy."
#| fig-height: 6
if (nrow(per_obs_cv) > 0L) {
  medians <- per_obs_cv |>
    group_by(polarity, strategy) |>
    summarise(median_cv = median(cv), .groups = "drop")

  ggplot(per_obs_cv |> filter(cv < 3),
         aes(x = cv, fill = strategy)) +
    geom_histogram(bins = 40, alpha = 0.8) +
    geom_vline(data = medians, aes(xintercept = median_cv),
               colour = "black", linewidth = 0.6) +
    geom_vline(xintercept = c(0.3, 0.5),
               linetype = "dashed", colour = "red") +
    facet_grid(strategy ~ polarity) +
    scale_fill_manual(values = strategy_palette) +
    labs(x = "Cross-lab CV per (compound, mixture)",
         y = "Observations",
         title = "Per-(compound, mixture) cross-lab CV") +
    theme_bw() +
    theme(legend.position = "none")
} else {
  message("No (compound, mixture, polarity) tuples with all 4 labs detected.")
}
```

# Notes

- Inputs: `annotation_chrom_metrics_normalized.csv` (P3 producer),
  `detected_peak_counts.csv` (P3 producer).
- The producer rebuilds the normalized CSV from scratch when any of its
  inputs change; this consumer is fast (CSV reads + dplyr ops only).
- Five strategies × two metrics × two compound-set policies × four labs
  = 16 main ICC numbers, expanded to 6 pairs × 6 strategies in §P3.6.
- The "Lab-median" strategy is computed in this file from the bulk
  distribution (matches 2_global's `tic_norm` exactly). The producer
  does **not** export Lab-median scaled values because the
  normalization is per-(lab, polarity) post-hoc, not per-row.
