---
title: "Reproducibility results — data-level progression and pairwise agreement"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
        embed-resources: true
---

# Introduction

The same 19 mixtures × 4 labs × 2 polarities are observable at **four
different data levels**, each built by a different stage of the
processing pipeline:

| Level | Built by | Feature space | Per sample summary |
|---|---|---|---|
| **Full**       | mzML, no peak detection | all MS1 ions | TIC, BPC |
| **Detected**   | xcms (per-lab, per-sample peak picking) | per-injection detected peaks | TIC, BPC restricted to detected ions |
| **Annotated**  | SIRIUS MS2 + SNR-aware chromExtract | the spike-in panel (~250 (mix × compound) tuples detected in ≥1 lab) | sum/max of maxIntensity over the panel |
| **Consensus**  | `4_consensus_peaks.qmd` cross-lab feature matching | NAPS-aligned features matched in all 4 labs | sample × consensus-peak matrix |

This file asks **one question**: *does each successive layer recover
biological signal from technical variance?* The answer is yes — see the
Phase 1 variance decomposition and PCA progression for the headline
story, and the Phase 5 Consensus-peak analysis for the per-peak picture
on the broader feature space (1.8k features vs ~250 in Annotated).

Other axes of the analysis live in dedicated files:

- **Normalization strategies + full ICC ladder + all-strategies
  pairwise ICC heatmap** → `5_normalization.qmd` §P3.1–§P3.6
- **NAPS RT alignment** → `3_rt_alignment.qmd`
- **Per-compound identity / RT / abundance / m/z diagnostics** →
  `8_per_compound_diagnostics.qmd`
- **Confidence tiering (A / B / C / D)** → `6_confidence_tier.qmd`

Data: 19 standard mixtures (HUMAN endosome) from 4 labs with harmonized
LC systems. No technical replicates; lab identity is confounded with
instrument platform.

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)
library(pheatmap)

base_path     <- here("5_downstream_analysis")
producer_dir  <- here("5_downstream_analysis", "object")
source(file.path(base_path, "helpers.R"))
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`, `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) %>%
  distinct() %>%
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) %>%
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

Each lab's protocol is unique; lab identity is completely confounded with
instrumental setup. Notable variations:

- MS1 scan rate
- Injection volume
- Solvent brand.

## Metrics Definition

Three main chromatographic metrics:

- **TIC**: Sum of all ion intensities
- **BPC**: Max intensity per scan

Three data levels (least → most processed):

- **Full Signal**: raw profile including baseline (`bpc_full` / `tic_full`)
- **Detected Signal**: peaks from centWave; isolates compound signal from
  baseline (`bpc_detect` / `tic_detect`)
- **Annotated Signal**: derived from
  `annotation_chrom_metrics.csv` (SIRIUS producer) restricted to
  `(mixture, compound)` pairs with non-NA `maxIntensity` in **all 4
  labs** at that polarity (recomputed each render — the
  "well-detected" set, not hardcoded). Per-sample bulk:
  `tic = sum(areaUnderTic)`; `bpc = max(maxIntensity)`. Polarity is
  collapsed per-side so each polarity has its own well-detected set.

Two polarities: Positive and Negative ionization modes.

```{r load-data-both-modes}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "object/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]
    load(file.path(base_path, "object/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "object/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]
    load(file.path(base_path, "object/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = ifelse(data_level == "full", "Full Signal", "Detected Signal")) %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm = bpc / median_bpc,
      tic_norm = tic / median_tic
    ) %>%
    ungroup()
}

compute_metrics_consensus <- function(polarity, polarity_name) {
  # Consensus Signal — TIC of the cross-lab matched feature signal per
  # (lab, mixture). Loaded from tic_consensus.RData (built in
  # 4_consensus_peaks.qmd / 1_full_detected_objects.qmd). BPC here is
  # derived as the per-chromatogram max-intensity (parallel to TIC = sum).
  load(file.path(base_path, "object/tic_consensus.RData"))
  tic <- tic_consensus[tic_consensus$polarity == polarity]
  ints <- intensity(tic)
  data.frame(
    lab = tic$lab,
    mixture = tic$mixture,
    polarity = polarity_name,
    bpc = vapply(ints, max, numeric(1), na.rm = TRUE),
    tic = vapply(ints, sum, numeric(1), na.rm = TRUE)
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = "Consensus Signal") %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm   = bpc / median_bpc,
      tic_norm   = tic / median_tic
    ) %>%
    ungroup()
}

compute_metrics_annotated <- function(polarity, polarity_name) {
  pol_str <- if (polarity == 1L) "pos" else "neg"

  # mixture forced to character — read.csv would otherwise coerce quoted
  # "2.10" → 2.1 (numeric), silently colliding with mixture "2.1".
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))
  chrom$mixture <- gsub("\\.", "_", chrom$mixture)

  well_detected <- chrom %>%
    filter(polarity == pol_str, !is.na(maxIntensity)) %>%
    distinct(mixture, compound, lab) %>%
    count(mixture, compound) %>%
    filter(n == 4L) %>%
    select(mixture, compound)

  bulk <- chrom %>%
    filter(polarity == pol_str, !is.na(maxIntensity)) %>%
    inner_join(well_detected, by = c("mixture", "compound")) %>%
    group_by(lab, mixture, compound) %>%
    slice_max(maxIntensity, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    group_by(lab, mixture) %>%
    summarise(bpc = max(maxIntensity, na.rm = TRUE),
              tic = sum(areaUnderTic, na.rm = TRUE),
              .groups = "drop")

  data.frame(
    lab = bulk$lab, mixture = bulk$mixture,
    polarity = polarity_name,
    bpc = bulk$bpc, tic = bulk$tic
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = "Annotated Signal") %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm   = bpc / median_bpc,
      tic_norm   = tic / median_tic
    ) %>%
    ungroup()
}

metrics_full      <- run_for_polarities(compute_metrics, "full")
metrics_detect    <- run_for_polarities(compute_metrics, "detect")
metrics_annotated <- run_for_polarities_names(compute_metrics_annotated)
metrics_consensus <- run_for_polarities_names(compute_metrics_consensus)

# Recompute the "275-set" size (per polarity) for the prose / sanity check.
.well_n <- function(p) {
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))
  chrom$mixture <- gsub("\\.", "_", chrom$mixture)
  chrom %>% filter(polarity == p, !is.na(maxIntensity)) %>%
    distinct(mixture, compound, lab) %>%
    count(mixture, compound) %>% filter(n == 4L) %>% nrow()
}
cat("well-detected (mixture, compound) pairs — pos:", .well_n("pos"),
    " | neg:", .well_n("neg"), "\n")

metrics_combined <- bind_rows(metrics_full, metrics_detect,
                              metrics_annotated, metrics_consensus) %>%
  mutate(data_level = factor(data_level,
                             # Display order = Full -> Detected -> Consensus
                             # -> Annotated. The three TIC-based untargeted
                             # levels appear first (Full, Detected, Consensus);
                             # Annotated (per-compound abundance sum) is
                             # conceptually different and kept rightmost.
                             levels = c("Full Signal",
                                        "Detected Signal",
                                        "Consensus Signal",
                                        "Annotated Signal"))) %>%
  group_by(lab, polarity, data_level) %>%
  mutate(
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE)
  ) %>%
  ungroup()

# Reusable ordered character vector for the few loops that iterate levels.
# Order: Full -> Detected -> Consensus -> Annotated (Annotated rightmost).
data_levels_all <- c("Full Signal", "Detected Signal",
                     "Consensus Signal", "Annotated Signal")
```

---

# Phase 1: The Problem — Raw Data Assessment {#phase1}

This phase characterizes the raw data to understand the magnitude of
inter-laboratory variability before any corrections are applied.

## 1.1 Distribution of Raw Metrics

### 1.1.1 Violin Plots: Overall Intensity Distribution

```{r violin-plots}
#| fig-cap: "BPC and TIC distributions by lab, polarity, and data level (Full → Detected → Consensus → Annotated, rightmost)."
#| fig-width: 18
#| fig-height: 7
for (metric in c("bpc", "tic")) print(plot_violin(metrics_combined, metric))
```

Detected signal recenters the negative and positive distributions.
The bimodal shape is evident.

### 1.1.2 Density Plots: Distribution Shape by Lab

```{r density-plots}
ggplot(metrics_combined %>% filter(data_level == "Full Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_bw() +
  labs(title = "TIC Distribution by Lab (Full Signal)",
       x = "log2(TIC)", y = "Density")

ggplot(metrics_combined %>% filter(data_level == "Detected Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_bw() +
  labs(title = "TIC Distribution by Lab (Detected Signal)",
       x = "log2(TIC)", y = "Density")
```

Full signal is more compressed than detected signal, likely because
each mixture has different numbers of compounds.

### 1.1.3 Mixture-Level Agreement Across Labs

```{r mixture-distribution}
#| fig-width: 14
#| fig-height: 7
plot_mixture_distribution <- function(data, metric) {
  if (metric %in% c("bpc", "tic")) {
    data <- data |> mutate(across(all_of(metric), log2))
    title_suffix <- "(log2)"
  } else {
    title_suffix <- ""
  }

  ggplot(data, aes(x = mixture, y = .data[[metric]], color = lab, group = lab)) +
    geom_line(alpha = 0.7) +
    facet_grid(data_level ~ polarity) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(title = paste("Mixture Distribution Across Labs -", metric, title_suffix),
         x = "Mixture", y = paste(metric, title_suffix), color = "Laboratory")
}

for (metric in c("bpc", "tic")) {
  print(plot_mixture_distribution(metrics_combined, metric))
}
```

There is good agreement between Cembio and Afekta, and some with ICL.
However, HMGU is consistently lower intensity.

## 1.2 Variance Decomposition: Quantifying the Problem

### 1.2.1 Linear Model Analysis

Let's run linear models to evaluate mixture effects and batch effects.
The goal is to quantify how much of the variance is due to lab vs mixture.
We use ANOVA to aggregate coefficients into Sum of Squares (SS), which
quantifies the relative power of each factor.

We run this analysis across all 4 conditions (positive/negative, full/detected):

```{r variance-decomposition-lm}
results_table <- data.frame()
residual_plots <- list()
all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in data_levels_all) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")

    residual_plots[[paste(pol, level, sep = " - ")]] <- result$residual_plot
    all_coefficients <- rbind(all_coefficients,
      result$coefficients %>% mutate(Polarity = pol, Data_Level = level))
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

### 1.2.2 Lab Effect Coefficients

```{r lab-coefficients}
#| fig-cap: "Lab effect coefficients from linear models. Deviation from reference lab (Afekta); negative = lower intensity."
lab_coefs <- all_coefficients %>%
  filter(type == "Lab Effect") %>%
  mutate(lab_name = gsub("lab", "", coefficient))

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_hue() +
  labs(title = "Lab Effect Coefficients from Linear Models",
       subtitle = "Deviation from reference lab (afekta); negative = lower intensity",
       x = "Laboratory", y = "Coefficient (log2 scale)", fill = "Laboratory") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.2.3 Variance Decomposition Summary

```{r variance-stacked-bar}
#| fig-cap: "Variance decomposition of TIC across all four data levels (Full → Detected → Consensus → Annotated) and both polarities. Lab variance dominates in raw data; subsequent steps progressively shift variance away from lab towards mixture and residual."
#| fig-width: 14
#| fig-height: 7
variance_long <- results_table %>%
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) %>%
  pivot_longer(cols = ends_with("_Pct"), names_to = "Source", values_to = "Percentage") %>%
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Data_Level = factor(Data_Level, levels = data_levels_all),
    R2_label = paste("R² =", round(R_squared, 3))
  )

text_labels <- variance_long %>% distinct(Polarity, Data_Level, R2_label)

ggplot(variance_long,
       aes(x = Data_Level, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.85, width = 0.7) +
  geom_text(aes(x = Data_Level, label = R2_label, y = 107), data = text_labels,
            size = 3.6, inherit.aes = FALSE) +
  facet_wrap(~ Polarity, nrow = 1) +
  labs(title = "Variance decomposition: TIC across all four data levels",
       subtitle = "Raw data shows massive lab variance — subsequent steps shift variance away from lab",
       x = NULL, y = "Variance contribution (%)", fill = "Source") +
  theme_bw(base_size = 13) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylim(0, 115)
```

### 1.2.4 ICC across the four data levels

The variance decomposition above splits total variance into Lab,
Mixture, and Residual components — informative but not directly
comparable to the per-strategy bulk ICC reported in §P3.5 of
`5_normalization.qmd`. Here we compute the **same ICC statistic**
(variance-component ratio on log2-bulk TIC) at each data level so the
progression Full → Detected → Consensus → Annotated is on one axis.

```{r icc-across-data-levels}
#| fig-cap: "Reproducibility ICC of bulk TIC across the four data levels. Higher = labs agree on a mixture's value as tightly as the mixtures themselves differ. Computed with the same variance-component formula used by 5_normalization.qmd §P3.5."
#| fig-height: 4
icc_one <- function(df) {
  if (!nrow(df) || all(is.na(df$tic))) return(NA_real_)
  df$lv <- log2(pmax(df$tic, 1e-9))
  between <- var(tapply(df$lv, df$mixture, mean,
                         na.rm = TRUE), na.rm = TRUE)
  within  <- mean(tapply(df$lv, df$mixture, var,
                          na.rm = TRUE), na.rm = TRUE)
  if (!is.finite(between + within) || (between + within) == 0)
    return(NA_real_)
  between / (between + within)
}

icc_by_level <- metrics_combined |>
  filter(!is.na(tic), tic > 0) |>
  group_by(polarity, data_level) |>
  group_modify(~ tibble(ICC = icc_one(.x))) |>
  ungroup() |>
  mutate(data_level = factor(data_level, levels = data_levels_all))

ggplot(icc_by_level, aes(x = data_level, y = ICC, fill = data_level)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f", ICC)),
            vjust = -0.4, size = 4) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  facet_wrap(~ polarity) +
  scale_fill_brewer(palette = "Set2") +
  ylim(0, 1.05) +
  labs(title = "Reproducibility (ICC) by data level",
       subtitle = "Bulk TIC, variance-component ICC. Red = ICC 0.5 moderate-reproducibility landmark.",
       x = NULL, y = "ICC") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))

knitr::kable(icc_by_level |>
               mutate(ICC = round(ICC, 3)) |>
               tidyr::pivot_wider(names_from = polarity,
                                   values_from = ICC),
             caption = "ICC of bulk TIC across the four data levels.")
```

### 1.2.5 Residual Diagnostics

```{r residual-plots}
#| fig-height: 10
residual_data <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in data_levels_all) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    residual_data <- rbind(residual_data, data.frame(
      Fitted = fitted(result$model), Residuals = residuals(result$model),
      Polarity = pol, Data_Level = level,
      Lab = data_subset$lab, Mixture = as.factor(data_subset$mixture)
    ))
  }
}

ggplot(residual_data %>%
         mutate(Data_Level = factor(Data_Level, levels = data_levels_all)),
       aes(x = Fitted, y = Residuals, color = Mixture, shape = Lab)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_color_hue() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(title = "Fitted vs Residuals: TIC Across All Conditions",
       subtitle = "Points should scatter randomly around zero for a good model fit",
       x = "Fitted Values", y = "Residuals", color = "Mixture", shape = "Laboratory") +
  theme_bw() +
  theme(legend.position = "bottom")

results_table
```

**Key Finding:** Overall variance is dominated by lab effects.
The residual plot clearly shows separation between labs vs HMGU and
still lab separation in others. Lab separation remains unexplained by
mixture. We cannot compare labs directly without normalization.

## 1.3 Multivariate visualization (PCA / UMAP / presence–absence)

Colour = mixture, shape = lab. Same-shape clustering means the lab
dominates the layout; same-colour clustering means mixture (biology)
dominates. The four-step progression
**Full → Detected → Annotated → Consensus** is the same problem from
the variance decomposition above, viewed in 2D.

### 1.3.1 Build PCA matrices

```{r pca-helpers}
prepare_pca_tic <- function(data_level, polarity_int) {
  if (data_level == "full") {
    load(file.path(base_path, "object/tic_full.RData"))
    chr_data <- tic_full[tic_full$polarity == polarity_int]
  } else {
    load(file.path(base_path, "object/tic_detect.RData"))
    chr_data <- tic_detect[tic_detect$polarity == polarity_int]
  }

  int_list <- intensity(chr_data); rt_list <- rtime(chr_data)
  all_rts  <- unlist(rt_list)
  rt_grid  <- seq(min(all_rts, na.rm = TRUE),
                  max(all_rts, na.rm = TRUE), length.out = 500)

  int_matrix <- t(sapply(seq_along(int_list), function(i) {
    approx(rt_list[[i]], int_list[[i]], xout = rt_grid,
           method = "linear", rule = 2)$y
  }))
  rownames(int_matrix) <- paste0(chr_data$lab, "_", chr_data$mixture)
  colnames(int_matrix) <- paste0("RT_", round(rt_grid, 1))

  list(data = int_matrix,
       meta = data.frame(sample  = rownames(int_matrix),
                         lab     = chr_data$lab,
                         mixture = chr_data$mixture))
}

# Annotated: sample × compound maxIntensity, restricted to the
# well-detected set (same definition as the bulk Annotated metric above).
prepare_pca_annotated <- function(polarity_int) {
  pol_str <- if (polarity_int == 1L) "pos" else "neg"
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))
  chrom$mixture <- gsub("\\.", "_", chrom$mixture)

  well_detected <- chrom |>
    filter(polarity == pol_str, !is.na(maxIntensity)) |>
    distinct(mixture, compound, lab) |>
    count(mixture, compound) |>
    filter(n == 4L) |>
    select(mixture, compound)

  wide <- chrom |>
    filter(polarity == pol_str, !is.na(maxIntensity)) |>
    inner_join(well_detected, by = c("mixture", "compound")) |>
    group_by(lab, mixture, compound) |>
    slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
    ungroup() |>
    mutate(sample = paste0(lab, "_", mixture)) |>
    select(sample, compound, maxIntensity, lab, mixture) |>
    pivot_wider(names_from = compound, values_from = maxIntensity,
                values_fill = 0) |>
    arrange(sample)

  data_mat <- wide |> select(-sample, -lab, -mixture) |> as.matrix()
  rownames(data_mat) <- wide$sample
  list(data = data_mat, meta = wide |> select(sample, lab, mixture))
}

# Consensus: sample × consensus-peak matrix from 3_peak_feature output.
prepare_pca_consensus <- function(polarity_int) {
  pk_metrics <- read.csv(file.path(base_path,
                                   "object/peak_metrics_detected_all4labs.csv"),
                         stringsAsFactors = FALSE,
                         colClasses = c(mixture = "character"))
  pk_metrics$mixture <- gsub("\\.", "_", pk_metrics$mixture)
  wide <- pk_metrics |>
    filter(as.integer(polarity) == polarity_int) |>
    mutate(sample = paste0(lab, "_", mixture)) |>
    select(sample, consensus_id, areaUnderTic, lab, mixture) |>
    pivot_wider(names_from = consensus_id, values_from = areaUnderTic,
                values_fill = 0, names_prefix = "peak_") |>
    arrange(sample)

  data_mat <- wide |> select(starts_with("peak_")) |> as.matrix()
  rownames(data_mat) <- wide$sample
  list(data = data_mat, meta = wide |> select(sample, lab, mixture))
}

# Run prcomp, drop zero-variance columns so scale.=TRUE doesn't error.
pca_scores_df <- function(pca_list, polarity_label, transform_fn = identity) {
  X <- transform_fn(pca_list$data)
  keep <- apply(X, 2, function(z) is.finite(var(z)) && var(z) > 0)
  res <- prcomp(X[, keep, drop = FALSE], scale. = TRUE, center = TRUE)
  var_expl <- (res$sdev^2 / sum(res$sdev^2)) * 100
  list(scores = data.frame(PC1 = res$x[, 1], PC2 = res$x[, 2],
                           lab = pca_list$meta$lab,
                           mixture = pca_list$meta$mixture,
                           polarity = polarity_label),
       var_expl = var_expl)
}

plot_pca_by_polarity <- function(pos, neg, title, subtitle) {
  scores <- bind_rows(pos$scores, neg$scores) |>
    mutate(polarity = factor(polarity, levels = c("Positive", "Negative")))
  panel_labels <- c(
    Positive = sprintf("Positive (PC1 %.1f%%, PC2 %.1f%%)",
                       pos$var_expl[1], pos$var_expl[2]),
    Negative = sprintf("Negative (PC1 %.1f%%, PC2 %.1f%%)",
                       neg$var_expl[1], neg$var_expl[2])
  )
  ggplot(scores, aes(x = PC1, y = PC2, color = mixture, shape = lab)) +
    geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    facet_wrap(~ polarity, scales = "free",
               labeller = labeller(polarity = panel_labels)) +
    scale_color_hue() +
    labs(title = title, subtitle = subtitle,
         x = "PC1", y = "PC2", color = "Mixture", shape = "Laboratory") +
    theme_bw() +
    theme(plot.title = element_text(face = "bold"),
          legend.position = "right")
}
```

```{r prepare-pca-data}
pca_full_pos      <- prepare_pca_tic("full",   1L)
pca_full_neg      <- prepare_pca_tic("full",   0L)
pca_detect_pos    <- prepare_pca_tic("detect", 1L)
pca_detect_neg    <- prepare_pca_tic("detect", 0L)
pca_annotated_pos <- prepare_pca_annotated(1L)
pca_annotated_neg <- prepare_pca_annotated(0L)
pca_consensus_pos <- prepare_pca_consensus(1L)
pca_consensus_neg <- prepare_pca_consensus(0L)
```

### 1.3.2 PCA progression Full → Detected → Annotated → Consensus

```{r pca-full-tic}
#| fig-cap: "PCA of Raw TIC profiles. Shapes cluster (lab dominates); colours scattered (mixture signal buried in technical variance)."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_full_pos, "Positive"),
  neg = pca_scores_df(pca_full_neg, "Negative"),
  title = "PCA — Full Signal (Raw TIC profiles)",
  subtitle = "Shapes cluster, colours scattered: technical noise dominates"
)
```

```{r pca-detected-tic}
#| fig-cap: "PCA of Detected TIC. Shape clusters loosen, colour grouping emerges — peak detection recovers biological signal."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_detect_pos, "Positive"),
  neg = pca_scores_df(pca_detect_neg, "Negative"),
  title = "PCA — Detected Signal",
  subtitle = "Shapes disperse, colours begin to group: mixture signal emerges"
)
```

```{r pca-annotated}
#| fig-cap: "PCA of Annotated signal (sample × compound maxIntensity over the well-detected set). Identity-keyed features concentrate biological variance further; mixture clusters tighten and lab spread within each mixture shrinks."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_annotated_pos, "Positive", transform_fn = log1p),
  neg = pca_scores_df(pca_annotated_neg, "Negative", transform_fn = log1p),
  title = "PCA — Annotated Signal (per-compound maxIntensity)",
  subtitle = "Restricted to (mixture, compound) pairs detected in all 4 labs"
)
```

```{r pca-consensus}
#| fig-cap: "PCA of Consensus Peaks (matched across all 4 labs). Same-colour points cluster across shapes — biology dominates. log1p compresses the presence/absence gap that would otherwise dominate scaling on the sparse matrix."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_consensus_pos, "Positive", transform_fn = log1p),
  neg = pca_scores_df(pca_consensus_neg, "Negative", transform_fn = log1p),
  title = "PCA — Consensus Peaks (cross-lab matched signal)",
  subtitle = "Colours cluster across shapes: biology dominates"
)
```

### 1.3.2.b Silhouette — unsupervised view of the same story

The PCA scatterplots above are the visual story; this is the numerical
companion. For each (data level × polarity) we take the matrix the PCA
sees, compute Euclidean distances between samples, and ask **silhouette
analysis** two questions in parallel:

- **Silhouette(mixture)** — using *mixture* as the cluster label. Mean
  ∈ \[−1, 1\]; high = same-mixture samples sit closer to each other than
  to other-mixture samples (good cross-lab agreement on what's in each
  mixture). Should **rise** as we move Full → Consensus.
- **Silhouette(lab)** — using *lab* as the cluster label. High = a lab's
  samples cluster together regardless of mixture (lab effect dominates,
  bad). Should **fall** as we move Full → Consensus.

No special handling of zeros, no model assumptions — just "for this
matrix, do samples cluster by mixture (what we want) or by lab (what we
don't)?".

```{r silhouette-by-level}
#| message: false
library(cluster)

# Pull the same input each PCA used (transform_fn applied where appropriate).
sil_inputs <- list(
  Full        = list(pos = pca_full_pos,      neg = pca_full_neg,      tx = identity),
  Detected    = list(pos = pca_detect_pos,    neg = pca_detect_neg,    tx = identity),
  Annotated   = list(pos = pca_annotated_pos, neg = pca_annotated_neg, tx = log1p),
  Consensus   = list(pos = pca_consensus_pos, neg = pca_consensus_neg, tx = log1p))

mean_silhouette <- function(pca_list, transform_fn) {
  X <- transform_fn(pca_list$data)
  # Drop zero-variance columns so distances are well-defined.
  keep <- apply(X, 2, function(z) is.finite(var(z)) && var(z) > 0)
  d <- dist(X[, keep, drop = FALSE])
  by_mix <- mean(silhouette(as.integer(factor(pca_list$meta$mixture)), d)[, "sil_width"])
  by_lab <- mean(silhouette(as.integer(factor(pca_list$meta$lab)),     d)[, "sil_width"])
  c(by_mix = by_mix, by_lab = by_lab)
}

sil_rows <- lapply(names(sil_inputs), function(lvl) {
  inp <- sil_inputs[[lvl]]
  pos <- mean_silhouette(inp$pos, inp$tx)
  neg <- mean_silhouette(inp$neg, inp$tx)
  data.frame(level = lvl,
             sil_mixture_pos = round(pos["by_mix"], 3),
             sil_lab_pos     = round(pos["by_lab"], 3),
             sil_mixture_neg = round(neg["by_mix"], 3),
             sil_lab_neg     = round(neg["by_lab"], 3))
})
sil_tbl <- do.call(rbind, sil_rows)
rownames(sil_tbl) <- NULL
knitr::kable(sil_tbl,
             col.names = c("Data level",
                           "Mixture (pos)", "Lab (pos)",
                           "Mixture (neg)", "Lab (neg)"),
             caption = "Mean silhouette per (data level, polarity). Higher Mixture column = same-mixture samples cluster together (cross-lab agreement). Higher Lab column = labs cluster apart (lab effect dominates). Healthy progression: Mixture rises, Lab falls.")
```

```{r silhouette-plot}
#| fig-cap: "Mean silhouette across the four data levels. Solid = labels by mixture (we want this to rise — same-mixture samples cluster). Dashed = labels by lab (we want this to fall — lab effect shrinks). Each line is one polarity."
#| fig-height: 4
sil_long <- sil_tbl |>
  tidyr::pivot_longer(-level, names_to = "key", values_to = "silhouette") |>
  tidyr::separate(key, into = c("metric", "by", "polarity"), sep = "_") |>
  mutate(level = factor(level, levels = c("Full", "Detected",
                                           "Annotated", "Consensus")),
         label = factor(by, levels = c("mixture", "lab"),
                        labels = c("by Mixture", "by Lab")),
         polarity = ifelse(polarity == "pos", "Positive", "Negative"))
ggplot(sil_long, aes(x = level, y = silhouette,
                     colour = polarity, linetype = label,
                     group = interaction(polarity, label))) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.4) +
  scale_color_manual(values = c(Positive = "#E64B35", Negative = "#3B7C92")) +
  labs(x = NULL, y = "Mean silhouette",
       colour = NULL, linetype = NULL,
       title = "Unsupervised clustering: do samples group by mixture (we want) or by lab (we don't)?",
       subtitle = "Healthy progression: solid (mixture) rises, dashed (lab) falls as we move Full -> Consensus") +
  theme_bw() +
  theme(legend.position = "top")
```

### 1.3.3 UMAP of Consensus Peaks

UMAP with cosine distance handles the sparse / presence–absence
consensus matrix better than Euclidean PCA — zero-inflation no longer
dominates the layout.

```{r umap-consensus}
#| fig-cap: "UMAP of Consensus Peaks. Tight per-mixture clusters across labs."
#| fig-width: 10
#| fig-height: 6
library(uwot)

run_umap <- function(pca_list, polarity_label) {
  um <- umap(log1p(pca_list$data),
             n_neighbors = 10, min_dist = 0.2,
             metric = "cosine", scale = TRUE)
  data.frame(UMAP1 = um[, 1], UMAP2 = um[, 2],
             lab = pca_list$meta$lab,
             mixture = pca_list$meta$mixture,
             polarity = polarity_label)
}

umap_scores <- bind_rows(
  run_umap(pca_consensus_pos, "Positive"),
  run_umap(pca_consensus_neg, "Negative")
) |>
  mutate(polarity = factor(polarity, levels = c("Positive", "Negative")))

ggplot(umap_scores,
       aes(x = UMAP1, y = UMAP2, color = mixture, shape = lab)) +
  geom_point(size = 3, alpha = 0.9) +
  facet_wrap(~ polarity) +
  scale_color_hue() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = "UMAP — Consensus Peaks",
       subtitle = "Sparse-aware projection: per-mixture clusters across labs",
       color = "Mixture", shape = "Laboratory") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")
```

### 1.3.4 Consensus presence/absence heatmap

Each consensus peak is observed in only one biological mixture across
labs — the matrix is highly sparse and block-structured. This is why
Euclidean PCA on the raw consensus matrix is dominated by zeros, and
why log1p / cosine alternatives perform better.

```{r consensus-presence-heatmap}
#| fig-cap: "Presence/absence heatmap of Consensus Peaks (positive mode). Rows = samples; columns = consensus peaks. Block structure shows each mixture has its own peak subset."
#| fig-width: 12
#| fig-height: 8
library(pheatmap)

consensus_binary <- ifelse(pca_consensus_pos$data > 0, 1, 0)
keep_cols <- colSums(consensus_binary) > 0
consensus_binary <- consensus_binary[, keep_cols]

annotation_df <- data.frame(
  Mixture = pca_consensus_pos$meta$mixture,
  Lab     = pca_consensus_pos$meta$lab
)
rownames(annotation_df) <- rownames(consensus_binary)

ann_colors <- list(
  Lab = c(afekta = "#E57373", hmgu = "#64B5F6",
          icl    = "#81C784", cembio = "#FFD54F")
)

pheatmap(consensus_binary,
         cluster_rows = FALSE, cluster_cols = FALSE,
         show_rownames = FALSE, show_colnames = FALSE,
         annotation_row = annotation_df,
         annotation_colors = ann_colors,
         main = "Consensus Peak Presence/Absence\nRows = Samples | Columns = Consensus Peaks",
         border_color = NA)
```

---

# Phase 5: Peak Feature Analysis — Consensus Peaks {#phase5}

This phase extends the global analysis by incorporating peak-level metrics
computed from EICs of consensus peaks detected across all 4 laboratories.
This provides a more refined view of inter-lab reproducibility at the
individual peak level.

**Scope note.** Phase 5 reads `peak_metrics_detected_all4labs.csv` from
`4_consensus_peaks.qmd`, which matches **xcms-detected** peaks across
labs without any spike-in identity. It is independent of the SIRIUS
producer cluster-RT algorithm — re-running the producer does **not**
require re-rendering `4_consensus_peaks.qmd`. Phase 5 answers a
different question: "for the same chromatographic feature detected in
all 4 labs, how reproducible is the peak shape / intensity?" — the
flip side of the Annotated row, where Phase 5 measures *detection-level*
agreement and the Annotated row measures *identity-level* abundance.

**Matching basis.** The peak_metrics table is produced by
`4_consensus_peaks.qmd` under **NAPS-aligned matching with raw-RT
fallback** (20 ppm / 15 s), which finds 3 490 four-lab features; after
the peak-quality filters here, 1 787 four-lab consensus peaks remain.
This is the detection-level (xcms peak shape/intensity) view; the
identity-level (SIRIUS-annotated, normalized) ICC ladder is reported
in `5_normalization.qmd` §P3.5–§P3.6.

## 5.1 Data Overview

The peak feature data comes from `4_consensus_peaks.qmd`, which:

1. Created a consensus peak table matching peaks across labs (mz tol = 20 ppm, rt tol = 15 s, on NAPS-aligned RT)
2. Filtered to peaks detected in all 4 labs (1787 consensus peaks after QC; 7148 lab rows)
3. Extracted EICs and computed peak metrics (FWHM, area, SNR, etc.)
4. Applied quality filters (prominence, peak width, FWHM validity)

```{r load-peak-metrics}
pk_metrics <- read.csv(file.path(base_path, "object/peak_metrics_detected_all4labs.csv"))

cat("Peak metrics dimensions:", nrow(pk_metrics), "rows x", ncol(pk_metrics), "columns\n")
cat("Unique consensus peaks:", length(unique(pk_metrics$consensus_id)), "\n")
cat("Labs represented:", paste(unique(pk_metrics$lab), collapse = ", "), "\n")
```

## 5.2 Spearman Intensity Rank Correlation

We use Spearman (rank-based) correlation to test whether peaks that are
relatively intense in one lab are also relatively intense in another.
Spearman is robust to the non-linear scale differences across labs without
requiring explicit normalization — it compares rank order, not absolute
values.

### 5.2.1 Intensity Correlation Matrix

```{r peak-metrics-correlation}
# Pivot to wide format for correlation
intensity_wide <- pk_metrics |>
  select(consensus_id, lab, areaUnderTic, maxIntensity, mixture, polarity) |>
  pivot_wider(
    id_cols = c(consensus_id, mixture, polarity),
    names_from = lab,
    values_from = c(areaUnderTic, maxIntensity)
  )

# Spearman correlation for areaUnderTic
area_cols <- paste0("areaUnderTic_", c("afekta", "hmgu", "icl", "cembio"))
area_matrix <- intensity_wide[, area_cols]
names(area_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_area <- cor(area_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("Spearman Correlation - Area Under TIC:\n")
print(round(cor_area, 3))

# Spearman correlation for maxIntensity
max_cols <- paste0("maxIntensity_", c("afekta", "hmgu", "icl", "cembio"))
max_matrix <- intensity_wide[, max_cols]
names(max_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_max <- cor(max_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("\nSpearman Correlation - Max Intensity:\n")
print(round(cor_max, 3))
```

### 5.2.2 Correlation Heatmap Visualization

```{r peak-correlation-heatmap}
#| fig-cap: "Spearman rank correlation heatmaps for recalculated peak intensities across 4 labs."
#| fig-height: 5
library(pheatmap)

pheatmap(cor_area,
         main = "Area Under TIC (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)

pheatmap(cor_max,
         main = "Max Intensity (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)
```

Afekta and Cembio show the strongest rank correlation (ρ ≈ 0.76),
consistent with their shared Agilent Q-TOF platform. All other pairs
are moderate (ρ ≈ 0.4–0.5), indicating that relative peak intensity
rankings are partially conserved across labs but with substantial
instrument-specific variation.

## 5.3 Peak Metric Distributions Across Labs

The following metrics are computed for each EIC from the filtered consensus
peaks:

- **Peak Width**: Time span (seconds) between the detected peak boundaries.
  Reflects chromatographic resolution and column performance.
- **FWHM** (Full Width at Half Maximum): Width of the peak at half its
  maximum intensity. A narrower, more symmetric measure of peak shape than
  peak width.
- **Signal-to-Noise Ratio (SNR)**: Ratio of peak maximum intensity to the
  estimated baseline. Higher SNR indicates a cleaner, more confident
  detection.

### 5.3.1 Peak Width Distribution

```{r peak-width-distribution}
ggplot(pk_metrics |> filter(!is.na(peakWidth)),
       aes(x = lab, y = peakWidth, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Peak Width Distribution by Lab",
   subtitle = "Consensus peaks detected in all 4 labs",
   x = "Laboratory",
   y = "Peak Width (seconds)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

### 5.3.2 FWHM Distribution

```{r fwhm-distribution}
ggplot(pk_metrics |> filter(!is.na(xicFwhm)), aes(x = lab, y = xicFwhm, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "FWHM Distribution by Lab",
   subtitle = "Full Width at Half Maximum of detected peaks",
   x = "Laboratory",
   y = "FWHM (seconds)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

### 5.3.3 Signal-to-Noise Ratio Distribution

```{r snr-distribution}
ggplot(pk_metrics |> filter(!is.na(signalToNoiseRatio)),
       aes(x = lab, y = log10(signalToNoiseRatio + 1), fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Signal-to-Noise Ratio Distribution by Lab",
   subtitle = "log10(SNR + 1) scale",
   x = "Laboratory",
   y = "log10(SNR + 1)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

## 5.4 Inter-Lab Variability for Consensus Peaks

For each consensus peak (detected in all 4 labs), we compute the coefficient
of variation (CV = standard deviation / mean) of each metric across the 4
labs. CV is inherently scale-normalized: it measures relative spread
regardless of absolute intensity differences, so no prior normalization of
intensity metrics is needed.

- **CV < 0.3 (30%)**: Good reproducibility
- **CV 0.3–0.5**: Moderate variability
- **CV > 0.5**: High variability

**IQR** (Interquartile Range): the difference between the 75th and 25th
percentiles of the CV distribution, measuring the spread of the middle 50%
of values.

### 5.4.1 CV Summary Table

```{r cv-summary-table}
cv_by_consensus <- pk_metrics |>
  group_by(consensus_id) |>
  summarise(
    n_labs = n(),
    cv_area = sd(areaUnderTic, na.rm = TRUE) / mean(areaUnderTic, na.rm = TRUE),
    cv_max = sd(maxIntensity, na.rm = TRUE) / mean(maxIntensity, na.rm = TRUE),
    cv_fwhm = sd(xicFwhm, na.rm = TRUE) / mean(xicFwhm, na.rm = TRUE),
    cv_peakwidth = sd(peakWidth, na.rm = TRUE) / mean(peakWidth, na.rm = TRUE),
    cv_snr = sd(signalToNoiseRatio, na.rm = TRUE) / mean(signalToNoiseRatio, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_labs == 4)

cv_summary <- cv_by_consensus |>
  summarise(
    across(starts_with("cv_"), list(
      median = ~median(.x, na.rm = TRUE),
      q25 = ~quantile(.x, 0.25, na.rm = TRUE),
      q75 = ~quantile(.x, 0.75, na.rm = TRUE)
    ))
  )

cv_table <- data.frame(
  Metric = c("Area Under TIC", "Max Intensity", "FWHM", "Peak Width", "SNR"),
  Median_CV = c(cv_summary$cv_area_median, cv_summary$cv_max_median,
                cv_summary$cv_fwhm_median, cv_summary$cv_peakwidth_median,
                cv_summary$cv_snr_median),
  IQR = c(cv_summary$cv_area_q75 - cv_summary$cv_area_q25,
          cv_summary$cv_max_q75 - cv_summary$cv_max_q25,
          cv_summary$cv_fwhm_q75 - cv_summary$cv_fwhm_q25,
          cv_summary$cv_peakwidth_q75 - cv_summary$cv_peakwidth_q25,
          cv_summary$cv_snr_q75 - cv_summary$cv_snr_q25)
)

knitr::kable(cv_table,
             caption = "Inter-Lab CV for Consensus Peaks (median and interquartile range)",
             digits = 3)
```

### 5.4.2 CV Distribution Visualization

```{r cv-distribution}
#| fig-cap: "CV distribution across consensus peaks for each metric. Red dashed line = 30% CV threshold."
cv_long <- cv_by_consensus |>
 pivot_longer(
   cols = starts_with("cv_"),
   names_to = "metric",
   names_prefix = "cv_",
   values_to = "cv"
 ) |>
 mutate(metric = case_when(
   metric == "area" ~ "Area Under TIC",
   metric == "max" ~ "Max Intensity",
   metric == "fwhm" ~ "FWHM",
   metric == "peakwidth" ~ "Peak Width",
   metric == "snr" ~ "SNR"
 ))

ggplot(cv_long |> filter(!is.na(cv) & cv < 2), aes(x = metric, y = cv, fill = metric)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 geom_hline(yintercept = 0.3, linetype = "dashed", color = "red") +
 annotate("text", x = 5.3, y = 0.35, label = "CV = 30%", color = "red", size = 3) +
 labs(
   title = "Inter-Lab Variability for Consensus Peaks",
   subtitle = "Lower CV = better reproducibility across labs",
   x = "Metric",
   y = "Coefficient of Variation"
 ) +
 theme_bw() +
 theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

Shape metrics (FWHM, peak width) show lower CV than intensity metrics
(area, max intensity), confirming that chromatographic behavior is more
reproducible across labs than ionization response. SNR shows the highest
variability, reflecting differences in baseline estimation across
instruments.

---

# Summary {#summary}

This file shows the **data-level progression** story:

1. **Lab variance dominates raw data**. At the Full Signal level,
   97–100 % of TIC variance is lab-attributable. PCA shape clusters
   (lab dominates), mixture colours scattered.
2. **Peak detection recovers some biology**. Detected Signal PCA shape
   clusters loosen, colour grouping emerges. xcms peak picking strips
   most of the lab-specific background.
3. **Annotation concentrates the biology further**. Annotated PCA
   (sample × compound maxIntensity) tightens mixture clusters; lab
   spread within each mixture shrinks. SIRIUS-keyed features carry
   identity-aware abundance.
4. **Consensus peaks deliver the cleanest picture**. The cross-lab
   matched peak set (~1.8 k features) gives tight per-mixture clusters
   in PCA + UMAP — same-mixture points across all four labs sit
   together. Per-peak inter-lab CV distribution shows the broader
   feature space behaves like the spike-in panel.

For the **normalization-strategy evaluation** (Raw / Aligned / Lab-
median / Global / Decile / LOESS with the full ICC ladder × pairwise
heatmap), see `5_normalization.qmd` §P3.4–§P3.6. For the **RT
alignment**, see `3_rt_alignment.qmd`. For the **per-compound
diagnostics**, see `8_per_compound_diagnostics.qmd`.

---

# Pairwise reproducibility on the winning normalization

The full strategy comparison (six normalizations × both metrics × both
compound-set policies) lives in `5_normalization.qmd` §P3.5–§P3.6.
Below: the same pairwise structure, **restricted to Raw vs Lab-median
only** (the §P3.4 winner for Annotated Signal). Two views:

1. **Bulk-level pairwise ICC** — per (lab pair × strategy), how well
   do the two labs agree on each mixture's total annotated signal?
2. **Per-compound pairwise correlation** — same lab pairs, but on
   *individual* (mixture × compound) intensities rather than summed
   bulk. Spearman ρ, so a pair can have high bulk ICC and low ρ if
   the labs agree on totals only because their compound-level errors
   average out.

```{r reprod-load-normalized}
#| message: false
source(here("5_downstream_analysis", "helpers.R"))

producer_dir <- here("5_downstream_analysis", "object")
ann_norm <- read.csv(
  file.path(producer_dir, "annotation_chrom_metrics_normalized.csv"),
  stringsAsFactors = FALSE,
  colClasses = c(mixture = "character"))
ann_norm$mixture <- gsub("\\.", "_", ann_norm$mixture)

# Strict subset (4/4 labs, all aligned).
well_detected_w <- ann_norm |>
  filter(!is.na(maxIntensity), maxIntensity > 0) |>
  distinct(mixture, lab, compound, polarity) |>
  group_by(mixture, compound, polarity) |>
  filter(n_distinct(lab) == 4L) |>
  ungroup() |>
  distinct(mixture, compound, polarity)
ann_well_w <- ann_norm |>
  inner_join(well_detected_w, by = c("mixture", "compound", "polarity")) |>
  group_by(lab, mixture, polarity, compound) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup()
well_aligned_w <- ann_well_w |>
  filter(!is.na(aligned_rt)) |>
  count(mixture, compound, polarity, name = "n") |>
  filter(n == 4L) |>
  select(mixture, compound, polarity)
ann_well_aligned_w <- ann_well_w |>
  inner_join(well_aligned_w, by = c("mixture", "compound", "polarity"))

cat("Strict (4/4 aligned) (mix, cpd, pol) tuples:",
    nrow(well_aligned_w), "\n")
```

```{r reprod-pairwise-bulk-raw-vs-labmed}
#| fig-cap: "Pairwise ICC on Raw vs Lab-median, Strict subset. ★ = same-platform pair (afekta + cembio). The Lab-median column shows what the §P3.4 winner buys for each lab pair: the closer a tile is to the ★ tile in the same column, the closer that pair gets to the same-platform ceiling."
#| fig-height: 5

lab_pairs_w <- list(
  c("afekta", "cembio"), c("afekta", "hmgu"), c("afekta", "icl"),
  c("cembio", "hmgu"),   c("cembio", "icl"),  c("hmgu",   "icl"))
agilent_w <- c("afekta", "cembio")

# Bulk per (lab, mixture, polarity, strategy=Raw|Lab-median).
bulk_raw <- ann_well_aligned_w |>
  group_by(lab, mixture, polarity) |>
  summarise(bulk = sum(maxIntensity, na.rm = TRUE),
            .groups = "drop") |>
  mutate(strategy = "Raw")
lab_med <- bulk_raw |>
  group_by(lab, polarity) |>
  summarise(med = median(bulk, na.rm = TRUE), .groups = "drop")
bulk_labmed <- bulk_raw |>
  left_join(lab_med, by = c("lab", "polarity")) |>
  mutate(bulk = bulk / med, strategy = "Lab-median") |>
  select(-med)
bulk_two <- bind_rows(bulk_raw, bulk_labmed)

icc_pairs_w <- bind_rows(lapply(lab_pairs_w, function(p) {
  pair_label <- paste(sort(p), collapse = " + ")
  bulk_pair  <- bulk_two |> filter(lab %in% p)
  bind_rows(lapply(c("pos", "neg"), function(pol_str) {
    pol_name <- if (pol_str == "pos") "Positive" else "Negative"
    bind_rows(lapply(c("Raw", "Lab-median"), function(s) {
      df <- bulk_pair |> filter(polarity == pol_str, strategy == s)
      if (!nrow(df) || all(is.na(df$bulk))) return(NULL)
      res <- compute_icc(df, "bulk")
      res$strategy      <- s
      res$polarity      <- pol_name
      res$pair          <- pair_label
      res$same_platform <- all(p %in% agilent_w)
      res
    }))
  }))
}))

ggplot(icc_pairs_w |>
         mutate(strategy = factor(strategy, levels = c("Raw", "Lab-median")),
                pair_label = ifelse(same_platform,
                                     paste(pair, "★"), pair)),
       aes(x = strategy, y = reorder(pair_label, ICC), fill = ICC)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", ICC)), size = 3.4) +
  facet_wrap(~ polarity) +
  scale_fill_gradient2(low = "white", mid = "#FDC086", high = "#1F77B4",
                       midpoint = 0.5, limits = c(0, 1), name = "ICC") +
  labs(x = NULL, y = "Lab pair (★ = same platform)",
       title = "Pairwise ICC — Raw vs Lab-median (the §P3.4 winner)",
       subtitle = sprintf("Strict subset (n = %d), maxIntensity",
                          nrow(well_aligned_w))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

## Pairwise agreement at the individual-compound level

The plot above computes pairwise ICC on **summed** bulk (one number
per mixture for each lab). Here we instead compute the Spearman ρ
across **individual** (mixture × compound) measurements — so a pair
can have high bulk ICC but low per-compound ρ if errors average out
in the sum.

```{r reprod-per-compound-pairwise}
#| fig-cap: "Spearman ρ between each lab pair, computed on per-(mixture × compound) maxIntensity and areaUnderTic — not bulk. ρ → 1 means the labs rank the individual spike-ins the same way; ρ near 0 means compound-level rankings disagree. Compare to the bulk pairwise ICC above to see when per-compound noise is averaged out by summing."
#| fig-height: 5

wide_per_compound_w <- ann_well_aligned_w |>
  select(mixture, compound, polarity, lab, maxIntensity, areaUnderTic) |>
  pivot_wider(names_from = lab,
              values_from = c(maxIntensity, areaUnderTic))

per_compound_pair_cor_w <- bind_rows(lapply(lab_pairs_w, function(p) {
  bind_rows(lapply(c("pos", "neg"), function(pol_str) {
    pol_name <- if (pol_str == "pos") "Positive" else "Negative"
    df <- wide_per_compound_w |> filter(polarity == pol_str)
    bind_rows(lapply(c("maxIntensity", "areaUnderTic"), function(m) {
      a <- df[[paste0(m, "_", p[1])]]
      b <- df[[paste0(m, "_", p[2])]]
      ok <- is.finite(a) & a > 0 & is.finite(b) & b > 0
      if (sum(ok) < 5L) return(NULL)
      tibble(pair          = paste(sort(p), collapse = " + "),
             polarity      = pol_name,
             metric        = m,
             n             = sum(ok),
             spearman      = cor(a[ok], b[ok], method = "spearman"),
             same_platform = all(p %in% agilent_w))
    }))
  }))
}))

ggplot(per_compound_pair_cor_w |>
         mutate(pair_label = ifelse(same_platform,
                                     paste(pair, "★"), pair)),
       aes(x = metric, y = reorder(pair_label, spearman), fill = spearman)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", spearman)),
            colour = "black", size = 3) +
  facet_wrap(~ polarity) +
  scale_fill_gradient2(low = "#B40426", mid = "white", high = "#3B4CC0",
                       midpoint = 0.5, limits = c(0, 1), name = "Spearman") +
  labs(x = NULL, y = "Lab pair (★ = same platform)",
       title = "Per-compound pairwise Spearman ρ between lab pairs",
       subtitle = "Computed on individual (mixture × compound) intensities, not summed bulk") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

knitr::kable(per_compound_pair_cor_w |>
               transmute(metric, polarity, pair,
                          Spearman = round(spearman, 3), n) |>
               arrange(metric, polarity, pair),
             caption = "Per-compound pairwise Spearman ρ.")
```

---

# Future directions

- Expand the four-level progression to the other HUMAN matrices
  (FE, Others) beyond HE endosome.
- Add a "post-normalization" data level (Annotated normalized) to the
  PCA progression once the §8 normalization output is the default
  consumer input.


