---
title: "Data-level progression: Full → Detected → Annotated → Consensus"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
---

# Introduction

The same 19 mixtures × 4 labs × 2 polarities are observable at **four
different data levels**, each built by a different stage of the
processing pipeline:

| Level | Built by | Feature space | Per sample summary |
|---|---|---|---|
| **Full**       | mzML, no peak detection | all MS1 ions | TIC, BPC |
| **Detected**   | xcms (per-lab, per-sample peak picking) | per-injection detected peaks | TIC, BPC restricted to detected ions |
| **Annotated**  | SIRIUS MS2 + SNR-aware chromExtract | the spike-in panel (~250 (mix × compound) tuples detected in ≥1 lab) | sum/max of maxIntensity over the panel |
| **Consensus**  | `2_consensus_peaks.qmd` cross-lab feature matching | ~1,787 features matched in all 4 labs | sample × consensus-peak matrix |

This file asks **one question**: *does each successive layer recover
biological signal from technical variance?* The answer is yes — see the
Phase 1 variance decomposition and PCA progression for the headline
story, and the Phase 5 Consensus-peak analysis for the per-peak picture
on the broader feature space (1.8k features vs ~250 in Annotated).

Other axes of the analysis live in dedicated files:

- **Normalization strategies (Global / Decile / LOESS / NAPS)** →
  `5_normalization.qmd` + `7_icc_reproducibility.qmd`
- **Cross-lab ICC ladder, pairwise reproducibility, 3-lab subset
  (no hmgu), Agilent ceiling** → `7_icc_reproducibility.qmd`
- **NAPS RT alignment** → `4_rt_alignment.qmd`
- **Per-compound identity / RT / abundance / m/z diagnostics** →
  `9_per_compound_diagnostics.qmd`
- **Confidence tiering (A / B / C / D)** → `6_confidence_tier.qmd`

Data: 19 standard mixtures (HUMAN endosome) from 4 labs with harmonized
LC systems. No technical replicates; lab identity is confounded with
instrument platform.

## Load Libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(here)
library(patchwork)
library(ggbeeswarm)
library(pheatmap)

base_path     <- here("5_downstream_analysis")
producer_dir  <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                      "results_all_ms2")
source(file.path(base_path, "global_analysis_helpers.R"))
```

## Setup Information

```{r load-setup}
setup <- read.csv(file.path(base_path, "lab_setup_phase1.csv"))

setup_simplified <- setup %>%
  filter(details == "harmonized") %>%
  select(lab, `LC.Instrument`, `MS.Instrument`, `Inj.vol..uL..pos`, `Inj.vol..uL..neg`) %>%
  distinct() %>%
  mutate(
    platform = case_when(
      grepl("Agilent 1290", `LC.Instrument`) ~ "Agilent",
      grepl("Acquity", `LC.Instrument`) ~ "Waters",
      grepl("ExionLC", `LC.Instrument`) ~ "Sciex",
      grepl("Bruker", `LC.Instrument`) ~ "Bruker",
      TRUE ~ "Other"
    ),
    ms_instrument = case_when(
      grepl("6546", `MS.Instrument`) ~ "Agilent_6546",
      grepl("6545", `MS.Instrument`) ~ "Agilent_6545",
      grepl("Synapt", `MS.Instrument`) ~ "Waters_Synapt",
      grepl("Zeno", `MS.Instrument`) ~ "Sciex_Zeno",
      grepl("TIMS", `MS.Instrument`) ~ "Bruker_TIMS",
      TRUE ~ "Other"
    ),
    inj_vol_pos = `Inj.vol..uL..pos`,
    inj_vol_neg = `Inj.vol..uL..neg`
  ) %>%
  select(lab, platform, ms_instrument, inj_vol_pos, inj_vol_neg)

setup_simplified
```

Each lab's protocol is unique; lab identity is completely confounded with
instrumental setup. Notable variations:

- MS1 scan rate
- Injection volume
- Solvent brand.

## Metrics Definition

Three main chromatographic metrics:

- **TIC**: Sum of all ion intensities
- **BPC**: Max intensity per scan

Three data levels (least → most processed):

- **Full Signal**: raw profile including baseline (`bpc_full` / `tic_full`)
- **Detected Signal**: peaks from centWave; isolates compound signal from
  baseline (`bpc_detect` / `tic_detect`)
- **Annotated Signal**: derived from
  `annotation_chrom_metrics.csv` (SIRIUS producer) restricted to
  `(mixture, compound)` pairs with non-NA `maxIntensity` in **all 4
  labs** at that polarity (recomputed each render — the
  "well-detected" set, not hardcoded). Per-sample bulk:
  `tic = sum(areaUnderTic)`; `bpc = max(maxIntensity)`. Polarity is
  collapsed per-side so each polarity has its own well-detected set.

Two polarities: Positive and Negative ionization modes.

```{r load-data-both-modes}
compute_metrics <- function(polarity, polarity_name, data_level = "full") {
  if (data_level == "full") {
    load(file.path(base_path, "object/bpc_full.RData"))
    bpc <- bpc_full[bpc_full$polarity == polarity]
    load(file.path(base_path, "object/tic_full.RData"))
    tic <- tic_full[tic_full$polarity == polarity]
  } else {
    load(file.path(base_path, "object/bpc_detect.RData"))
    bpc <- bpc_detect[bpc_detect$polarity == polarity]
    load(file.path(base_path, "object/tic_detect.RData"))
    tic <- tic_detect[tic_detect$polarity == polarity]
  }

  data.frame(
    lab = bpc$lab,
    mixture = bpc$mixture,
    polarity = polarity_name,
    bpc = vapply(intensity(bpc), sum, numeric(1), na.rm = TRUE),
    tic = vapply(intensity(tic), sum, numeric(1), na.rm = TRUE)
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = ifelse(data_level == "full", "Full Signal", "Detected Signal")) %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm = bpc / median_bpc,
      tic_norm = tic / median_tic
    ) %>%
    ungroup()
}

compute_metrics_annotated <- function(polarity, polarity_name) {
  pol_str <- if (polarity == 1L) "pos" else "neg"

  # mixture forced to character — read.csv would otherwise coerce quoted
  # "2.10" → 2.1 (numeric), silently colliding with mixture "2.1".
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))

  well_detected <- chrom %>%
    filter(polarity == pol_str, !is.na(maxIntensity)) %>%
    distinct(mixture, compound, lab) %>%
    count(mixture, compound) %>%
    filter(n == 4L) %>%
    select(mixture, compound)

  bulk <- chrom %>%
    filter(polarity == pol_str, !is.na(maxIntensity)) %>%
    inner_join(well_detected, by = c("mixture", "compound")) %>%
    group_by(lab, mixture, compound) %>%
    slice_max(maxIntensity, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    group_by(lab, mixture) %>%
    summarise(bpc = max(maxIntensity, na.rm = TRUE),
              tic = sum(areaUnderTic, na.rm = TRUE),
              .groups = "drop")

  data.frame(
    lab = bulk$lab, mixture = bulk$mixture,
    polarity = polarity_name,
    bpc = bulk$bpc, tic = bulk$tic
  ) %>%
    left_join(setup_simplified, by = "lab") %>%
    mutate(data_level = "Annotated Signal") %>%
    group_by(lab, polarity) %>%
    mutate(
      median_bpc = median(bpc, na.rm = TRUE),
      median_tic = median(tic, na.rm = TRUE),
      bpc_norm   = bpc / median_bpc,
      tic_norm   = tic / median_tic
    ) %>%
    ungroup()
}

metrics_full      <- run_for_polarities(compute_metrics, "full")
metrics_detect    <- run_for_polarities(compute_metrics, "detect")
metrics_annotated <- run_for_polarities_names(compute_metrics_annotated)

# Recompute the "275-set" size (per polarity) for the prose / sanity check.
.well_n <- function(p) {
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))
  chrom %>% filter(polarity == p, !is.na(maxIntensity)) %>%
    distinct(mixture, compound, lab) %>%
    count(mixture, compound) %>% filter(n == 4L) %>% nrow()
}
cat("well-detected (mixture, compound) pairs — pos:", .well_n("pos"),
    " | neg:", .well_n("neg"), "\n")

metrics_combined <- bind_rows(metrics_full, metrics_detect, metrics_annotated) %>%
  mutate(data_level = factor(data_level,
                             levels = c("Full Signal",
                                        "Detected Signal",
                                        "Annotated Signal"))) %>%
  group_by(lab, polarity, data_level) %>%
  mutate(
    bpc_median_scaled = bpc / median(bpc, na.rm = TRUE),
    tic_median_scaled = tic / median(tic, na.rm = TRUE),
    bpc_autoscaled = (bpc - mean(bpc, na.rm = TRUE)) / mad(bpc, na.rm = TRUE),
    tic_autoscaled = (tic - mean(tic, na.rm = TRUE)) / mad(tic, na.rm = TRUE)
  ) %>%
  ungroup()

# Reusable ordered character vector for the few loops that iterate levels.
data_levels_all <- c("Full Signal", "Detected Signal", "Annotated Signal")
```

---

# Phase 1: The Problem — Raw Data Assessment {#phase1}

This phase characterizes the raw data to understand the magnitude of
inter-laboratory variability before any corrections are applied.

## 1.1 Distribution of Raw Metrics

### 1.1.1 Violin Plots: Overall Intensity Distribution

```{r violin-plots}
#| fig-cap: "BPC and TIC distributions by lab, polarity, and data level."
for (metric in c("bpc", "tic")) print(plot_violin(metrics_combined, metric))
```

Detected signal recenters the negative and positive distributions.
The bimodal shape is evident.

### 1.1.2 Density Plots: Distribution Shape by Lab

```{r density-plots}
ggplot(metrics_combined %>% filter(data_level == "Full Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_bw() +
  labs(title = "TIC Distribution by Lab (Full Signal)",
       x = "log2(TIC)", y = "Density")

ggplot(metrics_combined %>% filter(data_level == "Detected Signal"),
       aes(x = log2(tic), fill = polarity)) +
  geom_density(alpha = 0.3) +
  facet_grid(lab ~ polarity) +
  theme_bw() +
  labs(title = "TIC Distribution by Lab (Detected Signal)",
       x = "log2(TIC)", y = "Density")
```

Full signal is more compressed than detected signal, likely because
each mixture has different numbers of compounds.

### 1.1.3 Mixture-Level Agreement Across Labs

```{r mixture-distribution}
plot_mixture_distribution <- function(data, metric) {
  if (metric %in% c("bpc", "tic")) {
    data <- data |> mutate(across(all_of(metric), log2))
    title_suffix <- "(log2)"
  } else {
    title_suffix <- ""
  }

  ggplot(data, aes(x = mixture, y = .data[[metric]], color = lab, group = lab)) +
    geom_line(alpha = 0.7) +
    facet_grid(data_level ~ polarity) +
    theme_bw() +
    labs(title = paste("Mixture Distribution Across Labs -", metric, title_suffix),
         x = "Mixture", y = paste(metric, title_suffix), color = "Laboratory")
}

for (metric in c("bpc", "tic")) {
  print(plot_mixture_distribution(metrics_combined, metric))
}
```

There is good agreement between Cembio and Afekta, and some with ICL.
However, HMGU is consistently lower intensity.

## 1.2 Variance Decomposition: Quantifying the Problem

### 1.2.1 Linear Model Analysis

Let's run linear models to evaluate mixture effects and batch effects.
The goal is to quantify how much of the variance is due to lab vs mixture.
We use ANOVA to aggregate coefficients into Sum of Squares (SS), which
quantifies the relative power of each factor.

We run this analysis across all 4 conditions (positive/negative, full/detected):

```{r variance-decomposition-lm}
results_table <- data.frame()
residual_plots <- list()
all_coefficients <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in data_levels_all) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")

    residual_plots[[paste(pol, level, sep = " - ")]] <- result$residual_plot
    all_coefficients <- rbind(all_coefficients,
      result$coefficients %>% mutate(Polarity = pol, Data_Level = level))
    results_table <- rbind(results_table, data.frame(
      Polarity = pol, Data_Level = level, R_squared = result$r_squared,
      Lab_Variance_Pct = result$variance_table$Percentage[1],
      Mixture_Variance_Pct = result$variance_table$Percentage[2],
      Residual_Variance_Pct = result$variance_table$Percentage[3]
    ))
  }
}
```

### 1.2.2 Lab Effect Coefficients

```{r lab-coefficients}
#| fig-cap: "Lab effect coefficients from linear models. Deviation from reference lab (Afekta); negative = lower intensity."
lab_coefs <- all_coefficients %>%
  filter(type == "Lab Effect") %>%
  mutate(lab_name = gsub("lab", "", coefficient))

ggplot(lab_coefs, aes(x = lab_name, y = estimate, fill = lab_name)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_fill_hue() +
  labs(title = "Lab Effect Coefficients from Linear Models",
       subtitle = "Deviation from reference lab (afekta); negative = lower intensity",
       x = "Laboratory", y = "Coefficient (log2 scale)", fill = "Laboratory") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.2.3 Variance Decomposition Summary

```{r variance-stacked-bar}
#| fig-cap: "Variance decomposition of TIC across all conditions. Lab variance dominates in raw data."
variance_long <- results_table %>%
  select(Polarity, Data_Level, R_squared, ends_with("_Pct")) %>%
  pivot_longer(cols = ends_with("_Pct"), names_to = "Source", values_to = "Percentage") %>%
  mutate(
    Source = gsub("_Variance_Pct|_Pct", "", Source),
    Label = paste(Polarity, Data_Level, sep = "\n"),
    R2_label = paste("R² =", round(R_squared, 3))
  )

text_labels <- variance_long %>% distinct(Label, R2_label)

ggplot(variance_long, aes(x = Label, y = Percentage, fill = Source)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(x = Label, label = R2_label, y = 105), data = text_labels,
            size = 3, inherit.aes = FALSE) +
  labs(title = "Variance Decomposition: TIC Across All Conditions",
       subtitle = "Raw data shows massive lab variance — the core problem",
       x = "Polarity & Data Level", y = "Variance Contribution (%)", fill = "Source") +
  theme_bw() +
  theme(legend.position = "top") +
  ylim(0, 110)
```

### 1.2.4 Residual Diagnostics

```{r residual-plots}
#| fig-height: 10
residual_data <- data.frame()

for (pol in c("Positive", "Negative")) {
  for (level in data_levels_all) {
    data_subset <- metrics_combined %>% filter(polarity == pol, data_level == level)
    result <- check_metric_quality(data_subset, "tic")
    residual_data <- rbind(residual_data, data.frame(
      Fitted = fitted(result$model), Residuals = residuals(result$model),
      Polarity = pol, Data_Level = level,
      Lab = data_subset$lab, Mixture = as.factor(data_subset$mixture)
    ))
  }
}

ggplot(residual_data %>%
         mutate(Data_Level = factor(Data_Level, levels = data_levels_all)),
       aes(x = Fitted, y = Residuals, color = Mixture, shape = Lab)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Data_Level ~ Polarity) +
  scale_color_hue() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(title = "Fitted vs Residuals: TIC Across All Conditions",
       subtitle = "Points should scatter randomly around zero for a good model fit",
       x = "Fitted Values", y = "Residuals", color = "Mixture", shape = "Laboratory") +
  theme_bw() +
  theme(legend.position = "bottom")

results_table
```

**Key Finding:** Overall variance is dominated by lab effects.
The residual plot clearly shows separation between labs vs HMGU and
still lab separation in others. Lab separation remains unexplained by
mixture. We cannot compare labs directly without normalization.

## 1.3 Multivariate visualization (PCA / UMAP / presence–absence)

Colour = mixture, shape = lab. Same-shape clustering means the lab
dominates the layout; same-colour clustering means mixture (biology)
dominates. The four-step progression
**Full → Detected → Annotated → Consensus** is the same problem from
the variance decomposition above, viewed in 2D.

### 1.3.1 Build PCA matrices

```{r pca-helpers}
prepare_pca_tic <- function(data_level, polarity_int) {
  if (data_level == "full") {
    load(file.path(base_path, "object/tic_full.RData"))
    chr_data <- tic_full[tic_full$polarity == polarity_int]
  } else {
    load(file.path(base_path, "object/tic_detect.RData"))
    chr_data <- tic_detect[tic_detect$polarity == polarity_int]
  }

  int_list <- intensity(chr_data); rt_list <- rtime(chr_data)
  all_rts  <- unlist(rt_list)
  rt_grid  <- seq(min(all_rts, na.rm = TRUE),
                  max(all_rts, na.rm = TRUE), length.out = 500)

  int_matrix <- t(sapply(seq_along(int_list), function(i) {
    approx(rt_list[[i]], int_list[[i]], xout = rt_grid,
           method = "linear", rule = 2)$y
  }))
  rownames(int_matrix) <- paste0(chr_data$lab, "_", chr_data$mixture)
  colnames(int_matrix) <- paste0("RT_", round(rt_grid, 1))

  list(data = int_matrix,
       meta = data.frame(sample  = rownames(int_matrix),
                         lab     = chr_data$lab,
                         mixture = chr_data$mixture))
}

# Annotated: sample × compound maxIntensity, restricted to the
# well-detected set (same definition as the bulk Annotated metric above).
prepare_pca_annotated <- function(polarity_int) {
  pol_str <- if (polarity_int == 1L) "pos" else "neg"
  chrom <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                    stringsAsFactors = FALSE,
                    colClasses = c(mixture = "character"))

  well_detected <- chrom |>
    filter(polarity == pol_str, !is.na(maxIntensity)) |>
    distinct(mixture, compound, lab) |>
    count(mixture, compound) |>
    filter(n == 4L) |>
    select(mixture, compound)

  wide <- chrom |>
    filter(polarity == pol_str, !is.na(maxIntensity)) |>
    inner_join(well_detected, by = c("mixture", "compound")) |>
    group_by(lab, mixture, compound) |>
    slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
    ungroup() |>
    mutate(sample = paste0(lab, "_", mixture)) |>
    select(sample, compound, maxIntensity, lab, mixture) |>
    pivot_wider(names_from = compound, values_from = maxIntensity,
                values_fill = 0) |>
    arrange(sample)

  data_mat <- wide |> select(-sample, -lab, -mixture) |> as.matrix()
  rownames(data_mat) <- wide$sample
  list(data = data_mat, meta = wide |> select(sample, lab, mixture))
}

# Consensus: sample × consensus-peak matrix from 3_peak_feature output.
prepare_pca_consensus <- function(polarity_int) {
  pk_metrics <- read.csv(file.path(base_path,
                                   "object/peak_metrics_detected_all4labs.csv"),
                         stringsAsFactors = FALSE,
                         colClasses = c(mixture = "character"))
  wide <- pk_metrics |>
    filter(as.integer(polarity) == polarity_int) |>
    mutate(sample = paste0(lab, "_", mixture)) |>
    select(sample, consensus_id, areaUnderTic, lab, mixture) |>
    pivot_wider(names_from = consensus_id, values_from = areaUnderTic,
                values_fill = 0, names_prefix = "peak_") |>
    arrange(sample)

  data_mat <- wide |> select(starts_with("peak_")) |> as.matrix()
  rownames(data_mat) <- wide$sample
  list(data = data_mat, meta = wide |> select(sample, lab, mixture))
}

# Run prcomp, drop zero-variance columns so scale.=TRUE doesn't error.
pca_scores_df <- function(pca_list, polarity_label, transform_fn = identity) {
  X <- transform_fn(pca_list$data)
  keep <- apply(X, 2, function(z) is.finite(var(z)) && var(z) > 0)
  res <- prcomp(X[, keep, drop = FALSE], scale. = TRUE, center = TRUE)
  var_expl <- (res$sdev^2 / sum(res$sdev^2)) * 100
  list(scores = data.frame(PC1 = res$x[, 1], PC2 = res$x[, 2],
                           lab = pca_list$meta$lab,
                           mixture = pca_list$meta$mixture,
                           polarity = polarity_label),
       var_expl = var_expl)
}

plot_pca_by_polarity <- function(pos, neg, title, subtitle) {
  scores <- bind_rows(pos$scores, neg$scores) |>
    mutate(polarity = factor(polarity, levels = c("Positive", "Negative")))
  panel_labels <- c(
    Positive = sprintf("Positive (PC1 %.1f%%, PC2 %.1f%%)",
                       pos$var_expl[1], pos$var_expl[2]),
    Negative = sprintf("Negative (PC1 %.1f%%, PC2 %.1f%%)",
                       neg$var_expl[1], neg$var_expl[2])
  )
  ggplot(scores, aes(x = PC1, y = PC2, color = mixture, shape = lab)) +
    geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    facet_wrap(~ polarity, scales = "free",
               labeller = labeller(polarity = panel_labels)) +
    scale_color_hue() +
    labs(title = title, subtitle = subtitle,
         x = "PC1", y = "PC2", color = "Mixture", shape = "Laboratory") +
    theme_bw() +
    theme(plot.title = element_text(face = "bold"),
          legend.position = "right")
}
```

```{r prepare-pca-data}
pca_full_pos      <- prepare_pca_tic("full",   1L)
pca_full_neg      <- prepare_pca_tic("full",   0L)
pca_detect_pos    <- prepare_pca_tic("detect", 1L)
pca_detect_neg    <- prepare_pca_tic("detect", 0L)
pca_annotated_pos <- prepare_pca_annotated(1L)
pca_annotated_neg <- prepare_pca_annotated(0L)
pca_consensus_pos <- prepare_pca_consensus(1L)
pca_consensus_neg <- prepare_pca_consensus(0L)
```

### 1.3.2 PCA progression Full → Detected → Annotated → Consensus

```{r pca-full-tic}
#| fig-cap: "PCA of Raw TIC profiles. Shapes cluster (lab dominates); colours scattered (mixture signal buried in technical variance)."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_full_pos, "Positive"),
  neg = pca_scores_df(pca_full_neg, "Negative"),
  title = "PCA — Full Signal (Raw TIC profiles)",
  subtitle = "Shapes cluster, colours scattered: technical noise dominates"
)
```

```{r pca-detected-tic}
#| fig-cap: "PCA of Detected TIC. Shape clusters loosen, colour grouping emerges — peak detection recovers biological signal."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_detect_pos, "Positive"),
  neg = pca_scores_df(pca_detect_neg, "Negative"),
  title = "PCA — Detected Signal",
  subtitle = "Shapes disperse, colours begin to group: mixture signal emerges"
)
```

```{r pca-annotated}
#| fig-cap: "PCA of Annotated signal (sample × compound maxIntensity over the well-detected set). Identity-keyed features concentrate biological variance further; mixture clusters tighten and lab spread within each mixture shrinks."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_annotated_pos, "Positive", transform_fn = log1p),
  neg = pca_scores_df(pca_annotated_neg, "Negative", transform_fn = log1p),
  title = "PCA — Annotated Signal (per-compound maxIntensity)",
  subtitle = "Restricted to (mixture, compound) pairs detected in all 4 labs"
)
```

```{r pca-consensus}
#| fig-cap: "PCA of Consensus Peaks (matched across all 4 labs). Same-colour points cluster across shapes — biology dominates. log1p compresses the presence/absence gap that would otherwise dominate scaling on the sparse matrix."
#| fig-width: 12
#| fig-height: 6
plot_pca_by_polarity(
  pos = pca_scores_df(pca_consensus_pos, "Positive", transform_fn = log1p),
  neg = pca_scores_df(pca_consensus_neg, "Negative", transform_fn = log1p),
  title = "PCA — Consensus Peaks (cross-lab matched signal)",
  subtitle = "Colours cluster across shapes: biology dominates"
)
```

### 1.3.3 UMAP of Consensus Peaks

UMAP with cosine distance handles the sparse / presence–absence
consensus matrix better than Euclidean PCA — zero-inflation no longer
dominates the layout.

```{r umap-consensus}
#| fig-cap: "UMAP of Consensus Peaks. Tight per-mixture clusters across labs."
#| fig-width: 10
#| fig-height: 6
library(uwot)

run_umap <- function(pca_list, polarity_label) {
  um <- umap(log1p(pca_list$data),
             n_neighbors = 10, min_dist = 0.2,
             metric = "cosine", scale = TRUE)
  data.frame(UMAP1 = um[, 1], UMAP2 = um[, 2],
             lab = pca_list$meta$lab,
             mixture = pca_list$meta$mixture,
             polarity = polarity_label)
}

umap_scores <- bind_rows(
  run_umap(pca_consensus_pos, "Positive"),
  run_umap(pca_consensus_neg, "Negative")
) |>
  mutate(polarity = factor(polarity, levels = c("Positive", "Negative")))

ggplot(umap_scores,
       aes(x = UMAP1, y = UMAP2, color = mixture, shape = lab)) +
  geom_point(size = 3, alpha = 0.9) +
  facet_wrap(~ polarity) +
  scale_color_hue() +
  guides(color = guide_legend(ncol = 2)) +
  labs(title = "UMAP — Consensus Peaks",
       subtitle = "Sparse-aware projection: per-mixture clusters across labs",
       color = "Mixture", shape = "Laboratory") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")
```

### 1.3.4 Consensus presence/absence heatmap

Each consensus peak is observed in only one biological mixture across
labs — the matrix is highly sparse and block-structured. This is why
Euclidean PCA on the raw consensus matrix is dominated by zeros, and
why log1p / cosine alternatives perform better.

```{r consensus-presence-heatmap}
#| fig-cap: "Presence/absence heatmap of Consensus Peaks (positive mode). Rows = samples; columns = consensus peaks. Block structure shows each mixture has its own peak subset."
#| fig-width: 12
#| fig-height: 8
library(pheatmap)

consensus_binary <- ifelse(pca_consensus_pos$data > 0, 1, 0)
keep_cols <- colSums(consensus_binary) > 0
consensus_binary <- consensus_binary[, keep_cols]

annotation_df <- data.frame(
  Mixture = pca_consensus_pos$meta$mixture,
  Lab     = pca_consensus_pos$meta$lab
)
rownames(annotation_df) <- rownames(consensus_binary)

ann_colors <- list(
  Lab = c(afekta = "#E57373", hmgu = "#64B5F6",
          icl    = "#81C784", cembio = "#FFD54F")
)

pheatmap(consensus_binary,
         cluster_rows = FALSE, cluster_cols = FALSE,
         show_rownames = FALSE, show_colnames = FALSE,
         annotation_row = annotation_df,
         annotation_colors = ann_colors,
         main = "Consensus Peak Presence/Absence\nRows = Samples | Columns = Consensus Peaks",
         border_color = NA)
```

---

# Phase 5: Peak Feature Analysis — Consensus Peaks {#phase5}

This phase extends the global analysis by incorporating peak-level metrics
computed from EICs of consensus peaks detected across all 4 laboratories.
This provides a more refined view of inter-lab reproducibility at the
individual peak level.

**Scope note.** Phase 5 reads `peak_metrics_detected_all4labs.csv` from
`2_consensus_peaks.qmd`, which matches **xcms-detected** peaks across
labs without any spike-in identity. It is independent of the SIRIUS
producer cluster-RT algorithm — re-running the producer does **not**
require re-rendering `2_consensus_peaks.qmd`. Phase 5 answers a
different question: "for the same chromatographic feature detected in
all 4 labs, how reproducible is the peak shape / intensity?" — the
flip side of the Annotated row, where Phase 5 measures *detection-level*
agreement and the Annotated row measures *identity-level* abundance.

## 5.1 Data Overview

The peak feature data comes from `2_consensus_peaks.qmd`, which:

1. Created a consensus peak table matching peaks across labs (mz tol = 10 ppm, rt tol = 10s)
2. Filtered to peaks detected in all 4 labs (1209 consensus peaks → 871 after QC)
3. Extracted EICs and computed peak metrics (FWHM, area, SNR, etc.)
4. Applied quality filters (prominence, peak width, FWHM validity)

```{r load-peak-metrics}
pk_metrics <- read.csv(file.path(base_path, "object/peak_metrics_detected_all4labs.csv"))

cat("Peak metrics dimensions:", nrow(pk_metrics), "rows x", ncol(pk_metrics), "columns\n")
cat("Unique consensus peaks:", length(unique(pk_metrics$consensus_id)), "\n")
cat("Labs represented:", paste(unique(pk_metrics$lab), collapse = ", "), "\n")
```

## 5.2 Spearman Intensity Rank Correlation

We use Spearman (rank-based) correlation to test whether peaks that are
relatively intense in one lab are also relatively intense in another.
Spearman is robust to the non-linear scale differences across labs without
requiring explicit normalization — it compares rank order, not absolute
values.

### 5.2.1 Intensity Correlation Matrix

```{r peak-metrics-correlation}
# Pivot to wide format for correlation
intensity_wide <- pk_metrics |>
  select(consensus_id, lab, areaUnderTic, maxIntensity, mixture, polarity) |>
  pivot_wider(
    id_cols = c(consensus_id, mixture, polarity),
    names_from = lab,
    values_from = c(areaUnderTic, maxIntensity)
  )

# Spearman correlation for areaUnderTic
area_cols <- paste0("areaUnderTic_", c("afekta", "hmgu", "icl", "cembio"))
area_matrix <- intensity_wide[, area_cols]
names(area_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_area <- cor(area_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("Spearman Correlation - Area Under TIC:\n")
print(round(cor_area, 3))

# Spearman correlation for maxIntensity
max_cols <- paste0("maxIntensity_", c("afekta", "hmgu", "icl", "cembio"))
max_matrix <- intensity_wide[, max_cols]
names(max_matrix) <- c("afekta", "hmgu", "icl", "cembio")

cor_max <- cor(max_matrix, method = "spearman", use = "pairwise.complete.obs")
cat("\nSpearman Correlation - Max Intensity:\n")
print(round(cor_max, 3))
```

### 5.2.2 Correlation Heatmap Visualization

```{r peak-correlation-heatmap}
#| fig-cap: "Spearman rank correlation heatmaps for recalculated peak intensities across 4 labs."
#| fig-height: 5
library(pheatmap)

pheatmap(cor_area,
         main = "Area Under TIC (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)

pheatmap(cor_max,
         main = "Max Intensity (Spearman)",
         display_numbers = TRUE,
         number_format = "%.2f",
         cluster_rows = FALSE,
         cluster_cols = FALSE)
```

Afekta and Cembio show the strongest rank correlation (ρ ≈ 0.76),
consistent with their shared Agilent Q-TOF platform. All other pairs
are moderate (ρ ≈ 0.4–0.5), indicating that relative peak intensity
rankings are partially conserved across labs but with substantial
instrument-specific variation.

## 5.3 Peak Metric Distributions Across Labs

The following metrics are computed for each EIC from the filtered consensus
peaks:

- **Peak Width**: Time span (seconds) between the detected peak boundaries.
  Reflects chromatographic resolution and column performance.
- **FWHM** (Full Width at Half Maximum): Width of the peak at half its
  maximum intensity. A narrower, more symmetric measure of peak shape than
  peak width.
- **Signal-to-Noise Ratio (SNR)**: Ratio of peak maximum intensity to the
  estimated baseline. Higher SNR indicates a cleaner, more confident
  detection.

### 5.3.1 Peak Width Distribution

```{r peak-width-distribution}
ggplot(pk_metrics, aes(x = lab, y = peakWidth, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Peak Width Distribution by Lab",
   subtitle = "Consensus peaks detected in all 4 labs",
   x = "Laboratory",
   y = "Peak Width (seconds)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

### 5.3.2 FWHM Distribution

```{r fwhm-distribution}
ggplot(pk_metrics |> filter(!is.na(xicFwhm)), aes(x = lab, y = xicFwhm, fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "FWHM Distribution by Lab",
   subtitle = "Full Width at Half Maximum of detected peaks",
   x = "Laboratory",
   y = "FWHM (seconds)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

### 5.3.3 Signal-to-Noise Ratio Distribution

```{r snr-distribution}
ggplot(pk_metrics, aes(x = lab, y = log10(signalToNoiseRatio + 1), fill = lab)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 facet_wrap(~factor(polarity, labels = c("Negative", "Positive"))) +
 labs(
   title = "Signal-to-Noise Ratio Distribution by Lab",
   subtitle = "log10(SNR + 1) scale",
   x = "Laboratory",
   y = "log10(SNR + 1)"
 ) +
 theme_bw() +
 theme(legend.position = "none")
```

## 5.4 Inter-Lab Variability for Consensus Peaks

For each consensus peak (detected in all 4 labs), we compute the coefficient
of variation (CV = standard deviation / mean) of each metric across the 4
labs. CV is inherently scale-normalized: it measures relative spread
regardless of absolute intensity differences, so no prior normalization of
intensity metrics is needed.

- **CV < 0.3 (30%)**: Good reproducibility
- **CV 0.3–0.5**: Moderate variability
- **CV > 0.5**: High variability

**IQR** (Interquartile Range): the difference between the 75th and 25th
percentiles of the CV distribution, measuring the spread of the middle 50%
of values.

### 5.4.1 CV Summary Table

```{r cv-summary-table}
cv_by_consensus <- pk_metrics |>
  group_by(consensus_id) |>
  summarise(
    n_labs = n(),
    cv_area = sd(areaUnderTic, na.rm = TRUE) / mean(areaUnderTic, na.rm = TRUE),
    cv_max = sd(maxIntensity, na.rm = TRUE) / mean(maxIntensity, na.rm = TRUE),
    cv_fwhm = sd(xicFwhm, na.rm = TRUE) / mean(xicFwhm, na.rm = TRUE),
    cv_peakwidth = sd(peakWidth, na.rm = TRUE) / mean(peakWidth, na.rm = TRUE),
    cv_snr = sd(signalToNoiseRatio, na.rm = TRUE) / mean(signalToNoiseRatio, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_labs == 4)

cv_summary <- cv_by_consensus |>
  summarise(
    across(starts_with("cv_"), list(
      median = ~median(.x, na.rm = TRUE),
      q25 = ~quantile(.x, 0.25, na.rm = TRUE),
      q75 = ~quantile(.x, 0.75, na.rm = TRUE)
    ))
  )

cv_table <- data.frame(
  Metric = c("Area Under TIC", "Max Intensity", "FWHM", "Peak Width", "SNR"),
  Median_CV = c(cv_summary$cv_area_median, cv_summary$cv_max_median,
                cv_summary$cv_fwhm_median, cv_summary$cv_peakwidth_median,
                cv_summary$cv_snr_median),
  IQR = c(cv_summary$cv_area_q75 - cv_summary$cv_area_q25,
          cv_summary$cv_max_q75 - cv_summary$cv_max_q25,
          cv_summary$cv_fwhm_q75 - cv_summary$cv_fwhm_q25,
          cv_summary$cv_peakwidth_q75 - cv_summary$cv_peakwidth_q25,
          cv_summary$cv_snr_q75 - cv_summary$cv_snr_q25)
)

knitr::kable(cv_table,
             caption = "Inter-Lab CV for Consensus Peaks (median and interquartile range)",
             digits = 3)
```

### 5.4.2 CV Distribution Visualization

```{r cv-distribution}
#| fig-cap: "CV distribution across consensus peaks for each metric. Red dashed line = 30% CV threshold."
cv_long <- cv_by_consensus |>
 pivot_longer(
   cols = starts_with("cv_"),
   names_to = "metric",
   names_prefix = "cv_",
   values_to = "cv"
 ) |>
 mutate(metric = case_when(
   metric == "area" ~ "Area Under TIC",
   metric == "max" ~ "Max Intensity",
   metric == "fwhm" ~ "FWHM",
   metric == "peakwidth" ~ "Peak Width",
   metric == "snr" ~ "SNR"
 ))

ggplot(cv_long |> filter(!is.na(cv) & cv < 2), aes(x = metric, y = cv, fill = metric)) +
 geom_violin(trim = FALSE, alpha = 0.6) +
 geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.5) +
 geom_hline(yintercept = 0.3, linetype = "dashed", color = "red") +
 annotate("text", x = 5.3, y = 0.35, label = "CV = 30%", color = "red", size = 3) +
 labs(
   title = "Inter-Lab Variability for Consensus Peaks",
   subtitle = "Lower CV = better reproducibility across labs",
   x = "Metric",
   y = "Coefficient of Variation"
 ) +
 theme_bw() +
 theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

Shape metrics (FWHM, peak width) show lower CV than intensity metrics
(area, max intensity), confirming that chromatographic behavior is more
reproducible across labs than ionization response. SNR shows the highest
variability, reflecting differences in baseline estimation across
instruments.

---

# Summary {#summary}

This file shows the **data-level progression** story:

1. **Lab variance dominates raw data**. At the Full Signal level,
   97–100 % of TIC variance is lab-attributable. PCA shape clusters
   (lab dominates), mixture colours scattered.
2. **Peak detection recovers some biology**. Detected Signal PCA shape
   clusters loosen, colour grouping emerges. xcms peak picking strips
   most of the lab-specific background.
3. **Annotation concentrates the biology further**. Annotated PCA
   (sample × compound maxIntensity) tightens mixture clusters; lab
   spread within each mixture shrinks. SIRIUS-keyed features carry
   identity-aware abundance.
4. **Consensus peaks deliver the cleanest picture**. The cross-lab
   matched peak set (~1.8 k features) gives tight per-mixture clusters
   in PCA + UMAP — same-mixture points across all four labs sit
   together. Per-peak inter-lab CV distribution shows the broader
   feature space behaves like the spike-in panel.

For the **quantitative reproducibility numbers** (ICC ladder, pairwise
ICC, same-platform Agilent ceiling, 3-lab subset), see
`7_icc_reproducibility.qmd`. For the **normalization strategy comparison**,
see `5_normalization.qmd` + `7_icc_reproducibility.qmd`. For the
**RT alignment**, see `4_rt_alignment.qmd`. For the **per-compound
diagnostics**, see `9_per_compound_diagnostics.qmd`.

---

# Future directions

- Expand the four-level progression to the other HUMAN matrices
  (FE, Others) beyond HE endosome.
- Add a "post-normalization" data level (Annotated normalized) to the
  PCA progression once the §8 normalization output is the default
  consumer input.


