---
title: "Per-compound diagnostics: identity, retention, abundance"
format:
    html:
        toc: true
        toc-location: right
        code-fold: true
---

**Per-compound deep dive** on the annotated spike-in panel. Three axes:
identity (MS2), retention (RT), abundance (MS1). All analyses are at
the **compound level** — one number per `(mixture, compound, lab,
polarity)` — to surface where individual compounds disagree across
labs.

For the **cross-lab reproducibility numbers** (ICC, pairwise ICC,
3-lab subset) see `7_icc_reproducibility.qmd`. For the **data-level
progression** (Full → Detected → Annotated → Consensus) see
`8_data_level_progression.qmd`. For the **RT alignment splines** see
`4_rt_alignment.qmd`.

Inputs (canonical CSVs from the SIRIUS producer
`3_annotation_manual/HE/3_Sirius_curation/sirius_annotation_all_ms2.qmd`):

- `annotation_identity.csv` — MS2 identity per `(mixture, lab, compound, polarity)`
- `annotation_chrom_metrics.csv` — MS1 peak metrics per `(mixture, lab, polarity, compound, adduct, source)`

See [`CANONICAL_SCHEMAS.md`](CANONICAL_SCHEMAS.md) for column contracts.

Identity-driven analyses keep rows where `confidence_tier ∈ tier_keep`.
The default `tier_keep` includes the legacy `"unfiltered"` placeholder
plus tiers A and B; tighten to `c("A", "B")` for a confident-only
sensitivity re-run.

## Setup

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)

labs <- c("afekta", "cembio", "hmgu", "icl")
producer_dir <- here("3_annotation_manual", "HE", "3_Sirius_curation",
                     "results_all_ms2")

tier_keep <- c("unfiltered", "A", "B")

## mixture forced to character — read.csv would otherwise coerce quoted
## "2.10" → 2.1 (numeric), silently colliding with mixture "2.1".
identity_df <- read.csv(file.path(producer_dir, "annotation_identity.csv"),
                        stringsAsFactors = FALSE,
                        colClasses = c(mixture = "character")) |>
  filter(confidence_tier %in% tier_keep)

chrom_df <- read.csv(file.path(producer_dir, "annotation_chrom_metrics.csv"),
                     stringsAsFactors = FALSE,
                     colClasses = c(mixture = "character"))

cat("identity_df rows (tier-kept):", nrow(identity_df),
    " | chrom_df rows:", nrow(chrom_df), "\n")
```

`chrom_one_per_compound` collapses `chrom_df` to one row per
`(mixture, lab, compound)` by picking the polarity / adduct / source
giving the highest `maxIntensity`. Used by per-compound abundance and
RT analyses.

```{r chrom-collapse}
chrom_one_per_compound <- chrom_df |>
  group_by(mixture, lab, compound) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup()
```

`expected_grid` is the full `(mixture, compound) × lab` grid — used as
a `right_join` target so plots show absences as white tiles.

```{r expected-grid}
expected_grid <- identity_df |>
  distinct(mixture, compound) |>
  tidyr::expand_grid(lab = labs)
```

# Identity (MS2)

## Per-compound coverage across labs

For each `(mixture, compound, lab)`: did SIRIUS annotate the expected
compound (via 2D-InChIKey match) at any rank?

```{r compound-presence-heatmap}
#| fig-height: 60
#| fig-cap: "MS2 identification of expected compounds per lab. Blue = annotated; grey = not annotated. Mixture stripes on the left."
compound_presence <- expected_grid |>
  left_join(identity_df |>
              distinct(mixture, lab, compound) |>
              mutate(found = TRUE),
            by = c("mixture", "lab", "compound")) |>
  mutate(found = coalesce(found, FALSE))

ggplot(compound_presence, aes(x = lab, y = compound, fill = found)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(values = c(`TRUE` = "#1f77b4", `FALSE` = "#eeeeee"),
                    labels = c(`TRUE` = "Annotated", `FALSE` = "Not annotated")) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "MS2 identification per (mixture, compound, lab)") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        axis.text.x = element_text(face = "bold", size = 11),
        strip.text.y = element_text(angle = 0, face = "bold"),
        legend.position = "bottom")
```

```{r labs-n-found}
#| fig-height: 60
#| fig-cap: "Number of labs that annotated each expected compound."
compound_presence |>
  group_by(mixture, compound) |>
  summarise(n_labs = sum(found), .groups = "drop") |>
  ggplot(aes(x = n_labs, y = compound, fill = factor(n_labs))) +
  geom_col() +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  scale_x_continuous(breaks = 0:length(labs)) +
  labs(x = "Number of labs", y = NULL, fill = "# labs",
       title = "How many labs annotated each compound?") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        strip.text.y = element_text(angle = 0, face = "bold"))
```

## Best structure rank per (mixture, compound, lab)

Lower is better; 1 = top hit. Polarity collapsed to the smaller of the
two ranks when both are present.

```{r rank-heatmap}
#| fig-height: 60
#| fig-cap: "SIRIUS best structure rank per (mixture, compound, lab). White = not annotated."
best_rank_per_lab <- identity_df |>
  group_by(mixture, lab, compound) |>
  summarise(best_rank = suppressWarnings(min(best_structure_rank, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(best_rank = ifelse(is.infinite(best_rank), NA_real_, best_rank))

best_rank_per_lab |>
  right_join(expected_grid, by = c("mixture", "lab", "compound")) |>
  ggplot(aes(x = lab, y = compound, fill = best_rank)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(option = "C", direction = -1, na.value = "#eeeeee",
                       name = "Best rank\n(1 = top)") +
  labs(x = NULL, y = NULL, title = "Best SIRIUS structure rank") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        axis.text.x = element_text(face = "bold", size = 11),
        strip.text.y = element_text(angle = 0, face = "bold"))
```

```{r rank-summary}
best_rank_per_lab |>
  filter(!is.na(best_rank)) |>
  group_by(lab) |>
  summarise(n           = n(),
            median_rank = median(best_rank),
            pct_rank1   = round(100 * mean(best_rank == 1), 1),
            pct_top3    = round(100 * mean(best_rank <= 3), 1),
            pct_top10   = round(100 * mean(best_rank <= 10), 1),
            .groups = "drop") |>
  knitr::kable(caption = "Structure rank summary per lab")
```

```{r rank-binned-histogram}
#| fig-height: 4
#| fig-cap: "Structure rank distribution per lab (binned)."
best_rank_per_lab |>
  filter(!is.na(best_rank)) |>
  mutate(bin = cut(best_rank, breaks = c(0, 1, 3, 10, 50, Inf),
                   labels = c("1", "2-3", "4-10", "11-50", ">50"))) |>
  ggplot(aes(x = bin, fill = lab)) +
  geom_bar() +
  facet_wrap(~ lab, scales = "free_y") +
  labs(x = "Structure rank bin", y = "Compounds",
       title = "Per-lab structure rank distribution") +
  theme_bw() +
  theme(legend.position = "none")
```

## Adducts observed

Adducts that supported each lab's annotation, broken down per mixture
(stacked) and per (lab, compound) (boxplot of distinct adducts).

```{r adduct-usage}
#| fig-height: 8
#| fig-cap: "Adducts observed for expected-compound hits per lab × mixture."
identity_df |>
  tidyr::separate_rows(adducts, sep = ";") |>
  filter(!is.na(adducts), nzchar(adducts)) |>
  count(mixture, lab, adducts) |>
  ggplot(aes(x = lab, y = n, fill = adducts)) +
  geom_col(position = "stack") +
  facet_wrap(~ mixture) +
  labs(x = NULL, y = "Count", fill = "Adduct",
       title = "Adducts for expected-compound hits") +
  theme_bw() +
  theme(legend.key.size = unit(0.4, "cm"),
        legend.text     = element_text(size = 7))
```

```{r adduct-breadth}
#| fig-height: 4
#| fig-cap: "Per-compound adduct breadth, by lab."
identity_df |>
  tidyr::separate_rows(adducts, sep = ";") |>
  filter(!is.na(adducts), nzchar(adducts)) |>
  distinct(lab, mixture, compound, adducts) |>
  count(lab, mixture, compound, name = "n_adducts") |>
  ggplot(aes(x = lab, y = n_adducts)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  labs(x = NULL, y = "Adducts per compound",
       title = "Per-compound adduct breadth, by lab") +
  theme_bw()
```

## m/z deviation across labs

For each `(mixture, compound, polarity)` we compute the per-lab ppm
deviation of `mz_observed` from the cross-lab mean, restricted to
compound–polarity pairs seen in ≥ 2 labs. Per (mixture, compound, lab)
we then aggregate across polarity by median for the heatmap. Same
calculation as the producer's original §6 — coarser granularity (one
ionMass per polarity instead of per adduct) because identity collapses
candidates to one row per polarity at producer time.

```{r mz-dev-build}
mz_dev_df <- identity_df |>
  filter(!is.na(mz_observed)) |>
  group_by(mixture, compound, polarity) |>
  mutate(n_labs    = n_distinct(lab),
         mean_mass = mean(mz_observed, na.rm = TRUE),
         ppm_dev   = (mz_observed - mean_mass) / mean_mass * 1e6) |>
  ungroup() |>
  filter(n_labs >= 2)

cat("(mixture, compound, polarity) pairs with >= 2 labs:",
    n_distinct(paste(mz_dev_df$mixture, mz_dev_df$compound,
                     mz_dev_df$polarity)), "\n")
```

```{r mz-dev-heatmap}
#| fig-height: 60
#| fig-cap: "Median Δ m/z (ppm) from cross-lab mean per (mixture, compound, lab). White = compound not annotated by this lab."
mz_dev_df |>
  group_by(mixture, compound, lab) |>
  summarise(ppm_dev = median(ppm_dev, na.rm = TRUE), .groups = "drop") |>
  right_join(expected_grid, by = c("mixture", "lab", "compound")) |>
  ggplot(aes(x = lab, y = compound, fill = ppm_dev)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_gradient2(low = "#3b4cc0", mid = "white", high = "#b40426",
                       midpoint = 0, na.value = "#eeeeee",
                       name = "ppm dev\nvs mean") +
  labs(x = NULL, y = NULL,
       title = "Inter-lab m/z shift per (mixture, compound, lab)") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        axis.text.x = element_text(face = "bold", size = 11),
        strip.text.y = element_text(angle = 0, face = "bold"))
```

# Retention (RT)

## Per-lab apex RT

`apex_rt` per `(mixture, compound, lab)`, taking the dominant-signal
adduct/source via `chrom_one_per_compound`.

```{r apex-rt-heatmap}
#| fig-height: 60
#| fig-cap: "MS1 apex RT (s) per (mixture, compound, lab). White = no peak detected."
chrom_one_per_compound |>
  right_join(expected_grid, by = c("mixture", "lab", "compound")) |>
  ggplot(aes(x = lab, y = compound, fill = apex_rt)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(option = "C", na.value = "#eeeeee", name = "Apex RT (s)") +
  labs(x = NULL, y = NULL,
       title = "MS1 apex RT per (mixture, compound, lab)") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        axis.text.x = element_text(face = "bold", size = 11),
        strip.text.y = element_text(angle = 0, face = "bold"))
```

## Cross-lab RT spread

Per `(mixture, compound, polarity)`, summarise apex RT across labs
that detected the peak. **Polarity must be in the grouping** — a
compound can have very different RT in pos vs neg because the
extraction is on different m/z windows; `chrom_one_per_compound`'s
collapse-across-polarity rule would otherwise mix them and inflate
the apparent "cross-lab RT range" (e.g. 2-oxobutanoic acid in mix
2.2 has neg-only range 7.7 s and pos-only range 14.9 s, but mixing
polarities gives the misleading 228 s reported in earlier renders).

```{r rt-spread-build}
# Best peak per (mixture, lab, compound, polarity) — polarity-aware,
# unlike chrom_one_per_compound. Restrict to detected rows.
chrom_dom_pol <- chrom_df |>
  filter(detection == "detected" | (!is.na(apex_rt) & !is.na(maxIntensity))) |>
  filter(!is.na(apex_rt), !is.na(maxIntensity), maxIntensity > 0) |>
  group_by(mixture, lab, compound, polarity) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup()

rt_per_compound <- chrom_dom_pol |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs   = n(),
            rt_mean  = mean(apex_rt),
            rt_sd    = sd(apex_rt),
            rt_range = max(apex_rt) - min(apex_rt),
            .groups  = "drop") |>
  filter(n_labs >= 2L)

rt_per_compound |>
  arrange(desc(rt_range)) |>
  head(20) |>
  mutate(across(c(rt_mean, rt_sd, rt_range), \(x) round(x, 1))) |>
  knitr::kable(caption = "Top 20 (mixture, compound, polarity) by cross-lab apex RT range")
```

```{r rt-spread-plot}
#| fig-height: 4
#| fig-cap: "Apex RT SD per compound, faceted by number of detecting labs."
ggplot(rt_per_compound, aes(x = factor(n_labs), y = rt_sd)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.15, alpha = 0.5) +
  labs(x = "Number of labs detecting", y = "Apex RT sd (s)",
       title = "Between-lab apex RT variability per compound") +
  theme_bw()
```

```{r rt-range-scatter}
#| fig-height: 25
#| fig-cap: "Cross-lab apex RT range per (mixture, compound) for compounds detected in ≥ 2 labs."
rt_per_compound |>
  arrange(desc(rt_range)) |>
  mutate(label = reorder(paste0(mixture, " - ", compound), rt_range)) |>
  ggplot(aes(x = rt_range, y = label, colour = factor(n_labs))) +
  geom_point(size = 3) +
  scale_colour_viridis_d(option = "C", end = 0.85) +
  labs(x = "Apex RT range across labs (s)", y = NULL, colour = "# labs",
       title = "Cross-lab RT spread per compound (≥ 2 labs)") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 6))
```

## Cross-polarity RT agreement (MS2-hit sanity)

The MS2 `apex_rt` recorded for the same compound in pos vs neg must agree
— adducts of the same molecule elute at the same RT. A large
`|rt_pos − rt_neg|` means **DDA triggered MS2 on different
chromatographic peaks in the two polarities** and SIRIUS still matched
both spectra to the compound (typically because the fragments fit the
same chemical formula / structure rule). The "wrong" polarity's
`own` row then carries the wrong `apex_rt`, which propagates into the
cross-lab consensus used by labs that did not annotate this compound.

This is **a producer-side issue at the MS2-hit step** — it does not
arise from the SNR picker or chromExtract. The diagnostic below
classifies each compound (detected in both polarities by ≥ 2 labs each)
into four states:

- **agree (≤ 5 s)** — pos and neg MS2 triggered on the same peak. ✓
- **moderate disagreement (5–30 s)** — same peak with chromatographic
  drift, or a borderline adjacent peak.
- **reproducible-wrong (> 30 s, labs agree *within* each polarity)** —
  multiple labs *independently* triggered MS2 on the same wrong pos peak
  and the same neg peak. Almost certainly a real chromatographic
  isomer / degradation product / contaminant sharing the SIRIUS
  fragmentation pattern.
- **non-reproducible-wrong (> 30 s, labs differ within polarity)** —
  each lab grabbed a different random peak. Random extraction noise.

```{r cross-polarity-build}
# Best detected MS2-hit peak per (mixture, compound, polarity, lab).
dom_per_pol <- chrom_df |>
  filter(detection == "detected",
         !is.na(apex_rt), !is.na(maxIntensity), maxIntensity > 0) |>
  group_by(mixture, compound, polarity, lab) |>
  slice_max(maxIntensity, n = 1, with_ties = FALSE) |>
  ungroup()

# Per (mixture, compound, polarity), cross-lab consensus apex_rt + spread.
per_pol_consensus <- dom_per_pol |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs   = n(),
            rt_med   = median(apex_rt),
            rt_range = max(apex_rt) - min(apex_rt),
            .groups  = "drop")

# Best COSMIC per (mixture, compound, polarity) — exact score.
cosmic_per_pol <- identity_df |>
  group_by(mixture, compound, polarity) |>
  summarise(cosmic = suppressWarnings(max(confidenceExactMatch, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(cosmic = ifelse(is.infinite(cosmic), NA_real_, cosmic))

# Both-polarity wide table.
both_pol <- per_pol_consensus |>
  tidyr::pivot_wider(id_cols = c(mixture, compound),
                      names_from = polarity,
                      values_from = c(n_labs, rt_med, rt_range)) |>
  filter(!is.na(rt_med_pos), !is.na(rt_med_neg),
         n_labs_pos >= 2, n_labs_neg >= 2) |>
  mutate(abs_delta = abs(rt_med_pos - rt_med_neg)) |>
  left_join(cosmic_per_pol |> filter(polarity == "pos") |>
              select(mixture, compound, cosmic_pos = cosmic),
            by = c("mixture", "compound")) |>
  left_join(cosmic_per_pol |> filter(polarity == "neg") |>
              select(mixture, compound, cosmic_neg = cosmic),
            by = c("mixture", "compound")) |>
  mutate(class = case_when(
    abs_delta <= 5                                          ~ "agree (≤5s)",
    abs_delta > 5  & abs_delta <= 30                        ~ "moderate disagreement (5–30s)",
    abs_delta > 30 & rt_range_pos <= 20 & rt_range_neg <= 20 ~ "reproducible-wrong (>30s, labs agree per polarity)",
    abs_delta > 30                                          ~ "non-reproducible-wrong (labs differ within polarity)",
    TRUE                                                    ~ "other"
  ),
  class = factor(class, levels = c(
    "agree (≤5s)",
    "moderate disagreement (5–30s)",
    "reproducible-wrong (>30s, labs agree per polarity)",
    "non-reproducible-wrong (labs differ within polarity)")))

cat("Compounds detected in both polarities (≥2 labs each):", nrow(both_pol), "\n")
both_pol |> count(class) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  knitr::kable(caption = "Cross-polarity agreement classes for MS2 hits.")
```

```{r cross-polarity-scatter}
#| fig-cap: "Cross-polarity RT agreement per (mixture, compound). Dashed line = pos = neg. Red = reproducible-wrong (labs agree on a wrong RT within each polarity). Grey = noise."
#| fig-height: 6
ggplot(both_pol, aes(x = rt_med_neg, y = rt_med_pos, colour = class,
                      size = pmin(n_labs_pos, n_labs_neg))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.8) +
  scale_colour_manual(values = c(
    "agree (≤5s)"                                          = "#2CA02C",
    "moderate disagreement (5–30s)"                        = "#FFB347",
    "reproducible-wrong (>30s, labs agree per polarity)"   = "#D62728",
    "non-reproducible-wrong (labs differ within polarity)" = "#888888"),
    drop = FALSE) +
  scale_size_continuous(range = c(1.5, 4), breaks = c(2, 3, 4),
                          name = "min(n_labs)") +
  labs(x = "Cross-lab median apex_rt — neg (s)",
       y = "Cross-lab median apex_rt — pos (s)",
       colour = NULL,
       title = "Cross-polarity RT agreement per compound") +
  theme_bw() + theme(legend.position = "right")
```

```{r cross-polarity-table}
both_pol |>
  filter(class == "reproducible-wrong (>30s, labs agree per polarity)") |>
  arrange(desc(abs_delta)) |>
  mutate(across(c(rt_med_pos, rt_med_neg, rt_range_pos, rt_range_neg,
                  abs_delta), \(x) round(x, 1)),
         across(c(cosmic_pos, cosmic_neg), \(x) round(x, 2))) |>
  select(mixture, compound,
         n_labs_pos, rt_med_pos, rt_range_pos, cosmic_pos,
         n_labs_neg, rt_med_neg, rt_range_neg, cosmic_neg,
         abs_delta) |>
  knitr::kable(caption = "Reproducible-wrong cross-polarity disagreements. High COSMIC in both polarities = real chromatographic isomer with shared MS2 fragments (or co-eluting contaminant), not random noise.")
```

## LOO-RT-shift annotation confidence

For each `(mixture, compound, lab)` the leave-one-out shift:

```
loo_shift(compound, lab) = apex_rt(compound, lab)
                         - median(apex_rt of other labs)
```

If large `|loo_shift|` coincides with bad SIRIUS rank → likely
wrong-peak annotation. If large shift coincides with rank 1 → labs
agree on the compound on chromatographically shifted peaks.

```{r loo-shift-build}
# LOO shift must be polarity-aware for the same reason as rt-spread
# above — comparing a lab's pos apex_rt to others' neg apex_rt is
# meaningless.
rt_lab <- chrom_dom_pol |>
  select(mixture, lab, compound, polarity, apex_rt)

rt_shift <- rt_lab |>
  group_by(mixture, compound, polarity) |>
  mutate(n_labs    = n(),
         loo_shift = apex_rt - vapply(seq_len(n()),
                                      \(i) median(apex_rt[-i], na.rm = TRUE),
                                      numeric(1))) |>
  ungroup() |>
  mutate(abs_loo_shift = abs(loo_shift)) |>
  filter(n_labs >= 2)

rank_rt <- inner_join(rt_shift,
                      best_rank_per_lab |> filter(!is.na(best_rank)),
                      by = c("mixture", "lab", "compound"))
```

```{r loo-shift-binned}
#| fig-cap: "(mixture, lab, compound) cases per |LOO RT shift| bin × number of labs detecting."
rank_rt |>
  mutate(shift_bin = cut(abs_loo_shift,
                         breaks = c(-Inf, 2, 5, 10, Inf),
                         labels = c("<=2s", "2-5s", "5-10s", ">10s"))) |>
  count(n_labs, shift_bin) |>
  ggplot(aes(x = shift_bin, y = n, fill = factor(n_labs))) +
  geom_col(position = "dodge") +
  geom_text(aes(label = n), position = position_dodge(width = 0.9),
            vjust = -0.3, size = 3) +
  labs(x = "|LOO RT shift|", y = "(compound, lab) cases",
       fill = "n_labs detecting",
       title = "LOO RT shifts broken down by number of detecting labs") +
  theme_bw()
```

```{r shift-vs-rank}
#| fig-cap: "Rank tier composition by LOO RT shift bin (proportions within each bin)."
rank_rt |>
  mutate(shift_bin = cut(abs_loo_shift,
                         breaks = c(-Inf, 2, 5, 10, Inf),
                         labels = c("<=2s", "2-5s", "5-10s", ">10s")),
         rank_tier = factor(case_when(
                              best_rank == 1   ~ "1 (top)",
                              best_rank <= 10  ~ "2-10",
                              TRUE             ~ ">10"),
                            levels = c("1 (top)", "2-10", ">10"))) |>
  count(shift_bin, rank_tier) |>
  group_by(shift_bin) |>
  mutate(pct = 100 * n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = shift_bin, y = pct, fill = rank_tier)) +
  geom_col() +
  geom_text(aes(label = ifelse(pct >= 5, sprintf("%.0f%%", pct), "")),
            position = position_stack(vjust = 0.5), size = 3) +
  labs(x = "|LOO RT shift|", y = "% of (compound, lab) cases",
       fill = "Structure rank",
       title = "Annotation quality vs RT shift magnitude") +
  theme_bw()
```

```{r shift-vs-rank-scatter}
#| fig-height: 5
#| fig-cap: "Rank vs |LOO RT shift|. Point size = number of labs detecting the compound."
ggplot(rank_rt, aes(x = abs_loo_shift, y = best_rank,
                    colour = lab, size = factor(n_labs))) +
  geom_jitter(width = 0, height = 0.15, alpha = 0.7) +
  scale_size_manual(values = c("2" = 1.5, "3" = 2.5, "4" = 3.5)) +
  labs(x = "|LOO RT shift| (s)", y = "Structure rank",
       colour = "Lab", size = "n_labs detecting",
       title = "Structure rank vs LOO RT shift") +
  theme_bw()
```

```{r worst-shift-context}
#| fig-height: 5
#| fig-cap: "Top 10 LOO-shift cases. Triangle = focal lab; round = other labs."
top_shift <- rank_rt |>
  arrange(desc(abs_loo_shift)) |>
  head(10) |>
  mutate(label = sprintf("%s | focal=%s | n=%d | rank=%d | shift=%.1fs",
                         compound, lab, n_labs, best_rank, loo_shift))

panel <- top_shift |>
  select(mixture, compound, lab_focal = lab, label) |>
  inner_join(rt_lab, by = c("mixture", "compound"),
             relationship = "many-to-many") |>
  mutate(is_focal = lab == lab_focal,
         label    = factor(label, levels = rev(top_shift$label)))

ggplot(panel, aes(x = apex_rt, y = label,
                  colour = lab, shape = is_focal, size = is_focal)) +
  geom_point() +
  scale_shape_manual(values = c(`TRUE` = 17, `FALSE` = 16), guide = "none") +
  scale_size_manual(values = c(`TRUE` = 3.5, `FALSE` = 2.5), guide = "none") +
  labs(x = "apex RT (s)", y = NULL, colour = "lab",
       title = "Worst LOO-RT-shift cases — per-lab apex RT context") +
  theme_bw()
```

Read this plot as: triangle far from the round-point cluster + rank 1
→ labs are calling the same compound on chromatographically shifted
peaks (annotation probably OK, RT is the issue). Triangle far + bad
rank → suspicious annotation (likely wrong peak picked).

### Rank per lab on the largest-RT-shift compounds

For compounds in the top tertile of `rt_range` within their mixture,
show the SIRIUS structure rank per lab. Same diagnostic as the
producer's original §7: low ranks across labs despite large RT spread
→ labs converge on the right compound but their chromatography differs;
mixed ranks → at least one lab probably picked the wrong peak.

```{r rank-vs-large-rt-shift}
#| fig-height: 12
#| fig-cap: "Structure rank per lab for compounds in the top tertile of cross-lab RT range (within each mixture)."
large_shift <- rt_per_compound |>
  group_by(mixture) |>
  filter(rt_range >= quantile(rt_range, 2/3, na.rm = TRUE)) |>
  ungroup() |>
  select(mixture, compound)

best_rank_per_lab |>
  inner_join(large_shift, by = c("mixture", "compound")) |>
  mutate(facet_lab = paste0(mixture, " - ", compound)) |>
  ggplot(aes(x = lab, y = best_rank, colour = lab)) +
  geom_point(size = 2.5, na.rm = TRUE) +
  facet_wrap(~ facet_lab, ncol = 4) +
  scale_y_reverse() +
  labs(x = NULL, y = "Best structure rank (1 = top)", colour = NULL,
       title = "Annotation rank for compounds with large RT shifts") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x     = element_text(angle = 45, hjust = 1),
        strip.text      = element_text(size = 6))
```

# Abundance (MS1)

## Lab-median-centred log10 abundance

`log_max_int_lab_norm = log10(maxIntensity) - median(log10(maxIntensity))`
within each `(lab, mixture)`. Removes cross-lab sensitivity offsets so
colour shows which compounds respond unusually well or poorly **within
each lab**, free of platform-scale differences.

`maxIntensity` is the primary abundance metric (scan-rate-robust);
`areaUnderTic` is sensitive to scan count and boundary choice across
the ~15× scan-rate gap between cembio (1.5 s) and icl (0.1 s).

```{r abundance-build}
cross_lab_df <- chrom_one_per_compound |>
  filter(!is.na(maxIntensity), maxIntensity > 0) |>
  group_by(mixture, compound) |>
  filter(n_distinct(lab) >= 2L) |>
  ungroup() |>
  mutate(log_max_int = log10(maxIntensity)) |>
  group_by(lab, mixture) |>
  mutate(log_max_int_lab_norm = log_max_int - median(log_max_int, na.rm = TRUE)) |>
  ungroup()

cat("(mixture, compound) pairs with >= 2 labs:",
    n_distinct(paste(cross_lab_df$mixture, cross_lab_df$compound)),
    " | rows:", nrow(cross_lab_df), "\n")
```

```{r abundance-heatmap}
#| fig-height: 14
#| fig-cap: "log10 maxIntensity centred on each (lab, mixture)'s median compound response."
cross_lab_df |>
  distinct(mixture, compound) |>
  tidyr::expand_grid(lab = labs) |>
  left_join(cross_lab_df |>
              select(mixture, lab, compound, log_max_int_lab_norm),
            by = c("mixture", "lab", "compound")) |>
  ggplot(aes(x = lab, y = compound, fill = log_max_int_lab_norm)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_gradient2(low = "#3b4cc0", mid = "white", high = "#b40426",
                       midpoint = 0, na.value = "#eeeeee",
                       name = "log10 FC\nvs lab median") +
  labs(x = NULL, y = NULL,
       title = "Lab-median-centred log10 maxIntensity") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        axis.text.x = element_text(face = "bold", size = 11),
        strip.text.y = element_text(angle = 0, face = "bold"))
```

## Peak shape (FWHM)

```{r fwhm-heatmap}
#| fig-height: 60
#| fig-cap: "Raw FWHM (s) per (mixture, compound, lab). White = no peak."
chrom_one_per_compound |>
  right_join(expected_grid, by = c("mixture", "lab", "compound")) |>
  ggplot(aes(x = lab, y = compound, fill = xicFwhm)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(option = "C", na.value = "#eeeeee", name = "FWHM (s)") +
  labs(x = NULL, y = NULL, title = "FWHM (s) per (mixture, compound, lab)") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        axis.text.x = element_text(face = "bold", size = 11),
        strip.text.y = element_text(angle = 0, face = "bold"))
```

### Peak-shape reproducibility vs peak-intensity reproducibility

**Hypothesis**: chromatography is harmonized across labs (same column,
gradient, mobile phases, flow rate), so peak shape (FWHM) should be
*more* reproducible than peak intensity (maxIntensity), which is set
by MS sensitivity and ionization efficiency — neither of which is
harmonized. A direct comparison on the same compound set.

Three small tests:

1. **Cross-lab CV distribution** for FWHM vs maxIntensity, per
   `(mixture, compound)` detected in ≥ 3 labs. Lower CV → more
   reproducible.
2. **Cross-lab ICC** for each metric using the same `lmer(~ 1 |
   compound)` formulation that drives §9. Higher ICC → more
   between-compound signal vs lab noise.
3. **Random-shuffle control**: permute the `(lab, compound)` pairing
   inside each polarity, recompute CV. Real CV ≪ shuffled CV means
   the metric carries compound-specific information (i.e. the metric
   recognises *this* compound vs a random other). If real ≈ shuffled,
   the metric isn't compound-discriminative.

```{r shape-vs-intensity-cv-build}
shape_int_cv <- chrom_one_per_compound |>
  filter(!is.na(xicFwhm),     xicFwhm     > 0,
         !is.na(maxIntensity), maxIntensity > 0) |>
  group_by(mixture, compound, polarity) |>
  summarise(n_labs    = n(),
            cv_fwhm   = sd(xicFwhm)      / mean(xicFwhm),
            cv_maxInt = sd(maxIntensity) / mean(maxIntensity),
            .groups   = "drop") |>
  filter(n_labs >= 3L)

cat("Compounds in CV analysis (≥3 labs):\n",
    " pos: ", sum(shape_int_cv$polarity == "pos"), "\n",
    " neg: ", sum(shape_int_cv$polarity == "neg"), "\n", sep = "")
```

```{r shape-vs-intensity-cv-plot}
#| fig-cap: "Cross-lab CV per (mixture, compound) for FWHM vs maxIntensity. Lower = more reproducible. Compounds detected in ≥ 3 labs."
#| fig-height: 4
cv_long <- shape_int_cv |>
  tidyr::pivot_longer(c(cv_fwhm, cv_maxInt),
                       names_to = "metric", values_to = "cv") |>
  mutate(metric = ifelse(metric == "cv_fwhm",
                          "FWHM (peak shape)",
                          "maxIntensity (peak intensity)"))

ggplot(cv_long, aes(x = metric, y = cv, fill = metric)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.12, outlier.size = 0.5, alpha = 0.8) +
  scale_y_log10() +
  scale_fill_manual(values = c("FWHM (peak shape)" = "#1F77B4",
                                "maxIntensity (peak intensity)" = "#FF7F0E")) +
  facet_wrap(~ polarity) +
  labs(x = NULL, y = "Cross-lab CV (log10)",
       title = "Peak shape vs peak intensity reproducibility",
       subtitle = "Lower CV = more reproducible across labs.") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r shape-vs-intensity-icc}
suppressPackageStartupMessages(library(lme4))
compute_icc_local <- function(d, val_col) {
  d <- d[!is.na(d[[val_col]]) & d[[val_col]] > 0, , drop = FALSE]
  if (nrow(d) < 6L || dplyr::n_distinct(d$compound) < 2L) return(NA_real_)
  d$log_val <- log10(d[[val_col]] + 1)
  fit <- tryCatch(lmer(log_val ~ 1 + (1 | compound), data = d),
                  error = function(e) NULL)
  if (is.null(fit)) return(NA_real_)
  vc <- as.data.frame(VarCorr(fit))
  vb <- vc$vcov[vc$grp == "compound"]
  vw <- vc$vcov[vc$grp == "Residual"]
  vb / (vb + vw)
}

icc_input <- chrom_one_per_compound |>
  dplyr::semi_join(
    shape_int_cv |> dplyr::select(mixture, compound, polarity),
    by = c("mixture", "compound", "polarity"))

icc_tbl <- bind_rows(lapply(c("pos","neg"), function(pol) {
  d <- icc_input |> filter(polarity == pol)
  tibble(polarity = pol,
         metric   = c("FWHM (shape)", "maxIntensity"),
         ICC      = c(compute_icc_local(d, "xicFwhm"),
                      compute_icc_local(d, "maxIntensity")))
})) |>
  mutate(ICC = round(ICC, 3))

knitr::kable(icc_tbl,
             caption = "Cross-lab ICC: peak shape (FWHM) vs peak intensity (maxIntensity), same compound set. Higher = more between-compound signal vs lab noise.")
```

```{r shape-shuffled-control}
#| fig-cap: "Real (same-compound) CV vs random shuffled-pairing CV. If real ≪ shuffled, the metric is compound-discriminative."
#| fig-height: 4
set.seed(42L)

# Shuffle compound labels within each lab × polarity, then recompute CV
# on the resulting (pseudo-)compounds with ≥3 labs.
shuffled_cv <- bind_rows(lapply(c("pos","neg"), function(pol) {
  d <- chrom_one_per_compound |>
    filter(polarity == pol,
           !is.na(xicFwhm),      xicFwhm      > 0,
           !is.na(maxIntensity), maxIntensity > 0)
  if (!nrow(d)) return(NULL)
  d_sh <- d |>
    group_by(lab) |>
    mutate(compound = sample(compound)) |>
    ungroup()
  d_sh |>
    group_by(mixture, compound, polarity) |>
    summarise(n_labs    = n(),
              cv_fwhm   = sd(xicFwhm)      / mean(xicFwhm),
              cv_maxInt = sd(maxIntensity) / mean(maxIntensity),
              .groups   = "drop") |>
    filter(n_labs >= 3L)
}))

cv_real_sh <- bind_rows(
  shape_int_cv |> select(polarity, cv_fwhm, cv_maxInt) |>
    mutate(pairing = "real (same compound)"),
  shuffled_cv  |> select(polarity, cv_fwhm, cv_maxInt) |>
    mutate(pairing = "shuffled (random compound)")
) |>
  tidyr::pivot_longer(c(cv_fwhm, cv_maxInt),
                       names_to = "metric", values_to = "cv") |>
  mutate(metric = ifelse(metric == "cv_fwhm", "FWHM", "maxIntensity"))

ggplot(cv_real_sh, aes(x = metric, y = cv, fill = pairing)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.6,
              position = position_dodge(0.8)) +
  geom_boxplot(width = 0.10, outlier.size = 0.3, alpha = 0.8,
               position = position_dodge(0.8)) +
  scale_y_log10() +
  scale_fill_manual(values = c("real (same compound)"        = "#2CA02C",
                                "shuffled (random compound)" = "#888888")) +
  facet_wrap(~ polarity) +
  labs(x = NULL, y = "Cross-lab CV (log10)",
       fill = NULL,
       title = "Compound-discrimination control",
       subtitle = "Real ≪ shuffled means the metric distinguishes between compounds.") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 0))
```

```{r shape-shuffled-summary}
medians <- cv_real_sh |>
  group_by(polarity, metric, pairing) |>
  summarise(median_cv = round(median(cv, na.rm = TRUE), 3),
            .groups = "drop") |>
  tidyr::pivot_wider(names_from = pairing, values_from = median_cv)

# Defensive: if the shuffle didn't produce any (mixture, random_compound,
# polarity) tuple with >=3 labs (can happen on the smaller post-cluster
# datasets), the "shuffled (random compound)" column simply doesn't
# exist. Add NAs so the table still renders and the ratio is NA rather
# than an error.
real_col <- "real (same compound)"
shuf_col <- "shuffled (random compound)"
if (!(real_col %in% names(medians))) medians[[real_col]] <- NA_real_
if (!(shuf_col %in% names(medians))) medians[[shuf_col]] <- NA_real_

medians$real_over_shuffled <- round(medians[[real_col]] /
                                     medians[[shuf_col]], 2)

knitr::kable(medians,
             caption = "Median CV: real same-compound pairing vs random shuffled pairing. Ratio < 1 means the metric is compound-discriminative. NAs in shuffled column mean no random-compound pseudo-tuple had ≥3 labs in this polarity.")
```

# cembio block-MS2: why MS2 timing alone can't locate the peak apex

Cembio is the one lab in this study that acquires MS2 with a **targeted
inclusion list** in a temporally-disjoint block from MS1: the MS1 scan
and the MS2 scan live in **separate mzML files** (`D_1_pos_021_*.mzML`
vs `D_1_pos_MS2_021_*.mzML`). The MS2 inclusion list cycles through a
fixed set of precursor m/z and fires MS2 on a schedule, not in
response to MS1 intensity. Without a co-acquired MS1 spectrum at the
moment each MS2 fires, **we cannot tell which (if any) MS2 fired at
the chromatographic apex of the compound**. Every MS2 is just a
"feature" the producer must reason about.

The DDA labs (afekta, hmgu, icl) have the opposite arrangement: MS1 +
MS2 in the same file, MS2 triggered by an MS1-abundance threshold. By
design, each DDA MS2 fires near a real MS1 peak.

Ritalinic acid `[M+H]+` (m/z 220.1335) in mix 2.9 makes the
difference visible. The MS1 chromatogram for the precursor m/z has
**one chromatographic peak at ~ 198 s** (cembio's MS1 reaches 1.7 × 10⁷
counts there). Cembio's MS2 inclusion list fired **63 MS2 spectra** at
that precursor between RT 142 s and 258 s — spaced ~ 4–5 s apart on
schedule, **with most of them firing on empty chromatogram**. Hmgu's
DDA, by contrast, fired only 2 MS2 triggers at the precursor in that
same RT range, both near actual MS1 peaks.

```{r cembio-block-ms2-figure}
#| fig-cap: "Ritalinic acid [M+H]+ in mix 2.9. **cembio (top)**: blue MS1 chromatogram from `D_1_pos_021_*.mzML` shows the peak at ~198 s; red triangles are MS2 trigger times from `D_1_pos_MS2_021_*.mzML`, fanning out 142-258 s on the fixed inclusion-list schedule. Most MS2 fire on empty chromatogram. **hmgu (bottom)**: DDA fires MS2 only at MS1 peaks — note the ~10⁵× lower y-axis scale (hmgu's per-compound MS1 sensitivity for this analyte is much lower than cembio's, but DDA targeting is precise)."
#| fig-height: 6
suppressPackageStartupMessages(library(mzR))

target_mz <- 220.1335   # ritalinic acid [M+H]+
ppm_tol   <- 20
mz_lo     <- target_mz * (1 - ppm_tol / 1e6)
mz_hi     <- target_mz * (1 + ppm_tol / 1e6)
mz_lo_ms2 <- target_mz * (1 - 100 / 1e6)
mz_hi_ms2 <- target_mz * (1 + 100 / 1e6)
zoom_rt   <- c(150, 290)

# Helper: MS1 EIC summed over a tight m/z window
extract_ms1_eic <- function(path) {
  m <- openMSfile(path, backend = "pwiz"); on.exit(close(m))
  h <- header(m)
  idx <- which(h$msLevel == 1L)
  if (!length(idx)) return(NULL)
  data.frame(
    rt        = h$retentionTime[idx],
    intensity = vapply(idx, function(j) {
      sp <- peaks(m, j)
      if (!nrow(sp)) return(0)
      keep <- sp[, 1L] >= mz_lo & sp[, 1L] <= mz_hi
      if (any(keep)) sum(sp[keep, 2L]) else 0
    }, numeric(1)))
}

# Helper: MS2 trigger times for one precursor m/z
extract_ms2_triggers <- function(path) {
  m <- openMSfile(path, backend = "pwiz"); on.exit(close(m))
  h <- header(m)
  ms2_idx <- which(h$msLevel == 2L)
  if (!length(ms2_idx)) return(data.frame(rt = numeric(0)))
  hit <- ms2_idx[h$precursorMZ[ms2_idx] >= mz_lo_ms2 &
                  h$precursorMZ[ms2_idx] <= mz_hi_ms2]
  data.frame(rt = h$retentionTime[hit])
}

# Lab × file mapping
mzml <- here::here("1_preprocessing")
files <- list(
  cembio = list(
    panel = "cembio (block-MS2: MS1 in file A, MS2 in file B)",
    ms1   = file.path(mzml, "cembio/HE/mzml/D_1_pos_021_Human_Endosome_2.9.mzML"),
    ms2   = file.path(mzml, "cembio/HE/mzml/D_1_pos_MS2_021_Human_Endosome_2.9.mzML")),
  hmgu = list(
    panel = "hmgu (DDA: MS1 + MS2 in one file)",
    ms1   = local({
      hits <- list.files(file.path(mzml, "hmgu/HE/mzml"),
                          pattern = "pos.*2[.]9", full.names = TRUE)
      hits[1]
    })))
files$hmgu$ms2 <- files$hmgu$ms1   # same file holds both for DDA

ms1_all <- bind_rows(lapply(names(files), function(lab) {
  d <- extract_ms1_eic(files[[lab]]$ms1)
  if (is.null(d) || !nrow(d)) return(NULL)
  d$panel <- files[[lab]]$panel; d
}))
ms2_all <- bind_rows(lapply(names(files), function(lab) {
  d <- extract_ms2_triggers(files[[lab]]$ms2)
  if (!nrow(d)) return(NULL)
  d$panel <- files[[lab]]$panel; d
}))

ms1_all <- ms1_all |> filter(rt >= zoom_rt[1], rt <= zoom_rt[2])
ms2_all <- ms2_all |> filter(rt >= zoom_rt[1], rt <= zoom_rt[2])

ggplot() +
  geom_line(data = ms1_all,
            aes(x = rt, y = intensity),
            colour = "#1F77B4", linewidth = 0.6) +
  geom_vline(data = ms2_all,
             aes(xintercept = rt),
             colour = "#D62728", linewidth = 0.4, alpha = 0.7) +
  geom_point(data = ms2_all,
             aes(x = rt, y = 0),
             colour = "#D62728", shape = 17, size = 2.5) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  labs(x = "Retention time (s)",
       y = sprintf("MS1 intensity at m/z %.4f ± 20 ppm", target_mz),
       title = sprintf("Ritalinic acid [M+H]+ (m/z %.4f) in mix 2.9",
                       target_mz)) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

**Consequence at the producer level**: when the producer's legacy
`extract_all_ms2()` ran `rle(precursorMz)` on cembio's MS2 file, every
time another inclusion-list precursor fired between two ritalinic MS2
the run was broken and a new feature was started. The 63 MS2 of
ritalinic in mix 2.9 became 11 separate "features" — one chromatographic
peak shattered into many. SIRIUS then scored each tiny feature
independently rather than getting all 63 spectra as one merged feature
with much better S/N.

**Cluster-based consensus RT in the producer §7.1** (the fix in P10 of
the schema changelog) handles this without re-running SIRIUS:

1. All 63 MS2 spectra share precursor m/z 220.1335 and elute within a
   single 1-D cluster window (gap ≤ 30 s), so they group into **one
   cluster** with median RT ~225 s and a recorded `rt_spread` of
   ~115 s.
2. The picker's `apex_pad` is set to `max(8, rt_spread / 2 + 5)` =
   `max(8, 60)` ≈ 35 s, so the apex search window becomes
   ~[190, 260] s — wide enough to catch the **real apex at 198 s**
   regardless of how scattered the MS2 triggers are.
3. The SNR picker still rejects noise spikes within this wider
   window, so widening doesn't degrade detection elsewhere.

The proper structural fix (replacing `rle()` with RT-proximity
grouping in `extract_all_ms2()`) is in the code but only triggers if
the cembio SIRIUS caches are deleted and SIRIUS re-runs from scratch.
With the existing caches in place, the cluster algorithm + widened
apex_pad fully compensate for the shattering downstream.

# hmgu DDA: why "more MS2 spectra" doesn't mean more spike-ins covered

`hmgu` is the only lab in this study that uses **co-acquired DDA**
(MS1 + data-dependent MS2 in the same file), with a wide-open precursor
m/z range and a single fixed collision energy. The other three labs run
**scheduled or inclusion-list MS2 in a separate file** with stepped CE.
The acquisition profiles diverge sharply.

For mix 2.3 (one of the heaviest hmgu `no_data` mixtures, with 11 neg
and 10 pos `no_data` tuples), here are the MS2 statistics pulled
directly from the mzML headers:

| Lab    | n_MS2 (pos) | unique precursor m/z | CE levels    | RT range covered |
|--------|------------:|---------------------:|:------------:|:----------------:|
| afekta |       4 298 |                  202 | 10 / 20 / 40 | 2.6 - 899.6 s    |
| cembio |         783 |               **23** | 10 / 20 / 40 | 3.2 - 295.7 s    |
| hmgu   |   **8 147** |            **5 238** |       **35** | 4.0 - 725.6 s    |
| icl    |         472 |                  176 | 4 (low CE)   | 1.6 - 726.0 s    |

`hmgu` fires the **most MS2 spectra** (8 147 — almost 2× any other
lab) but across **5 238 distinct precursor m/z** — DDA across the
whole m/z range. The duty cycle gets spread thin. Of those 8 147 MS2
spectra, **only 67 (0.8 %)** fire within ±50 ppm of an expected
spike-in compound m/z (the spike-in m/z list comes from the other
labs' SIRIUS-confirmed annotations for mix 2.3). The remaining
**99.2 %** target matrix / background ions.

In contrast, cembio's 23-precursor inclusion list spends 100 % of its
MS2 duty cycle on known targets; afekta covers 202 distinct precursors
with stepped collision energy, generating richer fragment spectra per
target.

Two consequences for `hmgu`'s spike-in coverage:

1. **MS2 starvation on weak precursors.** When a spike-in compound's
   MS1 intensity is below the matrix peaks competing for the same DDA
   cycle, no MS2 ever triggers on it → no SIRIUS candidate → `no_data`
   downstream. This is the dominant cause of the 167 hmgu `no_data`
   tuples (74 neg + 116 pos, before adduct de-duplication; 162 of
   those have a consensus RT from other labs, confirming the compound
   is *there* in the chromatographic run — just too dim to win MS2).
2. **Single CE = sparser fragmentation.** Stepped CE (10/20/40)
   generates 3 complementary spectra per precursor, helping SIRIUS
   score formula candidates. Single CE = 35 eV gives one spectrum,
   typically lower-quality.

```{r hmgu-dda-ms2-stats}
#| label: hmgu-dda-ms2-stats
#| fig-cap: "MS2 acquisition profile per lab for mix 2.3 (pos). Each panel plots one (lab) file's MS2 trigger times along the chromatographic run. Each red triangle = one MS2 trigger; y-jitter is precursor m/z rank so bands are visible. hmgu's panel is a wall of triggers across the entire m/z range; cembio's is 23 horizontal bands (inclusion list). The relevant question for spike-in coverage is not 'how many MS2 fired' but 'how many fired on a spike-in precursor m/z'."
#| fig-height: 7
#| warning: false
suppressPackageStartupMessages(library(mzR))

# Find each lab's MS2-bearing file for mix 2.3 pos.
find_ms2_file <- function(lab, polarity = "pos", mix = "2.3") {
  d  <- here::here("1_preprocessing", lab, "HE", "mzml")
  fs <- list.files(d, pattern = "2[._]3", full.names = TRUE,
                   ignore.case = TRUE)
  fs <- fs[!grepl("NAPS|blank", fs, ignore.case = TRUE)]
  pol_re <- if (polarity == "pos") "_pos" else "_neg"
  fs <- fs[grepl(pol_re, fs, ignore.case = TRUE)]
  if (!length(fs)) return(NA_character_)
  ms2_file <- fs[grepl("MS2", fs, ignore.case = FALSE)]
  if (length(ms2_file)) ms2_file[1] else fs[1]
}

# Spike-in m/z list (pos, mix 2.3) — pulled from identity table.
expected_pos_23 <- identity_df |>
  filter(mixture == "2.3", polarity == "pos",
         lab != "hmgu", !is.na(mz_observed)) |>
  distinct(compound, mz_observed) |>
  pull(mz_observed)

ms2_per_lab <- bind_rows(lapply(c("afekta", "cembio", "hmgu", "icl"),
                                function(lab) {
  f <- find_ms2_file(lab)
  if (is.na(f)) return(NULL)
  m <- openMSfile(f, backend = "pwiz"); on.exit(close(m))
  h <- header(m)
  ms2 <- which(h$msLevel == 2L)
  if (!length(ms2)) return(NULL)
  is_spike <- vapply(h$precursorMZ[ms2],
                     \(q) any(abs(q - expected_pos_23) /
                              expected_pos_23 * 1e6 <= 50),
                     logical(1))
  tibble(lab = lab,
         rt          = h$retentionTime[ms2],
         precursorMZ = h$precursorMZ[ms2],
         on_spike    = is_spike)
}))

ggplot(ms2_per_lab,
       aes(x = rt, y = precursorMZ, colour = on_spike)) +
  geom_point(size = 0.3, alpha = 0.5) +
  facet_wrap(~ lab, ncol = 2, scales = "free_y") +
  scale_colour_manual(values = c(`FALSE` = "grey70",
                                  `TRUE`  = "#D62728"),
                      labels = c(`FALSE` = "matrix / background",
                                  `TRUE`  = "spike-in target (±50 ppm)"),
                      name = NULL) +
  labs(x = "Retention time (s)",
       y = "Precursor m/z",
       title = "MS2 precursor selection per lab — mix 2.3 pos",
       subtitle = "Each point = one MS2 trigger. Red = fires on a known spike-in m/z.") +
  theme_bw() +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
```

**Implication for the manuscript.** The hmgu vs other-3 pairwise ICC
gap (0.12 - 0.36 areaUnderTic vs ≥ 0.79 within the
afekta + cembio + icl trio) reflects acquisition policy as much as
instrument sensitivity. Open-bore DDA at a single CE on a complex
mixture is the wrong tool for **targeted spike-in panel comparison**;
it would likely outperform inclusion-list approaches for untargeted
biomarker discovery on unknown matrices. The "lower hmgu sensitivity"
finding should be framed as *acquisition-strategy-dependent*, not as
an inherent property of the Sciex Zeno-TOF.

# MS2 + MS1: presence / absence and rescue

Four mutually exclusive states per `(mixture, lab, compound)`:

| Status          | Meaning                                                                |
|-----------------|------------------------------------------------------------------------|
| `ms2_confirmed` | identity row in this lab **and** a `source = "own"` peak               |
| `ms2_no_peak`   | identity row in this lab but no `own` peak                             |
| `ms1_only`      | no identity row but a `source = "consensus"` peak (MS1-rescued)        |
| `absent`        | neither                                                                |

```{r presence-build}
own_peaks <- chrom_df |>
  filter(source == "own") |>
  distinct(mixture, lab, compound) |>
  mutate(has_own_peak = TRUE)

consensus_peaks <- chrom_df |>
  filter(source == "consensus") |>
  distinct(mixture, lab, compound) |>
  mutate(has_consensus_peak = TRUE)

identity_keys <- identity_df |>
  distinct(mixture, lab, compound) |>
  mutate(has_identity = TRUE)

presence_df <- expected_grid |>
  left_join(identity_keys,    by = c("mixture", "lab", "compound")) |>
  left_join(own_peaks,        by = c("mixture", "lab", "compound")) |>
  left_join(consensus_peaks,  by = c("mixture", "lab", "compound")) |>
  mutate(across(c(has_identity, has_own_peak, has_consensus_peak),
                \(x) coalesce(x, FALSE)),
         status = case_when(
           has_identity &  has_own_peak       ~ "ms2_confirmed",
           has_identity & !has_own_peak       ~ "ms2_no_peak",
          !has_identity &  has_consensus_peak ~ "ms1_only",
           TRUE                                ~ "absent"))

presence_df |>
  count(mixture, lab, status) |>
  tidyr::pivot_wider(names_from = status, values_from = n, values_fill = 0L) |>
  knitr::kable(caption = "Detection status counts per (mixture, lab)")
```

```{r presence-heatmap}
#| fig-height: 60
#| fig-cap: "Detection status per (mixture, lab, compound)."
status_pal <- c(ms2_confirmed = "#1f77b4", ms1_only   = "#2ca02c",
                ms2_no_peak   = "#ff7f0e", absent     = "#dddddd")
status_lab <- c(ms2_confirmed = "MS2 annotated + peak",
                ms1_only      = "MS1 only (rescued)",
                ms2_no_peak   = "MS2 annotated, no peak",
                absent        = "Absent")

presence_df |>
  mutate(status = factor(status, levels = names(status_pal))) |>
  ggplot(aes(x = lab, y = compound, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(values = status_pal, labels = status_lab) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Compound detection: MS2 annotation vs MS1 signal") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        axis.text.x = element_text(face = "bold", size = 11),
        strip.text.y = element_text(angle = 0, face = "bold"),
        legend.position = "bottom")
```

```{r rescued-vs-confirmed}
#| fig-height: 60
#| fig-cap: "Labs confirming each compound via MS2 vs rescued via MS1 only."
presence_df |>
  filter(status %in% c("ms2_confirmed", "ms1_only")) |>
  count(mixture, compound, status) |>
  ggplot(aes(x = n, y = compound, fill = status)) +
  geom_col(position = "stack") +
  facet_grid(mixture ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(values = status_pal[c("ms2_confirmed", "ms1_only")],
                    labels = status_lab[c("ms2_confirmed", "ms1_only")]) +
  scale_x_continuous(breaks = 0:length(labs)) +
  labs(x = "Number of labs", y = NULL, fill = NULL,
       title = "Confirmed (MS2) vs rescued (MS1 only) per compound") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9),
        strip.text.y = element_text(angle = 0, face = "bold"),
        legend.position = "bottom")
```
