---
title: "Polarity Comparison Report: Positive vs Negative Ionization Modes"
subtitle: "Ring Trial Lab Effects Analysis - Hardware and Protocol Impacts"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
editor_options:
  chunk_output_type: console
---

# Executive Summary

This report compares lab and hardware effects across **positive and negative ionization modes** to understand:

1. **Are lab differences universal or compound-class specific?**
2. **Which RT regions show consistent patterns across polarities?**
3. **Do protocols affect both modes equally?**
4. **Are standards (references) equally consistent in both modes?**

## Key Findings at a Glance

| Aspect | Positive Mode | Negative Mode | Comparison |
|--------|--------------|---------------|------------|
| Strongest Lab Effect | RT Q4 (R²=0.53) | RT Q4 (check) | Consistent? |
| Protocol Effect (Agilent) | Significant (p=3.8e-6) | Check | Mode-specific? |
| Q3 Robustness | Yes (p=0.169) | Check | Universal? |
| Standards Consistency | ✓ (R²=0.09, NS) | Check | Both modes? |

---

# Setup and Data Loading

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(here)

base_path <- here("5_downstream_analysis")
```

# Load Pre-computed Results

Since the full analyses have been run, we'll load the key summary statistics from both modes.

```{r load-summaries}
# These would be extracted from the rendered analyses
# For now, we'll create them manually based on the reports

# POSITIVE MODE RESULTS
pos_friedman <- data.frame(
  Mode = "Positive",
  Metric = c("BPC/TIC Ratio", "Ratio RT Q1", "Ratio RT Q2", "Ratio RT Q3", "Ratio RT Q4"),
  Chi_squared = c(16.6, 35.0, 25.1, 21.0, 46.1),
  P_value = c(8.6e-4, 1.2e-7, 1.5e-5, 1.1e-4, 5.5e-10)
)

pos_lm_rsq <- data.frame(
  Mode = "Positive",
  Metric = c("BPC", "TIC", "BPC/TIC Ratio", "Ratio RT Q1", "Ratio RT Q2", "Ratio RT Q3", "Ratio RT Q4"),
  R_squared = c(0.943, 0.995, 0.144, 0.237, 0.281, 0.263, 0.530)
)

pos_agilent <- data.frame(
  Mode = "Positive",
  Metric = c("BPC/TIC Ratio", "Ratio RT Q1", "Ratio RT Q2", "Ratio RT Q3", "Ratio RT Q4"),
  P_value = c(3.8e-6, 3.8e-6, 3.8e-6, 0.169, 3.8e-6)
)

pos_annotated <- data.frame(
  Mode = "Positive",
  Metric = "BPC/TIC Ratio",
  R_squared = 0.09,
  P_value = 0.077
)

# NEGATIVE MODE RESULTS
# These will be populated from the negative mode analysis
# For demonstration, using placeholder values - replace with actual results

neg_friedman <- data.frame(
  Mode = "Negative",
  Metric = c("BPC/TIC Ratio", "Ratio RT Q1", "Ratio RT Q2", "Ratio RT Q3", "Ratio RT Q4"),
  Chi_squared = c(38.94, 34.01, 7.36, 45.00, 54.66),
  P_value = c(1.79e-8, 1.97e-7, 0.0613, 9.25e-10, 8.10e-12)
)

neg_lm_rsq <- data.frame(
  Mode = "Negative",
  Metric = c("BPC", "TIC", "BPC/TIC Ratio", "Ratio RT Q1", "Ratio RT Q2", "Ratio RT Q3", "Ratio RT Q4"),
  R_squared = c(0.934, 0.973, 0.636, 0.470, 0.012, 0.771, 0.981)
)

neg_agilent <- data.frame(
  Mode = "Negative",
  Metric = c("BPC/TIC Ratio", "Ratio RT Q1", "Ratio RT Q2", "Ratio RT Q3", "Ratio RT Q4"),
  P_value = c(5.23e-4, 3.8e-6, 0.798, 3.8e-6, 3.8e-6)
)

neg_annotated <- data.frame(
  Mode = "Negative",
  Metric = "BPC/TIC Ratio",
  R_squared = 0.082,
  P_value = 0.104
)

# Combine for comparison
friedman_combined <- rbind(pos_friedman, neg_friedman)
lm_combined <- rbind(pos_lm_rsq, neg_lm_rsq)
agilent_combined <- rbind(pos_agilent, neg_agilent)
annotated_combined <- rbind(pos_annotated, neg_annotated)
```

# Comparison 1: Overall Lab Effects (Friedman Tests)

## What This Shows
Friedman test assesses whether all 4 labs differ in signal characteristics while controlling for mixture composition. Higher Chi² and lower p-value indicate stronger lab differentiation.

```{r friedman-comparison-table}
friedman_table <- friedman_combined %>%
  select(Mode, Metric, Chi_squared, P_value) %>%
  mutate(
    Chi_squared = sprintf("%.1f", Chi_squared),
    P_value_fmt = case_when(
      P_value < 0.001 ~ sprintf("%.2e ***", P_value),
      P_value < 0.01 ~ sprintf("%.3f **", P_value),
      P_value < 0.05 ~ sprintf("%.3f *", P_value),
      TRUE ~ sprintf("%.3f", P_value)
    )
  ) %>%
  select(-P_value) %>%
  rename(`P-value` = P_value_fmt, `χ²` = Chi_squared)

kable(friedman_table, caption = "Friedman Test Results: Lab Differences by Polarity and RT Quartile")
```

## Visualization: Lab Effect Strength by Polarity

```{r friedman-plot, fig.width=10, fig.height=6}
friedman_plot_data <- friedman_combined %>%
  filter(!is.na(P_value))

if (nrow(friedman_plot_data) > 0) {
  ggplot(friedman_plot_data, aes(x = Metric, y = Chi_squared, fill = Mode)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_text(aes(label = sprintf("χ²=%.1f", Chi_squared)), 
              position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
    scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "darkorange")) +
    theme_bw() +
    labs(
      title = "Lab Differentiation Strength: Positive vs Negative Mode",
      subtitle = "Friedman Chi-squared values (higher = stronger lab differences)",
      x = "Metric",
      y = "Chi-squared statistic",
      fill = "Ionization Mode"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
} else {
  plot.new()
  text(0.5, 0.5, "Awaiting negative mode data", cex = 1.5)
}
```

## Interpretation: Lab Effects Across Polarities

**Expected Patterns:**

- **If Chi² values are similar**: Lab effects are universal across compound types
- **If positive > negative**: Basic compounds more sensitive to lab protocols
- **If negative > positive**: Acidic compounds more sensitive to lab protocols
- **RT Q4 consistently highest**: Late-eluting (lipophilic) compounds most affected

---

# Comparison 2: Linear Model R² Values (Variance Explained)

## What This Shows
R² indicates the proportion of variance explained by lab and hardware setup. Higher R² means lab choice matters more; lower R² means within-lab variation dominates.

```{r lm-comparison-table}
lm_table <- lm_combined %>%
  mutate(
    R_squared_fmt = sprintf("%.3f", R_squared),
    Percent = sprintf("%.1f%%", R_squared * 100)
  ) %>%
  select(Mode, Metric, R_squared_fmt, Percent) %>%
  rename(`R²` = R_squared_fmt, `Variance Explained` = Percent)

kable(lm_table, caption = "Linear Model R²: Lab/Setup Explains This Much Variation")
```

## Visualization: Variance Explained by Lab Choice

```{r lm-rsq-plot, fig.width=10, fig.height=6}
lm_plot_data <- lm_combined %>%
  filter(!is.na(R_squared))

if (nrow(lm_plot_data) > 0) {
  ggplot(lm_plot_data, aes(x = reorder(Metric, R_squared), y = R_squared, fill = Mode)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.2f", R_squared)), 
              position = position_dodge(width = 0.9), hjust = -0.1, size = 3) +
    coord_flip() +
    scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "darkorange")) +
    ylim(0, 1.1) +
    theme_bw() +
    labs(
      title = "Lab/Setup Variance Explained: Positive vs Negative Mode",
      subtitle = "R² from linear models (0 = no lab effect, 1 = all variance from labs)",
      x = "Metric",
      y = "R² (proportion of variance explained)",
      fill = "Ionization Mode"
    )
} else {
  plot.new()
  text(0.5, 0.5, "Awaiting negative mode data", cex = 1.5)
}
```

## Key Insights: R² Patterns

### Expected Observations

1. **TIC and BPC**: Should be very high (>0.9) in both modes - instrument-driven
2. **RT Q4**: Should show strongest lab effect in both modes
3. **BPC/TIC Ratio**: Moderate effect, represents peak complexity
4. **Comparison**: If one mode shows consistently higher R², that compound class is more lab-sensitive

---

# Comparison 3: Agilent Protocol Effect (Hardware Constant)

## What This Shows
Afekta and Cembio both use Agilent 1290 LC but have different MS tuning protocols. This comparison isolates **protocol effects** holding hardware constant. Lower p-value means protocol matters.

```{r agilent-comparison-table}
agilent_table <- agilent_combined %>%
  mutate(
    P_value_fmt = case_when(
      P_value < 0.001 ~ sprintf("%.2e ***", P_value),
      P_value < 0.01 ~ sprintf("%.3f **", P_value),
      P_value < 0.05 ~ sprintf("%.3f *", P_value),
      TRUE ~ sprintf("%.3f NS", P_value)
    ),
    Significant = ifelse(P_value < 0.05, "Yes", "No")
  ) %>%
  select(Mode, Metric, P_value_fmt, Significant) %>%
  rename(`P-value` = P_value_fmt)

kable(agilent_table, caption = "Agilent Protocol Comparison (Afekta vs Cembio): Identical Hardware, Different Tuning")
```

## Visualization: Protocol Effect Significance

```{r agilent-plot, fig.width=10, fig.height=6}
agilent_plot_data <- agilent_combined %>%
  filter(!is.na(P_value))

if (nrow(agilent_plot_data) > 0) {
  ggplot(agilent_plot_data, aes(x = Metric, y = -log10(P_value), fill = Mode)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red", linewidth = 1) +
    geom_text(aes(label = ifelse(P_value < 0.05, "Sig", "NS")), 
              position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
    scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "darkorange")) +
    annotate("text", x = 0.5, y = -log10(0.05), label = "p = 0.05 threshold", 
             hjust = -0.1, vjust = -0.5, color = "red", size = 3.5) +
    theme_bw() +
    labs(
      title = "Agilent Protocol Effect: Positive vs Negative Mode",
      subtitle = "Same hardware (Agilent 1290), different MS tuning protocols",
      x = "Metric",
      y = "-log10(P-value) [higher = stronger protocol difference]",
      fill = "Ionization Mode"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
} else {
  plot.new()
  text(0.5, 0.5, "Awaiting negative mode data", cex = 1.5)
}
```

## Critical Finding: Q3 Robustness

**In Positive Mode**: RT Q3 shows NO significant protocol difference (p = 0.169)

**Question for Negative Mode**: Does Q3 remain protocol-robust in negative ionization?

- **If YES**: Q3 contains universally robust compounds (independent of ionization chemistry)
- **If NO**: Q3 robustness is compound-class specific (only for positive mode compounds)

---

# Comparison 4: Standards Validation (Annotated Data)

## What This Shows
Standards are reference compounds with known identity. If labs measure standards consistently despite protocol differences, it validates that unknowns' variation is fixable.

```{r annotated-table}
annotated_table <- annotated_combined %>%
  mutate(
    R_squared_fmt = sprintf("%.3f", R_squared),
    P_value_fmt = ifelse(P_value < 0.05, 
                         sprintf("%.3f *", P_value),
                         sprintf("%.3f NS", P_value)),
    Quality = ifelse(P_value >= 0.05, "✓ Consistent", "⚠ Variable")
  ) %>%
  select(Mode, Metric, R_squared_fmt, P_value_fmt, Quality) %>%
  rename(`R²` = R_squared_fmt, `P-value` = P_value_fmt, `Quality Assessment` = Quality)

kable(annotated_table, caption = "Standards Consistency: Lab Effects on Annotated Reference Compounds")
```

## Interpretation: Quality Assurance by Mode

**Positive Mode**: R² = 0.09, p = 0.077 (NOT significant) ✓✓
→ Standards measured consistently across labs

**Negative Mode**: *To be confirmed*

### If Both Modes Show Weak Effects
✅ Ring trial quality assurance is successful  
✅ Reference compounds validate protocols  
✅ Unknown compound variation is protocol-driven (fixable)

---

# Comparison 5: RT Quartile Pattern Analysis

## What This Reveals
Does the RT-resolved pattern (Q4 > Q3 > Q2 > Q1) hold across both polarities?

```{r rt-pattern-comparison, fig.width=12, fig.height=6}
rt_comparison <- friedman_combined %>%
  filter(grepl("Ratio RT Q", Metric)) %>%
  mutate(
    Quartile = gsub("Ratio RT ", "", Metric),
    Quartile = factor(Quartile, levels = c("Q1", "Q2", "Q3", "Q4"))
  )

if (nrow(rt_comparison %>% filter(!is.na(Chi_squared))) > 0) {
  ggplot(rt_comparison %>% filter(!is.na(Chi_squared)), 
         aes(x = Quartile, y = Chi_squared, color = Mode, group = Mode)) +
    geom_line(linewidth = 1.5, alpha = 0.8) +
    geom_point(size = 4, alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f", Chi_squared)), vjust = -1, size = 3.5) +
    scale_color_manual(values = c("Positive" = "steelblue", "Negative" = "darkorange")) +
    theme_bw() +
    labs(
      title = "RT Quartile Pattern: Lab Effect Strength Across Chromatographic Separation",
      subtitle = "Does late-elution sensitivity hold in both ionization modes?",
      x = "RT Quartile (Q1=0-25%, Q2=25-50%, Q3=50-75%, Q4=75-100%)",
      y = "Friedman Chi-squared (lab differentiation strength)",
      color = "Ionization Mode"
    ) +
    theme(legend.position = "top")
} else {
  plot.new()
  text(0.5, 0.5, "Awaiting negative mode data", cex = 1.5)
}
```

## Expected Hypotheses

### Hypothesis 1: Universal Pattern (Q4 > Q1 in both modes)
→ Late-eluting compounds are inherently more sensitive to instrument tuning  
→ **Mechanism**: Lipophilic compounds have variable ionization efficiency  
→ **Action**: Harmonize Q4 tuning in BOTH modes

### Hypothesis 2: Mode-Specific Patterns
→ Positive Q4 > Q1, but Negative Q1 > Q4  
→ **Mechanism**: Different compound classes peak at different RT regions  
→ **Action**: Mode-specific harmonization strategies

---

# Comparison 6: Data Processing Progression

## Full → Detected → Annotated Pattern

```{r processing-progression}
# Create a comparison of R² across processing levels
processing_comparison <- data.frame(
  Mode = rep(c("Positive", "Negative"), each = 3),
  Level = rep(c("Full", "Detected", "Annotated"), 2),
  BPC_TIC_Rsq = c(0.144, 0.210, 0.09,  # Positive
                  NA, NA, NA)           # Negative - to be filled
)

processing_comparison <- processing_comparison %>%
  mutate(Level = factor(Level, levels = c("Full", "Detected", "Annotated")))

if (sum(!is.na(processing_comparison$BPC_TIC_Rsq)) > 3) {
  ggplot(processing_comparison %>% filter(!is.na(BPC_TIC_Rsq)), 
         aes(x = Level, y = BPC_TIC_Rsq, color = Mode, group = Mode)) +
    geom_line(linewidth = 1.5, alpha = 0.8) +
    geom_point(size = 4, alpha = 0.8) +
    geom_text(aes(label = sprintf("%.2f", BPC_TIC_Rsq)), vjust = -1, size = 3.5) +
    scale_color_manual(values = c("Positive" = "steelblue", "Negative" = "darkorange")) +
    theme_bw() +
    labs(
      title = "Data Processing Effect: BPC/TIC Ratio R² Across Processing Levels",
      subtitle = "Lab effects decrease as data is refined (Full → Detected → Annotated)",
      x = "Processing Level",
      y = "R² (lab variance explained)",
      color = "Ionization Mode"
    ) +
    ylim(0, 0.4)
} else {
  cat("Awaiting negative mode data for processing progression comparison\n")
}
```

## Interpretation: What This Pattern Means

**Expected Pattern**: Full > Detected > Annotated

- **Full**: Includes noise → highest lab variation
- **Detected**: True peaks only → moderate lab variation
- **Annotated**: Standards only → lowest lab variation

**If this holds in both modes**: Lab differences are real instrumental effects, not random noise

---

# Summary and Recommendations

## Key Questions Answered

### 1. Are lab differences universal or compound-class specific?

**Positive Mode**: Strong lab effects (Friedman p < 0.001 across all quartiles)  
**Negative Mode**: *Check after data extraction*

**Expected Answer**: If both show strong effects, lab differences are **universal instrument performance issues**, not chemistry-specific.

### 2. Which RT regions show consistent patterns?

**Positive Mode**: Q4 (late-elution) shows **2.2× stronger** effect than Q1  
**Negative Mode**: *Check Q4 vs Q1 ratio*

**Expected Answer**: If Q4 > Q1 in both, focus harmonization on **late-eluting, lipophilic compounds** regardless of charge state.

### 3. Do protocols affect both modes equally?

**Positive Mode**: Significant Agilent protocol difference (p = 3.8e-6), except Q3 (p = 0.169)  
**Negative Mode**: *Check protocol p-values and Q3 status*

**Expected Answer**: If Q3 remains robust in both modes, these compounds are **naturally protocol-tolerant** and should be studied.

### 4. Are standards equally consistent?

**Positive Mode**: Yes (R² = 0.09, p = 0.077 NS) ✓  
**Negative Mode**: *Check annotated R² and p-value*

**Expected Answer**: If both show weak effects, **quality assurance is validated** across both ionization chemistries.

---

## Actionable Recommendations (Pending Full Data)

### If Patterns are SIMILAR Across Modes

1. **Single Harmonization Strategy** can address both polarities
2. **Focus on RT Q4** (late-eluting compounds) in BOTH modes
3. **Protocol standardization** is priority #1 (not hardware replacement)
4. **Q3 compounds** should be studied as naturally robust references

### If Patterns are DIFFERENT Across Modes

1. **Mode-specific harmonization** required
2. **Compound class analysis** needed to understand which types are problematic
3. **Different tuning protocols** may be optimal for positive vs negative
4. **Prioritize the mode with stronger effects** for immediate action

---

## Next Steps

1. **Extract negative mode summary statistics** from rendered HTML report
2. **Populate all NA values** in this comparison document
3. **Re-render** to generate complete comparison plots
4. **Identify mode-specific vs universal patterns**
5. **Draft targeted harmonization protocols** based on findings

---

## Appendix: Data Extraction Notes

To populate negative mode results, extract from `2_global_signal_analysis_neg_simplified.html`:

- **Friedman Summary Table**: Chi² and p-values for BPC/TIC and RT quartiles
- **Linear Model Summary**: R² values for all metrics
- **Agilent Comparison**: p-values for protocol effect
- **Annotated Summary**: R² and p-value for standards

**Extraction Method**: Open HTML → Search for "Friedman Test Summary" → Copy values → Paste into R code above

---

*Report generated: `r Sys.Date()`*  
*Analysis files: `2_global_signal_analysis_pos_simplified.qmd` and `2_global_signal_analysis_neg_simplified.qmd`*
